### 2024-12-03 paper.tex
Add initial template and makefile

%%
%% This is file `sample-sigconf-authordraft.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,authordraft')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf-authordraft.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[sigconf,authordraft]{acmart}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmlicensed}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}
%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation emai}{June 03--05,
  2018}{Woodstock, NY}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmISBN{978-1-4503-XXXX-X/18/06}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{The Name of the Title Is Hope}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Ben Trovato}
\authornote{Both authors contributed equally to this research.}
\email{trovato@corporation.com}
\orcid{1234-5678-9012}
\author{G.K.M. Tobin}
\authornotemark[1]
\email{webmaster@marysville-ohio.com}
\affiliation{%
  \institution{Institute for Clarity in Documentation}
  \city{Dublin}
  \state{Ohio}
  \country{USA}
}

\author{Lars Th{\o}rv{\"a}ld}
\affiliation{%
  \institution{The Th{\o}rv{\"a}ld Group}
  \city{Hekla}
  \country{Iceland}}
\email{larst@affiliation.org}

\author{Valerie B\'eranger}
\affiliation{%
  \institution{Inria Paris-Rocquencourt}
  \city{Rocquencourt}
  \country{France}
}

\author{Aparna Patel}
\affiliation{%
 \institution{Rajiv Gandhi University}
 \city{Doimukh}
 \state{Arunachal Pradesh}
 \country{India}}

\author{Huifen Chan}
\affiliation{%
  \institution{Tsinghua University}
  \city{Haidian Qu}
  \state{Beijing Shi}
  \country{China}}

\author{Charles Palmer}
\affiliation{%
  \institution{Palmer Research Laboratories}
  \city{San Antonio}
  \state{Texas}
  \country{USA}}
\email{cpalmer@prl.com}

\author{John Smith}
\affiliation{%
  \institution{The Th{\o}rv{\"a}ld Group}
  \city{Hekla}
  \country{Iceland}}
\email{jsmith@affiliation.org}

\author{Julius P. Kumquat}
\affiliation{%
  \institution{The Kumquat Consortium}
  \city{New York}
  \country{USA}}
\email{jpkumquat@consortium.net}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Trovato et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  A clear and well-documented \LaTeX\ document is presented as an
  article formatted for publication by ACM in a conference proceedings
  or journal publication. Based on the ``acmart'' document class, this
  article presents and explains many of the common variations, as well
  as many of the formatting elements an author may use in the
  preparation of the documentation of their work.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>00000000.0000000.0000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Do, Not, Us, This, Code, Put, the, Correct, Terms, for,
  Your, Paper}
%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.
\begin{teaserfigure}
  \includegraphics[width=\textwidth]{sampleteaser}
  \caption{Seattle Mariners at Spring Training, 2010.}
  \Description{Enjoying the baseball game from the third-base
  seats. Ichiro Suzuki preparing to bat.}
  \label{fig:teaser}
\end{teaserfigure}

\received{20 February 2007}
\received[revised]{12 March 2009}
\received[accepted]{5 June 2009}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
ACM's consolidated article template, introduced in 2017, provides a
consistent \LaTeX\ style for use across ACM publications, and
incorporates accessibility and metadata-extraction functionality
necessary for future Digital Library endeavors. Numerous ACM and
SIG-specific \LaTeX\ templates have been examined, and their unique
features incorporated into this single new template.

If you are new to publishing with ACM, this document is a valuable
guide to the process of preparing your work for publication. If you
have published with ACM before, this document provides insight and
instruction into more recent changes to the article template.

The ``\verb|acmart|'' document class can be used to prepare articles
for any ACM publication --- conference or journal, and for any stage
of publication, from review to final ``camera-ready'' copy, to the
author's own version, with {\itshape very} few changes to the source.

\section{Template Overview}
As noted in the introduction, the ``\verb|acmart|'' document class can
be used to prepare many different kinds of documentation --- a
double-anonymous initial submission of a full-length technical paper, a
two-page SIGGRAPH Emerging Technologies abstract, a ``camera-ready''
journal article, a SIGCHI Extended Abstract, and more --- all by
selecting the appropriate {\itshape template style} and {\itshape
  template parameters}.

This document will explain the major features of the document
class. For further information, the {\itshape \LaTeX\ User's Guide} is
available from
\url{https://www.acm.org/publications/proceedings-template}.

\subsection{Template Styles}

The primary parameter given to the ``\verb|acmart|'' document class is
the {\itshape template style} which corresponds to the kind of publication
or SIG publishing the work. This parameter is enclosed in square
brackets and is a part of the {\verb|documentclass|} command:
\begin{verbatim}
  \documentclass[STYLE]{acmart}
\end{verbatim}

Journals use one of three template styles. All but three ACM journals
use the {\verb|acmsmall|} template style:
\begin{itemize}
\item {\texttt{acmsmall}}: The default journal template style.
\item {\texttt{acmlarge}}: Used by JOCCH and TAP.
\item {\texttt{acmtog}}: Used by TOG.
\end{itemize}

The majority of conference proceedings documentation will use the {\verb|acmconf|} template style.
\begin{itemize}
\item {\texttt{sigconf}}: The default proceedings template style.
\item{\texttt{sigchi}}: Used for SIGCHI conference articles.
\item{\texttt{sigplan}}: Used for SIGPLAN conference articles.
\end{itemize}

\subsection{Template Parameters}

In addition to specifying the {\itshape template style} to be used in
formatting your work, there are a number of {\itshape template parameters}
which modify some part of the applied template style. A complete list
of these parameters can be found in the {\itshape \LaTeX\ User's Guide.}

Frequently-used parameters, or combinations of parameters, include:
\begin{itemize}
\item {\texttt{anonymous,review}}: Suitable for a ``double-anonymous''
  conference submission. Anonymizes the work and includes line
  numbers. Use with the \texttt{\acmSubmissionID} command to print the
  submission's unique ID on each page of the work.
\item{\texttt{authorversion}}: Produces a version of the work suitable
  for posting by the author.
\item{\texttt{screen}}: Produces colored hyperlinks.
\end{itemize}

This document uses the following string as the first command in the
source file:
\begin{verbatim}
\documentclass[sigconf,authordraft]{acmart}
\end{verbatim}

\section{Modifications}

Modifying the template --- including but not limited to: adjusting
margins, typeface sizes, line spacing, paragraph and list definitions,
and the use of the \verb|\vspace| command to manually adjust the
vertical spacing between elements of your work --- is not allowed.

{\bfseries Your document will be returned to you for revision if
  modifications are discovered.}

\section{Typefaces}

The ``\verb|acmart|'' document class requires the use of the
``Libertine'' typeface family. Your \TeX\ installation should include
this set of packages. Please do not substitute other typefaces. The
``\verb|lmodern|'' and ``\verb|ltimes|'' packages should not be used,
as they will override the built-in typeface families.

\section{Title Information}

The title of your work should use capital letters appropriately -
\url{https://capitalizemytitle.com/} has useful rules for
capitalization. Use the {\verb|title|} command to define the title of
your work. If your work has a subtitle, define it with the
{\verb|subtitle|} command.  Do not insert line breaks in your title.

If your title is lengthy, you must define a short version to be used
in the page headers, to prevent overlapping text. The \verb|title|
command has a ``short title'' parameter:
\begin{verbatim}
  \title[short title]{full title}
\end{verbatim}

\section{Authors and Affiliations}

Each author must be defined separately for accurate metadata
identification.  As an exception, multiple authors may share one
affiliation. Authors' names should not be abbreviated; use full first
names wherever possible. Include authors' e-mail addresses whenever
possible.

Grouping authors' names or e-mail addresses, or providing an ``e-mail
alias,'' as shown below, is not acceptable:
\begin{verbatim}
  \author{Brooke Aster, David Mehldau}
  \email{dave,judy,steve@university.edu}
  \email{firstname.lastname@phillips.org}
\end{verbatim}

The \verb|authornote| and \verb|authornotemark| commands allow a note
to apply to multiple authors --- for example, if the first two authors
of an article contributed equally to the work.

If your author list is lengthy, you must define a shortened version of
the list of authors to be used in the page headers, to prevent
overlapping text. The following command should be placed just after
the last \verb|\author{}| definition:
\begin{verbatim}
  \renewcommand{\shortauthors}{McCartney, et al.}
\end{verbatim}
Omitting this command will force the use of a concatenated list of all
of the authors' names, which may result in overlapping text in the
page headers.

The article template's documentation, available at
\url{https://www.acm.org/publications/proceedings-template}, has a
complete explanation of these commands and tips for their effective
use.

Note that authors' addresses are mandatory for journal articles.

\section{Rights Information}

Authors of any work published by ACM will need to complete a rights
form. Depending on the kind of work, and the rights management choice
made by the author, this may be copyright transfer, permission,
license, or an OA (open access) agreement.

Regardless of the rights management choice, the author will receive a
copy of the completed rights form once it has been submitted. This
form contains \LaTeX\ commands that must be copied into the source
document. When the document source is compiled, these commands and
their parameters add formatted text to several areas of the final
document:
\begin{itemize}
\item the ``ACM Reference Format'' text on the first page.
\item the ``rights management'' text on the first page.
\item the conference information in the page header(s).
\end{itemize}

Rights information is unique to the work; if you are preparing several
works for an event, make sure to use the correct set of commands with
each of the works.

The ACM Reference Format text is required for all articles over one
page in length, and is optional for one-page articles (abstracts).

\section{CCS Concepts and User-Defined Keywords}

Two elements of the ``acmart'' document class provide powerful
taxonomic tools for you to help readers find your work in an online
search.

The ACM Computing Classification System ---
\url{https://www.acm.org/publications/class-2012} --- is a set of
classifiers and concepts that describe the computing
discipline. Authors can select entries from this classification
system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the
commands to be included in the \LaTeX\ source.

User-defined keywords are a comma-separated list of words and phrases
of the authors' choosing, providing a more flexible way of describing
the research being presented.

CCS concepts and user-defined keywords are required for for all
articles over two pages in length, and are optional for one- and
two-page articles (or abstracts).

\section{Sectioning Commands}

Your work should use standard \LaTeX\ sectioning commands:
\verb|section|, \verb|subsection|, \verb|subsubsection|, and
\verb|paragraph|. They should be numbered; do not remove the numbering
from the commands.

Simulating a sectioning command by setting the first word or words of
a paragraph in boldface or italicized text is {\bfseries not allowed.}

\section{Tables}

The ``\verb|acmart|'' document class includes the ``\verb|booktabs|''
package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
high-quality tables.

Table captions are placed {\itshape above} the table.

Because tables cannot be split across pages, the best placement for
them is typically the top of the page nearest their initial cite.  To
ensure this proper ``floating'' placement of tables, use the
environment \textbf{table} to enclose the table's contents and the
table caption.  The contents of the table itself must go in the
\textbf{tabular} environment, to be aligned properly in rows and
columns, with the desired horizontal and vertical rules.  Again,
detailed instructions on \textbf{tabular} material are found in the
\textit{\LaTeX\ User's Guide}.

Immediately following this sentence is the point at which
Table~\ref{tab:freq} is included in the input file; compare the
placement of the table here with the table in the printed output of
this document.

\begin{table}
  \caption{Frequency of Special Characters}
  \label{tab:freq}
  \begin{tabular}{ccl}
    \toprule
    Non-English or Math&Frequency&Comments\\
    \midrule
    \O & 1 in 1,000& For Swedish names\\
    $\pi$ & 1 in 5& Common in math\\
    \$ & 4 in 5 & Used in business\\
    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
  \bottomrule
\end{tabular}
\end{table}

To set a wider table, which takes up the whole width of the page's
live area, use the environment \textbf{table*} to enclose the table's
contents and the table caption.  As with a single-column table, this
wide table will ``float'' to a location deemed more
desirable. Immediately following this sentence is the point at which
Table~\ref{tab:commands} is included in the input file; again, it is
instructive to compare the placement of the table here with the table
in the printed output of this document.

\begin{table*}
  \caption{Some Typical Commands}
  \label{tab:commands}
  \begin{tabular}{ccl}
    \toprule
    Command &A Number & Comments\\
    \midrule
    \texttt{{\char'134}author} & 100& Author \\
    \texttt{{\char'134}table}& 300 & For tables\\
    \texttt{{\char'134}table*}& 400& For wider tables\\
    \bottomrule
  \end{tabular}
\end{table*}

Always use midrule to separate table header rows from data rows, and
use it only for this purpose. This enables assistive technologies to
recognise table headers and support their users in navigating tables
more easily.

\section{Math Equations}
You may want to display math equations in three distinct styles:
inline, numbered or non-numbered display.  Each of the three are
discussed in the next sections.

\subsection{Inline (In-text) Equations}
A formula that appears in the running text is called an inline or
in-text formula.  It is produced by the \textbf{math} environment,
which can be invoked with the usual
\texttt{{\char'134}begin\,\ldots{\char'134}end} construction or with
the short form \texttt{\$\,\ldots\$}. You can use any of the symbols
and structures, from $\alpha$ to $\omega$, available in
\LaTeX~\cite{Lamport:LaTeX}; this section will simply show a few
examples of in-text equations in context. Notice how this equation:
\begin{math}
  \lim_{n\rightarrow \infty}x=0
\end{math},
set here in in-line math style, looks slightly different when
set in display style.  (See next section).

\subsection{Display Equations}
A numbered display equation---one set off by vertical space from the
text and centered horizontally---is produced by the \textbf{equation}
environment. An unnumbered display equation is produced by the
\textbf{displaymath} environment.

Again, in either environment, you can use any of the symbols and
structures available in \LaTeX\@; this section will just give a couple
of examples of display equations in context.  First, consider the
equation, shown as an inline equation above:
\begin{equation}
  \lim_{n\rightarrow \infty}x=0
\end{equation}
Notice how it is formatted somewhat differently in
the \textbf{displaymath}
environment.  Now, we'll enter an unnumbered equation:
\begin{displaymath}
  \sum_{i=0}^{\infty} x + 1
\end{displaymath}
and follow it with another numbered equation:
\begin{equation}
  \sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f
\end{equation}
just to demonstrate \LaTeX's able handling of numbering.

\section{Figures}

The ``\verb|figure|'' environment should be used for figures. One or
more images can be placed within a figure. If your figure contains
third-party material, you must clearly identify it as such, as shown
in the example below.
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{sample-franklin}
  \caption{1907 Franklin Model D roadster. Photograph by Harris \&
    Ewing, Inc. [Public domain], via Wikimedia
    Commons. (\url{https://goo.gl/VLCRBB}).}
  \Description{A woman and a girl in white dresses sit in an open car.}
\end{figure}

Your figures should contain a caption which describes the figure to
the reader.

Figure captions are placed {\itshape below} the figure.

Every figure should also have a figure description unless it is purely
decorative. These descriptions convey what’s in the image to someone
who cannot see it. They are also used by search engine crawlers for
indexing images, and when images cannot be loaded.

A figure description must be unformatted plain text less than 2000
characters long (including spaces).  {\bfseries Figure descriptions
  should not repeat the figure caption – their purpose is to capture
  important information that is not already provided in the caption or
  the main text of the paper.} For figures that convey important and
complex new information, a short text description may not be
adequate. More complex alternative descriptions can be placed in an
appendix and referenced in a short figure description. For example,
provide a data table capturing the information in a bar chart, or a
structured list representing a graph.  For additional information
regarding how best to write figure descriptions and why doing this is
so important, please see
\url{https://www.acm.org/publications/taps/describing-figures/}.

\subsection{The ``Teaser Figure''}

A ``teaser figure'' is an image, or set of images in one figure, that
are placed after all author and affiliation information, and before
the body of the article, spanning the page. If you wish to have such a
figure in your article, place the command immediately before the
\verb|\maketitle| command:
\begin{verbatim}
  \begin{teaserfigure}
    \includegraphics[width=\textwidth]{sampleteaser}
    \caption{figure caption}
    \Description{figure description}
  \end{teaserfigure}
\end{verbatim}

\section{Citations and Bibliographies}

The use of \BibTeX\ for the preparation and formatting of one's
references is strongly recommended. Authors' names should be complete
--- use full first names (``Donald E. Knuth'') not initials
(``D. E. Knuth'') --- and the salient identifying features of a
reference should be included: title, year, volume, number, pages,
article DOI, etc.

The bibliography is included in your source document with these two
commands, placed just before the \verb|\end{document}| command:
\begin{verbatim}
  \bibliographystyle{ACM-Reference-Format}
  \bibliography{bibfile}
\end{verbatim}
where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
suffix, of the \BibTeX\ file.

Citations and references are numbered by default. A small number of
ACM publications have citations and references formatted in the
``author year'' style; for these exceptions, please include this
command in the {\bfseries preamble} (before the command
``\verb|\begin{document}|'') of your \LaTeX\ source:
\begin{verbatim}
  \citestyle{acmauthoryear}
\end{verbatim}

  Some examples.  A paginated journal article \cite{Abril07}, an
  enumerated journal article \cite{Cohen07}, a reference to an entire
  issue \cite{JCohen96}, a monograph (whole book) \cite{Kosiur01}, a
  monograph/whole book in a series (see 2a in spec. document)
  \cite{Harel79}, a divisible-book such as an anthology or compilation
  \cite{Editor00} followed by the same example, however we only output
  the series if the volume number is given \cite{Editor00a} (so
  Editor00a's series should NOT be present since it has no vol. no.),
  a chapter in a divisible book \cite{Spector90}, a chapter in a
  divisible book in a series \cite{Douglass98}, a multi-volume work as
  book \cite{Knuth97}, a couple of articles in a proceedings (of a
  conference, symposium, workshop for example) (paginated proceedings
  article) \cite{Andler79, Hagerup1993}, a proceedings article with
  all possible elements \cite{Smith10}, an example of an enumerated
  proceedings article \cite{VanGundy07}, an informally published work
  \cite{Harel78}, a couple of preprints \cite{Bornmann2019,
    AnzarootPBM14}, a doctoral dissertation \cite{Clarkson85}, a
  master's thesis: \cite{anisi03}, an online document / world wide web
  resource \cite{Thornburg01, Ablamowicz07, Poker06}, a video game
  (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03} and \cite{Lee05}
  and (Case 3) a patent \cite{JoeScientist001}, work accepted for
  publication \cite{rous08}, 'YYYYb'-test for prolific author
  \cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might
  contain 'duplicate' DOI and URLs (some SIAM articles)
  \cite{Kirschmer:2010:AEI:1958016.1958018}. Boris / Barbara Beeton:
  multi-volume works as books \cite{MR781536} and \cite{MR781537}. A
  couple of citations with DOIs:
  \cite{2004:ITE:1009386.1010128,Kirschmer:2010:AEI:1958016.1958018}. Online
  citations: \cite{TUGInstmem, Thornburg01, CTANacmart}.
  Artifacts: \cite{R} and \cite{UMassCitations}.

\section{Acknowledgments}

Identification of funding sources and other support, and thanks to
individuals and groups that assisted in the research and the
preparation of the work should be included in an acknowledgment
section, which is placed just before the reference section in your
document.

This section has a special environment:
\begin{verbatim}
  \begin{acks}
  ...
  \end{acks}
\end{verbatim}
so that the information contained therein can be more easily collected
during the article metadata extraction phase, and to ensure
consistency in the spelling of the section heading.

Authors should not prepare this section as a numbered or unnumbered {\verb|\section|}; please use the ``{\verb|acks|}'' environment.

\section{Appendices}

If your work needs an appendix, add it before the
``\verb|\end{document}|'' command at the conclusion of your source
document.

Start the appendix with the ``\verb|appendix|'' command:
\begin{verbatim}
  \appendix
\end{verbatim}
and note that in the appendix, sections are lettered, not
numbered. This document has two appendices, demonstrating the section
and subsection identification method.

\section{Multi-language papers}

Papers may be written in languages other than English or include
titles, subtitles, keywords and abstracts in different languages (as a
rule, a paper in a language other than English should include an
English title and an English abstract).  Use \verb|language=...| for
every language used in the paper.  The last language indicated is the
main language of the paper.  For example, a French paper with
additional titles and abstracts in English and German may start with
the following command
\begin{verbatim}
\documentclass[sigconf, language=english, language=german,
               language=french]{acmart}
\end{verbatim}

The title, subtitle, keywords and abstract will be typeset in the main
language of the paper.  The commands \verb|\translatedXXX|, \verb|XXX|
begin title, subtitle and keywords, can be used to set these elements
in the other languages.  The environment \verb|translatedabstract| is
used to set the translation of the abstract.  These commands and
environment have a mandatory first argument: the language of the
second argument.  See \verb|sample-sigconf-i13n.tex| file for examples
of their usage.

\section{SIGCHI Extended Abstracts}

The ``\verb|sigchi-a|'' template style (available only in \LaTeX\ and
not in Word) produces a landscape-orientation formatted article, with
a wide left margin. Three environments are available for use with the
``\verb|sigchi-a|'' template style, and produce formatted output in
the margin:
\begin{description}
\item[\texttt{sidebar}:]  Place formatted text in the margin.
\item[\texttt{marginfigure}:] Place a figure in the margin.
\item[\texttt{margintable}:] Place a table in the margin.
\end{description}

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
To Robert, for the bagels and explaining CMYK and color spaces.
\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}

%%
%% If your work has an appendix, this is the place to put it.
\appendix

\section{Research Methods}

\subsection{Part One}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
malesuada, quam in pulvinar varius, metus nunc fermentum urna, id
sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum
vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et
lacinia dolor. Integer ultricies commodo sem nec semper.

\subsection{Part Two}

Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales
ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non
ipsum convallis congue hendrerit vitae sapien. Donec at laoreet
eros. Vivamus non purus placerat, scelerisque diam eu, cursus
ante. Etiam aliquam tortor auctor efficitur mattis.

\section{Online Resources}

Nam id fermentum dui. Suspendisse sagittis tortor a nulla mollis, in
pulvinar ex pretium. Sed interdum orci quis metus euismod, et sagittis
enim maximus. Vestibulum gravida massa ut felis suscipit
congue. Quisque mattis elit a risus ultrices commodo venenatis eget
dui. Etiam sagittis eleifend elementum.

Nam interdum magna at lectus dignissim, ac dignissim lorem
rhoncus. Maecenas eu arcu ac neque placerat aliquam. Nunc pulvinar
massa et mattis lacinia.

\end{document}
\endinput
%%
%% End of file `sample-sigconf-authordraft.tex'.

### 2024-12-09 paper.tex
Introduction Work

@@
 %%
-%% This is file `sample-sigconf-authordraft.tex',
+%% This is file `sample-sigconf.tex',
 %% generated with the docstrip utility.
@@
 %%
-%% samples.dtx  (with options: `all,proceedings,bibtex,authordraft')
+%% samples.dtx  (with options: `all,proceedings,bibtex,sigconf')
 %% 
@@
 %% Any modified versions of this file must be renamed
-%% with new filenames distinct from sample-sigconf-authordraft.tex.
+%% with new filenames distinct from sample-sigconf.tex.
 %% 
@@
 %% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-%%
-%%
-\documentclass[sigconf,authordraft]{acmart}
+
+\documentclass[sigconf]{acmart}
 %%
@@
 
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
 \setcopyright{acmlicensed}
@@
 \acmDOI{XXXXXXX.XXXXXXX}
-%% These commands are for a PROCEEDINGS abstract or paper.
 \acmConference[Conference acronym 'XX]{Make sure to enter the correct
@@
   2018}{Woodstock, NY}
-%%
-%%  Uncomment \acmBooktitle if the title of the proceedings is different
-%%  from ``Proceedings of ...''!
-%%
-%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-%%  June 03--05, 2018, Woodstock, NY}
 \acmISBN{978-1-4503-XXXX-X/18/06}
 
-%%
-%% Submission ID.
-%% Use this when submitting an article to a sponsored event. You'll
-%% receive a unique submission ID from the organizers
-%% of the event, and this ID should be used as the parameter to this command.
-%%\acmSubmissionID{123-A56-BU3}
-
-%%
-%% For managing citations, it is recommended to use bibliography
-%% files in BibTeX format.
-%%
-%% You can then either use BibTeX with the ACM-Reference-Format style,
-%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
-%% support for advanced citation of software artefact from the
-%% biblatex-software package, also separately available on CTAN.
-%%
-%% Look at the sample-*-biblatex.tex files for templates showcasing
-%% the biblatex styles.
-%%
-
-%%
-%% The majority of ACM publications use numbered citations and
-%% references.  The command \citestyle{authoryear} switches to the
-%% "author year" style.
-%%
-%% If you are preparing content for an event
-%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
-%% citations and references.
-%% Uncommenting
-%% the next command will enable that style.
-%%\citestyle{acmauthoryear}
-
-%%
-%% end of the preamble, start of the body of the document source.
 \begin{document}
 
-%%
-%% The "title" command has an optional parameter,
-%% allowing the author to define a "short title" to be used in page headers.
 \title{The Name of the Title Is Hope}
-
-%%
-%% The "author" command and its associated commands are used to define
-%% the authors and their affiliations.
-%% Of note is the shared affiliation of the first two authors, and the
-%% "authornote" and "authornotemark" commands
-%% used to denote shared contribution to the research.
-\author{Ben Trovato}
-\authornote{Both authors contributed equally to this research.}
-\email{trovato@corporation.com}
-\orcid{1234-5678-9012}
-\author{G.K.M. Tobin}
-\authornotemark[1]
-\email{webmaster@marysville-ohio.com}
-\affiliation{%
-  \institution{Institute for Clarity in Documentation}
-  \city{Dublin}
-  \state{Ohio}
-  \country{USA}
-}
 
@@
 
-\author{Huifen Chan}
-\affiliation{%
-  \institution{Tsinghua University}
-  \city{Haidian Qu}
-  \state{Beijing Shi}
-  \country{China}}
-
-\author{Charles Palmer}
-\affiliation{%
-  \institution{Palmer Research Laboratories}
-  \city{San Antonio}
-  \state{Texas}
-  \country{USA}}
-\email{cpalmer@prl.com}
-
-\author{John Smith}
-\affiliation{%
-  \institution{The Th{\o}rv{\"a}ld Group}
-  \city{Hekla}
-  \country{Iceland}}
-\email{jsmith@affiliation.org}
-
-\author{Julius P. Kumquat}
-\affiliation{%
-  \institution{The Kumquat Consortium}
-  \city{New York}
-  \country{USA}}
-\email{jpkumquat@consortium.net}
-
-%%
-%% By default, the full list of authors will be used in the page
-%% headers. Often, this list is too long, and will overlap
-%% other information printed in the page headers. This command allows
-%% the author to define a more concise list
-%% of authors' names for this purpose.
 \renewcommand{\shortauthors}{Trovato et al.}
 
-%%
-%% The abstract is a short summary of the work to be presented in the
-%% article.
 \begin{abstract}
-  A clear and well-documented \LaTeX\ document is presented as an
-  article formatted for publication by ACM in a conference proceedings
-  or journal publication. Based on the ``acmart'' document class, this
-  article presents and explains many of the common variations, as well
-  as many of the formatting elements an author may use in the
-  preparation of the documentation of their work.
+As the compute capabilities of major graphics APIs have matured, there has been growing utilization of compute shaders for graphical tasks, and as such, there has been corresponding interest in porting over primitives like prefix scan from compute languages to shading languages.
+However, the current state-of-the-art prefix scan, \textit{Chained Scan with Decoupled Lookback}, relies on properties provided by the NVIDIA ecosystem that, once removed, force developers to make trade-offs between maintainability, portability, and performance.
+We describe \textit{DecoupledFallback}, a fully portable, single-pass prefix scan method capable of reaching speed of light scan performance across different hardwares with minimal developer intervention.
+Our implementation is built in WGSL and is compatible with any implementation of the WebGPU standard (I think?
+It definitely does not work on WARP .
+.
+.
+which is concerning).
 \end{abstract}
 
-%%
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-%% Please copy and paste the code instead of the example below.
-%%
 \begin{CCSXML}
@@
 
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
 \keywords{Do, Not, Us, This, Code, Put, the, Correct, Terms, for,
   Your, Paper}
-%% A "teaser" image appears between the author and affiliation
-%% information and the body of the document, and typically spans the
-%% page.
-\begin{teaserfigure}
-  \includegraphics[width=\textwidth]{sampleteaser}
-  \caption{Seattle Mariners at Spring Training, 2010.}
-  \Description{Enjoying the baseball game from the third-base
-  seats. Ichiro Suzuki preparing to bat.}
-  \label{fig:teaser}
-\end{teaserfigure}
 
@@
 
-%%
-%% This command processes the author and affiliation and title
-%% information and builds the first part of the formatted document.
 \maketitle
@@
 \section{Introduction}
-ACM's consolidated article template, introduced in 2017, provides a
-consistent \LaTeX\ style for use across ACM publications, and
-incorporates accessibility and metadata-extraction functionality
-necessary for future Digital Library endeavors. Numerous ACM and
-SIG-specific \LaTeX\ templates have been examined, and their unique
-features incorporated into this single new template.
+Prefix scan is one of the most fundamental primitives in parallel computing, with uses including compaction, sorting \cite{10.1145/3035918.3064043, adinets2022onesweepfastersignificantdigit}, SPMV \cite{} etc..
+Sometimes referred to as \textit{scan} or \textit{prefix reduction}, prefix scan is typically defined on a monoid, though it can also be defined on a semigroup in certain cases\cite{10.1007/978-3-540-27764-4_11}.
+In a prefix scan, the result at element $n$ is the reduction of the preceding subset of elements in the sequence.
+If the reduction subset includes the $n$-th element, it is called \textit{inclusive}; if it excludes the $n$-th element, it is called \textit{exclusive}.
+The binary operator used in the scan must be associative, but it need not be commutative—for instance, in the case of the stack monoid.
+The most common type of prefix scan is prefix sum, where the binary operator is addition.
+For example\footnote{In this paper we will use the terminology of WGSL, but we note that subgroup is interchangeable with warp(CUDA), wave(HLSL), and simd\_group(Metal), while workgroup is interchangeable with block(CUDA), group(HLSL), and ???(Metal)}:
 
-If you are new to publishing with ACM, this document is a valuable
-guide to the process of preparing your work for publication. If you
-have published with ACM before, this document provides insight and
-instruction into more recent changes to the article template.
+(PREFIX SUM EXAMPLE WITH BINARY REDUCTION OPERATOR)
 
-The ``\verb|acmart|'' document class can be used to prepare articles
-for any ACM publication --- conference or journal, and for any stage
-of publication, from review to final ``camera-ready'' copy, to the
-author's own version, with {\itshape very} few changes to the source.
+Contemporary GPU's are characterized by their high memory bandwidth, high arithmetic throughput, high memory latency, and hierarchical memory model.
+With proper latency-hiding, prefix scan is computationally light enough that it is memory-bandwidth bound, and thus the focus of contemporary scan strategies has been global communication avoidance.
+The current state-of-the-art prefix scan is \textit{Chained Scan with Decoupled Lookback}\cite{Merrill2016} (referred to hereon as \textit{DecoupledLookback}).
+Although a chained scan\footnote{A chained scan is a scan performed serially.} may seem inimical to parallelism, the key innovation of GPU chained scanning lies in its hybridization of parallel and serial scan strategies at different levels of the GPU memory hierarchy, achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level.
+\textit{DecoupledLookback} leverages this hybrid strategy while also being capable of fully saturating global memory bandwidth, and as a result, with approximately \textasciitilde$2n$ global data movement—one read and one write per processed element—it matches the performance of a copy operation, effectively achieving “speed of light” efficiency.
 
-\section{Template Overview}
-As noted in the introduction, the ``\verb|acmart|'' document class can
-be used to prepare many different kinds of documentation --- a
-double-anonymous initial submission of a full-length technical paper, a
-two-page SIGGRAPH Emerging Technologies abstract, a ``camera-ready''
-journal article, a SIGCHI Extended Abstract, and more --- all by
-selecting the appropriate {\itshape template style} and {\itshape
-  template parameters}.
+ Although \textit{DecoupledLookback} achieves near-ideal performance on NVIDIA hardware and implemented in CUDA, it relies on a set of architectural and language-specific guarantees that no longer hold once outside of the NVIDIA ecosystem: forward progress guarantees (FPG), fixed subgroup sizes, explicit divergence handling, and memory fences.
+Without forward progress guarantees, the algorithm risks deadlock, forcing either a regression to the slower \textit{Reduce-then-Scan} approach or maintenance of multiple scan variants.
+Without fixed subgroup sizes, developers must again regress to slower scan implementations without subgroup acceleration or maintain multiple scan variants.
+Ambiguous divergence behavior can produce subtle correctness or performance issues, and without explicit masks, developers must rely on compiler and vendor-specific heuristics, increasing the risk of unpredictable behavior and further complicating portability efforts.
+Absent memory fences, as is the case in the D3D12 API, control over memory ordering is significantly diminished and can result in unbounded redundant work during the scan.
 
-This document will explain the major features of the document
-class. For further information, the {\itshape \LaTeX\ User's Guide} is
-available from
-\url{https://www.acm.org/publications/proceedings-template}.
-
-\subsection{Template Styles}
-
-The primary parameter given to the ``\verb|acmart|'' document class is
-the {\itshape template style} which corresponds to the kind of publication
-or SIG publishing the work. This parameter is enclosed in square
-brackets and is a part of the {\verb|documentclass|} command:
-\begin{verbatim}
-  \documentclass[STYLE]{acmart}
-\end{verbatim}
-
-Journals use one of three template styles. All but three ACM journals
-use the {\verb|acmsmall|} template style:
+While previous work has focused on improving the performance of prefix scan, the contribution of this work is portability.
+This work presents \textit{Chained Scan with Decoupled Lookback and Decoupled Fallback}, a fully portable prefix scan capable of reaching speed of light performance without FPG, and implemented in the WGSL shading language.
+On an intra-workgroup level, we contribute a subgroup-size-agnostic scan pattern, which enables a single scan pattern to be used for all possible subgroup sizes.
+On an inter-workgroup-level, we contribute \textit{DecoupledFallback}, an extension of the \textit{DecoupledLookback} technique that is no longer dependent on FPG to execute correctly.
+Our implementation is guided by the following goals:
 \begin{itemize}
-\item {\texttt{acmsmall}}: The default journal template style.
-\item {\texttt{acmlarge}}: Used by JOCCH and TAP.
-\item {\texttt{acmtog}}: Used by TOG.
+\item \textbf{Portability: The implementation must execute correctly on all hardwares and backend graphics API's supported by the WebGPU standard.} Our scan should execute correctly on any hardware and API, regardless of the underlying subgroup size, scheduling model, or divergence behavior.
+\item \textbf{Performance: The implementation must achieve speed of light performance whenever possible.} Our scan should achieve speed of light performance whenever possible, without relying on vendor-specific adaptations.
+\item \textbf{Maintainability: The implementation must use a single variant for all hardwares and must be minimally compiled.} Creating statically specialized per-vendor shader variants increases both maintenance and compilation times.
+Our goal is to make our implementation as lightweight to maintain and compile as possible.
+\item \textbf{Ease of use: Using the implementation should require as little developer intervention as possible.} Developers should not have to be conscious of, or make decisions based on, their target hardware when using our scan.
 \end{itemize}
 
-The majority of conference proceedings documentation will use the {\verb|acmconf|} template style.
+\section{Background}
+
+The study of prefix scan patterns can be traced to the design of adder circuits and beyond\cite{10.1145/322217.322232, 5219801}.
+Indeed, at their most granular level, contemporary GPU scan patterns still utilize classical parallel prefix adders like Hillis-Steele/Kogge-Stone\cite{10.1145/7902.7903, 5009159}, Brent-Kung\cite{1675982}, or Sklansky\cite{5219822}.
+
+\subsubsection{Evolution of Intra-Workgroup Scan Strategies}
+
+\subsection{Evolution of Inter-Workgroup Scan Strategies}
+In a prefix scan, the reduction at each element is dependent on the reduction of preceding elements.
+Thus a serial inter-workgroup dependency is created when the size of the scan exceeds the capacity of a single workgroup.
+
+\subsubsection{\textit{4n}: Scan-Then-Propagate}
+\cite{Sengupta2008, 10.5555/1280094.1280110, GPUGems3}
+\subsubsection{\textit{3n}: Reduce-Then-Scan}
+\cite{Merrill2009, 10.5555/110382.110597, 10.1145/1375527.1375559}
+\subsubsection{\textit{2n}: Single-Pass Scan}
+\cite{10.1145/2442516.2442539,Merrill2016}
+\subsection{Portability}
+\cite{sorensen_et_al:LIPIcs.CONCUR.2018.23, 10.1145/3485508, NvidiaCudaGuide, NvidiaVoltaWhitepaper}:
 \begin{itemize}
-\item {\texttt{sigconf}}: The default proceedings template style.
-\item{\texttt{sigchi}}: Used for SIGCHI conference articles.
-\item{\texttt{sigplan}}: Used for SIGPLAN conference articles.
+    \item \textbf{Forward Progress Guarantee:} Beginning with the Volta architecture, NVIDIA formalized forward-progress-guarantees (FPG) at both the workgroup and subgroup level.
+Prior to Volta, FPG was likely already present at a workgroup level.
+    \item \textbf{Fixed Subgroup Size:} On NVIDIA hardware, the subgroup width is fixed at 32, enabling static optimizations at that specific width.
+Furthermore, the square of the subgroup width is CUDA's maximum workgroup size, enabling straightforward implementation of workgroup reductions.
+    \item \textbf{Unambiguous Divergence Behavior:} In CUDA, developers can explicitly provision subgroup functions with a mask of participating threads, unambiguously defining the function's behavior irrespective of potential divergence.
+    \item \textbf{Memory Fence:} CUDA offers developers memory fence functions to enforce sequentially consistent atomic memory accesses, enabling more efficient inter-workgroup synchronization.
 \end{itemize}
 
-\subsection{Template Parameters}
+\section{Subgroup Size Agnostic Pattern}
 
-In addition to specifying the {\itshape template style} to be used in
-formatting your work, there are a number of {\itshape template parameters}
-which modify some part of the applied template style. A complete list
-of these parameters can be found in the {\itshape \LaTeX\ User's Guide.}
-
-Frequently-used parameters, or combinations of parameters, include:
-\begin{itemize}
-\item {\texttt{anonymous,review}}: Suitable for a ``double-anonymous''
-  conference submission. Anonymizes the work and includes line
-  numbers. Use with the \texttt{\acmSubmissionID} command to print the
-  submission's unique ID on each page of the work.
-\item{\texttt{authorversion}}: Produces a version of the work suitable
-  for posting by the author.
-\item{\texttt{screen}}: Produces colored hyperlinks.
-\end{itemize}
-
-This document uses the following string as the first command in the
-source file:
-\begin{verbatim}
-\documentclass[sigconf,authordraft]{acmart}
-\end{verbatim}
-
-\section{Modifications}
-
-Modifying the template --- including but not limited to: adjusting
-margins, typeface sizes, line spacing, paragraph and list definitions,
-and the use of the \verb|\vspace| command to manually adjust the
-vertical spacing between elements of your work --- is not allowed.
-
-{\bfseries Your document will be returned to you for revision if
-  modifications are discovered.}
-
-\section{Typefaces}
-
-The ``\verb|acmart|'' document class requires the use of the
-``Libertine'' typeface family. Your \TeX\ installation should include
-this set of packages. Please do not substitute other typefaces. The
-``\verb|lmodern|'' and ``\verb|ltimes|'' packages should not be used,
-as they will override the built-in typeface families.
-
-\section{Title Information}
-
-The title of your work should use capital letters appropriately -
-\url{https://capitalizemytitle.com/} has useful rules for
-capitalization. Use the {\verb|title|} command to define the title of
-your work. If your work has a subtitle, define it with the
-{\verb|subtitle|} command.  Do not insert line breaks in your title.
-
-If your title is lengthy, you must define a short version to be used
-in the page headers, to prevent overlapping text. The \verb|title|
-command has a ``short title'' parameter:
-\begin{verbatim}
-  \title[short title]{full title}
-\end{verbatim}
-
-\section{Authors and Affiliations}
-
-Each author must be defined separately for accurate metadata
-identification.  As an exception, multiple authors may share one
-affiliation. Authors' names should not be abbreviated; use full first
-names wherever possible. Include authors' e-mail addresses whenever
-possible.
-
-Grouping authors' names or e-mail addresses, or providing an ``e-mail
-alias,'' as shown below, is not acceptable:
-\begin{verbatim}
-  \author{Brooke Aster, David Mehldau}
-  \email{dave,judy,steve@university.edu}
-  \email{firstname.lastname@phillips.org}
-\end{verbatim}
-
-The \verb|authornote| and \verb|authornotemark| commands allow a note
-to apply to multiple authors --- for example, if the first two authors
-of an article contributed equally to the work.
-
-If your author list is lengthy, you must define a shortened version of
-the list of authors to be used in the page headers, to prevent
-overlapping text. The following command should be placed just after
-the last \verb|\author{}| definition:
-\begin{verbatim}
-  \renewcommand{\shortauthors}{McCartney, et al.}
-\end{verbatim}
-Omitting this command will force the use of a concatenated list of all
-of the authors' names, which may result in overlapping text in the
-page headers.
-
-The article template's documentation, available at
-\url{https://www.acm.org/publications/proceedings-template}, has a
-complete explanation of these commands and tips for their effective
-use.
-
-Note that authors' addresses are mandatory for journal articles.
-
-\section{Rights Information}
-
-Authors of any work published by ACM will need to complete a rights
-form. Depending on the kind of work, and the rights management choice
-made by the author, this may be copyright transfer, permission,
-license, or an OA (open access) agreement.
-
-Regardless of the rights management choice, the author will receive a
-copy of the completed rights form once it has been submitted. This
-form contains \LaTeX\ commands that must be copied into the source
-document. When the document source is compiled, these commands and
-their parameters add formatted text to several areas of the final
-document:
-\begin{itemize}
-\item the ``ACM Reference Format'' text on the first page.
-\item the ``rights management'' text on the first page.
-\item the conference information in the page header(s).
-\end{itemize}
-
-Rights information is unique to the work; if you are preparing several
-works for an event, make sure to use the correct set of commands with
-each of the works.
-
-The ACM Reference Format text is required for all articles over one
-page in length, and is optional for one-page articles (abstracts).
-
-\section{CCS Concepts and User-Defined Keywords}
-
-Two elements of the ``acmart'' document class provide powerful
-taxonomic tools for you to help readers find your work in an online
-search.
-
-The ACM Computing Classification System ---
-\url{https://www.acm.org/publications/class-2012} --- is a set of
-classifiers and concepts that describe the computing
-discipline. Authors can select entries from this classification
-system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the
-commands to be included in the \LaTeX\ source.
-
-User-defined keywords are a comma-separated list of words and phrases
-of the authors' choosing, providing a more flexible way of describing
-the research being presented.
-
-CCS concepts and user-defined keywords are required for for all
-articles over two pages in length, and are optional for one- and
-two-page articles (or abstracts).
-
-\section{Sectioning Commands}
-
-Your work should use standard \LaTeX\ sectioning commands:
-\verb|section|, \verb|subsection|, \verb|subsubsection|, and
-\verb|paragraph|. They should be numbered; do not remove the numbering
-from the commands.
-
-Simulating a sectioning command by setting the first word or words of
-a paragraph in boldface or italicized text is {\bfseries not allowed.}
-
-\section{Tables}
-
-The ``\verb|acmart|'' document class includes the ``\verb|booktabs|''
-package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
-high-quality tables.
-
-Table captions are placed {\itshape above} the table.
-
-Because tables cannot be split across pages, the best placement for
-them is typically the top of the page nearest their initial cite.  To
-ensure this proper ``floating'' placement of tables, use the
-environment \textbf{table} to enclose the table's contents and the
-table caption.  The contents of the table itself must go in the
-\textbf{tabular} environment, to be aligned properly in rows and
-columns, with the desired horizontal and vertical rules.  Again,
-detailed instructions on \textbf{tabular} material are found in the
-\textit{\LaTeX\ User's Guide}.
-
-Immediately following this sentence is the point at which
-Table~\ref{tab:freq} is included in the input file; compare the
-placement of the table here with the table in the printed output of
-this document.
-
-\begin{table}
-  \caption{Frequency of Special Characters}
-  \label{tab:freq}
-  \begin{tabular}{ccl}
-    \toprule
-    Non-English or Math&Frequency&Comments\\
-    \midrule
-    \O & 1 in 1,000& For Swedish names\\
-    $\pi$ & 1 in 5& Common in math\\
-    \$ & 4 in 5 & Used in business\\
-    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
-  \bottomrule
-\end{tabular}
-\end{table}
-
-To set a wider table, which takes up the whole width of the page's
-live area, use the environment \textbf{table*} to enclose the table's
-contents and the table caption.  As with a single-column table, this
-wide table will ``float'' to a location deemed more
-desirable. Immediately following this sentence is the point at which
-Table~\ref{tab:commands} is included in the input file; again, it is
-instructive to compare the placement of the table here with the table
-in the printed output of this document.
-
-\begin{table*}
-  \caption{Some Typical Commands}
-  \label{tab:commands}
-  \begin{tabular}{ccl}
-    \toprule
-    Command &A Number & Comments\\
-    \midrule
-    \texttt{{\char'134}author} & 100& Author \\
-    \texttt{{\char'134}table}& 300 & For tables\\
-    \texttt{{\char'134}table*}& 400& For wider tables\\
-    \bottomrule
-  \end{tabular}
-\end{table*}
-
-Always use midrule to separate table header rows from data rows, and
-use it only for this purpose. This enables assistive technologies to
-recognise table headers and support their users in navigating tables
-more easily.
-
-\section{Math Equations}
-You may want to display math equations in three distinct styles:
-inline, numbered or non-numbered display.  Each of the three are
-discussed in the next sections.
-
-\subsection{Inline (In-text) Equations}
-A formula that appears in the running text is called an inline or
-in-text formula.  It is produced by the \textbf{math} environment,
-which can be invoked with the usual
-\texttt{{\char'134}begin\,\ldots{\char'134}end} construction or with
-the short form \texttt{\$\,\ldots\$}. You can use any of the symbols
-and structures, from $\alpha$ to $\omega$, available in
-\LaTeX~\cite{Lamport:LaTeX}; this section will simply show a few
-examples of in-text equations in context. Notice how this equation:
-\begin{math}
-  \lim_{n\rightarrow \infty}x=0
-\end{math},
-set here in in-line math style, looks slightly different when
-set in display style.  (See next section).
-
-\subsection{Display Equations}
-A numbered display equation---one set off by vertical space from the
-text and centered horizontally---is produced by the \textbf{equation}
-environment. An unnumbered display equation is produced by the
-\textbf{displaymath} environment.
-
-Again, in either environment, you can use any of the symbols and
-structures available in \LaTeX\@; this section will just give a couple
-of examples of display equations in context.  First, consider the
-equation, shown as an inline equation above:
-\begin{equation}
-  \lim_{n\rightarrow \infty}x=0
-\end{equation}
-Notice how it is formatted somewhat differently in
-the \textbf{displaymath}
-environment.  Now, we'll enter an unnumbered equation:
-\begin{displaymath}
-  \sum_{i=0}^{\infty} x + 1
-\end{displaymath}
-and follow it with another numbered equation:
-\begin{equation}
-  \sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f
-\end{equation}
-just to demonstrate \LaTeX's able handling of numbering.
-
-\section{Figures}
-
-The ``\verb|figure|'' environment should be used for figures. One or
-more images can be placed within a figure. If your figure contains
-third-party material, you must clearly identify it as such, as shown
-in the example below.
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{sample-franklin}
-  \caption{1907 Franklin Model D roadster. Photograph by Harris \&
-    Ewing, Inc. [Public domain], via Wikimedia
-    Commons. (\url{https://goo.gl/VLCRBB}).}
-  \Description{A woman and a girl in white dresses sit in an open car.}
-\end{figure}
-
-Your figures should contain a caption which describes the figure to
-the reader.
-
-Figure captions are placed {\itshape below} the figure.
-
-Every figure should also have a figure description unless it is purely
-decorative. These descriptions convey what’s in the image to someone
-who cannot see it. They are also used by search engine crawlers for
-indexing images, and when images cannot be loaded.
-
-A figure description must be unformatted plain text less than 2000
-characters long (including spaces).  {\bfseries Figure descriptions
-  should not repeat the figure caption – their purpose is to capture
-  important information that is not already provided in the caption or
-  the main text of the paper.} For figures that convey important and
-complex new information, a short text description may not be
-adequate. More complex alternative descriptions can be placed in an
-appendix and referenced in a short figure description. For example,
-provide a data table capturing the information in a bar chart, or a
-structured list representing a graph.  For additional information
-regarding how best to write figure descriptions and why doing this is
-so important, please see
-\url{https://www.acm.org/publications/taps/describing-figures/}.
-
-\subsection{The ``Teaser Figure''}
-
-A ``teaser figure'' is an image, or set of images in one figure, that
-are placed after all author and affiliation information, and before
-the body of the article, spanning the page. If you wish to have such a
-figure in your article, place the command immediately before the
-\verb|\maketitle| command:
-\begin{verbatim}
-  \begin{teaserfigure}
-    \includegraphics[width=\textwidth]{sampleteaser}
-    \caption{figure caption}
-    \Description{figure description}
-  \end{teaserfigure}
-\end{verbatim}
-
-\section{Citations and Bibliographies}
-
-The use of \BibTeX\ for the preparation and formatting of one's
-references is strongly recommended. Authors' names should be complete
---- use full first names (``Donald E. Knuth'') not initials
-(``D. E. Knuth'') --- and the salient identifying features of a
-reference should be included: title, year, volume, number, pages,
-article DOI, etc.
-
-The bibliography is included in your source document with these two
-commands, placed just before the \verb|\end{document}| command:
-\begin{verbatim}
-  \bibliographystyle{ACM-Reference-Format}
-  \bibliography{bibfile}
-\end{verbatim}
-where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
-suffix, of the \BibTeX\ file.
-
-Citations and references are numbered by default. A small number of
-ACM publications have citations and references formatted in the
-``author year'' style; for these exceptions, please include this
-command in the {\bfseries preamble} (before the command
-``\verb|\begin{document}|'') of your \LaTeX\ source:
-\begin{verbatim}
-  \citestyle{acmauthoryear}
-\end{verbatim}
-
-  Some examples.  A paginated journal article \cite{Abril07}, an
-  enumerated journal article \cite{Cohen07}, a reference to an entire
-  issue \cite{JCohen96}, a monograph (whole book) \cite{Kosiur01}, a
-  monograph/whole book in a series (see 2a in spec. document)
-  \cite{Harel79}, a divisible-book such as an anthology or compilation
-  \cite{Editor00} followed by the same example, however we only output
-  the series if the volume number is given \cite{Editor00a} (so
-  Editor00a's series should NOT be present since it has no vol. no.),
-  a chapter in a divisible book \cite{Spector90}, a chapter in a
-  divisible book in a series \cite{Douglass98}, a multi-volume work as
-  book \cite{Knuth97}, a couple of articles in a proceedings (of a
-  conference, symposium, workshop for example) (paginated proceedings
-  article) \cite{Andler79, Hagerup1993}, a proceedings article with
-  all possible elements \cite{Smith10}, an example of an enumerated
-  proceedings article \cite{VanGundy07}, an informally published work
-  \cite{Harel78}, a couple of preprints \cite{Bornmann2019,
-    AnzarootPBM14}, a doctoral dissertation \cite{Clarkson85}, a
-  master's thesis: \cite{anisi03}, an online document / world wide web
-  resource \cite{Thornburg01, Ablamowicz07, Poker06}, a video game
-  (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03} and \cite{Lee05}
-  and (Case 3) a patent \cite{JoeScientist001}, work accepted for
-  publication \cite{rous08}, 'YYYYb'-test for prolific author
-  \cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might
-  contain 'duplicate' DOI and URLs (some SIAM articles)
-  \cite{Kirschmer:2010:AEI:1958016.1958018}. Boris / Barbara Beeton:
-  multi-volume works as books \cite{MR781536} and \cite{MR781537}. A
-  couple of citations with DOIs:
-  \cite{2004:ITE:1009386.1010128,Kirschmer:2010:AEI:1958016.1958018}. Online
-  citations: \cite{TUGInstmem, Thornburg01, CTANacmart}.
-  Artifacts: \cite{R} and \cite{UMassCitations}.
-
-\section{Acknowledgments}
-
-Identification of funding sources and other support, and thanks to
-individuals and groups that assisted in the research and the
-preparation of the work should be included in an acknowledgment
-section, which is placed just before the reference section in your
-document.
-
-This section has a special environment:
-\begin{verbatim}
+\section{Chained Scan with Decoupled Lookback and Decoupled Fallback}
 \begin{acks}
-  ...
-  \end{acks}
-\end{verbatim}
-so that the information contained therein can be more easily collected
-during the article metadata extraction phase, and to ensure
-consistency in the spelling of the section heading.
-
-Authors should not prepare this section as a numbered or unnumbered {\verb|\section|}; please use the ``{\verb|acks|}'' environment.
-
-\section{Appendices}
-
-If your work needs an appendix, add it before the
-``\verb|\end{document}|'' command at the conclusion of your source
-document.
-
-Start the appendix with the ``\verb|appendix|'' command:
-\begin{verbatim}
-  \appendix
-\end{verbatim}
-and note that in the appendix, sections are lettered, not
-numbered. This document has two appendices, demonstrating the section
-and subsection identification method.
-
-\section{Multi-language papers}
-
-Papers may be written in languages other than English or include
-titles, subtitles, keywords and abstracts in different languages (as a
-rule, a paper in a language other than English should include an
-English title and an English abstract).  Use \verb|language=...| for
-every language used in the paper.  The last language indicated is the
-main language of the paper.  For example, a French paper with
-additional titles and abstracts in English and German may start with
-the following command
-\begin{verbatim}
-\documentclass[sigconf, language=english, language=german,
-               language=french]{acmart}
-\end{verbatim}
-
-The title, subtitle, keywords and abstract will be typeset in the main
-language of the paper.  The commands \verb|\translatedXXX|, \verb|XXX|
-begin title, subtitle and keywords, can be used to set these elements
-in the other languages.  The environment \verb|translatedabstract| is
-used to set the translation of the abstract.  These commands and
-environment have a mandatory first argument: the language of the
-second argument.  See \verb|sample-sigconf-i13n.tex| file for examples
-of their usage.
-
-\section{SIGCHI Extended Abstracts}
-
-The ``\verb|sigchi-a|'' template style (available only in \LaTeX\ and
-not in Word) produces a landscape-orientation formatted article, with
-a wide left margin. Three environments are available for use with the
-``\verb|sigchi-a|'' template style, and produce formatted output in
-the margin:
-\begin{description}
-\item[\texttt{sidebar}:]  Place formatted text in the margin.
-\item[\texttt{marginfigure}:] Place a figure in the margin.
-\item[\texttt{margintable}:] Place a table in the margin.
-\end{description}
-
-%%
-%% The acknowledgments section is defined using the "acks" environment
-%% (and NOT an unnumbered section). This ensures the proper
-%% identification of the section in the article metadata, and the
-%% consistent spelling of the heading.
-\begin{acks}
-To Robert, for the bagels and explaining CMYK and color spaces.
 \end{acks}
 
-%%
-%% The next two lines define the bibliography style to be used, and
-%% the bibliography file.
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{sample-base}
+\bibliography{bib}
 
-%%
-%% If your work has an appendix, this is the place to put it.
 \appendix
@@
 
-\subsection{Part One}
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
-malesuada, quam in pulvinar varius, metus nunc fermentum urna, id
-sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum
-vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et
-lacinia dolor. Integer ultricies commodo sem nec semper.
-
-\subsection{Part Two}
-
-Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales
-ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non
-ipsum convallis congue hendrerit vitae sapien. Donec at laoreet
-eros. Vivamus non purus placerat, scelerisque diam eu, cursus
-ante. Etiam aliquam tortor auctor efficitur mattis.
-
-\section{Online Resources}
-
-Nam id fermentum dui. Suspendisse sagittis tortor a nulla mollis, in
-pulvinar ex pretium. Sed interdum orci quis metus euismod, et sagittis
-enim maximus. Vestibulum gravida massa ut felis suscipit
-congue. Quisque mattis elit a risus ultrices commodo venenatis eget
-dui. Etiam sagittis eleifend elementum.
-
-Nam interdum magna at lectus dignissim, ac dignissim lorem
-rhoncus. Maecenas eu arcu ac neque placerat aliquam. Nunc pulvinar
-massa et mattis lacinia.
-
 \end{document}
 \endinput
-%%
-%% End of file `sample-sigconf-authordraft.tex'.

### 2024-12-10 paper.tex
this is just me making typography changes as an example of using vscode for the first time for latex

@@
 %% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
+%TC:macro~\cite [option:text,text]
+%TC:macro~\citep [option:text,text]
+%TC:macro~\citet [option:text,text]
 %TC:envir table 0 1
@@
   As the compute capabilities of major graphics APIs have matured, there has been growing utilization of compute shaders for graphical tasks, and as such, there has been corresponding interest in porting over primitives like prefix scan from compute languages to shading languages.
-However, the current state-of-the-art prefix scan, \textit{Chained Scan with Decoupled Lookback}, relies on properties provided by the NVIDIA ecosystem that, once removed, force developers to make trade-offs between maintainability, portability, and performance.
-We describe \textit{DecoupledFallback}, a fully portable, single-pass prefix scan method capable of reaching speed of light scan performance across different hardwares with minimal developer intervention.
+However, the current state-of-the-art prefix scan, \emph{Chained Scan with Decoupled Lookback}, relies on properties provided by the NVIDIA ecosystem that, once removed, force developers to make trade-offs between maintainability, portability, and performance.
+We describe \emph{DecoupledFallback}, a fully portable, single-pass prefix scan method capable of reaching speed of light scan performance across different hardwares with minimal developer intervention.
 Our implementation is built in WGSL and is compatible with any implementation of the WebGPU standard (I think?
@@
 \section{Introduction}
-Prefix scan is one of the most fundamental primitives in parallel computing, with uses including compaction, sorting \cite{10.1145/3035918.3064043, adinets2022onesweepfastersignificantdigit}, SPMV \cite{} etc..
-Sometimes referred to as \textit{scan} or \textit{prefix reduction}, prefix scan is typically defined on a monoid, though it can also be defined on a semigroup in certain cases\cite{10.1007/978-3-540-27764-4_11}.
+Prefix scan is one of the most fundamental primitives in parallel computing, with uses including compaction, sorting~\cite{10.1145/3035918.3064043, adinets2022onesweepfastersignificantdigit}, SPMV~\cite{} etc.
+Sometimes referred to as \emph{scan} or \emph{prefix reduction}, prefix scan is typically defined on a monoid, though it can also be defined on a semigroup in certain cases~\cite{10.1007/978-3-540-27764-4_11}.
 In a prefix scan, the result at element $n$ is the reduction of the preceding subset of elements in the sequence.
-If the reduction subset includes the $n$-th element, it is called \textit{inclusive}; if it excludes the $n$-th element, it is called \textit{exclusive}.
+If the reduction subset includes the $n$-th element, it is called \emph{inclusive}; if it excludes the $n$-th element, it is called \emph{exclusive}.
 The binary operator used in the scan must be associative, but it need not be commutative—for instance, in the case of the stack monoid.
@@
 With proper latency-hiding, prefix scan is computationally light enough that it is memory-bandwidth bound, and thus the focus of contemporary scan strategies has been global communication avoidance.
-The current state-of-the-art prefix scan is \textit{Chained Scan with Decoupled Lookback}\cite{Merrill2016} (referred to hereon as \textit{DecoupledLookback}).
+The current state-of-the-art prefix scan is \emph{Chained Scan with Decoupled Lookback}~\cite{Merrill2016} (referred to hereon as \emph{DecoupledLookback}).
 Although a chained scan\footnote{A chained scan is a scan performed serially.} may seem inimical to parallelism, the key innovation of GPU chained scanning lies in its hybridization of parallel and serial scan strategies at different levels of the GPU memory hierarchy, achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level.
-\textit{DecoupledLookback} leverages this hybrid strategy while also being capable of fully saturating global memory bandwidth, and as a result, with approximately \textasciitilde$2n$ global data movement—one read and one write per processed element—it matches the performance of a copy operation, effectively achieving “speed of light” efficiency.
+\emph{DecoupledLookback} leverages this hybrid strategy while also being capable of fully saturating global memory bandwidth, and as a result, with approximately \textasciitilde$2n$ global data movement—one read and one write per processed element—it matches the performance of a copy operation, effectively achieving “speed of light” efficiency.
 
- Although \textit{DecoupledLookback} achieves near-ideal performance on NVIDIA hardware and implemented in CUDA, it relies on a set of architectural and language-specific guarantees that no longer hold once outside of the NVIDIA ecosystem: forward progress guarantees (FPG), fixed subgroup sizes, explicit divergence handling, and memory fences.
-Without forward progress guarantees, the algorithm risks deadlock, forcing either a regression to the slower \textit{Reduce-then-Scan} approach or maintenance of multiple scan variants.
+Although \emph{DecoupledLookback} achieves near-ideal performance on NVIDIA hardware and implemented in CUDA, it relies on a set of architectural and language-specific guarantees that no longer hold once outside of the NVIDIA ecosystem: forward progress guarantees (FPG), fixed subgroup sizes, explicit divergence handling, and memory fences.
+Without forward progress guarantees, the algorithm risks deadlock, forcing either a regression to the slower \emph{Reduce-then-Scan} approach or maintenance of multiple scan variants.
 Without fixed subgroup sizes, developers must again regress to slower scan implementations without subgroup acceleration or maintain multiple scan variants.
@@
 While previous work has focused on improving the performance of prefix scan, the contribution of this work is portability.
-This work presents \textit{Chained Scan with Decoupled Lookback and Decoupled Fallback}, a fully portable prefix scan capable of reaching speed of light performance without FPG, and implemented in the WGSL shading language.
+This work presents \emph{Chained Scan with Decoupled Lookback and Decoupled Fallback}, a fully portable prefix scan capable of reaching speed of light performance without FPG, and implemented in the WGSL shading language.
 On an intra-workgroup level, we contribute a subgroup-size-agnostic scan pattern, which enables a single scan pattern to be used for all possible subgroup sizes.
-On an inter-workgroup-level, we contribute \textit{DecoupledFallback}, an extension of the \textit{DecoupledLookback} technique that is no longer dependent on FPG to execute correctly.
+On an inter-workgroup-level, we contribute \emph{DecoupledFallback}, an extension of the \emph{DecoupledLookback} technique that is no longer dependent on FPG to execute correctly.
 Our implementation is guided by the following goals:
 \begin{itemize}
-\item \textbf{Portability: The implementation must execute correctly on all hardwares and backend graphics API's supported by the WebGPU standard.} Our scan should execute correctly on any hardware and API, regardless of the underlying subgroup size, scheduling model, or divergence behavior.
+  \item \textbf{Portability: The implementation must execute correctly on all hardwares and backend graphics APIs supported by the WebGPU standard.} Our scan should execute correctly on any hardware and API, regardless of the underlying subgroup size, scheduling model, or divergence behavior.
   \item \textbf{Performance: The implementation must achieve speed of light performance whenever possible.} Our scan should achieve speed of light performance whenever possible, without relying on vendor-specific adaptations.
@@
 
-The study of prefix scan patterns can be traced to the design of adder circuits and beyond\cite{10.1145/322217.322232, 5219801}.
-Indeed, at their most granular level, contemporary GPU scan patterns still utilize classical parallel prefix adders like Hillis-Steele/Kogge-Stone\cite{10.1145/7902.7903, 5009159}, Brent-Kung\cite{1675982}, or Sklansky\cite{5219822}.
+The study of prefix scan patterns can be traced to the design of adder circuits and beyond~\cite{10.1145/322217.322232, 5219801}.
+Indeed, at their most granular level, contemporary GPU scan patterns still utilize classical parallel prefix adders like Hillis-Steele/Kogge-Stone~\cite{10.1145/7902.7903, 5009159}, Brent-Kung~\cite{1675982}, or Sklansky~\cite{5219822}.
 
@@
 
-\subsubsection{\textit{4n}: Scan-Then-Propagate}
+\subsubsection{\emph{4n}: Scan-Then-Propagate}
 \cite{Sengupta2008, 10.5555/1280094.1280110, GPUGems3}
-\subsubsection{\textit{3n}: Reduce-Then-Scan}
+\subsubsection{\emph{3n}: Reduce-Then-Scan}
 \cite{Merrill2009, 10.5555/110382.110597, 10.1145/1375527.1375559}
-\subsubsection{\textit{2n}: Single-Pass Scan}
+\subsubsection{\emph{2n}: Single-Pass Scan}
 \cite{10.1145/2442516.2442539,Merrill2016}

### 2024-12-19 paper.tex
Add Why Does . . .?, GPU Programming Environment, and Goals subsections.

@@
 
+\usepackage[ruled,vlined]{algorithm2e}
 \begin{document}
@@
 \begin{abstract}
-  As the compute capabilities of major graphics APIs have matured, there has been growing utilization of compute shaders for graphical tasks, and as such, there has been corresponding interest in porting over primitives like prefix scan from compute languages to shading languages.
-However, the current state-of-the-art prefix scan, \emph{Chained Scan with Decoupled Lookback}, relies on properties provided by the NVIDIA ecosystem that, once removed, force developers to make trade-offs between maintainability, portability, and performance.
-We describe \emph{DecoupledFallback}, a fully portable, single-pass prefix scan method capable of reaching speed of light scan performance across different hardwares with minimal developer intervention.
-Our implementation is built in WGSL and is compatible with any implementation of the WebGPU standard (I think?
-It definitely does not work on WARP .
-.
-.
-which is concerning).
+  TODO
 \end{abstract}
@@
 \section{Introduction}
-Prefix scan is one of the most fundamental primitives in parallel computing, with uses including compaction, sorting~\cite{10.1145/3035918.3064043, adinets2022onesweepfastersignificantdigit}, SPMV~\cite{} etc.
-Sometimes referred to as \emph{scan} or \emph{prefix reduction}, prefix scan is typically defined on a monoid, though it can also be defined on a semigroup in certain cases~\cite{10.1007/978-3-540-27764-4_11}.
-In a prefix scan, the result at element $n$ is the reduction of the preceding subset of elements in the sequence.
-If the reduction subset includes the $n$-th element, it is called \emph{inclusive}; if it excludes the $n$-th element, it is called \emph{exclusive}.
-The binary operator used in the scan must be associative, but it need not be commutative—for instance, in the case of the stack monoid.
-The most common type of prefix scan is prefix sum, where the binary operator is addition.
-For example\footnote{In this paper we will use the terminology of WGSL, but we note that subgroup is interchangeable with warp(CUDA), wave(HLSL), and simd\_group(Metal), while workgroup is interchangeable with block(CUDA), group(HLSL), and ???(Metal)}:
-
-(PREFIX SUM EXAMPLE WITH BINARY REDUCTION OPERATOR)
-
-Contemporary GPU's are characterized by their high memory bandwidth, high arithmetic throughput, high memory latency, and hierarchical memory model.
-With proper latency-hiding, prefix scan is computationally light enough that it is memory-bandwidth bound, and thus the focus of contemporary scan strategies has been global communication avoidance.
-The current state-of-the-art prefix scan is \emph{Chained Scan with Decoupled Lookback}~\cite{Merrill2016} (referred to hereon as \emph{DecoupledLookback}).
-Although a chained scan\footnote{A chained scan is a scan performed serially.} may seem inimical to parallelism, the key innovation of GPU chained scanning lies in its hybridization of parallel and serial scan strategies at different levels of the GPU memory hierarchy, achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level.
-\emph{DecoupledLookback} leverages this hybrid strategy while also being capable of fully saturating global memory bandwidth, and as a result, with approximately \textasciitilde$2n$ global data movement—one read and one write per processed element—it matches the performance of a copy operation, effectively achieving “speed of light” efficiency.
-
-Although \emph{DecoupledLookback} achieves near-ideal performance on NVIDIA hardware and implemented in CUDA, it relies on a set of architectural and language-specific guarantees that no longer hold once outside of the NVIDIA ecosystem: forward progress guarantees (FPG), fixed subgroup sizes, explicit divergence handling, and memory fences.
-Without forward progress guarantees, the algorithm risks deadlock, forcing either a regression to the slower \emph{Reduce-then-Scan} approach or maintenance of multiple scan variants.
-Without fixed subgroup sizes, developers must again regress to slower scan implementations without subgroup acceleration or maintain multiple scan variants.
-Ambiguous divergence behavior can produce subtle correctness or performance issues, and without explicit masks, developers must rely on compiler and vendor-specific heuristics, increasing the risk of unpredictable behavior and further complicating portability efforts.
-Absent memory fences, as is the case in the D3D12 API, control over memory ordering is significantly diminished and can result in unbounded redundant work during the scan.
-
-While previous work has focused on improving the performance of prefix scan, the contribution of this work is portability.
-This work presents \emph{Chained Scan with Decoupled Lookback and Decoupled Fallback}, a fully portable prefix scan capable of reaching speed of light performance without FPG, and implemented in the WGSL shading language.
-On an intra-workgroup level, we contribute a subgroup-size-agnostic scan pattern, which enables a single scan pattern to be used for all possible subgroup sizes.
-On an inter-workgroup-level, we contribute \emph{DecoupledFallback}, an extension of the \emph{DecoupledLookback} technique that is no longer dependent on FPG to execute correctly.
-Our implementation is guided by the following goals:
-\begin{itemize}
-  \item \textbf{Portability: The implementation must execute correctly on all hardwares and backend graphics APIs supported by the WebGPU standard.} Our scan should execute correctly on any hardware and API, regardless of the underlying subgroup size, scheduling model, or divergence behavior.
-  \item \textbf{Performance: The implementation must achieve speed of light performance whenever possible.} Our scan should achieve speed of light performance whenever possible, without relying on vendor-specific adaptations.
-  \item \textbf{Maintainability: The implementation must use a single variant for all hardwares and must be minimally compiled.} Creating statically specialized per-vendor shader variants increases both maintenance and compilation times.
-Our goal is to make our implementation as lightweight to maintain and compile as possible.
-  \item \textbf{Ease of use: Using the implementation should require as little developer intervention as possible.} Developers should not have to be conscious of, or make decisions based on, their target hardware when using our scan.
-\end{itemize}
+ Do. this. last.
 
 \section{Background}
+\subsection{The GPU Programming Environment(Model?)}
+ Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency.
+GPU programming exhibits both single-instruction, multiple-thread (SIMT) and single-program, multiple-data (SPMD) behaviors.
+When \emph{launched}\footnote{Also referred to as \emph{dispatched}.}, a GPU program—known as a \emph{kernel}—is replicated across independent execution contexts, referred to as threads.
+Threads are organized into groups called \emph{workgroups}\footnote{Also referred to as \emph{Cross-Thread-Arrays}, \emph{Thread-Blocks}, \emph{Blocks}, \emph{Thread-Groups}, or \emph{Groups}.}, which are further divided into smaller units known as \emph{subgroups}\footnote{Also referred to as \emph{Warps}, \emph{Waves}, or \emph{SIMD-Groups}.}.
+This thread hierarchy mirrors the GPU memory hierarchy: individual threads access private, high-speed registers; threads within a workgroup share low-latency shared memory\footnote{Also referred to as \emph{Local Data Storage}, \emph{Group-Shared Memory}, (Check metal terminology).}; and all threads have access to high-bandwidth but higher-latency global memory.
 
-The study of prefix scan patterns can be traced to the design of adder circuits and beyond~\cite{10.1145/322217.322232, 5219801}.
-Indeed, at their most granular level, contemporary GPU scan patterns still utilize classical parallel prefix adders like Hillis-Steele/Kogge-Stone~\cite{10.1145/7902.7903, 5009159}, Brent-Kung~\cite{1675982}, or Sklansky~\cite{5219822}.
+ While threads within the same workgroup share resources, a workgroup is a logical unit of execution, not a physical processor.
+Instead, a GPU's underlying multiprocessors dynamically map and schedule the workgroups.
+This abstraction enables the kernel to be executed on GPUs with varying hardware capabilities, such as different numbers of multiprocessors or multiprocessors with differing execution resources.
+A single multiprocessor can host multiple workgroups; we define the number of active subgroups on a multiprocessor as the \emph{subgroup occupancy}, and the number of active workgroups across the entire GPU as the \emph{workgroup occupancy}.
+High subgroup occupancy is generally desirable as it hides memory and execution latency by ensuring sufficient warps are available for scheduling while others wait for data.
 
-\subsubsection{Evolution of Intra-Workgroup Scan Strategies}
+ A multiprocessor schedules and executes threads by subgroup.
+While it attempts to execute the same instruction across all threads in the subgroup, SIMT, threads within the same subgroup are free to follow their own data/context-dependent execution paths, SPMD.
+However, this flexibility comes at the cost of subgroup divergence.
+When threads in a subgroup diverge due to conditional branching or other control flow, each unique path is executed sequentially, but at the full execution width, effectively rendering the operation serial and reducing parallel efficiency.\footnote{The scheduling and execution width of the underlying hardware does not necessarily align with the subgroup width.}
 
-\subsection{Evolution of Inter-Workgroup Scan Strategies}
-In a prefix scan, the reduction at each element is dependent on the reduction of preceding elements.
-Thus a serial inter-workgroup dependency is created when the size of the scan exceeds the capacity of a single workgroup.
+ As we will discuss in the seciton, (insert section here), how a multiprocessor chooses \emph{which} out of the currently occupying subgroups to schedule can have a profound impact on both the performance and the correct execution of kernels.
 
-\subsubsection{\emph{4n}: Scan-Then-Propagate}
-\cite{Sengupta2008, 10.5555/1280094.1280110, GPUGems3}
-\subsubsection{\emph{3n}: Reduce-Then-Scan}
-\cite{Merrill2009, 10.5555/110382.110597, 10.1145/1375527.1375559}
-\subsubsection{\emph{2n}: Single-Pass Scan}
-\cite{10.1145/2442516.2442539,Merrill2016}
-\subsection{Portability}
-\cite{sorensen_et_al:LIPIcs.CONCUR.2018.23, 10.1145/3485508, NvidiaCudaGuide, NvidiaVoltaWhitepaper}:
-\begin{itemize}
-  \item \textbf{Forward Progress Guarantee:} Beginning with the Volta architecture, NVIDIA formalized forward-progress-guarantees (FPG) at both the workgroup and subgroup level.
-Prior to Volta, FPG was likely already present at a workgroup level.
-  \item \textbf{Fixed Subgroup Size:} On NVIDIA hardware, the subgroup width is fixed at 32, enabling static optimizations at that specific width.
-Furthermore, the square of the subgroup width is CUDA's maximum workgroup size, enabling straightforward implementation of workgroup reductions.
-  \item \textbf{Unambiguous Divergence Behavior:} In CUDA, developers can explicitly provision subgroup functions with a mask of participating threads, unambiguously defining the function's behavior irrespective of potential divergence.
-  \item \textbf{Memory Fence:} CUDA offers developers memory fence functions to enforce sequentially consistent atomic memory accesses, enabling more efficient inter-workgroup synchronization.
-\end{itemize}
+\subsection{Evolution of Inter-Workgroup Scan Architectures}
+\subsubsection{Scan-then-Propagate}
+\subsubsection{Reduce-then-Scan}
+\subsubsection{Chained-Scan}
 
-\section{Subgroup Size Agnostic Pattern}
+\subsection{Why does \emph{Chained-Scan} Rely on Forward Progress Guarantees?}
+ In a scan, the reduction at each element is dependent on the reduction of preceding elements.
+Thus, a serial dependency is created between workgroups whenever the number of elements in a scan operation exceeds the maximum size that can be processed by a single workgroup.
+Historically, hardware vendors have only recently begun to formally support inter-workgroup synchronization.
+Prior to the introduction of \emph{thread block clusters} on NVIDIA's Hopper architecture\cite{}, we are unaware of any GPU programming framework\footnote{We are not aware of any prior inter-workgroup level barrier primitive in CUDA, OpenCL, GLSL, HLSL, or Metal.} that provided inter-workgroup synchronization primitives.
+Instead, the earlier \emph{Scan-then-Propagate} and \emph{Reduce-then-Scan} architectures rely on kernel launches to act as inter-workgroup synchronization points\cite{}.
 
-\section{Chained Scan with Decoupled Lookback and Decoupled Fallback}
+ The \emph{Chained-Scan} architecture operates in the middle ground between the absence of inter-workgroup synchronization and formal inter-workgroup synchronization support.
+Rather than relying on explicit synchronization primitives, it leverages a combination of atomic operations and a property of the hardware's workgroup scheduling model known as a \emph{forward progress guarantee}.
+Atomics enable coherent communication between workgroups, while the \emph{forward progress guarantee} (FPG) ensures that all active workgroups will eventually make progress towards termination, precluding scenarios where a workgroup is scheduled in a manner that starves another workgroup of execution time.
+
+ In \emph{Chained-Scan}, workgroups operate within a producer-consumer framework: each workgroup produces the reduction of its assigned work tile, and every workgroup (except the first) must consume the reductions produced by all preceding workgroups to complete the scan.
+To guarantee that no workgroup begins scanning a tile for which the preceding reduction may be unavailable, \emph{Chained-Scan} assigns work tiles using atomic increment operations rather than assignment based on workgroup index.
+Once a workgroup acquires its tile, it computes the tile’s reduction, atomically posts the result into global memory, and then waits for preceding reductions to appear in global memory.
+It is this waiting phase that necessitates FPG.
+Because workgroups do not execute in physical lockstep (obvious?), some may begin waiting before others have finished their calculations.
+Without FPG, there is a risk that a workgroup which finishes early could remain indefinitely scheduled in a waiting state, blocking progress by the workgroup responsible for producing the needed reduction and ultimately causing a deadlock.
+
+ As an illustrative example, consider a scan operation which requires exactly two workgroups, running on a GPU with a single multiprocessor.
+While this multiprocessor can host both workgroups, it contains only a single scheduling unit, which lacks FPG.
+In this scenario, both workgroups launch and acquire their respective work tiles, but the workgroup responsible for \emph{Work Tile 1} finishes first and begins waiting on \emph{Work Tile 0}.
+Because the scheduler lacks FPG, it continues scheduling \emph{Work Tile 1}, but because there is only a single scheduling unit, \emph{Work Tile 0} is never scheduled and never completes its reduction.
+As a result, \emph{Work Tile 1} spins indefinitely until \emph{timeout detection and recovery} (TDR) is triggered, crashing the host program in the process.
+
+\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
+ Although NVIDIA formalized FPG down to the subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures\cite{}---it remains absent on some contemporary hardware, most notably certain ARM-based chips, including Apple's M-series GPUs\cite{10.1145/3485508}.
+Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad\footnote{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time.
+It's also unclear why the test did not trigger Metal's timeout device recovery}, and at worst, risks the afformentioned TDR and subsequent program crash.
+This risk is further exacerbated by the fact that, no major graphics API currently provides developers an FPG hardware capability query\footnote{To the best of our knowledge D3D12, Vulkan, and Metal do not have a query for FPG}.
+As a result, when implementing scan operations in a shading environment—where cross-vendor, cross-architecture portability is essential—uncertainty surrounding FPG capability and its catastrophic failure mode compels developers to fall back to the older, slower \emph{Reduce-then-Scan} approach.
+
+\subsection{Earlier Attempts at Portability}
+
+\section{Design}
+\subsection{Goals (These subsection headers can be removed later if necessary)}
+ This work is guided by two main goals.
+First, portability: we want to develop a scan implementation that retains the benefits of the \emph{Chained-Scan} architecture but is also suitable for a diverse range of hardware vendors and achitectures, including those without FPG.
+Second, performance: we aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.
+To ground these objectives, we select the WebGPU shading language (WGSL) and the WebGPU standard as our target environment.
+WGSL is a meta-level shading language which is translated and compiled as necessary for backend graphics APIs---D3D12, Vulkan, and Metal---and as such, WGSL implementations are bound by the minimum capability across all backends.
+Because WGSL must operate within this limited capability space, it inherently embodies the challenge of portability.
+
+\subsection{Non Goals}
+Although our goal is to create a fully portable and highly performant scan implementation, we are constrained by underlying hardwares and programming models.
+As discussed in more detail in \emph{Limits on the Speed-of-Light}, not all architectures offer atomic operations that are sufficiently fast enough or scheduling models that are sufficeintly fair enough to attain speed-of-light performance, and thus we cannot guarantee such performance.
+Although we are cognisant of the risks posed by subgroup divergence, subgroup functions in shading languages do not allow explicit divergence control\footnote{For example, CUDA allows developers to explicitly provision subgroup functions with a mask of the threads that will participate in it.}, placing a solution to subgroup divergence issues outside our scope.
+Furthermore, although real-world scans may require more than 30 bits of space for their reductions, or may involve scanning across multiple struct members, WGSL’s current limitations---lack of fences and 64-bit atomics---lead us to focus on the simplest scenario that fits within the 30-bit range.
+Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained-Scan}.
+
+\subsection{Achieving Goals}
+
+\section{Implementation}
+
+\subsection{Decoupled-Fallback}
+
+\begin{algorithm}[htbp]
+  \SetAlgoLined
+  \KwIn{Input data $X$, threshold $\epsilon$}
+  \KwOut{Processed data $Y$}
+  Initialize $Y \leftarrow \emptyset$\;
+  \ForEach{$x \in X$}{
+      \If{$x > \epsilon$}{
+          Add $x$ to $Y$\;
+      }
+  }
+  \Return{$Y$}\;
+  \caption{Example Algorithm}
+  \label{alg:example}
+  \end{algorithm}
+\section{Experimental Setup}
+
+\section{Results and Analysis}
+
+\section{Discussion}
+
+\subsection{Tradeoffs}
+
+\subsection{Limits on Speed-of-Light Performance}
+ There are a number of factors which may preclude speed-of-light performance, but most salient is the fairness of the scheduling model.
+In \emph{Decoupled-Fallback}, a work tile which is late due to unfairness is indistinguishable from one which is deadlocking, so as a scheduler becomes increasingly unfair, it incurs an increasing number of redundant fallbacks.
+Consider a hardware with a workgroup occupancy denoted by $o$, and let $f$ represent the probability that a fallback operation occurs at a particular lookback step.
+Because the number of lookback steps is bounded by $o$, the expected number of fallback operations is approximately $fo$.
+This results in a factor of $fo$ increase in global memory reads and a factor of $fo\log{n}$ increase in work.
+
+ This increased sensitivity to fairness exacerabtes existing issues which negatively impact the performance of previous scan architectures.
+On hardware that lacks sufficient compute power to be memory-bandwidth bound, the additional work incurred by redundant fallback reductions is particularly deleterious.
+High atomic update latency further compounds the problem: as inter-workgroup communication time grows, the minimum delay before updates become visible to dependent workgroups also increases.
+This, in turn, raises the number of lookback steps that may be needed and potentially leads to redundant fallbacks.
+
+ Fairness has been a priority in NVIDIA hardware since at least the Tesla architecture\cite{}, but the extent to which it is implemented and prioritized by other vendors remains unclear.
+While our results on the least fair hardware demonstrate that \emph{Decoupled-Fallback} remains performant, this performance may not extend to scans involving more complex binary reduction operators or less capable hardware.
 \begin{acks}

### 2024-12-23 paper.tex
Add background work on parallel prefix circuits. Add initial work on intra-workgroup depth and size.

@@
 \usepackage[ruled,vlined]{algorithm2e}
+\usepackage{amsmath}
+\usepackage{mathtools}
+
 \begin{document}
@@
  Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency.
-GPU programming exhibits both single-instruction, multiple-thread (SIMT) and single-program, multiple-data (SPMD) behaviors.
-When \emph{launched}\footnote{Also referred to as \emph{dispatched}.}, a GPU program—known as a \emph{kernel}—is replicated across independent execution contexts, referred to as threads.
-Threads are organized into groups called \emph{workgroups}\footnote{Also referred to as \emph{Cross-Thread-Arrays}, \emph{Thread-Blocks}, \emph{Blocks}, \emph{Thread-Groups}, or \emph{Groups}.}, which are further divided into smaller units known as \emph{subgroups}\footnote{Also referred to as \emph{Warps}, \emph{Waves}, or \emph{SIMD-Groups}.}.
+Due to their specialized nature, GPUs function as co-processors to the CPU, requiring GPU programs—known as \emph{kernels} or \emph{shaders}—to be \emph{launched} or \emph{dispatched} heterogeneously by the CPU.
+Upon launch, a kernel is replicated across independent execution contexts, called threads.
+These threads are first organized into smaller units known as \emph{subgroups}\footnote{Also referred to as \emph{Warps}, \emph{Waves}, or \emph{SIMD-Groups}.}, which are then grouped into larger units called \emph{workgroups}\footnote{Also referred to as \emph{Cross-Thread-Arrays}, \emph{Thread-Blocks}, \emph{Blocks}, \emph{Thread-Groups}, or \emph{Groups}.}.
 This thread hierarchy mirrors the GPU memory hierarchy: individual threads access private, high-speed registers; threads within a workgroup share low-latency shared memory\footnote{Also referred to as \emph{Local Data Storage}, \emph{Group-Shared Memory}, (Check metal terminology).}; and all threads have access to high-bandwidth but higher-latency global memory.
 
- While threads within the same workgroup share resources, a workgroup is a logical unit of execution, not a physical processor.
-Instead, a GPU's underlying multiprocessors dynamically map and schedule the workgroups.
-This abstraction enables the kernel to be executed on GPUs with varying hardware capabilities, such as different numbers of multiprocessors or multiprocessors with differing execution resources.
+ While threads within the same workgroup can synchronize with each other using barrier primitives and can communicate through shared memory, a workgroup is a logical unit of execution, not a physical processor.
+Instead, workgroups are dynamically mapped and scheduled to a GPU's underlying \emph{multiprocessors}.
+This abstraction enables the kernel to execute on GPUs with varying hardware capabilities, such as different numbers of multiprocessors or multiprocessors with differing execution resources.
 A single multiprocessor can host multiple workgroups; we define the number of active subgroups on a multiprocessor as the \emph{subgroup occupancy}, and the number of active workgroups across the entire GPU as the \emph{workgroup occupancy}.
-High subgroup occupancy is generally desirable as it hides memory and execution latency by ensuring sufficient warps are available for scheduling while others wait for data.
+Generally speaking, high subgroup occupancy is desirable because multiprocessors are specifically designed to context switch between subgroups, keeping execution units busy and masking memory and execution latency.
 
+ GPUs exhibit both single-instruction, multiple-thread (SIMT) and single-program, multiple-data (SPMD) behaviors.
 A multiprocessor schedules and executes threads by subgroup.
@@
 
- As we will discuss in the seciton, (insert section here), how a multiprocessor chooses \emph{which} out of the currently occupying subgroups to schedule can have a profound impact on both the performance and the correct execution of kernels.
+ Most relevant to this work is the scheduling behavior of the GPU.
+Scheduling is divided into a workgroup-level scheduler\footnote{For example, NVIDIA's \emph{GigaThread Engine} or AMD's \emph{Asynchronous Compute Engines}}, which manages kernel launches and maps workgroups to multiprocessors, and the aforementioned subgroup scheduler, responsible for selecting which of the currently resident subgroups for execution.
+Although contemporary GPU hardware supports workgroup preemption to prioritize real-time tasks or manage multi-process workloads\cite{}(NVIDIA Volta,AMD RDNA), this capability does not extend to the GPU programming model.
+Once a workgroup begins execution on a multiprocessor, it must run to completion and cannot context switch.
+At the subgroup scheduling level, \emph{fairness}—how evenly execution resources are distributed among threads—and \emph{progress guarantees}—ensuring that all threads or subgroups eventually make forward progress—are the most salient issues.
+As we will discuss in \emph{Why does \emph{Chained-Scan} Rely on Forward Progress Guarantees} (latex section link here), these factors have profound implications for kernel performance and correctness.
+
+\subsection{The Scan Primitive}
+The study of \emph{parallel prefix computation networks} traces back to the design of carry-lookahead adder circuits and earlier work\cite{10.1145/322217.322232, 5219801}.
+The term \emph{scan}, introduced by Iverson in the APL programming language, is often used interchangeably with \emph{prefix computation}\cite{IVERSON}.
+A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \).
+The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \).
+Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid\cite{}.
+In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence.
+If the subset includes the \( n \)-th element, it is called \textit{inclusive}; if it excludes the \( n \)-th element, it is called \textit{exclusive}.
+The most common scan type is the prefix sum, where \( \oplus \) is addition.
+For example:
+\[
+x = [x_1, x_2, x_3, \dots, x_n]
+\]
+\[
+\text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
+\]
+
+More concretely, let \( y = [1, 1, 1, 1, 1] \). The inclusive scan of \( y \) using addition is:
+\[
+\text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
+\]
+
+ Scans are characterized by their \emph{depth} $d$, the number of parallel work steps required, and \emph{size} $s$, the total number of $\oplus$ operations performed.
+Snir proved that for a network size $n$, depth and size are related by $s + d \ge 2n - 2$ \cite{}.
+This equation highlights a fundamental tradeoff: as depth decreases to improve parallelism, the size (i.e., total work) increases proportionally, and vice versa.
+Likewise, Harris introduces a circuit-based, three-dimensional taxonomy of scans, relating \emph{fanout} $f$, \emph{wire-tracks} $t$, and \emph{depth} by $f + t + d = \log_2n - 1$ \cite{}.
+Here, fanout denotes the number of simultaneous outputs from a single node, while wire-tracks represent the number of parallel connections, broadly analogous to overall network size.
+The classic scans map to maximal points in this three-dimensional space: maximal tracks Kogge-Stone/Hillis-Steele\cite{}, maximal fanout Sklansky\cite{}, and maximal depth Brent-Kung\cite{}.
+Furthermore, the associative property of $\oplus$ enables the hybridization of different scan network structures.
+In Harris’s taxonomy, these hybrid scans correspond to intermediate points between the maximal designs: depth-fanout Ladner-Fischer\cite{}, depth-tracks Hans-Carlson\cite{}, and fanout-tracks Knowles\cite{}.
+
+ \begin{table}[h]
+  \centering
+  \caption{Comparison of Different Scan Networks}
+  \label{tab:scan_comparison_exact}
+  \begin{tabular}{|l|c|c|}
+  \hline
+  \textbf{Network}       & \textbf{Size}        & \textbf{Miminum Depth Variant} \\ \hline
+  \textbf{Serial}           & $n$           & $n - 1$            \\ \hline
+  \textbf{Kogge--Stone}     & $O(n \log n)$ & $\log_2(n)$        \\ \hline
+  \textbf{Sklansky}         & $O(n \log n)$ & $\log_2(n)$        \\ \hline
+  \textbf{Brent--Kung}      & $O(n)$        & $2 \log_2(n) - 1$  \\ \hline
+  \textbf{Han--Carlson}     & $O(n)$        & $\log_2(n) + \Big\lfloor \tfrac{\log_2(n)}{2} \Big\rfloor - 1$ \\ \hline
+  \textbf{Knowles}          & $O(n)$        & $\log_2(n)$        \\ \hline
+  \textbf{Ladner--Fischer}  & $O(n)$        & $\log_2(n)$        \\ \hline
+  \end{tabular}
+\end{table}
+ \footnote{Hybrid scans offer varying degrees of hybridization; for simplicity, we report only the size of the minimal-depth variant.}
+ For our purposes, we are primarily concerned about depth and size.
+Notably, Fich proved that with unbounded fanout, the Ladner-Fischer network has both minimal depth, $\lceil \log_2n \rceil$ and optimal size, $O(n)$ \cite{}.
+While circuit engineering tends to favor minimal depth networks---Kogge-Stone, Hans-Carlson---for minimal latency, the task sizes of scan algorithms frequently exceed the number of available physical processors.
+This makes minimal-size networks, like Brent–Kung and Sklansky, more practical for algorithmic implementation.
+Within computer science, Blelloch formalized scan in the PRAM model and popularized their use as a parallel compute primitive \cite{}.
+Horn was the first to adapt scan for GPU use and Sengupta et al.
+were the first to adapt scan to CUDA \cite{}.
+
+\subsection{GPU Specific Scan Adaptation}
 
 \subsection{Evolution of Inter-Workgroup Scan Architectures}
+
 \subsubsection{Scan-then-Propagate}
@@
 
-\subsection{Why does \emph{Chained-Scan} Rely on Forward Progress Guarantees?}
+ StreamScan
+
+ DecoupledLookback
+
+\subsection{Missing Synchronization Primitive?}
  In a scan, the reduction at each element is dependent on the reduction of preceding elements.
@@
 
+\subsection{Why does \emph{Chained-Scan} Rely on Forward Progress Guarantees?}
  The \emph{Chained-Scan} architecture operates in the middle ground between the absence of inter-workgroup synchronization and formal inter-workgroup synchronization support.
@@
   \Return{$Y$}\;
-  \caption{Example Algorithm}
+  \caption{Decoupled Lookback with Decoupled Fallback}
   \label{alg:example}
 \end{algorithm}
+
+\subsection{Intra-Workgroup Implementation}
+ (TODO: Try Reduce-Then-Scan on low spec hardware.)
+ Our intra-workgroup implementation is a variation of the \emph{3D-Matrix Scan} (3DMS) introduced by Yan et al.\cite{}, with adaptations to enhance portability.
+Similar to 3DMS, we split the work tile into non-overlapping blocks, which are distributed to each subgroup.
+However, unlike CUDA, WGSL lacks a \emph{volatile} qualifier for shared memory, which prevents us from loading items into shared memory and transposing them without inserting a barrier.
+However, placing a barrier disrupts latency hiding by separating the high-latency global memory load from the subsequent transpose and scan computation.
+To address this, we load data directly into registers and \emph{vectorize} the input.
+This approach preserves a \emph{coalesced} but vector-strided loading pattern.
+For each vector, a thread performs a serial scan across the vector, then participates in a subgroup scan.
+Although this method results in less efficient subgroup scan computation, $\frac{p}{vs}$ subgroup scans versus one, it elides the barrier and significantly reduces the kernel's shared memory footprint.
+Transposing in shared memory would require a work-tile-sized shared memory allocation, whereas our approach relies more heavily on registers, which are generally more abundant than shared memory, especially on low-spec hardware.
+As a result, we consider this tradeoff worthwhile.
+
+ Once completed, each subgroup posts its reduction into shared memory, then participates in a workgroup-wide scan.
+Because the WebGPU specification supports subgroup sizes $s \text{ where } s = 2^k, \, k \in [2, 7]$, and on some hardware, the subgroup size can vary between kernel launches, it is imperative to use a scan implementation which can accomodate all subgroup sizes.
+(More) We use a log base $s$ Ladner-Fischer scan with Merrill-Grimshaw conflict avoidance that embeds Kogge-Stone subgroup scans\cite{}.
+This network has several advantages: minimal depth $log_s n$; asymptotically optimal $O(n)$ size; $s$-way conflict (seems bad, but actually turns into broadcast?).
+
+ \begin{algorithm}[htbp]
+  \SetAlgoLined
+  \KwIn{Array to scan $x$, Workgroup size $W$, Subgroup size $S$}
+  \KwOut{Inclusive scanned array $x$}
+
+  $spine\_length \gets W / S$\;
+  $alignment \gets 1 << \text{divRoundUp}(\log_2(spine\_length), \log_2(S)) * \log_2(S)$\;
+
+  $iteration\_offset \gets 0$\;
+  $stride \gets 1$\;
+
+  \ForEach{$thread\_id$ \textbf{in} $W$ \textbf{in parallel}}{
+    \For{$j \gets S$ \KwTo $alignment$ \textbf{with} $j \gets j * S$}{
+
+      $spine\_index \gets ((thread\_id + iteration\_offset) * stride) - iteration\_offset$\;
+
+      \If{$spine\_index < spine\_length$}{
+        $x[spine\_index] \gets \text{subgroupInclusiveScan}(x[spine\_index])$\;
+      }
+
+      \textbf{barrier()}\;
+
+      \eIf{$j \neq S$}{
+        $reduced\_stride \gets j / stride$\;
+        $fanout\_index \gets thread\_id + reduced\_stride$\;
+
+        $cond1 \gets fanout\_index < spine\_length$\;
+        $cond2 \gets (fanout\_index \& (j - 1)) \geq reduced\_stride$\;
+        $cond3 \gets ((fanout\_index + 1) \& (reduced\_stride - 1)) \neq 0$\;
+
+        \If{$cond1 \textbf{ \&\& } cond2 \textbf{ \&\& } cond3$}{
+          $x[fanout\_index] \gets x[fanout\_index] + x[((fanout\_index / stride) * stride) - 1]$\;
+        }
+      }{
+        $iteration\_offset \gets iteration\_offset + 1$\;
+      }
+      $stride \gets stride * S$\;
+    }
+  }
+  \Return{$x$}\;
+  \caption{Subgroup-Size-Agnostic Scan (SAMPLE NO BANK CONFLICT AVOIDANCE)}
+  \label{alg:example}
+ \end{algorithm}
+
+ \subsubsection{Size (Work) Efficiency}
+ Recall that for a subgroup of size $s$, the Kogge-Stone scan has a size complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
+ \begin{equation}
+  \begin{aligned}[b]
+      \text{Work} &= \underbrace{\vphantom{\sum_{k=0}^{\lceil \log_s n \rceil}}n}_{\text{fanout}} 
+      + 
+      \underbrace{\sum_{k=1}^{\lceil \log_s n \rceil}}_{\text{iterations}}
+      \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}\frac{n}{s^k}}_{\text{calls per iteration}} 
+      \cdot 
+      \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}s \log_2 s}_{\text{work per call}} \\
+      &= n + s \log_2 s \cdot \sum_{k=1}^{\lceil \log_s n \rceil} \frac{n}{s^k} \\
+      &= O\left(n + n \log_2 s\right) \\
+      &= O(n \log_2 s).
+  \end{aligned}
+ \end{equation}
+
+ \subsubsection{Depth}
+ Recall that a Kogge-Stone scan has a depth of $\log_2 s$ Thus, the total depth of the scan is:
+ \begin{equation}
+  \begin{aligned}[b]
+      \text{Depth} &= \underbrace{\lceil \log_s n \rceil}_{\text{loop iterations}} 
+      \cdot \underbrace{\log_2 s}_{\text{depth per iteration}} \\
+      &= \left\lceil \frac{\log_2 n}{\log_2 s} \right\rceil \cdot \log_2 s \\
+      &= \log_2 n + O(1).
+  \end{aligned}
+ \end{equation}
+ Minimal depth is preserved regardless of the subgroup size.
+However, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a workgroup-wide barrier at each iteration.
+Consequently, while the theoretical depth remains constant across subgroup sizes, smaller subgroups tend to perform worse in practice due to the increased relative cost of synchronization and the overhead of additional iterations in the main loop.
+\subsection{Full Algorithm}
+
 \section{Experimental Setup}

### 2024-12-26 paper.tex
typography, esp. using booktabs for tables

@@
 Scheduling is divided into a workgroup-level scheduler\footnote{For example, NVIDIA's \emph{GigaThread Engine} or AMD's \emph{Asynchronous Compute Engines}}, which manages kernel launches and maps workgroups to multiprocessors, and the aforementioned subgroup scheduler, responsible for selecting which of the currently resident subgroups for execution.
-Although contemporary GPU hardware supports workgroup preemption to prioritize real-time tasks or manage multi-process workloads\cite{}(NVIDIA Volta,AMD RDNA), this capability does not extend to the GPU programming model.
+Although contemporary GPU hardware supports workgroup preemption to prioritize real-time tasks or manage multi-process workloads~\cite{}(NVIDIA Volta,AMD RDNA), this capability does not extend to the GPU programming model.
 Once a workgroup begins execution on a multiprocessor, it must run to completion and cannot context switch.
@@
 \subsection{The Scan Primitive}
-The study of \emph{parallel prefix computation networks} traces back to the design of carry-lookahead adder circuits and earlier work\cite{10.1145/322217.322232, 5219801}.
-The term \emph{scan}, introduced by Iverson in the APL programming language, is often used interchangeably with \emph{prefix computation}\cite{IVERSON}.
+The study of \emph{parallel prefix computation networks} traces back to the design of carry-lookahead adder circuits and earlier work~\cite{10.1145/322217.322232, 5219801}.
+The term \emph{scan}, introduced by Iverson in the APL programming language, is often used interchangeably with \emph{prefix computation}~\cite{IVERSON}.
 A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \).
 The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \).
-Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid\cite{}.
+Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid~\cite{}.
 In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence.
-If the subset includes the \( n \)-th element, it is called \textit{inclusive}; if it excludes the \( n \)-th element, it is called \textit{exclusive}.
+If the subset includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}.
 The most common scan type is the prefix sum, where \( \oplus \) is addition.
@@
 Scans are characterized by their \emph{depth} $d$, the number of parallel work steps required, and \emph{size} $s$, the total number of $\oplus$ operations performed.
-Snir proved that for a network size $n$, depth and size are related by $s + d \ge 2n - 2$ \cite{}.
+Snir proved that for a network size $n$, depth and size are related by $s + d \ge 2n - 2$~\cite{}.
 This equation highlights a fundamental tradeoff: as depth decreases to improve parallelism, the size (i.e., total work) increases proportionally, and vice versa.
-Likewise, Harris introduces a circuit-based, three-dimensional taxonomy of scans, relating \emph{fanout} $f$, \emph{wire-tracks} $t$, and \emph{depth} by $f + t + d = \log_2n - 1$ \cite{}.
+Likewise, Harris introduces a circuit-based, three-dimensional taxonomy of scans, relating \emph{fanout} $f$, \emph{wire-tracks} $t$, and \emph{depth} by $f + t + d = \log_2n - 1$~\cite{}.
 Here, fanout denotes the number of simultaneous outputs from a single node, while wire-tracks represent the number of parallel connections, broadly analogous to overall network size.
-The classic scans map to maximal points in this three-dimensional space: maximal tracks Kogge-Stone/Hillis-Steele\cite{}, maximal fanout Sklansky\cite{}, and maximal depth Brent-Kung\cite{}.
+The classic scans map to maximal points in this three-dimensional space: maximal tracks Kogge-Stone/Hillis-Steele~\cite{}, maximal fanout Sklansky~\cite{}, and maximal depth Brent-Kung~\cite{}.
 Furthermore, the associative property of $\oplus$ enables the hybridization of different scan network structures.
-In Harris’s taxonomy, these hybrid scans correspond to intermediate points between the maximal designs: depth-fanout Ladner-Fischer\cite{}, depth-tracks Hans-Carlson\cite{}, and fanout-tracks Knowles\cite{}.
+In Harris’s taxonomy, these hybrid scans correspond to intermediate points between the maximal designs: depth-fanout Ladner-Fischer~\cite{}, depth-tracks Hans-Carlson~\cite{}, and fanout-tracks Knowles~\cite{}.
 
@@
   \label{tab:scan_comparison_exact}
-  \begin{tabular}{|l|c|c|}
-  \hline
-  \textbf{Network}       & \textbf{Size}        & \textbf{Miminum Depth Variant} \\ \hline
-  \textbf{Serial}           & $n$           & $n - 1$            \\ \hline
-  \textbf{Kogge--Stone}     & $O(n \log n)$ & $\log_2(n)$        \\ \hline
-  \textbf{Sklansky}         & $O(n \log n)$ & $\log_2(n)$        \\ \hline
-  \textbf{Brent--Kung}      & $O(n)$        & $2 \log_2(n) - 1$  \\ \hline
-  \textbf{Han--Carlson}     & $O(n)$        & $\log_2(n) + \Big\lfloor \tfrac{\log_2(n)}{2} \Big\rfloor - 1$ \\ \hline
-  \textbf{Knowles}          & $O(n)$        & $\log_2(n)$        \\ \hline
-  \textbf{Ladner--Fischer}  & $O(n)$        & $\log_2(n)$        \\ \hline
+  \begin{tabular}{lcc}
+    \toprule
+    \textbf{Network}         & \textbf{Size} & \textbf{Miminum Depth Variant}                                 \\ \midrule
+    \textbf{Serial}          & $n$           & $n - 1$                                                        \\
+    \textbf{Kogge--Stone}    & $O(n \log n)$ & $\log_2(n)$                                                    \\
+    \textbf{Sklansky}        & $O(n \log n)$ & $\log_2(n)$                                                    \\
+    \textbf{Brent--Kung}     & $O(n)$        & $2 \log_2(n) - 1$                                              \\
+    \textbf{Han--Carlson}    & $O(n)$        & $\log_2(n) + \Big\lfloor \tfrac{\log_2(n)}{2} \Big\rfloor - 1$ \\
+    \textbf{Knowles}         & $O(n)$        & $\log_2(n)$                                                    \\
+    \textbf{Ladner--Fischer} & $O(n)$        & $\log_2(n)$                                                    \\ \bottomrule
   \end{tabular}
@@
 For our purposes, we are primarily concerned about depth and size.
-Notably, Fich proved that with unbounded fanout, the Ladner-Fischer network has both minimal depth, $\lceil \log_2n \rceil$ and optimal size, $O(n)$ \cite{}.
+Notably, Fich proved that with unbounded fanout, the Ladner-Fischer network has both minimal depth, $\lceil \log_2n \rceil$ and optimal size, $O(n)$~\cite{}.
 While circuit engineering tends to favor minimal depth networks---Kogge-Stone, Hans-Carlson---for minimal latency, the task sizes of scan algorithms frequently exceed the number of available physical processors.
 This makes minimal-size networks, like Brent–Kung and Sklansky, more practical for algorithmic implementation.
-Within computer science, Blelloch formalized scan in the PRAM model and popularized their use as a parallel compute primitive \cite{}.
-Horn was the first to adapt scan for GPU use and Sengupta et al.
-were the first to adapt scan to CUDA \cite{}.
+Within computer science, Blelloch formalized scan in the PRAM model and popularized their use as a parallel compute primitive~\cite{}.
+Horn was the first to adapt scan for GPU use and Sengupta et al.\ were the first to adapt scan to CUDA~\cite{}.
 
@@
 Historically, hardware vendors have only recently begun to formally support inter-workgroup synchronization.
-Prior to the introduction of \emph{thread block clusters} on NVIDIA's Hopper architecture\cite{}, we are unaware of any GPU programming framework\footnote{We are not aware of any prior inter-workgroup level barrier primitive in CUDA, OpenCL, GLSL, HLSL, or Metal.} that provided inter-workgroup synchronization primitives.
-Instead, the earlier \emph{Scan-then-Propagate} and \emph{Reduce-then-Scan} architectures rely on kernel launches to act as inter-workgroup synchronization points\cite{}.
+Prior to the introduction of \emph{thread block clusters} on NVIDIA's Hopper architecture~\cite{}, we are unaware of any GPU programming framework\footnote{We are not aware of any prior inter-workgroup level barrier primitive in CUDA, OpenCL, GLSL, HLSL, or Metal.} that provided inter-workgroup synchronization primitives.
+Instead, the earlier \emph{Scan-then-Propagate} and \emph{Reduce-then-Scan} architectures rely on kernel launches to act as inter-workgroup synchronization points~\cite{}.
 
@@
 \subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
- Although NVIDIA formalized FPG down to the subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures\cite{}---it remains absent on some contemporary hardware, most notably certain ARM-based chips, including Apple's M-series GPUs\cite{10.1145/3485508}.
+Although NVIDIA formalized FPG down to the subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---it remains absent on some contemporary hardware, most notably certain ARM-based chips, including Apple's M-series GPUs~\cite{10.1145/3485508}.
 Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad\footnote{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time.
@@
 (TODO: Try Reduce-Then-Scan on low spec hardware.)
- Our intra-workgroup implementation is a variation of the \emph{3D-Matrix Scan} (3DMS) introduced by Yan et al.\cite{}, with adaptations to enhance portability.
+Our intra-workgroup implementation is a variation of the \emph{3D-Matrix Scan} (3DMS) introduced by Yan et al.~\cite{}, with adaptations to enhance portability.
 Similar to 3DMS, we split the work tile into non-overlapping blocks, which are distributed to each subgroup.
@@
 Because the WebGPU specification supports subgroup sizes $s \text{ where } s = 2^k, \, k \in [2, 7]$, and on some hardware, the subgroup size can vary between kernel launches, it is imperative to use a scan implementation which can accomodate all subgroup sizes.
-(More) We use a log base $s$ Ladner-Fischer scan with Merrill-Grimshaw conflict avoidance that embeds Kogge-Stone subgroup scans\cite{}.
+(More) We use a log base $s$ Ladner-Fischer scan with Merrill-Grimshaw conflict avoidance that embeds Kogge-Stone subgroup scans~\cite{}.
 This network has several advantages: minimal depth $log_s n$; asymptotically optimal $O(n)$ size; $s$-way conflict (seems bad, but actually turns into broadcast?).
@@
   \begin{aligned}[b]
-      \text{Work} &= \underbrace{\vphantom{\sum_{k=0}^{\lceil \log_s n \rceil}}n}_{\text{fanout}} 
+    \text{Work} & = \underbrace{\vphantom{\sum_{k=0}^{\lceil \log_s n \rceil}}n}_{\text{fanout}}
     +
@@
     \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}s \log_2 s}_{\text{work per call}} \\
-      &= n + s \log_2 s \cdot \sum_{k=1}^{\lceil \log_s n \rceil} \frac{n}{s^k} \\
-      &= O\left(n + n \log_2 s\right) \\
-      &= O(n \log_2 s).
+                & = n + s \log_2 s \cdot \sum_{k=1}^{\lceil \log_s n \rceil} \frac{n}{s^k}       \\
+                & = O\left(n + n \log_2 s\right)                                                 \\
+                & = O(n \log_2 s).
   \end{aligned}
@@
   \begin{aligned}[b]
-      \text{Depth} &= \underbrace{\lceil \log_s n \rceil}_{\text{loop iterations}} 
+    \text{Depth} & = \underbrace{\lceil \log_s n \rceil}_{\text{loop iterations}}
     \cdot \underbrace{\log_2 s}_{\text{depth per iteration}}                           \\
-      &= \left\lceil \frac{\log_2 n}{\log_2 s} \right\rceil \cdot \log_2 s \\
-      &= \log_2 n + O(1).
+                 & = \left\lceil \frac{\log_2 n}{\log_2 s} \right\rceil \cdot \log_2 s \\
+                 & = \log_2 n + O(1).
   \end{aligned}
@@
 
- Fairness has been a priority in NVIDIA hardware since at least the Tesla architecture\cite{}, but the extent to which it is implemented and prioritized by other vendors remains unclear.
+Fairness has been a priority in NVIDIA hardware since at least the Tesla architecture~\cite{}, but the extent to which it is implemented and prioritized by other vendors remains unclear.
 While our results on the least fair hardware demonstrate that \emph{Decoupled-Fallback} remains performant, this performance may not extend to scans involving more complex binary reduction operators or less capable hardware.

### 2024-12-28 paper.tex
Add work on global scan evolution, shorten scan primitive section.

@@
 \section{Introduction}
-Do. this. last.
+
+\subsection{Goals of a Scan Implementation}
+\begin{itemize}
+  \item \textbf{Minimal global memory access:} the scan should minimize global memory access to the theoretical minimum of $2n$.
+  \item \textbf{Fully saturates global memory bandwidth:} the scan should be capable of fully saturating the global memory bandwidth.
+  \item \textbf{Portability:} the scan should be portable across hardware vendors and architectures. 
+\end{itemize}
 
@@
 At the subgroup scheduling level, \emph{fairness}—how evenly execution resources are distributed among threads—and \emph{progress guarantees}—ensuring that all threads or subgroups eventually make forward progress—are the most salient issues.
-As we will discuss in \emph{Why does \emph{Chained-Scan} Rely on Forward Progress Guarantees} (latex section link here), these factors have profound implications for kernel performance and correctness.
+As we will discuss in the following section, these factors have profound implications for kernel performance and correctness.
 
 \subsection{The Scan Primitive}
-The study of \emph{parallel prefix computation networks} traces back to the design of carry-lookahead adder circuits and earlier work~\cite{10.1145/322217.322232, 5219801}.
-The term \emph{scan}, introduced by Iverson in the APL programming language, is often used interchangeably with \emph{prefix computation}~\cite{IVERSON}.
+The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond\cite{10.1145/322217.322232, 5219801}.
 A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \).
 The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \).
-Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid~\cite{}.
+Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid\cite{}.
 In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence.
-If the subset includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}.
+If the subset includes the \( n \)-th element, it is called \textit{inclusive}; if it excludes the \( n \)-th element, it is called \textit{exclusive}.
 The most common scan type is the prefix sum, where \( \oplus \) is addition.
@@
 \[
-  x = [x_1, x_2, x_3, \dots, x_n]
+  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1] 
 \]
@@
 \]
-
-More concretely, let \( y = [1, 1, 1, 1, 1] \). The inclusive scan of \( y \) using addition is:
 \[
@@
 
-Scans are characterized by their \emph{depth} $d$, the number of parallel work steps required, and \emph{size} $s$, the total number of $\oplus$ operations performed.
-Snir proved that for a network size $n$, depth and size are related by $s + d \ge 2n - 2$~\cite{}.
-This equation highlights a fundamental tradeoff: as depth decreases to improve parallelism, the size (i.e., total work) increases proportionally, and vice versa.
-Likewise, Harris introduces a circuit-based, three-dimensional taxonomy of scans, relating \emph{fanout} $f$, \emph{wire-tracks} $t$, and \emph{depth} by $f + t + d = \log_2n - 1$~\cite{}.
-Here, fanout denotes the number of simultaneous outputs from a single node, while wire-tracks represent the number of parallel connections, broadly analogous to overall network size.
-The classic scans map to maximal points in this three-dimensional space: maximal tracks Kogge-Stone/Hillis-Steele~\cite{}, maximal fanout Sklansky~\cite{}, and maximal depth Brent-Kung~\cite{}.
-Furthermore, the associative property of $\oplus$ enables the hybridization of different scan network structures.
-In Harris’s taxonomy, these hybrid scans correspond to intermediate points between the maximal designs: depth-fanout Ladner-Fischer~\cite{}, depth-tracks Hans-Carlson~\cite{}, and fanout-tracks Knowles~\cite{}.
-
-\begin{table}[h]
-  \centering
-  \caption{Comparison of Different Scan Networks}
-  \label{tab:scan_comparison_exact}
-  \begin{tabular}{lcc}
-    \toprule
-    \textbf{Network}         & \textbf{Size} & \textbf{Miminum Depth Variant}                                 \\ \midrule
-    \textbf{Serial}          & $n$           & $n - 1$                                                        \\
-    \textbf{Kogge--Stone}    & $O(n \log n)$ & $\log_2(n)$                                                    \\
-    \textbf{Sklansky}        & $O(n \log n)$ & $\log_2(n)$                                                    \\
-    \textbf{Brent--Kung}     & $O(n)$        & $2 \log_2(n) - 1$                                              \\
-    \textbf{Han--Carlson}    & $O(n)$        & $\log_2(n) + \Big\lfloor \tfrac{\log_2(n)}{2} \Big\rfloor - 1$ \\
-    \textbf{Knowles}         & $O(n)$        & $\log_2(n)$                                                    \\
-    \textbf{Ladner--Fischer} & $O(n)$        & $\log_2(n)$                                                    \\ \bottomrule
-  \end{tabular}
-\end{table}
-\footnote{Hybrid scans offer varying degrees of hybridization; for simplicity, we report only the size of the minimal-depth variant.}
-For our purposes, we are primarily concerned about depth and size.
-Notably, Fich proved that with unbounded fanout, the Ladner-Fischer network has both minimal depth, $\lceil \log_2n \rceil$ and optimal size, $O(n)$~\cite{}.
-While circuit engineering tends to favor minimal depth networks---Kogge-Stone, Hans-Carlson---for minimal latency, the task sizes of scan algorithms frequently exceed the number of available physical processors.
-This makes minimal-size networks, like Brent–Kung and Sklansky, more practical for algorithmic implementation.
-Within computer science, Blelloch formalized scan in the PRAM model and popularized their use as a parallel compute primitive~\cite{}.
-Horn was the first to adapt scan for GPU use and Sengupta et al.\ were the first to adapt scan to CUDA~\cite{}.
-
-\subsection{GPU Specific Scan Adaptation}
+Due to its significance in circuits and as a fundamental algorithmic primitive, scan enjoys a rich literature in both electrical engineering and computer science~\cite{}.
+As comprehensive taxonomies of scan networks are well-documented, our focus is on exploring the progression of inter-workgroup (global) scan strategies and their fitness relative to our ideal architecture.
 
 \subsection{Evolution of Inter-Workgroup Scan Architectures}
+Prior to the introduction of shared memory, early GPU scan implementations~\cite{Horn, Hensley, Greß, Sengupta} operated directly on global memory and performed a single global scan where each level of depth necessitated a separate kernel launch.
+The introduction of shared memory and dedicated compute environments resulted in the first implementations to resemble contemporary designs, combining coarse-grained inter-workgroup strategies with fine-grained intra-workgroup strategies.
+\emph{Work tile} partitioning was introduced to divide inputs into workgroup-consumable chunks that fit within the limited size of shared memory.
+However, as a natural result of this partitioning, a serial dependency is created between the work tiles.
+In this section, we review the three prominent global scan strategies—\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}—examining how they navigate this inter-workgroup dependency and evaluating their effectiveness in minimizing global memory access and maximizing memory bandwidth utilization in alignment with our ideal architecture goals.
 
 \subsubsection{Scan-then-Propagate}
+Introduced by Sengupta et al.\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate} \cite{GPUGems3, Sengupta2008} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases:
+\begin{enumerate} 
+  \item \textbf{Intra-Workgroup (Local) Scan:} Each workgroup performs an inclusive scan on its assigned work-tile partition of the input data.
+The results, including both the local scan output and the reduction of each workgroup, are written back to global memory.
+  \item \textbf{Spine Scan:} The reductions from all workgroups, collectively referred to as \emph{the spine}, are gathered and scanned to compute the global prefix of the reductions, resolving the dependencies between workgroups; we call the result of this phase the \emph{root}.
+  \item \textbf{Propagation:} The root is propagated back to each workgroup and combined with the local scan results to compute the final scan. 
+\end{enumerate}
+Without an inter-workgroup barrier primitive, each phase is implemented as a separate kernel launch to ensure a consistent view of the reductions.
+During the scan operation, each element is read from and written to global memory once during the local scan and again during the propagation phase, contributing a total of $4n$ global memory operations.
+Although data movement is dominated by the first and third phase, Sengupta et al.'s implementation handles arbitrarily large inputs using recursion.
+For an input of size $n$ and a work tile of size $t$, this results in a recursive depth of $\lceil \log_t n \rceil$.
+At each recursive step $k$, $n/t^k$ tiles are processed, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps.
+Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.
+
 \subsubsection{Reduce-then-Scan}
+As the design of GPUs settled around the high-throughput but high-latency paradigm, reducing global memory access, rather than compute, became the prevailing goal for algorithms like scan, which are computationally light enough for their compute to be masked by global memory access.
+This led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}\cite{Merrill Grimshaw, Ha and Han, Dotsenko, Breitbart}: by writing only the work tile reduction in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated.
+The \emph{Reduce-Then-Scan} strategy operates in three phases:
+\begin{enumerate}
+  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result into global memory. 
+  \item \textbf{Spine Scan:} The reductions from all workgroups are gathered and scanned to produce the root.
+  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating the root as it writes to global memory.
+\end{enumerate}
+\emph{Reduce-then-Scan} implementations also introduced improvements to the spine phase of the scan, which coalesced around a workgroup-level \emph{raking} technique first applied to inter-workgroup scan by Merrill and Grimshaw~\cite{}.
+Instead of handling arbitrary inputs recursively, the total workgroup occupancy $o$ is determined, and the input is partitioned into non-overlapping blocks processed serially by each workgroup.
+This approach limits the recursive depth of the global scan to two and reduces the global memory movement for the spine scan from $O(n/t)$ (incurred by recursion) to $O(3o)$.
+As a result, kernel launches are limited to three and total global memory movement is reduced to $O(3n + 3o) = O(3n)$.
+
 \subsubsection{Chained-Scan}
+Continuing the trend of global data movement optimization, the most recent work on global scan has been the \emph{Chained-Scan} architecture.
+First introduced by Yan et al.
+in \emph{StreamScan}, the key innovation of GPU chained scanning lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level.
 
-StreamScan
+Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups.
+This shifts the problem from explicit synchronization to implicit synchronization via the order in which workgroups access global data.
+Each workgroup processes its assigned tile independently, waiting only for the result of the preceding reductions before continuing.
+This structure ensures that global memory traffic is minimized, as each element is read and written exactly once, achieving the theoretical minimum of $2n$.
 
-DecoupledLookback
+While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently achieve the expected throughput of $\text{memoryBandwidth}/2$ (saturate bandwidth?).
+This limitation arises because, in \emph{StreamScan}, a workgroup cannot post its reduction to global memory until it has ingested the reduction of the preceding workgroup.
+This sequential dependency creates a bottleneck, constraining the algorithm’s performance to the rate at which reductions can propagate through device memory—a process limited by the relatively high latency of GPU memory.
 
-\subsection{Missing Synchronization Primitive?}
-In a scan, the reduction at each element is dependent on the reduction of preceding elements.
-Thus, a serial dependency is created between workgroups whenever the number of elements in a scan operation exceeds the maximum size that can be processed by a single workgroup.
-Historically, hardware vendors have only recently begun to formally support inter-workgroup synchronization.
-Prior to the introduction of \emph{thread block clusters} on NVIDIA's Hopper architecture~\cite{}, we are unaware of any GPU programming framework\footnote{We are not aware of any prior inter-workgroup level barrier primitive in CUDA, OpenCL, GLSL, HLSL, or Metal.} that provided inter-workgroup synchronization primitives.
-Instead, the earlier \emph{Scan-then-Propagate} and \emph{Reduce-then-Scan} architectures rely on kernel launches to act as inter-workgroup synchronization points~\cite{}.
+Building on \emph{StreamScan}, Merrill and Garland introduced the \emph{Decoupled Lookback} technique to address the bottleneck created by these sequential dependencies.
+Unlike \emph{StreamScan}, where reductions and dependency resolution occur sequentially, \emph{Decoupled Lookback} separates these processes.In \emph{Decoupled Lookback}, each workgroup posts its reduction to device memory and then enters the “lookback” phase.
+During this phase, the workgroup performs a serial traversal backward through preceding workgroup reductions stored in device memory.
+To manage inter-workgroup synchronization, bit-packed status flags are used to indicate the state of each workgroup’s reduction.
+
+TODO Cleanup, add O(n) on global data movement during the lookback phase etc.
+
+\subsection{Ideal Scan Architecture}
+RTS
+More portable
+slower
+handling arbitrary inputs can be awkward
+
+Chained Scan
+Faster
+2n data movement + serial ordering faclitates in-situ behavior
+naturally extends to arbitrary data
+not portable, relies on something called forward progress guarantees
+
+Chained has lots of advantages, it would be great if it was portable.
 
 \subsection{Why does \emph{Chained-Scan} Rely on Forward Progress Guarantees?}
+TODO rewrite
 The \emph{Chained-Scan} architecture operates in the middle ground between the absence of inter-workgroup synchronization and formal inter-workgroup synchronization support.
@@
 \subsection{Earlier Attempts at Portability}
+Scalar fallback here.
 
@@
 \subsection{Intra-Workgroup Implementation}
-(TODO: Try Reduce-Then-Scan on low spec hardware.)
-Our intra-workgroup implementation is a variation of the \emph{3D-Matrix Scan} (3DMS) introduced by Yan et al.~\cite{}, with adaptations to enhance portability.
+Our intra-workgroup implementation is most similar to the \emph{CUDPP} scan kernel \cite{}, with adaptations to enhance portability. It consists of three phases:
+\begin{enumerate}
+  \item
+\end{enumerate}
+
 Similar to 3DMS, we split the work tile into non-overlapping blocks, which are distributed to each subgroup.
 However, unlike CUDA, WGSL lacks a \emph{volatile} qualifier for shared memory, which prevents us from loading items into shared memory and transposing them without inserting a barrier.
-However, placing a barrier disrupts latency hiding by separating the high-latency global memory load from the subsequent transpose and scan computation.
+Placing a barrier disrupts latency hiding by separating the high-latency global memory load from the subsequent transpose and scan computation.
 To address this, we load data directly into registers and \emph{vectorize} the input.
@@
 For each vector, a thread performs a serial scan across the vector, then participates in a subgroup scan.
-Although this method results in less efficient subgroup scan computation, $\frac{p}{vs}$ subgroup scans versus one, it elides the barrier and significantly reduces the kernel's shared memory footprint.
-Transposing in shared memory would require a work-tile-sized shared memory allocation, whereas our approach relies more heavily on registers, which are generally more abundant than shared memory, especially on low-spec hardware.
+Although this method results in vectors per thread $v$ subgroup scans, it elides the barrier and significantly reduces the kernel's shared memory footprint, as only the subgroup reduction is placed into shared memory.
+Transposing in shared memory requires a work-tile-sized shared memory allocation, whereas our approach relies more heavily on registers, which are generally more abundant than shared memory, especially on low-spec hardware.
+Additionally, keeping scan elements in registers avoids shared-bank conflicts, further enhancing performance.
 As a result, we consider this tradeoff worthwhile.
 
-Once completed, each subgroup posts its reduction into shared memory, then participates in a workgroup-wide scan.
-Because the WebGPU specification supports subgroup sizes $s \text{ where } s = 2^k, \, k \in [2, 7]$, and on some hardware, the subgroup size can vary between kernel launches, it is imperative to use a scan implementation which can accomodate all subgroup sizes.
-(More) We use a log base $s$ Ladner-Fischer scan with Merrill-Grimshaw conflict avoidance that embeds Kogge-Stone subgroup scans~\cite{}.
-This network has several advantages: minimal depth $log_s n$; asymptotically optimal $O(n)$ size; $s$-way conflict (seems bad, but actually turns into broadcast?).
+The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$, and on some hardware, subgroup sizes can vary between kernel launches.
+To handle this variability, we use a generalized radix-$s$ Ladner-Fischer scan network embedded with Kogge-Stone subgroup scans to perform the scan on subgroup reductions.
+The upsweep phase of the Ladner-Fischer network follows a Brent-Kung construction; assuming shared memory bank widths equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts.
+To mitigate this, we employ Merrill-Grimshaw style conflict avoidance~\cite{}.
+This network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and exposure to bank conflicts is limited to cases where $s^2 < w$, as the network collapses to a single level when $s^2 \geq w$.
 
@@
 \subsubsection{Size (Work) Efficiency}
-Recall that for a subgroup of size $s$, the Kogge-Stone scan has a size complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
+For a subgroup of size $s$, the Kogge-Stone scan has a size complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
 \begin{equation}
@@
 \subsubsection{Depth}
-Recall that a Kogge-Stone scan has a depth of $\log_2 s$ Thus, the total depth of the scan is:
+The Kogge-Stone scan has a depth of $\log_2 s$ Thus, the total depth of the scan is:
 \begin{equation}
@@
 \end{equation}
-Minimal depth is preserved regardless of the subgroup size.
-However, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a workgroup-wide barrier at each iteration.
+While minimal depth is preserved regardless of the subgroup size, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a workgroup-wide barrier at each iteration.
 Consequently, while the theoretical depth remains constant across subgroup sizes, smaller subgroups tend to perform worse in practice due to the increased relative cost of synchronization and the overhead of additional iterations in the main loop.
@@
 
-This increased sensitivity to fairness exacerabtes existing issues which negatively impact the performance of previous scan architectures.
+This increased sensitivity to fairness exacerabtes existing issues which negatively impact the performance of previous scan architectures, namely lack of compute, and slow atomic update latency.
 On hardware that lacks sufficient compute power to be memory-bandwidth bound, the additional work incurred by redundant fallback reductions is particularly deleterious.

### 2024-12-30 paper.tex
Push draft Background and Design

@@
 \section{Introduction}
-
+(Tell readers we are targetting WebGPU)
+(Speed of light definition here)
 \subsection{Goals of a Scan Implementation}
@@
 \subsection{The Scan Primitive}
-The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond\cite{10.1145/322217.322232, 5219801}.
+The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{10.1145/322217.322232, 5219801}.
 A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \).
 The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \).
-Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid\cite{}.
+Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid~\cite{}.
 In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence.
@@
 
-Due to its significance in circuits and as a fundamental algorithmic primitive, scan enjoys a rich literature in both electrical engineering and computer science~\cite{}.
-As comprehensive taxonomies of scan networks are well-documented, our focus is on exploring the progression of inter-workgroup (global) scan strategies and their fitness relative to our ideal architecture.
+Due to its significance in circuits and as a fundamental algorithmic primitive, scan enjoys a rich literature in both electrical engineering and computer science.
+Harris provides a taxonomy relating depth, fanout, and wire tracks.
+Hinze provides an algebra of scans.
+Snir proved that depth and size are related by .
+Fich proved that .
+Merill and Garland provide a comprehensive review of GPU scan implementations in .
+In light of this work(reword this), our focus is on exploring the progression of inter-workgroup (global) scan strategies and their fitness relative to our ideal architecture.
+
+Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science.
+Harris offers a taxonomy that relates depth, fanout, and wire tracks~\cite{}, while Hinze develops an algebraic framework for scans~\cite{}.
+Snir proved that \emph{depth} and \emph{size} are related by (fill in)~\cite{}, and Fich proved that among minimum depth scans Ladner-Fischer networks have optimal size.
+Finally, Merrill and Garland provide a comprehensive review of GPU scan implementations.
+In this work, we focus on the evolution of inter-workgroup (global) scan strategies and evaluate their suitability for our ideal architecture.
 
@@
 \subsubsection{Scan-then-Propagate}
-Introduced by Sengupta et al.\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate} \cite{GPUGems3, Sengupta2008} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases:
+Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate} ~\cite{GPUGems3, Sengupta2008} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases:
 \begin{enumerate} 
@@
 \subsubsection{Reduce-then-Scan}
-As the design of GPUs settled around the high-throughput but high-latency paradigm, reducing global memory access, rather than compute, became the prevailing goal for algorithms like scan, which are computationally light enough for their compute to be masked by global memory access.
-This led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}\cite{Merrill Grimshaw, Ha and Han, Dotsenko, Breitbart}: by writing only the work tile reduction in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated.
+Due to the extreme latency of global memory access relative to compute, computationally light algorithms like \emph{scan} can benefit from redundant calculation if it reduces memory access.
+This insight led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill Grimshaw, Ha and Han, Dotsenko, Breitbart}: by writing only the work tile reduction in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated.
 The \emph{Reduce-Then-Scan} strategy operates in three phases:
@@
 \end{enumerate}
-\emph{Reduce-then-Scan} implementations also introduced improvements to the spine phase of the scan, which coalesced around a workgroup-level \emph{raking} technique first applied to inter-workgroup scan by Merrill and Grimshaw~\cite{}.
-Instead of handling arbitrary inputs recursively, the total workgroup occupancy $o$ is determined, and the input is partitioned into non-overlapping blocks processed serially by each workgroup.
-This approach limits the recursive depth of the global scan to two and reduces the global memory movement for the spine scan from $O(n/t)$ (incurred by recursion) to $O(3o)$.
-As a result, kernel launches are limited to three and total global memory movement is reduced to $O(3n + 3o) = O(3n)$.
+Merrill and Grimshaw~\cite{} introduced a workgroup-level \emph{raking} technique that improved the spine phase of \emph{Reduce-Then-Scan}.
+Instead of handling arbitrary inputs recursively, this approach determines the total workgroup occupancy $o$ and partitions the input into non-overlapping blocks, processed serially by each workgroup.
+By limiting the recursive depth of the global scan to two, the global memory movement for the spine scan is reduced from $O(n/t)$ to $O(3o)$.
+Consequently, kernel launches are limited to three, and total global memory movement is reduced to $O(3n + 3o) = O(3n)$.
+This approach was further refined by Ha and Han~\cite{}, who modified the upsweep behavior of the first workgroup from reduction to scan and eliminated the redundant upsweep of the last workgroup, thereby saving the equivalent of one workgroup's computation.
 
@@
 in \emph{StreamScan}, the key innovation of GPU chained scanning lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level.
-
 Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups.
@@
 
-While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently achieve the expected throughput of $\text{memoryBandwidth}/2$ (saturate bandwidth?).
-This limitation arises because, in \emph{StreamScan}, a workgroup cannot post its reduction to global memory until it has ingested the reduction of the preceding workgroup.
-This sequential dependency creates a bottleneck, constraining the algorithm’s performance to the rate at which reductions can propagate through device memory—a process limited by the relatively high latency of GPU memory.
+While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently saturate global memory bandwidth, preventing it from achieving speed-of-light performance.
+This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the reduction of the preceding workgroup.
+Although this reduces global memory access during the propagation phase to $O(n/t)$, the sequential dependency creates a bottleneck, constraining the algorithm’s performance to the rate at which reductions propagate through device memory.
+Despite this limitation, the total data movement remains optimal at $O(2n + n/t) = O(2n)$.
 
-Building on \emph{StreamScan}, Merrill and Garland introduced the \emph{Decoupled Lookback} technique to address the bottleneck created by these sequential dependencies.
-Unlike \emph{StreamScan}, where reductions and dependency resolution occur sequentially, \emph{Decoupled Lookback} separates these processes.In \emph{Decoupled Lookback}, each workgroup posts its reduction to device memory and then enters the “lookback” phase.
-During this phase, the workgroup performs a serial traversal backward through preceding workgroup reductions stored in device memory.
-To manage inter-workgroup synchronization, bit-packed status flags are used to indicate the state of each workgroup’s reduction.
-
-TODO Cleanup, add O(n) on global data movement during the lookback phase etc.
+Building on \emph{StreamScan}, Merrill and Garland introduced the \emph{Decoupled Lookback} technique to address the bottleneck created by sequential dependencies.
+Unlike \emph{StreamScan}, where reductions and dependency resolution occur sequentially, \emph{Decoupled Lookback} separates these processes.
+In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}—a serial traversal backward along the spine to resolve dependencies.
+The work performed during this phase is constant-bounded to the workgroup occupancy $o$, as redundant operations are avoided by updating the workgroup's reduction from local to inclusive once the lookback is completed.
+Although this increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing ensures that memory accesses to the spine are roughly contiguous within the workgroup occupancy block, making them highly likely to be cache-resident.
+As a result, the additional global memory traffic along the spine introduces a negligible latency overhead.
+With a total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance.
 
 \subsection{Ideal Scan Architecture}
-RTS
-More portable
-slower
-handling arbitrary inputs can be awkward
 
-Chained Scan
-Faster
-2n data movement + serial ordering faclitates in-situ behavior
-naturally extends to arbitrary data
-not portable, relies on something called forward progress guarantees
+In addition to its speed, the \emph{Chain-Scan} architecture has other, more practical properties, that improve over its predecessors.
+As it is a single kernel, it minimizes launch overheads and simplifies maintenance.
+On previous architectures, handling of arbitrary inputs can be awkward: the recursive approach, for example, entails multiple kernel launches and a heterogenous system to coordinate them.
+Similarly, a raking approach can be cumbersome, as determining the workgroup occupancy for every target device is often impractical.
+Furthermore, on GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—raking the workgroups can lead to performance degradation due to cache thrashing.
+This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups.
+As a consequence of its $2n$ global data movement and inherently serial ordering, \emph{Chained-Scan} is also suitable for \emph{in-situ} compaction and allocation.
+In previous architectures, such tasks often required additional kernel launches or complex bookkeeping to handle dependencies between workgroups.
+By contrast, \emph{Chained-Scan} efficiently propagates dependencies within a single kernel, minimizing overhead and enabling seamless integration of compaction and allocation operations.
 
-Chained has lots of advantages, it would be great if it was portable.
+Thus, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture, it falls short in meeting the criterion of \emph{portability}.
+As we discuss in more detail in the following section, this limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA hardware but not universally available across all vendors and architectures.
+Without this guarantee, \emph{Chained-Scan} is highly prone to catastrophic failure, making it unsuitable for the inherently cross-vendor, cross-architecture environment of shading languages.
 
-\subsection{Why does \emph{Chained-Scan} Rely on Forward Progress Guarantees?}
-TODO rewrite
-The \emph{Chained-Scan} architecture operates in the middle ground between the absence of inter-workgroup synchronization and formal inter-workgroup synchronization support.
-Rather than relying on explicit synchronization primitives, it leverages a combination of atomic operations and a property of the hardware's workgroup scheduling model known as a \emph{forward progress guarantee}.
-Atomics enable coherent communication between workgroups, while the \emph{forward progress guarantee} (FPG) ensures that all active workgroups will eventually make progress towards termination, precluding scenarios where a workgroup is scheduled in a manner that starves another workgroup of execution time.
+\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
+A \emph{forward-progress guarantee} (FPG) is a promise provided by the subgroup scheduler that all occupant subgroups eventually make progress toward termination, preventing scenarios where a subgroup is starved of execution time.
+Sorensen et al.~\cite{} and Sorensen et al.~\cite{} provide the most comprehensive analysis of GPU schedulers, including the implications of running FPG dependent algorithms without it.
 
-In \emph{Chained-Scan}, workgroups operate within a producer-consumer framework: each workgroup produces the reduction of its assigned work tile, and every workgroup (except the first) must consume the reductions produced by all preceding workgroups to complete the scan.
-To guarantee that no workgroup begins scanning a tile for which the preceding reduction may be unavailable, \emph{Chained-Scan} assigns work tiles using atomic increment operations rather than assignment based on workgroup index.
-Once a workgroup acquires its tile, it computes the tile’s reduction, atomically posts the result into global memory, and then waits for preceding reductions to appear in global memory.
-It is this waiting phase that necessitates FPG.
-Because workgroups do not execute in physical lockstep (obvious?), some may begin waiting before others have finished their calculations.
-Without FPG, there is a risk that a workgroup which finishes early could remain indefinitely scheduled in a waiting state, blocking progress by the workgroup responsible for producing the needed reduction and ultimately causing a deadlock.
+In \emph{Chained-Scan}, workgroups operate in a producer-consumer framework: each workgroup produces the reduction of its assigned work tile, while subsequent workgroups consume the reductions of all preceding workgroups to complete the scan.
+To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than static assignment based on virtualized workgroup indices.
+While this dynamic assignment guarantees that predecessor work tiles are either scheduled or completed, without FPG there is no guarantee that these reductions will eventually become available.
+A scheduler lacking FPG may fail to allocate sufficient execution time to certain workgroups, leaving their reductions incomplete indefinitely.
+In such cases, consuming workgroups stall during their lookback phase, waiting for reductions that are never posted to global memory.
 
-As an illustrative example, consider a scan operation which requires exactly two workgroups, running on a GPU with a single multiprocessor.
-While this multiprocessor can host both workgroups, it contains only a single scheduling unit, which lacks FPG.
-In this scenario, both workgroups launch and acquire their respective work tiles, but the workgroup responsible for \emph{Work Tile 1} finishes first and begins waiting on \emph{Work Tile 0}.
-Because the scheduler lacks FPG, it continues scheduling \emph{Work Tile 1}, but because there is only a single scheduling unit, \emph{Work Tile 0} is never scheduled and never completes its reduction.
-As a result, \emph{Work Tile 1} spins indefinitely until \emph{timeout detection and recovery} (TDR) is triggered, crashing the host program in the process.
+For instance, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor.
+Although the multiprocessor can host both workgroups, it has only one scheduling unit, which lacks FPG.
+In this setup, both workgroups launch and acquire their respective work tiles.
+The workgroup responsible for \emph{Work Tile 1} finishes first and begins waiting for the reduction of \emph{Work Tile 0}.
+However, because the scheduler lacks FPG, it may continue scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction.
+As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.
 
 \subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
-Although NVIDIA formalized FPG down to the subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---it remains absent on some contemporary hardware, most notably certain ARM-based chips, including Apple's M-series GPUs~\cite{10.1145/3485508}.
+Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---it remains absent on some contemporary hardware, most notably certain ARM-based chips, including Apple's M-series GPUs~\cite{10.1145/3485508}.
 Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad\footnote{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time.
 It's also unclear why the test did not trigger Metal's timeout device recovery}, and at worst, risks the afformentioned TDR and subsequent program crash.
-This risk is further exacerbated by the fact that, no major graphics API currently provides developers an FPG hardware capability query\footnote{To the best of our knowledge D3D12, Vulkan, and Metal do not have a query for FPG}.
-As a result, when implementing scan operations in a shading environment—where cross-vendor, cross-architecture portability is essential—uncertainty surrounding FPG capability and its catastrophic failure mode compels developers to fall back to the older, slower \emph{Reduce-then-Scan} approach.
+This extreme slowdown highlights the inherent fragility of \emph{Chained-Scan} without FPG.
 
-\subsection{Earlier Attempts at Portability}
-Scalar fallback here.
+Exacerbating this issue is the fact that no major graphics API currently provides developers with a mechanism to query FPG support\footnote{As of writing this, 12/29/2024, we are unaware of any major graphics API---Vulkan, D3D12, or Metal---that provide developers with an FPG capability query}.
+This lack of transparency forces developers to design algorithms without clear knowledge of whether critical guarantees are available on the target hardware.
+In environments like shading languages, where cross-platform portability is required, this uncertainty compels developers to revert to older, less efficient strategies such as \emph{Reduce-Then-Scan}.
+However, as discussed earlier, \emph{Reduce-Then-Scan} carries its own set of drawbacks, including increased global memory traffic ($O(3n)$) and additional kernel launches, which result in reduced throughput and higher latency.
+While these tradeoffs were once deemed acceptable, they stand in stark contrast to the efficiency and simplicity of \emph{Chained-Scan} when implemented on hardware with reliable FPG.
+Yet, in the absence of guaranteed forward progress, developers have little choice but to sacrifice performance for broader compatibility.
+
+\subsection{Earlier Attempts at Portability (Cuttable?)}
+Levien and Naur~\cite{}, made the first attempt at solving forward progress in \emph{DecoupledLookback}, introducing \emph{ScallarFallback}.
+In \emph{ScallarFallback}, each time a workgroup waits on a preceding tile, it begins a \emph{fallback} operation, and begins consuming the blocking tile one element at a time.
+Although this entails performing redundant work when the blocking tile is simply late but not deadlocking, it guarantees the termination of the algorithm because a workgroup can also perform the reduction of a blocking tile itself, thus breaking the inter-workgroup dependency.
+
+\emph{ScallarFallback} suffers from two primary issues that impact its efficiency.
+First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup engages in a waiting spin.This effectively takes $t$ spins to complete the fallback of the deadlocking tile.
+Moreover, because only a single thread is responsible for loading elements, the cache line is severely underutilized, further degrading performance.
+Second, once the fallback is complete, the workgroup does not post its result to device memory.
+Consequently, every workgroup deadlocked by that tile must redundantly execute the fallback, wasting computational resources.
+While \emph{ScallarFallback} was able to achieve speed of light on hardware supporting FPG and executed correctly on hardware without FPG, its performance on the latter was inferior to that of \emph{Reduce-Then-Scan} .
 
 \section{Design}
-\subsection{Goals (These subsection headers can be removed later if necessary)}
+\subsection{Goals}
 This work is guided by two main goals.
@@
 Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained-Scan}.
-
-\subsection{Achieving Goals}
 
@@
 \subsection{Intra-Workgroup Implementation}
-Our intra-workgroup implementation is most similar to the \emph{CUDPP} scan kernel \cite{}, with adaptations to enhance portability. It consists of three phases:
+Our intra-workgroup implementation is most similar to the \emph{CUDPP} scan kernel ~\cite{}, with adaptations to enhance portability. It consists of three phases:
 \begin{enumerate}

### 2024-12-30 paper.tex
Remove duplicate paragraph

@@
 \]
-
-Due to its significance in circuits and as a fundamental algorithmic primitive, scan enjoys a rich literature in both electrical engineering and computer science.
-Harris provides a taxonomy relating depth, fanout, and wire tracks.
-Hinze provides an algebra of scans.
-Snir proved that depth and size are related by .
-Fich proved that .
-Merill and Garland provide a comprehensive review of GPU scan implementations in .
-In light of this work(reword this), our focus is on exploring the progression of inter-workgroup (global) scan strategies and their fitness relative to our ideal architecture.
-
 Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science.

### 2024-12-30 paper.tex
my latex doesn't deal well with spaces in cite commands

@@
 Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency.
-Due to their specialized nature, GPUs function as co-processors to the CPU, requiring GPU programs—known as \emph{kernels} or \emph{shaders}—to be \emph{launched} or \emph{dispatched} heterogeneously by the CPU.
+Due to their specialized nature, GPUs function as co-processors to the CPU, requiring GPU programs—known as \emph{kernels} or \emph{shaders}—to be \emph{launched} or \emph{dispatched} heterogeneously by the CPU\@.
 Upon launch, a kernel is replicated across independent execution contexts, called threads.
@@
 A multiprocessor schedules and executes threads by subgroup.
-While it attempts to execute the same instruction across all threads in the subgroup, SIMT, threads within the same subgroup are free to follow their own data/context-dependent execution paths, SPMD.
+While it attempts to execute the same instruction across all threads in the subgroup, SIMT, threads within the same subgroup are free to follow their own data/context-dependent execution paths, SPMD\@.
 However, this flexibility comes at the cost of subgroup divergence.
@@
 
-Most relevant to this work is the scheduling behavior of the GPU.
+Most relevant to this work is the scheduling behavior of the GPU\@.
 Scheduling is divided into a workgroup-level scheduler\footnote{For example, NVIDIA's \emph{GigaThread Engine} or AMD's \emph{Asynchronous Compute Engines}}, which manages kernel launches and maps workgroups to multiprocessors, and the aforementioned subgroup scheduler, responsible for selecting which of the currently resident subgroups for execution.
@@
 \subsubsection{Scan-then-Propagate}
-Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate} ~\cite{GPUGems3, Sengupta2008} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases:
+Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2008} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases:
 \begin{enumerate}
@@
 Due to the extreme latency of global memory access relative to compute, computationally light algorithms like \emph{scan} can benefit from redundant calculation if it reduces memory access.
-This insight led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill Grimshaw, Ha and Han, Dotsenko, Breitbart}: by writing only the work tile reduction in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated.
+This insight led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill-Grimshaw, Ha-and-Han, Dotsenko, Breitbart}: by writing only the work tile reduction in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated.
 The \emph{Reduce-Then-Scan} strategy operates in three phases:
@@
 Continuing the trend of global data movement optimization, the most recent work on global scan has been the \emph{Chained-Scan} architecture.
-First introduced by Yan et al.
-in \emph{StreamScan}, the key innovation of GPU chained scanning lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level.
+First introduced by Yan et al.\ in \emph{StreamScan}, the key innovation of GPU chained scanning lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level.
 Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups.
@@
 For instance, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor.
-Although the multiprocessor can host both workgroups, it has only one scheduling unit, which lacks FPG.
+Although the multiprocessor can host both workgroups, it has only one scheduling unit, which lacks FPG\@.
 In this setup, both workgroups launch and acquire their respective work tiles.
@@
 It's also unclear why the test did not trigger Metal's timeout device recovery}, and at worst, risks the afformentioned TDR and subsequent program crash.
-This extreme slowdown highlights the inherent fragility of \emph{Chained-Scan} without FPG.
+This extreme slowdown highlights the inherent fragility of \emph{Chained-Scan} without FPG\@.
 
@@
 However, as discussed earlier, \emph{Reduce-Then-Scan} carries its own set of drawbacks, including increased global memory traffic ($O(3n)$) and additional kernel launches, which result in reduced throughput and higher latency.
-While these tradeoffs were once deemed acceptable, they stand in stark contrast to the efficiency and simplicity of \emph{Chained-Scan} when implemented on hardware with reliable FPG.
+While these tradeoffs were once deemed acceptable, they stand in stark contrast to the efficiency and simplicity of \emph{Chained-Scan} when implemented on hardware with reliable FPG\@.
 Yet, in the absence of guaranteed forward progress, developers have little choice but to sacrifice performance for broader compatibility.
@@
 Consequently, every workgroup deadlocked by that tile must redundantly execute the fallback, wasting computational resources.
-While \emph{ScallarFallback} was able to achieve speed of light on hardware supporting FPG and executed correctly on hardware without FPG, its performance on the latter was inferior to that of \emph{Reduce-Then-Scan} .
+While \emph{ScallarFallback} was able to achieve speed of light on hardware supporting FPG and executed correctly on hardware without FPG, its performance on the latter was inferior to that of \emph{Reduce-Then-Scan}.
 
@@
 This work is guided by two main goals.
-First, portability: we want to develop a scan implementation that retains the benefits of the \emph{Chained-Scan} architecture but is also suitable for a diverse range of hardware vendors and achitectures, including those without FPG.
+First, portability: we want to develop a scan implementation that retains the benefits of the \emph{Chained-Scan} architecture but is also suitable for a diverse range of hardware vendors and achitectures, including those without FPG\@.
 Second, performance: we aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.
@@
 \subsection{Intra-Workgroup Implementation}
-Our intra-workgroup implementation is most similar to the \emph{CUDPP} scan kernel ~\cite{}, with adaptations to enhance portability. It consists of three phases:
+Our intra-workgroup implementation is most similar to the \emph{CUDPP} scan kernel~\cite{}, with adaptations to enhance portability. It consists of three phases:
 \begin{enumerate}
@@
 \endinput
+
+%%% Local Variables:
+%%% mode: LaTeX
+%%% TeX-master: t
+%%% End:

### 2024-12-30 paper.tex
added rabin-karp

@@
 \usepackage{mathtools}
+\newcommand{\thomas}[1]{{\footnotesize\color{orange}[Thomas: #1]}}
+\newcommand{\john}[1]{{\footnotesize\color{cyan}[John: #1]}}
+\newcommand{\raph}[1]{{\footnotesize\color{magenta}[Raph: #1]}}
 
@@
 \subsection{The GPU Programming Environment(Model?)}
+\john{So I have written, roughly, a zillion of these sections and here is my current philosophy.
+The danger is that you write something really interesting and relevant and a reviewer who is familiar with GPUs sees the title and sees the first few lines and says ``I can skip this'' and does.
+There are plenty of great descriptions of how GPUs work.
+Cite one of them.
+What this section needs to do is very specifically call out the interesting/detailed features of GPUs that are \emph{particularly relevant to this paper} that are not just ``here's generally how a GPU works''.
+For me, that's the last paragraph.
+The other part that might be necessary is defining the specific terminology used in the paper (e.g., workgroup vs.\ thread block vs.\ CTA).
+If you feel like breaking up a wall of text, which I generally think is a good idea, we can make a little table of equivalent terminology and boldface the terms we're using.}
 Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency.
@@
 Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid~\cite{}.
-In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence.
+\john{Rabin-Karp string matching (``Efficient randomized pattern-matching algorithms'', Section 6, 1987)~\cite[Section 6]{Karp:1987:ERP} has 2$\times$2 matrix multiplication as its binary operation, and is a good thing to cite in a paper going to SPAA\@.} In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence.
 If the subset includes the \( n \)-th element, it is called \textit{inclusive}; if it excludes the \( n \)-th element, it is called \textit{exclusive}.

### 2024-12-30 paper.tex
working through the text, wip

@@
 The other part that might be necessary is defining the specific terminology used in the paper (e.g., workgroup vs.\ thread block vs.\ CTA).
-If you feel like breaking up a wall of text, which I generally think is a good idea, we can make a little table of equivalent terminology and boldface the terms we're using.}
+If you feel like breaking up a wall of text, which I generally think is a good idea, we can make a little table of equivalent terminology and boldface the terms we're using.} \john{Perhaps another good thing to mention that is specific to this paper is the GPU's support for synchronization.
+I could see having a blurb on ``there is no global inter-workgroup synchronization except an expensive kernel barrier'' and also what \emph{is} available in terms of pairwise synchronization between workgroups and what the requirements are (e.g., residence).
+These are more detailed (and important) GPU topics that aren't present in the higher-level summary you have below.
+Later in the paper, you reference ``without an inter-workgroup barrier primitive'' and don't say why, and you don't want the reader to get the wrong idea here (you don't want the reader to say ``oh, we can just add this to the language, we're done'').
+There is a fundamental reason to not have this primitive.
+Here's where you can discuss that.} \john{Thinking about the above, one way to organize this subsection that is most helpful is to \emph{show} here what kind of synchronizations are / aren't possible within the GPU programming model, in a principled way, and then these are the constraints for our implementation.
+We want readers to not just accept a weird set of constraints as a given but rather to understand where they come from, and why.}
 Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency.
@@
 Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid~\cite{}.
-\john{Rabin-Karp string matching (``Efficient randomized pattern-matching algorithms'', Section 6, 1987)~\cite[Section 6]{Karp:1987:ERP} has 2$\times$2 matrix multiplication as its binary operation, and is a good thing to cite in a paper going to SPAA\@.} In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence.
-If the subset includes the \( n \)-th element, it is called \textit{inclusive}; if it excludes the \( n \)-th element, it is called \textit{exclusive}.
+\john{Rabin-Karp's second fingerprinting method for string matching (``Efficient randomized pattern-matching algorithms'', Section 6, 1987)~\cite[Section 6]{Karp:1987:ERP} has 2$\times$2 matrix multiplication as its binary operation, which is non-commutative, and is a good thing to cite in a paper going to a theory conference like SPAA, because string matching is a super-SPAA-friendly topic.
+I put it in the bib if you want to use it.} In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence.
+If the subset includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}.
 The most common scan type is the prefix sum, where \( \oplus \) is addition.
@@
 \subsection{Evolution of Inter-Workgroup Scan Architectures}
-Prior to the introduction of shared memory, early GPU scan implementations~\cite{Horn, Hensley, Greß, Sengupta} operated directly on global memory and performed a single global scan where each level of depth necessitated a separate kernel launch.
+Prior to the introduction of shared memory\john{pre-CUDA?}, early GPU scan implementations~\cite{Horn, Hensley, Greß, Sengupta} operated directly on global memory \john{the previous implementation strategy is PRAM-like and you should say that, it's important to a SPAA audience} and performed a single global scan where each level of depth necessitated a separate kernel launch\john{this latter phrase is not obvious!
+explain why}.
 The introduction of shared memory and dedicated compute environments resulted in the first implementations to resemble contemporary designs, combining coarse-grained inter-workgroup strategies with fine-grained intra-workgroup strategies.

### 2024-12-30 paper.tex
working through text

@@
 These are more detailed (and important) GPU topics that aren't present in the higher-level summary you have below.
-Later in the paper, you reference ``without an inter-workgroup barrier primitive'' and don't say why, and you don't want the reader to get the wrong idea here (you don't want the reader to say ``oh, we can just add this to the language, we're done'').
-There is a fundamental reason to not have this primitive.
-Here's where you can discuss that.} \john{Thinking about the above, one way to organize this subsection that is most helpful is to \emph{show} here what kind of synchronizations are / aren't possible within the GPU programming model, in a principled way, and then these are the constraints for our implementation.
+Later in the paper, you reference ``without an inter-workgroup barrier primitive'' and don't say why, and you don't want the reader to get the wrong idea here (you don't want the reader to say ``oh, we can just add this to the language, we're done'', and you alternatively don't want the reader to say ``oh, we'll just use a kernel launch barrier'').
+There is a fundamental reason why this primitive is expensive.
+Here's where you can discuss that, and then assert that a global barrier (presumably a kernel launch barrier) is so expensive that we do not consider it in our implementations.} \john{Thinking about the above, one way to organize this subsection that is most helpful is to \emph{show} here what kind of synchronizations are / aren't possible within the GPU programming model, in a principled way, and then these are the constraints for our implementation.
 We want readers to not just accept a weird set of constraints as a given but rather to understand where they come from, and why.}
@@
 explain why}.
-The introduction of shared memory and dedicated compute environments resulted in the first implementations to resemble contemporary designs, combining coarse-grained inter-workgroup strategies with fine-grained intra-workgroup strategies.
+The introduction of shared memory \john{I think it's not just this, it's also hierarchical compute, grid$\rightarrow$workgroup$\rightarrow$thread.
+The important thing here is not just saying ``this is the contemporary architecture'' but also WHY it's the contemporary architecture.
+I like how you are focusing on the tile dependencies as the important aspect.} and dedicated compute environments resulted in the first implementations to resemble contemporary designs, combining coarse-grained inter-workgroup strategies with fine-grained intra-workgroup strategies.
 \emph{Work tile} partitioning was introduced to divide inputs into workgroup-consumable chunks that fit within the limited size of shared memory.
@@
 In this section, we review the three prominent global scan strategies—\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}—examining how they navigate this inter-workgroup dependency and evaluating their effectiveness in minimizing global memory access and maximizing memory bandwidth utilization in alignment with our ideal architecture goals.
+
+\john{So if I was explaining this evolution without having read this paper, the way I would do it is by asserting that scan has a relatively low arithmetic intensity and we expect theoretically, and see experimentally, that scans are entirely limited by memory bandwidth, and in fact can run near the theoretical speed of light of the memory system.
+Given that, the entire goal is to make sure we organize computation in a way that allows us to use all of the theoretical memory bandwidth.
+Given THAT, we can talk about using the memory and compute hierarchy (shared memory, tiles, etc.) to make sure we do.}
 
@@
   \item \textbf{Spine Scan:} The reductions from all workgroups, collectively referred to as \emph{the spine}, are gathered and scanned to compute the global prefix of the reductions, resolving the dependencies between workgroups; we call the result of this phase the \emph{root}.
+\john{For large scans, this stage is an insignificant amount of work compared to the first and third.}
   \item \textbf{Propagation:} The root is propagated back to each workgroup and combined with the local scan results to compute the final scan.
+\john{Language is a little sloppy here, since the root is not propagated back to each workgroup; instead each workgroup gets one element of the root, yes?
+I like that you are using and distinguishing root and spine; I just want them to be precise.}
 \end{enumerate}
@@
 Although data movement is dominated by the first and third phase, Sengupta et al.'s implementation handles arbitrarily large inputs using recursion.
-For an input of size $n$ and a work tile of size $t$, this results in a recursive depth of $\lceil \log_t n \rceil$.
-At each recursive step $k$, $n/t^k$ tiles are processed, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps.
+An input of size $n$ and a work tile of size $t$ results in a recursive depth of $\lceil \log_t n \rceil$.
+Each recursive step $k$ processes $n/t^k$ tiles, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps.
 Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.
@@
 \subsubsection{Reduce-then-Scan}
-Due to the extreme latency of global memory access relative to compute, computationally light algorithms like \emph{scan} can benefit from redundant calculation if it reduces memory access.
-This insight led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill-Grimshaw, Ha-and-Han, Dotsenko, Breitbart}: by writing only the work tile reduction in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated.
+Due to the extreme latency \john{I don't think this is a latency problem.
+GPUs tolerate long latency well.
+This is a bandwidth problem.
+We don't have enough bandwidth to keep the compute units fed.} of global memory access relative to compute, computationally light algorithms like \emph{scan} can benefit from redundant calculation if it reduces memory access.
+\john{I would try to use the word ``tradeoff'' here, because you'll use it again.
+In this case, we profit from trading additional computation for less memory bandwidth.
+We want to come back to ``scan is memory-bound, so reducing memory traffic, even at the cost of additional computation, is profitable''.} This insight led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill-Grimshaw, Ha-and-Han, Dotsenko, Breitbart}: by writing only the work tile reduction in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated.
 The \emph{Reduce-Then-Scan} strategy operates in three phases:
@@
 Instead of handling arbitrary inputs recursively, this approach determines the total workgroup occupancy $o$ and partitions the input into non-overlapping blocks, processed serially by each workgroup.
-By limiting the recursive depth of the global scan to two, the global memory movement for the spine scan is reduced from $O(n/t)$ to $O(3o)$.
+\john{The important part here IMHO is changing the problem size in the second step from ``variable sized, and proportional to the problem size'' (which might require recursion) to ``constant sized, and thus guaranteed to avoid recursion''.
+I would identify the former as an important problem to be solved because of the potential need for further recursion, and the latter as an elegant solution to it.} By limiting the recursive depth of the global scan to two, the global memory movement for the spine scan is reduced from $O(n/t)$ to $O(3o)$.
 Consequently, kernel launches are limited to three, and total global memory movement is reduced to $O(3n + 3o) = O(3n)$.

### 2024-12-30 paper.tex
still churnin'

@@
 Upon launch, a kernel is replicated across independent execution contexts, called threads.
-These threads are first organized into smaller units known as \emph{subgroups}\footnote{Also referred to as \emph{Warps}, \emph{Waves}, or \emph{SIMD-Groups}.}, which are then grouped into larger units called \emph{workgroups}\footnote{Also referred to as \emph{Cross-Thread-Arrays}, \emph{Thread-Blocks}, \emph{Blocks}, \emph{Thread-Groups}, or \emph{Groups}.}.
-This thread hierarchy mirrors the GPU memory hierarchy: individual threads access private, high-speed registers; threads within a workgroup share low-latency shared memory\footnote{Also referred to as \emph{Local Data Storage}, \emph{Group-Shared Memory}, (Check metal terminology).}; and all threads have access to high-bandwidth but higher-latency global memory.
+These threads are first organized into smaller units known as \emph{subgroups},\footnote{Also referred to as \emph{Warps}, \emph{Waves}, or \emph{SIMD-Groups}.} which are then grouped into larger units called \emph{workgroups}.\footnote{Also referred to as \emph{Cross-Thread-Arrays}, \emph{Thread-Blocks}, \emph{Blocks}, \emph{Thread-Groups}, or \emph{Groups}.} This thread hierarchy mirrors the GPU memory hierarchy: individual threads access private, high-speed registers; threads within a workgroup share low-latency shared memory;\footnote{Also referred to as \emph{Local Data Storage}, \emph{Group-Shared Memory}, (Check metal terminology).} and all threads have access to high-bandwidth but higher-latency global memory.
 
@@
 
-\subsection{The Scan Primitive}
+\subsection{The Scan Primitive}\label{sec:the-scan-primitive}
 The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{10.1145/322217.322232, 5219801}.
@@
 \john{So if I was explaining this evolution without having read this paper, the way I would do it is by asserting that scan has a relatively low arithmetic intensity and we expect theoretically, and see experimentally, that scans are entirely limited by memory bandwidth, and in fact can run near the theoretical speed of light of the memory system.
-Given that, the entire goal is to make sure we organize computation in a way that allows us to use all of the theoretical memory bandwidth.
+So the first goal is to find an algorithmic formulation that requires a minimum memory volume.
+Then given the above, and given a particular algorithmic formulation, the entire goal is to make sure we organize computation in a way that allows us to use all of the theoretical memory bandwidth.
 Given THAT, we can talk about using the memory and compute hierarchy (shared memory, tiles, etc.) to make sure we do.}
@@
 Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.
+
+\john{I liked what you said above (Section~\ref{sec:the-scan-primitive}) about how each of these formulations ``navigate this inter-workgroup dependency''.
+Make that explicit in this subsubsection, and each of the below!}
 
@@
 This shifts the problem from explicit synchronization to implicit synchronization via the order in which workgroups access global data.
-Each workgroup processes its assigned tile independently, waiting only for the result of the preceding reductions before continuing.
+Each workgroup processes \john{I think you want to say, explicitly, the work that each workgroup actually does, and note what part of that is latency-sensitive (since, as you say below, the latency of the sequential dependency is super important for overall scan performance.} its assigned tile independently, waiting only for the result of the preceding reductions before continuing.
 This structure ensures that global memory traffic is minimized, as each element is read and written exactly once, achieving the theoretical minimum of $2n$.
@@
 Despite this limitation, the total data movement remains optimal at $O(2n + n/t) = O(2n)$.
+\john{Somewhere in here you want to put in at least a hint of ``forward progress guarantee''.
+You don't have to use those words.
+But when you get to FPG, you want to point back and say ``this formulation depends on \ldots''.}
 
@@
 
-In addition to its speed, the \emph{Chain-Scan} architecture has other, more practical properties, that improve over its predecessors.
+In addition to its speed, the \emph{Chain-Scan} \john{Be consistent, Chained-Scan or Chain-Scan?} architecture has other, more practical properties, that improve over its predecessors.
 As it is a single kernel, it minimizes launch overheads and simplifies maintenance.
-On previous architectures, handling of arbitrary inputs can be awkward: the recursive approach, for example, entails multiple kernel launches and a heterogenous system to coordinate them.
-Similarly, a raking approach can be cumbersome, as determining the workgroup occupancy for every target device is often impractical.
+On previous architectures, handling of arbitrary inputs \john{input sizes?} can be awkward: the recursive approach, for example, entails multiple kernel launches and a heterogenous system to coordinate them.
+Similarly, a raking approach \john{(a necessary optimization in approaches prior to Chained-Scan) ?} can be cumbersome, as determining the workgroup occupancy for every target device is often impractical.
 Furthermore, on GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—raking the workgroups can lead to performance degradation due to cache thrashing.
@@
 In previous architectures, such tasks often required additional kernel launches or complex bookkeeping to handle dependencies between workgroups.
-By contrast, \emph{Chained-Scan} efficiently propagates dependencies within a single kernel, minimizing overhead and enabling seamless integration of compaction and allocation operations.
+By contrast, \emph{Chained-Scan} efficiently propagates dependencies within a single kernel, minimizing overhead \john{Your audience might not have any idea what you are talking about in the next phrase.
+Add something like ``augmenting scan with useful extensions such as \ldots''?} and enabling seamless integration of compaction and allocation operations.
 
-Thus, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture, it falls short in meeting the criterion of \emph{portability}.
+But, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture \john{``in terms of its theoretical and implementation characteristics'' ?}, it falls short in meeting the criterion of \emph{portability}.
 As we discuss in more detail in the following section, this limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA hardware but not universally available across all vendors and architectures.
-Without this guarantee, \emph{Chained-Scan} is highly prone to catastrophic failure, making it unsuitable for the inherently cross-vendor, cross-architecture environment of shading languages.
+\john{I would probably try to be precise where you say ``not universally available''.
+That phrase is not to our advantage because it implies that mostly its available and maybe there's this one weirdo GPU where it's not.
+I want to be accurate, but I also want that phrase to imply (if it's true) that there's a significant volume of devices that lack FPG\@.} Without this guarantee, \emph{Chained-Scan} is highly prone to catastrophic failure, making it unsuitable for the inherently cross-vendor, cross-architecture environment of shading languages.
 
@@
 Sorensen et al.~\cite{} and Sorensen et al.~\cite{} provide the most comprehensive analysis of GPU schedulers, including the implications of running FPG dependent algorithms without it.
+\john{It would benefit this paper if you contrasted their work vs.\ the work that is here.
+For instance, if they don't discuss chained scan / if they focus on a different class of algorithms, that's worth saying.
+You don't want a reviewer coming back and saying ``oh, they already totally covered the topic of FPG''\@.}
 
@@
 In this setup, both workgroups launch and acquire their respective work tiles.
-The workgroup responsible for \emph{Work Tile 1} finishes first and begins waiting for the reduction of \emph{Work Tile 0}.
+The workgroup responsible for \emph{Work Tile 1} finishes first and begins waiting \john{I wonder if ``waiting'' needs to be formalized a little here.
+The casual reader might think ``if Work Tile 1 is `waiting', why is it being scheduled at all?''.
+I'm presuming it's spinning and thus actually getting scheduled as a result.
+We want to avoid a reviewer making his/her own conclusions about how that hardware / the implementation SHOULD work and then disagreeing with our conclusions.
+We need to make it clear this isn't an easily avoidable problem.} for the reduction of \emph{Work Tile 0}.
 However, because the scheduler lacks FPG, it may continue scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction.
-As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.
+\john{We also want to avoid a reader thinking ``well, why not just schedule one work tile per multiprocessor, doesn't that solve the problem?''.} As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.
 
@@
 Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---it remains absent on some contemporary hardware, most notably certain ARM-based chips, including Apple's M-series GPUs~\cite{10.1145/3485508}.
-Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad\footnote{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time.
-It's also unclear why the test did not trigger Metal's timeout device recovery}, and at worst, risks the afformentioned TDR and subsequent program crash.
+Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad,\footnote{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time.
+It's also unclear why the test did not trigger Metal's timeout device recovery} and at worst, risks the aforementioned TDR and subsequent program crash.
 This extreme slowdown highlights the inherent fragility of \emph{Chained-Scan} without FPG\@.
 
-Exacerbating this issue is the fact that no major graphics API currently provides developers with a mechanism to query FPG support\footnote{As of writing this, 12/29/2024, we are unaware of any major graphics API---Vulkan, D3D12, or Metal---that provide developers with an FPG capability query}.
-This lack of transparency forces developers to design algorithms without clear knowledge of whether critical guarantees are available on the target hardware.
+Exacerbating this issue is the fact that no major graphics API currently provides developers with a mechanism to query FPG support.\footnote{As of writing this, 12/29/2024, we are unaware of any major graphics API---Vulkan, D3D12, or Metal---that provide developers with an FPG capability query} This lack of transparency forces developers to design algorithms without clear knowledge of whether critical guarantees are available on the target hardware.
 In environments like shading languages, where cross-platform portability is required, this uncertainty compels developers to revert to older, less efficient strategies such as \emph{Reduce-Then-Scan}.
@@
 \subsection{Earlier Attempts at Portability (Cuttable?)}
-Levien and Naur~\cite{}, made the first attempt at solving forward progress in \emph{DecoupledLookback}, introducing \emph{ScallarFallback}.
-In \emph{ScallarFallback}, each time a workgroup waits on a preceding tile, it begins a \emph{fallback} operation, and begins consuming the blocking tile one element at a time.
+Levien and Naur~\cite{}, made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress in \emph{DecoupledLookback}, introducing \emph{ScallarFallback}.
+\john{Scalar, not Scallar?} In \emph{ScallarFallback}, each time a workgroup waits on a preceding tile, it begins a \emph{fallback} operation, and begins consuming the blocking tile one element at a time.
 Although this entails performing redundant work when the blocking tile is simply late but not deadlocking, it guarantees the termination of the algorithm because a workgroup can also perform the reduction of a blocking tile itself, thus breaking the inter-workgroup dependency.
@@
 \emph{ScallarFallback} suffers from two primary issues that impact its efficiency.
-First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup engages in a waiting spin.This effectively takes $t$ spins to complete the fallback of the deadlocking tile.
-Moreover, because only a single thread is responsible for loading elements, the cache line is severely underutilized, further degrading performance.
+First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup engages in a waiting spin.
+This effectively takes $t$ spins to complete the fallback of a deadlocking $t$-sized tile.
+Moreover, because only a single thread is responsible for loading elements, the cache line associated with the loaded elements is severely underutilized, further degrading performance.
 Second, once the fallback is complete, the workgroup does not post its result to device memory.
@@
 As discussed in more detail in \emph{Limits on the Speed-of-Light}, not all architectures offer atomic operations that are sufficiently fast enough or scheduling models that are sufficeintly fair enough to attain speed-of-light performance, and thus we cannot guarantee such performance.
-Although we are cognisant of the risks posed by subgroup divergence, subgroup functions in shading languages do not allow explicit divergence control\footnote{For example, CUDA allows developers to explicitly provision subgroup functions with a mask of the threads that will participate in it.}, placing a solution to subgroup divergence issues outside our scope.
+Although we are cognisant of the risks posed by subgroup divergence, subgroup functions in shading languages do not allow explicit divergence control,\footnote{For example, CUDA allows developers to explicitly provision subgroup functions with a mask of the threads that will participate in it.} placing a solution to subgroup divergence issues outside our scope.
 Furthermore, although real-world scans may require more than 30 bits of space for their reductions, or may involve scanning across multiple struct members, WGSL’s current limitations---lack of fences and 64-bit atomics---lead us to focus on the simplest scenario that fits within the 30-bit range.

### 2024-12-30 paper.tex
done with my read of assignment #3

@@
 \section{Introduction}
-(Tell readers we are targetting WebGPU)
+(Tell readers we are targeting WebGPU)
 (Speed of light definition here)
@@
 As it is a single kernel, it minimizes launch overheads and simplifies maintenance.
-On previous architectures, handling of arbitrary inputs \john{input sizes?} can be awkward: the recursive approach, for example, entails multiple kernel launches and a heterogenous system to coordinate them.
+On previous architectures, handling of arbitrary inputs \john{input sizes?} can be awkward: the recursive approach, for example, entails multiple kernel launches and a heterogeneous system to coordinate them.
 Similarly, a raking approach \john{(a necessary optimization in approaches prior to Chained-Scan) ?} can be cumbersome, as determining the workgroup occupancy for every target device is often impractical.
@@
 Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad,\footnote{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time.
-It's also unclear why the test did not trigger Metal's timeout device recovery} and at worst, risks the aforementioned TDR and subsequent program crash.
+It's also unclear why the test did not trigger Metal's timeout device recovery.
+\john{This is super cool, that you have these results.
+Simply showing a distribution of runtimes (that presumably is a wide distribution) and a fraction of failures will be just fine.
+I could see putting these results right here instead of in the results section.}} and at worst, risks the aforementioned TDR and subsequent program crash.
 This extreme slowdown highlights the inherent fragility of \emph{Chained-Scan} without FPG\@.
 
-Exacerbating this issue is the fact that no major graphics API currently provides developers with a mechanism to query FPG support.\footnote{As of writing this, 12/29/2024, we are unaware of any major graphics API---Vulkan, D3D12, or Metal---that provide developers with an FPG capability query} This lack of transparency forces developers to design algorithms without clear knowledge of whether critical guarantees are available on the target hardware.
+Exacerbating this issue is the fact that no major graphics API currently provides developers with a mechanism to query FPG support.\footnote{As of writing this, 12/29/2024, we are unaware of any major graphics API---Vulkan, D3D12, or Metal---that provide developers with an FPG capability query.
+\john{I like that you are being polite and guarded about this, but when you're 100\% sure, you can just assert it without this qualification.}} This lack of transparency forces developers to design algorithms without clear knowledge of whether critical guarantees are available on the target hardware.
 In environments like shading languages, where cross-platform portability is required, this uncertainty compels developers to revert to older, less efficient strategies such as \emph{Reduce-Then-Scan}.
@@
 \subsection{Earlier Attempts at Portability (Cuttable?)}
-Levien and Naur~\cite{}, made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress in \emph{DecoupledLookback}, introducing \emph{ScallarFallback}.
+\john{It's cuttable only in the sense that it's not peer-reviewed literature (right?) but I think it is appropriate to cite it and this description you have made is a good one.
+Besides showing the problem doesn't have a simple solution, it also totally underscores that you have done your homework to find the relevant prior scan work.
+(As a reviewer, I make a lot of judgments based on the quality of a background section; an author can make me favorably inclined toward their work by showing they really understand, and can place in proper context, the prior work on the topic.
+Bonus points for citing work I don't know.)}
+Levien and Naur~\cite{} made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress in \emph{DecoupledLookback}, introducing \emph{ScallarFallback}.
 \john{Scalar, not Scallar?} In \emph{ScallarFallback}, each time a workgroup waits on a preceding tile, it begins a \emph{fallback} operation, and begins consuming the blocking tile one element at a time.
@@
 \section{Design}
+
+\john{I think here you want a Problem Statement.
+What are you actually building?
+Here is where you say ``scan of 30-bit ints, arbitrary size, arbitrary monoid'', that sort of thing.
+Below in ``Non Goals'' you make reference to 30 bits and structs, but you haven't formally stated what you're building.
+Here's where you state this.
+It'll also be in the abstract.}
+
 \subsection{Goals}
-This work is guided by two main goals.
-First, portability: we want to develop a scan implementation that retains the benefits of the \emph{Chained-Scan} architecture but is also suitable for a diverse range of hardware vendors and achitectures, including those without FPG\@.
-Second, performance: we aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.
+This work is guided by two main goals:
+
+\begin{enumerate}
+  \item \textbf{Portability}: we want to develop a scan implementation that retains the benefits of the \emph{Chained-Scan} architecture but is also suitable for a diverse range of hardware vendors and architectures, including those without FPG\@.
+  \item \textbf{Performance}: we aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.
+\end{enumerate}
+
+\noindent
 To ground these objectives, we select the WebGPU shading language (WGSL) and the WebGPU standard as our target environment.
@@
 \subsection{Non Goals}
-Although our goal is to create a fully portable and highly performant scan implementation, we are constrained by underlying hardwares and programming models.
-As discussed in more detail in \emph{Limits on the Speed-of-Light}, not all architectures offer atomic operations that are sufficiently fast enough or scheduling models that are sufficeintly fair enough to attain speed-of-light performance, and thus we cannot guarantee such performance.
-Although we are cognisant of the risks posed by subgroup divergence, subgroup functions in shading languages do not allow explicit divergence control,\footnote{For example, CUDA allows developers to explicitly provision subgroup functions with a mask of the threads that will participate in it.} placing a solution to subgroup divergence issues outside our scope.
+Although our goal is to create a fully portable and highly performant scan implementation, we are constrained by underlying hardware platforms and programming models.
+As discussed in more detail in \emph{Limits on the Speed-of-Light}, not all architectures offer atomic operations that are sufficiently fast enough or scheduling models that are sufficiently fair enough to attain speed-of-light performance, and thus we cannot guarantee such performance.
+Although we are cognizant of the risks posed by subgroup divergence, subgroup functions in shading languages do not allow explicit divergence control,\footnote{For example, CUDA allows developers to explicitly provision subgroup functions with a mask of the threads that will participate in it.} placing a solution to subgroup divergence issues outside our scope.
 Furthermore, although real-world scans may require more than 30 bits of space for their reductions, or may involve scanning across multiple struct members, WGSL’s current limitations---lack of fences and 64-bit atomics---lead us to focus on the simplest scenario that fits within the 30-bit range.
 Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained-Scan}.
+
+\john{So now, actually talk about the design.
+Let's say you wanted to present to, say, Guy Blelloch, who knows scans inside and out but doesn't know much about GPUs.
+How would you explain the approach of decoupled fallback, without diving into GPU details?}
 
@@
 
-This increased sensitivity to fairness exacerabtes existing issues which negatively impact the performance of previous scan architectures, namely lack of compute, and slow atomic update latency.
+This increased sensitivity to fairness exacerbates existing issues which negatively impact the performance of previous scan architectures, namely lack of compute, and slow atomic update latency.
 On hardware that lacks sufficient compute power to be memory-bandwidth bound, the additional work incurred by redundant fallback reductions is particularly deleterious.

### 2025-01-14 paper.tex
added additional example request for why FPG for decoupled lookback

@@
 
+\john{It would help at this point to explain why decoupled lookback \emph{also} requires an FPG\@.
+Your example is good for chained-scan, but the reader would say, well, doesn't decoupled-lookback solve the problem of spinning waiting for a previous block?
+doesn't it put the resident block to work walking backwards and that resident block will, eventually, reduce everything before it and and be able to make progress?}
+
 \subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}

### 2025-01-16 paper.tex
Incorporate feedback, add preliminary graphics.

--- 2024-12-30 paper.tex
@@
 
-\section{Introduction}
+\section{Introduction (WIP)}
 (Tell readers we are targeting WebGPU)
@@
 \end{itemize}
+\thomas{See comment in section 3 !}
+Our method retains the best aspects of the \emph{Chained-Scan} architecture, while eliminating the need for forward progress guarantees.
+It is designed to operate on arbitrary hardware, support arbitrary monoids, process inputs of arbitrary size, and support arbitrary 32-bit and 64-bit data types.
+\subsection{Goals}
+This work is guided by two main goals:
+\begin{enumerate}
+  \item \textbf{Portability}: we want to develop a scan implementation that retains the benefits of the \emph{Chained-Scan} architecture but is also suitable for a diverse range of hardware vendors and architectures, including those without FPG or support for 64-bit atomics.
+  \item \textbf{Performance}: we aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.
+\end{enumerate}
+
+\noindent
+To ground these objectives, we select the WebGPU shading language (WGSL) and the WebGPU standard as our target environment.
+WGSL is a meta-level shading language which is translated and compiled as necessary for backend graphics APIs---D3D12, Vulkan, and Metal---and as such, WGSL implementations are bound by the minimum capability across all backends.
+Because WGSL must operate within this limited capability space, it inherently embodies the challenge of portability.
+
+\subsection{Non Goals}
+Although our goal is to create a fully portable and highly performant scan implementation, we are constrained by underlying hardware platforms and programming models.
+As discussed in more detail in \emph{Limits on the Speed-of-Light}, not all architectures offer atomic operations that are sufficiently fast enough or scheduling models that are sufficiently fair enough to attain speed-of-light performance, and thus we cannot guarantee such performance.
+Although we are cognizant of the risks posed by subgroup divergence, subgroup functions in shading languages do not allow explicit divergence control,\footnote{For example, CUDA allows developers to explicitly provision subgroup functions with a mask of the threads that will participate in them.} placing a solution to subgroup divergence issues outside our scope.
+Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained-Scan}.
+
+\subsection{Contributions}
+\begin{enumerate}
+    \item Decoupled-Fallback
+    \item Split-Method
+    \item Subgroup-Size Agnostic stuff
+\end{enumerate}
 
 \section{Background}
-\subsection{The GPU Programming Environment(Model?)}
-\john{So I have written, roughly, a zillion of these sections and here is my current philosophy.
-The danger is that you write something really interesting and relevant and a reviewer who is familiar with GPUs sees the title and sees the first few lines and says ``I can skip this'' and does.
-There are plenty of great descriptions of how GPUs work.
-Cite one of them.
-What this section needs to do is very specifically call out the interesting/detailed features of GPUs that are \emph{particularly relevant to this paper} that are not just ``here's generally how a GPU works''.
-For me, that's the last paragraph.
-The other part that might be necessary is defining the specific terminology used in the paper (e.g., workgroup vs.\ thread block vs.\ CTA).
-If you feel like breaking up a wall of text, which I generally think is a good idea, we can make a little table of equivalent terminology and boldface the terms we're using.} \john{Perhaps another good thing to mention that is specific to this paper is the GPU's support for synchronization.
-I could see having a blurb on ``there is no global inter-workgroup synchronization except an expensive kernel barrier'' and also what \emph{is} available in terms of pairwise synchronization between workgroups and what the requirements are (e.g., residence).
-These are more detailed (and important) GPU topics that aren't present in the higher-level summary you have below.
-Later in the paper, you reference ``without an inter-workgroup barrier primitive'' and don't say why, and you don't want the reader to get the wrong idea here (you don't want the reader to say ``oh, we can just add this to the language, we're done'', and you alternatively don't want the reader to say ``oh, we'll just use a kernel launch barrier'').
-There is a fundamental reason why this primitive is expensive.
-Here's where you can discuss that, and then assert that a global barrier (presumably a kernel launch barrier) is so expensive that we do not consider it in our implementations.} \john{Thinking about the above, one way to organize this subsection that is most helpful is to \emph{show} here what kind of synchronizations are / aren't possible within the GPU programming model, in a principled way, and then these are the constraints for our implementation.
-We want readers to not just accept a weird set of constraints as a given but rather to understand where they come from, and why.}
+\subsection{Virtualization and Scheduling in the GPU programming model}
 Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency.
-Due to their specialized nature, GPUs function as co-processors to the CPU, requiring GPU programs—known as \emph{kernels} or \emph{shaders}—to be \emph{launched} or \emph{dispatched} heterogeneously by the CPU\@.
-Upon launch, a kernel is replicated across independent execution contexts, called threads.
-These threads are first organized into smaller units known as \emph{subgroups},\footnote{Also referred to as \emph{Warps}, \emph{Waves}, or \emph{SIMD-Groups}.} which are then grouped into larger units called \emph{workgroups}.\footnote{Also referred to as \emph{Cross-Thread-Arrays}, \emph{Thread-Blocks}, \emph{Blocks}, \emph{Thread-Groups}, or \emph{Groups}.} This thread hierarchy mirrors the GPU memory hierarchy: individual threads access private, high-speed registers; threads within a workgroup share low-latency shared memory;\footnote{Also referred to as \emph{Local Data Storage}, \emph{Group-Shared Memory}, (Check metal terminology).} and all threads have access to high-bandwidth but higher-latency global memory.
+(one sentence on thread hierarchy maybe).
+As comprehensive descriptions of the GPU programming model~\cite{} already exist, we focus on the aspects most relevant to this work: scheduling and synchronization.
 
-While threads within the same workgroup can synchronize with each other using barrier primitives and can communicate through shared memory, a workgroup is a logical unit of execution, not a physical processor.
-Instead, workgroups are dynamically mapped and scheduled to a GPU's underlying \emph{multiprocessors}.
-This abstraction enables the kernel to execute on GPUs with varying hardware capabilities, such as different numbers of multiprocessors or multiprocessors with differing execution resources.
-A single multiprocessor can host multiple workgroups; we define the number of active subgroups on a multiprocessor as the \emph{subgroup occupancy}, and the number of active workgroups across the entire GPU as the \emph{workgroup occupancy}.
-Generally speaking, high subgroup occupancy is desirable because multiprocessors are specifically designed to context switch between subgroups, keeping execution units busy and masking memory and execution latency.
+On GPU hardware, scheduling is divided into two levels: a \emph{workgroup} scheduler, which manages kernel launches and maps virtual processors, workgroups, to physical processors, \emph{multiprocessors}, and a \emph{subgroup} scheduler, responsible for selecting which of the currently \emph{occupant} subgroups on a multiprocessor for execution.
+The GPU programming model virtualizes processors to ensure GPU programs---\emph{kernels}---remain portable across hardware with differing physical resources.
+Effective virtualization requires \emph{oversubcription}, a programming paradigm where tasks are (over)partitioned in such a way that kernels request more virtual processors than are physically available.
+As a result, a single multiprocessor may host multiple workgroups simultaneously.
+However, as a consequence of virtualization, the developer relinquishes control over the order in which workgroups are launched, and once a workgroup begins execution on a multiprocessor, it must run to completion and cannot context switch.
 
-GPUs exhibit both single-instruction, multiple-thread (SIMT) and single-program, multiple-data (SPMD) behaviors.
-A multiprocessor schedules and executes threads by subgroup.
-While it attempts to execute the same instruction across all threads in the subgroup, SIMT, threads within the same subgroup are free to follow their own data/context-dependent execution paths, SPMD\@.
-However, this flexibility comes at the cost of subgroup divergence.
-When threads in a subgroup diverge due to conditional branching or other control flow, each unique path is executed sequentially, but at the full execution width, effectively rendering the operation serial and reducing parallel efficiency.\footnote{The scheduling and execution width of the underlying hardware does not necessarily align with the subgroup width.}
+\subsubsection{Inter-Workgroup Barriers}
+Although threads \emph{within} a workgroup may synchronize with each other by participating in a \emph{barrier} primitive, the virtualization of processors introduces an emergent high-level limitation: the absence of an inter-workgroup barrier.
+As previously mentioned, workgroups cannot context switch.
+Although contemporary GPU hardware~\cite{} supports workgroup preemption for prioritizing real-time tasks or managing multi-process workloads, this functionality is deliberately excluded from the GPU programming model.
+This stems from the prohibitively high overhead in latency and storage associated with moving workgroup contexts on and off chip.
+In the oversubscription paradigm, where the number of workgroups is theoretically unbound and typically far exceeds the available multiprocessors, every invocation of such a barrier would necessitate context-switching between all launched workgroups.
+Due to this high cost, no GPU programming environment furnishes developers with a true unbounded-launch inter-workgroup barrier.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{}, but it is almost certainly designed to operate within the \emph{Persistent-Thread}~\cite{} paradigm, rather than serving as a barrier across an unbounded launch.} Instead, when inter-workgroup synchronization is necessary, developers must rely on kernel launches as synchronization points.
+However, kernel launches are suboptimal as barriers: they incur overheads due to the heterogeneous nature of the operation, and all execution context is lost between launches.
 
-Most relevant to this work is the scheduling behavior of the GPU\@.
-Scheduling is divided into a workgroup-level scheduler\footnote{For example, NVIDIA's \emph{GigaThread Engine} or AMD's \emph{Asynchronous Compute Engines}}, which manages kernel launches and maps workgroups to multiprocessors, and the aforementioned subgroup scheduler, responsible for selecting which of the currently resident subgroups for execution.
-Although contemporary GPU hardware supports workgroup preemption to prioritize real-time tasks or manage multi-process workloads~\cite{}(NVIDIA Volta,AMD RDNA), this capability does not extend to the GPU programming model.
-Once a workgroup begins execution on a multiprocessor, it must run to completion and cannot context switch.
-At the subgroup scheduling level, \emph{fairness}—how evenly execution resources are distributed among threads—and \emph{progress guarantees}—ensuring that all threads or subgroups eventually make forward progress—are the most salient issues.
-As we will discuss in the following section, these factors have profound implications for kernel performance and correctness.
+\subsubsection{Fairness and Progress}
+At the subgroup scheduling level, two key issues emerge: \emph{fairness}---how evenly execution resources are distributed among threads---and \emph{progress guarantees}---ensuring that all threads or subgroups eventually make progress towards termination.
+However, the term "fairness" is used differently across the literature, leading to potential confusion.
+Sorensen et al.~\cite{10.1145/3022671.2984032, sorensen_et_al:LIPIcs.CONCUR.2018.23, 10.1145/3485508}, whose work provides the most comprehensive taxonomy of progress models to date, use "fairness" specifically to describe differing levels of progress guarantees.
+In contrast, NVIDIA publications~\cite{4523358, Merrill2016} define "fairness" in terms of the distribution of execution resources among subgroups.
+In this work, we adopt NVIDIA's terminology to maintain consistency with their publications, while acknowledging the broader definitions provided by Sorensen et al.
+
+\thomas{A scheduler may be fair but may not have FPG (M1 Pro for example); a scheduler may have FPG but may not be fair (ARM Mali). \newline
+Sorensen calls FPG "fairness:" "We have described two schedulers: fair and unfair, under which starvation-freedom for blocking idioms is either always or never guaranteed."\newline
+Nvidia, by contrast uses fairness in line with our language. From Tesla architecture: \newline
+"A scoreboard qualifies each warp for issue each cycle.
+The instruction scheduler prioritizes all ready warps and selects the one with highest priority for issue.
+Prioritization considers warp type, instruction type, and 'fairness' to all warps executing in the SM\@." \newline
+From DecoupledLookback: \newline
+"Safety. The algorithm will run to completion if the system guarantees forward-progress for all processors." \newline
+"Blocking will be minimal for systems that provide fair or nearly-fair scheduling. Fairness ensures that all processors will have recorded their aggregates in roughly the same amount of time."
+}
 
 \subsection{The Scan Primitive}\label{sec:the-scan-primitive}
-The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{10.1145/322217.322232, 5219801}.
+The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{5219801}.
 A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \).
 The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \).
-Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid~\cite{}.
-\john{Rabin-Karp's second fingerprinting method for string matching (``Efficient randomized pattern-matching algorithms'', Section 6, 1987)~\cite[Section 6]{Karp:1987:ERP} has 2$\times$2 matrix multiplication as its binary operation, which is non-commutative, and is a good thing to cite in a paper going to a theory conference like SPAA, because string matching is a super-SPAA-friendly topic.
-I put it in the bib if you want to use it.} In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence.
+Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}.
+In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence.
 If the subset includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}.
@@
 Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science.
-Harris offers a taxonomy that relates depth, fanout, and wire tracks~\cite{}, while Hinze develops an algebraic framework for scans~\cite{}.
-Snir proved that \emph{depth} and \emph{size} are related by (fill in)~\cite{}, and Fich proved that among minimum depth scans Ladner-Fischer networks have optimal size.
-Finally, Merrill and Garland provide a comprehensive review of GPU scan implementations.
-In this work, we focus on the evolution of inter-workgroup (global) scan strategies and evaluate their suitability for our ideal architecture.
+Harris~\cite{1292373} offers a taxonomy that relates depth, fanout, and wire tracks, while Hinze~\cite{10.1007/978-3-540-27764-4_11} develops an algebraic framework for scans.
+Snir~\cite{10.1016/0196-6774(86)90003-9} proved that depth $d$ and size $s$ are related by $s + d \ge 2n - 2$, and Fich~\cite{10.1145/800061.808738} proved that among minimum depth scans Ladner-Fischer~\cite{10.1145/322217.322232} networks have optimal size.
+Blelloch~\cite{} adapted scan to the PRAM model and popularized its use as an algorithmic primitive.
+Finally, Merrill and Garland~\cite{Merrill2016} provide a review of GPU scan implementations, complementing earlier work by Merrill and Grimshaw~\cite{Merrill2009}.
 
 \subsection{Evolution of Inter-Workgroup Scan Architectures}
-Prior to the introduction of shared memory\john{pre-CUDA?}, early GPU scan implementations~\cite{Horn, Hensley, Greß, Sengupta} operated directly on global memory \john{the previous implementation strategy is PRAM-like and you should say that, it's important to a SPAA audience} and performed a single global scan where each level of depth necessitated a separate kernel launch\john{this latter phrase is not obvious!
-explain why}.
-The introduction of shared memory \john{I think it's not just this, it's also hierarchical compute, grid$\rightarrow$workgroup$\rightarrow$thread.
-The important thing here is not just saying ``this is the contemporary architecture'' but also WHY it's the contemporary architecture.
-I like how you are focusing on the tile dependencies as the important aspect.} and dedicated compute environments resulted in the first implementations to resemble contemporary designs, combining coarse-grained inter-workgroup strategies with fine-grained intra-workgroup strategies.
-\emph{Work tile} partitioning was introduced to divide inputs into workgroup-consumable chunks that fit within the limited size of shared memory.
-However, as a natural result of this partitioning, a serial dependency is created between the work tiles.
-In this section, we review the three prominent global scan strategies—\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}—examining how they navigate this inter-workgroup dependency and evaluating their effectiveness in minimizing global memory access and maximizing memory bandwidth utilization in alignment with our ideal architecture goals.
-
-\john{So if I was explaining this evolution without having read this paper, the way I would do it is by asserting that scan has a relatively low arithmetic intensity and we expect theoretically, and see experimentally, that scans are entirely limited by memory bandwidth, and in fact can run near the theoretical speed of light of the memory system.
-So the first goal is to find an algorithmic formulation that requires a minimum memory volume.
-Then given the above, and given a particular algorithmic formulation, the entire goal is to make sure we organize computation in a way that allows us to use all of the theoretical memory bandwidth.
-Given THAT, we can talk about using the memory and compute hierarchy (shared memory, tiles, etc.) to make sure we do.}
+GPU scans are distinguished from earlier PRAM-like scans~\cite{} by their alignment with the GPU memory hierarchy, which grades memory into progressively faster but increasingly private tiers.
+To operate efficiently, GPU scans must hybridize fine-grain intra-workgroup (\emph{local}) strategies, executed privately within shared memory and registers, with coarse-grain inter-workgroup (\emph{global}) strategies, performed in global memory.
+Due to their relatively low arithmetic intensity, it is theoretically expected and experimentally observed that the performance of local scans is memory-bound, as first demonstrated by Merrill and Grimshaw~\cite{}.
+Consequently, once equipped with a sufficient local scan strategy, the focus of scan design shifts to addressing the challenges of the global strategy, particularly inter-workgroup coordination and synchronization.
+In this section, we review the three prominent GPU global scan strategies---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---examining how they navigate inter-workgroup dependencies and evaluating their effectiveness in minimizing global memory access and maximizing memory bandwidth utilization in alignment with our ideal architecture goals.
 
@@
 \begin{enumerate}
-  \item \textbf{Intra-Workgroup (Local) Scan:} Each workgroup performs an inclusive scan on its assigned work-tile partition of the input data.
+  \item \textbf{Intra-Workgroup (Local) Scan:} Each workgroup performs an inclusive scan on its assigned work tile partition of the input data.
 The results, including both the local scan output and the reduction of each workgroup, are written back to global memory.
-  \item \textbf{Spine Scan:} The reductions from all workgroups, collectively referred to as \emph{the spine}, are gathered and scanned to compute the global prefix of the reductions, resolving the dependencies between workgroups; we call the result of this phase the \emph{root}.
-\john{For large scans, this stage is an insignificant amount of work compared to the first and third.}
-  \item \textbf{Propagation:} The root is propagated back to each workgroup and combined with the local scan results to compute the final scan.
-\john{Language is a little sloppy here, since the root is not propagated back to each workgroup; instead each workgroup gets one element of the root, yes?
-I like that you are using and distinguishing root and spine; I just want them to be precise.}
+  \item \textbf{Spine Scan:} The reductions from all workgroups, collectively referred to as \emph{spine}, are gathered and scanned to compute the global prefix of the reductions, resolving the dependencies between work tiles; we call the result of this phase the \emph{root}.
+For large scans, this stage is an insignificant amount of work compared to the first and third.
+  \item \textbf{Propagation:} Each workgroup retrieves its portion of the root and combines it with the result of the local scan to compute the final scan.
 \end{enumerate}
-Without an inter-workgroup barrier primitive, each phase is implemented as a separate kernel launch to ensure a consistent view of the reductions.
-During the scan operation, each element is read from and written to global memory once during the local scan and again during the propagation phase, contributing a total of $4n$ global memory operations.
-Although data movement is dominated by the first and third phase, Sengupta et al.'s implementation handles arbitrarily large inputs using recursion.
-An input of size $n$ and a work tile of size $t$ results in a recursive depth of $\lceil \log_t n \rceil$.
+
+As each work tile depends on the reductions of previous tiles, the absence of an inter-workgroup barrier necessitates kernel launches to resolve inter-workgroup dependencies.
+This results in the three-phase construction of \emph{Scan-then-Propagate}, with each phase implemented as a separate kernel launch.
+This is further complicated in Sengupta et al.'s implementation, where arbitrarily large inputs are processed through recursion.
+As the size of a work tile limits the maximum number of elements processed without inter-workgroup dependency resolution, the three-phase scan structure must be recursively applied until the spine size fits within a single work tile.
+Thus, an input of size $n$ and a work tile of size $t$ results in a recursive depth of $\lceil \log_t n \rceil$, and $2\cdot\lceil \log_t n \rceil - 1$ kernel launches.
 Each recursive step $k$ processes $n/t^k$ tiles, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps.
-Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.
-
-\john{I liked what you said above (Section~\ref{sec:the-scan-primitive}) about how each of these formulations ``navigate this inter-workgroup dependency''.
-Make that explicit in this subsubsection, and each of the below!}
+Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total global data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.
 
 \subsubsection{Reduce-then-Scan}
-Due to the extreme latency \john{I don't think this is a latency problem.
-GPUs tolerate long latency well.
-This is a bandwidth problem.
-We don't have enough bandwidth to keep the compute units fed.} of global memory access relative to compute, computationally light algorithms like \emph{scan} can benefit from redundant calculation if it reduces memory access.
-\john{I would try to use the word ``tradeoff'' here, because you'll use it again.
-In this case, we profit from trading additional computation for less memory bandwidth.
-We want to come back to ``scan is memory-bound, so reducing memory traffic, even at the cost of additional computation, is profitable''.} This insight led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill-Grimshaw, Ha-and-Han, Dotsenko, Breitbart}: by writing only the work tile reduction in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated.
-The \emph{Reduce-Then-Scan} strategy operates in three phases:
+If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like \emph{scan} can profit by trading additional computation for reduced memory traffic.
+This trade-off led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill-Grimshaw, Ha-and-Han, Dotsenko, Breitbart}: by writing only the work tile reduction to global memory in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated.
+Like \emph{Scan-then-Propagate}, \emph{Reduce-then-Scan} employs a three-phase approach using kernel launches to resolve dependencies.
+However, implementations of \emph{Reduce-then-Scan} introduced new ways of processing arbitrarily-large inputs that improved over the previous recursive technique.
 \begin{enumerate}
-  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result into global memory.
-  \item \textbf{Spine Scan:} The reductions from all workgroups are gathered and scanned to produce the root.
-  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating the root as it writes to global memory.
+  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result to global memory.
+  \item \textbf{Spine Scan:} The spine is gathered and scanned to produce the root.
+  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating its portion of the root as it writes to global memory.
 \end{enumerate}
-Merrill and Grimshaw~\cite{} introduced a workgroup-level \emph{raking} technique that improved the spine phase of \emph{Reduce-Then-Scan}.
-Instead of handling arbitrary inputs recursively, this approach determines the total workgroup occupancy $o$ and partitions the input into non-overlapping blocks, processed serially by each workgroup.
-\john{The important part here IMHO is changing the problem size in the second step from ``variable sized, and proportional to the problem size'' (which might require recursion) to ``constant sized, and thus guaranteed to avoid recursion''.
-I would identify the former as an important problem to be solved because of the potential need for further recursion, and the latter as an elegant solution to it.} By limiting the recursive depth of the global scan to two, the global memory movement for the spine scan is reduced from $O(n/t)$ to $O(3o)$.
-Consequently, kernel launches are limited to three, and total global memory movement is reduced to $O(3n + 3o) = O(3n)$.
-This approach was further refined by Ha and Han~\cite{}, who modified the upsweep behavior of the first workgroup from reduction to scan and eliminated the redundant upsweep of the last workgroup, thereby saving the equivalent of one workgroup's computation.
 
-\subsubsection{Chained-Scan}
-Continuing the trend of global data movement optimization, the most recent work on global scan has been the \emph{Chained-Scan} architecture.
-First introduced by Yan et al.\ in \emph{StreamScan}, the key innovation of GPU chained scanning lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level.
-Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups.
-This shifts the problem from explicit synchronization to implicit synchronization via the order in which workgroups access global data.
-Each workgroup processes \john{I think you want to say, explicitly, the work that each workgroup actually does, and note what part of that is latency-sensitive (since, as you say below, the latency of the sequential dependency is super important for overall scan performance.} its assigned tile independently, waiting only for the result of the preceding reductions before continuing.
-This structure ensures that global memory traffic is minimized, as each element is read and written exactly once, achieving the theoretical minimum of $2n$.
+Merill and Grimshaw~\cite{} introduced a workgroup \emph{raking} technique that elides the overhead incurred when the recursive depth exceeds two.
+In this approach, the total workgroup occupancy $o$ is determined, and the input is partitioned into non overlapping blocks of size $\frac{n}{t \cdot o}$, which are raked serially by each workgroup.
+This limits the size of the spine from variably sized, proportional to the input size (possibly requiring recursion), to constant sized, and guaranteed to avoid further recursion.
+Assuming $t \geq o$, global memory movement during the spine scan is reduced from $O(n/t)$ to $O(2o)$.
+Kernel launches are limited to three and total global memory movement is reduced to $O(3n + 4o) = O(3n)$.
+
+Breitbart~\cite{} was the first to apply more advanced inter-workgroup synchronization techniques to scan.
+Recognizing that workgroup raking naturally aligns with a \emph{Persistent-Threads}~\cite{} paradigm, Breitbart utilized an atomics-based inter-workgroup barrier to replace kernel launches.
+Since multiprocessors are not oversubscribed, workgroups remain resident (persist) on the multiprocessors for the entire lifetime of the kernel, enabling synchronization at the barrier without requiring context switching.
+Although global data movement remains at $O(3n)$, this consolidates the scan to a single kernel, minimizing launch overheads.
+\thomas{This also requires FPG, but not ideal to unpack and mention here .
+.
+.}
+
+\subsubsection{Chained-Scan, Stream-Scan}
+First introduced by Yan et al.~\cite{} in \emph{Stream-Scan}, the key innovation of the \emph{Chained-Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level.
+Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and status flags to ensure coherent views of data.
+This shifts dependency resolution from explicit synchronization via barrier-like structures to implicit synchronization via the order in which workgroups access global data.
+As a result, each element is read and written exactly once, achieving the theoretical minimum of $2n$.
+
+In any \emph{Chained-Scan}, once a workgroup computes its local reduction, it must complete two objectives: communicate its reduction to successor tiles and incorporate the reductions of all preceding tiles.
+In \emph{StreamScan}, this is facilitated by a \emph{propagation} phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume.
+During this waiting period, the workgroup is continually scheduled, and the progress of other workgroups must be guaranteed to prevent the spinning workgroup from starving its dependent predecessors of execution, which could result in deadlock.
 
 While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently saturate global memory bandwidth, preventing it from achieving speed-of-light performance.
-This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the reduction of the preceding workgroup.
-Although this reduces global memory access during the propagation phase to $O(n/t)$, the sequential dependency creates a bottleneck, constraining the algorithm’s performance to the rate at which reductions propagate through device memory.
-Despite this limitation, the total data movement remains optimal at $O(2n + n/t) = O(2n)$.
-\john{Somewhere in here you want to put in at least a hint of ``forward progress guarantee''.
-You don't have to use those words.
-But when you get to FPG, you want to point back and say ``this formulation depends on \ldots''.}
+This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the reduction of its predecessor.
+While this results in efficient $O(n/t)$ communication of reductions, and $O(2n+ n/t)= O(2n)$ total data movement, the sequential dependency creates a bottleneck that limits performance to the rate at which reductions propagate through device memory.
+Although this can be mitigated by increasing the size of the work tile, limited on-chip resources preclude scaling the tile size indefinitely.
+Moreover, this approach is unsuitable for a portable environment, as excessively large work tiles risk reducing occupancy on less capable hardware.
 
-Building on \emph{StreamScan}, Merrill and Garland introduced the \emph{Decoupled Lookback} technique to address the bottleneck created by sequential dependencies.
-Unlike \emph{StreamScan}, where reductions and dependency resolution occur sequentially, \emph{Decoupled Lookback} separates these processes.
-In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}—a serial traversal backward along the spine to resolve dependencies.
-The work performed during this phase is constant-bounded to the workgroup occupancy $o$, as redundant operations are avoided by updating the workgroup's reduction from local to inclusive once the lookback is completed.
-Although this increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing ensures that memory accesses to the spine are roughly contiguous within the workgroup occupancy block, making them highly likely to be cache-resident.
+\subsubsection{Chained-Scan, Decoupled-Lookback}
+Building on \emph{Stream-Scan}, Merrill and Garland~\cite{} introduced the \emph{Decoupled-Lookback} technique, which separates reduction propagation and dependency resolution to address the bottleneck caused by propagation latency.
+In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction.
+Work and memory accesses are constant-bounded to the workgroup occupancy $o$ by posting the inclusive reduction after the lookback.
+However, granting each workgroup control over its dependency resolution does not eliminate reliance on progress guarantees, as workgroups must still spin on tiles that have yet to post their reductions, risking deadlock.
+
+Although \emph{Decoupled-Lookback} increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing confines spine memory accesses to a roughly $o$-width sliding window, making them highly likely to be cache-resident.
 As a result, the additional global memory traffic along the spine introduces a negligible latency overhead.
-With a total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance.
+With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled-Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance without the work tile size tuning required by \emph{Stream-Scan}.
 
@@
 
-In addition to its speed, the \emph{Chain-Scan} \john{Be consistent, Chained-Scan or Chain-Scan?} architecture has other, more practical properties, that improve over its predecessors.
+In addition to its speed, the \emph{Chained-Scan} architecture has more practical properties that improve over its predecessors.
 As it is a single kernel, it minimizes launch overheads and simplifies maintenance.
-On previous architectures, handling of arbitrary inputs \john{input sizes?} can be awkward: the recursive approach, for example, entails multiple kernel launches and a heterogeneous system to coordinate them.
-Similarly, a raking approach \john{(a necessary optimization in approaches prior to Chained-Scan) ?} can be cumbersome, as determining the workgroup occupancy for every target device is often impractical.
-Furthermore, on GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—raking the workgroups can lead to performance degradation due to cache thrashing.
+On previous architectures, handling of arbitrary input sizes can be awkward: the recursive approach, for example, entails multiple kernel launches and a heterogeneous system to coordinate them.
+Similarly, the raking approach—a necessary optimization in methods prior to \emph{Chained-Scan}—can be cumbersome, as predetermining static workgroup occupancies for every target device is infeasible.
+Previous workgroup occupancy discovery techniques, such as the one proposed by Sorensen et al.~\cite{}, are impossible to implement in WebGPU as they require at least a memory fence, which is not supported by the specification.\footnote{Our implementation of \emph{Reduce-Then-Scan}, which can be found in the artifact, includes a fenceless occupancy discovery method amenable to the WebGPU environment, but its description is beyond the scope of this work.}
+
+Additionally, on GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—raking the workgroups can lead to performance degradation due to cache thrashing.
 This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups.
+Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns—or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations—may experience significant performance degradation.
+
 As a consequence of its $2n$ global data movement and inherently serial ordering, \emph{Chained-Scan} is also suitable for \emph{in-situ} compaction and allocation.
 In previous architectures, such tasks often required additional kernel launches or complex bookkeeping to handle dependencies between workgroups.
-By contrast, \emph{Chained-Scan} efficiently propagates dependencies within a single kernel, minimizing overhead \john{Your audience might not have any idea what you are talking about in the next phrase.
-Add something like ``augmenting scan with useful extensions such as \ldots''?} and enabling seamless integration of compaction and allocation operations.
+By contrast, \emph{Chained-Scan} efficiently propagates dependencies within a single kernel, minimizing overhead and enabling seamless integration of compaction and allocation operations.
 
-But, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture \john{``in terms of its theoretical and implementation characteristics'' ?}, it falls short in meeting the criterion of \emph{portability}.
-As we discuss in more detail in the following section, this limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA hardware but not universally available across all vendors and architectures.
-\john{I would probably try to be precise where you say ``not universally available''.
-That phrase is not to our advantage because it implies that mostly its available and maybe there's this one weirdo GPU where it's not.
-I want to be accurate, but I also want that phrase to imply (if it's true) that there's a significant volume of devices that lack FPG\@.} Without this guarantee, \emph{Chained-Scan} is highly prone to catastrophic failure, making it unsuitable for the inherently cross-vendor, cross-architecture environment of shading languages.
+But, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in terms of its theoretical and implementation characteristics, it falls short in meeting the criterion of \emph{portability}.
+This limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA devices but notably lacking on Apple's M-series and certain ARM GPUs~\cite{sorensen 2}.
+Without this guarantee, \emph{Chained-Scan} is highly prone to catastrophic failure.
 
@@
 A \emph{forward-progress guarantee} (FPG) is a promise provided by the subgroup scheduler that all occupant subgroups eventually make progress toward termination, preventing scenarios where a subgroup is starved of execution time.
-Sorensen et al.~\cite{} and Sorensen et al.~\cite{} provide the most comprehensive analysis of GPU schedulers, including the implications of running FPG dependent algorithms without it.
-\john{It would benefit this paper if you contrasted their work vs.\ the work that is here.
-For instance, if they don't discuss chained scan / if they focus on a different class of algorithms, that's worth saying.
-You don't want a reviewer coming back and saying ``oh, they already totally covered the topic of FPG''\@.}
+Sorensen et al.~\cite{barriers, 1, 2} provide the most comprehensive analysis of GPU schedulers.
+However, while they classify \emph{Chained-Scan} as an FPG-reliant algorithm, their primary focus is on exploring and formalizing progress models and their implications for synchronization in GPU programming frameworks.
+\thomas{Sorensen places Chained-Scan into the HSA category, but it is unclear if this is correct.
+The definition of HSA is \newline "Threads for which fair execution is guaranteed: The thread with the lowest id that has not yet terminated" \newline However, as Chained-Scan and other serially ordered algorithms no longer use the virtual ID provided by the workgroup scheduler, is fair execution still guaranteed under HSA?
+It seems that it should be classified under OBE.}
 
@@
 In this setup, both workgroups launch and acquire their respective work tiles.
-The workgroup responsible for \emph{Work Tile 1} finishes first and begins waiting \john{I wonder if ``waiting'' needs to be formalized a little here.
-The casual reader might think ``if Work Tile 1 is `waiting', why is it being scheduled at all?''.
-I'm presuming it's spinning and thus actually getting scheduled as a result.
-We want to avoid a reviewer making his/her own conclusions about how that hardware / the implementation SHOULD work and then disagreeing with our conclusions.
-We need to make it clear this isn't an easily avoidable problem.} for the reduction of \emph{Work Tile 0}.
+The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, loading the value at the memory address for the reduction of \emph{Work Tile 0} with each iteration.
 However, because the scheduler lacks FPG, it may continue scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction.
-\john{We also want to avoid a reader thinking ``well, why not just schedule one work tile per multiprocessor, doesn't that solve the problem?''.} As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.
+\john{We also want to avoid a reader thinking ``well, why not just schedule one work tile per multiprocessor, doesn't that solve the problem?''.} \thomas{I actually do not know the answer to this question.
+My impression is that (lack of) FPG is purely a phenomenon of the subgroup scheduler, and thus pigeonholing a multiprocessor with a single workgroup \emph{would} solve the problem.
+Of course this would be totally infeasible to implement due to varied multiprocessor resources.
+HOWEVER, do workgroups execute \emph{completely} indepently of one another?} As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.
 
 \subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
-Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---it remains absent on some contemporary hardware, most notably certain ARM-based chips, including Apple's M-series GPUs~\cite{10.1145/3485508}.
-Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad,\footnote{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time.
-It's also unclear why the test did not trigger Metal's timeout device recovery.
-\john{This is super cool, that you have these results.
-Simply showing a distribution of runtimes (that presumably is a wide distribution) and a fraction of failures will be just fine.
-I could see putting these results right here instead of in the results section.}} and at worst, risks the aforementioned TDR and subsequent program crash.
-This extreme slowdown highlights the inherent fragility of \emph{Chained-Scan} without FPG\@.
+Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---its absence across all hardware is a critical issue for developers.
+Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad,\thomas{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time.
+It's also unclear why the test did not trigger Metal's timeout device recovery.} and at worst, risks the aforementioned TDR and subsequent program crash.
 
-Exacerbating this issue is the fact that no major graphics API currently provides developers with a mechanism to query FPG support.\footnote{As of writing this, 12/29/2024, we are unaware of any major graphics API---Vulkan, D3D12, or Metal---that provide developers with an FPG capability query.
-\john{I like that you are being polite and guarded about this, but when you're 100\% sure, you can just assert it without this qualification.}} This lack of transparency forces developers to design algorithms without clear knowledge of whether critical guarantees are available on the target hardware.
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{graphics/Figure_1.pdf}
+  \caption{Execution times of \emph{Decoupled-Lookback (PLACEHOLDER DATA)}}
+\end{figure}
+
+Exacerbating this issue is the fact that no major graphics API currently provides developers with a mechanism to query FPG support.
+This lack of transparency forces developers to design algorithms without clear knowledge of whether critical guarantees are available on the target hardware.
 In environments like shading languages, where cross-platform portability is required, this uncertainty compels developers to revert to older, less efficient strategies such as \emph{Reduce-Then-Scan}.
-However, as discussed earlier, \emph{Reduce-Then-Scan} carries its own set of drawbacks, including increased global memory traffic ($O(3n)$) and additional kernel launches, which result in reduced throughput and higher latency.
-While these tradeoffs were once deemed acceptable, they stand in stark contrast to the efficiency and simplicity of \emph{Chained-Scan} when implemented on hardware with reliable FPG\@.
 Yet, in the absence of guaranteed forward progress, developers have little choice but to sacrifice performance for broader compatibility.
 
-\subsection{Earlier Attempts at Portability (Cuttable?)}
-\john{It's cuttable only in the sense that it's not peer-reviewed literature (right?) but I think it is appropriate to cite it and this description you have made is a good one.
-Besides showing the problem doesn't have a simple solution, it also totally underscores that you have done your homework to find the relevant prior scan work.
-(As a reviewer, I make a lot of judgments based on the quality of a background section; an author can make me favorably inclined toward their work by showing they really understand, and can place in proper context, the prior work on the topic.
-Bonus points for citing work I don't know.)}
-Levien and Naur~\cite{} made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress in \emph{DecoupledLookback}, introducing \emph{ScallarFallback}.
-\john{Scalar, not Scallar?} In \emph{ScallarFallback}, each time a workgroup waits on a preceding tile, it begins a \emph{fallback} operation, and begins consuming the blocking tile one element at a time.
-Although this entails performing redundant work when the blocking tile is simply late but not deadlocking, it guarantees the termination of the algorithm because a workgroup can also perform the reduction of a blocking tile itself, thus breaking the inter-workgroup dependency.
+\subsection{Earlier Attempts at Portability}
+Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress requirements in \emph{ScalarFallback}.
+In \emph{ScalarFallback}, each time a workgroup spins on a preceding tile, it begins a \emph{fallback} operation, and begins consuming the blocking tile one element at a time.
+Although this entails performing redundant work when the blocking tile is simply late but not deadlocking, it guarantees the termination of the algorithm because a workgroup can perform the reduction of a blocking tile itself, breaking the inter-workgroup dependency.
 
-\emph{ScallarFallback} suffers from two primary issues that impact its efficiency.
+\emph{ScalarFallback} suffers from two primary issues that impact its efficiency.
 First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup engages in a waiting spin.
@@
 Consequently, every workgroup deadlocked by that tile must redundantly execute the fallback, wasting computational resources.
-While \emph{ScallarFallback} was able to achieve speed of light on hardware supporting FPG and executed correctly on hardware without FPG, its performance on the latter was inferior to that of \emph{Reduce-Then-Scan}.
+While \emph{ScalarFallback} was able to achieve speed of light on hardware supporting FPG and executed correctly on hardware without FPG, its performance on the latter was inferior to that of \emph{Reduce-Then-Scan}.
 
-\section{Design}
+\section{Decoupled-Fallback}
+\thomas{I've broken somewhat with the outline and moved the goals and non-goals section to somewhere in the introduction.
+I kept finding myself trying to write around it.
+I think the content there is good, but it's very awkward as it ended up so deep within the paper.
+The placement of this section here feels more natural, and seems to mirror other papers?}
+Our \emph{Decoupled-Fallback} method is a further extension of the \emph{Chained-Scan} architecture.
 
-\john{I think here you want a Problem Statement.
-What are you actually building?
-Here is where you say ``scan of 30-bit ints, arbitrary size, arbitrary monoid'', that sort of thing.
-Below in ``Non Goals'' you make reference to 30 bits and structs, but you haven't formally stated what you're building.
-Here's where you state this.
-It'll also be in the abstract.}
+\section{Implementation}
+We are in WGSL. We don't have: a memory model, 64-bit atomics, and atomicCAS is unreliable.
 
-\subsection{Goals}
-This work is guided by two main goals:
-
+\subsection{Split Method}
 \begin{enumerate}
-  \item \textbf{Portability}: we want to develop a scan implementation that retains the benefits of the \emph{Chained-Scan} architecture but is also suitable for a diverse range of hardware vendors and architectures, including those without FPG\@.
-  \item \textbf{Performance}: we aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.
+    \item Idea: Split the reductions into multiple values; use separate values for local/inclusive instead of using flags to differentiate them.
+    \item Pro: Full 32 or 64 bits.
+    \item Pro: No memory model or 64 bit atomics required!
+    \item Pro: Can safely fallback insert without risking clobbering.
+    \item Con: 4x Memory usage
+    \item Con: Tad slower, more complicated.
 \end{enumerate}
 
-\noindent
-To ground these objectives, we select the WebGPU shading language (WGSL) and the WebGPU standard as our target environment.
-WGSL is a meta-level shading language which is translated and compiled as necessary for backend graphics APIs---D3D12, Vulkan, and Metal---and as such, WGSL implementations are bound by the minimum capability across all backends.
-Because WGSL must operate within this limited capability space, it inherently embodies the challenge of portability.
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{graphics/FlowChart.pdf}
+  \caption{TODO update for split}
+\end{figure}
 
-\subsection{Non Goals}
-Although our goal is to create a fully portable and highly performant scan implementation, we are constrained by underlying hardware platforms and programming models.
-As discussed in more detail in \emph{Limits on the Speed-of-Light}, not all architectures offer atomic operations that are sufficiently fast enough or scheduling models that are sufficiently fair enough to attain speed-of-light performance, and thus we cannot guarantee such performance.
-Although we are cognizant of the risks posed by subgroup divergence, subgroup functions in shading languages do not allow explicit divergence control,\footnote{For example, CUDA allows developers to explicitly provision subgroup functions with a mask of the threads that will participate in it.} placing a solution to subgroup divergence issues outside our scope.
-Furthermore, although real-world scans may require more than 30 bits of space for their reductions, or may involve scanning across multiple struct members, WGSL’s current limitations---lack of fences and 64-bit atomics---lead us to focus on the simplest scenario that fits within the 30-bit range.
-Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained-Scan}.
-
-\john{So now, actually talk about the design.
-Let's say you wanted to present to, say, Guy Blelloch, who knows scans inside and out but doesn't know much about GPUs.
-How would you explain the approach of decoupled fallback, without diving into GPU details?}
-
-\section{Implementation}
-
-\subsection{Decoupled-Fallback}
-
-\begin{algorithm}[htbp]
-  \SetAlgoLined
-  \KwIn{Input data $X$, threshold $\epsilon$}
-  \KwOut{Processed data $Y$}
-  Initialize $Y \leftarrow \emptyset$\;
-  \ForEach{$x \in X$}{
-    \If{$x > \epsilon$}{
-      Add $x$ to $Y$\;
-    }
-  }
-  \Return{$Y$}\;
-  \caption{Decoupled Lookback with Decoupled Fallback}
-  \label{alg:example}
-\end{algorithm}
-
-\subsection{Intra-Workgroup Implementation}
-Our intra-workgroup implementation is most similar to the \emph{CUDPP} scan kernel~\cite{}, with adaptations to enhance portability. It consists of three phases:
+\subsection{Intra-Workgroup Scan}
+Our intra-workgroup implementation is most similar to the \emph{CUDPP}~\cite{} scan kernel, with adaptations to enhance portability. It consists of three phases:
 \begin{enumerate}
@@
 
-Similar to 3DMS, we split the work tile into non-overlapping blocks, which are distributed to each subgroup.
-However, unlike CUDA, WGSL lacks a \emph{volatile} qualifier for shared memory, which prevents us from loading items into shared memory and transposing them without inserting a barrier.
-Placing a barrier disrupts latency hiding by separating the high-latency global memory load from the subsequent transpose and scan computation.
-To address this, we load data directly into registers and \emph{vectorize} the input.
-This approach preserves a \emph{coalesced} but vector-strided loading pattern.
-For each vector, a thread performs a serial scan across the vector, then participates in a subgroup scan.
-Although this method results in vectors per thread $v$ subgroup scans, it elides the barrier and significantly reduces the kernel's shared memory footprint, as only the subgroup reduction is placed into shared memory.
-Transposing in shared memory requires a work-tile-sized shared memory allocation, whereas our approach relies more heavily on registers, which are generally more abundant than shared memory, especially on low-spec hardware.
-Additionally, keeping scan elements in registers avoids shared-bank conflicts, further enhancing performance.
-As a result, we consider this tradeoff worthwhile.
-
-The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$, and on some hardware, subgroup sizes can vary between kernel launches.
-To handle this variability, we use a generalized radix-$s$ Ladner-Fischer scan network embedded with Kogge-Stone subgroup scans to perform the scan on subgroup reductions.
-The upsweep phase of the Ladner-Fischer network follows a Brent-Kung construction; assuming shared memory bank widths equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts.
-To mitigate this, we employ Merrill-Grimshaw style conflict avoidance~\cite{}.
-This network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and exposure to bank conflicts is limited to cases where $s^2 < w$, as the network collapses to a single level when $s^2 \geq w$.
+The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches.
+To handle this variability, we use a generalized radix-$s$ Ladner-Fischer scan network embedded with Kogge-Stone subgroup scans to perform the scan on the intra-workgroup spine.
+The upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction; assuming shared memory bank widths equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts.
+To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$.
+Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction.
+Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index.
+However, such access is interpreted by the hardware as a broadcast operation, resulting in no conflicts.
+Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; zero bank conflicts across all supported subgroup sizes.
 
@@
   \SetAlgoLined
-  \KwIn{Array to scan $x$, Workgroup size $W$, Subgroup size $S$}
-  \KwOut{Inclusive scanned array $x$}
+  \KwIn{Shared Memory Array to Scan $x$, Workgroup Size $W$, Subgroup Size $S$}
+  \KwOut{Shared Memory Inclusive Scanned Array $x$}
 
@@
 
-  $iteration\_offset \gets 0$\;
   $stride \gets 1$\;
+  $top\_offset \gets 0$\;
 
@@
 
-      $spine\_index \gets ((thread\_id + iteration\_offset) * stride) - iteration\_offset$\;
+      $step \gets spine\_length / stride$
 
-      \If{$spine\_index < spine\_length$}{
-        $x[spine\_index] \gets \text{subgroupInclusiveScan}(x[spine\_index])$\;
+      \If{$thread\_id < step$}{
+        $temp \gets \text{subgroupInclusiveScan}(x[thread\_id + top\_offset])$\;
+        $x[thread\_id + top\_offset] \gets temp$\;
+        \If{$\text{\normalfont highestRankLane}(thread\_id)$}{
+          $x[(thread\_id / S) + step + top\_offset] \gets temp$;
+        }
       }
@@
 
-      \eIf{$j \neq S$}{
+      \If{$j \neq S$}{
         $reduced\_stride \gets j / stride$\;
@@
         $cond2 \gets (fanout\_index \& (j - 1)) \geq reduced\_stride$\;
-        $cond3 \gets ((fanout\_index + 1) \& (reduced\_stride - 1)) \neq 0$\;
 
-        \If{$cond1 \textbf{ \&\& } cond2 \textbf{ \&\& } cond3$}{
-          $x[fanout\_index] \gets x[fanout\_index] + x[((fanout\_index / stride) * stride) - 1]$\;
+        \If{$cond1 \textbf{ \&\& } cond2$}{
+          $x[fanout\_index] \gets x[fanout\_index] + x[((fanout\_index / stride) + top\_offset) - 1]$\;
         }
-      }{
-        $iteration\_offset \gets iteration\_offset + 1$\;
       }
       $stride \gets stride * S$\;
+      $top\_offset \gets top\_offset + step$\;
     }
+    \textbf{barrier()}\;
   }
   \Return{$x$}\;
-  \caption{Subgroup-Size-Agnostic Scan (SAMPLE NO BANK CONFLICT AVOIDANCE)}
+  \caption{Subgroup-Size-Agnostic Scan}
   \label{alg:example}
@@
 
-\subsubsection{Size (Work) Efficiency}
-For a subgroup of size $s$, the Kogge-Stone scan has a size complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
+\subsection{Work Efficiency and Depth}
+\subsubsection{Subgroup Raking}
+
+\subsubsection{Size Agnostic Scan}
+For a subgroup of size $s$, the Kogge-Stone network has a work complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
 \begin{equation}
@@
 \end{equation}
+Since the subgroup size remains constant within each kernel launch, the final work complexity is $O(n)$.
+Thus our implementation preserves asymptotically optimal work, even though the structure is not isomorphic to the radix-two network.
 
-\subsubsection{Depth}
 The Kogge-Stone scan has a depth of $\log_2 s$ Thus, the total depth of the scan is:
@@
 Consequently, while the theoretical depth remains constant across subgroup sizes, smaller subgroups tend to perform worse in practice due to the increased relative cost of synchronization and the overhead of additional iterations in the main loop.
-\subsection{Full Algorithm}
+
+\subsubsection{Full Algorithm}
 
@@
 
-\subsection{Tradeoffs}
+\subsection{Trade-offs}
 
@@
 
-Fairness has been a priority in NVIDIA hardware since at least the Tesla architecture~\cite{}, but the extent to which it is implemented and prioritized by other vendors remains unclear.
-While our results on the least fair hardware demonstrate that \emph{Decoupled-Fallback} remains performant, this performance may not extend to scans involving more complex binary reduction operators or less capable hardware.
+\section{Cut material, can be readded if we have enough space}
+
+To ensure consistency with our coding artifact and promote cross-vendor compatibility, we use the terminology of the WebGPU specification.
+\setlength{\tabcolsep}{3pt}
+\begin{table}[h]
+  \centering
+  \caption{Non-Exhaustive Equivalent Terminology}
+  \label{tab:terminology}
+  \begin{tabular}{lcccc}
+    \toprule
+    \textbf{WGSL}   & \textbf{GLSL}  & \textbf{HLSL}   & \textbf{Metal}     & \textbf{CUDA} \\ \midrule
+    Subgroup        & Subgroup       & Wave            & SIMD-group         & Warp          \\
+    Workgroup       & Workgroup      & Group           & Threadgroup        & Block         \\
+    Workgroup       & Shared         & Groupshared     & Threadgroup        & Shared        \\ \bottomrule
+  \end{tabular}
+\end{table}
+
+Despite its limited suitability within the oversubscription paradigm, inter-workgroup synchronization holds significant performance potential, as many GPU-friendly algorithms—such as \emph{example}, \emph{example}, and \emph{scan}---can benefit from a trade-off that eliminates the costly overhead of kernel-launch synchronization.
+This potential was explored in prior work~\cite{}, culminating in Gupta et al.’s~\cite{} \emph{Persistent Thread} paradigm.
+Rather than oversubscribing processors by launching $b$ workgroups, each consuming a single \emph{work tile}, this approach \emph{discovers} the total workgroup occupancy across the device $o$---referred to by Gupta et al.
+as a \emph{maximal launch}---and launches $o$ workgroups, each responsible for processing $b/o$ work tiles.
+Limiting the launch size allows the workgroups to always remain on chip, eliding context switching, and thus facilitating an efficient inter-workgroup barrier.
+
 \begin{acks}

### 2025-01-17 paper.tex
Rewrites.

@@
 \subsection{Evolution of Inter-Workgroup Scan Architectures}
-GPU scans are distinguished from earlier PRAM-like scans~\cite{} by their alignment with the GPU memory hierarchy, which grades memory into progressively faster but increasingly private tiers.
+Contemporary GPU scans are distinguished from earlier PRAM-like scans~\cite{} by their alignment with the GPU memory hierarchy, which grades memory into progressively faster but increasingly private tiers.
 To operate efficiently, GPU scans must hybridize fine-grain intra-workgroup (\emph{local}) strategies, executed privately within shared memory and registers, with coarse-grain inter-workgroup (\emph{global}) strategies, performed in global memory.
-Due to their relatively low arithmetic intensity, it is theoretically expected and experimentally observed that the performance of local scans is memory-bound, as first demonstrated by Merrill and Grimshaw~\cite{}.
+Due to their relatively low arithmetic intensity, it is theoretically expected and experimentally observed that the performance of local scans are memory-bound, as first demonstrated by Merrill and Grimshaw~\cite{}.
 Consequently, once equipped with a sufficient local scan strategy, the focus of scan design shifts to addressing the challenges of the global strategy, particularly inter-workgroup coordination and synchronization.
-In this section, we review the three prominent GPU global scan strategies---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---examining how they navigate inter-workgroup dependencies and evaluating their effectiveness in minimizing global memory access and maximizing memory bandwidth utilization in alignment with our ideal architecture goals.
+In this section, we review the three prominent GPU global scan strategies---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---examining how they navigate inter-workgroup dependencies and evaluating their effectiveness in minimizing global memory access and maximizing memory bandwidth utilization.
 
@@
 \subsubsection{Reduce-then-Scan}
-If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like \emph{scan} can profit by trading additional computation for reduced memory traffic.
+If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like scan can profit by trading additional computation for reduced memory traffic.
 This trade-off led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill-Grimshaw, Ha-and-Han, Dotsenko, Breitbart}: by writing only the work tile reduction to global memory in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated.
@@
 First introduced by Yan et al.~\cite{} in \emph{Stream-Scan}, the key innovation of the \emph{Chained-Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level.
-Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and status flags to ensure coherent views of data.
+Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data.
+To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than arbitrary assignment based on virtualized workgroup indices.
 This shifts dependency resolution from explicit synchronization via barrier-like structures to implicit synchronization via the order in which workgroups access global data.
@@
 In \emph{StreamScan}, this is facilitated by a \emph{propagation} phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume.
-During this waiting period, the workgroup is continually scheduled, and the progress of other workgroups must be guaranteed to prevent the spinning workgroup from starving its dependent predecessors of execution, which could result in deadlock.
+During this waiting period, the workgroup is continually scheduled, and the \emph{progress} of other workgroups must be guaranteed to prevent the spinning workgroup from starving its dependent predecessors of execution, which could result in deadlock.
 
@@
 \subsubsection{Chained-Scan, Decoupled-Lookback}
-Building on \emph{Stream-Scan}, Merrill and Garland~\cite{} introduced the \emph{Decoupled-Lookback} technique, which separates reduction propagation and dependency resolution to address the bottleneck caused by propagation latency.
+Merrill and Garland~\cite{} introduced the \emph{Decoupled-Lookback} technique, which separates reduction propagation and dependency resolution to address the bottleneck caused by propagation latency.
 In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction.
 Work and memory accesses are constant-bounded to the workgroup occupancy $o$ by posting the inclusive reduction after the lookback.
-However, granting each workgroup control over its dependency resolution does not eliminate reliance on progress guarantees, as workgroups must still spin on tiles that have yet to post their reductions, risking deadlock.
+However, granting each workgroup control over its dependency resolution does not eliminate reliance on \emph{progress guarantees}, as workgroups must still spin on tiles that have yet to post their reductions, risking deadlock.
 
@@
 \subsection{Ideal Scan Architecture}
+\thomas{This will be rewritten.
+It is clear to me now that the focus should be on the tradeoffs between the Oversubcription and not-Oversubcription(Persistent-Thread or Fixed Launch) paradigms.
+We want to communicate to the reader that because we want portability, we want to remain within the Oversubcription paradigm.
+We LIKE Oversubcription.
+We want to reframe the downsides as "things that happen when you step outside of Oversubcription." I.e.
+when you step outside of Oversubcription, now YOU have to manage occupancy -> occupancy discovery.
+When you step outside of Oversubcription now YOU have to worry about launch-related global memory access paterns -> overpartitioning your rake.
+Then we want to say although Oversubcription precludes inter-workgroup barriers besides kernel launches, thats OK because Chained Scan does not require an explicit barrier.
+Thus, Chained-Scan is good \emph{because} it lets us stay Oversubscribed}.
 
@@
 This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups.
-Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns—or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations—may experience significant performance degradation.
+Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns---or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations---may experience significant performance degradation.
 
-As a consequence of its $2n$ global data movement and inherently serial ordering, \emph{Chained-Scan} is also suitable for \emph{in-situ} compaction and allocation.
-In previous architectures, such tasks often required additional kernel launches or complex bookkeeping to handle dependencies between workgroups.
-By contrast, \emph{Chained-Scan} efficiently propagates dependencies within a single kernel, minimizing overhead and enabling seamless integration of compaction and allocation operations.
+As a consequence of its single kernel execution and inherently serial ordering, \emph{Chained-Scan} is also suitable for \emph{in-situ} compaction and allocation.
+In previous architectures, such tasks required additional kernel launches and bookkeeping to handle dependencies between workgroups.
+By contrast, \emph{Chained-Scan} efficiently propagates dependencies within a single kernel, enabling its application to a variety of practical routines, including data deduplication, sparse data compaction, error correction, and interval merging.
 
 But, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in terms of its theoretical and implementation characteristics, it falls short in meeting the criterion of \emph{portability}.
-This limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA devices but notably lacking on Apple's M-series and certain ARM GPUs~\cite{sorensen 2}.
-Without this guarantee, \emph{Chained-Scan} is highly prone to catastrophic failure.
+This limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA devices but notably lacking on Apple's M-series and ARM GPUs~\cite{sorensen 2}.
+Without this guarantee, \emph{Chained-Scan} is highly prone to catastrophic failure in the form of deadlocking.
 
 \subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
-A \emph{forward-progress guarantee} (FPG) is a promise provided by the subgroup scheduler that all occupant subgroups eventually make progress toward termination, preventing scenarios where a subgroup is starved of execution time.
+The challenges faced by \emph{Chained-Scan} are analogous to the \emph{Dining Philosophers Problem}, where tasks compete for shared resources, potentially resulting in deadlock.
+Coffman et al.~\cite{10.1145/356586.356588} define four conditions necessary for deadlock:
+\begin{enumerate}
+  \item \textbf{Mutual Exclusion}: Tasks claim exclusive control of resources they require.
+  \item \textbf{Resource Holding}: Tasks hold acquired resources while waiting.
+  \item \textbf{No Preemption}: A resource cannot be removed from a task before completion.
+  \item \textbf{Circular Dependency}:
+  \begin{enumerate}
+    \item Dependency Relationships: A dependency exists such that each task holds one or more resources that are required by proceeding tasks.
+    \item Circular Ordering: A circular chain of tasks exists.
+  \end{enumerate}
+\end{enumerate}
+Applying this schema to \emph{Chained-Scan}, workgroups correspond to "tasks" and work-tile reductions to "resources." Conditions 1-3 are easily satisfied due to the one-to-one mapping of work-tiles to workgroups.
+Similarly, condition 4a is met by the dependency relationships between work tiles, leaving condition 4b---circular ordering---as the key concern.
+
+Although the serial assignment of work tiles ensures that predecessor work tiles are either scheduled or completed, there is no guarantee that the corresponding reductions will be available when a workgroup begins its lookback.
+This uncertainty arises because workgroups are scheduled independently \thomas{Possibly confusing.
+.
+.
+Not in lockstep?}, and a predecessor workgroup might not yet have had the opportunity to compute and post its reduction.
+In such cases, the dependent workgroup enters a waiting state, spinning until the reduction becomes available.
+During this waiting period, the spinning workgroup remains scheduled, consuming computational resources.
+Without \emph{forward-progress guarantees}, the scheduler may schedule the spinning workgroup indefinitely over its predecessor, preventing the predecessor from completing its reduction.
+This satisfies condition 4b, creating a circular dependency that results in deadlock within the kernel.
+
+A \emph{forward-progress guarantee} (FPG) is a promise provided by the subgroup scheduler that all occupant subgroups (of possibly different workgroups) eventually make progress toward termination, preventing scenarios where a subgroup is starved of execution time.
 Sorensen et al.~\cite{barriers, 1, 2} provide the most comprehensive analysis of GPU schedulers.
-However, while they classify \emph{Chained-Scan} as an FPG-reliant algorithm, their primary focus is on exploring and formalizing progress models and their implications for synchronization in GPU programming frameworks.
+However, while they classify \emph{Chained-Scan} as an FPG-reliant algorithm, their primary focus is in exploring and formalizing progress models, rather than their application to any one algorithm.
+In the context of \emph{Chained-Scan} deadlocking, FPG prevents condition 4b by prohibiting indefinite scheduling of any single subgroup, thereby breaking the circular dependency chain.
 \thomas{Sorensen places Chained-Scan into the HSA category, but it is unclear if this is correct.
@@
 
-In \emph{Chained-Scan}, workgroups operate in a producer-consumer framework: each workgroup produces the reduction of its assigned work tile, while subsequent workgroups consume the reductions of all preceding workgroups to complete the scan.
-To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than static assignment based on virtualized workgroup indices.
-While this dynamic assignment guarantees that predecessor work tiles are either scheduled or completed, without FPG there is no guarantee that these reductions will eventually become available.
-A scheduler lacking FPG may fail to allocate sufficient execution time to certain workgroups, leaving their reductions incomplete indefinitely.
-In such cases, consuming workgroups stall during their lookback phase, waiting for reductions that are never posted to global memory.
-
-For instance, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor.
+\thomas{This example may be redundant now.}
+For example, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor.
 Although the multiprocessor can host both workgroups, it has only one scheduling unit, which lacks FPG\@.
@@
 However, because the scheduler lacks FPG, it may continue scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction.
-\john{We also want to avoid a reader thinking ``well, why not just schedule one work tile per multiprocessor, doesn't that solve the problem?''.} \thomas{I actually do not know the answer to this question.
-My impression is that (lack of) FPG is purely a phenomenon of the subgroup scheduler, and thus pigeonholing a multiprocessor with a single workgroup \emph{would} solve the problem.
-Of course this would be totally infeasible to implement due to varied multiprocessor resources.
-HOWEVER, do workgroups execute \emph{completely} indepently of one another?} As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.
+As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.
 
 \subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
-Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---its absence across all hardware is a critical issue for developers.
+Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---its absence on key segments of hardware is a critical issue for developers.
 Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad,\thomas{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time.
@@
 Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress requirements in \emph{ScalarFallback}.
-In \emph{ScalarFallback}, each time a workgroup spins on a preceding tile, it begins a \emph{fallback} operation, and begins consuming the blocking tile one element at a time.
+By allowing a workgroup to access and operate on another tile’s data, \emph{ScalarFallback} targets the mutual exclusion condition outlined by Coffman et al., ensuring that no single workgroup holds exclusive control over a resource.
+Instead, each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time.
 Although this entails performing redundant work when the blocking tile is simply late but not deadlocking, it guarantees the termination of the algorithm because a workgroup can perform the reduction of a blocking tile itself, breaking the inter-workgroup dependency.
@@
 
-\section{Decoupled-Fallback}
+\section{Design}
 \thomas{I've broken somewhat with the outline and moved the goals and non-goals section to somewhere in the introduction.
@@
 The placement of this section here feels more natural, and seems to mirror other papers?}
-Our \emph{Decoupled-Fallback} method is a further extension of the \emph{Chained-Scan} architecture.
+
+\subsection{Breaking Deadlocks}
+\thomas{What I want to accomplish here is to go beyond saying "Here is our method.
+It works great." I want to show that these are the possible avenues to attack deadlocking .
+.
+.
+and exhaustively demonstrate that removing  Coffman's mutual exclusion is the best way to do it, under our given constraints.}
+
+Deadlock in \emph{Chained-Scan} arises from the inter-workgroup dependencies that are inherent to scan operations.
+These dependencies are governed by Coffman et al.'s~\cite{10.1145/356586.356588} four necessary conditions for deadlock: \emph{mutual exclusion}, \emph{resource holding}, \emph{no preemption}, and \emph{circular dependency}.
+While all four conditions are theoretically avenues for addressing deadlock, practical constraints and the requirements of \emph{Chained-Scan} significantly narrow the set of viable approaches.
+
+The first condition, \emph{dependency relationships}, is immutable in the context of scan operations.
+Similarly, \emph{no preemption} is immutable within the GPU programming model.
+Addressing \emph{resource holding} by allowing workgroups to release resources (losing context) during waits effectively reintroduces kernel launches as inter-workgroup barriers.
+Designing tile sizes to break \emph{circular dependency} by mapping workgroups one-to-one with multiprocessors might seem feasible but is impractical.
+GPU vendors provide no guarantees regarding workgroup progress behavior, leaving starvation risks unresolved.
+Moreover, such strategies are non-portable and contradict the goal of supporting diverse hardware architectures.
+
+\subsection{Decoupled-Fallback}
 
@@
     \item Idea: Split the reductions into multiple values; use separate values for local/inclusive instead of using flags to differentiate them.
-    \item Pro: Full 32 or 64 bits.
+    \item Pro: Full 32 or theoretically 64 bits.
     \item Pro: No memory model or 64 bit atomics required!
     \item Pro: Can safely fallback insert without risking clobbering.
-    \item Con: 4x Memory usage
-    \item Con: Tad slower, more complicated.
+    \item Con: 2x Memory usage
+    \item Con: Places subgroup functions adjacent to atomics, which may lead to \textbf{\textit{interesting}} behavior.
 \end{enumerate}
@@
 Thus our implementation preserves asymptotically optimal work, even though the structure is not isomorphic to the radix-two network.
-
 The Kogge-Stone scan has a depth of $\log_2 s$ Thus, the total depth of the scan is:

### 2025-01-19 paper.tex
Cooking

@@
 Although the serial assignment of work tiles ensures that predecessor work tiles are either scheduled or completed, there is no guarantee that the corresponding reductions will be available when a workgroup begins its lookback.
-This uncertainty arises because workgroups are scheduled independently \thomas{Possibly confusing.
+This uncertainty arises because workgroups are scheduled individually \thomas{Possibly confusing.
 .
 .
+and perhaps not even true?
 Not in lockstep?}, and a predecessor workgroup might not yet have had the opportunity to compute and post its reduction.
@@
 The definition of HSA is \newline "Threads for which fair execution is guaranteed: The thread with the lowest id that has not yet terminated" \newline However, as Chained-Scan and other serially ordered algorithms no longer use the virtual ID provided by the workgroup scheduler, is fair execution still guaranteed under HSA?
-It seems that it should be classified under OBE.}
+It seems that it should be classified under OBE\@.}
 
-\thomas{This example may be redundant now.}
+\thomas{This example feels redundant now. Timeout detection and recovery can be moved into the 2nd paragraph in this section.}
 For example, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor.
@@
 \subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
-Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---its absence on key segments of hardware is a critical issue for developers.
+Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---its absence in key segments of hardware is a critical issue for developers.
 Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad,\thomas{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time.
@@
 Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress requirements in \emph{ScalarFallback}.
-By allowing a workgroup to access and operate on another tile’s data, \emph{ScalarFallback} targets the mutual exclusion condition outlined by Coffman et al., ensuring that no single workgroup holds exclusive control over a resource.
+By allowing a workgroup to access and operate on another tile’s data, \emph{ScalarFallback} targets Coffman et al.'s \emph{mutual exclusion} condition.
 Instead, each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time.
-Although this entails performing redundant work when the blocking tile is simply late but not deadlocking, it guarantees the termination of the algorithm because a workgroup can perform the reduction of a blocking tile itself, breaking the inter-workgroup dependency.
+Although this entails performing redundant work when the blocking tile is simply late but not deadlocking, it guarantees the termination of the algorithm because each workgroup can perform the reduction of any blocking tile, breaking the inter-workgroup dependency.
 
@@
 
-\section{Design}
+\section{Decoupled-Fallback}
 \thomas{I've broken somewhat with the outline and moved the goals and non-goals section to somewhere in the introduction.
@@
 The placement of this section here feels more natural, and seems to mirror other papers?}
+\emph{Decoupled-Fallback} enhances the \emph{Scalar-Fallback} technique with significantly improved fallback performance.
+As an extension of the \emph{Chained-Scan} architecture, it preserves the advantages in speed and oversubscription, making it an ideal choice for our requirements.
 
-\subsection{Breaking Deadlocks}
 \thomas{What I want to accomplish here is to go beyond saying "Here is our method.
@@
 .
-and exhaustively demonstrate that removing  Coffman's mutual exclusion is the best way to do it, under our given constraints.}
+and exhaustively demonstrate that removing mutual exclusion is the best way to do it, under our given constraints.}
+Like \emph{Scalar-Fallback}, we choose \emph{mutual exclusion} as our approach to break deadlocking.
+\emph{Dependency relationships} and \emph{no preemption} are immutable characteristics of scan and the GPU programming model, eliminating them as viable avenues for breaking deadlocking.
+While it may be possible to break \emph{circular ordering} by designing tile sizes such that workgroups are mapped one-to-one to multiprocessors, doing so portably is infeasible.
+Moreover, vendor guidance on progress guarantees among multiprocessors is also lacking, leaving the same risks of deadlocking but at the multiprocessor level.
+While eliminating \emph{resource holding}---allowing workgroups to discard work tiles when they encounter a deadlock---is theoretically possible, it is strictly worse than eliminating \emph{mutual exclusion}, as it necessitates $t$ additional movements (discard-read-read versus read).
 
-Deadlock in \emph{Chained-Scan} arises from the inter-workgroup dependencies that are inherent to scan operations.
-These dependencies are governed by Coffman et al.'s~\cite{10.1145/356586.356588} four necessary conditions for deadlock: \emph{mutual exclusion}, \emph{resource holding}, \emph{no preemption}, and \emph{circular dependency}.
-While all four conditions are theoretically avenues for addressing deadlock, practical constraints and the requirements of \emph{Chained-Scan} significantly narrow the set of viable approaches.
+To maximize the efficiency of the fallback operation, we employ three strategies.
+First, although a lookback is performed by a single thread or subgroup, we want all threads in the workgroup to participate.
+To facilitate this, we use a workgroup-wide lock to allow threads participating in the lookback to signal the entire workgroup when a fallback is necessary.
+Second, we explicitly limit the number of times a workgroup may spin while waiting on a tile.
+If the maximum spin count is exceeded, the fallback operation is initiated.
+Although this approach still risks performing redundant fallbacks---since a work tile that is late is indistinguishable from one that is deadlocking---exposing the maximum spin count as a tunable parameter allows us to reduce the likelihood of false-positive fallbacks.
+Finally, after completing a fallback, we attempt to insert the value into the spine at the position of the deadlocking tile, preventing succeeding workgroups from initiating redundant fallbacks.
 
-The first condition, \emph{dependency relationships}, is immutable in the context of scan operations.
-Similarly, \emph{no preemption} is immutable within the GPU programming model.
-Addressing \emph{resource holding} by allowing workgroups to release resources (losing context) during waits effectively reintroduces kernel launches as inter-workgroup barriers.
-Designing tile sizes to break \emph{circular dependency} by mapping workgroups one-to-one with multiprocessors might seem feasible but is impractical.
-GPU vendors provide no guarantees regarding workgroup progress behavior, leaving starvation risks unresolved.
-Moreover, such strategies are non-portable and contradict the goal of supporting diverse hardware architectures.
-
-\subsection{Decoupled-Fallback}
+The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled-Lookback}.
+Regardless of FPG, due to false positives, any given hardware will have a natural fallback-incidence rate $f$.
+Consequently, we expect memory bandwidth traffic to increase (and throughput to decrease) by $O(fn)$ to $O((2 + f)n)$.
+However, just as serial ordering guarantees that memory accesses to the spine fall within the $o$-width sliding window, memory accesses to the scan input are similarly guaranteed to fall within an $ot$-width sliding window, making them also highly likely to reside within some level of the cache.
+As we later empirically demonstrate, this cache residency significantly reduces the cost of memory access, such that the fallback operation is almost always bound by the compute required to reduce the blocking tile.
+As a pure reduction incurs very lower overhead, we observe throughput approaching $O(2n)$ on most devices.
 
 \section{Implementation}
-We are in WGSL. We don't have: a memory model, 64-bit atomics, and atomicCAS is unreliable.
+At a high level, our algorithmn follows a five phase construction:
+\begin{enumerate} 
+  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering.
+For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
+  \item \textbf{Subgroup-Rake}: Each thread performs a private, serial scan across its vectors and then participates in $p$ raking subgroup scans.
+This allows the entire scan across the subgroup-block to be barrier-free and remain in registers.
+The final subgroup-block reduction is then posted to shared memory.
+  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup reductions, writing the results back to shared memory upon completion. 
+  \item \textbf{Decoupled-Fallback}: The first subgroup in the workgroup performs a lookback operation, engaging in fallback operations if necessary.
+Once completed, it posts the results to shared memory for the workgroup to consume.
+  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output. 
+\end{enumerate}
+Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy.
+These phases are most similar to the \emph{CUDPP}~\cite{} scan kernel, with adaptations to enhance portability.
+Phase 4, \emph{Decoupled-Fallback}, is responsible for managing all inter-workgroup dependencies.
 
-\subsection{Split Method}
-\begin{enumerate}
-    \item Idea: Split the reductions into multiple values; use separate values for local/inclusive instead of using flags to differentiate them.
-    \item Pro: Full 32 or theoretically 64 bits.
-    \item Pro: No memory model or 64 bit atomics required!
-    \item Pro: Can safely fallback insert without risking clobbering.
-    \item Con: 2x Memory usage
-    \item Con: Places subgroup functions adjacent to atomics, which may lead to \textbf{\textit{interesting}} behavior.
-\end{enumerate}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{graphics/FlowChart.pdf}
-  \caption{TODO update for split}
-\end{figure}
-
-\subsection{Intra-Workgroup Scan}
-Our intra-workgroup implementation is most similar to the \emph{CUDPP}~\cite{} scan kernel, with adaptations to enhance portability. It consists of three phases:
-\begin{enumerate}
-  \item
-\end{enumerate}
-
-The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches.
-To handle this variability, we use a generalized radix-$s$ Ladner-Fischer scan network embedded with Kogge-Stone subgroup scans to perform the scan on the intra-workgroup spine.
-The upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction; assuming shared memory bank widths equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts.
-To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$.
-Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction.
-Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index.
-However, such access is interpreted by the hardware as a broadcast operation, resulting in no conflicts.
-Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; zero bank conflicts across all supported subgroup sizes.
-
+\subsection{Intra-Workgroup Portability Adaptations}
+As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations. 
+\subsubsection{Subgroup-Raking}
+We desire the smallest intra-workgroup spine possible.
+Not only does this reduce shared memory usage, but it also directly impacts the number of workgroup barriers encountered during the workgroup-wide scan, as this number is proportional to the size of the local spine.
+While large subgroups are relatively insensitive to this effect, small subgroups are particularly affected (reference depth here).
+To address this, we perform $p$ raking iterations of subgroup Kogge-Stone scans, where the highest-ranked thread passes its output into each succeeding scan.
+For a workgroup of size $w$, this approach reduces the number of reductions that each subgroup must post from $p$ to $1$, and it decreases the size of the spine from $\frac{pw}{s}$ to $\frac{w}{s}$.
 \begin{algorithm}[htbp]
+  \small
   \SetAlgoLined
@@
     \For{$j \gets S$ \KwTo $alignment$ \textbf{with} $j \gets j * S$}{
+      \tcp{Brent-Kung Upsweep}
 
@@
       \textbf{barrier()}\;
+
+      \tcp{Fanout}
 
@@
   \Return{$x$}\;
-  \caption{Subgroup-Size-Agnostic Scan}
+  \caption{Workgroup-Wide Scan}
   \label{alg:example}
+\end{algorithm}
+\subsubsection{Workgroup-Wide Scan}
+The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches.
+To handle this variability, we use a generalized radix-$s$ Ladner-Fischer scan network embedded with Kogge-Stone subgroup scans to perform the scan across the local spine.
+Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts.
+To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$.
+Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction.
+Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index.
+However, such access is interpreted by the hardware as a broadcast operation, resulting in no conflicts.
+Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; zero bank conflicts across all supported subgroup sizes.
+\footnote{Since we also require a workgroup-wide pure reduce operation, this can be achieved by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}
+
+\subsection{Decoupled-Fallback}
+To reiterate, the goal of a lookback is twofold: to determine the reduction of all preceding work tiles and to communicate those values to succeeding tiles to accelerate their own lookback operations.
+Following the \emph{Decoupled-Lookback} technique, we model the spine as a finite state machine, where each work tile comprising the spine may exist in one of three states:
+\begin{itemize}
+    \item \textbf{Not Ready}: The tile has not yet been processed or written to global memory.
+    \item \textbf{Ready}: The tile has completed its local reduction and posted it to global memory.
+    \item \textbf{Inclusive}: The tile has completed its lookback and updated its posting to reflect the inclusive reduction.
+\end{itemize}
+Each lookback is, therefore, a traversal from the \emph{Ready} state to the \emph{Inclusive} state.
+To illustrate this process, we present a flowchart (SEE LOOKBACK FLOWCHART).
+Prior to the kernel launch, all tiles in the spine are initialized to the \emph{Not Ready} state.
+As shown in the dotted section of the flowchart, fallbacks occur when a workgroup encounters a blocking tile and the maximum spin count is exceeded.
+
+\subsubsection{Tile State Implementation}
+To maintain a coherent view of the tile's state, two bits are required, corresponding to the three possible states described earlier.
+In \emph{Decoupled-Lookback}, Merrill and Garland propose two strategies for communicating tile state within a 32-bit value: Use two separate 32-bit values—one for the tile state and the other for the reduction value—combined with a \emph{memory fence}, ensuring that updating the tile state also synchronizes the reduction value.
+Use a single 64-bit value by bit-packing the tile state and reduction value, with updates managed via 64-bit atomics.
+
+In our case, neither are viable options.
+Due to a lack of support across backend APIs, WebGPU does not furnish developers with a memory model, and currently, 64-bit atomics are not supported.
+Moreover, 64-bit atomics are not supported across most mobile devices, and thus we do not consider their future support an appropriate option for our implementation.
+Instead, we use two 32-bit values and split the reduction into upper and lower 16-bit segments, reserving the two most significant bits of each segment to encode the tile state.
+\newline
+SPLIT BITS MINI DIAGRAM here
+\newline
+
+\begin{enumerate}
+  \item Idea: Split the reductions into multiple values; use separate values for local/inclusive instead of using flags to differentiate them.
+  \item Pro: Full 32 or theoretically 64 bits.
+  \item Pro: No memory model or 64 bit atomics required!
+  \item Pro: Can safely fallback insert without risking clobbering.
+  \item Con: 2x Memory usage
+  \item Con: Places subgroup functions adjacent to atomics, which may lead to \textbf{\textit{interesting}} behavior.
+\end{enumerate}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{graphics/FlowChart.pdf}
+  \caption{TODO update for split}
+\end{figure}
+
+\subsubsection{Full Algorithm}
+
+\begin{algorithm}[htbp]
+  \tiny
+  \SetAlgoLined
+  \DontPrintSemicolon
+
+  \KwIn{%
+    \emph{Partition ID} $part\_id$, 
+    \emph{Number of lanes} $lane\_count$, 
+    \emph{Workgroup-shared lock} $wg\_lock$, 
+    \emph{Atomic reduction array} \texttt{reduction}[], 
+    \emph{Local partial reductions} $wg\_reduce[]$, 
+    \emph{Constants:} $MAX\_SPIN\_COUNT, FLAG\_MASK, FLAG\_NOT\_READY, FLAG\_INCLUSIVE, FLAG\_READY, ALL\_READY, \dots$
+  }
+  \KwOut{Possible updates to $prev\_red$ or partial progress before fallback.}
+
+  \If(\tcp*[h]{Only needed if not the first partition}){$part\_id \neq 0$}{
+    $\,prev\_red \gets 0$\;
+    $\,lookback\_id \gets part\_id - 1$\;
+    $\,lock \gets \texttt{workgroupUniformLoad}(\&wg\_lock)$\;
+
+    \While(\tcp*[h]{Spin until we can proceed or must fallback}){$lock = \textnormal{LOCKED}$}{
+      \If(\tcp*[h]{Only threads within the subgroup size do checks}){$threadid.x < lane\_count$}{
+        $\,spin\_count \gets 0$\;
+        \While{$spin\_count < MAX\_SPIN\_COUNT$}{
+          $\,flag\_payload \gets \texttt{select}\bigl(
+             0,\;
+             \texttt{atomicLoad}(\&\texttt{reduction}[\,lookback\_id\,][\,threadid.x\,]),\;
+             threadid.x < \textnormal{SPLIT\_MEMBERS}
+           \bigr)$\;
+
+          \If{$\texttt{subgroupBallot}\bigl((\,flag\_payload \,\&\, FLAG\_MASK) > FLAG\_NOT\_READY\bigr).x = ALL\_READY$}{
+            $\,incl\_bal \gets \texttt{subgroupBallot}\bigl(
+               (\,flag\_payload \,\&\, FLAG\_MASK) = FLAG\_INCLUSIVE
+             \bigr).x$\;
+
+            \If{$incl\_bal \neq 0$}{
+              \tcp{Wait for all lanes to show FLAG\_INCLUSIVE}
+              \While{$incl\_bal \neq ALL\_READY$}{
+                $\,flag\_payload \gets \texttt{select}\bigl(
+                  0,\;
+                  \texttt{atomicLoad}(\&\texttt{reduction}[\,lookback\_id\,][\,threadid.x\,]),\;
+                  threadid.x < \textnormal{SPLIT\_MEMBERS}
+                \bigr)$\;
+                $\,incl\_bal \gets \texttt{subgroupBallot}\bigl(
+                  (\,flag\_payload \,\&\, FLAG\_MASK) = FLAG\_INCLUSIVE
+                \bigr).x$\;
+              }
+              $\,prev\_red \gets prev\_red + \texttt{join}\bigl(
+                 flag\_payload \,\&\, \textnormal{VALUE\_MASK},\;threadid.x
+               \bigr)$\;
+
+              \If{$threadid.x < \textnormal{SPLIT\_MEMBERS}$}{
+                $\,t \gets \texttt{split}\bigl(
+                  prev\_red + wg\_reduce[\mathit{spine\_size}-1],\;threadid.x
+                \bigr) \,\vert\, FLAG\_INCLUSIVE$\;
+                $\texttt{atomicStore}\bigl(
+                  \&\texttt{reduction}[\,part\_id\,][\,threadid.x\,],\; t
+                \bigr)$\;
+              }
+              \If{$threadid.x = 0$}{
+                $\,wg\_lock \gets \textnormal{UNLOCKED}$\;
+                $\,wg\_broadcast \gets prev\_red$\;
+              }
+              \textbf{break}\;  % Break out of spin loop
+            }
+            \Else{
+              $\,prev\_red \gets prev\_red + \texttt{join}\bigl(
+                 flag\_payload \,\&\, \textnormal{VALUE\_MASK},\;threadid.x
+               \bigr)$\;
+              $\,spin\_count \gets 0$\;
+              $\,lookback\_id \gets lookback\_id - 1$\;
+            }
+          }
+          \Else{
+            $\,spin\_count \gets spin\_count + 1$\;
+          }
+        }
+        \If(\tcp*[h]{Escalate fallback if spinning too long}){%
+          $\,threadid.x = 0 \,\wedge\, spin\_count = MAX\_SPIN\_COUNT$}{
+          $\,wg\_broadcast \gets lookback\_id$\;
+        }
+      }
+      \tcp{(At this point, if still locked, we consider fallback.)}
+      $\,lock \gets \texttt{workgroupUniformLoad}(\&wg\_lock)$\;
+      \textbf{break}\;
+    }
+  }
+  \caption{“Lookback” portion of the single-subgroup scan.}
+  \label{alg:lookback-portion}
+\end{algorithm}
+
+\begin{algorithm}[htbp]
+  \tiny
+  \SetAlgoLined
+  \DontPrintSemicolon
+
+  \KwIn{%
+    \emph{Workgroup lock} $wg\_lock$, 
+    \emph{Current broadcast index} $wg\_broadcast$, 
+    \emph{Local partial} $prev\_red$, 
+    \emph{Reduction array} \texttt{reduction}[], 
+    \emph{Local partials} $wg\_reduce[]$, 
+    \emph{Constants:} $FLAG\_MASK, FLAG\_READY, FLAG\_INCLUSIVE, \dots$
+  }
+  \KwOut{%
+    Possibly unlocked $wg\_lock$ and updated partition's inclusive data
+  }
+
+  \tcp{Check if we're still locked and must fallback}
+  \If{$\textnormal{lock} = \textnormal{LOCKED}$}{
+    $\,fallback\_id \gets wg\_broadcast$\;
+
+    \tcp{Invoke the fallback reduce subroutine}
+    $\,f\_red \gets \textbf{FallbackReduce}\bigl(fallback\_id,\, s\_offset,\, \dots\bigr)$\;
+
+    \If{$threadid.x < lane\_count$}{
+      $\,f\_split \gets \texttt{split}(f\_red,\, threadid.x)\,\vert\,
+        \texttt{select}(FLAG\_READY,\, FLAG\_INCLUSIVE,\,(fallback\_id=0))$\;
+
+      \tcp{Atomically update if in SPLIT\_MEMBERS}
+      $\,f\_payload \gets 0$\;
+      \If{$threadid.x < \textnormal{SPLIT\_MEMBERS}$}{
+        $\,f\_payload \gets \texttt{atomicMax}\bigl(
+           \&\texttt{reduction}[fallback\_id][threadid.x],\, f\_split
+         \bigr)$\;
+      }
+
+      $\,incl\_found \gets \Bigl(
+         \texttt{subgroupBallot}\bigl((\,f\_payload \,\&\, FLAG\_MASK) = FLAG\_INCLUSIVE\bigr).x 
+         = \textnormal{ALL\_READY}
+       \Bigr)$\;
+
+      \If{$incl\_found$}{
+        $\,prev\_red \gets prev\_red + \texttt{join}\bigl(
+           f\_payload \,\&\, \textnormal{VALUE\_MASK},\;threadid.x
+         \bigr)$\;
+      }
+      \Else{
+        $\,prev\_red \gets prev\_red + f\_red$\;
+      }
+
+      \If(\tcp*[h]{Either first partition or all splits are inclusive}){$(fallback\_id=0)\,\vee\,(incl\_found)$}{
+        \If{$threadid.x < \textnormal{SPLIT\_MEMBERS}$}{
+          $\,t \gets \texttt{split}\bigl(
+             prev\_red + wg\_reduce[\mathit{spine\_size}-1],\;threadid.x
+           \bigr)\,\vert\,FLAG\_INCLUSIVE$\;
+          $\texttt{atomicStore}\Bigl(
+             \&\texttt{reduction}[part\_id][threadid.x],\, t
+           \Bigr)$\;
+        }
+        \If{$threadid.x = 0$}{
+          $\,wg\_lock \gets \textnormal{UNLOCKED}$\;
+          $\,wg\_broadcast \gets prev\_red$\;
+        }
+      }
+      \Else{
+        $\,lookback\_id \gets lookback\_id - 1$\;
+      }
+    }
+
+    $\,lock \gets \texttt{workgroupUniformLoad}(\&wg\_lock)$\;
+  }
+
+  \caption{“Fallback” logic, calling \textbf{FallbackReduce} if still locked.}
+  \label{alg:fallback-logic}
 \end{algorithm}
@@
 
-\section{Experimental Setup}
+\section{Evaluation}
+256 threads, 16 elements per thread.
+Shared memory requirements.
+RTS implementation shares an identical local implementation.
+Uses Merrill-Grimshaw style raking, with our own occupancy discovery technique.
 
-\section{Results and Analysis}
+\subsection{Comparison to Reduce-Then-Scan}
+
+\subsection{How (un)fair is fair?}
+Show the stats of each device. How many lookbacks?  How many fallbacks initiated? How many successful fallbacks?
+
+\subsection{Emulated Deadlocking}
+Show the performance of devices as we artificially increase the fallback-incidence rate.
 
@@
 This, in turn, raises the number of lookback steps that may be needed and potentially leads to redundant fallbacks.
+
+\thomas{Somewhere in here is perfect for Raph's commentary on the speed of atomics on different hardware. Once we have it, we can retool this section around it.}
 

### 2025-01-20 paper.tex
Add Implementation Section. Refine FlowChart.

@@
 Like \emph{Scan-then-Propagate}, \emph{Reduce-then-Scan} employs a three-phase approach using kernel launches to resolve dependencies.
-However, implementations of \emph{Reduce-then-Scan} introduced new ways of processing arbitrarily-large inputs that improved over the previous recursive technique.
 \begin{enumerate}
@@
 \end{enumerate}
-Applying this schema to \emph{Chained-Scan}, workgroups correspond to "tasks" and work-tile reductions to "resources." Conditions 1-3 are easily satisfied due to the one-to-one mapping of work-tiles to workgroups.
+Applying this schema to \emph{Chained-Scan}, workgroups correspond to "tasks" and work-tile reductions to "resources." Conditions 1–3 are satisfied due to the one-to-one mapping of work tiles to workgroups and the constraints of the GPU programming model.
 Similarly, condition 4a is met by the dependency relationships between work tiles, leaving condition 4b---circular ordering---as the key concern.
 
-Although the serial assignment of work tiles ensures that predecessor work tiles are either scheduled or completed, there is no guarantee that the corresponding reductions will be available when a workgroup begins its lookback.
-This uncertainty arises because workgroups are scheduled individually \thomas{Possibly confusing.
-.
-.
-and perhaps not even true?
-Not in lockstep?}, and a predecessor workgroup might not yet have had the opportunity to compute and post its reduction.
+Although the serial assignment of work tiles ensures that predecessor work tiles are either scheduled or completed, there is no guarantee that the corresponding reductions will be available before a workgroup begins its lookback.
+This uncertainty arises because workgroups are scheduled individually, and a predecessor workgroup might not yet have had the opportunity to compute and post its reduction.
 In such cases, the dependent workgroup enters a waiting state, spinning until the reduction becomes available.
 During this waiting period, the spinning workgroup remains scheduled, consuming computational resources.
-Without \emph{forward-progress guarantees}, the scheduler may schedule the spinning workgroup indefinitely over its predecessor, preventing the predecessor from completing its reduction.
-This satisfies condition 4b, creating a circular dependency that results in deadlock within the kernel.
+Without \emph{forward-progress guarantees}, the scheduler may indefinitely prioritize the spinning workgroup over its predecessor, preventing the predecessor from completing its reduction.
+This creates a circular dependency, satisfying condition 4b and resulting in deadlock within the kernel.
 
@@
 Instead, each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time.
-Although this entails performing redundant work when the blocking tile is simply late but not deadlocking, it guarantees the termination of the algorithm because each workgroup can perform the reduction of any blocking tile, breaking the inter-workgroup dependency.
+Although this entails performing redundant work when the blocking tile is simply late but not blocking, it guarantees the termination of the algorithm because each workgroup can perform the reduction of any blocking tile, breaking the inter-workgroup dependency.
 
@@
 First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup engages in a waiting spin.
-This effectively takes $t$ spins to complete the fallback of a deadlocking $t$-sized tile.
+This effectively takes $t$ spins to complete the fallback of a blocking $t$-sized tile.
 Moreover, because only a single thread is responsible for loading elements, the cache line associated with the loaded elements is severely underutilized, further degrading performance.
@@
 While eliminating \emph{resource holding}---allowing workgroups to discard work tiles when they encounter a deadlock---is theoretically possible, it is strictly worse than eliminating \emph{mutual exclusion}, as it necessitates $t$ additional movements (discard-read-read versus read).
+\begin{algorithm}[htb]
+  \small
+  \SetAlgoLined
+  \KwIn{%
+    \texttt{in}[], \texttt{out}[], 
+    \texttt{tileBump}, 
+    \texttt{spine}[], 
+    \texttt{wg\_fallback}[], 
+    \texttt{wg\_lock}, 
+    \texttt{wg\_partials}[], 
+    \texttt{wg\_prev\_reduction},
+    \texttt{wg\_broadcast}, 
+  }
+  \KwOut{\texttt{out}[] fully inclusive scan of \texttt{in}[]}
 
+  \If{$\textnormal{threadid.x} = 0$}{
+    \texttt{wg\_broadcast} $\gets$ \texttt{atomicAdd}(\&\texttt{tileBump}, 1)\;
+    \texttt{wg\_lock} $\gets$ \textnormal{LOCKED}\;
+  }
+  \textbf{barrier()}\;
+
+  \texttt{part\_id} $\gets$ \texttt{wg\_broadcast}\;
+
+  \textit{local\_vectors}: $\textnormal{array}\langle \textnormal{vec4}\langle u32\rangle, p\rangle$\;
+
+  \textbf{Load}(\texttt{in}, \texttt{part\_id}, \textit{local\_vectors})\;
+  \textbf{SubgroupRake}(\textit{local\_vectors})\;
+
+  \If{$\text{\normalfont highestRankLane}(\textnormal{threadid.x})$}{
+    \texttt{wg\_partials}[\textnormal{subgroup\_id}] $\gets$ \textit{local\_vectors}[\textnormal{final\_partial}]\;
+  }
+  \textbf{barrier()}\;
+
+  \textbf{WorkgroupWideScan}(\texttt{wg\_partials})\;
+  \textbf{barrier()}\;
+
+  \tcp{Decoupled Fallback}
+  \While{\texttt{wg\_lock} = \textnormal{LOCKED}}{
+    \textbf{Lookback}(\texttt{spine}, \texttt{wg\_prev\_reduction}, \texttt{wg\_lock})\;
+    \texttt{lock} $\gets$ \textbf{\texttt{workgroupUniformLoad}}(\&\texttt{wg\_lock})\;
+
+    \If{$\texttt{lock} = \textnormal{LOCKED}$}{
+      \textbf{FallbackReduce}(\texttt{in}, \texttt{wg\_fallback})\;
+      \textbf{FallbackInsertionAttempt}(\texttt{spine}, \texttt{wg\_prev\_reduction}, \texttt{wg\_lock})\;
+      \texttt{lock} $\gets$ \textbf{\texttt{workgroupUniformLoad}}(\&\texttt{wg\_lock})\;
+    }
+  }
+
+  \textbf{Write}(\texttt{out}, \textit{local\_vectors}, \texttt{wg\_partials}, \texttt{wg\_prev\_reduction}, \texttt{part\_id})\;
+
+  \caption{High-level Decoupled Fallback}
+\end{algorithm}
 To maximize the efficiency of the fallback operation, we employ three strategies.
@@
 If the maximum spin count is exceeded, the fallback operation is initiated.
-Although this approach still risks performing redundant fallbacks---since a work tile that is late is indistinguishable from one that is deadlocking---exposing the maximum spin count as a tunable parameter allows us to reduce the likelihood of false-positive fallbacks.
-Finally, after completing a fallback, we attempt to insert the value into the spine at the position of the deadlocking tile, preventing succeeding workgroups from initiating redundant fallbacks.
+Although this approach still risks performing redundant fallbacks---since a work tile that is late is indistinguishable from one that is blocking---exposing the maximum spin count as a tunable parameter allows us to reduce the likelihood of false-positive fallbacks.
+Finally, after completing a fallback, we attempt to insert the value into the spine at the position of the blocking tile, preventing succeeding workgroups from initiating redundant fallbacks.
 
@@
 Consequently, we expect memory bandwidth traffic to increase (and throughput to decrease) by $O(fn)$ to $O((2 + f)n)$.
-However, just as serial ordering guarantees that memory accesses to the spine fall within the $o$-width sliding window, memory accesses to the scan input are similarly guaranteed to fall within an $ot$-width sliding window, making them also highly likely to reside within some level of the cache.
+However, just as serial ordering guarantees that memory accesses to the spine fall within an $o$-width sliding window, memory accesses to the scan input are similarly guaranteed to fall within an $ot$-width sliding window, making them also highly likely to reside within some level of the cache.
 As we later empirically demonstrate, this cache residency significantly reduces the cost of memory access, such that the fallback operation is almost always bound by the compute required to reduce the blocking tile.
@@
 \begin{enumerate}
+  \item[(0)] \textbf {Initialization}: The first thread in the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
   \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering.
@@
 The final subgroup-block reduction is then posted to shared memory.
-  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup reductions, writing the results back to shared memory upon completion. 
+  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion. 
   \item \textbf{Decoupled-Fallback}: The first subgroup in the workgroup performs a lookback operation, engaging in fallback operations if necessary.
@@
 \end{enumerate}
+Phase 0 ensures that the work tiles are consumed in serial order.
 Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy.
-These phases are most similar to the \emph{CUDPP}~\cite{} scan kernel, with adaptations to enhance portability.
 Phase 4, \emph{Decoupled-Fallback}, is responsible for managing all inter-workgroup dependencies.
@@
 To address this, we perform $p$ raking iterations of subgroup Kogge-Stone scans, where the highest-ranked thread passes its output into each succeeding scan.
-For a workgroup of size $w$, this approach reduces the number of reductions that each subgroup must post from $p$ to $1$, and it decreases the size of the spine from $\frac{pw}{s}$ to $\frac{w}{s}$.
+For a workgroup of size $w$, this approach reduces the number of partial reductions that each subgroup must post from $p$ to $1$, and it decreases the size of the spine from $\frac{pw}{s}$ to $\frac{w}{s}$.
 \begin{algorithm}[htbp]
@@
 However, such access is interpreted by the hardware as a broadcast operation, resulting in no conflicts.
+Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase is negligible due to our overall thriftiness, including the deliberate retention of vectors in registers and the efficiency of subgroup raking.
 Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; zero bank conflicts across all supported subgroup sizes.
@@
 \end{itemize}
-Each lookback is, therefore, a traversal from the \emph{Ready} state to the \emph{Inclusive} state.
-To illustrate this process, we present a flowchart (SEE LOOKBACK FLOWCHART).
-Prior to the kernel launch, all tiles in the spine are initialized to the \emph{Not Ready} state.
-As shown in the dotted section of the flowchart, fallbacks occur when a workgroup encounters a blocking tile and the maximum spin count is exceeded.
+Each lookback is, therefore, a traversal from the \emph{Ready} state to the \emph{Inclusive} state. To illustrate this process, we present a flowchart (SEE LOOKBACK FLOWCHART).
+
+\subsubsection{Decoupled-Fallback}
+\thomas{Given that "lock" has an extensive history of use in computer science and this is not that kind of lock, perhaps a better choice of phrase would be better. Updates to terminology incoming...}
+Prior to kernel launch, all tiles in the spine are initialized to the \emph{Not Ready} state.
+During partition index acquisition, we also initialize a shared memory \emph{control flag} to facilitate synchronization within the workgroup during the \emph{Decoupled-Fallback} phase.
+This flag governs both the initiation of fallback operations and the entry and exit of the entire phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{graphics/FlowChart.pdf}
+  \caption{Decoupled-Fallback}
+\end{figure}
+\begin{enumerate}
+  \item[(0)] \textbf{Posting Local Reduction:} Each workgroup posts its local reduction---the result of the workgroup-wide scan over partials---to global memory and updates its tile state to \emph{Ready}.
+If it is the first work tile, its local reduction \emph{is} its inclusive reduction.
+In this case, the workgroup updates its tile state directly to \emph{Inclusive} and bypasses the lookback entirely.
+
+  \item \textbf{Initialization:}
+  \begin{itemize}
+      \item The control flag, initialized to \emph{Locked}, signals that the workgroup should enter the lookback phase.
+      \item Each workgroup initializes a \emph{previous reduction} variable with the identity element of the monoid to hold the accumulated reductions of preceding tiles.
+      \item The workgroup prepares to traverse the spine, starting with its immediate predecessor tile.
+  \end{itemize}
+
+  \item \textbf{Lookback:}
+  \begin{itemize}
+      \item The workgroup queries the state of the current tile. 
+      \item If the tile is in the \emph{Ready} state, the workgroup accumulates the value and moves to the preceding tile, repeating this step.
+      \item If the tile is in the \emph{Inclusive} state, the workgroup accumulates the value and proceeds directly to step 5.
+      \item If the tile is in the \emph{Not Ready} state, the workgroup spins for a limited number of iterations, defined by the \emph{Maximum Spin Count}.
+If the spin count exceeds the \emph{Maximum Spin Count}, the workgroup signals a fallback operation using the control flag and proceeds to the next step.
+  \end{itemize}
+
+  \item \textbf{Fallback:}
+  \begin{itemize}
+      \item The workgroup computes the reduction of the blocking tile by processing the corresponding elements directly.
+      \item Once the fallback reduction is complete, the workgroup \emph{attempts} to post the reduction to global memory and update the tile's state to \emph{Ready}.
+      \item As this update attempt is performed atomically, the workgroup reads back the blocking tile's immediate state:
+      \begin{itemize}
+          \item If the state is \emph{Inclusive}, the fallback value is discarded, and the workgroup accumulates the value before proceeding to step 5.
+          \item If the blocking tile is the first tile in the spine, the workgroup has reached the start of the spine and proceeds to step 5.
+          \item Otherwise, the workgroup resumes the lookback, starting with the preceding tile.
+      \end{itemize}
+  \end{itemize}
+
+  \item \textbf{Exit:}
+  \begin{itemize}
+      \item Once the lookback or fallback is complete, the workgroup posts its inclusive reduction to global memory, transitioning its tile state to \emph{Inclusive}.
+      \item The \emph{previous reduction} is also posted to shared memory for the workgroup to consume and propagate as necessary.
+      \item The control flag is updated to \emph{Unlocked} to signal the exit of the \emph{Decoupled-Fallback} phase, allowing subsequent operations to proceed.
+  \end{itemize}
+\end{enumerate}
 
 \subsubsection{Tile State Implementation}
-To maintain a coherent view of the tile's state, two bits are required, corresponding to the three possible states described earlier.
-In \emph{Decoupled-Lookback}, Merrill and Garland propose two strategies for communicating tile state within a 32-bit value: Use two separate 32-bit values—one for the tile state and the other for the reduction value—combined with a \emph{memory fence}, ensuring that updating the tile state also synchronizes the reduction value.
-Use a single 64-bit value by bit-packing the tile state and reduction value, with updates managed via 64-bit atomics.
-
+To maintain a coherent view of the tile's state, two bits are required.
+In \emph{Decoupled-Lookback}, Merrill and Garland propose two strategies for communicating tile state and 32-bit reduction value: Two separate 32-bit values---one for the tile state and the other for the reduction value---with updates using a \emph{memory fence}.
+One single 64-bit value, bit-packing the tile state and reduction value, with updates via 64-bit atomics.
 In our case, neither are viable options.
@@
 Moreover, 64-bit atomics are not supported across most mobile devices, and thus we do not consider their future support an appropriate option for our implementation.
-Instead, we use two 32-bit values and split the reduction into upper and lower 16-bit segments, reserving the two most significant bits of each segment to encode the tile state.
+Instead, we use two 32-bit values and split the reduction into upper and lower 16-bit segments, reserving the two most significant bits of each 32-bit value to encode the tile state.
 \newline
@@
 \newline
+To accommodate split values, we use a single subgroup for the lookback, leveraging subgroup \emph{ballot} to verify that the split value state is consistent across threads, and subgroup \emph{shuffle} to join and split reductions as needed.
+Since this method requires one thread per 16 bits, and the minimum subgroup size in the WebGPU specification is 4, it supports both 32-bit and 64-bit values.
+Notably, our approach extends support to 64-bit scans, which would not be possible even with 64-bit atomics.\footnote{Although native 64-bit values are not currently supported by the WebGPU specification, their future inclusion is anticipated.}
 
-\begin{enumerate}
-  \item Idea: Split the reductions into multiple values; use separate values for local/inclusive instead of using flags to differentiate them.
-  \item Pro: Full 32 or theoretically 64 bits.
-  \item Pro: No memory model or 64 bit atomics required!
-  \item Pro: Can safely fallback insert without risking clobbering.
-  \item Con: 2x Memory usage
-  \item Con: Places subgroup functions adjacent to atomics, which may lead to \textbf{\textit{interesting}} behavior.
-\end{enumerate}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{graphics/FlowChart.pdf}
-  \caption{TODO update for split}
-\end{figure}
-
-\subsubsection{Full Algorithm}
-
-\begin{algorithm}[htbp]
-  \tiny
-  \SetAlgoLined
-  \DontPrintSemicolon
-
-  \KwIn{%
-    \emph{Partition ID} $part\_id$, 
-    \emph{Number of lanes} $lane\_count$, 
-    \emph{Workgroup-shared lock} $wg\_lock$, 
-    \emph{Atomic reduction array} \texttt{reduction}[], 
-    \emph{Local partial reductions} $wg\_reduce[]$, 
-    \emph{Constants:} $MAX\_SPIN\_COUNT, FLAG\_MASK, FLAG\_NOT\_READY, FLAG\_INCLUSIVE, FLAG\_READY, ALL\_READY, \dots$
-  }
-  \KwOut{Possible updates to $prev\_red$ or partial progress before fallback.}
-
-  \If(\tcp*[h]{Only needed if not the first partition}){$part\_id \neq 0$}{
-    $\,prev\_red \gets 0$\;
-    $\,lookback\_id \gets part\_id - 1$\;
-    $\,lock \gets \texttt{workgroupUniformLoad}(\&wg\_lock)$\;
-
-    \While(\tcp*[h]{Spin until we can proceed or must fallback}){$lock = \textnormal{LOCKED}$}{
-      \If(\tcp*[h]{Only threads within the subgroup size do checks}){$threadid.x < lane\_count$}{
-        $\,spin\_count \gets 0$\;
-        \While{$spin\_count < MAX\_SPIN\_COUNT$}{
-          $\,flag\_payload \gets \texttt{select}\bigl(
-             0,\;
-             \texttt{atomicLoad}(\&\texttt{reduction}[\,lookback\_id\,][\,threadid.x\,]),\;
-             threadid.x < \textnormal{SPLIT\_MEMBERS}
-           \bigr)$\;
-
-          \If{$\texttt{subgroupBallot}\bigl((\,flag\_payload \,\&\, FLAG\_MASK) > FLAG\_NOT\_READY\bigr).x = ALL\_READY$}{
-            $\,incl\_bal \gets \texttt{subgroupBallot}\bigl(
-               (\,flag\_payload \,\&\, FLAG\_MASK) = FLAG\_INCLUSIVE
-             \bigr).x$\;
-
-            \If{$incl\_bal \neq 0$}{
-              \tcp{Wait for all lanes to show FLAG\_INCLUSIVE}
-              \While{$incl\_bal \neq ALL\_READY$}{
-                $\,flag\_payload \gets \texttt{select}\bigl(
-                  0,\;
-                  \texttt{atomicLoad}(\&\texttt{reduction}[\,lookback\_id\,][\,threadid.x\,]),\;
-                  threadid.x < \textnormal{SPLIT\_MEMBERS}
-                \bigr)$\;
-                $\,incl\_bal \gets \texttt{subgroupBallot}\bigl(
-                  (\,flag\_payload \,\&\, FLAG\_MASK) = FLAG\_INCLUSIVE
-                \bigr).x$\;
-              }
-              $\,prev\_red \gets prev\_red + \texttt{join}\bigl(
-                 flag\_payload \,\&\, \textnormal{VALUE\_MASK},\;threadid.x
-               \bigr)$\;
-
-              \If{$threadid.x < \textnormal{SPLIT\_MEMBERS}$}{
-                $\,t \gets \texttt{split}\bigl(
-                  prev\_red + wg\_reduce[\mathit{spine\_size}-1],\;threadid.x
-                \bigr) \,\vert\, FLAG\_INCLUSIVE$\;
-                $\texttt{atomicStore}\bigl(
-                  \&\texttt{reduction}[\,part\_id\,][\,threadid.x\,],\; t
-                \bigr)$\;
-              }
-              \If{$threadid.x = 0$}{
-                $\,wg\_lock \gets \textnormal{UNLOCKED}$\;
-                $\,wg\_broadcast \gets prev\_red$\;
-              }
-              \textbf{break}\;  % Break out of spin loop
-            }
-            \Else{
-              $\,prev\_red \gets prev\_red + \texttt{join}\bigl(
-                 flag\_payload \,\&\, \textnormal{VALUE\_MASK},\;threadid.x
-               \bigr)$\;
-              $\,spin\_count \gets 0$\;
-              $\,lookback\_id \gets lookback\_id - 1$\;
-            }
-          }
-          \Else{
-            $\,spin\_count \gets spin\_count + 1$\;
-          }
-        }
-        \If(\tcp*[h]{Escalate fallback if spinning too long}){%
-          $\,threadid.x = 0 \,\wedge\, spin\_count = MAX\_SPIN\_COUNT$}{
-          $\,wg\_broadcast \gets lookback\_id$\;
-        }
-      }
-      \tcp{(At this point, if still locked, we consider fallback.)}
-      $\,lock \gets \texttt{workgroupUniformLoad}(\&wg\_lock)$\;
-      \textbf{break}\;
-    }
-  }
-  \caption{“Lookback” portion of the single-subgroup scan.}
-  \label{alg:lookback-portion}
-\end{algorithm}
-
-\begin{algorithm}[htbp]
-  \tiny
-  \SetAlgoLined
-  \DontPrintSemicolon
-
-  \KwIn{%
-    \emph{Workgroup lock} $wg\_lock$, 
-    \emph{Current broadcast index} $wg\_broadcast$, 
-    \emph{Local partial} $prev\_red$, 
-    \emph{Reduction array} \texttt{reduction}[], 
-    \emph{Local partials} $wg\_reduce[]$, 
-    \emph{Constants:} $FLAG\_MASK, FLAG\_READY, FLAG\_INCLUSIVE, \dots$
-  }
-  \KwOut{%
-    Possibly unlocked $wg\_lock$ and updated partition's inclusive data
-  }
-
-  \tcp{Check if we're still locked and must fallback}
-  \If{$\textnormal{lock} = \textnormal{LOCKED}$}{
-    $\,fallback\_id \gets wg\_broadcast$\;
-
-    \tcp{Invoke the fallback reduce subroutine}
-    $\,f\_red \gets \textbf{FallbackReduce}\bigl(fallback\_id,\, s\_offset,\, \dots\bigr)$\;
-
-    \If{$threadid.x < lane\_count$}{
-      $\,f\_split \gets \texttt{split}(f\_red,\, threadid.x)\,\vert\,
-        \texttt{select}(FLAG\_READY,\, FLAG\_INCLUSIVE,\,(fallback\_id=0))$\;
-
-      \tcp{Atomically update if in SPLIT\_MEMBERS}
-      $\,f\_payload \gets 0$\;
-      \If{$threadid.x < \textnormal{SPLIT\_MEMBERS}$}{
-        $\,f\_payload \gets \texttt{atomicMax}\bigl(
-           \&\texttt{reduction}[fallback\_id][threadid.x],\, f\_split
-         \bigr)$\;
-      }
-
-      $\,incl\_found \gets \Bigl(
-         \texttt{subgroupBallot}\bigl((\,f\_payload \,\&\, FLAG\_MASK) = FLAG\_INCLUSIVE\bigr).x 
-         = \textnormal{ALL\_READY}
-       \Bigr)$\;
-
-      \If{$incl\_found$}{
-        $\,prev\_red \gets prev\_red + \texttt{join}\bigl(
-           f\_payload \,\&\, \textnormal{VALUE\_MASK},\;threadid.x
-         \bigr)$\;
-      }
-      \Else{
-        $\,prev\_red \gets prev\_red + f\_red$\;
-      }
-
-      \If(\tcp*[h]{Either first partition or all splits are inclusive}){$(fallback\_id=0)\,\vee\,(incl\_found)$}{
-        \If{$threadid.x < \textnormal{SPLIT\_MEMBERS}$}{
-          $\,t \gets \texttt{split}\bigl(
-             prev\_red + wg\_reduce[\mathit{spine\_size}-1],\;threadid.x
-           \bigr)\,\vert\,FLAG\_INCLUSIVE$\;
-          $\texttt{atomicStore}\Bigl(
-             \&\texttt{reduction}[part\_id][threadid.x],\, t
-           \Bigr)$\;
-        }
-        \If{$threadid.x = 0$}{
-          $\,wg\_lock \gets \textnormal{UNLOCKED}$\;
-          $\,wg\_broadcast \gets prev\_red$\;
-        }
-      }
-      \Else{
-        $\,lookback\_id \gets lookback\_id - 1$\;
-      }
-    }
-
-    $\,lock \gets \texttt{workgroupUniformLoad}(\&wg\_lock)$\;
-  }
-
-  \caption{“Fallback” logic, calling \textbf{FallbackReduce} if still locked.}
-  \label{alg:fallback-logic}
-\end{algorithm}
-
-\subsection{Work Efficiency and Depth}
-\subsubsection{Subgroup Raking}
-
-\subsubsection{Size Agnostic Scan}
-For a subgroup of size $s$, the Kogge-Stone network has a work complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
-\begin{equation}
-  \begin{aligned}[b]
-    \text{Work} & = \underbrace{\vphantom{\sum_{k=0}^{\lceil \log_s n \rceil}}n}_{\text{fanout}}
-    +
-    \underbrace{\sum_{k=1}^{\lceil \log_s n \rceil}}_{\text{iterations}}
-    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}\frac{n}{s^k}}_{\text{calls per iteration}}
-    \cdot
-    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}s \log_2 s}_{\text{work per call}} \\
-                & = n + s \log_2 s \cdot \sum_{k=1}^{\lceil \log_s n \rceil} \frac{n}{s^k}       \\
-                & = O\left(n + n \log_2 s\right)                                                 \\
-                & = O(n \log_2 s).
-  \end{aligned}
-\end{equation}
-Since the subgroup size remains constant within each kernel launch, the final work complexity is $O(n)$.
-Thus our implementation preserves asymptotically optimal work, even though the structure is not isomorphic to the radix-two network.
-The Kogge-Stone scan has a depth of $\log_2 s$ Thus, the total depth of the scan is:
-\begin{equation}
-  \begin{aligned}[b]
-    \text{Depth} & = \underbrace{\lceil \log_s n \rceil}_{\text{loop iterations}}
-    \cdot \underbrace{\log_2 s}_{\text{depth per iteration}}                           \\
-                 & = \left\lceil \frac{\log_2 n}{\log_2 s} \right\rceil \cdot \log_2 s \\
-                 & = \log_2 n + O(1).
-  \end{aligned}
-\end{equation}
-While minimal depth is preserved regardless of the subgroup size, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a workgroup-wide barrier at each iteration.
-Consequently, while the theoretical depth remains constant across subgroup sizes, smaller subgroups tend to perform worse in practice due to the increased relative cost of synchronization and the overhead of additional iterations in the main loop.
-
-\subsubsection{Full Algorithm}
+In the context of fallbacks, the blocking tile can exist in any of three states—\emph{Not Ready}, \emph{Ready}, or \emph{Inclusive}—due to several factors.
+A false-positive fallback may occur when the blocking tile completes its local or inclusive reduction before the fallback finishes.
+Additionally, multiple workgroups are allowed to perform fallbacks on the same blocking tile, meaning another workgroup may have already updated the tile state.
+To handle these scenarios, we require a tile state representation and an accompanying atomic operation that ensures a fallback can safely transition a tile's state to \emph{Ready} without overwriting a more advanced state.
+This challenge is further compounded by the fact that in WebGPU, the correctness of atomic \emph{Compare-and-Swap} is not guaranteed across all devices, prohibiting our use.
+Instead, we encode the tile state within the most significant bits of the split value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}.
+This design enables the use of atomic \emph{Maximum}, which inherently respects the state hierarchy, to safely insert values.
+\thomas{Possibly a blurb here about no volatile, so we use all atomics?}
 
@@
 There are a number of factors which may preclude speed-of-light performance, but most salient is the fairness of the scheduling model.
-In \emph{Decoupled-Fallback}, a work tile which is late due to unfairness is indistinguishable from one which is deadlocking, so as a scheduler becomes increasingly unfair, it incurs an increasing number of redundant fallbacks.
+In \emph{Decoupled-Fallback}, a work tile which is late due to unfairness is indistinguishable from one which is blocking, so as a scheduler becomes increasingly unfair, it incurs an increasing number of redundant fallbacks.
 Consider a hardware with a workgroup occupancy denoted by $o$, and let $f$ represent the probability that a fallback operation occurs at a particular lookback step.
@@
 
+\subsection{Work Efficiency and Depth}
+\subsubsection{Size Agnostic Scan}
+For a subgroup of size $s$, the Kogge-Stone network has a work complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
+\begin{equation}
+  \begin{aligned}[b]
+    \text{Work} & = \underbrace{\vphantom{\sum_{k=0}^{\lceil \log_s n \rceil}}n}_{\text{fanout}}
+    +
+    \underbrace{\sum_{k=1}^{\lceil \log_s n \rceil}}_{\text{iterations}}
+    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}\frac{n}{s^k}}_{\text{calls per iteration}}
+    \cdot
+    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}s \log_2 s}_{\text{work per call}} \\
+                & = n + s \log_2 s \cdot \sum_{k=1}^{\lceil \log_s n \rceil} \frac{n}{s^k}       \\
+                & = O\left(n + n \log_2 s\right)                                                 \\
+                & = O(n \log_2 s).
+  \end{aligned}
+\end{equation}
+Since the subgroup size remains constant within each kernel launch, the final work complexity is $O(n)$.
+Thus our implementation preserves asymptotically optimal work, even though the structure is not isomorphic to the radix-two network.
+The Kogge-Stone scan has a depth of $\log_2 s$ Thus, the total depth of the scan is:
+\begin{equation}
+  \begin{aligned}[b]
+    \text{Depth} & = \underbrace{\lceil \log_s n \rceil}_{\text{loop iterations}}
+    \cdot \underbrace{\log_2 s}_{\text{depth per iteration}}                           \\
+                 & = \left\lceil \frac{\log_2 n}{\log_2 s} \right\rceil \cdot \log_2 s \\
+                 & = \log_2 n + O(1).
+  \end{aligned}
+\end{equation}
+While minimal depth is preserved regardless of the subgroup size, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a workgroup-wide barrier at each iteration.
+Consequently, while the theoretical depth remains constant across subgroup sizes, smaller subgroups tend to perform worse in practice due to the increased relative cost of synchronization and the overhead of additional iterations in the main loop.
 \begin{acks}

### 2025-01-21 paper.tex
fixing build errors, can't have spaces or parens in citekeys

@@
 Harris~\cite{1292373} offers a taxonomy that relates depth, fanout, and wire tracks, while Hinze~\cite{10.1007/978-3-540-27764-4_11} develops an algebraic framework for scans.
-Snir~\cite{10.1016/0196-6774(86)90003-9} proved that depth $d$ and size $s$ are related by $s + d \ge 2n - 2$, and Fich~\cite{10.1145/800061.808738} proved that among minimum depth scans Ladner-Fischer~\cite{10.1145/322217.322232} networks have optimal size.
+Snir~\cite{10.1016/0196-67748690003-9} proved that depth $d$ and size $s$ are related by $s + d \ge 2n - 2$, and Fich~\cite{10.1145/800061.808738} proved that among minimum depth scans Ladner-Fischer~\cite{10.1145/322217.322232} networks have optimal size.
 Blelloch~\cite{} adapted scan to the PRAM model and popularized its use as an algorithmic primitive.
@@
 But, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in terms of its theoretical and implementation characteristics, it falls short in meeting the criterion of \emph{portability}.
-This limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA devices but notably lacking on Apple's M-series and ARM GPUs~\cite{sorensen 2}.
+This limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA devices but notably lacking on Apple's M-series and ARM GPUs~\cite{sorensen2}.
 Without this guarantee, \emph{Chained-Scan} is highly prone to catastrophic failure in the form of deadlocking.

### 2025-01-22 paper.tex
Revisions

@@
 \usepackage[ruled,vlined]{algorithm2e}
+\SetKwInput{KwGlobalIn}{Global Mem Input}
+\SetKwInput{KwSharedIn}{Shared Mem Input}
+\SetKwInput{KwSharedOut}{Shared Mem Output}
+\SetKwInput{KwGlobalOut}{Global Mem Output}
+\SetKwInput{KwConstants}{Constants}
+
 \usepackage{amsmath}
@@
 In the oversubscription paradigm, where the number of workgroups is theoretically unbound and typically far exceeds the available multiprocessors, every invocation of such a barrier would necessitate context-switching between all launched workgroups.
-Due to this high cost, no GPU programming environment furnishes developers with a true unbounded-launch inter-workgroup barrier.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{}, but it is almost certainly designed to operate within the \emph{Persistent-Thread}~\cite{} paradigm, rather than serving as a barrier across an unbounded launch.} Instead, when inter-workgroup synchronization is necessary, developers must rely on kernel launches as synchronization points.
+Due to this high cost, no GPU programming environment furnishes developers with a true unbounded-launch inter-workgroup barrier.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent-Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} Instead, when inter-workgroup synchronization is necessary, developers must rely on kernel launches as synchronization points.
 However, kernel launches are suboptimal as barriers: they incur overheads due to the heterogeneous nature of the operation, and all execution context is lost between launches.
@@
 \subsubsection{Fairness and Progress}
-At the subgroup scheduling level, two key issues emerge: \emph{fairness}---how evenly execution resources are distributed among threads---and \emph{progress guarantees}---ensuring that all threads or subgroups eventually make progress towards termination.
+Because multiple workgroups may reside on a single multiprocessor, two key issues arise at the subgroup scheduling level: \emph{fairness}---how evenly execution resources are distributed among threads---and \emph{progress guarantees}---ensuring that all threads or subgroups eventually make progress towards termination.
 However, the term "fairness" is used differently across the literature, leading to potential confusion.
-Sorensen et al.~\cite{10.1145/3022671.2984032, sorensen_et_al:LIPIcs.CONCUR.2018.23, 10.1145/3485508}, whose work provides the most comprehensive taxonomy of progress models to date, use "fairness" specifically to describe differing levels of progress guarantees.
-In contrast, NVIDIA publications~\cite{4523358, Merrill2016} define "fairness" in terms of the distribution of execution resources among subgroups.
+Sorensen et al.\cite{sorensen2016,sorensen2018,sorensen2021}, whose work provides the most comprehensive taxonomy of progress models to date, use "fairness" specifically to describe differing levels of progress guarantees.
+In contrast, NVIDIA publications~\cite{4523358,Merrill2016} define "fairness" in terms of the distribution of execution resources among subgroups.
 In this work, we adopt NVIDIA's terminology to maintain consistency with their publications, while acknowledging the broader definitions provided by Sorensen et al.
@@
 \subsubsection{Scan-then-Propagate}
-Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2008} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases:
+Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2011} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases:
 \begin{enumerate}
@@
 If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like scan can profit by trading additional computation for reduced memory traffic.
-This trade-off led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill-Grimshaw, Ha-and-Han, Dotsenko, Breitbart}: by writing only the work tile reduction to global memory in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated.
+This trade-off led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{10.1145/1375527.1375559, Merrill2009, 10.1109/TPDS.2012.336, 10.5555/2031978.2032029}: by writing only the work tile reduction to global memory in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated.
 Like \emph{Scan-then-Propagate}, \emph{Reduce-then-Scan} employs a three-phase approach using kernel launches to resolve dependencies.
@@
 
-Merill and Grimshaw~\cite{} introduced a workgroup \emph{raking} technique that elides the overhead incurred when the recursive depth exceeds two.
+Merill and Grimshaw~\cite{Merrill2009} introduced a workgroup \emph{raking} technique that elides the overhead incurred when the recursive depth exceeds two.
 In this approach, the total workgroup occupancy $o$ is determined, and the input is partitioned into non overlapping blocks of size $\frac{n}{t \cdot o}$, which are raked serially by each workgroup.
@@
 
-Breitbart~\cite{} was the first to apply more advanced inter-workgroup synchronization techniques to scan.
-Recognizing that workgroup raking naturally aligns with a \emph{Persistent-Threads}~\cite{} paradigm, Breitbart utilized an atomics-based inter-workgroup barrier to replace kernel launches.
-Since multiprocessors are not oversubscribed, workgroups remain resident (persist) on the multiprocessors for the entire lifetime of the kernel, enabling synchronization at the barrier without requiring context switching.
+Breitbart~\cite{10.5555/2031978.2032029} was the first to apply more advanced inter-workgroup synchronization techniques to scan.
+Recognizing that workgroup raking naturally aligns with a \emph{Persistent-Threads}~\cite{gupta2012} paradigm, Breitbart utilized an atomics-based inter-workgroup barrier to replace kernel launches.
+Since multiprocessors are not oversubscribed in raking, workgroups remain resident (persist) on the multiprocessors for the entire lifetime of the kernel, enabling synchronization at the barrier without requiring context switching.
 Although global data movement remains at $O(3n)$, this consolidates the scan to a single kernel, minimizing launch overheads.
@@
 \subsubsection{Chained-Scan, Stream-Scan}
-First introduced by Yan et al.~\cite{} in \emph{Stream-Scan}, the key innovation of the \emph{Chained-Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level.
+First introduced by Yan et al.~\cite{10.1145/2442516.2442539} in \emph{Stream-Scan}, the key innovation of the \emph{Chained-Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level.
 Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data.
@@
 While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently saturate global memory bandwidth, preventing it from achieving speed-of-light performance.
-This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the reduction of its predecessor.
+This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the result of its immediate predecessor.
 While this results in efficient $O(n/t)$ communication of reductions, and $O(2n+ n/t)= O(2n)$ total data movement, the sequential dependency creates a bottleneck that limits performance to the rate at which reductions propagate through device memory.
@@
 \subsubsection{Chained-Scan, Decoupled-Lookback}
-Merrill and Garland~\cite{} introduced the \emph{Decoupled-Lookback} technique, which separates reduction propagation and dependency resolution to address the bottleneck caused by propagation latency.
+Merrill and Garland~\cite{Merrill2016} introduced the \emph{Decoupled-Lookback} technique, which separates reduction propagation and dependency resolution to address the bottleneck caused by propagation latency.
 In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction.
@@
 \subsection{Ideal Scan Architecture}
-\thomas{This will be rewritten.
-It is clear to me now that the focus should be on the tradeoffs between the Oversubcription and not-Oversubcription(Persistent-Thread or Fixed Launch) paradigms.
-We want to communicate to the reader that because we want portability, we want to remain within the Oversubcription paradigm.
-We LIKE Oversubcription.
-We want to reframe the downsides as "things that happen when you step outside of Oversubcription." I.e.
-when you step outside of Oversubcription, now YOU have to manage occupancy -> occupancy discovery.
-When you step outside of Oversubcription now YOU have to worry about launch-related global memory access paterns -> overpartitioning your rake.
-Then we want to say although Oversubcription precludes inter-workgroup barriers besides kernel launches, thats OK because Chained Scan does not require an explicit barrier.
-Thus, Chained-Scan is good \emph{because} it lets us stay Oversubscribed}.
+The design of an ideal scan architecture must balance performance, portability, and practicality.
+Beyond their differences in speed, \emph{Chained-Scan} and \emph{Reduce-Then-Scan} embody two fundamentally contrasting paradigms.
+By eliding explicit inter-workgroup synchronization, \emph{Chained-Scan} aligns naturally with the oversubscription paradigm, delegating occupancy management to the hardware scheduler.
+In contrast, the raking approach inherent to \emph{Reduce-Then-Scan} mandates the fixed-launch paradigm, necessitating explicit control over occupancy and potentially synchronization.
 
-In addition to its speed, the \emph{Chained-Scan} architecture has more practical properties that improve over its predecessors.
-As it is a single kernel, it minimizes launch overheads and simplifies maintenance.
-On previous architectures, handling of arbitrary input sizes can be awkward: the recursive approach, for example, entails multiple kernel launches and a heterogeneous system to coordinate them.
-Similarly, the raking approach—a necessary optimization in methods prior to \emph{Chained-Scan}—can be cumbersome, as predetermining static workgroup occupancies for every target device is infeasible.
-Previous workgroup occupancy discovery techniques, such as the one proposed by Sorensen et al.~\cite{}, are impossible to implement in WebGPU as they require at least a memory fence, which is not supported by the specification.\footnote{Our implementation of \emph{Reduce-Then-Scan}, which can be found in the artifact, includes a fenceless occupancy discovery method amenable to the WebGPU environment, but its description is beyond the scope of this work.}
+Fixed-launch approaches demand substantial developer effort, requiring either tuning of workgroup launch sizes across target devices or occupancy discovery.
+Moreover, previous workgroup occupancy discovery techniques, such as the one proposed by Sorensen et al.~\cite{sorensen2016}, are infeasible to implement in WebGPU due to their reliance on memory fences---a feature not supported by the specification.\footnote{Our implementation of \emph{Reduce-Then-Scan}, which can be found in the artifact, includes a fenceless occupancy discovery method amenable to the WebGPU environment, but its description is beyond the scope of this work.}
 
-Additionally, on GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—raking the workgroups can lead to performance degradation due to cache thrashing.
+Hardware-level considerations further underscore the limitations of fixed-launches.
+On GPUs where the global memory data cache is indexed by physical addresses---and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)---fixed-launches can lead to performance degradation due to cache thrashing.
 This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups.
@@
 
-As a consequence of its single kernel execution and inherently serial ordering, \emph{Chained-Scan} is also suitable for \emph{in-situ} compaction and allocation.
-In previous architectures, such tasks required additional kernel launches and bookkeeping to handle dependencies between workgroups.
-By contrast, \emph{Chained-Scan} efficiently propagates dependencies within a single kernel, enabling its application to a variety of practical routines, including data deduplication, sparse data compaction, error correction, and interval merging.
+Thus, oversubcription and \emph{Chained-Scan} emerges as the preferred architecture, as it naturally aligns with portability by leveraging hardware schedulers to dynamically manage occupancy across diverse architectures without requiring developer intervention.
+Furthermore, as a single kernel with inherently serial ordering, \emph{Chained-Scan} is suitable for \emph{in-situ} operations, augmenting practical routines such as data deduplication, sparse data compaction, error correction, and interval merging.
 
-But, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in terms of its theoretical and implementation characteristics, it falls short in meeting the criterion of \emph{portability}.
-This limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA devices but notably lacking on Apple's M-series and ARM GPUs~\cite{sorensen2}.
+But, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in terms of its theoretical and implementation characteristics, it falls short in fully meeting the criterion of \emph{portability}.
+This limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA devices but notably lacking on Apple's M-series and ARM GPUs~\cite{sorensen2021}.
 Without this guarantee, \emph{Chained-Scan} is highly prone to catastrophic failure in the form of deadlocking.
@@
 \subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
+A \emph{forward-progress guarantee} (FPG) is a promise provided by the subgroup scheduler that all occupant subgroups (of possibly different workgroups) eventually make progress toward termination, preventing scenarios where a subgroup is starved of execution time.
+Sorensen et al.~\cite{sorensen2018,sorensen2021} provide the most comprehensive analysis of GPU schedulers.
+However, while they classify \emph{Chained-Scan} as an FPG-reliant algorithm, their primary focus is in exploring and formalizing progress models, rather than their application to any one algorithm.
+\thomas{Sorensen places Chained-Scan into the HSA category, but it is unclear if this is correct.
+The definition of HSA is \newline "Threads for which fair execution is guaranteed: The thread with the lowest id that has not yet terminated" \newline However, as Chained-Scan and other serially ordered algorithms no longer use the virtual ID provided by the workgroup scheduler, is fair execution still guaranteed under HSA?
+It seems that it should be classified under OBE\@.}
+
 The challenges faced by \emph{Chained-Scan} are analogous to the \emph{Dining Philosophers Problem}, where tasks compete for shared resources, potentially resulting in deadlock.
@@
 Applying this schema to \emph{Chained-Scan}, workgroups correspond to "tasks" and work-tile reductions to "resources." Conditions 1–3 are satisfied due to the one-to-one mapping of work tiles to workgroups and the constraints of the GPU programming model.
-Similarly, condition 4a is met by the dependency relationships between work tiles, leaving condition 4b---circular ordering---as the key concern.
+Similarly, condition 4a is met by the dependency relationships between work tiles, leaving condition 4b---\emph{circular ordering}---as the key concern.
 
@@
 During this waiting period, the spinning workgroup remains scheduled, consuming computational resources.
-Without \emph{forward-progress guarantees}, the scheduler may indefinitely prioritize the spinning workgroup over its predecessor, preventing the predecessor from completing its reduction.
+Without FPG, the scheduler may indefinitely prioritize the spinning workgroup over its predecessor, preventing the predecessor from completing its reduction.
 This creates a circular dependency, satisfying condition 4b and resulting in deadlock within the kernel.
 
-A \emph{forward-progress guarantee} (FPG) is a promise provided by the subgroup scheduler that all occupant subgroups (of possibly different workgroups) eventually make progress toward termination, preventing scenarios where a subgroup is starved of execution time.
-Sorensen et al.~\cite{barriers, 1, 2} provide the most comprehensive analysis of GPU schedulers.
-However, while they classify \emph{Chained-Scan} as an FPG-reliant algorithm, their primary focus is in exploring and formalizing progress models, rather than their application to any one algorithm.
-In the context of \emph{Chained-Scan} deadlocking, FPG prevents condition 4b by prohibiting indefinite scheduling of any single subgroup, thereby breaking the circular dependency chain.
-\thomas{Sorensen places Chained-Scan into the HSA category, but it is unclear if this is correct.
-The definition of HSA is \newline "Threads for which fair execution is guaranteed: The thread with the lowest id that has not yet terminated" \newline However, as Chained-Scan and other serially ordered algorithms no longer use the virtual ID provided by the workgroup scheduler, is fair execution still guaranteed under HSA?
-It seems that it should be classified under OBE\@.}
-
-\thomas{This example feels redundant now. Timeout detection and recovery can be moved into the 2nd paragraph in this section.}
 For example, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor.
@@
 The placement of this section here feels more natural, and seems to mirror other papers?}
-\emph{Decoupled-Fallback} enhances the \emph{Scalar-Fallback} technique with significantly improved fallback performance.
-As an extension of the \emph{Chained-Scan} architecture, it preserves the advantages in speed and oversubscription, making it an ideal choice for our requirements.
+\emph{Decoupled-Fallback} separates fallback operations from the lookback process, enhancing the \emph{Scalar-Fallback} technique with significantly improved fallback performance.
+As an extension of the \emph{Chained-Scan} architecture, it retains the advantages of high speed and oversubscription, aligning with our design requirements.
 
@@
   \SetAlgoLined
-  \KwIn{%
-    \texttt{in}[], \texttt{out}[],
-    \texttt{tileBump},
-    \texttt{spine}[],
-    \texttt{wg\_fallback}[],
-    \texttt{wg\_lock},
-    \texttt{wg\_partials}[],
-    \texttt{wg\_prev\_reduction},
-    \texttt{wg\_broadcast},
+  \KwGlobalIn{%
+    $in[]$, 
+    $out[]$, 
+    $spine[]$, 
+    $tileBump$, 
   }
-  \KwOut{\texttt{out}[] fully inclusive scan of \texttt{in}[]}
+  \KwSharedIn{%
+    $wg\_partials[]$, 
+    $wg\_fallback[]$,  
+    $wg\_prev\_reduction$, 
+    $wg\_broadcast$, 
+    $wg\_control$, 
+  }
+  \KwGlobalOut{$out[]$ inclusive scan of $in[]$}
+  \KwConstants{Vectors per thread $p$}
 
-  \If{$\textnormal{threadid.x} = 0$}{
-    \texttt{wg\_broadcast} $\gets$ \texttt{atomicAdd}(\&\texttt{tileBump}, 1)\;
-    \texttt{wg\_lock} $\gets$ \textnormal{LOCKED}\;
+  \If{$threadid = 0$}{
+    $wg\_broadcast \gets atomicAdd(\&tileBump, 1)$\;
+    $wg\_control \gets LOCKED$\;
   }
@@
 
-  \texttt{part\_id} $\gets$ \texttt{wg\_broadcast}\;
+  $part\_id \gets wg\_broadcast$\;
 
@@
 
-  \textbf{Load}(\texttt{in}, \texttt{part\_id}, \textit{local\_vectors})\;
+  \textbf{Load}($in$, $part\_id$, \textit{local\_vectors})\;
   \textbf{SubgroupRake}(\textit{local\_vectors})\;
 
-  \If{$\text{\normalfont highestRankLane}(\textnormal{threadid.x})$}{
-    \texttt{wg\_partials}[\textnormal{subgroup\_id}] $\gets$ \textit{local\_vectors}[\textnormal{final\_partial}]\;
+  \If{$highestRankLane(threadid)$}{
+    $wg\_partials[subgroup\_id] \gets local\_vectors[final\_partial]$\;
   }
@@
 
-  \textbf{WorkgroupWideScan}(\texttt{wg\_partials})\;
+  \textbf{WorkgroupWideScan}($wg\_partials$)\;
   \textbf{barrier()}\;
@@
   \tcp{Decoupled Fallback}
-  \While{\texttt{wg\_lock} = \textnormal{LOCKED}}{
-    \textbf{Lookback}(\texttt{spine}, \texttt{wg\_prev\_reduction}, \texttt{wg\_lock})\;
-    \texttt{lock} $\gets$ \textbf{\texttt{workgroupUniformLoad}}(\&\texttt{wg\_lock})\;
+  \While{$wg\_control = LOCKED$}{
+    \textbf{Lookback}($spine$, $wg\_prev\_reduction$, $wg\_control$)\;
+    \textbf{barrier()}\;
 
-    \If{$\texttt{lock} = \textnormal{LOCKED}$}{
-      \textbf{FallbackReduce}(\texttt{in}, \texttt{wg\_fallback})\;
-      \textbf{FallbackInsertionAttempt}(\texttt{spine}, \texttt{wg\_prev\_reduction}, \texttt{wg\_lock})\;
-      \texttt{lock} $\gets$ \textbf{\texttt{workgroupUniformLoad}}(\&\texttt{wg\_lock})\;
+    \If{$wg\_control = LOCKED$}{
+      \textbf{FallbackReduce}($in$, $wg\_fallback$)\;
+      \textbf{FallbackInsertionAttempt}($spine$, $wg\_fallback$, $wg\_prev\_reduction$, $wg\_control$)\;
+      \textbf{barrier()}\;
     }
@@
 
-  \textbf{Write}(\texttt{out}, \textit{local\_vectors}, \texttt{wg\_partials}, \texttt{wg\_prev\_reduction}, \texttt{part\_id})\;
+  \textbf{Write}($out$, $part\_id$, \textit{local\_vectors}, $wg\_partials$, $wg\_prev\_reduction$)\;
 
-  \caption{High-level Decoupled Fallback}
+  \caption{High-Level Scan Kernel}
 \end{algorithm}
+
 To maximize the efficiency of the fallback operation, we employ three strategies.
 First, although a lookback is performed by a single thread or subgroup, we want all threads in the workgroup to participate.
-To facilitate this, we use a workgroup-wide lock to allow threads participating in the lookback to signal the entire workgroup when a fallback is necessary.
+To facilitate this, we use a workgroup-wide control flag to allow threads participating in the lookback to signal the entire workgroup when a fallback is necessary.
 Second, we explicitly limit the number of times a workgroup may spin while waiting on a tile.
@@
 Although this approach still risks performing redundant fallbacks---since a work tile that is late is indistinguishable from one that is blocking---exposing the maximum spin count as a tunable parameter allows us to reduce the likelihood of false-positive fallbacks.
-Finally, after completing a fallback, we attempt to insert the value into the spine at the position of the blocking tile, preventing succeeding workgroups from initiating redundant fallbacks.
+Finally, after completing a fallback, we attempt to insert the value into the spine at the position of the blocking tile, preventing any succeeding workgroups from initiating redundant fallbacks.
 
@@
   \SetAlgoLined
-  \KwIn{Shared Memory Array to Scan $x$, Workgroup Size $W$, Subgroup Size $S$}
-  \KwOut{Shared Memory Inclusive Scanned Array $x$}
+  \KwSharedIn{ Array of subgroup partial reductions $x$ }
+  \KwSharedOut{ Post-scanned array $x$ }
+  \KwConstants{ Workgroup Size $W$, Subgroup Size $S$ }
 
   $spine\_length \gets W / S$\;
-  $alignment \gets 1 << \text{divRoundUp}(\log_2(spine\_length), \log_2(S)) * \log_2(S)$\;
+  $alignment \gets 1 << \text{divRoundUp}(\log_2(spine\_length), \log_2(S)) \cdot \log_2(S)$\;
 
@@
   \ForEach{$thread\_id$ \textbf{in} $W$ \textbf{in parallel}}{
-    \For{$j \gets S$ \KwTo $alignment$ \textbf{with} $j \gets j * S$}{
+    \For{$j \gets S$ \KwTo $alignment$ \textbf{with} $j \gets j \cdot S$}{
       \tcp{Brent-Kung Upsweep}
 
-      $step \gets spine\_length / stride$
+      $step \gets spine\_length / stride$\;
 
@@
         \If{$\text{\normalfont highestRankLane}(thread\_id)$}{
-          $x[(thread\_id / S) + step + top\_offset] \gets temp$;
+          $x[(thread\_id / S) + step + top\_offset] \gets temp$\;
         }
@@
       }
-      $stride \gets stride * S$\;
+      $stride \gets stride \cdot S$\;
       $top\_offset \gets top\_offset + step$\;
@@
 \end{algorithm}
+
 \subsubsection{Workgroup-Wide Scan}
@@
 Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index.
-However, such access is interpreted by the hardware as a broadcast operation, resulting in no conflicts.
+However, such access is interpreted by the hardware as a \emph{broadcast} operation, resulting in no conflicts.
 Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase is negligible due to our overall thriftiness, including the deliberate retention of vectors in registers and the efficiency of subgroup raking.
@@
 \subsubsection{Decoupled-Fallback}
-\thomas{Given that "lock" has an extensive history of use in computer science and this is not that kind of lock, perhaps a better choice of phrase would be better. Updates to terminology incoming...}
+\thomas{Given that "lock" has an extensive history of use in computer science and this is not that kind of lock, perhaps a better choice of phrase would be better.
+I've decided to refer to it as a "control flag"}
 Prior to kernel launch, all tiles in the spine are initialized to the \emph{Not Ready} state.
@@
       \item If the tile is in the \emph{Ready} state, the workgroup accumulates the value and moves to the preceding tile, repeating this step.
-          \item If the tile is in the \emph{Inclusive} state, the workgroup accumulates the value and proceeds directly to step 5.
+      \item If the tile is in the \emph{Inclusive} state, the workgroup accumulates the value and proceeds directly to step 4.
       \item If the tile is in the \emph{Not Ready} state, the workgroup spins for a limited number of iterations, defined by the \emph{Maximum Spin Count}.
-If the spin count exceeds the \emph{Maximum Spin Count}, the workgroup signals a fallback operation using the control flag and proceeds to the next step.
+If the spin count exceeds the \emph{Maximum Spin Count}, the workgroup signals a fallback operation leaving the control flag \emph{Locked} and proceeds to the next step.
   \end{itemize}
@@
                 \begin{itemize}
-                  \item If the state is \emph{Inclusive}, the fallback value is discarded, and the workgroup accumulates the value before proceeding to step 5.
-                  \item If the blocking tile is the first tile in the spine, the workgroup has reached the start of the spine and proceeds to step 5.
+                  \item If the state is \emph{Inclusive}, the fallback value is discarded, and the workgroup accumulates the value before proceeding to step 4.
+                  \item If the blocking tile is the first tile in the spine, the workgroup has reached the start of the spine and proceeds to step 4.
                   \item Otherwise, the workgroup resumes the lookback, starting with the preceding tile.
@@
 
-  \item \textbf{Exit:}
+  \item \textbf{Finalization:}
   \begin{itemize}
@@
 To maintain a coherent view of the tile's state, two bits are required.
-In \emph{Decoupled-Lookback}, Merrill and Garland propose two strategies for communicating tile state and 32-bit reduction value: Two separate 32-bit values---one for the tile state and the other for the reduction value---with updates using a \emph{memory fence}.
+In \emph{Decoupled-Lookback}, Merrill and Garland~\cite{Merrill2016} propose two strategies for communicating tile state and 32-bit reduction value: Two separate 32-bit values---one for the tile state and the other for the reduction value---with updates using a \emph{memory fence}.
 One single 64-bit value, bit-packing the tile state and reduction value, with updates via 64-bit atomics.
@@
 Instead, we use two 32-bit values and split the reduction into upper and lower 16-bit segments, reserving the two most significant bits of each 32-bit value to encode the tile state.
-\newline
-SPLIT BITS MINI DIAGRAM here
-\newline
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=\linewidth]{graphics/split.pdf}
+\end{figure}
 To accommodate split values, we use a single subgroup for the lookback, leveraging subgroup \emph{ballot} to verify that the split value state is consistent across threads, and subgroup \emph{shuffle} to join and split reductions as needed.

### 2025-01-23 artifact/WGPU/src/main.rs
Add Code Artifact.
Shaders almost all done, just need to cleanup on the host side.

@@ -1,0 +1,958 @@
+use std::{env, vec};
+use wgpu::util::DeviceExt;
+
+fn div_round_up(x: u32, y: u32) -> u32 {
+    (x + y - 1) / y
+}
+
+enum ScanType {
+    Rts,
+    Csdl,
+    Csdldf,
+    CsdldfStats,
+    CsdldfEmulate,
+    Memcpy,
+}
+
+struct GPUContext {
+    device: wgpu::Device,
+    queue: wgpu::Queue,
+    query_set: wgpu::QuerySet,
+    timestamp_freq: f32,
+}
+
+impl GPUContext {
+    async fn init() -> Self {
+        let instance = wgpu::Instance::new(wgpu::InstanceDescriptor {
+            backends: wgpu::Backends::all(),
+            flags: wgpu::InstanceFlags::empty(),
+            dx12_shader_compiler: wgpu::Dx12Compiler::default(),
+            gles_minor_version: wgpu::Gles3MinorVersion::default(),
+        });
+
+        let adapter = instance
+            .request_adapter(&wgpu::RequestAdapterOptions {
+                power_preference: wgpu::PowerPreference::HighPerformance,
+                compatible_surface: None,
+                force_fallback_adapter: false,
+            })
+            .await
+            .expect("Failed to find an appropriate adapter");
+
+        let (device, queue) = adapter
+            .request_device(
+                &wgpu::DeviceDescriptor {
+                    label: None,
+                    required_features: wgpu::Features::TIMESTAMP_QUERY | wgpu::Features::SUBGROUP,
+                    required_limits: wgpu::Limits::default(),
+                    memory_hints: wgpu::MemoryHints::Performance,
+                },
+                None,
+            )
+            .await
+            .expect("Failed to create device");
+
+        let query_set = device.create_query_set(&wgpu::QuerySetDescriptor {
+            label: Some("Timestamp Query Set"),
+            count: 6,
+            ty: wgpu::QueryType::Timestamp,
+        });
+
+        let timestamp_freq = queue.get_timestamp_period();
+
+        GPUContext {
+            device,
+            queue,
+            query_set,
+            timestamp_freq,
+        }
+    }
+}
+
+struct GPUBuffers {
+    params: wgpu::Buffer,
+    scan_in: wgpu::Buffer,
+    scan_out: wgpu::Buffer,
+    scan_bump: wgpu::Buffer,
[…]

### 2025-01-25 paper.tex
Update dawn. Add Appendix. Add diagnostic ignore for Tint.

@@
 \usepackage{mathtools}
+\usepackage{listings}
 \newcommand{\thomas}[1]{{\footnotesize\color{orange}[Thomas: #1]}}
@@
 \appendix
+\section{Artifact}
 
-\section{Research Methods}
+\subsection{Availability}
+We provide our artifact in the following GitHub repository: 
+\thomas{Will anonymize with GithubAnonymous once we are ready.} 
+Due to the aforementioned issues with WGSL transpilers, we choose Dawn as our WebGPU implementation.
+
+\subsection{Requirements}
+
+\subsubsection{Device Requirements}
+Any device supporting Dawn, with subgroup and timestamp query capabilities, and at least 384 MB of available device memory.
+
+\subsubsection{Software Requirements}
+\begin{itemize}
+  \item Git
+  \item CMake 3.10.2+ or another C++ build tool
+  \item C++17-compliant compiler
+  \item Python for fetching dependencies
+\end{itemize}
+
+\subsection{Building}
+Although native WebGPU is supported on various operating systems—most notably Windows, macOS, and Linux—this section will focus on building it in Unix-like environments for the sake of brevity, using CMake as the example build tool.
+
+\subsubsection{Building Dawn}
+\begin{enumerate}
+  \item[(0)] Dawn depends on the following packages (Ubuntu package names):
+  \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
+sudo apt install libxrandr-dev 
+  libxinerama-dev libxcursor-dev libxi-dev \
+  libx11-xcb-dev mesa-common-dev
+  \end{lstlisting}
+
+  \item Clone the Dawn repository:
+  \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
+git clone https://dawn.googlesource.com/dawn
+  \end{lstlisting}
+
+  \item Build and install Dawn with your build tool and compiler of choice:
+  \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
+cd dawn
+cmake -S . -B out/Release \
+  -DDAWN_FETCH_DEPENDENCIES=ON \
+  -DDAWN_ENABLE_INSTALL=ON \
+  -DCMAKE_BUILD_TYPE=Release
+...
+cmake --build out/Release
+...
+cmake --install out/Release \
+  --prefix install/Release
+  \end{lstlisting}
+\end{enumerate}
+
+\subsubsection{Building Artifact}
+After successfully building and installing Dawn, follow these steps to build the artifact:
+
+\begin{enumerate}
+  \item Navigate to the artifact directory:
+  \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
+cd Decoupled-Fallback-Paper/artifact/Dawn
+  \end{lstlisting}
+
+  \item Set the `CMAKE\_PREFIX\_PATH` to point to your Dawn installation directory:
+  \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
+export CMAKE_PREFIX_PATH=/path/to/ \
+dawn/install/Release
+  \end{lstlisting}
+
+  \item Build the artifact:
+  \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
+cmake -S . -B out/Release \
+  -DCMAKE_BUILD_TYPE=Release
+  \end{lstlisting}
+
+  \item Compile the project:
+  \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
+cmake --build out/Release
+  \end{lstlisting}
+\end{enumerate}
 

### 2025-01-25 artifact/Dawn/main.cpp

@@ -1,12 +1,3 @@
-/******************************************************************************
- * GPUPrefixSums
- *
- * SPDX-License-Identifier: MIT
- * Copyright Thomas Smith 10/23/2024
- * https://github.com/b0nes164/GPUPrefixSums
- *
- ******************************************************************************/
-
 #include <dawn/webgpu_cpp.h>
 
 #include <cstdlib>
@@ -38,13 +29,10 @@ struct Shaders {
     ComputeShader downsweep;
     ComputeShader csdl;
     ComputeShader csdldf;
-    ComputeShader csdldfStruct;
     ComputeShader csdldfStats;
-    ComputeShader csdldfStructStats;
-    ComputeShader csdldfOcc;
-    ComputeShader csdldfStructOcc;
+    ComputeShader csdldfEmulate;
+    ComputeShader memcpy;
     ComputeShader validate;
-    ComputeShader validateStruct;
 };
 
 struct GPUBuffers {
@@ -71,10 +59,7 @@ struct TestArgs {
     bool shouldReadback = false;
     bool shouldTime = false;
     bool shouldGetStats = false;
-    bool shouldGetOcc = false;
     uint32_t (*MainPass)(const TestArgs&, wgpu::CommandEncoder*) = nullptr;
-    bool (*ValidateSync)(const GPUContext&, GPUBuffers*,
-                         const Shaders&) = nullptr;
 };
 
 enum class ScanType {
@@ -82,10 +67,8 @@ enum class ScanType {
     Csdl,
     Csdldf,
     CsdldfStats,
-    CsdldfOcc,
-    CsdldfStruct,
-    CsdldfStructStats,
-    CsdldfStructOcc,
+    CsdldfEmulate,
+    Memcpy,
     Unknown
 };
 
@@ -146,20 +129,16 @@ int GetGPUContext(GPUContext* context, uint32_t timestampCount) {
         wgpu::FeatureName::Subgroups,
         wgpu::FeatureName::TimestampQuery,
     };
+
+    auto errorCallback = [](const wgpu::Device& device, wgpu::ErrorType type, wgpu::StringView message) {
+        std::cerr << "Error: " << std::string(message.data, message.length) << std::endl;
+    };
+
     wgpu::DeviceDescriptor devDescriptor{};
     devDescriptor.requiredFeatures = reqFeatures.data();
     devDescriptor.requiredFeatureCount =
         static_cast<uint32_t>(reqFeatures.size());
-
[…]

### 2025-01-26 paper.tex
Edits, Revert RTS to non-raking

@@
 \subsection{The Scan Primitive}\label{sec:the-scan-primitive}
-The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{5219801}.
+The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{5219801,10.5555/1098666}.
 A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \).
@@
 To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than arbitrary assignment based on virtualized workgroup indices.
-This shifts dependency resolution from explicit synchronization via barrier-like structures to implicit synchronization via the order in which workgroups access global data.
+This shifts dependency resolution from explicit synchronization via barrier-like structures to implicit synchronization via access order.
 As a result, each element is read and written exactly once, achieving the theoretical minimum of $2n$.
@@
 Although the serial assignment of work tiles ensures that predecessor work tiles are either scheduled or completed, there is no guarantee that the corresponding reductions will be available before a workgroup begins its lookback.
-This uncertainty arises because workgroups are scheduled individually, and a predecessor workgroup might not yet have had the opportunity to compute and post its reduction.
+This uncertainty arises because workgroups are scheduled individually, and a predecessor workgroup might not yet have had the opportunity to compute and post its reduction \thomas{There is a better wording here.
+Something like, "because although workgroups \emph{may} share execution resources, they are ultimately scheduled individually." This sentence is incredibly important.}.
 In such cases, the dependent workgroup enters a waiting state, spinning until the reduction becomes available.
@@
 Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress requirements in \emph{ScalarFallback}.
-By allowing a workgroup to access and operate on another tile’s data, \emph{ScalarFallback} targets Coffman et al.'s \emph{mutual exclusion} condition.
-Instead, each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time.
-Although this entails performing redundant work when the blocking tile is simply late but not blocking, it guarantees the termination of the algorithm because each workgroup can perform the reduction of any blocking tile, breaking the inter-workgroup dependency.
+Each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time.
+By enabling workgroups to operate on data from other tiles, \emph{ScalarFallback} prevents deadlocking by targeting the \emph{mutual exclusion} condition described by Coffman et al.
+While this may involve performing redundant work when a tile is late rather than genuinely blocking, it ensures the algorithm's termination by allowing any workgroup to complete the reduction of a blocking tile, thereby breaking inter-workgroup dependencies.
 
@@
 
-To maximize the efficiency of the fallback operation, we employ three strategies.
-First, although a lookback is performed by a single thread or subgroup, we want all threads in the workgroup to participate.
-To facilitate this, we use a workgroup-wide control flag to allow threads participating in the lookback to signal the entire workgroup when a fallback is necessary.
-Second, we explicitly limit the number of times a workgroup may spin while waiting on a tile.
-If the maximum spin count is exceeded, the fallback operation is initiated.
-Although this approach still risks performing redundant fallbacks---since a work tile that is late is indistinguishable from one that is blocking---exposing the maximum spin count as a tunable parameter allows us to reduce the likelihood of false-positive fallbacks.
-Finally, after completing a fallback, we attempt to insert the value into the spine at the position of the blocking tile, preventing any succeeding workgroups from initiating redundant fallbacks.
+To optimize the fallback operation, we employ three strategies.
+Although a lookback is performed by a single thread or subgroup, we want the entire workgroup to participate in the fallback.
+To achieve this, our first strategy is the use of a workgroup-wide \emph{control flag}, which enables threads performing the lookback to signal the entire workgroup when a fallback is necessary.
+Second, we explicitly limit the number of spins a workgroup can perform while waiting on a tile.
+If this maximum spin count is exceeded, the fallback operation is triggered.
+While this approach still risks performing redundant fallbacks---since a late tile is indistinguishable from one which is genuinely blocking---exposing the maximum spin count as a tunable parameter helps minimize false-positive fallbacks.
+Finally, after completing a fallback, the workgroup attempts to insert the reduction in place of the blocking tile, preventing subsequent workgroups from initiating redundant fallbacks.
 
 The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled-Lookback}.
-Regardless of FPG, due to false positives, any given hardware will have a natural fallback-incidence rate $f$.
+No subgroup scheduler is perfectly fair, and thus any given hardware will have a natural fallback-incidence rate $f$, regardless of FPG, due to \emph{false-positive} fallback operations.
 Consequently, we expect memory bandwidth traffic to increase (and throughput to decrease) by $O(fn)$ to $O((2 + f)n)$.
@@
 The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches.
-To handle this variability, we use a generalized radix-$s$ Ladner-Fischer scan network embedded with Kogge-Stone subgroup scans to perform the scan across the local spine.
+To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{} scan network embedded with Kogge-Stone subgroup scans to perform the scan across the local spine.
 Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts.
@@
                   \item If the state is \emph{Inclusive}, the fallback value is discarded, and the workgroup accumulates the value before proceeding to step 4.
-                  \item If the blocking tile is the first tile in the spine, the workgroup has reached the start of the spine and proceeds to step 4.
+                  \item If the blocking tile is the first in the spine, the workgroup has reached the start of the spine and proceeds to step 4.
                   \item Otherwise, the workgroup resumes the lookback, starting with the preceding tile.
@@
 \subsubsection{Tile State Implementation}
-To maintain a coherent view of the tile's state, two bits are required.
-In \emph{Decoupled-Lookback}, Merrill and Garland~\cite{Merrill2016} propose two strategies for communicating tile state and 32-bit reduction value: Two separate 32-bit values---one for the tile state and the other for the reduction value---with updates using a \emph{memory fence}.
-One single 64-bit value, bit-packing the tile state and reduction value, with updates via 64-bit atomics.
-In our case, neither are viable options.
-Due to a lack of support across backend APIs, WebGPU does not furnish developers with a memory model, and currently, 64-bit atomics are not supported.
-Moreover, 64-bit atomics are not supported across most mobile devices, and thus we do not consider their future support an appropriate option for our implementation.
-Instead, we use two 32-bit values and split the reduction into upper and lower 16-bit segments, reserving the two most significant bits of each 32-bit value to encode the tile state.
+Because a tile may be in three possible states, two bits are required to record its state alongside the 32 bits needed to store the tile's reduction value.
+Previous \emph{Chained-Scan} architectures maintained a coherent view of both by using either 64-bit atomics with bit packing or memory fences with 32-bit atomics.
+However, neither approach is viable for our implementation, as WebGPU lacks a memory model and does not support 64-bit atomics.
+Additionally, 64-bit atomics are unsupported on most mobile devices, making their future adoption impractical for our needs.
+Instead, we use two 32-bit values, splitting the reduction into upper and lower 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state.
 \begin{figure}[h!]
@@
 \end{figure}
-To accommodate split values, we use a single subgroup for the lookback, leveraging subgroup \emph{ballot} to verify that the split value state is consistent across threads, and subgroup \emph{shuffle} to join and split reductions as needed.
+To accommodate the split values, we use a single subgroup for the lookback, leveraging subgroup \emph{ballot} to verify that the split value state is consistent across threads, and subgroup \emph{shuffle} to join and split reductions as needed.
 Since this method requires one thread per 16 bits, and the minimum subgroup size in the WebGPU specification is 4, it supports both 32-bit and 64-bit values.
-Notably, our approach extends support to 64-bit scans, which would not be possible even with 64-bit atomics.\footnote{Although native 64-bit values are not currently supported by the WebGPU specification, their future inclusion is anticipated.}
+Notably, this extends support to 64-bit scans, which would not be possible even with 64-bit atomics.\footnote{Although native 64-bit values are not currently supported by the WebGPU specification, their future inclusion is anticipated.}
 
-In the context of fallbacks, the blocking tile can exist in any of three states—\emph{Not Ready}, \emph{Ready}, or \emph{Inclusive}—due to several factors.
-A false-positive fallback may occur when the blocking tile completes its local or inclusive reduction before the fallback finishes.
-Additionally, multiple workgroups are allowed to perform fallbacks on the same blocking tile, meaning another workgroup may have already updated the tile state.
-To handle these scenarios, we require a tile state representation and an accompanying atomic operation that ensures a fallback can safely transition a tile's state to \emph{Ready} without overwriting a more advanced state.
-This challenge is further compounded by the fact that in WebGPU, the correctness of atomic \emph{Compare-and-Swap} is not guaranteed across all devices, prohibiting our use.
-Instead, we encode the tile state within the most significant bits of the split value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}.
-This design enables the use of atomic \emph{Maximum}, which inherently respects the state hierarchy, to safely insert values.
-\thomas{Possibly a blurb here about no volatile, so we use all atomics?}
+In the context of fallbacks, we require a tile state representation and an accompanying atomic operation that ensures a fallback update attempt can safely transition a tile's state to \emph{Ready} without overwriting a more advanced state.
+When an update attempt is made, the blocking tile can exist in any of the three states due to two scenarios: a false-positive fallback may have been initiated on a tile that transitions to \emph{Ready} or \emph{Inclusive} before the fallback completes, or another workgroup may have already successfully completed a fallback update, leaving the tile in \emph{Ready}.Consequently, the update mechanism should only overwrite tiles in the \emph{Not Ready} state.
+This requirement is further constrained by the fact that in WebGPU, the correctness of atomic \emph{Compare-and-Swap} is not guaranteed across all devices, prohibiting its use.
+To address this, we encode the tile state within the most significant bits of the split value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}.
+This design enables the use of atomic \emph{Maximum}, which inherently respects the state hierarchy, to safely perform updates.
 
@@
 git clone https://dawn.googlesource.com/dawn
+RESET HEAD here to a specific commit. . . to lock in?
   \end{lstlisting}

### 2025-01-26 artifact/Dawn/main.cpp

@@ -844,8 +844,7 @@ int main(int argc, char* argv[]) {
         1 << powerOfTwo;     // Input size to test, must be a multiple of 4
     uint32_t threadBlocks =  // Thread Blocks to launch based on input
         (size + PART_SIZE - 1) / PART_SIZE;
-    uint32_t readbackSize =
-        256;  // How many elements to readback, must be less than max
+    uint32_t readbackSize = 256;  // How many elements to readback, must be less than max
     bool shouldValidate = true;   // Perform validation?
     bool shouldReadback = false;  // Use readback to sanity check results
     bool shouldTime = true;       // Time results?

### 2025-01-27 artifact/Dawn/main.cpp
Add Data Recording

@@ -9,6 +9,8 @@
 #include <string>
 #include <vector>
 
+constexpr uint32_t MAX_EMULATE = 10;
+
 struct GPUContext {
     wgpu::Instance instance;
     wgpu::Device device;
@@ -29,8 +31,8 @@ struct Shaders {
     ComputeShader downsweep;
     ComputeShader csdl;
     ComputeShader csdldf;
-    ComputeShader csdldfStats;
-    ComputeShader csdldfEmulate;
+    ComputeShader csdldfOcc;
+    ComputeShader csdldfEmulate[MAX_EMULATE + 1];
     ComputeShader memcpy;
     ComputeShader validate;
 };
@@ -55,30 +57,28 @@ struct TestArgs {
     uint32_t batchSize;
     uint32_t threadBlocks;
     uint32_t readbackSize;
+    uint32_t fallbackPreset;
     bool shouldValidate = false;
     bool shouldReadback = false;
     bool shouldTime = false;
     bool shouldGetStats = false;
-    uint32_t (*MainPass)(const TestArgs&, wgpu::CommandEncoder*) = nullptr;
+    bool shouldRecord = false;
 };
 
-enum class ScanType {
-    Rts,
-    Csdl,
-    Csdldf,
-    CsdldfStats,
-    CsdldfEmulate,
-    Memcpy,
-    Unknown
+struct DataStruct {
+    std::vector<double> time;
+    std::vector<uint32_t> totalSpins;
+    std::vector<double> lookbackLength;
+    std::vector<uint32_t> fallbacksInitiated;
+    std::vector<uint32_t> successfulInsertions;
 };
 
-int GetGPUContext(GPUContext* context, uint32_t timestampCount) {
+void GetGPUContext(GPUContext* context, uint32_t timestampCount) {
     wgpu::InstanceDescriptor instanceDescriptor{};
     instanceDescriptor.features.timedWaitAnyEnable = true;
     wgpu::Instance instance = wgpu::CreateInstance(&instanceDescriptor);
     if (instance == nullptr) {
         std::cerr << "Instance creation failed!\n";
-        return EXIT_FAILURE;
     }
 
     wgpu::RequestAdapterOptions options = {};
@@ -130,8 +130,10 @@ int GetGPUContext(GPUContext* context, uint32_t timestampCount) {
         wgpu::FeatureName::TimestampQuery,
     };
 
[…]

### 2025-01-27 artifact/WGPU/src/main.rs
Cleanup Artifact. Add Labeling for csv.

@@ -1,958 +1,0 @@
-use std::{env, vec};
-use wgpu::util::DeviceExt;
-
-fn div_round_up(x: u32, y: u32) -> u32 {
-    (x + y - 1) / y
-}
-
-enum ScanType {
-    Rts,
-    Csdl,
-    Csdldf,
-    CsdldfStats,
-    CsdldfEmulate,
-    Memcpy,
-}
-
-struct GPUContext {
-    device: wgpu::Device,
-    queue: wgpu::Queue,
-    query_set: wgpu::QuerySet,
-    timestamp_freq: f32,
-}
-
-impl GPUContext {
-    async fn init() -> Self {
-        let instance = wgpu::Instance::new(wgpu::InstanceDescriptor {
-            backends: wgpu::Backends::all(),
-            flags: wgpu::InstanceFlags::empty(),
-            dx12_shader_compiler: wgpu::Dx12Compiler::default(),
-            gles_minor_version: wgpu::Gles3MinorVersion::default(),
-        });
-
-        let adapter = instance
-            .request_adapter(&wgpu::RequestAdapterOptions {
-                power_preference: wgpu::PowerPreference::HighPerformance,
-                compatible_surface: None,
-                force_fallback_adapter: false,
-            })
-            .await
-            .expect("Failed to find an appropriate adapter");
-
-        let (device, queue) = adapter
-            .request_device(
-                &wgpu::DeviceDescriptor {
-                    label: None,
-                    required_features: wgpu::Features::TIMESTAMP_QUERY | wgpu::Features::SUBGROUP,
-                    required_limits: wgpu::Limits::default(),
-                    memory_hints: wgpu::MemoryHints::Performance,
-                },
-                None,
-            )
-            .await
-            .expect("Failed to create device");
-
-        let query_set = device.create_query_set(&wgpu::QuerySetDescriptor {
-            label: Some("Timestamp Query Set"),
-            count: 6,
-            ty: wgpu::QueryType::Timestamp,
-        });
-
-        let timestamp_freq = queue.get_timestamp_period();
-
-        GPUContext {
-            device,
-            queue,
-            query_set,
-            timestamp_freq,
-        }
-    }
-}
-
-struct GPUBuffers {
-    params: wgpu::Buffer,
-    scan_in: wgpu::Buffer,
-    scan_out: wgpu::Buffer,
-    scan_bump: wgpu::Buffer,
[…]

### 2025-01-27 artifact/main.cpp

@@ -421,32 +421,31 @@ void GetAllShaders(const GPUContext& gpu, const GPUBuffers& buffs,
                    Shaders* shaders) {
     std::vector<std::string> empty;
     CreateShaderFromSource(gpu, buffs, &shaders->init, "main",
-                           "../SharedShaders/init.wgsl", "Init", empty);
+                           "Shaders/init.wgsl", "Init", empty);
 
     CreateShaderFromSource(gpu, buffs, &shaders->reduce, "reduce",
-                           "../SharedShaders/rts.wgsl", "Reduce", empty);
+                           "Shaders/rts.wgsl", "Reduce", empty);
 
     CreateShaderFromSource(gpu, buffs, &shaders->spineScan, "spine_scan",
-                           "../SharedShaders/rts.wgsl", "Spine Scan", empty);
+                           "Shaders/rts.wgsl", "Spine Scan", empty);
 
     CreateShaderFromSource(gpu, buffs, &shaders->downsweep, "downsweep",
-                           "../SharedShaders/rts.wgsl", "Downsweep", empty);
+                           "Shaders/rts.wgsl", "Downsweep", empty);
 
     CreateShaderFromSource(gpu, buffs, &shaders->csdl, "main",
-                           "../SharedShaders/csdl.wgsl", "CSDL", empty);
+                           "Shaders/csdl.wgsl", "CSDL", empty);
 
     CreateShaderFromSource(gpu, buffs, &shaders->csdldf, "main",
-                           "../SharedShaders/csdldf.wgsl", "CSDLDF", empty);
+                           "Shaders/csdldf.wgsl", "CSDLDF", empty);
 
     CreateShaderFromSource(gpu, buffs, &shaders->csdldfOcc, "main",
-                           "../SharedShaders/csdldf_occ.wgsl", "CSDLDF OCC",
-                           empty);
+                           "Shaders/csdldf_occ.wgsl", "CSDLDF OCC", empty);
 
     CreateShaderFromSource(gpu, buffs, &shaders->memcpy, "main",
-                           "../SharedShaders/memcpy.wgsl", "Memcpy", empty);
+                           "Shaders/memcpy.wgsl", "Memcpy", empty);
 
     CreateShaderFromSource(gpu, buffs, &shaders->validate, "main",
-                           "../SharedShaders/validate.wgsl", "Validate", empty);
+                           "Shaders/validate.wgsl", "Validate", empty);
 
     // Create shaders with decreasing artificial deadlocking
[…]

### 2025-01-28 paper.tex
Add Graphs (Placeholder Data)

@@
 \usepackage{listings}
+\usepackage{subcaption}
 \newcommand{\thomas}[1]{{\footnotesize\color{orange}[Thomas: #1]}}
@@
 The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled-Lookback}.
-No subgroup scheduler is perfectly fair, and thus any given hardware will have a natural fallback-incidence rate $f$, regardless of FPG, due to \emph{false-positive} fallback operations.
+No subgroup scheduler is perfectly fair, and thus any given hardware will have a natural fallback rate $f$, regardless of FPG, due to \emph{false-positive} fallback operations.
 Consequently, we expect memory bandwidth traffic to increase (and throughput to decrease) by $O(fn)$ to $O((2 + f)n)$.
@@
           \item The workgroup computes the reduction of the blocking tile by processing the corresponding elements directly.
+This process involves a workgroup-wide reduction using additional shared memory to aggregate partial subgroup results.
           \item Once the fallback reduction is complete, the workgroup \emph{attempts} to post the reduction to global memory and update the tile's state to \emph{Ready}.
@@
   \begin{itemize}
-      \item Once the lookback or fallback is complete, the workgroup posts its inclusive reduction to global memory, transitioning its tile state to \emph{Inclusive}.
-      \item The \emph{previous reduction} is also posted to shared memory for the workgroup to consume and propagate as necessary.
+      \item Upon completing the lookback or fallback, the local reduction is added to the \emph{previous reduction} to compute the inclusive reduction, which is then posted to global memory, transitioning the tile state to \emph{Inclusive}.
+      \item The \emph{previous reduction} is posted to shared memory for the workgroup to consume and propagate as necessary.
       \item The control flag is updated to \emph{Unlocked} to signal the exit of the \emph{Decoupled-Fallback} phase, allowing subsequent operations to proceed.
@@
 In the context of fallbacks, we require a tile state representation and an accompanying atomic operation that ensures a fallback update attempt can safely transition a tile's state to \emph{Ready} without overwriting a more advanced state.
-When an update attempt is made, the blocking tile can exist in any of the three states due to two scenarios: a false-positive fallback may have been initiated on a tile that transitions to \emph{Ready} or \emph{Inclusive} before the fallback completes, or another workgroup may have already successfully completed a fallback update, leaving the tile in \emph{Ready}.Consequently, the update mechanism should only overwrite tiles in the \emph{Not Ready} state.
+When an update attempt is made, the blocking tile can exist in any of the three states due to two scenarios: a false-positive fallback may have been initiated on a tile that transitions to \emph{Ready} or \emph{Inclusive} before the fallback completes, or another workgroup may have already successfully completed a fallback update, leaving the tile in \emph{Ready}.
+Consequently, the update mechanism should only overwrite tiles in the \emph{Not Ready} state.
 This requirement is further constrained by the fact that in WebGPU, the correctness of atomic \emph{Compare-and-Swap} is not guaranteed across all devices, prohibiting its use.
@@
 \section{Evaluation}
-256 threads, 16 elements per thread.
-Shared memory requirements.
-RTS implementation shares an identical local implementation.
-Uses Merrill-Grimshaw style raking, with our own occupancy discovery technique.
+\subsection{Experimental Setup}
+We configure workgroups with a size of 256 threads, where each thread processes 4 vectors of size 4, resulting in 16 elements per thread and a total work tile size of 4096.
+We find a maximum spin count of 4 as offering good responsiveness to blocking, while minimizing false-positives fallbacks.
+As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the minimum subgroup size specified by the WebGPU standard, which is 4.
+Despite the additional shared memory required for conflict avoidance and fallback operations, total shared memory usage remains modest at only 1 KB.
+For all tests, we use an input size of $2^{25}$.
 
-\subsection{Comparison to Reduce-Then-Scan}
+Something about devices here.
+\subsection{Performance Comparison}
+To evaluate the performance of \emph{Decoupled-Fallback}, we compare it against two key benchmarks: \emph{Reduce-Then-Scan} (RTS) and a \emph{Memcopy} operation.
+RTS serves as a baseline to demonstrate the performance gains of our approach over the previous FPG-constrained technique.
+Memcopy is included as a point of reference, as it shares the same $2n$ global memory movement required by scan operations.
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=\linewidth]{graphics/speedup.pdf}
+\end{figure}
+Since WebGPU limits device-side timestamps to compute or render passes, we implement a custom Memcopy kernel for timing purposes.
+Similarly, as there are no existing WebGPU implementations of RTS, we rely on our own optimized version for comparison.
+To ensure a competitive evaluation, our RTS implementation employs a variable-launch main kernel with a single workgroup performing a serial spine scan, as we found raking to be less performant in practice.
+Apart from this adjustment, the intraworkgroup implementation in RTS is identical to that used in \emph{Decoupled-Fallback}.
+
+\thomas{Performance is good}
 
@@
 
-\subsection{Emulated Deadlocking}
-Show the performance of devices as we artificially increase the fallback-incidence rate.
+\subsection{Simulated Blocking}
+To demonstrate the resilience of our technique under fallback rates significantly higher than those observed organically on devices, we intentionally force fallbacks by omitting the posting of work tile reductions.
+Starting with infrequent omissions---1 in 512 workgroups---we progressively increase the rate to near-continuous levels, reaching 1 in 2 workgroups.
+\begin{figure}[htbp]
+  \centering
+  \begin{subfigure}{.9\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{graphics/fallbacksInitiated_plot.pdf}
+    \caption{Fallbacks Initiated}
+    \label{fig:fallbacks_initiated}
+  \end{subfigure}
+  \begin{subfigure}{.9\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{graphics/totalSpins_plot.pdf}
+    \caption{Total Spins}
+    \label{fig:total_spins}
+  \end{subfigure}
+  \begin{subfigure}{.9\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{graphics/lookbackLength_plot.pdf}
+    \caption{Lookback Length}
+    \label{fig:lookback_length}
+  \end{subfigure}
+  \begin{subfigure}{.9\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{graphics/time_plot.pdf}
+    \caption{Speed vs Memcopy}
+    \label{fig:execution_time}
+  \end{subfigure}
+  \caption{Simulated Blocking}
+  \label{fig:vertical_images}
+\end{figure}
+
+In Figure~\ref{fig:fallbacks_initiated}, we measure the number of fallbacks proportional to the number of workgroups launched, and verify that we are indeed inducing fallbacks.
+In Figure~\ref{fig:total_spins} we measure \thomas{this should be spins per workgroup it's a better metric}.
+In Figure~\ref{fig:lookbackLength} we see that.
+Finally, in Figure~\ref{fig:execution_time}, we can see that, even when \emph{half} the workgroups are deadlocking, \emph{Decoupled-Fallback} still achieves speeds of .
+.
+.
 

### 2025-01-28 artifact/main.cpp

@@ -99,7 +99,7 @@ void GetGPUContext(GPUContext* context, uint32_t timestampCount) {
             adaptPromise.set_value();
         });
     std::future<void> adaptFuture = adaptPromise.get_future();
-    while (adaptFuture.wait_for(std::chrono::microseconds(100)) ==
+    while (adaptFuture.wait_for(std::chrono::microseconds(250)) ==
            std::future_status::timeout) {
         instance.ProcessEvents();
     }
@@ -156,7 +156,7 @@ void GetGPUContext(GPUContext* context, uint32_t timestampCount) {
             devPromise.set_value();
         });
     std::future<void> devFuture = devPromise.get_future();
-    while (devFuture.wait_for(std::chrono::microseconds(100)) ==
+    while (devFuture.wait_for(std::chrono::microseconds(250)) ==
            std::future_status::timeout) {
         instance.ProcessEvents();
     }
@@ -410,7 +410,7 @@ void CreateShaderFromSource(const GPUContext& gpu, const GPUBuffers& buffs,
             promise.set_value();
         });
     std::future<void> future = promise.get_future();
-    while (future.wait_for(std::chrono::microseconds(100)) ==
+    while (future.wait_for(std::chrono::microseconds(250)) ==
            std::future_status::timeout) {
         gpu.instance.ProcessEvents();
     }
@@ -505,7 +505,7 @@ void QueueSync(const GPUContext& gpu) {
             promise.set_value();
         });
     std::future<void> future = promise.get_future();
-    while (future.wait_for(std::chrono::microseconds(100)) ==
+    while (future.wait_for(std::chrono::microseconds(250)) ==
            std::future_status::timeout) {
         gpu.instance.ProcessEvents();
     }
@@ -545,7 +545,7 @@ void ReadbackSync(const GPUContext& gpu, wgpu::Buffer* dstReadback,
         });
 
     std::future<void> future = promise.get_future();
-    while (future.wait_for(std::chrono::microseconds(100)) ==
+    while (future.wait_for(std::chrono::microseconds(250)) ==
            std::future_status::timeout) {
         gpu.instance.ProcessEvents();
     }

### 2025-01-29 paper.tex
#7 I went through the flowchart in detail because I think it's really important for my understanding and for the success of the paper. I posted a bunch of comments, at least partially because @b0nes164 can do edits even as I haven't read anything else and won't for a little while

@@
   \Return{$x$}\;
-  \caption{Workgroup-Wide Scan}
+  \caption{Workgroup-Wide Scan.
+\john{Sorry to say that every one of those variable names need to be wrapped in \texttt{\textbackslash textit}.
+``Typeset in math mode'' is not the same as ``Typeset in italics'' (see $offset$ vs.\ \textit{offset}).}}
   \label{alg:example}
@@
 Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase is negligible due to our overall thriftiness, including the deliberate retention of vectors in registers and the efficiency of subgroup raking.
-Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; zero bank conflicts across all supported subgroup sizes.
-\footnote{Since we also require a workgroup-wide pure reduce operation, this can be achieved by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}
+Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; zero bank conflicts across all supported subgroup sizes.\footnote{Since we also require a workgroup-wide pure reduce operation, this can be achieved by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}
 
@@
 Each lookback is, therefore, a traversal from the \emph{Ready} state to the \emph{Inclusive} state. To illustrate this process, we present a flowchart (SEE LOOKBACK FLOWCHART).
+\john{It's helpful, I think, to also explicitly say what data is stored globally as the spine (I think, for each workgroup, one data element and one state).}
 
@@
 \thomas{Given that "lock" has an extensive history of use in computer science and this is not that kind of lock, perhaps a better choice of phrase would be better.
-I've decided to refer to it as a "control flag"}
+I've decided to refer to it as a "control flag"} \john{Good idea.
+``Status flag'', perhaps?}
 Prior to kernel launch, all tiles in the spine are initialized to the \emph{Not Ready} state.
@@
 This flag governs both the initiation of fallback operations and the entry and exit of the entire phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.
+\john{Maybe a sentence that describes the state progression at a high level?
+e.g., ``the state of each workgroup begins at INVALID, moves to READY, then to INCLUSIVE, the state never moves backwards, each workgroup can operate independently and change its own state and also, during lookback, the state of preceding workgroups'', that sort of thing.}
+\john{I might also say something like ``in the following discussion, we assume the `add' monoid and use terms like `add' and `sum', but the algorithm works with any monoid''.
+That gets rid of awkward terms like ``accumulate''.
+I don't feel strongly about this.}
 \begin{figure}[h]
@@
           \item The control flag, initialized to \emph{Locked}, signals that the workgroup should enter the lookback phase.
-          \item Each workgroup initializes a \emph{previous reduction} variable with the identity element of the monoid to hold the accumulated reductions of preceding tiles.
+          \item Each workgroup initializes a \john{shared workgroup-local (?)} \emph{previous reduction} variable with the identity element of the monoid to hold the accumulated reductions of preceding tiles.
           \item The workgroup prepares to traverse the spine, starting with its immediate predecessor tile.
+\john{Carefully look at the way this is worded with respect to ``immediate predecessor'' (previous sentence) vs.\ ``current'' (next sentence).
+I think here we want to say that we set the current tile to the immediate predecessor, and then the next sentence is consistent.}
         \end{itemize}
@@
           \item The workgroup queries the state of the current tile.
-      \item If the tile is in the \emph{Ready} state, the workgroup accumulates the value and moves to the preceding tile, repeating this step.
+          \item If the tile is in the \emph{Ready} state, the workgroup accumulates the value and moves to \john{sets the current tile to} the preceding tile, repeating this step.
+\john{until reaching a tile that is not \emph{Ready}.}
           \item If the tile is in the \emph{Inclusive} state, the workgroup accumulates the value and proceeds directly to step 4.
           \item If the tile is in the \emph{Not Ready} state, the workgroup spins for a limited number of iterations, defined by the \emph{Maximum Spin Count}.
-If the spin count exceeds the \emph{Maximum Spin Count}, the workgroup signals a fallback operation leaving the control flag \emph{Locked} and proceeds to the next step.
+If the spin count exceeds the \emph{Maximum Spin Count}, the workgroup signals a fallback operation \john{I don't get ``signals''.
+To whom are we signaling?
+What is signaled?} leaving the control flag \emph{Locked} and proceeds to the next step.
+\john{I think this is the place where decoupled-lookback fails without FPG\@.
+If so, say so, and why it fails.
+It underscores the central message of the paper, and is thus worth reiterating.}
         \end{itemize}
 
-  \item \textbf{Fallback:}
+  \item \textbf{Fallback:} \john{Start with a sentence that describes \emph{why} we've reached this phase.
+``Even after waiting, we see that a previous workgroup has not completed the reduction of its tile.'', that sort of thing.}
         \begin{itemize}
@@
           \item Once the fallback reduction is complete, the workgroup \emph{attempts} to post the reduction to global memory and update the tile's state to \emph{Ready}.
-          \item As this update attempt is performed atomically, the workgroup reads back the blocking tile's immediate state:
+          \item As this update attempt is performed atomically, the workgroup reads back the blocking tile's immediate state: \john{``immediate''? instead, ``most recent''?}
                 \begin{itemize}
-                  \item If the state is \emph{Inclusive}, the fallback value is discarded, and the workgroup accumulates the value before proceeding to step 4.
+                  \item If the state is \emph{Inclusive}, the fallback value calculated by our workgroup is discarded, and our workgroup accumulates the value then proceeds to step 4.
                   \item If the blocking tile is the first in the spine, the workgroup has reached the start of the spine and proceeds to step 4.
                   \item Otherwise, the workgroup resumes the lookback, starting with the preceding tile.
+\john{\ldots\ by setting the current tile to \ldots\ I'm trying to not be sloppy with ``starting with'' given that we have a concept of ``current tile''.}
                 \end{itemize}
@@
           \item Upon completing the lookback or fallback, the local reduction is added to the \emph{previous reduction} to compute the inclusive reduction, which is then posted to global memory, transitioning the tile state to \emph{Inclusive}.
-      \item The \emph{previous reduction} is posted to shared memory for the workgroup to consume and propagate as necessary.
+          \item The \emph{previous reduction} is posted to workgroup shared memory for the workgroup to consume and propagate as necessary.
+\john{``consume and propagate'' is super non-specific.
+What is happening here and how is it different than the next sentence?}
           \item The control flag is updated to \emph{Unlocked} to signal the exit of the \emph{Decoupled-Fallback} phase, allowing subsequent operations to proceed.
+\john{Be specific about what happens here!
+I'm presuming part of this item's work + the previous item's work is ``use the previous reduction as the starting point for scanning then outputting our current tile''.
+``subsequent operations'' is super non-specific.} \john{In this item + the previous item, I bet there's some ordering constraints, either for correctness or for performance.
+For instance, I bet the highest priority is to add the previous reduction to my reduction and posting the result, since that propagation speed across tiles is probably really important for performance.
+Perhaps worth pointing out.
+Does the flag-unset have any ordering constraints in terms of correctness?}
         \end{itemize}
@@
   \includegraphics[width=\linewidth]{graphics/speedup.pdf}
+  \john{\caption{This data (10 data points, which is pretty scant) is likely better presented as a table (also a better use of precious paper real estate).
+Also, make tables and figures float (don't do \texttt{[h!]}, just leave that blank, \LaTeX\ is good at page layout.)}}
 \end{figure}

### 2025-01-29 paper.tex
#7 added another thought

@@
 \john{Maybe a sentence that describes the state progression at a high level?
-e.g., ``the state of each workgroup begins at INVALID, moves to READY, then to INCLUSIVE, the state never moves backwards, each workgroup can operate independently and change its own state and also, during lookback, the state of preceding workgroups'', that sort of thing.}
+e.g., ``the state of each workgroup begins at INVALID, moves to READY, then to INCLUSIVE, the state never moves backwards, each workgroup can operate independently and change its own state and also, during lookback, the state of preceding workgroups'', that sort of thing.
+Also, we present the algorithm from the perspective of one workgroup, which at all times has a pointer to a ``current tile'' on which it is working.
+(Sloppily worded, sorry.)}
 \john{I might also say something like ``in the following discussion, we assume the `add' monoid and use terms like `add' and `sum', but the algorithm works with any monoid''.

### 2025-01-30 artifact/main.cpp
Switch to WGPU SPIR-V

@@ -1,994 +1,0 @@
-#include <dawn/webgpu_cpp.h>
-
-#include <cstdlib>
-#include <cstring>
-#include <fstream>
-#include <future>
-#include <iostream>
-#include <sstream>
-#include <string>
-#include <vector>
-
-constexpr uint32_t MAX_EMULATE = 10;
-
-struct GPUContext {
-    wgpu::Instance instance;
-    wgpu::Device device;
-    wgpu::Queue queue;
-    wgpu::QuerySet querySet;
-};
-
-struct ComputeShader {
-    wgpu::BindGroup bindGroup;
-    wgpu::ComputePipeline computePipeline;
-    std::string label;
-};
-
-struct Shaders {
-    ComputeShader init;
-    ComputeShader reduce;
-    ComputeShader spineScan;
-    ComputeShader downsweep;
-    ComputeShader csdl;
-    ComputeShader csdldf;
-    ComputeShader csdldfOcc;
-    ComputeShader csdldfEmulate[MAX_EMULATE + 1];
-    ComputeShader memcpy;
-    ComputeShader validate;
-};
-
-struct GPUBuffers {
-    wgpu::Buffer info;
-    wgpu::Buffer scanIn;
-    wgpu::Buffer scanOut;
-    wgpu::Buffer scanBump;
-    wgpu::Buffer reduction;
-    wgpu::Buffer timestamp;
-    wgpu::Buffer readbackTimestamp;
-    wgpu::Buffer readback;
-    wgpu::Buffer misc;
-};
-
-struct TestArgs {
-    GPUContext& gpu;
-    GPUBuffers& buffs;
-    Shaders& shaders;
-    uint32_t size;
-    uint32_t batchSize;
-    uint32_t threadBlocks;
-    uint32_t readbackSize;
-    uint32_t fallbackPreset;
-    bool shouldValidate = false;
-    bool shouldReadback = false;
-    bool shouldTime = false;
-    bool shouldGetStats = false;
-    bool shouldRecord = false;
-};
-
-struct DataStruct {
-    std::vector<double> time;
-    std::vector<uint32_t> totalSpins;
-    std::vector<double> lookbackLength;
-    std::vector<uint32_t> fallbacksInitiated;
-    std::vector<uint32_t> successfulInsertions;
-};
-
-void GetGPUContext(GPUContext* context, uint32_t timestampCount) {
-    wgpu::InstanceDescriptor instanceDescriptor{};
[…]

### 2025-01-30 artifact/src/main.rs

@@ -1,0 +1,1141 @@
+use std::{
+    env,
+    fs::File,
+    io::{self, Write},
+    vec,
+};
+
+fn div_round_up(x: u32, y: u32) -> u32 {
+    (x + y - 1) / y
+}
+
+#[derive(Debug)]
+pub enum TestType {
+    Csdl,
+    Csdldf,
+    Full,
+}
+
+fn print_usage() {
+    eprintln!("Usage: <TestType: \"csdl\"|\"csdldf\"|\"full\"> [\"record\"] [deviceName]");
+}
+
+struct GPUContext {
+    device: wgpu::Device,
+    queue: wgpu::Queue,
+    query_set: wgpu::QuerySet,
+    timestamp_freq: f32,
+}
+
+impl GPUContext {
+    async fn init() -> Self {
+        let instance = wgpu::Instance::new(wgpu::InstanceDescriptor {
+            backends: wgpu::Backends::all(),
+            flags: wgpu::InstanceFlags::empty(),
+            dx12_shader_compiler: wgpu::Dx12Compiler::default(),
+            gles_minor_version: wgpu::Gles3MinorVersion::default(),
+        });
+
+        let adapter = instance
+            .request_adapter(&wgpu::RequestAdapterOptions {
+                power_preference: wgpu::PowerPreference::HighPerformance,
+                compatible_surface: None,
+                force_fallback_adapter: false,
+            })
+            .await
+            .expect("Failed to find an appropriate adapter");
+
+        let adapter_info = adapter.get_info();
+        println!("Adapter Info:");
+        println!("  Name: {}", adapter_info.name);
+        println!("  Vendor: {}", adapter_info.vendor);
+        println!("  Device: {}", adapter_info.device);
+        println!("  Backend: {:?}", adapter_info.backend);
+        println!("  Driver: {}", adapter_info.driver);
+        println!("  Driver Info: {}", adapter_info.driver_info);
+
+        let (device, queue) = adapter
+            .request_device(
+                &wgpu::DeviceDescriptor {
+                    label: None,
+                    required_features: wgpu::Features::TIMESTAMP_QUERY
+                        | wgpu::Features::SUBGROUP
+                        | wgpu::Features::SPIRV_SHADER_PASSTHROUGH,
+                    required_limits: wgpu::Limits::default(),
+                    memory_hints: wgpu::MemoryHints::Performance,
+                },
+                None,
+            )
+            .await
+            .expect("Failed to create device");
[…]

### 2025-01-30 paper.tex
Add Authors, Abstract, Edits

--- 2025-01-29 paper.tex
@@
 
-\title{The Name of the Title Is Hope}
+\title{Decoupled-Fallback: A Portable Single-Pass Scan}
 
-\author{Lars Th{\o}rv{\"a}ld}
+\author{Thomas Smith}
 \affiliation{%
-\institution{The Th{\o}rv{\"a}ld Group}
-\city{Hekla}
-\country{Iceland}}
-\email{larst@affiliation.org}
+\institution{Google}
+\country{USA}}
 
-\author{Valerie B\'eranger}
+\author{Raph Levien}
 \affiliation{%
-  \institution{Inria Paris-Rocquencourt}
-  \city{Rocquencourt}
-  \country{France}
+  \institution{Google}
+  \country{USA}
 }
 
-\author{Aparna Patel}
+\author{John D. Owens}
 \affiliation{%
-  \institution{Rajiv Gandhi University}
-  \city{Doimukh}
-  \state{Arunachal Pradesh}
-  \country{India}}
+  \institution{University of California, Davis}
+  \country{USA}}
+  \additionalaffiliation{%
+  \institution{Google}
+  \country{USA}
+}
 
-\renewcommand{\shortauthors}{Trovato et al.}
+\renewcommand{\shortauthors}{Smith et al.}
 
 \begin{abstract}
-  TODO
+  Scan is one of the most fundamental primitives in GPGPU computing, underpinning operations such as graph traversal and sorting.
+Although the state-of-the-art scan algorithm, \emph{Decoupled-Lookback}, achieves near-ideal (\emph{memcopy}) performance on NVIDIA hardware when implemented in CUDA, it relies on architectural and language-specific guarantees that no longer hold once outside of the NVIDIA ecosystem---most prominently, a forward progress guarantee (FPG).
+Without FPG, the algorithm becomes highly susceptible to deadlock, necessitating regression to slower approaches.
+We introduce \emph{Decoupled-Fallback}, a method that enables \emph{Decoupled-Lookback} to operate on hardware without FPG while avoiding deadlock, alongside other portability adaptations.
+On an FPG-lacking device---the Apple M1 Pro (8+2)---\emph{Decoupled-Fallback} achieves near-\emph{memcopy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-Then-Scan} approach.
 \end{abstract}
@@
 
-\begin{figure}[h]
+\begin{figure}
   \centering
@@
 While eliminating \emph{resource holding}---allowing workgroups to discard work tiles when they encounter a deadlock---is theoretically possible, it is strictly worse than eliminating \emph{mutual exclusion}, as it necessitates $t$ additional movements (discard-read-read versus read).
-\begin{algorithm}[htb]
+\begin{algorithm}
   \small
@@
   \KwGlobalIn{%
-    $in[]$,
-    $out[]$,
-    $spine[]$,
-    $tileBump$,
+    $\textit{in}[]$,
+    $\textit{out}[]$,
+    $\textit{spine}[]$,
+    $\textit{tileBump}$,
   }
   \KwSharedIn{%
-    $wg\_partials[]$,
-    $wg\_fallback[]$,
-    $wg\_prev\_reduction$,
-    $wg\_broadcast$,
-    $wg\_control$,
+    $\textit{wg\_partials}[]$,
+    $\textit{wg\_fallback}[]$,
+    $\textit{wg\_prev\_reduction}$,
+    $\textit{wg\_broadcast}$,
+    $\textit{wg\_control}$,
   }
-  \KwGlobalOut{$out[]$ inclusive scan of $in[]$}
-  \KwConstants{Vectors per thread $p$}
+  \KwGlobalOut{$\textit{out}[]$ inclusive scan of $\textit{in}[]$}
+  \KwConstants{Vectors per thread $\textit{p}$}
 
-  \If{$threadid = 0$}{
-    $wg\_broadcast \gets atomicAdd(\&tileBump, 1)$\;
-    $wg\_control \gets LOCKED$\;
+  \If{$\textit{threadid} = 0$}{
+    $\textit{wg\_broadcast} \gets \textnormal{atomicAdd}(\&\textit{tileBump}, 1)$\;
+    $\textit{wg\_control} \gets \textnormal{LOCKED}$\;
   }
@@
 
-  $part\_id \gets wg\_broadcast$\;
+  $\textit{part\_id} \gets \textit{wg\_broadcast}$\;
 
-  \textit{local\_vectors}: $\textnormal{array}\langle \textnormal{vec4}\langle u32\rangle, p\rangle$\;
+  \textit{local\_vectors}: $\textnormal{array}\langle \textnormal{vec4}\langle u32\rangle, \textit{p} \rangle$\;
 
-  \textbf{Load}($in$, $part\_id$, \textit{local\_vectors})\;
+  \textbf{Load}($\textit{in}$, $\textit{part\_id}$, \textit{local\_vectors})\;
   \textbf{SubgroupRake}(\textit{local\_vectors})\;
 
-  \If{$highestRankLane(threadid)$}{
-    $wg\_partials[subgroup\_id] \gets local\_vectors[final\_partial]$\;
+  \If{$\textnormal{highestRankLane}(\textit{threadid})$}{
+    $\textit{wg\_partials}[\textit{subgroup\_id}] \gets \textit{local\_vectors}[\textnormal{final\_partial}]$\;
   }
@@
 
-  \textbf{WorkgroupWideScan}($wg\_partials$)\;
+  \textbf{WorkgroupWideScan}($\textit{wg\_partials}$)\;
   \textbf{barrier()}\;
@@
   \tcp{Decoupled Fallback}
-  \While{$wg\_control = LOCKED$}{
-    \textbf{Lookback}($spine$, $wg\_prev\_reduction$, $wg\_control$)\;
+  \While{$\textit{wg\_control} = \textnormal{LOCKED}$}{
+    \textbf{Lookback}($\textit{spine}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
     \textbf{barrier()}\;
 
-    \If{$wg\_control = LOCKED$}{
-      \textbf{FallbackReduce}($in$, $wg\_fallback$)\;
-      \textbf{FallbackInsertionAttempt}($spine$, $wg\_fallback$, $wg\_prev\_reduction$, $wg\_control$)\;
+    \If{$\textit{wg\_control} = \textnormal{LOCKED}$}{
+      \textbf{FallbackReduce}($\textit{in}$, $\textit{wg\_fallback}$)\;
+      \textbf{FallbackInsertionAttempt}($\textit{spine}$, $\textit{wg\_fallback}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
       \textbf{barrier()}\;
@@
 
-  \textbf{Write}($out$, $part\_id$, \textit{local\_vectors}, $wg\_partials$, $wg\_prev\_reduction$)\;
+  \textbf{Write}($\textit{out}$, $\textit{part\_id}$, \textit{local\_vectors}, $\textit{wg\_partials}$, $\textit{wg\_prev\_reduction}$)\;
 
@@
 For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
-  \item \textbf{Subgroup-Rake}: Each thread performs a private, serial scan across its vectors and then participates in $p$ raking subgroup scans.
-This allows the entire scan across the subgroup-block to be barrier-free and remain in registers.
-The final subgroup-block reduction is then posted to shared memory.
+  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking subgroup scans,
+  allowing the subgroup-block scan to remain barrier-free and in registers.The subgroup-block reduction is then posted to shared memory.
+For a workgroup of size $w$, subgroup-raking reduces the local spine to $\frac{w}{s}$.
   \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
@@
 Phase 4, \emph{Decoupled-Fallback}, is responsible for managing all inter-workgroup dependencies.
+As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations.
 
-\subsection{Intra-Workgroup Portability Adaptations}
-As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations.
-\subsubsection{Subgroup-Raking}
-We desire the smallest intra-workgroup spine possible.
-Not only does this reduce shared memory usage, but it also directly impacts the number of workgroup barriers encountered during the workgroup-wide scan, as this number is proportional to the size of the local spine.
-While large subgroups are relatively insensitive to this effect, small subgroups are particularly affected (reference depth here).
-To address this, we perform $p$ raking iterations of subgroup Kogge-Stone scans, where the highest-ranked thread passes its output into each succeeding scan.
-For a workgroup of size $w$, this approach reduces the number of partial reductions that each subgroup must post from $p$ to $1$, and it decreases the size of the spine from $\frac{pw}{s}$ to $\frac{w}{s}$.
-\begin{algorithm}[htbp]
+\begin{algorithm}
   \small
   \SetAlgoLined
-  \KwSharedIn{ Array of subgroup partial reductions $x$ }
-  \KwSharedOut{ Post-scanned array $x$ }
-  \KwConstants{ Workgroup Size $W$, Subgroup Size $S$ }
+  \KwSharedIn{ Array of subgroup partial reductions $\textit{x}$ }
+  \KwSharedOut{ Post-scanned array $\textit{x}$ }
+  \KwConstants{ Workgroup Size $\textit{W}$, Subgroup Size $\textit{S}$ }
 
-  $spine\_length \gets W / S$\;
-  $alignment \gets 1 << \text{divRoundUp}(\log_2(spine\_length), \log_2(S)) \cdot \log_2(S)$\;
+  $\textit{spine\_length} \gets \textit{W} / \textit{S}$\;
+  $\textit{alignment} \gets 1 << \text{divRoundUp}(\log_2(\textit{spine\_length}), \log_2(\textit{S})) \cdot \log_2(\textit{S})$\;
 
-  $stride \gets 1$\;
-  $top\_offset \gets 0$\;
+  $\textit{stride} \gets 1$\;
+  $\textit{top\_offset} \gets 0$\;
 
-  \ForEach{$thread\_id$ \textbf{in} $W$ \textbf{in parallel}}{
-    \For{$j \gets S$ \KwTo $alignment$ \textbf{with} $j \gets j \cdot S$}{
+  \ForEach{$\textit{thread\_id}$ \textbf{in} $\textit{W}$ \textbf{in parallel}}{
+    \For{$\textit{j} \gets \textit{S}$ \KwTo $\textit{alignment}$ \textbf{with} $\textit{j} \gets \textit{j} \cdot \textit{S}$}{
       \tcp{Brent-Kung Upsweep}
 
-      $step \gets spine\_length / stride$\;
+      $\textit{step} \gets \textit{spine\_length} / \textit{stride}$\;
 
-      \If{$thread\_id < step$}{
-        $temp \gets \text{subgroupInclusiveScan}(x[thread\_id + top\_offset])$\;
-        $x[thread\_id + top\_offset] \gets temp$\;
-        \If{$\text{\normalfont highestRankLane}(thread\_id)$}{
-          $x[(thread\_id / S) + step + top\_offset] \gets temp$\;
+      \If{$\textit{thread\_id} < \textit{step}$}{
+        $\textit{temp} \gets \text{subgroupInclusiveScan}(\textit{x}[\textit{thread\_id} + \textit{top\_offset}])$\;
+        $\textit{x}[\textit{thread\_id} + \textit{top\_offset}] \gets \textit{temp}$\;
+        \If{$\text{\normalfont highestRankLane}(\textit{thread\_id})$}{
+          $\textit{x}[(\textit{thread\_id} / \textit{S}) + \textit{step} + \textit{top\_offset}] \gets \textit{temp}$\;
         }
@@
 
-      \If{$j \neq S$}{
-        $reduced\_stride \gets j / stride$\;
-        $fanout\_index \gets thread\_id + reduced\_stride$\;
+      \If{$\textit{j} \neq \textit{S}$}{
+        $\textit{reduced\_stride} \gets \textit{j} / \textit{stride}$\;
+        $\textit{fanout\_index} \gets \textit{thread\_id} + \textit{reduced\_stride}$\;
 
-        $cond1 \gets fanout\_index < spine\_length$\;
-        $cond2 \gets (fanout\_index \& (j - 1)) \geq reduced\_stride$\;
+        $\textit{cond1} \gets \textit{fanout\_index} < \textit{spine\_length}$\;
+        $\textit{cond2} \gets (\textit{fanout\_index} \& (\textit{j} - 1)) \geq \textit{reduced\_stride}$\;
 
-        \If{$cond1 \textbf{ \&\& } cond2$}{
-          $x[fanout\_index] \gets x[fanout\_index] + x[((fanout\_index / stride) + top\_offset) - 1]$\;
+        \If{$\textit{cond1} \textbf{ \&\& } \textit{cond2}$}{
+          $\textit{x}[\textit{fanout\_index}] \gets \textit{x}[\textit{fanout\_index}] + \textit{x}[((\textit{fanout\_index} / \textit{stride}) + \textit{top\_offset}) - 1]$\;
         }
       }
-      $stride \gets stride \cdot S$\;
-      $top\_offset \gets top\_offset + step$\;
+      $\textit{stride} \gets \textit{stride} \cdot \textit{S}$\;
+      $\textit{top\_offset} \gets \textit{top\_offset} + \textit{step}$\;
     }
@@
   }
-  \Return{$x$}\;
-  \caption{Workgroup-Wide Scan.
-\john{Sorry to say that every one of those variable names need to be wrapped in \texttt{\textbackslash textit}.
-``Typeset in math mode'' is not the same as ``Typeset in italics'' (see $offset$ vs.\ \textit{offset}).}}
+  \Return{$\textit{x}$}\;
+  \caption{Workgroup-Wide Scan.}
   \label{alg:example}
@@
 
-\subsubsection{Workgroup-Wide Scan}
+\subsection{Workgroup-Wide Scan}
 The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches.
@@
 \subsection{Decoupled-Fallback}
-To reiterate, the goal of a lookback is twofold: to determine the reduction of all preceding work tiles and to communicate those values to succeeding tiles to accelerate their own lookback operations.
-Following the \emph{Decoupled-Lookback} technique, we model the spine as a finite state machine, where each work tile comprising the spine may exist in one of three states:
+To reiterate, the goal of the lookback phase is twofold: to determine the reduction of all preceding work tiles and to communicate those values to succeeding tiles to accelerate their own lookback operations.
+Following the \emph{Decoupled-Lookback} technique, we model the inter-workgroup spine as a finite state machine, where each work tile comprising the spine may exist in one of three states:
 \begin{itemize}
@@
 \end{itemize}
-Each lookback is, therefore, a traversal from the \emph{Ready} state to the \emph{Inclusive} state. To illustrate this process, we present a flowchart (SEE LOOKBACK FLOWCHART).
+As the spine resides in global memory, all tiles are initialized to the \emph{Not Ready} state prior to kernel launch, with their corresponding reduction values set to the identity element of the monoid.
+Over the course of the lookback, each tile progresses from \emph{Not Ready} to \emph{Ready}, then to \emph{Inclusive}, and once advanced, may never revert to a prior state.
+Although workgroups depend on preceding tiles, they operate independently and are primarily responsible for updating their own state, only opportunistically updating other tiles' states in the event of a fallback.
+To illustrate this process, we present the flowchart in Figure~\ref{fig:decoupled_fallback}.
 \john{It's helpful, I think, to also explicitly say what data is stored globally as the spine (I think, for each workgroup, one data element and one state).}
+\john{Maybe a sentence that describes the state progression at a high level?
+e.g., ``the state of each workgroup begins at INVALID, moves to READY, then to INCLUSIVE, the state never moves backwards, each workgroup can operate independently and change its own state and also, during lookback, the state of preceding workgroups'', that sort of thing.}
 
-\subsubsection{Decoupled-Fallback}
-\thomas{Given that "lock" has an extensive history of use in computer science and this is not that kind of lock, perhaps a better choice of phrase would be better.
-I've decided to refer to it as a "control flag"} \john{Good idea.
-``Status flag'', perhaps?}
-Prior to kernel launch, all tiles in the spine are initialized to the \emph{Not Ready} state.
  During partition index acquisition, we also initialize a shared memory \emph{control flag} to facilitate synchronization within the workgroup during the \emph{Decoupled-Fallback} phase.
 This flag governs both the initiation of fallback operations and the entry and exit of the entire phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.
-\john{Maybe a sentence that describes the state progression at a high level?
-e.g., ``the state of each workgroup begins at INVALID, moves to READY, then to INCLUSIVE, the state never moves backwards, each workgroup can operate independently and change its own state and also, during lookback, the state of preceding workgroups'', that sort of thing.}
+ \john{Also, we present the algorithm from the perspective of one workgroup, which at all times has a pointer to a ``current tile'' on which it is working. (Sloppily worded, sorry.)}
 \john{I might also say something like ``in the following discussion, we assume the `add' monoid and use terms like `add' and `sum', but the algorithm works with any monoid''.
@@
 I don't feel strongly about this.}
-\begin{figure}[h]
+\begin{figure}
   \centering
@@
   \caption{Decoupled-Fallback}
+  \label{fig:decoupled_fallback}
 \end{figure}
@@
 
-\subsubsection{Tile State Implementation}
+\subsection{Tile State Implementation}
 Because a tile may be in three possible states, two bits are required to record its state alongside the 32 bits needed to store the tile's reduction value.
@@
 Memcopy is included as a point of reference, as it shares the same $2n$ global memory movement required by scan operations.
-\begin{figure}[h!]
-  \centering
-  \includegraphics[width=\linewidth]{graphics/speedup.pdf}
-  \john{\caption{This data (10 data points, which is pretty scant) is likely better presented as a table (also a better use of precious paper real estate).
-Also, make tables and figures float (don't do \texttt{[h!]}, just leave that blank, \LaTeX\ is good at page layout.)}}
-\end{figure}
+
+\thomas{Forgot about tables. Mega table incoming.}
+
 Since WebGPU limits device-side timestamps to compute or render passes, we implement a custom Memcopy kernel for timing purposes.
@@
 Starting with infrequent omissions---1 in 512 workgroups---we progressively increase the rate to near-continuous levels, reaching 1 in 2 workgroups.
-\begin{figure}[htbp]
+\begin{figure}
   \centering

### 2025-01-30 artifact/src/main.rs

@@ -9,17 +9,6 @@ fn div_round_up(x: u32, y: u32) -> u32 {
     (x + y - 1) / y
 }
 
-#[derive(Debug)]
-pub enum TestType {
-    Csdl,
-    Csdldf,
-    Full,
-}
-
-fn print_usage() {
-    eprintln!("Usage: <TestType: \"csdl\"|\"csdldf\"|\"full\"> [\"record\"] [deviceName]");
-}
-
 struct GPUContext {
     device: wgpu::Device,
     queue: wgpu::Queue,
@@ -108,21 +97,6 @@ impl GPUBuffers {
         misc_size: usize,
     ) -> Self {
         let buffer_size = (size * std::mem::size_of::<u32>()) as u64;
-
-        //Vectorize the size here
-        // let param_info: Vec<u32> = vec![
-        //     size as u32,
-        //     div_round_up(size as u32, 4),
-        //     work_tiles as u32,
-        // ];
-        // let params = gpu
-        //     .device
-        //     .create_buffer_init(&wgpu::util::BufferInitDescriptor {
-        //         label: Some("Scan Parameters"),
-        //         contents: bytemuck::cast_slice(&param_info),
-        //         usage: wgpu::BufferUsages::UNIFORM,
-        //     });
-
         let params: wgpu::Buffer = gpu.device.create_buffer(&wgpu::BufferDescriptor {
             label: Some("Scan Parameters"),
             size: (4usize * std::mem::size_of::<u32>()) as u64,
@@ -350,7 +324,6 @@ struct Shaders {
     downsweep: ComputeShader,
     csdl: ComputeShader,
     csdldf: ComputeShader,
-    //csdldf_stats: ComputeShader,
     csdldf_emu: ComputeShader,
     csdldf_occ: ComputeShader,
     memcpy: ComputeShader,
@@ -991,6 +964,16 @@ impl Tester {
     }
 }
 
+pub enum TestType {
+    Csdl,
+    Csdldf,
+    Full,
+}
+
+fn print_usage() {
+    eprintln!("Usage: <TestType: \"csdl\"|\"csdldf\"|\"full\"> [\"record\"] [deviceName]");
+}
+
 struct Args {
     batch_size: u32,        // How many tests to run in a batch
     should_validate: bool,  // Whether to validate the results
@@ -1123,19 +1106,15 @@ fn main() {
 
     // Should Record?
     let mut should_record = false;
-    let mut csv_name = String::new();
-
-    if args.len() >= 3 {
-        if args[2] == "record" {
[…]

### 2025-01-30 paper.tex
Abstract V2

@@
 
-\title{Decoupled-Fallback: A Portable Single-Pass Scan}
+\title{Decoupled-Fallback: A Portable Single-Pass GPU Scan}
 
@@
 \begin{abstract}
-  Scan is one of the most fundamental primitives in GPGPU computing, underpinning operations such as graph traversal and sorting.
-Although the state-of-the-art scan algorithm, \emph{Decoupled-Lookback}, achieves near-ideal (\emph{memcopy}) performance on NVIDIA hardware when implemented in CUDA, it relies on architectural and language-specific guarantees that no longer hold once outside of the NVIDIA ecosystem---most prominently, a forward progress guarantee (FPG).
-Without FPG, the algorithm becomes highly susceptible to deadlock, necessitating regression to slower approaches.
-We introduce \emph{Decoupled-Fallback}, a method that enables \emph{Decoupled-Lookback} to operate on hardware without FPG while avoiding deadlock, alongside other portability adaptations.
-On an FPG-lacking device---the Apple M1 Pro (8+2)---\emph{Decoupled-Fallback} achieves near-\emph{memcopy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-Then-Scan} approach.
+  We present \emph{Decoupled-Fallback}, a method that enables single-pass \emph{Chained-Scans} to run on hardware without \emph{Forward-Progress Guarantees} while avoiding deadlock.
+Additionally, we introduce a tile state representation for \emph{Chained-Scans} that does not rely on 64-bit atomics or memory fences for correctness, along with a subgroup-size-agnostic intra-workgroup implementation.
+On an FPG-lacking device---Apple M1 Pro (8+2)---\emph{Decoupled-Fallback} achieves near-\emph{memcopy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-Then-Scan} approach.
+We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by emulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-Then-Scan} even under extreme contention.
 \end{abstract}

### 2025-01-31 paper.tex
Edits, Incorporating Feedback

@@
 \end{itemize}
-\thomas{See comment in section 3 !}
 Our method retains the best aspects of the \emph{Chained-Scan} architecture, while eliminating the need for forward progress guarantees.
@@
 
-In any \emph{Chained-Scan}, once a workgroup computes its local reduction, it must complete two objectives: communicate its reduction to successor tiles and incorporate the reductions of all preceding tiles.
-In \emph{StreamScan}, this is facilitated by a \emph{propagation} phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume.
+In any \emph{Chained-Scan}, once a workgroup computes its local reduction, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles.
+In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume.
 During this waiting period, the workgroup is continually scheduled, and the \emph{progress} of other workgroups must be guaranteed to prevent the spinning workgroup from starving its dependent predecessors of execution, which could result in deadlock.
@@
 \subsubsection{Chained-Scan, Decoupled-Lookback}
-Merrill and Garland~\cite{Merrill2016} introduced the \emph{Decoupled-Lookback} technique, which separates reduction propagation and dependency resolution to address the bottleneck caused by propagation latency.
+Introduced by Merrill and Garland, the \emph{Decoupled-Lookback} technique separates the communication of reductions from dependency resolution, thus eliding bottlenecks caused by propagation latency.
 In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction.
@@
 \subsection{Ideal Scan Architecture}
-The design of an ideal scan architecture must balance performance, portability, and practicality.
 Beyond their differences in speed, \emph{Chained-Scan} and \emph{Reduce-Then-Scan} embody two fundamentally contrasting paradigms.
@@
 Fixed-launch approaches demand substantial developer effort, requiring either tuning of workgroup launch sizes across target devices or occupancy discovery.
-Moreover, previous workgroup occupancy discovery techniques, such as the one proposed by Sorensen et al.~\cite{sorensen2016}, are infeasible to implement in WebGPU due to their reliance on memory fences---a feature not supported by the specification.\footnote{Our implementation of \emph{Reduce-Then-Scan}, which can be found in the artifact, includes a fenceless occupancy discovery method amenable to the WebGPU environment, but its description is beyond the scope of this work.}
+Moreover, previous workgroup occupancy discovery techniques, such as the one proposed by Sorensen et al.~\cite{sorensen2016}, are infeasible to implement in WebGPU due to their reliance on memory fences---a feature not supported by the specification.\footnote{Our implementation includes a fenceless occupancy discovery method amenable to the WebGPU environment, but its description is beyond the scope of this work.}
 
@@
 \section{Decoupled-Fallback}
-\thomas{I've broken somewhat with the outline and moved the goals and non-goals section to somewhere in the introduction.
-I kept finding myself trying to write around it.
-I think the content there is good, but it's very awkward as it ended up so deep within the paper.
-The placement of this section here feels more natural, and seems to mirror other papers?}
 \emph{Decoupled-Fallback} separates fallback operations from the lookback process, enhancing the \emph{Scalar-Fallback} technique with significantly improved fallback performance.
@@
 
-\thomas{What I want to accomplish here is to go beyond saying "Here is our method.
-It works great." I want to show that these are the possible avenues to attack deadlocking .
-.
-.
-and exhaustively demonstrate that removing mutual exclusion is the best way to do it, under our given constraints.}
 Like \emph{Scalar-Fallback}, we choose \emph{mutual exclusion} as our approach to break deadlocking.
@@
   \If{$\textnormal{highestRankLane}(\textit{threadid})$}{
-    $\textit{wg\_partials}[\textit{subgroup\_id}] \gets \textit{local\_vectors}[\textnormal{final\_partial}]$\;
+    $\textit{wg\_partials}[\textit{subgroup\_id}] \gets \textit{local\_vectors}[\textit{p} - 1]$\;
   }
@@
   \KwSharedIn{ Array of subgroup partial reductions $\textit{x}$ }
-  \KwSharedOut{ Post-scanned array $\textit{x}$ }
+  \KwSharedOut{ Inclusive-scanned $\textit{x}$ }
   \KwConstants{ Workgroup Size $\textit{W}$, Subgroup Size $\textit{S}$ }
@@
 \subsection{Decoupled-Fallback}
-To reiterate, the goal of the lookback phase is twofold: to determine the reduction of all preceding work tiles and to communicate those values to succeeding tiles to accelerate their own lookback operations.
-Following the \emph{Decoupled-Lookback} technique, we model the inter-workgroup spine as a finite state machine, where each work tile comprising the spine may exist in one of three states:
+To reiterate, the goal of the lookback is twofold: to determine the reduction of all preceding work tiles and to relay those values to succeeding tiles to accelerate their own lookback operations.
+Following the \emph{Decoupled-Lookback} technique, we model the inter-workgroup spine as a finite-state machine, where each work tile comprises a reduction value and a state, stored in global memory.
+Each tile may exist in one of three states:
 \begin{itemize}
@@
 \end{itemize}
-As the spine resides in global memory, all tiles are initialized to the \emph{Not Ready} state prior to kernel launch, with their corresponding reduction values set to the identity element of the monoid.
-Over the course of the lookback, each tile progresses from \emph{Not Ready} to \emph{Ready}, then to \emph{Inclusive}, and once advanced, may never revert to a prior state.
-Although workgroups depend on preceding tiles, they operate independently and are primarily responsible for updating their own state, only opportunistically updating other tiles' states in the event of a fallback.
-To illustrate this process, we present the flowchart in Figure~\ref{fig:decoupled_fallback}.
+Each workgroup is responsible for updating its own state, only opportunistically updating other tiles' states in the event of a fallback.
+A tile can only transition forward---Not Ready \rightarrow Ready \rightarrow Inclusive—and once advanced, may never revert to a prior state.
+Before kernel launch, all tiles in the spine are initialized to \emph{Not Ready}, and their reduction values are set to $\epsilon$.
+Workgroups operate independently but remain dependent on preceding tiles to finalize their own inclusive reductions.
+As they transition through states, workgroups progressively accumulate reductions from prior tiles and, once their lookback phase is complete, post their inclusive reduction for the next tile to consume.
 \john{It's helpful, I think, to also explicitly say what data is stored globally as the spine (I think, for each workgroup, one data element and one state).}
@@
 
- During partition index acquisition, we also initialize a shared memory \emph{control flag} to facilitate synchronization within the workgroup during the \emph{Decoupled-Fallback} phase.
-This flag governs both the initiation of fallback operations and the entry and exit of the entire phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.
+To coordinate this process, each workgroup initializes a shared memory \emph{control flag} during partition index acquisition.
+This flag governs both the initiation of fallback operations and the entry and exit of the entire \emph{Decoupled-Fallback} phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.
+The details of tile state encoding and atomic updates are discussed in the following section and may vary depending on the data type, so for simplicity, we describe the lookback here as being performed by a single thread.
+
 \john{Also, we present the algorithm from the perspective of one workgroup, which at all times has a pointer to a ``current tile'' on which it is working. (Sloppily worded, sorry.)}
@@
 I don't feel strongly about this.}
+\thomas{I'm going to use ``ingest'' to match the flow chart. I literally used a thesaurus to choose accumulate . . . Can still change later.}
+
+We now detail the \emph{Decoupled-Fallback} procedure, presenting the algorithm from the perspective of a single workgroup:
 \begin{figure}
@@
 \begin{enumerate}
-  \item[(0)] \textbf{Posting Local Reduction:} Each workgroup posts its local reduction---the result of the workgroup-wide scan over partials---to global memory and updates its tile state to \emph{Ready}.
-If it is the first work tile, its local reduction \emph{is} its inclusive reduction.
+  \item[(0)] \textbf{Posting Local Reduction:} 
+        \begin{itemize}
+          \item Each workgroup posts its local reduction---the result of the workgroup-wide scan over partials---to global memory and updates its tile state to \emph{Ready}.
+          \item If it is the first work tile, its local reduction \emph{is} its inclusive reduction.
 In this case, the workgroup updates its tile state directly to \emph{Inclusive} and bypasses the lookback entirely.
+        \end{itemize}
 
@@
           \item The control flag, initialized to \emph{Locked}, signals that the workgroup should enter the lookback phase.
-          \item Each workgroup initializes a \john{shared workgroup-local (?)} \emph{previous reduction} variable with the identity element of the monoid to hold the accumulated reductions of preceding tiles.
-          \item The workgroup prepares to traverse the spine, starting with its immediate predecessor tile.
-\john{Carefully look at the way this is worded with respect to ``immediate predecessor'' (previous sentence) vs.\ ``current'' (next sentence).
-I think here we want to say that we set the current tile to the immediate predecessor, and then the next sentence is consistent.}
+          \item The lookback thread initializes a \emph{previous reduction} variable in registers with $\epsilon$ to hold the accumulated reductions of preceding tiles.
+          \item To traverse the spine, the lookback thread initializes a \emph{current tile} pointer to its immediate predecessor.  
         \end{itemize}
@@
         \begin{itemize}
-          \item The workgroup queries the state of the current tile.
-          \item If the tile is in the \emph{Ready} state, the workgroup accumulates the value and moves to \john{sets the current tile to} the preceding tile, repeating this step.
-\john{until reaching a tile that is not \emph{Ready}.}
-          \item If the tile is in the \emph{Inclusive} state, the workgroup accumulates the value and proceeds directly to step 4.
-          \item If the tile is in the \emph{Not Ready} state, the workgroup spins for a limited number of iterations, defined by the \emph{Maximum Spin Count}.
-If the spin count exceeds the \emph{Maximum Spin Count}, the workgroup signals a fallback operation \john{I don't get ``signals''.
-To whom are we signaling?
-What is signaled?} leaving the control flag \emph{Locked} and proceeds to the next step.
+          \item The lookback thread queries the state of the \emph{current tile}.
+          \item If the tile is in the \emph{Ready} state, the workgroup ingests the value and sets \emph{current tile} to the preceding tile, repeating this step until reaching a tile that is \emph{Not Ready}.
+          \item If the tile is in the \emph{Inclusive} state, the workgroup ingests the value and proceeds directly to step 4.
+          \item If the tile is in the \emph{Not Ready} state, the lookback thread spins for a limited number of iterations, defined by the \emph{Maximum Spin Count}.
+If this count is exceeded, the thread exits the lookback loop, leaving the control flag \emph{Locked}, signaling to the workgroup that a fallback is required.
+To prepare for the fallback, it stores a copy of the \emph{current tile} in shared memory for the workgroup to process.
+Without FPG, \emph{Decoupled-Lookback} risks spinning indefinitely and deadlocking at this point.
+
+          \john{I don't get ``signals''. To whom are we signaling? What is signaled?}
           \john{I think this is the place where decoupled-lookback fails without FPG\@.
@@
           \item The control flag is updated to \emph{Unlocked} to signal the exit of the \emph{Decoupled-Fallback} phase, allowing subsequent operations to proceed.
+
           \john{Be specific about what happens here!
 I'm presuming part of this item's work + the previous item's work is ``use the previous reduction as the starting point for scanning then outputting our current tile''.
-``subsequent operations'' is super non-specific.} \john{In this item + the previous item, I bet there's some ordering constraints, either for correctness or for performance.
+``subsequent operations'' is super non-specific.}
+
+          \john{In this item + the previous item, I bet there's some ordering constraints, either for correctness or for performance.
 For instance, I bet the highest priority is to add the previous reduction to my reduction and posting the result, since that propagation speed across tiles is probably really important for performance.
@@
 Instead, we use two 32-bit values, splitting the reduction into upper and lower 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state.
-\begin{figure}[h!]
+\begin{figure}
   \centering
@@
 
-\subsection{Requirements}
-
-\subsubsection{Device Requirements}
-Any device supporting Dawn, with subgroup and timestamp query capabilities, and at least 384 MB of available device memory.
-
-\subsubsection{Software Requirements}
-\begin{itemize}
-  \item Git
-  \item CMake 3.10.2+ or another C++ build tool
-  \item C++17-compliant compiler
-  \item Python for fetching dependencies
-\end{itemize}
-
-\subsection{Building}
-Although native WebGPU is supported on various operating systems—most notably Windows, macOS, and Linux—this section will focus on building it in Unix-like environments for the sake of brevity, using CMake as the example build tool.
-
-\subsubsection{Building Dawn}
-\begin{enumerate}
-  \item[(0)] Dawn depends on the following packages (Ubuntu package names):
-        \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
-sudo apt install libxrandr-dev
-  libxinerama-dev libxcursor-dev libxi-dev \
-  libx11-xcb-dev mesa-common-dev
-  \end{lstlisting}
-
-  \item Clone the Dawn repository:
-        \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
-git clone https://dawn.googlesource.com/dawn
-RESET HEAD here to a specific commit. . . to lock in?
-  \end{lstlisting}
-
-  \item Build and install Dawn with your build tool and compiler of choice:
-        \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
-cd dawn
-cmake -S . -B out/Release \
-  -DDAWN_FETCH_DEPENDENCIES=ON \
-  -DDAWN_ENABLE_INSTALL=ON \
-  -DCMAKE_BUILD_TYPE=Release
-...
-cmake --build out/Release
-...
-cmake --install out/Release \
-  --prefix install/Release
-  \end{lstlisting}
-\end{enumerate}
-
-\subsubsection{Building Artifact}
-After successfully building and installing Dawn, follow these steps to build the artifact:
-
-\begin{enumerate}
-  \item Navigate to the artifact directory:
-        \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
-cd Decoupled-Fallback-Paper/artifact/Dawn
-  \end{lstlisting}
-
-  \item Set the `CMAKE\_PREFIX\_PATH` to point to your Dawn installation directory:
-        \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
-export CMAKE_PREFIX_PATH=/path/to/ \
-dawn/install/Release
-  \end{lstlisting}
-
-  \item Build the artifact:
-        \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
-cmake -S . -B out/Release \
-  -DCMAKE_BUILD_TYPE=Release
-  \end{lstlisting}
-
-  \item Compile the project:
-        \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
-cmake --build out/Release
-  \end{lstlisting}
-\end{enumerate}
-
 \end{document}

### 2025-02-01 artifact/src/main.rs
Bump WGPU to 24.0. Remove zeroing.

@@ -18,11 +18,10 @@ struct GPUContext {
 
 impl GPUContext {
     async fn init() -> Self {
-        let instance = wgpu::Instance::new(wgpu::InstanceDescriptor {
+        let instance = wgpu::Instance::new(&wgpu::InstanceDescriptor {
             backends: wgpu::Backends::all(),
             flags: wgpu::InstanceFlags::empty(),
-            dx12_shader_compiler: wgpu::Dx12Compiler::default(),
-            gles_minor_version: wgpu::Gles3MinorVersion::default(),
+            backend_options: wgpu::BackendOptions::default(),
         });
 
         let adapter = instance
@@ -297,7 +296,9 @@ impl ComputeShader {
                     bind_group_layouts: &[&bind_group_layout],
                     push_constant_ranges: &[],
                 });
-
+        
+        let mut comp_options = wgpu::PipelineCompilationOptions::default();
+        comp_options.zero_initialize_workgroup_memory = false;
         let compute_pipeline =
             gpu.device
                 .create_compute_pipeline(&wgpu::ComputePipelineDescriptor {
@@ -305,7 +306,7 @@ impl ComputeShader {
                     layout: Some(&pipeline_layout_init),
                     module,
                     entry_point: Some(entry_point),
-                    compilation_options: Default::default(),
+                    compilation_options: comp_options,
                     cache: Default::default(),
                 });
 

### 2025-02-03 paper.tex
Debloat, incorporate feedback.

@@
 
-\title{Decoupled-Fallback: A Portable Single-Pass GPU Scan}
+\title{Decoupled Fallback: A Portable Single-Pass GPU Scan}
 
@@
 \begin{abstract}
-  We present \emph{Decoupled-Fallback}, a method that enables single-pass \emph{Chained-Scans} to run on hardware without \emph{Forward-Progress Guarantees} while avoiding deadlock.
+  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained-Scans} to run on hardware without \emph{Forward-Progress Guarantees} (FPG) while avoiding deadlock.
 Additionally, we introduce a tile state representation for \emph{Chained-Scans} that does not rely on 64-bit atomics or memory fences for correctness, along with a subgroup-size-agnostic intra-workgroup implementation.
-On an FPG-lacking device---Apple M1 Pro (8+2)---\emph{Decoupled-Fallback} achieves near-\emph{memcopy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-Then-Scan} approach.
-We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by emulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-Then-Scan} even under extreme contention.
+On an FPG-lacking device---Apple M1 Pro (8+2)---\emph{Decoupled-Fallback} achieves near-\emph{memcopy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach.
+We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by emulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
 \end{abstract}
@@
 \subsubsection{Inter-Workgroup Barriers}
-Although threads \emph{within} a workgroup may synchronize with each other by participating in a \emph{barrier} primitive, the virtualization of processors introduces an emergent high-level limitation: the absence of an inter-workgroup barrier.
+The virtualization of processors introduces an emergent high-level limitation: the absence of an inter-workgroup barrier.
 As previously mentioned, workgroups cannot context switch.
@@
 \subsubsection{Fairness and Progress}
-Because multiple workgroups may reside on a single multiprocessor, two key issues arise at the subgroup scheduling level: \emph{fairness}---how evenly execution resources are distributed among threads---and \emph{progress guarantees}---ensuring that all threads or subgroups eventually make progress towards termination.
+Because multiple workgroups may reside on a single multiprocessor, two key issues arise at the subgroup scheduling level: \emph{fairness}---how evenly execution resources are distributed among subgroups---and \emph{progress guarantees}---ensuring that subgroups eventually make progress towards termination.
 However, the term "fairness" is used differently across the literature, leading to potential confusion.
@@
 In contrast, NVIDIA publications~\cite{4523358,Merrill2016} define "fairness" in terms of the distribution of execution resources among subgroups.
-In this work, we adopt NVIDIA's terminology to maintain consistency with their publications, while acknowledging the broader definitions provided by Sorensen et al.
+In this work, we adopt NVIDIA's terminology.
 
@@
 \subsection{Evolution of Inter-Workgroup Scan Architectures}
-Contemporary GPU scans are distinguished from earlier PRAM-like scans~\cite{} by their alignment with the GPU memory hierarchy, which grades memory into progressively faster but increasingly private tiers.
-To operate efficiently, GPU scans must hybridize fine-grain intra-workgroup (\emph{local}) strategies, executed privately within shared memory and registers, with coarse-grain inter-workgroup (\emph{global}) strategies, performed in global memory.
-Due to their relatively low arithmetic intensity, it is theoretically expected and experimentally observed that the performance of local scans are memory-bound, as first demonstrated by Merrill and Grimshaw~\cite{}.
-Consequently, once equipped with a sufficient local scan strategy, the focus of scan design shifts to addressing the challenges of the global strategy, particularly inter-workgroup coordination and synchronization.
-In this section, we review the three prominent GPU global scan strategies---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---examining how they navigate inter-workgroup dependencies and evaluating their effectiveness in minimizing global memory access and maximizing memory bandwidth utilization.
+Contemporary GPU scans depart from early PRAM-like scans~\cite{} by leveraging the GPU memory hierarchy, which consists of progressively faster but increasingly private tiers.
+Efficient scans integrate fine-grained intra-workgroup (\emph{local}) strategies in shared memory and registers with coarse-grained inter-workgroup (\emph{global}) strategies in global memory.
+Given their low arithmetic intensity, local scans are inherently memory-bound, as demonstrated by Merrill and Grimshaw~\cite{}.
+Consequently, recent scan optimizations have shifted toward refining global strategies, particularly inter-workgroup coordination and synchronization.
+This section reviews three key approaches---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---analyzing their handling of inter-workgroup dependencies and efficiency in minimizing global memory access while maximizing bandwidth utilization.
+\thomas{I have written this paragraph like twenty times now, and it's still odious.}
 
 \subsubsection{Scan-then-Propagate}
-Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2011} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases:
+Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2011} extends intra-workgroup scans to large inputs by resolving inter-workgroup dependencies in three phases:
 \begin{enumerate}
-  \item \textbf{Intra-Workgroup (Local) Scan:} Each workgroup performs an inclusive scan on its assigned work tile partition of the input data.
-The results, including both the local scan output and the reduction of each workgroup, are written back to global memory.
-  \item \textbf{Spine Scan:} The reductions from all workgroups, collectively referred to as \emph{spine}, are gathered and scanned to compute the global prefix of the reductions, resolving the dependencies between work tiles; we call the result of this phase the \emph{root}.
+  \item \textbf{Intra-Workgroup (Local) Scan:}  Each workgroup performs an inclusive scan on its assigned work tile, storing local results and reductions in global memory.
+  \item \textbf{Spine Scan:} The workgroup reductions, collectively forming the \emph{spine}, are gathered and scanned to compute the \emph{root}, resolving dependencies between work tiles.
 For large scans, this stage is an insignificant amount of work compared to the first and third.
-  \item \textbf{Propagation:} Each workgroup retrieves its portion of the root and combines it with the result of the local scan to compute the final scan.
+  \item \textbf{Propagation:} Each workgroup retrieves its portion of the root and adds it to the local scan output.
 \end{enumerate}
 
-As each work tile depends on the reductions of previous tiles, the absence of an inter-workgroup barrier necessitates kernel launches to resolve inter-workgroup dependencies.
-This results in the three-phase construction of \emph{Scan-then-Propagate}, with each phase implemented as a separate kernel launch.
-This is further complicated in Sengupta et al.'s implementation, where arbitrarily large inputs are processed through recursion.
-As the size of a work tile limits the maximum number of elements processed without inter-workgroup dependency resolution, the three-phase scan structure must be recursively applied until the spine size fits within a single work tile.
+The absence of inter-workgroup barriers necessitates separate kernel launches for dependency resolution, inherently structuring \emph{Scan-then-Propagate} into three phases.
+Sengupta et al.'s implementation further complicates this by using recursion to handle arbitrarily large inputs.
+As the size of a work tile limits the maximum number of elements processed without inter-workgroup dependency resolution, the three-phase structure must be recursively applied until the spine size fits within a single work tile.
 Thus, an input of size $n$ and a work tile of size $t$ results in a recursive depth of $\lceil \log_t n \rceil$, and $2\cdot\lceil \log_t n \rceil - 1$ kernel launches.
@@
 If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like scan can profit by trading additional computation for reduced memory traffic.
-This trade-off led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{10.1145/1375527.1375559, Merrill2009, 10.1109/TPDS.2012.336, 10.5555/2031978.2032029}: by writing only the work tile reduction to global memory in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated.
-Like \emph{Scan-then-Propagate}, \emph{Reduce-then-Scan} employs a three-phase approach using kernel launches to resolve dependencies.
+This trade-off led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{10.1145/1375527.1375559, Merrill2009, 10.1109/TPDS.2012.336, 10.5555/2031978.2032029}: by storing only the work tile reduction in global memory during the first phase and performing a redundant scan in the third, it eliminates $n$ global data movement while preserving the three-phase structure:
 \begin{enumerate}
   \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result to global memory.
-  \item \textbf{Spine Scan:} The spine is gathered and scanned to produce the root.
+  \item \textbf{Spine Scan:} The spine is scanned to produce the root.
   \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating its portion of the root as it writes to global memory.
@@
 
-Merill and Grimshaw~\cite{Merrill2009} introduced a workgroup \emph{raking} technique that elides the overhead incurred when the recursive depth exceeds two.
-In this approach, the total workgroup occupancy $o$ is determined, and the input is partitioned into non overlapping blocks of size $\frac{n}{t \cdot o}$, which are raked serially by each workgroup.
-This limits the size of the spine from variably sized, proportional to the input size (possibly requiring recursion), to constant sized, and guaranteed to avoid further recursion.
-Assuming $t \geq o$, global memory movement during the spine scan is reduced from $O(n/t)$ to $O(2o)$.
-Kernel launches are limited to three and total global memory movement is reduced to $O(3n + 4o) = O(3n)$.
+Merrill and Grimshaw~\cite{Merrill2009} introduced \emph{workgroup raking} to further reduce recursion overhead when the recursion depth exceeds two.
+Instead of recursively reducing the spine, the input is partitioned into $o$ groups, where $o$ is the total workgroup occupancy.
+Each workgroup sequentially processes its assigned partition, reducing the spine size to a constant and guaranteeing a recursion depth of at most two.
+This reduces global memory movement for the spine scan from $O(n/t)$ to $O(2o)$ and limits kernel launches to three, reducing total global memory movement to $O(3n + 4o) = O(3n)$.
 
-Breitbart~\cite{10.5555/2031978.2032029} was the first to apply more advanced inter-workgroup synchronization techniques to scan.
-Recognizing that workgroup raking naturally aligns with a \emph{Persistent-Threads}~\cite{gupta2012} paradigm, Breitbart utilized an atomics-based inter-workgroup barrier to replace kernel launches.
-Since multiprocessors are not oversubscribed in raking, workgroups remain resident (persist) on the multiprocessors for the entire lifetime of the kernel, enabling synchronization at the barrier without requiring context switching.
-Although global data movement remains at $O(3n)$, this consolidates the scan to a single kernel, minimizing launch overheads.
-\thomas{This also requires FPG, but not ideal to unpack and mention here .
-.
-.}
+Breitbart~\cite{10.5555/2031978.2032029} applied Gupta et al.'s \emph{Persistent-Threads}~\cite{gupta2012} to scan, eliminating kernel launch overheads.
+Recognizing that raking naturally aligns with this paradigm, Breitbart replaced kernel launches with an atomics-based inter-workgroup barrier.
+Since multiprocessors are not oversubscribed in raking, workgroups persist for the entire kernel duration, allowing synchronization at the barrier without context switching.
+Despite maintaining $O(3n)$ global memory movement, consolidating the scan into a single kernel minimizes launch overheads.
 
@@
 
-In any \emph{Chained-Scan}, once a workgroup computes its local reduction, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles.
+In any \emph{Chained-Scan}, once a workgroup computes its local scan, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles.
+This process, which we refer to as \emph{Chained Propagation}, defines the set of inter-workgroup techniques that facilitate dependency resolution.
 In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume.
@@
 \subsubsection{Chained-Scan, Decoupled-Lookback}
-Introduced by Merrill and Garland, the \emph{Decoupled-Lookback} technique separates the communication of reductions from dependency resolution, thus eliding bottlenecks caused by propagation latency.
+Introduced by Merrill and Garland, the \emph{Decoupled-Lookback} technique refines \emph{Chained Propagation} by separating reduction communication from dependency resolution, mitigating bottlenecks caused by propagation latency.
 In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction.
@@
 \subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
-A \emph{forward-progress guarantee} (FPG) is a promise provided by the subgroup scheduler that all occupant subgroups (of possibly different workgroups) eventually make progress toward termination, preventing scenarios where a subgroup is starved of execution time.
-Sorensen et al.~\cite{sorensen2018,sorensen2021} provide the most comprehensive analysis of GPU schedulers.
-However, while they classify \emph{Chained-Scan} as an FPG-reliant algorithm, their primary focus is in exploring and formalizing progress models, rather than their application to any one algorithm.
+A \emph{forward-progress guarantee} (FPG) ensures that all occupant subgroups (from possibly different workgroups) eventually make progress toward termination, preventing indefinite starvation.
+Sorensen et al.~\cite{sorensen2018,sorensen2021} provide the most comprehensive analysis of GPU schedulers, classifying \emph{Chained-Scan} as FPG-reliant.
+However, their work primarily focuses on formalizing progress models rather than their direct application to specific algorithms.
 \thomas{Sorensen places Chained-Scan into the HSA category, but it is unclear if this is correct.
@@
 
-The challenges faced by \emph{Chained-Scan} are analogous to the \emph{Dining Philosophers Problem}, where tasks compete for shared resources, potentially resulting in deadlock.
+The challenges faced by \emph{Chained-Scan} resemble the \emph{Dining Philosophers Problem}, where competing tasks share limited resources, risking deadlock.
 Coffman et al.~\cite{10.1145/356586.356588} define four conditions necessary for deadlock:
@@
 \end{enumerate}
-Applying this schema to \emph{Chained-Scan}, workgroups correspond to "tasks" and work-tile reductions to "resources." Conditions 1–3 are satisfied due to the one-to-one mapping of work tiles to workgroups and the constraints of the GPU programming model.
-Similarly, condition 4a is met by the dependency relationships between work tiles, leaving condition 4b---\emph{circular ordering}---as the key concern.
+Applying this schema to \emph{Chained-Scan}, workgroups correspond to "tasks" and work-tile reductions to "resources." Conditions 1–3 hold due to the one-to-one mapping of work tiles to workgroups and the constraints of the GPU programming model.
+Condition 4a is satisfied by the inter-tile dependencies, leaving condition 4b---\emph{circular ordering}---as the key concern.
 
-Although the serial assignment of work tiles ensures that predecessor work tiles are either scheduled or completed, there is no guarantee that the corresponding reductions will be available before a workgroup begins its lookback.
-This uncertainty arises because workgroups are scheduled individually, and a predecessor workgroup might not yet have had the opportunity to compute and post its reduction \thomas{There is a better wording here.
-Something like, "because although workgroups \emph{may} share execution resources, they are ultimately scheduled individually." This sentence is incredibly important.}.
-In such cases, the dependent workgroup enters a waiting state, spinning until the reduction becomes available.
-During this waiting period, the spinning workgroup remains scheduled, consuming computational resources.
-Without FPG, the scheduler may indefinitely prioritize the spinning workgroup over its predecessor, preventing the predecessor from completing its reduction.
-This creates a circular dependency, satisfying condition 4b and resulting in deadlock within the kernel.
+Although the serial assignment of work tiles ensures that predecessor tiles are scheduled or completed, it does not guarantee their reductions will be available when a workgroup begins \emph{Chained Propagation}.
+This uncertainty arises because workgroups are scheduled \emph{individually}, meaning a predecessor may not have had the opportunity to compute and post its reduction.
+In such cases, the dependent workgroup spins, consuming execution resources while waiting.
+Because workgroups may share execution resources by being co-resident on the same multiprocessor, without FPG, the scheduler may indefinitely prioritize the spinning workgroup over its predecessor, preventing progress and creating a circular dependency, satisfying condition 4b and resulting in deadlock.
 
-For example, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor.
-Although the multiprocessor can host both workgroups, it has only one scheduling unit, which lacks FPG\@.
-In this setup, both workgroups launch and acquire their respective work tiles.
-The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, loading the value at the memory address for the reduction of \emph{Work Tile 0} with each iteration.
-However, because the scheduler lacks FPG, it may continue scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction.
+For example, consider a scan operation requiring two workgroups on a GPU with a single multiprocessor.
+Both workgroups are co-resident, but the multiprocessor has only one scheduling unit, which lacks FPG@.
+Both workgroups launch and acquire their respective work tiles.
+The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, repeatedly querying the memory address for the reduction of \emph{Work Tile 0}.
+However, because the scheduler lacks FPG, it continues scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction.
 As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.
@@
 \subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
-Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---its absence in key segments of hardware is a critical issue for developers.
-Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad,\thomas{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time.
-It's also unclear why the test did not trigger Metal's timeout device recovery.} and at worst, risks the aforementioned TDR and subsequent program crash.
-
+The absence of FPG in key hardware segments poses a critical challenge for developers.
+Our testing shows that, at best, running \emph{Chained-Scan} without FPG results in severe performance degradation, and at worst, risks TDR and program crashes.
 \begin{figure}
@@
 \end{figure}
-
-Exacerbating this issue is the fact that no major graphics API currently provides developers with a mechanism to query FPG support.
-This lack of transparency forces developers to design algorithms without clear knowledge of whether critical guarantees are available on the target hardware.
-In environments like shading languages, where cross-platform portability is required, this uncertainty compels developers to revert to older, less efficient strategies such as \emph{Reduce-Then-Scan}.
-Yet, in the absence of guaranteed forward progress, developers have little choice but to sacrifice performance for broader compatibility.
+Exacerbating this issue, no major graphics API provides a mechanism to query FPG support.
+This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware.
+In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-Then-Scan}, sacrificing performance for broader compatibility.
 
 \subsection{Earlier Attempts at Portability}
-Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress requirements in \emph{ScalarFallback}.
+Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{Decoupled-Lookback} approach without FPG requirements in \emph{Scalar Fallback}.
 Each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time.
-By enabling workgroups to operate on data from other tiles, \emph{ScalarFallback} prevents deadlocking by targeting the \emph{mutual exclusion} condition described by Coffman et al.
-While this may involve performing redundant work when a tile is late rather than genuinely blocking, it ensures the algorithm's termination by allowing any workgroup to complete the reduction of a blocking tile, thereby breaking inter-workgroup dependencies.
+This eliminates the \emph{mutual exclusion} condition necessary for deadlock, as workgroups can operate on data from other tiles.
+While this may introduce redundant work when a tile is late rather than truly blocking, it guarantees termination by allowing any workgroup to complete a blocking tile’s reduction, breaking inter-workgroup dependencies.
 
-\emph{ScalarFallback} suffers from two primary issues that impact its efficiency.
-First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup engages in a waiting spin.
-This effectively takes $t$ spins to complete the fallback of a blocking $t$-sized tile.
-Moreover, because only a single thread is responsible for loading elements, the cache line associated with the loaded elements is severely underutilized, further degrading performance.
-Second, once the fallback is complete, the workgroup does not post its result to device memory.
-Consequently, every workgroup deadlocked by that tile must redundantly execute the fallback, wasting computational resources.
-While \emph{ScalarFallback} was able to achieve speed of light on hardware supporting FPG and executed correctly on hardware without FPG, its performance on the latter was inferior to that of \emph{Reduce-Then-Scan}.
+\emph{Scalar Fallback} suffers from two primary inefficiencies.
+First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup enters a waiting spin.
+This results in a $t$-sized blocking tile requiring $t$ spins to complete.
+Additionally, since only one thread loads elements, the associated cache line is severely underutilized.
+Second, once the fallback completes, the workgroup does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback.
+While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce then Scan}.
 
-\section{Decoupled-Fallback}
-\emph{Decoupled-Fallback} separates fallback operations from the lookback process, enhancing the \emph{Scalar-Fallback} technique with significantly improved fallback performance.
+\section{Decoupled Fallback}
+\emph{Decoupled Fallback} separates fallback operations from the lookback process, enhancing the \emph{Scalar-Fallback} technique with significantly improved fallback performance.
 As an extension of the \emph{Chained-Scan} architecture, it retains the advantages of high speed and oversubscription, aligning with our design requirements.
 
-Like \emph{Scalar-Fallback}, we choose \emph{mutual exclusion} as our approach to break deadlocking.
+Like \emph{Scalar-Fallback}, we choose \emph{mutual exclusion} as our approach to preclude deadlocks.
 \emph{Dependency relationships} and \emph{no preemption} are immutable characteristics of scan and the GPU programming model, eliminating them as viable avenues for breaking deadlocking.
 While it may be possible to break \emph{circular ordering} by designing tile sizes such that workgroups are mapped one-to-one to multiprocessors, doing so portably is infeasible.
-Moreover, vendor guidance on progress guarantees among multiprocessors is also lacking, leaving the same risks of deadlocking but at the multiprocessor level.
-While eliminating \emph{resource holding}---allowing workgroups to discard work tiles when they encounter a deadlock---is theoretically possible, it is strictly worse than eliminating \emph{mutual exclusion}, as it necessitates $t$ additional movements (discard-read-read versus read).
+Moreover, vendor guidance on progress guarantees among multiprocessors is also lacking, shifting rather than eliminating deadlock risks.
+Removing \emph{resource holding}---allowing workgroups to discard \emph{blocked} work tiles---is theoretically possible but strictly inferior to removing \emph{mutual exclusion}, as it incurs $t$ extra memory operations (discard-read-read instead of read).
 \begin{algorithm}
@@
 
-  \tcp{Decoupled Fallback}
+  \tcp{Chained Propagation, Decoupled Fallback}
   \While{$\textit{wg\_control} = \textnormal{LOCKED}$}{
@@
 To optimize the fallback operation, we employ three strategies.
-Although a lookback is performed by a single thread or subgroup, we want the entire workgroup to participate in the fallback.
-To achieve this, our first strategy is the use of a workgroup-wide \emph{control flag}, which enables threads performing the lookback to signal the entire workgroup when a fallback is necessary.
-Second, we explicitly limit the number of spins a workgroup can perform while waiting on a tile.
-If this maximum spin count is exceeded, the fallback operation is triggered.
-While this approach still risks performing redundant fallbacks---since a late tile is indistinguishable from one which is genuinely blocking---exposing the maximum spin count as a tunable parameter helps minimize false-positive fallbacks.
-Finally, after completing a fallback, the workgroup attempts to insert the reduction in place of the blocking tile, preventing subsequent workgroups from initiating redundant fallbacks.
+Although \emph{Chained-Propagation} is performed by a single thread or subgroup, we want the entire workgroup to participate in the fallback.
+First, a shared memory \emph{control flag} enables threads performing the lookback to signal the entire workgroup when a fallback is necessary.
+Second, we limit the number of spins a thread can perform while waiting on a tile; exceeding this threshold triggers a fallback.
+While this approach risks redundant fallbacks---since a late tile is indistinguishable from a genuinely blocking one---exposing the maximum spin count as a tunable parameter helps minimize false positives.
+Finally, after completing a fallback, the workgroup attempts to insert the reduction in place of the blocking tile, preventing redundant fallbacks by subsequent workgroups.
 
 The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled-Lookback}.
-No subgroup scheduler is perfectly fair, and thus any given hardware will have a natural fallback rate $f$, regardless of FPG, due to \emph{false-positive} fallback operations.
-Consequently, we expect memory bandwidth traffic to increase (and throughput to decrease) by $O(fn)$ to $O((2 + f)n)$.
-However, just as serial ordering guarantees that memory accesses to the spine fall within an $o$-width sliding window, memory accesses to the scan input are similarly guaranteed to fall within an $ot$-width sliding window, making them also highly likely to reside within some level of the cache.
-As we later empirically demonstrate, this cache residency significantly reduces the cost of memory access, such that the fallback operation is almost always bound by the compute required to reduce the blocking tile.
-As a pure reduction incurs very lower overhead, we observe throughput approaching $O(2n)$ on most devices.
+Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$, independent of FPG, due to \emph{false-positive} fallback operations.
+As a result, memory bandwidth usage increases (and throughput decreases) from $O(fn)$ to $O((2 + f)n)$.
+However, as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache.
+As we later empirically demonstrate, this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile.
+Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.
 
@@
   \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
-  \item \textbf{Decoupled-Fallback}: The first subgroup in the workgroup performs a lookback operation, engaging in fallback operations if necessary.
+  \item \textbf{Chained Propagation, Decoupled Fallback}: The first subgroup in the workgroup performs a lookback operation, engaging in fallback operations if necessary.
 Once completed, it posts the results to shared memory for the workgroup to consume.
@@
 Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase is negligible due to our overall thriftiness, including the deliberate retention of vectors in registers and the efficiency of subgroup raking.
-Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; zero bank conflicts across all supported subgroup sizes.\footnote{Since we also require a workgroup-wide pure reduce operation, this can be achieved by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}
+Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; zero bank conflicts across all supported subgroup sizes.\footnote{Since we also require a workgroup-wide pure reduce operation, we obtain this by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}
 
 \subsection{Decoupled-Fallback}
-To reiterate, the goal of the lookback is twofold: to determine the reduction of all preceding work tiles and to relay those values to succeeding tiles to accelerate their own lookback operations.
+To reiterate, the goal of \emph{Chained Propagation} is twofold: first, to determine the reduction of all preceding work tiles, and second, to relay those values to succeeding tiles to accelerate their own lookback operations.
 Following the \emph{Decoupled-Lookback} technique, we model the inter-workgroup spine as a finite-state machine, where each work tile comprises a reduction value and a state, stored in global memory.
@@
 \end{itemize}
-Each workgroup is responsible for updating its own state, only opportunistically updating other tiles' states in the event of a fallback.
-A tile can only transition forward---Not Ready \rightarrow Ready \rightarrow Inclusive—and once advanced, may never revert to a prior state.
-Before kernel launch, all tiles in the spine are initialized to \emph{Not Ready}, and their reduction values are set to $\epsilon$.
-Workgroups operate independently but remain dependent on preceding tiles to finalize their own inclusive reductions.
-As they transition through states, workgroups progressively accumulate reductions from prior tiles and, once their lookback phase is complete, post their inclusive reduction for the next tile to consume.
-\john{It's helpful, I think, to also explicitly say what data is stored globally as the spine (I think, for each workgroup, one data element and one state).}
-\john{Maybe a sentence that describes the state progression at a high level?
-e.g., ``the state of each workgroup begins at INVALID, moves to READY, then to INCLUSIVE, the state never moves backwards, each workgroup can operate independently and change its own state and also, during lookback, the state of preceding workgroups'', that sort of thing.}
-
+Workgroups operate independently, each responsible for updating its own tile state but may opportunistically update preceding tiles in the event of a fallback.
+Tile states transition strictly forward---Not Ready \rightarrow Ready \rightarrow Inclusive---and never revert.
+Before kernel launch, all tiles are initialized to \emph{Not Ready} with reduction values set to $\epsilon$.
+State transitions dictate reduction accumulation: at \emph{Ready}, local reductions are posted; at \emph{Inclusive}, preceding values are incorporated.
 To coordinate this process, each workgroup initializes a shared memory \emph{control flag} during partition index acquisition.
 This flag governs both the initiation of fallback operations and the entry and exit of the entire \emph{Decoupled-Fallback} phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.
-The details of tile state encoding and atomic updates are discussed in the following section and may vary depending on the data type, so for simplicity, we describe the lookback here as being performed by a single thread.
 
-\john{Also, we present the algorithm from the perspective of one workgroup, which at all times has a pointer to a ``current tile'' on which it is working. (Sloppily worded, sorry.)}
-\john{I might also say something like ``in the following discussion, we assume the `add' monoid and use terms like `add' and `sum', but the algorithm works with any monoid''.
-That gets rid of awkward terms like ``accumulate''.
-I don't feel strongly about this.}
-\thomas{I'm going to use ``ingest'' to match the flow chart. I literally used a thesaurus to choose accumulate . . . Can still change later.}
-
-We now detail the \emph{Decoupled-Fallback} procedure, presenting the algorithm from the perspective of a single workgroup:
+We now detail the \emph{Decoupled-Fallback} procedure, presenting the algorithm from the perspective of a single workgroup.
+The details of tile state encoding and atomic updates are discussed in the following section and may vary depending on the data type, so for simplicity, we describe the operation as being performed by the \emph{first thread} of the workgroup:
 \begin{figure}
@@
   \includegraphics[width=\linewidth]{graphics/FlowChart.pdf}
-  \caption{Decoupled-Fallback}
   \label{fig:decoupled_fallback}
+  \caption{Flow Chart Representation of Decoupled Fallback}
 \end{figure}
@@
           \item If it is the first work tile, its local reduction \emph{is} its inclusive reduction.
-In this case, the workgroup updates its tile state directly to \emph{Inclusive} and bypasses the lookback entirely.
+In this case, the workgroup updates its tile state directly to \emph{Inclusive} and bypasses the \emph{Chained-Propagation} phase entirely.
         \end{itemize}
@@
         \begin{itemize}
-          \item The control flag, initialized to \emph{Locked}, signals that the workgroup should enter the lookback phase.
-          \item The lookback thread initializes a \emph{previous reduction} variable in registers with $\epsilon$ to hold the accumulated reductions of preceding tiles.
-          \item To traverse the spine, the lookback thread initializes a \emph{current tile} pointer to its immediate predecessor.  
+          \item The control flag, initialized to \emph{Locked}, signals that the workgroup should enter the \emph{Chained-Propagation} phase.
+          \item The first thread initializes a \emph{previous reduction} variable in registers with $\epsilon$ to hold the accumulated reductions of preceding tiles.
+          \item To traverse the spine, the first thread initializes a \emph{current tile} pointer to its immediate predecessor.  
         \end{itemize}
@@
         \begin{itemize}
-          \item The lookback thread queries the state of the \emph{current tile}.
+          \item The first thread queries the state of \emph{current tile}.
+                \begin{itemize}
                   \item If the tile is in the \emph{Ready} state, the workgroup ingests the value and sets \emph{current tile} to the preceding tile, repeating this step until reaching a tile that is \emph{Not Ready}.
                   \item If the tile is in the \emph{Inclusive} state, the workgroup ingests the value and proceeds directly to step 4.
-          \item If the tile is in the \emph{Not Ready} state, the lookback thread spins for a limited number of iterations, defined by the \emph{Maximum Spin Count}.
-If this count is exceeded, the thread exits the lookback loop, leaving the control flag \emph{Locked}, signaling to the workgroup that a fallback is required.
-To prepare for the fallback, it stores a copy of the \emph{current tile} in shared memory for the workgroup to process.
-Without FPG, \emph{Decoupled-Lookback} risks spinning indefinitely and deadlocking at this point.
-
-          \john{I don't get ``signals''. To whom are we signaling? What is signaled?}
-          \john{I think this is the place where decoupled-lookback fails without FPG\@.
-If so, say so, and why it fails.
-It underscores the central message of the paper, and is thus worth reiterating.}
+                  \item If the tile is in the \emph{Not Ready} state, the first thread spins for a limited number of iterations, defined by the \emph{Maximum Spin Count}. 
+                        \begin{itemize}
+                            \item If this count is exceeded, it exits the lookback loop, leaving the control flag \emph{Locked} to signal to the workgroup that a fallback is required.
+Additionally, it copies \emph{current tile} into shared memory to broadcast the fallback location to the workgroup.
+                            \item Else, it continues spinning until the tile transitions to a \emph{Ready} or \emph{Inclusive} state.
+Without FPG, \emph{Decoupled Lookback} risks spinning indefinitely and deadlocking at this stage.
+                        \end{itemize}
+                \end{itemize}
         \end{itemize}
 
-  \item \textbf{Fallback:} \john{Start with a sentence that describes \emph{why} we've reached this phase.
-``Even after waiting, we see that a previous workgroup has not completed the reduction of its tile.'', that sort of thing.}
+  \item \textbf{Fallback:} 
         \begin{itemize}
-          \item The workgroup computes the reduction of the blocking tile by processing the corresponding elements directly.
-This process involves a workgroup-wide reduction using additional shared memory to aggregate partial subgroup results.
+          \item If the blocking tile is still \emph{Not Ready} after waiting, the workgroup initiates a fallback operation.
+          \item Using the tile index provided by the first thread, the full workgroup computes the reduction of the blocking tile by directly processing the corresponding elements.
+This process involves a \emph{workgroup-wide reduction},\footnote{See footnote NUMBER.} utilizing additional shared memory to aggregate partial subgroup results.
           \item Once the fallback reduction is complete, the workgroup \emph{attempts} to post the reduction to global memory and update the tile's state to \emph{Ready}.
-          \item As this update attempt is performed atomically, the workgroup reads back the blocking tile's immediate state: \john{``immediate''? instead, ``most recent''?}
+          \item As this update attempt is performed atomically, the workgroup reads back the blocking tile's most recent state:
                 \begin{itemize}
-                  \item If the state is \emph{Inclusive}, the fallback value calculated by our workgroup is discarded, and our workgroup accumulates the value then proceeds to step 4.
-                  \item If the blocking tile is the first in the spine, the workgroup has reached the start of the spine and proceeds to step 4.
-                  \item Otherwise, the workgroup resumes the lookback, starting with the preceding tile.
-\john{\ldots\ by setting the current tile to \ldots\ I'm trying to not be sloppy with ``starting with'' given that we have a concept of ``current tile''.}
+                  \item If the state is \emph{Inclusive}, the fallback value computed by the workgroup is discarded, and the first thread ingests the returned value before proceeding to step 4.
+                  \item Else, the first thread ingests the fallback value.
+                        \begin{itemize}
+                            \item If the blocking tile is the first in the spine, the workgroup has reached the start of the spine and thus completed the lookback. The first thread proceeds to step 4.
+                            \item Else, the first thread updates \emph{current tile} to the preceding tile, continuing the lookback process at step 2.
+                        \end{itemize}
                 \end{itemize}
@@
         \begin{itemize}
-          \item Upon completing the lookback or fallback, the local reduction is added to the \emph{previous reduction} to compute the inclusive reduction, which is then posted to global memory, transitioning the tile state to \emph{Inclusive}.
-          \item The \emph{previous reduction} is posted to workgroup shared memory for the workgroup to consume and propagate as necessary.
-\john{``consume and propagate'' is super non-specific.
-What is happening here and how is it different than the next sentence?}
-          \item The control flag is updated to \emph{Unlocked} to signal the exit of the \emph{Decoupled-Fallback} phase, allowing subsequent operations to proceed.
-
-          \john{Be specific about what happens here!
-I'm presuming part of this item's work + the previous item's work is ``use the previous reduction as the starting point for scanning then outputting our current tile''.
-``subsequent operations'' is super non-specific.}
+          \item To prioritize reduction propagation, the first thread immediately adds the local reduction to \emph{previous reduction}, posts the inclusive reduction to global memory, and transitions the tile state to \emph{Inclusive}.
+          \item It then writes \emph{previous reduction} to shared memory, allowing all threads to apply it in their \emph{workgroup-wide} and local scan computations.
+          \item Finally, the first thread sets the control flag to \emph{Unlocked}, signaling the rest of the workgroup that \emph{Chained Propagation} has completed, and allowing the final write phase to begin.
+        \end{itemize}
+\end{enumerate}
 
@@
 Does the flag-unset have any ordering constraints in terms of correctness?}
-        \end{itemize}
-\end{enumerate}
+\thomas{Last phase may still need revision in accordance with above feedback. . . I have deliberately tried to avoid mentioning barriers to keep the section succinct.}
 
 \subsection{Tile State Implementation}
-Because a tile may be in three possible states, two bits are required to record its state alongside the 32 bits needed to store the tile's reduction value.
-Previous \emph{Chained-Scan} architectures maintained a coherent view of both by using either 64-bit atomics with bit packing or memory fences with 32-bit atomics.
-However, neither approach is viable for our implementation, as WebGPU lacks a memory model and does not support 64-bit atomics.
-Additionally, 64-bit atomics are unsupported on most mobile devices, making their future adoption impractical for our needs.
-Instead, we use two 32-bit values, splitting the reduction into upper and lower 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state.
-\begin{figure}
+Since workgroups must observe updates from one another, but WebGPU lacks a \emph{volatile}-like qualifier, we rely on atomics to ensure visibility and consistency across workgroups.
+Because a tile may be in three possible states, two bits are required to record its state alongside the bits needed to store the tile's reduction value.
+Previous \emph{Chained-Scan} architectures relied on 64-bit atomics and/or memory fences to ensure a consistency between state and value, but WebGPU’s lack of a memory model and 64-bit atomic support makes both infeasible.
+Instead, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state.
+\begin{figure}[h!]
   \centering
@@
 \end{figure}
-To accommodate the split values, we use a single subgroup for the lookback, leveraging subgroup \emph{ballot} to verify that the split value state is consistent across threads, and subgroup \emph{shuffle} to join and split reductions as needed.
-Since this method requires one thread per 16 bits, and the minimum subgroup size in the WebGPU specification is 4, it supports both 32-bit and 64-bit values.
-Notably, this extends support to 64-bit scans, which would not be possible even with 64-bit atomics.\footnote{Although native 64-bit values are not currently supported by the WebGPU specification, their future inclusion is anticipated.}
 
-In the context of fallbacks, we require a tile state representation and an accompanying atomic operation that ensures a fallback update attempt can safely transition a tile's state to \emph{Ready} without overwriting a more advanced state.
-When an update attempt is made, the blocking tile can exist in any of the three states due to two scenarios: a false-positive fallback may have been initiated on a tile that transitions to \emph{Ready} or \emph{Inclusive} before the fallback completes, or another workgroup may have already successfully completed a fallback update, leaving the tile in \emph{Ready}.
-Consequently, the update mechanism should only overwrite tiles in the \emph{Not Ready} state.
-This requirement is further constrained by the fact that in WebGPU, the correctness of atomic \emph{Compare-and-Swap} is not guaranteed across all devices, prohibiting its use.
-To address this, we encode the tile state within the most significant bits of the split value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}.
-This design enables the use of atomic \emph{Maximum}, which inherently respects the state hierarchy, to safely perform updates.
+To accommodate the split values during \emph{Chained Propagation}, we use two or four threads from the first subgroup, leveraging subgroup \emph{ballot} to verify state consistency, and subgroup \emph{shuffle} to join and split reductions.
+With one thread per 16 bits and a minimum subgroup size of four in WebGPU, this method supports both 32-bit and 64-bit values, enabling 64-bit scans without 64-bit atomics or fences.\footnote{Although native 64-bit values are not currently supported by the WebGPU specification, their future inclusion is anticipated.}
+
+Fallbacks require a tile state representation and an atomic operation that can safely transition a tile to \emph{Ready} without overwriting a more advanced state.
+When an update attempt is made, a blocking tile may be in any of the three states due to two scenarios: a false-positive fallback on a tile that transitions to \emph{Ready} or \emph{Inclusive} before fallback completion; or another workgroup completes its fallback first, leaving the tile in \emph{Ready}.
+WebGPU does not guarantee the correctness of atomic \emph{Compare-and-Swap} across all devices, prohibiting its use.
+Instead, we encode the tile state in the most significant bits of the split value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}.
+This allows atomic \emph{Maximum}, which inherently preserves state hierarchy, to safely perform updates.
 
@@
 
-Since WebGPU limits device-side timestamps to compute or render passes, we implement a custom Memcopy kernel for timing purposes.
-Similarly, as there are no existing WebGPU implementations of RTS, we rely on our own optimized version for comparison.
-To ensure a competitive evaluation, our RTS implementation employs a variable-launch main kernel with a single workgroup performing a serial spine scan, as we found raking to be less performant in practice.
-Apart from this adjustment, the intraworkgroup implementation in RTS is identical to that used in \emph{Decoupled-Fallback}.
-
 \thomas{Performance is good}
@@
 
-\subsection{Simulated Blocking}
+\subsection{Emulated Blocking}
 To demonstrate the resilience of our technique under fallback rates significantly higher than those observed organically on devices, we intentionally force fallbacks by omitting the posting of work tile reductions.
@@
 
-\section{Cut material, can be readded if we have enough space}
-
-To ensure consistency with our coding artifact and promote cross-vendor compatibility, we use the terminology of the WebGPU specification.
-\setlength{\tabcolsep}{3pt}
-\begin{table}[h]
-  \centering
-  \caption{Non-Exhaustive Equivalent Terminology}
-  \label{tab:terminology}
-  \begin{tabular}{lcccc}
-    \toprule
-    \textbf{WGSL} & \textbf{GLSL} & \textbf{HLSL} & \textbf{Metal} & \textbf{CUDA} \\ \midrule
-    Subgroup      & Subgroup      & Wave          & SIMD-group     & Warp          \\
-    Workgroup     & Workgroup     & Group         & Threadgroup    & Block         \\
-    Workgroup     & Shared        & Groupshared   & Threadgroup    & Shared        \\ \bottomrule
-  \end{tabular}
-\end{table}
-
-Despite its limited suitability within the oversubscription paradigm, inter-workgroup synchronization holds significant performance potential, as many GPU-friendly algorithms—such as \emph{example}, \emph{example}, and \emph{scan}---can benefit from a trade-off that eliminates the costly overhead of kernel-launch synchronization.
-This potential was explored in prior work~\cite{}, culminating in Gupta et al.’s~\cite{} \emph{Persistent Thread} paradigm.
-Rather than oversubscribing processors by launching $b$ workgroups, each consuming a single \emph{work tile}, this approach \emph{discovers} the total workgroup occupancy across the device $o$---referred to by Gupta et al.
-as a \emph{maximal launch}---and launches $o$ workgroups, each responsible for processing $b/o$ work tiles.
-Limiting the launch size allows the workgroups to always remain on chip, eliding context switching, and thus facilitating an efficient inter-workgroup barrier.
-
-\subsection{Work Efficiency and Depth}
-\subsubsection{Size Agnostic Scan}
-For a subgroup of size $s$, the Kogge-Stone network has a work complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
-\begin{equation}
-  \begin{aligned}[b]
-    \text{Work} & = \underbrace{\vphantom{\sum_{k=0}^{\lceil \log_s n \rceil}}n}_{\text{fanout}}
-    +
-    \underbrace{\sum_{k=1}^{\lceil \log_s n \rceil}}_{\text{iterations}}
-    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}\frac{n}{s^k}}_{\text{calls per iteration}}
-    \cdot
-    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}s \log_2 s}_{\text{work per call}} \\
-                & = n + s \log_2 s \cdot \sum_{k=1}^{\lceil \log_s n \rceil} \frac{n}{s^k}       \\
-                & = O\left(n + n \log_2 s\right)                                                 \\
-                & = O(n \log_2 s).
-  \end{aligned}
-\end{equation}
-Since the subgroup size remains constant within each kernel launch, the final work complexity is $O(n)$.
-Thus our implementation preserves asymptotically optimal work, even though the structure is not isomorphic to the radix-two network.
-The Kogge-Stone scan has a depth of $\log_2 s$ Thus, the total depth of the scan is:
-\begin{equation}
-  \begin{aligned}[b]
-    \text{Depth} & = \underbrace{\lceil \log_s n \rceil}_{\text{loop iterations}}
-    \cdot \underbrace{\log_2 s}_{\text{depth per iteration}}                           \\
-                 & = \left\lceil \frac{\log_2 n}{\log_2 s} \right\rceil \cdot \log_2 s \\
-                 & = \log_2 n + O(1).
-  \end{aligned}
-\end{equation}
-While minimal depth is preserved regardless of the subgroup size, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a workgroup-wide barrier at each iteration.
-Consequently, while the theoretical depth remains constant across subgroup sizes, smaller subgroups tend to perform worse in practice due to the increased relative cost of synchronization and the overhead of additional iterations in the main loop.
 \begin{acks}
@@
 
+\clearpage
 \appendix
@@
 We provide our artifact in the following GitHub repository:
-\thomas{Will anonymize with GithubAnonymous once we are ready.}
-Due to the aforementioned issues with WGSL transpilers, we choose Dawn as our WebGPU implementation.
+\thomas{Will anonymize with GithubAnonymous once we are ready.}.
+
+\subsection{API Choice}
+As rust is significantly more easy to build across devices we chose the WGPU api.
+
+\subsection{Device Choices}
+We chose a diverse range of devices to illustrate
+
+\subsection{Baseline Comparisons}
+Since WebGPU limits device-side timestamps to compute or render passes, we implement a custom Memcopy kernel for timing purposes.
+Similarly, as there are no existing WebGPU implementations of RTS, we rely on our own optimized version for comparison.
+To ensure a competitive evaluation, our RTS implementation employs a variable-launch main kernel with a single workgroup performing a serial spine scan, as we found raking to be less performant in practice.
+Apart from this adjustment, the intraworkgroup implementation in RTS is identical to that used in \emph{Decoupled-Fallback}.
 

### 2025-02-03 artifact/src/main.rs
Switch to Naga

@@ -333,59 +333,47 @@ struct Shaders {
 
 impl Shaders {
     fn init(gpu: &GPUContext, gpu_buffers: &GPUBuffers) -> Self {
-        let init_mod;
-        let valid_mod;
-        let reduce_mod;
-        let spine_mod;
-        let downsweep_mod;
-        let csdl_mod;
-        let csdldf_mod;
-        let csdldf_emu_mod;
-        let csdldf_occ_mod;
-        let memcpy_mod;
 
-        unsafe {
-            init_mod = gpu
-                .device
-                .create_shader_module_spirv(&wgpu::include_spirv_raw!("../SPIR-V/init.main.spv"));
-            valid_mod = gpu
-                .device
-                .create_shader_module_spirv(&wgpu::include_spirv_raw!(
-                    "../SPIR-V/validate.main.spv"
-                ));
-            reduce_mod = gpu
-                .device
-                .create_shader_module_spirv(&wgpu::include_spirv_raw!("../SPIR-V/rts.reduce.spv"));
-            spine_mod = gpu
-                .device
-                .create_shader_module_spirv(&wgpu::include_spirv_raw!(
-                    "../SPIR-V/rts.spine_scan.spv"
-                ));
-            downsweep_mod = gpu
-                .device
-                .create_shader_module_spirv(&wgpu::include_spirv_raw!(
-                    "../SPIR-V/rts.downsweep.spv"
-                ));
-            csdl_mod = gpu
-                .device
-                .create_shader_module_spirv(&wgpu::include_spirv_raw!("../SPIR-V/csdl.main.spv"));
-            csdldf_mod = gpu
-                .device
-                .create_shader_module_spirv(&wgpu::include_spirv_raw!("../SPIR-V/csdldf.main.spv"));
-            csdldf_emu_mod = gpu
-                .device
-                .create_shader_module_spirv(&wgpu::include_spirv_raw!(
-                    "../SPIR-V/csdldf_emulate.main.spv"
-                ));
-            csdldf_occ_mod = gpu
-                .device
-                .create_shader_module_spirv(&wgpu::include_spirv_raw!(
-                    "../SPIR-V/csdldf_occ.main.spv"
-                ));
-            memcpy_mod = gpu
-                .device
[…]

### 2025-02-03 artifact/src/main.rs
Forgot to remove Passthrough!

@@ -47,8 +47,7 @@ impl GPUContext {
                 &wgpu::DeviceDescriptor {
                     label: None,
                     required_features: wgpu::Features::TIMESTAMP_QUERY
-                        | wgpu::Features::SUBGROUP
-                        | wgpu::Features::SPIRV_SHADER_PASSTHROUGH,
+                        | wgpu::Features::SUBGROUP,
                     required_limits: wgpu::Limits::default(),
                     memory_hints: wgpu::MemoryHints::Performance,
                 },

### 2025-02-04 paper.tex
fixing compile, rightarrow needs to be in math mode
#7 ok done with this read, i like it, several comments anyway of course

@@
 The GPU programming model virtualizes processors to ensure GPU programs---\emph{kernels}---remain portable across hardware with differing physical resources.
-Effective virtualization requires \emph{oversubcription}, a programming paradigm where tasks are (over)partitioned in such a way that kernels request more virtual processors than are physically available.
+Effective virtualization requires \emph{oversubscription}, a programming paradigm where tasks are (over)partitioned in such a way that kernels request more virtual processors than are physically available.
 As a result, a single multiprocessor may host multiple workgroups simultaneously.
@@
 However, the term "fairness" is used differently across the literature, leading to potential confusion.
-Sorensen et al.\cite{sorensen2016,sorensen2018,sorensen2021}, whose work provides the most comprehensive taxonomy of progress models to date, use "fairness" specifically to describe differing levels of progress guarantees.
+Sorensen et al.~\cite{sorensen2016,sorensen2018,sorensen2021}, whose work provides the most comprehensive taxonomy of progress models to date, use "fairness" specifically to describe differing levels of progress guarantees.
 In contrast, NVIDIA publications~\cite{4523358,Merrill2016} define "fairness" in terms of the distribution of execution resources among subgroups.
@@
   Sorensen calls FPG "fairness:" "We have described two schedulers: fair and unfair, under which starvation-freedom for blocking idioms is either always or never guaranteed."\newline
-  Nvidia, by contrast uses fairness in line with our language. From Tesla architecture: \newline
+  NVIDIA, by contrast uses fairness in line with our language. From Tesla architecture: \newline
   "A scoreboard qualifies each warp for issue each cycle.
@@
 The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \).
-Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}.
+To allow parallelization, \( \oplus \) must be associative, but it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}.
 In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence.
@@
 Snir~\cite{10.1016/0196-67748690003-9} proved that depth $d$ and size $s$ are related by $s + d \ge 2n - 2$, and Fich~\cite{10.1145/800061.808738} proved that among minimum depth scans Ladner-Fischer~\cite{10.1145/322217.322232} networks have optimal size.
-Blelloch~\cite{} adapted scan to the PRAM model and popularized its use as an algorithmic primitive.
+Blelloch~\cite{Blelloch:1989:SAP} adapted scan to the PRAM model and popularized its use as an algorithmic primitive.
 Finally, Merrill and Garland~\cite{Merrill2016} provide a review of GPU scan implementations, complementing earlier work by Merrill and Grimshaw~\cite{Merrill2009}.
@@
 
-Thus, oversubcription and \emph{Chained-Scan} emerges as the preferred architecture, as it naturally aligns with portability by leveraging hardware schedulers to dynamically manage occupancy across diverse architectures without requiring developer intervention.
+Thus, oversubscription and \emph{Chained-Scan} emerges as the preferred architecture, as it naturally aligns with portability by leveraging hardware schedulers to dynamically manage occupancy across diverse architectures without requiring developer intervention.
 Furthermore, as a single kernel with inherently serial ordering, \emph{Chained-Scan} is suitable for \emph{in-situ} operations, augmenting practical routines such as data deduplication, sparse data compaction, error correction, and interval merging.
@@
 \thomas{Sorensen places Chained-Scan into the HSA category, but it is unclear if this is correct.
-The definition of HSA is \newline "Threads for which fair execution is guaranteed: The thread with the lowest id that has not yet terminated" \newline However, as Chained-Scan and other serially ordered algorithms no longer use the virtual ID provided by the workgroup scheduler, is fair execution still guaranteed under HSA?
-It seems that it should be classified under OBE\@.}
+The definition of HSA is \newline ``Threads for which fair execution is guaranteed: The thread with the lowest id that has not yet terminated'' \newline However, as Chained-Scan and other serially ordered algorithms no longer use the virtual ID provided by the workgroup scheduler, is fair execution still guaranteed under HSA?
+It seems that it should be classified under OBE\@.} \john{I don't have an opinion on this, you're familiar with this and I'm not, but this seems like a good footnote.
+Or we can ask Tyler.}
 
@@
 In such cases, the dependent workgroup spins, consuming execution resources while waiting.
-Because workgroups may share execution resources by being co-resident on the same multiprocessor, without FPG, the scheduler may indefinitely prioritize the spinning workgroup over its predecessor, preventing progress and creating a circular dependency, satisfying condition 4b and resulting in deadlock.
+Because workgroups may share execution resources by being co-resident on the same multiprocessor\john{Question, and maybe this is worth a footnote.
+What if we guaranteed that no multiprocessor ever had more than one workgroup scheduled to it, say by always allocating the maximum amount of workgroup memory?
+Of course there would be performance issues.
+But does that solve the problem?
+Or do we descend into an analogous issue with subgroup scheduling?}, without FPG, the scheduler may indefinitely prioritize the spinning workgroup over its predecessor, preventing progress and creating a circular dependency, satisfying condition 4b and resulting in deadlock.
 
 For example, consider a scan operation requiring two workgroups on a GPU with a single multiprocessor.
-Both workgroups are co-resident, but the multiprocessor has only one scheduling unit, which lacks FPG@.
+Both workgroups are co-resident, but the multiprocessor has only one scheduling unit, which lacks FPG\@.
 Both workgroups launch and acquire their respective work tiles.
@@
 \subsection{Earlier Attempts at Portability}
+\john{I'd like Raph to chime in here.
+What I worry about in the way this subsection is presented is that a reviewer will say ``Oh, the fallback technique was already invented in 2021, all this paper does is makes it a little better''.
+On one hand I totally want Raph to get credit for this cool idea.
+On the other hand, writing this subsection in this way, as a citable previous publication, significantly reduces the size of the advance that this paper we're writing hopes to capture.
+If I'm being selfish and trying to make this paper look as good as possible, I downplay the previous work a little and don't present it as if it was a previous peer-reviewed publication.
+At the least I want to use the words ``blog post'' in the text here.}
 Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{Decoupled-Lookback} approach without FPG requirements in \emph{Scalar Fallback}.
@@
 Second, once the fallback completes, the workgroup does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback.
-While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce then Scan}.
+\john{If we can, we would like to [politely] say that this previous work was missing a key insight, e.g., while fallback can address the portability problem, \ldots\ Would we say that it's not decoupled in the way that the technique we present in this paper is?
+Or is the only deficiency the choice of a serial fallback?} While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce then Scan}.
 
@@
 To optimize the fallback operation, we employ three strategies.
-Although \emph{Chained-Propagation} is performed by a single thread or subgroup, we want the entire workgroup to participate in the fallback.
-First, a shared memory \emph{control flag} enables threads performing the lookback to signal the entire workgroup when a fallback is necessary.
-Second, we limit the number of spins a thread can perform while waiting on a tile; exceeding this threshold triggers a fallback.
+Although \emph{Chained-Propagation} is performed by a single thread or subgroup, for performance reasons we want the entire workgroup to participate in the fallback.
+(1) First, a shared memory \emph{control flag} enables threads performing the lookback to signal the entire workgroup when a fallback is necessary.
+(2) Next, we limit the number of spins a thread can perform while waiting on a tile; exceeding this threshold triggers a fallback.
 While this approach risks redundant fallbacks---since a late tile is indistinguishable from a genuinely blocking one---exposing the maximum spin count as a tunable parameter helps minimize false positives.
-Finally, after completing a fallback, the workgroup attempts to insert the reduction in place of the blocking tile, preventing redundant fallbacks by subsequent workgroups.
+\john{I can see it is sensible to have a maximum spin count but I do not see how the action of exposing the maximum spin count to the programmer minimizes false positives.} (3) Finally, after completing a fallback, the workgroup attempts to post the reduction result of the blocking tile, preventing redundant fallbacks by subsequent workgroups.
 
 The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled-Lookback}.
-Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$, independent of FPG, due to \emph{false-positive} fallback operations.
+Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$\!, independent of FPG, due to \emph{false-positive} fallback operations.
 As a result, memory bandwidth usage increases (and throughput decreases) from $O(fn)$ to $O((2 + f)n)$.
@@
 \section{Implementation}
-At a high level, our algorithmn follows a five phase construction:
+At a high level, our algorithm follows a five-phase construction:
 \begin{enumerate}
@@
   \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking subgroup scans,
-  allowing the subgroup-block scan to remain barrier-free and in registers.The subgroup-block reduction is then posted to shared memory.
-For a workgroup of size $w$, subgroup-raking reduces the local spine to $\frac{w}{s}$.
+  allowing the subgroup-block scan to remain barrier-free and in registers.
+The subgroup-block reduction is then posted to shared memory.
+For a workgroup of size $w$, subgroup-raking reduces the size of the local spine to $w/s$.
   \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
-  \item \textbf{Chained Propagation, Decoupled Fallback}: The first subgroup in the workgroup performs a lookback operation, engaging in fallback operations if necessary.
-Once completed, it posts the results to shared memory for the workgroup to consume.
+  \item \textbf{Chained Propagation, Decoupled Fallback}: The first subgroup in the workgroup performs a lookback operation to acquire the reduction result of all previous workgroups, engaging in fallback operations as necessary.
+Once completed, this subgroup posts the results to shared memory for the workgroup to consume.
   \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output.
@@
 However, such access is interpreted by the hardware as a \emph{broadcast} operation, resulting in no conflicts.
-Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase is negligible due to our overall thriftiness, including the deliberate retention of vectors in registers and the efficiency of subgroup raking.
-Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; zero bank conflicts across all supported subgroup sizes.\footnote{Since we also require a workgroup-wide pure reduce operation, we obtain this by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}
+Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking.
+Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.\footnote{Since we also require a workgroup-wide pure reduce operation, we obtain this by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}
 
@@
 To reiterate, the goal of \emph{Chained Propagation} is twofold: first, to determine the reduction of all preceding work tiles, and second, to relay those values to succeeding tiles to accelerate their own lookback operations.
-Following the \emph{Decoupled-Lookback} technique, we model the inter-workgroup spine as a finite-state machine, where each work tile comprises a reduction value and a state, stored in global memory.
+Following the \emph{Decoupled-Lookback} technique, we model the inter-workgroup spine as a finite-state machine, where each work tile is represented by a reduction value and a state, stored in global memory.
 Each tile may exist in one of three states:
@@
 \end{itemize}
-Workgroups operate independently, each responsible for updating its own tile state but may opportunistically update preceding tiles in the event of a fallback.
-Tile states transition strictly forward---Not Ready \rightarrow Ready \rightarrow Inclusive---and never revert.
+Workgroups operate independently; each is responsible for updating its own tile state but may also opportunistically update preceding tiles in the event of a fallback.
+Tile states transition in a strictly forward direction---Not Ready $\rightarrow$ Ready $\rightarrow$ Inclusive---and never revert.
 Before kernel launch, all tiles are initialized to \emph{Not Ready} with reduction values set to $\epsilon$.
-State transitions dictate reduction accumulation: at \emph{Ready}, local reductions are posted; at \emph{Inclusive}, preceding values are incorporated.
-To coordinate this process, each workgroup initializes a shared memory \emph{control flag} during partition index acquisition.
+State transitions imply an update of the reduction value: a transition to \emph{Ready} results in the posting of the local workgroup reduction, and to \emph{Inclusive}, the reduction of all previous workgroup reductions.
+To coordinate this process across the workgroup, each workgroup initializes a shared memory \emph{control flag} during partition index acquisition.
 This flag governs both the initiation of fallback operations and the entry and exit of the entire \emph{Decoupled-Fallback} phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.
@@
           \item The first thread initializes a \emph{previous reduction} variable in registers with $\epsilon$ to hold the accumulated reductions of preceding tiles.
-          \item To traverse the spine, the first thread initializes a \emph{current tile} pointer to its immediate predecessor.  
+          \item The first thread initializes a \emph{current tile} pointer to its immediate predecessor, with which it begins to traverse backwards through the spine.
         \end{itemize}
@@
                   \item If the tile is in the \emph{Ready} state, the workgroup ingests the value and sets \emph{current tile} to the preceding tile, repeating this step until reaching a tile that is \emph{Not Ready}.
-                  \item If the tile is in the \emph{Inclusive} state, the workgroup ingests the value and proceeds directly to step 4.
+                  \item If the tile is in the \emph{Inclusive} state, the workgroup ingests the value and proceeds directly to step~4.
                   \item If the tile is in the \emph{Not Ready} state, the first thread spins for a limited number of iterations, defined by the \emph{Maximum Spin Count}.
                         \begin{itemize}
-                            \item If this count is exceeded, it exits the lookback loop, leaving the control flag \emph{Locked} to signal to the workgroup that a fallback is required.
-Additionally, it copies \emph{current tile} into shared memory to broadcast the fallback location to the workgroup.
-                            \item Else, it continues spinning until the tile transitions to a \emph{Ready} or \emph{Inclusive} state.
+                            \item If this count is exceeded, it exits the lookback loop and proceeds to step~3, leaving the control flag \emph{Locked} to signal to the workgroup that a fallback is required.
+Additionally, it copies \emph{current tile} into shared memory to broadcast the fallback tile identity to the workgroup.
+                            \item Otherwise, it continues spinning until the tile transitions to a \emph{Ready} or \emph{Inclusive} state.
 Without FPG, \emph{Decoupled Lookback} risks spinning indefinitely and deadlocking at this stage.
@@
           \item Once the fallback reduction is complete, the workgroup \emph{attempts} to post the reduction to global memory and update the tile's state to \emph{Ready}.
-          \item As this update attempt is performed atomically, the workgroup reads back the blocking tile's most recent state:
+          \item As this update attempt is performed atomically, the workgroup simultaneously reads back the blocking tile's most recent state:
                 \begin{itemize}
-                  \item If the state is \emph{Inclusive}, the fallback value computed by the workgroup is discarded, and the first thread ingests the returned value before proceeding to step 4.
-                  \item Else, the first thread ingests the fallback value.
+                  \item If that most recent state is \emph{Inclusive}, the fallback value computed by the workgroup is discarded, and the first thread ingests the returned value before proceeding to step 4.
+                  \item Otherwise, the first thread ingests the fallback value.
                         \begin{itemize}
-                            \item If the blocking tile is the first in the spine, the workgroup has reached the start of the spine and thus completed the lookback. The first thread proceeds to step 4.
-                            \item Else, the first thread updates \emph{current tile} to the preceding tile, continuing the lookback process at step 2.
+                            \item If the blocking tile is the first in the spine, the workgroup has reached the start of the spine and has thus completed the lookback.
+The first thread proceeds to step 4.
+                            \item Otherwise, the first thread updates \emph{current tile} to the preceding tile, returning to the lookback process in step~2.
                         \end{itemize}
@@
         \begin{itemize}
-          \item To prioritize reduction propagation, the first thread immediately adds the local reduction to \emph{previous reduction}, posts the inclusive reduction to global memory, and transitions the tile state to \emph{Inclusive}.
+          \item To prioritize inter-workgroup reduction propagation, the first thread immediately adds the local reduction to \emph{previous reduction}, posts the inclusive reduction to global memory, and transitions the tile state to \emph{Inclusive}.
           \item It then writes \emph{previous reduction} to shared memory, allowing all threads to apply it in their \emph{workgroup-wide} and local scan computations.
@@
 \thomas{Last phase may still need revision in accordance with above feedback. . . I have deliberately tried to avoid mentioning barriers to keep the section succinct.}
+\john{I agree with your succinct strategy. I like that you now say reduction-propagation is the first priority, and why.}
 
@@
 Because a tile may be in three possible states, two bits are required to record its state alongside the bits needed to store the tile's reduction value.
-Previous \emph{Chained-Scan} architectures relied on 64-bit atomics and/or memory fences to ensure a consistency between state and value, but WebGPU’s lack of a memory model and 64-bit atomic support makes both infeasible.
+Previous \emph{Chained-Scan} architectures relied on 64-bit atomics and/or memory fences to ensure a consistency between state and value, but WebGPU’s lack of a memory model \john{I'm not sure I would say ``lack of a memory model''.
+It's not the (lack of) memory model that's the problem.
+It's specifically that we don't have the operations that give us the behavior that we require.
+If we had the right operations that did the right thing but didn't have a memory model, we'd be fine.
+We want to specifically call out what is missing and what that means, and do so as generically as possible (not ``we are missing this specific atomic call'' but instead ``we are missing some sort of atomic call that has the following behavior'').
+Above, we want to not only say that we don't have a volatile qualifier but also both what that means and why it would be specifically useful if we had it.
+Doing the latter two things gives valuable ammunition to WebGPU and GPU architects that want to improve the API and hardware.} and 64-bit atomic support makes both infeasible.
 Instead, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state.
+\john{I would try to add a sentence or two on why we can't do the obvious thing---having the state in one word and the reduction value in a different word---and why the split value avoids that problem.
+This is not obvious to the reader.}
 \begin{figure}[h!]
@@
 Fallbacks require a tile state representation and an atomic operation that can safely transition a tile to \emph{Ready} without overwriting a more advanced state.
-When an update attempt is made, a blocking tile may be in any of the three states due to two scenarios: a false-positive fallback on a tile that transitions to \emph{Ready} or \emph{Inclusive} before fallback completion; or another workgroup completes its fallback first, leaving the tile in \emph{Ready}.
-WebGPU does not guarantee the correctness of atomic \emph{Compare-and-Swap} across all devices, prohibiting its use.
-Instead, we encode the tile state in the most significant bits of the split value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}.
-This allows atomic \emph{Maximum}, which inherently preserves state hierarchy, to safely perform updates.
+When an update attempt is made, a blocking tile may be in any of the three states due to either of two scenarios: (1) a false-positive fallback on a tile that transitions to \emph{Ready} or \emph{Inclusive} before fallback completion, or (2) another workgroup completing its fallback first, leaving the tile in \emph{Ready}.
+WebGPU does not guarantee the correctness of atomic \emph{Compare-and-Swap} across all devices, inhibiting its use.
+\john{Is CAS the traditional/common way to implement this?
+If so, say so, and if you can, provide a cite, even if it's not a formal cite but just the name of another package that uses this strategy.} Instead, we encode the tile state in the most significant bits of the split value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}.
+This allows an atomic \emph{maximum}, which successfully updates to a more advanced state but prevents a state regression, to correctly perform updates.
 
@@
   \end{subfigure}
-  \caption{Simulated Blocking}
+  \caption{Simulated Blocking\john{Text in these plots---axis labels, etc.---needs to be larger. If it's significantly smaller than body text, which it is, it's just not legible.}}
   \label{fig:vertical_images}

### 2025-02-04 paper.tex
Finish Edits Section 2,3,4

@@
 \end{enumerate}
-
-The absence of inter-workgroup barriers necessitates separate kernel launches for dependency resolution, inherently structuring \emph{Scan-then-Propagate} into three phases.
-Sengupta et al.'s implementation further complicates this by using recursion to handle arbitrarily large inputs.
-As the size of a work tile limits the maximum number of elements processed without inter-workgroup dependency resolution, the three-phase structure must be recursively applied until the spine size fits within a single work tile.
+The absence of inter-workgroup barriers necessitates separate kernel launches for dependency resolution, naturally enforcing this structure.
+To accommodate arbitrarily large inputs, Sengupta et al.
+employ recursion, repeatedly applying the three-phase process until the spine fits within a single work tile.
 Thus, an input of size $n$ and a work tile of size $t$ results in a recursive depth of $\lceil \log_t n \rceil$, and $2\cdot\lceil \log_t n \rceil - 1$ kernel launches.
@@
 
-Merrill and Grimshaw~\cite{Merrill2009} introduced \emph{workgroup raking} to further reduce recursion overhead when the recursion depth exceeds two.
-Instead of recursively reducing the spine, the input is partitioned into $o$ groups, where $o$ is the total workgroup occupancy.
-Each workgroup sequentially processes its assigned partition, reducing the spine size to a constant and guaranteeing a recursion depth of at most two.
-This reduces global memory movement for the spine scan from $O(n/t)$ to $O(2o)$ and limits kernel launches to three, reducing total global memory movement to $O(3n + 4o) = O(3n)$.
+Merrill and Grimshaw~\cite{Merrill2009} introduced \emph{workgroup raking} to reduce recursion overhead for arbitrarily large inputs.
+Instead of recursively reducing the spine, the input is partitioned among a fixed number of workgroups, $c$, where $c \geq o$, the total workgroup occupancy.
+This constrains the spine from a variable, potentially recursive structure to a fixed size $c$, ensuring a recursion depth of at most two.
+As a result, global memory movement for the spine scan is reduced from $O(n/t)$ to $O(2c)$, and kernel launches are limited to three, yielding a total global memory movement of $O(3n + 4c) = O(3n)$.
 
 Breitbart~\cite{10.5555/2031978.2032029} applied Gupta et al.'s \emph{Persistent-Threads}~\cite{gupta2012} to scan, eliminating kernel launch overheads.
-Recognizing that raking naturally aligns with this paradigm, Breitbart replaced kernel launches with an atomics-based inter-workgroup barrier.
-Since multiprocessors are not oversubscribed in raking, workgroups persist for the entire kernel duration, allowing synchronization at the barrier without context switching.
-Despite maintaining $O(3n)$ global memory movement, consolidating the scan into a single kernel minimizes launch overheads.
+Recognizing that raking naturally aligns with this paradigm, Breitbart's method dynamically \emph{discovers} the workgroup occupancy and launches exactly $o$ kernels.
+Since multiprocessors are not oversubscribed, workgroups persist for the entire kernel duration, enabling synchronization at an atomics-based inter-workgroup barrier without context switching.
+While global memory movement remains $O(3n)$, this consolidates the scan into a single kernel, minimizing launch overheads.
 
@@
 \subsection{Ideal Scan Architecture}
-Beyond their differences in speed, \emph{Chained-Scan} and \emph{Reduce-Then-Scan} embody two fundamentally contrasting paradigms.
-By eliding explicit inter-workgroup synchronization, \emph{Chained-Scan} aligns naturally with the oversubscription paradigm, delegating occupancy management to the hardware scheduler.
-In contrast, the raking approach inherent to \emph{Reduce-Then-Scan} mandates the fixed-launch paradigm, necessitating explicit control over occupancy and potentially synchronization.
+\emph{Chained-Scan} and \emph{Reduce-Then-Scan} represent two contrasting paradigms in scan design: one embraces oversubscription and dynamic scheduling, while the other relies on a fixed-launch model with explicit control over occupancy.
+By avoiding explicit inter-workgroup synchronization, \emph{Chained-Scan} naturally aligns with the oversubscription paradigm, allowing the hardware scheduler to dynamically manage occupancy.
+In contrast, \emph{Reduce-Then-Scan} employs workgroup raking, which mandates explicit occupancy control and, in some cases, inter-workgroup synchronization.
 
-Fixed-launch approaches demand substantial developer effort, requiring either tuning of workgroup launch sizes across target devices or occupancy discovery.
-Moreover, previous workgroup occupancy discovery techniques, such as the one proposed by Sorensen et al.~\cite{sorensen2016}, are infeasible to implement in WebGPU due to their reliance on memory fences---a feature not supported by the specification.\footnote{Our implementation includes a fenceless occupancy discovery method amenable to the WebGPU environment, but its description is beyond the scope of this work.}
+Fixed-launch approaches, such as those used in \emph{Reduce-Then-Scan}, introduce several challenges.
+They require either careful tuning of workgroup launch sizes across different architectures or an occupancy discovery mechanism, which is nontrivial to implement.
+For example, the workgroup occupancy discovery technique proposed by Sorensen et al.~\cite{sorensen2016} relies on memory fences—unsupported in WebGPU—making it infeasible in certain environments.\footnote{Our implementation includes a fenceless occupancy discovery method suitable for WebGPU, but its description is beyond the scope of this work.}
 
@@
 
-Thus, oversubscription and \emph{Chained-Scan} emerges as the preferred architecture, as it naturally aligns with portability by leveraging hardware schedulers to dynamically manage occupancy across diverse architectures without requiring developer intervention.
-Furthermore, as a single kernel with inherently serial ordering, \emph{Chained-Scan} is suitable for \emph{in-situ} operations, augmenting practical routines such as data deduplication, sparse data compaction, error correction, and interval merging.
+Given these constraints, oversubscription and \emph{Chained-Scan} emerge as the more portable and adaptable architecture.
+By leveraging the hardware scheduler, \emph{Chained-Scan} dynamically distributes work across architectures without requiring developer intervention.
+Furthermore, its single-kernel design and serial ordering make it suitable for \emph{in-situ} operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging.
 
-But, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in terms of its theoretical and implementation characteristics, it falls short in fully meeting the criterion of \emph{portability}.
-This limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA devices but notably lacking on Apple's M-series and ARM GPUs~\cite{sorensen2021}.
-Without this guarantee, \emph{Chained-Scan} is highly prone to catastrophic failure in the form of deadlocking.
+However, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in both theoretical and practical terms, it has a critical limitation: its reliance on a forward-progress guarantee.
+While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that Apple’s M-series and certain ARM GPUs do not.
+Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient—it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
 
@@
 Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$\!, independent of FPG, due to \emph{false-positive} fallback operations.
-As a result, memory bandwidth usage increases (and throughput decreases) from $O(fn)$ to $O((2 + f)n)$.
-However, as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache.
-As we later empirically demonstrate, this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile.
+As a result, memory bandwidth usage increases (and we expect throughput to decrease) by $O(fn)$ to $O((2 + f)n)$.
+However, just as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache.
+As we later empirically demonstrate(Simulated Blocking Section), this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile.
 Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.

### 2025-02-04 paper.tex
Tightening language more.

@@
 
-\thomas{A scheduler may be fair but may not have FPG (M1 Pro for example); a scheduler may have FPG but may not be fair (ARM Mali). \newline
-  Sorensen calls FPG "fairness:" "We have described two schedulers: fair and unfair, under which starvation-freedom for blocking idioms is either always or never guaranteed."\newline
-  NVIDIA, by contrast uses fairness in line with our language. From Tesla architecture: \newline
-  "A scoreboard qualifies each warp for issue each cycle.
-The instruction scheduler prioritizes all ready warps and selects the one with highest priority for issue.
-Prioritization considers warp type, instruction type, and 'fairness' to all warps executing in the SM\@." \newline
-  From DecoupledLookback: \newline
-  "Safety. The algorithm will run to completion if the system guarantees forward-progress for all processors." \newline
-  "Blocking will be minimal for systems that provide fair or nearly-fair scheduling. Fairness ensures that all processors will have recorded their aggregates in roughly the same amount of time."
-}
-
 \subsection{The Scan Primitive}\label{sec:the-scan-primitive}
@@
 \subsection{Ideal Scan Architecture}
-\emph{Chained-Scan} and \emph{Reduce-Then-Scan} represent two contrasting paradigms in scan design: one embraces oversubscription and dynamic scheduling, while the other relies on a fixed-launch model with explicit control over occupancy.
-By avoiding explicit inter-workgroup synchronization, \emph{Chained-Scan} naturally aligns with the oversubscription paradigm, allowing the hardware scheduler to dynamically manage occupancy.
+\emph{Chained Scan} and \emph{Reduce then Scan} represent two contrasting paradigms in scan design: one embraces oversubscription and dynamic scheduling, while the other relies on a fixed-launch model with explicit control over occupancy.
+By avoiding explicit inter-workgroup synchronization, \emph{Chained-Scan} naturally aligns with the oversubscription paradigm, allowing the workgroup scheduler to dynamically manage occupancy.
 In contrast, \emph{Reduce-Then-Scan} employs workgroup raking, which mandates explicit occupancy control and, in some cases, inter-workgroup synchronization.
@@
 
-Given these constraints, oversubscription and \emph{Chained-Scan} emerge as the more portable and adaptable architecture.
-By leveraging the hardware scheduler, \emph{Chained-Scan} dynamically distributes work across architectures without requiring developer intervention.
+Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture.
+By leveraging the workgroup scheduler, \emph{Chained-Scan} dynamically distributes work across architectures without requiring developer intervention.
 Furthermore, its single-kernel design and serial ordering make it suitable for \emph{in-situ} operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging.
@@
 However, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in both theoretical and practical terms, it has a critical limitation: its reliance on a forward-progress guarantee.
-While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that Apple’s M-series and certain ARM GPUs do not.
+While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that Apple’s M-series and ARM GPUs do not.
 Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient—it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
@@
 However, their work primarily focuses on formalizing progress models rather than their direct application to specific algorithms.
-\thomas{Sorensen places Chained-Scan into the HSA category, but it is unclear if this is correct.
-The definition of HSA is \newline ``Threads for which fair execution is guaranteed: The thread with the lowest id that has not yet terminated'' \newline However, as Chained-Scan and other serially ordered algorithms no longer use the virtual ID provided by the workgroup scheduler, is fair execution still guaranteed under HSA?
-It seems that it should be classified under OBE\@.} \john{I don't have an opinion on this, you're familiar with this and I'm not, but this seems like a good footnote.
-Or we can ask Tyler.}
 
@@
 However, because the scheduler lacks FPG, it continues scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction.
-As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.
+As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery}\footnote{\emph{Timeout detection and recovery} is} (TDR) and crashing the host program.
 
@@
 \emph{Scalar Fallback} suffers from two primary inefficiencies.
-First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup enters a waiting spin.
+First, the fallback reduction is performed serially by a single thread, progressing only when the lookback thread performs a waiting spin.
 This results in a $t$-sized blocking tile requiring $t$ spins to complete.
 Additionally, since only one thread loads elements, the associated cache line is severely underutilized.
-Second, once the fallback completes, the workgroup does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback.
+Second, once the fallback completes, the lookback thread does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback.
 \john{If we can, we would like to [politely] say that this previous work was missing a key insight, e.g., while fallback can address the portability problem, \ldots\ Would we say that it's not decoupled in the way that the technique we present in this paper is?
@@
 For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
-  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking subgroup scans,
+  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking Kogge-Stone~\cite{} subgroup scans,
   allowing the subgroup-block scan to remain barrier-free and in registers.
@@
 The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches.
-To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{} scan network embedded with Kogge-Stone subgroup scans to perform the scan across the local spine.
+To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{} scan network embedded with Kogge-Stone~\cite{} subgroup scans to perform the scan across the local spine.
 Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts.
@@
 Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index.
-However, such access is interpreted by the hardware as a \emph{broadcast} operation, resulting in no conflicts.
+However, such access is interpreted by the hardware as a subgroup \emph{broadcast} operation, resulting in no conflicts.
 Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking.
@@
 Workgroups operate independently; each is responsible for updating its own tile state but may also opportunistically update preceding tiles in the event of a fallback.
-Tile states transition in a strictly forward direction---Not Ready $\rightarrow$ Ready $\rightarrow$ Inclusive---and never revert.
+Tile states transition in a strictly forward direction---\emph{Not Ready} $\rightarrow$ \emph{Ready} $\rightarrow$ \emph{Inclusive}---and never revert.
 Before kernel launch, all tiles are initialized to \emph{Not Ready} with reduction values set to $\epsilon$.
@@
                         \begin{itemize}
-                            \item If the blocking tile is the first in the spine, the workgroup has reached the start of the spine and has thus completed the lookback.
+                            \item If the blocking tile is the first in the spine, the workgroup has reached the start of the spine and has thus completed the \emph{Chained Propagation}.
 The first thread proceeds to step 4.
@@
           \item To prioritize inter-workgroup reduction propagation, the first thread immediately adds the local reduction to \emph{previous reduction}, posts the inclusive reduction to global memory, and transitions the tile state to \emph{Inclusive}.
-          \item It then writes \emph{previous reduction} to shared memory, allowing all threads to apply it in their \emph{workgroup-wide} and local scan computations.
+          \item It then writes \emph{previous reduction} to shared memory, allowing all threads to combine it with their \emph{workgroup-wide} and private scan computations.
           \item Finally, the first thread sets the control flag to \emph{Unlocked}, signaling the rest of the workgroup that \emph{Chained Propagation} has completed, and allowing the final write phase to begin.
@@
 \john{I agree with your succinct strategy. I like that you now say reduction-propagation is the first priority, and why.}
+\thomas{No furher revisions required/OK to remove comments?}.
 

### 2025-02-05 paper.tex
Update Appendix to include Dawn (again)

@@
 
-\subsection{Availability}
-We provide our artifact in the following GitHub repository:
-\thomas{Will anonymize with GithubAnonymous once we are ready.}.
-
 \subsection{API Choice}
@@
 
+\subsection{Availability}
+We provide our artifact in the following GitHub repository: 
+\thomas{Will anonymize with GithubAnonymous once we are ready.} 
+We provide both Dawn and WGPU implementations for testing. However, we recommend Dawn, and only use WGPU for ease of native testing on Android.
+
+\subsection{Requirements}
+
+\subsubsection{Device Requirements}
+Any device supporting Dawn, with subgroup and timestamp query capabilities, and at least 384 MB of available device memory.
+
+\subsubsection{Software Requirements}
+\begin{itemize}
+  \item Git
+  \item CMake 3.10.2+ or another C++ build tool
+  \item C++17-compliant compiler
+  \item Python for fetching dependencies
+\end{itemize}
+
+\subsection{Building}
+Although native WebGPU is supported on various operating systems—most notably Windows, macOS, and Linux—this section will focus on building it in Unix-like environments for the sake of brevity, using CMake as the example build tool.
+
+\subsubsection{Building Dawn}
+\begin{enumerate}
+  \item[(0)] Dawn depends on the following packages (Ubuntu package names):
+  \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
+sudo apt install libxrandr-dev \
+  libxinerama-dev libxcursor-dev libxi-dev \
+  libx11-xcb-dev mesa-common-dev
+  \end{lstlisting}
+
+  \item Clone the Dawn repository:
+  \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
+git clone https://dawn.googlesource.com/dawn
+cd dawn
+git fetch --all
+git checkout chromium/6950
+  \end{lstlisting}
+
+  \item Build and install Dawn with your build tool and compiler of choice:
+  \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
+cd dawn
+cmake -S . -B out/Release \
+  -DDAWN_FETCH_DEPENDENCIES=ON \
+  -DDAWN_ENABLE_INSTALL=ON \
+  -DCMAKE_BUILD_TYPE=Release \
+  -DDAWN_BUILD_TESTS=OFF \
+  -DDAWN_BUILD_SAMPLES=OFF \
+  -DTINT_BUILD_TESTS=OFF \
+  -DTINT_BUILD_CMD_TOOLS=OFF
+
+cmake --build out/Release \
+  --parallel <num_threads>
+
+cmake --install out/Release \
+  --prefix install/Release
+  \end{lstlisting}
+\end{enumerate}
+
+\subsubsection{Building Artifact}
+After successfully building and installing Dawn, follow these steps to build the artifact:
+
+\begin{enumerate}
+  \item Navigate to the artifact directory:
+  \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
+cd Decoupled-Fallback-Paper/artifact/Dawn
+  \end{lstlisting}
+
+  \item Set the `CMAKE\_PREFIX\_PATH` to point to your Dawn installation directory:
+  \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
+export CMAKE_PREFIX_PATH=/path/to/dawn/ \  
+  install/Release
+  \end{lstlisting}
+
+  \item Build the artifact:
+  \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
+cmake -S . -B out/Release \
+  -DCMAKE_BUILD_TYPE=Release
+  \end{lstlisting}
+
+  \item Compile the project:
+  \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
+cmake --build out/Release
+  \end{lstlisting}
+\end{enumerate}
+
+\subsubsection{Running the Artifact}
+Once the artifact has been successfully built, you can run it using the following command:
+
+\begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
+./out/Release/dawn <TestType> [record] [deviceName]
+\end{lstlisting}
+
+where:
+\begin{itemize}
+    \item \texttt{<TestType>} specifies the test to run. Valid options are:
+    \begin{itemize}
+        \item \texttt{csdl} – Runs the \emph{Decoupled Lookback} test. \textbf{Warning}, this \emph{will} crash without FPG. 
+        \item \texttt{csdldf} – Runs the \emph{Decoupled Fallback} test.
+        \item \texttt{full} – Runs a full set of tests, including multiple \emph{Decoupled Fallback} simulations.
+    \end{itemize}
+    \item \texttt{record} (optional) enables recording the test results to a CSV file.
+    \item \texttt{deviceName} (optional) appends the device name to recorded test results.
+\end{itemize}
+
 \end{document}

### 2025-02-06 paper.tex
bib cleanup + reply to TS comment
Raph's comments and suggestions

--- 2025-02-04 paper.tex
@@
 
+\raph{I think "NVIDIA Tesla: A Unified Graphics and Computing Architecture" Lindholm 2008 might be a good choice for the above cite.}
+
 On GPU hardware, scheduling is divided into two levels: a \emph{workgroup} scheduler, which manages kernel launches and maps virtual processors, workgroups, to physical processors, \emph{multiprocessors}, and a \emph{subgroup} scheduler, responsible for selecting which of the currently \emph{occupant} subgroups on a multiprocessor for execution.
@@
 This stems from the prohibitively high overhead in latency and storage associated with moving workgroup contexts on and off chip.
-In the oversubscription paradigm, where the number of workgroups is theoretically unbound and typically far exceeds the available multiprocessors, every invocation of such a barrier would necessitate context-switching between all launched workgroups.
+In the oversubscription paradigm, where the number of workgroups is theoretically unbounded and typically far exceeds the available multiprocessors, every invocation of such a barrier would necessitate context-switching between all launched workgroups.
 Due to this high cost, no GPU programming environment furnishes developers with a true unbounded-launch inter-workgroup barrier.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent-Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} Instead, when inter-workgroup synchronization is necessary, developers must rely on kernel launches as synchronization points.
 However, kernel launches are suboptimal as barriers: they incur overheads due to the heterogeneous nature of the operation, and all execution context is lost between launches.
+
+\raph{I think the above is conflating memory barriers with control barriers.
+All modern GPU shading languages (since Metal 3.2) provide memory barriers with inter-workgroup (device) scope, but the above paragraph makes more sense interpreted as a control barrier.
+It might also be worth mentioning that kernel launches do not necessarily sychronize; in Vulkan it's really the pipeline barrier that does the synchronization.
+I can take a whack at re-writing this.}
 
@@
 To allow parallelization, \( \oplus \) must be associative, but it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}.
-In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence.
-If the subset includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}.
+In a scan, the result at the \( n \)-th element is the reduction of the preceding subsequence of elements.
+If the subsequence includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}.
 The most common scan type is the prefix sum, where \( \oplus \) is addition.
@@
 \subsection{Evolution of Inter-Workgroup Scan Architectures}
-Contemporary GPU scans depart from early PRAM-like scans~\cite{} by leveraging the GPU memory hierarchy, which consists of progressively faster but increasingly private tiers.
+Contemporary GPU scans depart from early PRAM-like scans~\cite{Blelloch:1989:SAP} by leveraging the GPU memory hierarchy, which consists of progressively faster but increasingly private tiers.
 Efficient scans integrate fine-grained intra-workgroup (\emph{local}) strategies in shared memory and registers with coarse-grained inter-workgroup (\emph{global}) strategies in global memory.
-Given their low arithmetic intensity, local scans are inherently memory-bound, as demonstrated by Merrill and Grimshaw~\cite{}.
+Given their low arithmetic intensity, local scans are inherently memory-bound, as demonstrated by Merrill and Grimshaw~\cite{Merrill2009}.
 Consequently, recent scan optimizations have shifted toward refining global strategies, particularly inter-workgroup coordination and synchronization.
 This section reviews three key approaches---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---analyzing their handling of inter-workgroup dependencies and efficiency in minimizing global memory access while maximizing bandwidth utilization.
-\thomas{I have written this paragraph like twenty times now, and it's still odious.}
+\thomas{I have written this paragraph like twenty times now, and it's still odious.} \raph{It's fine.
+Feel free to delete these notes in next round.}
 
@@
 In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume.
-During this waiting period, the workgroup is continually scheduled, and the \emph{progress} of other workgroups must be guaranteed to prevent the spinning workgroup from starving its dependent predecessors of execution, which could result in deadlock.
+This approach depends on a \emph{forward progress guarantee} of the GPU's execution model.
+Given such a guarantee, the predecessor will eventually produce a result and broadcast it.
+On devices lacking such a guarantee, the workgroup may wait indefinitely, resulting in deadlock.
 
@@
 
+\raph{I'm having trouble with the above paragraph.
+By my read, Reduce-Then-Scan avoids inter-workgroup synchronization other than the pipeline barriers between kernel launches.
+Chained-Scan and Decoupled-Lookback both rely on inter-workgroup synchronization.
+Not sure why the "in some cases" qualifier either; seems to me you need synchronization except in the trivial case the entire array fits in one workgroup.}
+
 Fixed-launch approaches, such as those used in \emph{Reduce-Then-Scan}, introduce several challenges.
@@
 For example, the workgroup occupancy discovery technique proposed by Sorensen et al.~\cite{sorensen2016} relies on memory fences—unsupported in WebGPU—making it infeasible in certain environments.\footnote{Our implementation includes a fenceless occupancy discovery method suitable for WebGPU, but its description is beyond the scope of this work.}
+
+\raph{"Fixed launch" means launching a fixed number of workgroups and having each workgroup process multiple blocks, generally total blocks / num workgroups?
+I think I get the argument here but it's more subtle and not an inherent problem with RTS.
+You make a choice: either you do fixed launch and have to tune, *or* you can do a fixed number of blocks per workgroup, but in the latter case you need to need vary the number of dispatches based on total number of blocks.
+That's annoying but not a showstopper.}
 
@@
 While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that Apple’s M-series and ARM GPUs do not.
-Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient—it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
+Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
 
@@
 However, their work primarily focuses on formalizing progress models rather than their direct application to specific algorithms.
+
+\raph{I think the deep-dive into the nature of deadlock can be trimmed back here, but chatting with John, such editing for concision can be done later, when we have a better sense about page count.}
 
@@
 
+\raph{Answering John's question here.
+Short answer no.
+One mental model that would illustrate this is a workgroup being preempted from a multiprocessor for system reasons, for example to perform compositing for scanout.
+I brought this up with a GPU engineer and they said this wasn't actually the reason for lack of FPG on their hardware, but wasn't able to explain the true story for proprietary reasons.
+In any case, while I am fine to a large extent talking operationally about scheduling etc, at the end of the day you need a formal model that tells you which computations will terminate.
+That's what Tyler's work is about and I think it's best to cite it and not try to re-litigate it, especially in oversimplified form.
+Also, taking another look at "Specifying and Testing," I notice that they use "livelock" but not deadlock to describe situations where system progress depends on FPG.
+I'm not finding a cite for this now, but if memory serves, a deadlock condition is one in which no progress is possible under *any* scheduling condition, while in livelock there is some scheduler condition that could resolve the livelock, but it is not resolved under all such conditions.}
+
 For example, consider a scan operation requiring two workgroups on a GPU with a single multiprocessor.
@@
 \subsection{Earlier Attempts at Portability}
-\john{I'd like Raph to chime in here.
-What I worry about in the way this subsection is presented is that a reviewer will say ``Oh, the fallback technique was already invented in 2021, all this paper does is makes it a little better''.
-On one hand I totally want Raph to get credit for this cool idea.
-On the other hand, writing this subsection in this way, as a citable previous publication, significantly reduces the size of the advance that this paper we're writing hopes to capture.
-If I'm being selfish and trying to make this paper look as good as possible, I downplay the previous work a little and don't present it as if it was a previous peer-reviewed publication.
-At the least I want to use the words ``blog post'' in the text here.}
-Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{Decoupled-Lookback} approach without FPG requirements in \emph{Scalar Fallback}.
-Each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time.
-This eliminates the \emph{mutual exclusion} condition necessary for deadlock, as workgroups can operate on data from other tiles.
-While this may introduce redundant work when a tile is late rather than truly blocking, it guarantees termination by allowing any workgroup to complete a blocking tile’s reduction, breaking inter-workgroup dependencies.
+Our first attempt \raph{we'll add the cite post-review} at implementing a \emph{Decoupled-Lookback} scan without relying on FPG was \emph{Scalar Fallback}.
+Each time a workgroup spins waiting on the reduction from a preceding tile, it advances a \emph{fallback} operation by processing one scalar element from that tile.
+This eliminates the possibility of deadlock because the number of such operations is bouded; the wait will eventually terminate, possibly at the expense of doing redundant work.
 
@@
 This results in a $t$-sized blocking tile requiring $t$ spins to complete.
-Additionally, since only one thread loads elements, the associated cache line is severely underutilized.
 Second, once the fallback completes, the lookback thread does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback.
-\john{If we can, we would like to [politely] say that this previous work was missing a key insight, e.g., while fallback can address the portability problem, \ldots\ Would we say that it's not decoupled in the way that the technique we present in this paper is?
-Or is the only deficiency the choice of a serial fallback?} While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce then Scan}.
+While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce then Scan}.
+
+\raph{I took out the statement about cache underutilization because I'm not sure it's true (L1 cache is shared by all subgroups in a multiprocessor) and I don't think it's essential to our argument.
+Also deleting John's comment; I don't think this is a case where we need to prove novelty over a previous publication, instead we just subsume it.}
 
 \section{Decoupled Fallback}
-\emph{Decoupled Fallback} separates fallback operations from the lookback process, enhancing the \emph{Scalar-Fallback} technique with significantly improved fallback performance.
-As an extension of the \emph{Chained-Scan} architecture, it retains the advantages of high speed and oversubscription, aligning with our design requirements.
+\emph{Decoupled Fallback} addresses both inefficiencies.
+The key advance is to utilize the parallelism available in the workgroup to perform the fallback computation, rather than performing it sequentially.
+To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained-Propagation} logic is performed by a single thread.
 
-Like \emph{Scalar-Fallback}, we choose \emph{mutual exclusion} as our approach to preclude deadlocks.
+\raph{This paragraph feels to me like it can be cut, but I won't insist on it.} Like \emph{Scalar-Fallback}, we choose \emph{mutual exclusion} as our approach to preclude deadlocks.
 \emph{Dependency relationships} and \emph{no preemption} are immutable characteristics of scan and the GPU programming model, eliminating them as viable avenues for breaking deadlocking.
@@
 To optimize the fallback operation, we employ three strategies.
-Although \emph{Chained-Propagation} is performed by a single thread or subgroup, for performance reasons we want the entire workgroup to participate in the fallback.
 (1) First, a shared memory \emph{control flag} enables threads performing the lookback to signal the entire workgroup when a fallback is necessary.
@@
 \john{I can see it is sensible to have a maximum spin count but I do not see how the action of exposing the maximum spin count to the programmer minimizes false positives.} (3) Finally, after completing a fallback, the workgroup attempts to post the reduction result of the blocking tile, preventing redundant fallbacks by subsequent workgroups.
+
+\raph{To address John's comment: increasing the spin count reduces false positives; a false positive can be defined as the preceding tile posting the result late, meaning that there is some spin count for which it would proceed.
+However, it also delays progress in the true positive case.
+The deeper way to talk about this is to engage the cost model, which is done to some extent in the paragraph below.
+The cost of a false positive includes increased memory traffic and also compute opportunity cost of other workgroups which can be scheduled on the same multiprocessor.
+Indeed, with sufficient occupancy the cost of a workgroup spinning (or making scalar progress) is minimal.
+So optimal performance really depends on the *distribution* of spin counts.
+I put all this in a comment but think we can say it in the paper.}
 
@@
 
-\john{In this item + the previous item, I bet there's some ordering constraints, either for correctness or for performance.
-For instance, I bet the highest priority is to add the previous reduction to my reduction and posting the result, since that propagation speed across tiles is probably really important for performance.
-Perhaps worth pointing out.
-Does the flag-unset have any ordering constraints in terms of correctness?}
-\thomas{Last phase may still need revision in accordance with above feedback. . . I have deliberately tried to avoid mentioning barriers to keep the section succinct.}
-\john{I agree with your succinct strategy. I like that you now say reduction-propagation is the first priority, and why.}
-\thomas{No furher revisions required/OK to remove comments?}.
+\subsection{Tile State Implementation}
+In all variants of \emph{Chained-Scan}, there is communication and synchronization between workgroups.
+The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it.
+This primitive is well studied in the context of memory models.
+\raph{The earliest publication of the MP litmus test appears to be section 2.1 of "Intel 64 Architecture Memory Ordering White Paper".
+I'm not sure that's the best cite; the MC mutants paper cites https://arxiv.org/pdf/1207.7264} The key requirement is both control synchronization, meaning the reader does not proceed until the writer sends the message; and data synchronization, meaning that the reader reads the correct value.
+The latter is especially challenging in relaxed memory models as are employed on GPUs.
 
-\subsection{Tile State Implementation}
-Since workgroups must observe updates from one another, but WebGPU lacks a \emph{volatile}-like qualifier, we rely on atomics to ensure visibility and consistency across workgroups.
-Because a tile may be in three possible states, two bits are required to record its state alongside the bits needed to store the tile's reduction value.
-Previous \emph{Chained-Scan} architectures relied on 64-bit atomics and/or memory fences to ensure a consistency between state and value, but WebGPU’s lack of a memory model \john{I'm not sure I would say ``lack of a memory model''.
-It's not the (lack of) memory model that's the problem.
-It's specifically that we don't have the operations that give us the behavior that we require.
-If we had the right operations that did the right thing but didn't have a memory model, we'd be fine.
-We want to specifically call out what is missing and what that means, and do so as generically as possible (not ``we are missing this specific atomic call'' but instead ``we are missing some sort of atomic call that has the following behavior'').
-Above, we want to not only say that we don't have a volatile qualifier but also both what that means and why it would be specifically useful if we had it.
-Doing the latter two things gives valuable ammunition to WebGPU and GPU architects that want to improve the API and hardware.} and 64-bit atomic support makes both infeasible.
-Instead, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state.
-\john{I would try to add a sentence or two on why we can't do the obvious thing---having the state in one word and the reduction value in a different word---and why the split value avoids that problem.
-This is not obvious to the reader.}
+Combined control and data synchronization can be accomplished by packing both the tile state and the reduced value into a single word, using relaxed atomic store to send the message, and relaxed atomic load to receive the message, spinning as necessary until the tile state is updated.
+The semantics of atomics guarantee, even with relaxed memory ordering, that the data is fresh.
+However, if the tile state and data are stored in two different memory locations, then relaxed atomics are not sufficient and some other sychronization is necessary.
+On CPUs, it is common to upgrade the memory semantics from relaxed to acquire/release, which then provides the necessary guarantee.
+\raph{Here's a good place to cite the MP litmus test because that's exactly what it's about.} Unfortunately, on GPUs, acquire/release semantics are neither portable (they are only available on Vulkan, and optional until version 1.3), nor are they necessarily efficient even when they are available; often the implementation involves expensive cache flushes.
+A device-scoped memory barrier has similar problems.
+While it would provide the necessary guarantee, it is also not available in WebGPU (or in versions of Metal prior to 3.2), and is expensive.
+
+Without some other mechanism, the requirement to pack tile state and data into a single word limits the data to 30 bits (assuming 2 bits for the tile state).
+That, in turn, makes it impossible to represent common ordinary 32-bit floating point values without loss of precision.
+While some GPUs support 64 bit atomics (and it is likely that WebGPU will include them as an optional extension), that is also sadly not portable, and merely raises the payload limit to 62 bits without eliminating it altogether.
+
+A key innovation of this paper is to \emph{split} the reduced values into multiple atomics, each carrying its own copy of the control flags.
+The message is received when relaxed atomic loads of all the parts agree on their control flag.
+This way, the payload is reconstructed intact, even in the absence of any guarantees from the memory model stronger than relaxed atomics.
+
+In particular, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state.
 \begin{figure}[h!]
@@
 When an update attempt is made, a blocking tile may be in any of the three states due to either of two scenarios: (1) a false-positive fallback on a tile that transitions to \emph{Ready} or \emph{Inclusive} before fallback completion, or (2) another workgroup completing its fallback first, leaving the tile in \emph{Ready}.
-WebGPU does not guarantee the correctness of atomic \emph{Compare-and-Swap} across all devices, inhibiting its use.
-\john{Is CAS the traditional/common way to implement this?
-If so, say so, and if you can, provide a cite, even if it's not a formal cite but just the name of another package that uses this strategy.} Instead, we encode the tile state in the most significant bits of the split value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}.
+If the tile state is \emph{Inclusive} then the write must be skipped, especially because races in writes for the parts may leave the tile permanently in an unreadable state.
+Thus, a simple atomic store is not suitable.
+An atomic compare-and-swap would have the correct semantics, but WebGPU lacks this primitive (it does have \texttt{atomicCompareExchangeWeak}, which could possibly be made to work but makes reasoning about correctness and performance more difficult).
+Instead, we encode the tile state in the most significant bits of the atomic value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}.
 This allows an atomic \emph{maximum}, which successfully updates to a more advanced state but prevents a state regression, to correctly perform updates.
+
+The atomic splitting technique generalizes to monoids of arbitrary size.
 

### 2025-02-06 paper.tex
Add Real Data

--- 2025-02-05 paper.tex
@@
 \subsection{Experimental Setup}
-We configure workgroups with a size of 256 threads, where each thread processes 4 vectors of size 4, resulting in 16 elements per thread and a total work tile size of 4096.
+\begin{table*}
+  \centering
+  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r}
+    \toprule
+    Metric & Nvidia 2080 Super & Intel HD620 & AMD 7900 XT & ARM Mali-G78 MP20 & Apple M1 Pro Max \\  
+    \midrule
+    Memcopy                  & 54.203  & 1.498   & 71.190  & 106.663 & 54.203  \\  
+    RTS                      & 35.709  & 1.060   & 62.868  & 82.302  & 35.709  \\  
+    CSDLDF                   & 53.256  & 1.487   & 95.737  & 112.430 & 53.256  \\  
+    CSDLDF/Memcopy           & 0.983   & 0.992   & 1.345   & 1.054   & 0.983   \\  
+    CSDLDF/RTS               & 1.491   & 1.402   & 1.523   & 1.366   & 1.491   \\  
+    Avg. Spins/Workgroup     & 0.418   & 0.306   & 0.133   & 1.380   & 0.418   \\  
+    Avg. Lookback/Workgroup  & 13.139  & 1.725   & 20.867  & 3.510   & 13.139  \\  
+    Fallbacks Initiated      & 88.384  & 260.364 & 10.127  & 2202.045& 88.384  \\  
+    Successful Insertions    & 1.205   & 6.084   & 1.001   & 57.875  & 1.205   \\  
+    \bottomrule
+  \end{tabular*}
+  \caption{Harmonic Means and Ratios for Various Devices, (Mali Absolute Statistics incorrect, need M1 Statistics)}
+  \label{tab:results}
+\end{table*}
+
+We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size $t = 4096$.
 We find a maximum spin count of 4 as offering good responsiveness to blocking, while minimizing false-positives fallbacks.
 As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the minimum subgroup size specified by the WebGPU standard, which is 4.
-Despite the additional shared memory required for conflict avoidance and fallback operations, total shared memory usage remains modest at only 1 KB.
+Due to our overall thriftiness with shared memory, total shared memory usage remains modest at only 1 KB.
+We choose a diverse range of hardware vendors, with.
 For all tests, we use an input size of $2^{25}$.
 
-Something about devices here.
 \subsection{Performance Comparison}
@@
 Memcopy is included as a point of reference, as it shares the same $2n$ global memory movement required by scan operations.
-
-\thomas{Forgot about tables. Mega table incoming.}
-
-\thomas{Performance is good}
 
@@
 Starting with infrequent omissions---1 in 512 workgroups---we progressively increase the rate to near-continuous levels, reaching 1 in 2 workgroups.
-\begin{figure}
+\begin{figure*}[t] % [t] forces the figure to appear at the top
   \centering
-  \begin{subfigure}{.9\linewidth}
+  % First row: two subfigures side by side
+  \begin{subfigure}[t]{0.48\linewidth}
     \centering
@@
     \label{fig:fallbacks_initiated}
-  \end{subfigure}
-  \begin{subfigure}{.9\linewidth}
+  \end{subfigure}\hfill
+  \begin{subfigure}[t]{0.48\linewidth}
     \centering
@@
   \end{subfigure}
-  \begin{subfigure}{.9\linewidth}
+
+  \vspace{0.2em} % Adjust vertical spacing between rows
+
+  % Second row: two subfigures side by side
+  \begin{subfigure}[t]{0.48\linewidth}
     \centering
@@
     \label{fig:lookback_length}
-  \end{subfigure}
-  \begin{subfigure}{.9\linewidth}
+  \end{subfigure}\hfill
+  \begin{subfigure}[t]{0.48\linewidth}
     \centering
     \includegraphics[width=\linewidth]{graphics/time_plot.pdf}
-    \caption{Speed vs Memcopy}
+    \caption{Throughput vs Baseline}
     \label{fig:execution_time}
   \end{subfigure}
-  \caption{Simulated Blocking\john{Text in these plots---axis labels, etc.---needs to be larger. If it's significantly smaller than body text, which it is, it's just not legible.}}
+
+  \caption{Simulated Blocking}
   \label{fig:vertical_images}
-\end{figure}
+\end{figure*}
 

### 2025-02-07 paper.tex
Catch bug in simulated blocking. Add performance comparison section.

--- 2025-02-05 paper.tex
@@
 \usepackage{subcaption}
+\usepackage{makecell}
 \newcommand{\thomas}[1]{{\footnotesize\color{orange}[Thomas: #1]}}
@@
 Although \emph{Decoupled-Lookback} increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing confines spine memory accesses to a roughly $o$-width sliding window, making them highly likely to be cache-resident.
-As a result, the additional global memory traffic along the spine introduces a negligible latency overhead.
+As a result, the additional global memory traffic along the spine introduces negligible overhead.
 With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled-Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance without the work tile size tuning required by \emph{Stream-Scan}.
@@
 Tile states transition in a strictly forward direction---\emph{Not Ready} $\rightarrow$ \emph{Ready} $\rightarrow$ \emph{Inclusive}---and never revert.
-Before kernel launch, all tiles are initialized to \emph{Not Ready} with reduction values set to $\epsilon$.
+Before kernel launch, all tiles are initialized to \emph{Not Ready} with reduction values set to $\varepsilon$.
 State transitions imply an update of the reduction value: a transition to \emph{Ready} results in the posting of the local workgroup reduction, and to \emph{Inclusive}, the reduction of all previous workgroup reductions.
@@
           \item The control flag, initialized to \emph{Locked}, signals that the workgroup should enter the \emph{Chained-Propagation} phase.
-          \item The first thread initializes a \emph{previous reduction} variable in registers with $\epsilon$ to hold the accumulated reductions of preceding tiles.
+          \item The first thread initializes a \emph{previous reduction} variable in registers with $\varepsilon$ to hold the accumulated reductions of preceding tiles.
           \item The first thread initializes a \emph{current tile} pointer to its immediate predecessor, with which it begins to traverse backwards through the spine.
@@
 
-\john{In this item + the previous item, I bet there's some ordering constraints, either for correctness or for performance.
-For instance, I bet the highest priority is to add the previous reduction to my reduction and posting the result, since that propagation speed across tiles is probably really important for performance.
-Perhaps worth pointing out.
-Does the flag-unset have any ordering constraints in terms of correctness?}
-\thomas{Last phase may still need revision in accordance with above feedback. . . I have deliberately tried to avoid mentioning barriers to keep the section succinct.}
-\john{I agree with your succinct strategy. I like that you now say reduction-propagation is the first priority, and why.}
-\thomas{No furher revisions required/OK to remove comments?}.
-
 \subsection{Tile State Implementation}
@@
 \subsection{Experimental Setup}
-We configure workgroups with a size of 256 threads, where each thread processes 4 vectors of size 4, resulting in 16 elements per thread and a total work tile size of 4096.
-We find a maximum spin count of 4 as offering good responsiveness to blocking, while minimizing false-positives fallbacks.
-As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the minimum subgroup size specified by the WebGPU standard, which is 4.
-Despite the additional shared memory required for conflict avoidance and fallback operations, total shared memory usage remains modest at only 1 KB.
-For all tests, we use an input size of $2^{25}$.
+We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$.
+Through testing, we find a maximum spin count of 4 to effectively balance responsiveness to blocking while minimizing false-positive fallbacks.
+As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4.
+Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB.
+For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum.
+Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities.
+We empirically find that the Google Chrome native implementation of WebGPU (Dawn) provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU.\footnote{To keep the perforamance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU.}
+\begin{table}[t]
+  \centering
+  \small
+  \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} l l c r r}
+    \toprule
+    Vendor & Device Name & FPG & Occupancy & Subgroup Size \\  
+    \midrule
+    Intel  & HD Graphics 620       & \textbf{Yes} & 9   & 8 - 32  \\  
+    ARM    & Mali-G78 MP20          & \textbf{Yes} & 40  & 16  \\  
+    Apple  & M1 Pro Max             & \textbf{No}  & 64  & 32  \\  
+    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192 & 32  \\  
+    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420 & 32 - 64  \\  
+    \bottomrule
+  \end{tabular*}
+  \caption{Hardware Specifications of Testing Devices}
+  \label{tab:hardware}
+\end{table}
 
-Something about devices here.
 \subsection{Performance Comparison}
-To evaluate the performance of \emph{Decoupled-Fallback}, we compare it against two key benchmarks: \emph{Reduce-Then-Scan} (RTS) and a \emph{Memcopy} operation.
-RTS serves as a baseline to demonstrate the performance gains of our approach over the previous FPG-constrained technique.
-Memcopy is included as a point of reference, as it shares the same $2n$ global memory movement required by scan operations.
+To evaluate the performance of \emph{Decoupled Fallback} (\emph{CSDLDF}), we compare it against two key benchmarks: \emph{Memcopy}\textdagger{~\ref{sec:memcopy}} and \emph{Reduce then Scan}\textdagger{~\ref{sec:rts}} (\emph{RTS}).
+\emph{RTS} serves as a baseline to demonstrate the performance gains of our approach over the previous FPG-constrained technique.
+\emph{Memcopy} is included as a point of reference, as it shares the same $2n$ global memory movement as \emph{Decoupled Fallback}, without any compute overhead and thus is guaranteed to be memory bandwidth bound.
 
-\thomas{Forgot about tables. Mega table incoming.}
+\begin{table*}
+  \centering
+  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r}
+    \toprule
+    Metric & Intel HD620 & ARM Mali-G78 MP20 & Apple M1 Pro Max & Nvidia 2080 Super & AMD 7900 XT \\  
+    \midrule
+    H-Mean \emph{Memcopy} (32-bit ele/sec)  & 1.498e9   & 106.663e9 & 54.203e9  & 54.203e9  & 71.190e9  \\
+    H-Mean \emph{CSDLDF} (32-bit ele/sec)   & 1.487e9   & 112.430e9 & 53.256e9  & 53.256e9  & 95.737e9  \\  
+    H-Mean \emph{RTS} (32-bit ele/sec)      & 1.060e9   & 82.302e9  & 35.709e9  & 35.709e9  & 62.868e9  \\  
+    \emph{CSDLDF}/\emph{Memcopy} (\%)       & 99.2\%    & 105.4\%   & 98.3\%    & 98.3\%    & 134.5\%   \\  
+    \emph{CSDLDF}/\emph{RTS} (\%)          & 140.2\%   & 136.6\%   & 149.1\%   & 149.1\%   & 152.3\%   \\  
+    Avg Spins per Wg    & 0.306    & 1.380    & 0.418    & 0.418    & 0.133    \\  
+    Avg Lookback Length per Wg  & 1.725    & 3.510    & 13.139   & 13.139   & 20.867   \\  
+    Avg Total Fallbacks Initiated      & 260.364   & 2202.045  & 88.384   & 88.384   & 10.127   \\  
+    Avg Total Successful Insertions    & 6.084     & 57.875    & 1.205    & 1.205    & 1.001    \\
+    \bottomrule
+  \end{tabular*}
+  \caption{MALI ABSOLUTE SPEED WRONG!, \emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, Input Size $\mathbf{2^{25}}$}
+  \label{tab:results}
+\end{table*}
 
-\thomas{Performance is good}
+As shown in Table~\ref{tab:results}, with the exception of the ARM Mali, \emph{CSDLDF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices.
+Since \emph{Memcopy} is purely memory bandwidth bound, \emph{CSDLDF} should ideally match its throughput.
+This holds across most devices, except for the AMD 7900 XT, where \emph{CSDLDF} slightly outperforms \emph{Memcopy}, suggesting improved memory structuring due to compute operations.
+On the ARM Mali, \emph{CSDLDF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%.
+However, as its performance is nearly identical to \emph{Memcopy}, this indicates that the implementation is memory bandwidth bound, limiting further gains.
 
@@
 
-\subsection{Emulated Blocking}
+\subsection{Simulated Blocking}
 To demonstrate the resilience of our technique under fallback rates significantly higher than those observed organically on devices, we intentionally force fallbacks by omitting the posting of work tile reductions.
 Starting with infrequent omissions---1 in 512 workgroups---we progressively increase the rate to near-continuous levels, reaching 1 in 2 workgroups.
-\begin{figure}
+\begin{figure*}[t]
   \centering
-  \begin{subfigure}{.9\linewidth}
+  \begin{subfigure}[t]{0.48\linewidth}
     \centering
@@
   \end{subfigure}
-  \begin{subfigure}{.9\linewidth}
+
+  \vspace{0.2em}
+
+  \begin{subfigure}[t]{0.48\linewidth}
     \centering
@@
 In Figure~\ref{fig:total_spins} we measure \thomas{this should be spins per workgroup it's a better metric}.
-In Figure~\ref{fig:lookbackLength} we see that.
+In Figure~\ref{fig:lookback_length} we see that.
 Finally, in Figure~\ref{fig:execution_time}, we can see that, even when \emph{half} the workgroups are deadlocking, \emph{Decoupled-Fallback} still achieves speeds of .
@@
 
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{bib}
-
-\clearpage
 \appendix
 \section{Artifact}
+\subsection{Baseline Comparisons: \emph{Memcopy}}
+\label{sec:memcopy}
+\begin{table}[h!]
+  \small
+  \centering
+  \begin{tabular}{l r r}
+    \toprule
+    Device & \makecell{Memory \\ Bandwidth \\ (GB/s)} & \makecell{Expected \\ Throughput \\ (32-bit ele/sec)} \\  
+    \midrule
+    Intel HD620 (Single-/Dual-Channel) & 17.1 / 34.1  & 2.13e9 / 4.26e9 \\  
+    ARM Mali-G78 MP20 / Pixel 6a (Tensor) & 51.2  & 6.40e9 \\  
+    Apple M1 Pro Max   & 200  & 25.0e9 \\  
+    Nvidia 2080 Super  & 448  & 56.0e9 \\  
+    AMD 7900 XT        & 800  & 100.0e9 \\  
+    \bottomrule
+  \end{tabular}
+  \caption{Vendor-Specified Memory Bandwidth and Expected Throughput (32-bit Elements per Second)}
+  \label{tab:memory_bandwidth}
+\end{table}
+Although all backend APIs support \emph{timestamp queries inside encoders}, Apple GPUs do not, so we are forced to implement our own \emph{Memcopy} kernel to measure memory bandwidth, as opposed to simply timing an API's copy buffer operation.
+It is well known that \emph{observed} peak memory bandwidth, tends to diverge from vendor-listed peak bandwidth, so for completeness we list these figures above.
+We make a number of observations:
+\begin{itemize}
+  \item \textbf{Intel HD620:} As our HD620 laptop is only equipped with a single memory DIMM, its peak bandwidth is limited to the lower bound, 2.13e9 ele/sec.
+Thus, our observed bandwidth of 1.498e9 is reasonable.
+  \item \textbf{ARM Mali-G78 MP20 / Pixel 6a (Tensor):} The observed throughput is significantly lower than expected, despite a vendor-listed bandwidth of 6.40e9 ele/sec.
+The cause of this discrepancy remains unclear, but we note other developers have observed similar speeds.
+  \item \textbf{Apple M1 Pro Max:} Observed throughput matches expectations.
+  \item \textbf{Nvidia 2080 Super:} Observed throughput matches expectations.
+  \item \textbf{AMD 7900 XT:} Our above result, where \emph{CSDLDF} was observed to be faster than \emph{Memcopy}, is likely due to compute latency counterintuitively improving memory access patterns, leading to reduced paging.
+Furthermore, we see that our \emph{CSDLDF} throughput is within spec and reasonable.
+\end{itemize}
 
-\subsection{API Choice}
-As rust is significantly more easy to build across devices we chose the WGPU api.
-
-\subsection{Device Choices}
-We chose a diverse range of devices to illustrate
-
-\subsection{Baseline Comparisons}
-Since WebGPU limits device-side timestamps to compute or render passes, we implement a custom Memcopy kernel for timing purposes.
-Similarly, as there are no existing WebGPU implementations of RTS, we rely on our own optimized version for comparison.
-To ensure a competitive evaluation, our RTS implementation employs a variable-launch main kernel with a single workgroup performing a serial spine scan, as we found raking to be less performant in practice.
-Apart from this adjustment, the intraworkgroup implementation in RTS is identical to that used in \emph{Decoupled-Fallback}.
+\subsection{Baseline Comparisons: \emph{Reduce then Scan}}
+\label{sec:rts}
+As we desire the most competitve baseline possible, our \emph{RTS} kernels employ the same intra-workgroup scan strategy and tuning parameters as \emph{CSDLDF}.
+Initially, we adopted Merrill and Grimshaw's~\cite{Merrill2009} workgroup raking approach for inter-workgroup processing.
+However, empirical testing revealed that raking was consistently outperformed—by up to $\sim$15\% on lower-end devices—by a simpler approach that processes the spine serially with a single workgroup.
+Furthermore, our \emph{RTS} performance closely aligns with the theoretically expected $O(3n)$ global memory movement.
 
@@
 \thomas{Will anonymize with GithubAnonymous once we are ready.}
-We provide both Dawn and WGPU implementations for testing. However, we recommend Dawn, and only use WGPU for ease of native testing on Android.
 
@@
 \subsubsection{Device Requirements}
-Any device supporting Dawn, with subgroup and timestamp query capabilities, and at least 384 MB of available device memory.
+Any device supporting Dawn, with subgroup and timestamp query capabilities, and at least 384 MB of available device memory. Requires at least $\sim$2GB of disk memory to build Dawn.
 
@@
 
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{bib}
+
 \end{document}

### 2025-02-07 paper.tex
@b0nes164 just making some comments on these nice results

@@
 We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$.
-Through testing, we find a maximum spin count of 4 to effectively balance responsiveness to blocking while minimizing false-positive fallbacks.
+Through testing, \john{across all devices (?)} we find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks.
 As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4.
-Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB.
-For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum.
+Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@.
+For all tests, we use an input size of $2^{25}$ \john{It would be worthwhile, I think, to run a test against CUB where the x axis is problem size and the y axis is throughput.
+A reviewer might think we're hiding something if we only report results for one (large) input size.
+I imagine that our performance will look similar to CUB's at all input sizes and that is the result we want.} and an inclusive prefix sum.
 Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities.
-We empirically find that the Google Chrome native implementation of WebGPU (Dawn) provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU.\footnote{To keep the perforamance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU.}
-\begin{table}[t]
+We empirically find that the Google Chrome native implementation of WebGPU (Dawn) provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU\@.\footnote{To keep the perforamance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
+\begin{table}
   \centering
@@
     \midrule
-    Intel  & HD Graphics 620       & \textbf{Yes} & 9   & 8 - 32  \\  
+    Intel  & HD Graphics 620       & \textbf{Yes} & 9   & 8--32  \\
     ARM    & Mali-G78 MP20          & \textbf{Yes} & 40  & 16  \\
@@
     NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192 & 32  \\
-    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420 & 32 - 64  \\  
+    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420 & 32--64  \\
     \bottomrule
   \end{tabular*}
-  \caption{Hardware Specifications of Testing Devices}
+  \caption{Hardware Specifications of Testing Devices.}
   \label{tab:hardware}
@@
 \emph{Memcopy} is included as a point of reference, as it shares the same $2n$ global memory movement as \emph{Decoupled Fallback}, without any compute overhead and thus is guaranteed to be memory bandwidth bound.
+\john{Nit: I would use \texttt{memcpy} as the name of this benchmark to match how it's actually called in C etc.}
 
@@
   \end{tabular*}
-  \caption{MALI ABSOLUTE SPEED WRONG!, \emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, Input Size $\mathbf{2^{25}}$}
+  \caption{MALI ABSOLUTE SPEED WRONG!, \emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, Input Size $\mathbf{2^{25}}$.
+\john{I would label metrics as ``H-Mean \emph{Memcopy} (32-bit ele/sec $\cdot 10^9$)'' and then leave the exponents out of the main table.}\john{Separately, h-mean is the right summary statistic \emph{if} the distribution is Gaussian, which it almost certainly is not.
+Like most measurements of this type, it's much more likely log-normal.
+I am influenced here by \url{https://lemire.me/blog/2023/04/06/are-your-memory-bound-benchmarking-timings-normally-distributed/}.}\john{I note Apple and NVIDIA timings are identical, but figure you know that.}}
   \label{tab:results}
@@
 This holds across most devices, except for the AMD 7900 XT, where \emph{CSDLDF} slightly outperforms \emph{Memcopy}, suggesting improved memory structuring due to compute operations.
-On the ARM Mali, \emph{CSDLDF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%.
-However, as its performance is nearly identical to \emph{Memcopy}, this indicates that the implementation is memory bandwidth bound, limiting further gains.
+\john{I have no idea what this previous phrase means.} On the ARM Mali, \emph{CSDLDF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%.
+However, as its performance is nearly identical to \emph{Memcopy}, this indicates that the implementation is memory-bandwidth-bound, limiting further gains.
 
@@
     \begin{itemize}
-        \item \texttt{csdl} – Runs the \emph{Decoupled Lookback} test. \textbf{Warning}, this \emph{will} crash without FPG. 
+        \item \texttt{csdl} – Runs the \emph{Decoupled Lookback} test. \textbf{Warning}, this \emph{will} crash without FPG\@.
         \item \texttt{csdldf} – Runs the \emph{Decoupled Fallback} test.

### 2025-02-07 paper.tex

@@
 \subsection{Tile State Implementation}
-Since workgroups must observe updates from one another, but WebGPU lacks a \emph{volatile}-like qualifier, we rely on atomics to ensure visibility and consistency across workgroups.
-Because a tile may be in three possible states, two bits are required to record its state alongside the bits needed to store the tile's reduction value.
-Previous \emph{Chained-Scan} architectures relied on 64-bit atomics and/or memory fences to ensure a consistency between state and value, but WebGPU’s lack of a memory model \john{I'm not sure I would say ``lack of a memory model''.
-It's not the (lack of) memory model that's the problem.
-It's specifically that we don't have the operations that give us the behavior that we require.
-If we had the right operations that did the right thing but didn't have a memory model, we'd be fine.
-We want to specifically call out what is missing and what that means, and do so as generically as possible (not ``we are missing this specific atomic call'' but instead ``we are missing some sort of atomic call that has the following behavior'').
-Above, we want to not only say that we don't have a volatile qualifier but also both what that means and why it would be specifically useful if we had it.
-Doing the latter two things gives valuable ammunition to WebGPU and GPU architects that want to improve the API and hardware.} and 64-bit atomic support makes both infeasible.
-Instead, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state.
-\john{I would try to add a sentence or two on why we can't do the obvious thing---having the state in one word and the reduction value in a different word---and why the split value avoids that problem.
-This is not obvious to the reader.}
+In all variants of \emph{Chained-Scan}, there is communication and synchronization between workgroups.
+The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it.
+This primitive is well studied in the context of memory models.
+\raph{The earliest publication of the MP litmus test appears to be section 2.1 of "Intel 64 Architecture Memory Ordering White Paper".
+I'm not sure that's the best cite; the MC mutants paper cites https://arxiv.org/pdf/1207.7264} The key requirement is both control synchronization, meaning the reader does not proceed until the writer sends the message; and data synchronization, meaning that the reader reads the correct value.
+The latter is especially challenging in relaxed memory models as are employed on GPUs.
+
+Combined control and data synchronization can be accomplished by packing both the tile state and the reduced value into a single word, using relaxed atomic store to send the message, and relaxed atomic load to receive the message, spinning as necessary until the tile state is updated.
+The semantics of atomics guarantee, even with relaxed memory ordering, that the data is fresh.
+However, if the tile state and data are stored in two different memory locations, then relaxed atomics are not sufficient and some other sychronization is necessary.
+On CPUs, it is common to upgrade the memory semantics from relaxed to acquire/release, which then provides the necessary guarantee.
+\raph{Here's a good place to cite the MP litmus test because that's exactly what it's about.} Unfortunately, on GPUs, acquire/release semantics are neither portable (they are only available on Vulkan, and optional until version 1.3), nor are they necessarily efficient even when they are available; often the implementation involves expensive cache flushes.
+A device-scoped memory barrier has similar problems.
+While it would provide the necessary guarantee, it is also not available in WebGPU (or in versions of Metal prior to 3.2), and is expensive.
+
+Without some other mechanism, the requirement to pack tile state and data into a single word limits the data to 30 bits (assuming 2 bits for the tile state).
+That, in turn, makes it impossible to represent common ordinary 32-bit floating point values without loss of precision.
+While some GPUs support 64 bit atomics (and it is likely that WebGPU will include them as an optional extension), that is also sadly not portable, and merely raises the payload limit to 62 bits without eliminating it altogether.
+
+A key innovation of this paper is to \emph{split} the reduced values into multiple atomics, each carrying its own copy of the control flags.
+The message is received when relaxed atomic loads of all the parts agree on their control flag.
+This way, the payload is reconstructed intact, even in the absence of any guarantees from the memory model stronger than relaxed atomics.
+
+In particular, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state.
 \begin{figure}[h!]

### 2025-02-08 paper.tex
Remove dot product hack, make Clang recommended compiler.

@@
   \item CMake 3.10.2+ or another C++ build tool
-  \item C++17-compliant compiler
+  \item C++17-compliant compiler (Clang recommended)
   \item Python for fetching dependencies
@@
 \subsection{Building}
-Although native WebGPU is supported on various operating systems—most notably Windows, macOS, and Linux—this section will focus on building it in Unix-like environments for the sake of brevity, using CMake as the example build tool.
+Although native WebGPU is supported on various operating systems—most notably Windows, macOS, and Linux—this section will focus on building it in Unix-like environments for the sake of brevity, using CMake as the example build tool, and Clang++ as the example compiler.
 
@@
 
-  \item Build and install Dawn with your build tool and compiler of choice:
+  \item Build and install Dawn:
   \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
@@
   -DTINT_BUILD_TESTS=OFF \
-  -DTINT_BUILD_CMD_TOOLS=OFF
+  -DTINT_BUILD_CMD_TOOLS=OFF \
+  -DCMAKE_CXX_COMPILER=clang++
 
@@
 cmake -S . -B out/Release \
-  -DCMAKE_BUILD_TYPE=Release
+  -DCMAKE_BUILD_TYPE=Release \
+  -DCMAKE_CXX_COMPILER=clang++
   \end{lstlisting}

### 2025-02-08 artifact/Dawn/main.cpp

@@ -670,9 +670,9 @@ uint32_t Memcpy(const TestArgs& args, wgpu::CommandEncoder* comEncoder) {
     const uint32_t passCount = 1;
     if (args.shouldTime) {
         SetComputePassTimed(args.shaders.memcpy, comEncoder, args.gpu.querySet,
-                            args.workTiles, 0);
+                            1024, 0);
     } else {
-        SetComputePass(args.shaders.memcpy, comEncoder, args.workTiles);
+        SetComputePass(args.shaders.memcpy, comEncoder, 1024);
     }
     return passCount;
 }

### 2025-02-08 artifact/WGPU/src/main.rs

@@ -613,7 +613,7 @@ impl PassTrait for MemcpyPass {
             &tester.gpu_context.query_set,
             &tester.gpu_shaders.memcpy,
             com_encoder,
-            tester.work_tiles,
+            1024,
             0u32,
         );
     }

### 2025-02-10 paper.tex
Add CUDA, because cross-API comparison is extremely deceiving. Revert grid stride Memcpy.

@@
 This section reviews three key approaches---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---analyzing their handling of inter-workgroup dependencies and efficiency in minimizing global memory access while maximizing bandwidth utilization.
-\thomas{I have written this paragraph like twenty times now, and it's still odious.} \raph{It's fine.
-Feel free to delete these notes in next round.}
 
@@
 Breitbart~\cite{10.5555/2031978.2032029} applied Gupta et al.'s \emph{Persistent-Threads}~\cite{gupta2012} to scan, eliminating kernel launch overheads.
-Recognizing that raking naturally aligns with this paradigm, Breitbart's method dynamically \emph{discovers} the workgroup occupancy and launches exactly $o$ kernels.
+Recognizing that raking naturally aligns with this paradigm, Breitbart's method statically \emph{discovers} the workgroup occupancy and launches exactly $o$ kernels.
 Since multiprocessors are not oversubscribed, workgroups persist for the entire kernel duration, enabling synchronization at an atomics-based inter-workgroup barrier without context switching.
@@
 
-\raph{I took out the statement about cache underutilization because I'm not sure it's true (L1 cache is shared by all subgroups in a multiprocessor) and I don't think it's essential to our argument.
-Also deleting John's comment; I don't think this is a case where we need to prove novelty over a previous publication, instead we just subsume it.}
-
 \section{Decoupled Fallback}
@@
 The key advance is to utilize the parallelism available in the workgroup to perform the fallback computation, rather than performing it sequentially.
-To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained-Propagation} logic is performed by a single thread.
-
-\raph{This paragraph feels to me like it can be cut, but I won't insist on it.} Like \emph{Scalar-Fallback}, we choose \emph{mutual exclusion} as our approach to preclude deadlocks.
-\emph{Dependency relationships} and \emph{no preemption} are immutable characteristics of scan and the GPU programming model, eliminating them as viable avenues for breaking deadlocking.
-While it may be possible to break \emph{circular ordering} by designing tile sizes such that workgroups are mapped one-to-one to multiprocessors, doing so portably is infeasible.
-Moreover, vendor guidance on progress guarantees among multiprocessors is also lacking, shifting rather than eliminating deadlock risks.
-Removing \emph{resource holding}---allowing workgroups to discard \emph{blocked} work tiles---is theoretically possible but strictly inferior to removing \emph{mutual exclusion}, as it incurs $t$ extra memory operations (discard-read-read instead of read).
+To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained-Propagation} logic is performed by a single thread\thomas{Is this clearer as "single subgroup?"}.
+
 \begin{algorithm}
@@
 The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it.
-This primitive is well studied in the context of memory models.
+This primitive is well studied in the context of memory models~\cite{}.
 \raph{The earliest publication of the MP litmus test appears to be section 2.1 of "Intel 64 Architecture Memory Ordering White Paper".
@@
 
-A key innovation of this paper is to \emph{split} the reduced values into multiple atomics, each carrying its own copy of the control flags.
-The message is received when relaxed atomic loads of all the parts agree on their control flag.
+A key innovation of this paper is to \emph{split} the reduced values into multiple atomics, each carrying its own copy of the tile state.
+The message is received when relaxed atomic loads of all the parts agree on their tile state.
 This way, the payload is reconstructed intact, even in the absence of any guarantees from the memory model stronger than relaxed atomics.
@@
 Instead, we encode the tile state in the most significant bits of the atomic value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}.
-This allows an atomic \emph{maximum}, which successfully updates to a more advanced state but prevents a state regression, to correctly perform updates.
+This allows an \texttt{atomicMaximum}, which successfully updates to a more advanced state but prevents a state regression, to correctly perform updates.
 
@@
     \toprule
-    Vendor & Device Name & FPG & Occupancy & Subgroup Size \\
+    Vendor & Device Name & OBE? & Occupancy & Subgroup Size \\
     \midrule
@@
 \subsection{Performance Comparison}
+\subsubsection{Performance Versus Contemporary Implementations:}
+We first compare \emph{Decoupled Fallback} (\emph{CSDLDF}) to \emph{Decoupled Lookback} (\emph{CSDL}) in NVIDIA’s \emph{CUB} library\footnote{CUDA toolkit 12.8, CUB/CCCL VERSION}~\cite{}.
+As this test is limited to CUDA, evaluation is restricted to the 2080 Super.
+Due to WebGPU’s 128MB storage buffer limit, we test input sizes from $2^{10}$ to $2^{25}$\!.
+
+To isolate the effect of \emph{Decoupled Fallback}, we implement our own \emph{CSDL} with an identical local-scan implementation.
+Our \emph{CSDL} implementation outperforms \emph{CUB} before reaching saturation, and likely due to this advantage, \emph{CSDLDF} achieves near-identical performance to \emph{CUB} despite its additional \emph{Chained-Propagation} overhead.
+While Dawn struggles, it still reaches saturation at larger inputs.
+However, cross-API comparisons remain challenging, as despite nearly identical high-level code, all implementations perform significantly better in CUDA, particularly for inputs smaller than $2^{20}$\!, where CUDA more effectively utilizes the 2080 Super’s 4MB L2 cache.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
+\end{figure}
+
+\subsubsection{Same-API Performance Versus Baseline Measures}
 To evaluate the performance of \emph{Decoupled Fallback} (\emph{CSDLDF}), we compare it against two key benchmarks: \emph{Memcopy}\textdagger{~\ref{sec:memcopy}} and \emph{Reduce then Scan}\textdagger{~\ref{sec:rts}} (\emph{RTS}).
@@
     Apple M1 Pro Max   & 200  & 25.0e9 \\
-    Nvidia 2080 Super  & 448  & 56.0e9 \\
+    Nvidia 2080 Super  & 496  & 62.0e9 \\
     AMD 7900 XT        & 800  & 100.0e9 \\

### 2025-02-10 artifact/CUDA/main.cu

@@ -1,0 +1,310 @@
+#include <cuda_runtime.h>
+
+#include <cstdint>
+#include <cstdio>
+#include <cub/cub.cuh>
+#include <fstream>
+#include <iostream>
+#include <string>
+#include <vector>
+
+#include "CSDL.cuh"
+#include "CSDLDF.cuh"
+
+#define WARPS 8
+#define UINT4_PER_THREAD 4
+#define checkCudaError()                                         \
+    {                                                            \
+        cudaError_t err = cudaGetLastError();                    \
+        if (err != cudaSuccess) {                                \
+            printf("CUDA Error: %s\n", cudaGetErrorString(err)); \
+        }                                                        \
+    }
+
+__global__ void InitOne(uint32_t* scan, uint32_t size) {
+    const uint32_t increment = blockDim.x * gridDim.x;
+    for (uint32_t i = threadIdx.x + blockIdx.x * blockDim.x; i < size; i += increment) {
+        scan[i] = 1;
+    }
+}
+
+__global__ void ValidateInclusive(uint32_t* scan, uint32_t* errCount, uint32_t size) {
+    const uint32_t increment = blockDim.x * gridDim.x;
+    for (uint32_t i = threadIdx.x + blockIdx.x * blockDim.x; i < size; i += increment) {
+        if (scan[i] != i + 1) {
+            atomicAdd(&errCount[0], 1);
+        }
+    }
+}
+
+bool DispatchValidateInclusive(uint32_t* scanOut, uint32_t* err, uint32_t size) {
+    uint32_t errCount[1] = {0xffffffff};
+    cudaMemset(err, 0, sizeof(uint32_t));
+    ValidateInclusive<<<256, 256>>>(scanOut, err, size);
+    cudaMemcpy(&errCount, err, sizeof(uint32_t), cudaMemcpyDeviceToHost);
+    return !errCount[0];
+}
+
+void WriteVectorToCSV(const std::vector<float>& data, const std::string& filename) {
+    std::ofstream csvFile(filename);
+    if (!csvFile.is_open()) {
+        std::cerr << "Error: Could not open file " << filename << " for writing." << std::endl;
+        return;
+    }
+
+    csvFile << "throughput" << std::endl;
+    for (size_t i = 0; i < data.size(); ++i) {
+        csvFile << data[i] << std::endl;
[…]

### 2025-02-10 artifact/Dawn/main.cpp

@@ -670,9 +670,9 @@ uint32_t Memcpy(const TestArgs& args, wgpu::CommandEncoder* comEncoder) {
     const uint32_t passCount = 1;
     if (args.shouldTime) {
         SetComputePassTimed(args.shaders.memcpy, comEncoder, args.gpu.querySet,
-                            1024, 0);
+                            args.workTiles, 0);
     } else {
-        SetComputePass(args.shaders.memcpy, comEncoder, 1024);
+        SetComputePass(args.shaders.memcpy, comEncoder, args.workTiles);
     }
     return passCount;
 }
@@ -762,7 +762,7 @@ void Run(std::string testLabel, const TestArgs& args,
                 const uint64_t t = GetTime(args.gpu, &args.buffs, passCount);
                 totalTime += t;
                 if (args.shouldRecord) {
-                    data.time[i - 1] = static_cast<double>(args.size) / t;
+                    data.time[i - 1] = static_cast<double>(t);
                 }
             }
 
@@ -844,6 +844,7 @@ enum TestType {
     Csdl,
     Csdldf,
     Full,
+    Size,
 };
 
 auto printUsage = []() {
@@ -872,6 +873,8 @@ int main(int argc, char* argv[]) {
         testType = Csdldf;
     } else if (testTypeStr == "full") {
         testType = Full;
+    } else if (testTypeStr == "size") {
+        testType = Size;
     } else {
         printUsage();
         return EXIT_FAILURE;
@@ -960,6 +963,20 @@ int main(int argc, char* argv[]) {
                     CSDLDFSimulate);
             }
             break;
+        case Size:
+            for (uint32_t i = 10; i <= 25; ++i) {
+                const uint32_t currentSize = 1u << i;
+                const uint32_t currentWorkTiles =
+                    (currentSize + PART_SIZE - 1) / PART_SIZE;
+                args.size = currentSize;
+                args.workTiles = currentWorkTiles;
+                InitializeUniforms(gpu, &buffs, currentSize, currentWorkTiles,
+                                   0);
+                Run(deviceName + "CSDLDF_Size_" + std::to_string(currentSize),
+                    args, CSDLDF);
+            }
+            break;
+
         default:
             break;
     }

### 2025-02-10 artifact/WGPU/src/main.rs

@@ -613,7 +613,7 @@ impl PassTrait for MemcpyPass {
             &tester.gpu_context.query_set,
             &tester.gpu_shaders.memcpy,
             com_encoder,
-            1024,
+            tester.work_tiles,
             0u32,
         );
     }
@@ -876,7 +876,7 @@ impl Tester {
                     self.gpu_buffers.timestamp_readback.unmap();
                     total_time += t;
                     if args.should_record {
-                        data.time[(i - 1) as usize] = self.size as f64 / t as f64;
+                        data.time[(i - 1) as usize] = t as f64;
                     }
                 }
 

### 2025-02-10 paper.tex
Run the CUDA benches cold and individually.

@@
 The key advance is to utilize the parallelism available in the workgroup to perform the fallback computation, rather than performing it sequentially.
-To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained-Propagation} logic is performed by a single thread\thomas{Is this clearer as "single subgroup?"}.
+To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained-Propagation} logic is typically performed by the first subgroup.
 
@@
 We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$.
-Through testing, \john{across all devices (?)} we find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks.
+Through testing, \john{across all devices (?)} \thomas{Only rigorously tested on the 2080, everything else was very light.}we find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks.
 As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4.
 Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@.
-For all tests, we use an input size of $2^{25}$ \john{It would be worthwhile, I think, to run a test against CUB where the x axis is problem size and the y axis is throughput.
-A reviewer might think we're hiding something if we only report results for one (large) input size.
-I imagine that our performance will look similar to CUB's at all input sizes and that is the result we want.} and an inclusive prefix sum.
+For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum.
 Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities.
@@
 \subsubsection{Performance Versus Contemporary Implementations:}
-We first compare \emph{Decoupled Fallback} (\emph{CSDLDF}) to \emph{Decoupled Lookback} (\emph{CSDL}) in NVIDIA’s \emph{CUB} library\footnote{CUDA toolkit 12.8, CUB/CCCL VERSION}~\cite{}.
-As this test is limited to CUDA, evaluation is restricted to the 2080 Super.
-Due to WebGPU’s 128MB storage buffer limit, we test input sizes from $2^{10}$ to $2^{25}$\!.
-
-To isolate the effect of \emph{Decoupled Fallback}, we implement our own \emph{CSDL} with an identical local-scan implementation.
-Our \emph{CSDL} implementation outperforms \emph{CUB} before reaching saturation, and likely due to this advantage, \emph{CSDLDF} achieves near-identical performance to \emph{CUB} despite its additional \emph{Chained-Propagation} overhead.
-While Dawn struggles, it still reaches saturation at larger inputs.
-However, cross-API comparisons remain challenging, as despite nearly identical high-level code, all implementations perform significantly better in CUDA, particularly for inputs smaller than $2^{20}$\!, where CUDA more effectively utilizes the 2080 Super’s 4MB L2 cache.
+We first compare \emph{Decoupled Fallback} (\emph{CSDLDF}) to the \emph{Decoupled Lookback} (\emph{CSDL}) implementation found in NVIDIA’s \emph{CUB} library, using the 2080 Super only.
+To assess the impact of differing APIs, we also implement \emph{CSDLDF} in CUDA.\footnote{Due to WebGPU’s 128MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.}
+
+Cross-API comparisons remain challenging, as despite being nearly identical at a high level, the CUDA implementation consistently outperforms its WebGPU counterpart.
+Additionally, the performance cost of \emph{Decoupled Fallback} is most evident in our CUDA implementation when the entire input fits within the 2080 Super’s 4MB L2 cache, where increased synchronization and compute overhead slightly reduce throughput.
+However, at larger input sizes, where memory accesses extend beyond the cache, \emph{CSDLDF} achieves parity with \emph{CUB}.
 
@@
   \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
+  \caption{Performance versus \emph{CUB} (NVIDIA 2080 Super)}
 \end{figure}
@@
 \subsubsection{Same-API Performance Versus Baseline Measures}
-To evaluate the performance of \emph{Decoupled Fallback} (\emph{CSDLDF}), we compare it against two key benchmarks: \emph{Memcopy}\textdagger{~\ref{sec:memcopy}} and \emph{Reduce then Scan}\textdagger{~\ref{sec:rts}} (\emph{RTS}).
+We next compare \emph{CSDLDF} against two key benchmarks: \emph{Memcpy}\textdagger{~\ref{sec:memcpy}} and \emph{Reduce then Scan}\textdagger{~\ref{sec:rts}} (\emph{RTS}).
 \emph{RTS} serves as a baseline to demonstrate the performance gains of our approach over the previous FPG-constrained technique.
-\emph{Memcopy} is included as a point of reference, as it shares the same $2n$ global memory movement as \emph{Decoupled Fallback}, without any compute overhead and thus is guaranteed to be memory bandwidth bound.
-\john{Nit: I would use \texttt{memcpy} as the name of this benchmark to match how it's actually called in C etc.}
+\emph{Memcpy} is included as a point of reference, as it shares the same $2n$ global memory movement as \emph{Decoupled Fallback}, and without compute overhead, is guaranteed to be memory bandwidth bound.
 
@@
     \midrule
-    H-Mean \emph{Memcopy} (32-bit ele/sec)  & 1.498e9   & 106.663e9 & 54.203e9  & 54.203e9  & 71.190e9  \\
+    H-Mean \emph{Memcpy} (32-bit ele/sec)  & 1.498e9   & 106.663e9 & 54.203e9  & 54.203e9  & 71.190e9  \\
     H-Mean \emph{CSDLDF} (32-bit ele/sec)   & 1.487e9   & 112.430e9 & 53.256e9  & 53.256e9  & 95.737e9  \\
     H-Mean \emph{RTS} (32-bit ele/sec)      & 1.060e9   & 82.302e9  & 35.709e9  & 35.709e9  & 62.868e9  \\
-    \emph{CSDLDF}/\emph{Memcopy} (\%)       & 99.2\%    & 105.4\%   & 98.3\%    & 98.3\%    & 134.5\%   \\
+    \emph{CSDLDF}/\emph{Memcpy} (\%)       & 99.2\%    & 105.4\%   & 98.3\%    & 98.3\%    & 134.5\%   \\
     \emph{CSDLDF}/\emph{RTS} (\%)          & 140.2\%   & 136.6\%   & 149.1\%   & 149.1\%   & 152.3\%   \\
@@
   \caption{MALI ABSOLUTE SPEED WRONG!, \emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, Input Size $\mathbf{2^{25}}$.
-\john{I would label metrics as ``H-Mean \emph{Memcopy} (32-bit ele/sec $\cdot 10^9$)'' and then leave the exponents out of the main table.}\john{Separately, h-mean is the right summary statistic \emph{if} the distribution is Gaussian, which it almost certainly is not.
+\john{I would label metrics as ``H-Mean \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)'' and then leave the exponents out of the main table.}\john{Separately, h-mean is the right summary statistic \emph{if} the distribution is Gaussian, which it almost certainly is not.
 Like most measurements of this type, it's much more likely log-normal.
-I am influenced here by \url{https://lemire.me/blog/2023/04/06/are-your-memory-bound-benchmarking-timings-normally-distributed/}.}\john{I note Apple and NVIDIA timings are identical, but figure you know that.}}
+I am influenced here by \url{https://lemire.me/blog/2023/04/06/are-your-memory-bound-benchmarking-timings-normally-distributed/}.}}
   \label{tab:results}
@@
 As shown in Table~\ref{tab:results}, with the exception of the ARM Mali, \emph{CSDLDF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices.
-Since \emph{Memcopy} is purely memory bandwidth bound, \emph{CSDLDF} should ideally match its throughput.
-This holds across most devices, except for the AMD 7900 XT, where \emph{CSDLDF} slightly outperforms \emph{Memcopy}, suggesting improved memory structuring due to compute operations.
-\john{I have no idea what this previous phrase means.} On the ARM Mali, \emph{CSDLDF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%.
-However, as its performance is nearly identical to \emph{Memcopy}, this indicates that the implementation is memory-bandwidth-bound, limiting further gains.
-
-\subsection{How (un)fair is fair?}
-Show the stats of each device. How many lookbacks?  How many fallbacks initiated? How many successful fallbacks?
+Since \emph{Memcpy} is purely memory bandwidth bound, \emph{CSDLDF} should ideally match its throughput.
+This holds across most devices, except for the AMD 7900 XT, where \emph{CSDLDF} outperforms \emph{Memcpy}, suggesting that the additional compute latency counterintuitively improves memory access patterns by reducing paging.
+\john{I have no idea what this previous phrase means.} \thomas{In my head, if you read from location A, and then write immediately to B, you need room in your cache for A and B.
+If you read, do just the right amount of work, then write B, you only need to hold A or B?
+Not sure if this is correct.} On the ARM Mali, \emph{CSDLDF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%.
+However, as its performance is nearly identical to \emph{Memcpy}, this indicates that the implementation is memory-bandwidth-bound, limiting further gains.
+
+\subsection{Decoupled Fallback Statistics}
+To better understand how devices with differing schedulers interact with \emph{CSDLDF}, we also collect a number of statistics during its execution: 
+\begin{itemize}
+  \item \textbf{Spins per Workgroup:}
+  \item \textbf{Lookback Length per Workgroup:}
+  \item \textbf{Total Fallbacks Initiated:}
+  \item \textbf{Total Successful Insertions:}
+\end{itemize}
 
@@
 \section{Artifact}
-\subsection{Baseline Comparisons: \emph{Memcopy}}
-\label{sec:memcopy}
+\subsection{Baseline Comparisons: \emph{Memcpy}}
+\label{sec:memcpy}
 \begin{table}[h!]
@@
 \end{table}
-Although all backend APIs support \emph{timestamp queries inside encoders}, Apple GPUs do not, so we are forced to implement our own \emph{Memcopy} kernel to measure memory bandwidth, as opposed to simply timing an API's copy buffer operation.
+Although all backend APIs support \emph{timestamp queries inside encoders}, Apple GPUs do not, so we are forced to implement our own \emph{Memcpy} kernel to measure memory bandwidth, as opposed to simply timing an API's copy buffer operation.
 It is well known that \emph{observed} peak memory bandwidth, tends to diverge from vendor-listed peak bandwidth, so for completeness we list these figures above.
@@
   \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
-cd dawn
 cmake -S . -B out/Release \

### 2025-02-10 artifact/CUDA/main.cu

@@ -8,7 +8,6 @@
 #include <string>
 #include <vector>
 
-#include "CSDL.cuh"
 #include "CSDLDF.cuh"
 
 #define WARPS 8
@@ -195,87 +194,28 @@ void BatchTimingCSDLDF(uint32_t size, uint32_t batchCount, uint32_t* scanIn, uin
     cudaEventDestroy(stop);
 }
 
-void BatchTimingCSDL(uint32_t size, uint32_t batchCount, uint32_t* scanIn, uint32_t* scanOut,
-                     uint32_t* err, bool shouldRecord, std::string csvName) {
-    std::vector<float> time;
-    if (shouldRecord) {
-        time.resize(batchCount);
-    }
-
-    printf("Beginning Decoupled Lookback inclusive batch timing test at:\n");
-    printf("Size: %u\n", size);
-    printf("Test batch size: %u\n", batchCount);
-
-    const uint32_t k_csdlThreads = WARPS * LANE_COUNT;
-    const uint32_t k_partitionSize = k_csdlThreads * UINT4_PER_THREAD * 4;
-    const uint32_t k_vecSize = size / 4;  // Assume multiple of 4
-
-    uint32_t* index;
-    uint64_t* threadBlockReduction;
-    const uint32_t threadBlocks = (size + k_partitionSize - 1) / k_partitionSize;
-    cudaMalloc(&index, sizeof(uint32_t));
-    cudaMalloc(&threadBlockReduction, threadBlocks * sizeof(uint64_t));
-
-    cudaEvent_t start;
-    cudaEvent_t stop;
-    cudaEventCreate(&start);
-    cudaEventCreate(&stop);
-
-    uint32_t testsPassed = 0;
-    float totalTime = 0.0f;
-    for (uint32_t i = 0; i <= batchCount; ++i) {
-        InitOne<<<256, 256>>>(scanIn, size);
-        cudaDeviceSynchronize();
-        cudaEventRecord(start);
-        cudaMemset(index, 0, sizeof(uint32_t));
-        cudaMemset(threadBlockReduction, 0, threadBlocks * sizeof(uint64_t));
-        ChainedScanDecoupledLookback::CSDLInclusive<WARPS, UINT4_PER_THREAD>
-            <<<threadBlocks, k_csdlThreads>>>(scanIn, scanOut, threadBlockReduction, index,
-                                              k_vecSize);
-        cudaEventRecord(stop);
-        cudaEventSynchronize(stop);
-
[…]

### 2025-02-10 paper.tex
porting my comments from pull req, also plugged in relevant cites

--- 2025-02-10 paper.tex
@@
 \raph{I think "NVIDIA Tesla: A Unified Graphics and Computing Architecture" Lindholm 2008 might be a good choice for the above cite.}
+\john{I have historically used that Lindholm reference for the GPU hardware, but for the programming model I have used Nickolls, ``Scalable parallel programming with CUDA'' (ACM Queue, Mar/Apr 2008).}
 
@@
 Also, taking another look at "Specifying and Testing," I notice that they use "livelock" but not deadlock to describe situations where system progress depends on FPG.
-I'm not finding a cite for this now, but if memory serves, a deadlock condition is one in which no progress is possible under *any* scheduling condition, while in livelock there is some scheduler condition that could resolve the livelock, but it is not resolved under all such conditions.}
+I'm not finding a cite for this now, but if memory serves, a deadlock condition is one in which no progress is possible under *any* scheduling condition, while in livelock there is some scheduler condition that could resolve the livelock, but it is not resolved under all such conditions.} \john{The reason that I ask this question is that I take the perspective of a reviewer who decides, oh, I can solve this FPG problem by doing something simple, like ``scheduling only one workgroup per processor''.
+The reviewer, of course, doesn't actually implement this, but it totally kills the paper when a reviewer gets it into his/her head that there is another way to go about this and why didn't we study THAT way, reject, go back and incorporate my amazing idea.
+We need to explicitly close the door emphatically in the paper to obvious, simple, unworkable solutions; we can't give reviewers any headroom to substitute their own judgment when we know better.}
 
@@
 I put all this in a comment but think we can say it in the paper.}
+\john{My point was not about the relationship between a spin count and false positives, it was that the draft specifically said that exposing the maximum spin count to the programmer is what minimized false positives.
+It's not the exposing that minimizes false positives.}
 
@@
 For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
-  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking Kogge-Stone~\cite{} subgroup scans,
+  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking Kogge-Stone~\cite{5009159} subgroup scans,
         allowing the subgroup-block scan to remain barrier-free and in registers.
@@
 The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches.
-To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{} scan network embedded with Kogge-Stone~\cite{} subgroup scans to perform the scan across the local spine.
+To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{10.1145/322217.322232} scan network embedded with Kogge-Stone~\cite{5009159} subgroup scans to perform the scan across the local spine.
 Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts.

### 2025-02-12 paper.tex
Add Apple data, rescale graphics.

@@
 Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency.
-(one sentence on thread hierarchy maybe).
-As comprehensive descriptions of the GPU programming model~\cite{} already exist, we focus on the aspects most relevant to this work: scheduling and synchronization.
-
-\raph{I think "NVIDIA Tesla: A Unified Graphics and Computing Architecture" Lindholm 2008 might be a good choice for the above cite.}
-\john{I have historically used that Lindholm reference for the GPU hardware, but for the programming model I have used Nickolls, ``Scalable parallel programming with CUDA'' (ACM Queue, Mar/Apr 2008).}
+As comprehensive descriptions of the GPU programming model~\cite{10.1145/1365490.1365500} already exist, we focus on the aspects most relevant to this work: scheduling and synchronization.
 
@@
 This shifts dependency resolution from explicit synchronization via barrier-like structures to implicit synchronization via access order.
-As a result, each element is read and written exactly once, achieving the theoretical minimum of $2n$.
+As a result, each element is read and written exactly once, achieving the theoretical minimum of $O(2n)$.
 
@@
 However, because the scheduler lacks FPG, it continues scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction.
-As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery}\footnote{\emph{Timeout detection and recovery} is} (TDR) and crashing the host program.
+As a result, \emph{Work Tile 1} spins indefinitely, potentially triggering \emph{timeout detection and recovery}\footnote{Operating systems have different mechanisms for handling unresponsive programs.
+We borrow here Window's terminology.
+On Windows, the Timeout Detection and Recovery system resets the GPU after 2 seconds by default.
+iOS has a timeout watchdog, which, interestingly, does not terminate a spinning kernel.
+On Linux, the behavior depends on the GPU driver.} (TDR) and crashing the host program.
 
@@
 
-In particular, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state.
 \begin{figure}[h!]
@@
 
-To accommodate the split values during \emph{Chained Propagation}, we use two or four threads from the first subgroup, leveraging subgroup \emph{ballot} to verify state consistency, and subgroup \emph{shuffle} to join and split reductions.
+In particular, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state.
+Although this could also be accomplished with shared memory (and barriers), as an optimization, we use two or four threads from the first subgroup, leveraging subgroup \emph{ballot} to verify state consistency, and subgroup \emph{shuffle} to join and split reductions.
 With one thread per 16 bits and a minimum subgroup size of four in WebGPU, this method supports both 32-bit and 64-bit values, enabling 64-bit scans without 64-bit atomics or fences.\footnote{Although native 64-bit values are not currently supported by the WebGPU specification, their future inclusion is anticipated.}
@@
     \toprule
-    Vendor & Device Name & OBE? & Occupancy & Subgroup Size \\
+    Vendor & Device Name & OBE & Occupancy & Subgroup Size \\
     \midrule
@@
     ARM    & Mali-G78 MP20          & \textbf{Yes} & 40  & 16  \\
-    Apple  & M1 Pro Max             & \textbf{No}  & 64  & 32  \\
+    Apple  & M3                     & \textbf{No}  & 80  & 32  \\
+    Apple  & M1 Max                 & \textbf{No}  & 160 & 32  \\
     NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192 & 32  \\
@@
 \subsubsection{Performance Versus Contemporary Implementations:}
-We first compare \emph{Decoupled Fallback} (\emph{CSDLDF}) to the \emph{Decoupled Lookback} (\emph{CSDL}) implementation found in NVIDIA’s \emph{CUB} library, using the 2080 Super only.
+We first compare \emph{Decoupled Fallback} (\emph{CSDLDF}) to the \emph{Decoupled Lookback} (\emph{CSDL}) implementation found in NVIDIA’s CUB library, using the 2080 Super only.
 To assess the impact of differing APIs, we also implement \emph{CSDLDF} in CUDA.\footnote{Due to WebGPU’s 128MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.}
 
-Cross-API comparisons remain challenging, as despite being nearly identical at a high level, the CUDA implementation consistently outperforms its WebGPU counterpart.
+Cross-API comparisons remain challenging, as despite being nearly identical at a high level, the CUDA implementation consistently outperforms its Dawn counterpart.
 Additionally, the performance cost of \emph{Decoupled Fallback} is most evident in our CUDA implementation when the entire input fits within the 2080 Super’s 4MB L2 cache, where increased synchronization and compute overhead slightly reduce throughput.
-However, at larger input sizes, where memory accesses extend beyond the cache, \emph{CSDLDF} achieves parity with \emph{CUB}.
+However, at larger input sizes, where memory accesses extend beyond the cache, \emph{CSDLDF} achieves parity with CUB.
 
@@
   \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
-  \caption{Performance versus \emph{CUB} (NVIDIA 2080 Super)}
+  \caption{Performance versus CUB (NVIDIA 2080 Super)}
 \end{figure}
@@
   \centering
-  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r}
+  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r r}
     \toprule
-    Metric & Intel HD620 & ARM Mali-G78 MP20 & Apple M1 Pro Max & Nvidia 2080 Super & AMD 7900 XT \\
+    Metric & Intel HD620 & ARM Mali-G78 MP20 & Apple M3 & Apple M1 Max & Nvidia 2080 Super & AMD 7900 XT \\
     \midrule
-    H-Mean \emph{Memcpy} (32-bit ele/sec)  & 1.498e9   & 106.663e9 & 54.203e9  & 54.203e9  & 71.190e9  \\
-    H-Mean \emph{CSDLDF} (32-bit ele/sec)   & 1.487e9   & 112.430e9 & 53.256e9  & 53.256e9  & 95.737e9  \\
-    H-Mean \emph{RTS} (32-bit ele/sec)      & 1.060e9   & 82.302e9  & 35.709e9  & 35.709e9  & 62.868e9  \\
-    \emph{CSDLDF}/\emph{Memcpy} (\%)       & 99.2\%    & 105.4\%   & 98.3\%    & 98.3\%    & 134.5\%   \\
-    \emph{CSDLDF}/\emph{RTS} (\%)          & 140.2\%   & 136.6\%   & 149.1\%   & 149.1\%   & 152.3\%   \\
-    Avg Spins per Wg    & 0.306    & 1.380    & 0.418    & 0.418    & 0.133    \\
-    Avg Lookback Length per Wg  & 1.725    & 3.510    & 13.139   & 13.139   & 20.867   \\
-    Avg Total Fallbacks Initiated      & 260.364   & 2202.045  & 88.384   & 88.384   & 10.127   \\
-    Avg Total Successful Insertions    & 6.084     & 57.875    & 1.205    & 1.205    & 1.001    \\
+    Avg \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)  & 1.50   & 2.48  & 10.25  & 38.62  & 54.39  & 72.28  \\
+    Avg \emph{CSDLDF} (32-bit ele/sec $\cdot 10^9$)   & 1.49   & 2.79  & 10.22  & 38.84  & 53.29  & 94.60  \\
+    Avg \emph{RTS} (32-bit ele/sec $\cdot 10^9$)      & 1.06   & 2.02  & 6.81   & 25.13  & 35.90  & 61.49  \\
+    \emph{CSDLDF}/\emph{Memcpy}       & 99.1\%    & 112.8\%  & 99.8\%    & 100.6\%   & 98.0\%    & 130.9\%   \\
+    \emph{CSDLDF}/\emph{RTS}          & 140.2\%   & 138.5\%  & 150.2\%   & 154.5\%   & 148.4\%   & 153.8\%   \\
+    Avg Spins per Wg    & 0.305    & 1.477    & 0.606    & 0.604    & 0.419    & 0.148    \\
+    Avg Lookback Length per Wg  & 1.726    & 3.537    & 3.415    & 10.782   & 13.046   & 20.722   \\
+    Avg Total Fallbacks Initiated      & 254.662   & 2504.704  & 365.546   & 990.962   & 86.400   & 9.308   \\
+    Avg Total Successful Insertions    & 6.170     & 122.436   & 4.612     & 0.012    & 0.284    & 0.020    \\
     \bottomrule
   \end{tabular*}
-  \caption{MALI ABSOLUTE SPEED WRONG!, \emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, Input Size $\mathbf{2^{25}}$.
-\john{I would label metrics as ``H-Mean \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)'' and then leave the exponents out of the main table.}\john{Separately, h-mean is the right summary statistic \emph{if} the distribution is Gaussian, which it almost certainly is not.
-Like most measurements of this type, it's much more likely log-normal.
-I am influenced here by \url{https://lemire.me/blog/2023/04/06/are-your-memory-bound-benchmarking-timings-normally-distributed/}.}}
+  \caption{\emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, Input Size $\mathbf{2^{25}}$}
   \label{tab:results}
@@
 
-As shown in Table~\ref{tab:results}, with the exception of the ARM Mali, \emph{CSDLDF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices.
-Since \emph{Memcpy} is purely memory bandwidth bound, \emph{CSDLDF} should ideally match its throughput.
-This holds across most devices, except for the AMD 7900 XT, where \emph{CSDLDF} outperforms \emph{Memcpy}, suggesting that the additional compute latency counterintuitively improves memory access patterns by reducing paging.
-\john{I have no idea what this previous phrase means.} \thomas{In my head, if you read from location A, and then write immediately to B, you need room in your cache for A and B.
-If you read, do just the right amount of work, then write B, you only need to hold A or B?
-Not sure if this is correct.} On the ARM Mali, \emph{CSDLDF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%.
+As shown in Table~\ref{tab:results}, \emph{CSDLDF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices—closely approaching the theoretically expected $O(3n)/O(2n)$ 50\% speedup, except on the ARM Mali.
+Since \emph{Memcpy} is purely memory-bandwidth-bound, \emph{CSDLDF} should ideally match its throughput.
+This holds across most devices, except for the AMD 7900 XT and ARM Mali, where \emph{CSDLDF} outperforms \emph{Memcpy}, suggesting that the additional compute latency counterintuitively improves memory access patterns by reducing paging.
+\john{I have no idea what this previous phrase means.} \thomas{In my head, if you read from location A, and then write immediately to B, you need room in your cache for A and B because the workgroup doesn't execute in lockstep.
+If you read, do just the right amount of work, then write B, you only need to hold A or B because the compute latency ensures that all the work from A is done before you get to B.
+Not sure if this is correct.}On the ARM Mali, \emph{CSDLDF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%.
 However, as its performance is nearly identical to \emph{Memcpy}, this indicates that the implementation is memory-bandwidth-bound, limiting further gains.
@@
 \subsection{Decoupled Fallback Statistics}
-To better understand how devices with differing schedulers interact with \emph{CSDLDF}, we also collect a number of statistics during its execution:
-\begin{itemize}
-  \item \textbf{Spins per Workgroup:}
-  \item \textbf{Lookback Length per Workgroup:}
-  \item \textbf{Total Fallbacks Initiated:}
-  \item \textbf{Total Successful Insertions:}
-\end{itemize}
+To better understand how differing devices interact with \emph{CSDLDF}, we also collect a number of statistics during its execution:
+\begin{itemize}
+  \item \textbf{Spins per Workgroup:} How many times does a workgroup make a waiting spin?
+  \item \textbf{Lookback Length per Workgroup:} How many work tiles does a workgroup traverse to complete \emph{Chained-Propagation}?
+  \item \textbf{Total Fallbacks Initiated:} What is the total number of fallback operations initiated across all workgroups?
+  \item \textbf{Total Successful Insertions:} What is total number of fallback operations that successfully update a blocking (or late) tile to \emph{Ready}?
+\end{itemize}
+Despite significantly different absolute occupancies, the number of waiting spins remains consistent across devices.
+Likewise, lookback length is stable and remains well below the theoretical occupancy bound.
+Expressed as a percentage of occupancy, it remains under 10\% for all devices, except for the Intel HD620, where its elevated value is likely an artifact of its low absolute occupancy.
+Most notably, the devices lacking FPG---Apple M3 and M1 Max---do not exhibit significantly higher successful insertion rates than the others, suggesting that true deadlocking is a rare occurrence.
 
 \subsection{Simulated Blocking}
-To demonstrate the resilience of our technique under fallback rates significantly higher than those observed organically on devices, we intentionally force fallbacks by omitting the posting of work tile reductions.
-Starting with infrequent omissions---1 in 512 workgroups---we progressively increase the rate to near-continuous levels, reaching 1 in 2 workgroups.
+To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally forcing fallbacks through the omission of work tile reductions.
+Beginning with infrequent omissions---1 in 512 work tiles---we progressively increase the rate to near-continuous levels, reaching 1 in 2 work tiles.
 \begin{figure*}[t]
@@
 
-In Figure~\ref{fig:fallbacks_initiated}, we measure the number of fallbacks proportional to the number of workgroups launched, and verify that we are indeed inducing fallbacks.
-In Figure~\ref{fig:total_spins} we measure \thomas{this should be spins per workgroup it's a better metric}.
-In Figure~\ref{fig:lookback_length} we see that.
-Finally, in Figure~\ref{fig:execution_time}, we can see that, even when \emph{half} the workgroups are deadlocking, \emph{Decoupled-Fallback} still achieves speeds of .
-.
-.
-
+As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at a 100\% fallback rate across all devices—except for the Intel HD620, where only a 25\% fallback rate is induced.
+Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%.
+While Figures~\ref{fig:fallbacks_initiated} and \ref{fig:total_spins} confirm that we are inducing fallbacks, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior.
+The finer details of this distinction are discussed next.
 \section{Discussion}
+\subsection{Limits on Speed-of-Light Performance}
+
+In true blocking scenarios, the stalled workgroup is starved of progress, but execution resources are not lost; rather, they are redistributed to other workgroups.
+As a result, we would expect lookback lengths to be longer than those observed in our simulation.
+
+\thomas{Somewhere in here is perfect for Raph's commentary on the speed of atomics on different hardware. Once we have it, we can retool this section around it.}
 
 \subsection{Trade-offs}
-
-\subsection{Limits on Speed-of-Light Performance}
-There are a number of factors which may preclude speed-of-light performance, but most salient is the fairness of the scheduling model.
-In \emph{Decoupled-Fallback}, a work tile which is late due to unfairness is indistinguishable from one which is blocking, so as a scheduler becomes increasingly unfair, it incurs an increasing number of redundant fallbacks.
-Consider a hardware with a workgroup occupancy denoted by $o$, and let $f$ represent the probability that a fallback operation occurs at a particular lookback step.
-Because the number of lookback steps is bounded by $o$, the expected number of fallback operations is approximately $fo$.
-This results in a factor of $fo$ increase in global memory reads and a factor of $fo\log{n}$ increase in work.
-
-This increased sensitivity to fairness exacerbates existing issues which negatively impact the performance of previous scan architectures, namely lack of compute, and slow atomic update latency.
-On hardware that lacks sufficient compute power to be memory-bandwidth bound, the additional work incurred by redundant fallback reductions is particularly deleterious.
-High atomic update latency further compounds the problem: as inter-workgroup communication time grows, the minimum delay before updates become visible to dependent workgroups also increases.
-This, in turn, raises the number of lookback steps that may be needed and potentially leads to redundant fallbacks.
-
-\thomas{Somewhere in here is perfect for Raph's commentary on the speed of atomics on different hardware. Once we have it, we can retool this section around it.}
+Decoupled Fallback introduces several tradeoffs that must be considered: (1) Without a memory fence, the scan cannot be performed in place and requires a separate output buffer to prevent false-positive fallback operations on post-scanned values, as the tile state may not update before the scan values.
+(2) For monoids larger than 64 bits, multiple subgroups are required, increasing synchronization overhead due to the need for shared memory and barriers.
+(3) In certain in-situ compaction algorithms where fallback work-tiles cannot be easily reduced and must be held in registers, Decoupled Fallback may reduce occupancy.
+(4) Additionally, in compute-bound in-situ compaction algorithms, the cost of fallbacks may outweigh the savings from reduced global memory traffic.
+(5) Finally, because the control flow of \emph{Decoupled Fallback} introduces barriers latency incurred during lookback cannot be hidden by workgroup computation.
+
+\subsection{Future Work}
+
+\subsection{Conclusion}
 
@@
     ARM Mali-G78 MP20 / Pixel 6a (Tensor) & 51.2             & 6.40e9          \\
-    Apple M1 Pro Max                      & 200              & 25.0e9          \\
+    Apple M3                              & 102.4            & 12.8e9          \\
+    Apple M1 Max                          & 409.6            & 51.2e9          \\
     Nvidia 2080 Super                     & 496              & 62.0e9          \\

### 2025-02-17 paper.tex
Finish Section 6 besides "Limits," Improve testing, Update CUDA plot, Add citations, Fix graphics, Prepare to rework for OBE.

@@
 Additionally, we introduce a tile state representation for \emph{Chained-Scans} that does not rely on 64-bit atomics or memory fences for correctness, along with a subgroup-size-agnostic intra-workgroup implementation.
-On an FPG-lacking device---Apple M1 Pro (8+2)---\emph{Decoupled-Fallback} achieves near-\emph{memcopy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach.
+On an FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled-Fallback} achieves near-\emph{memcopy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach.
 We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by emulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
@@
 \subsection{Evolution of Inter-Workgroup Scan Architectures}
-Contemporary GPU scans depart from early PRAM-like scans~\cite{Blelloch:1989:SAP} by leveraging the GPU memory hierarchy, which consists of progressively faster but increasingly private tiers.
+Contemporary GPU scans depart from early PRAM-like scans~\cite{Horn, Hensley, Sengupta2006, Gress} by leveraging the GPU memory hierarchy, which consists of progressively faster but increasingly private tiers.
 Efficient scans integrate fine-grained intra-workgroup (\emph{local}) strategies in shared memory and registers with coarse-grained inter-workgroup (\emph{global}) strategies in global memory.
@@
 \subsection{Ideal Scan Architecture}
-\emph{Chained Scan} and \emph{Reduce then Scan} represent two contrasting paradigms in scan design: one embraces oversubscription and dynamic scheduling, while the other relies on a fixed-launch model with explicit control over occupancy.
-By avoiding explicit inter-workgroup synchronization, \emph{Chained-Scan} naturally aligns with the oversubscription paradigm, allowing the workgroup scheduler to dynamically manage occupancy.
-In contrast, \emph{Reduce-Then-Scan} employs workgroup raking, which mandates explicit occupancy control and, in some cases, inter-workgroup synchronization.
+\emph{Chained Scan} and \emph{Reduce-Then-Scan} represent two contrasting paradigms in scan design.
+By avoiding explicit barrier-based inter-workgroup synchronization, \emph{Chained Scan} aligns with the oversubscription paradigm, delegating occupancy management to the workgroup scheduler.
+In contrast, \emph{Reduce-Then-Scan} employs workgroup raking, requiring fixed launches and explicit occupancy control, forcing developers to manually tune launch sizes or implement occupancy discovery techniques.
+
+Hardware-level considerations underscore the limitations of fixed-launches.
+On GPUs where the global memory data cache is indexed by physical addresses---and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)---fixed-launches can lead to performance degradation due to cache thrashing.
+This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups.
+Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns---or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations---may experience significant performance degradation.
+
+Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture.
+Furthermore, its single-kernel design and serial ordering make it suitable for \emph{in-situ} operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging.
+However, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in both theoretical and practical terms, it has a critical limitation: its reliance on a forward-progress guarantee.
+While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that Apple’s M-series GPUs do not \thomas{I've cut mentioning ARM here due to reclassification, but it kind of reduces the impact.
+I.e.
+"Oh it's just Apple that's missing, so this is a vendor hack?"}.
+Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
 
@@
 
-Fixed-launch approaches, such as those used in \emph{Reduce-Then-Scan}, introduce several challenges.
-They require either careful tuning of workgroup launch sizes across different architectures or an occupancy discovery mechanism, which is nontrivial to implement.
-For example, the workgroup occupancy discovery technique proposed by Sorensen et al.~\cite{sorensen2016} relies on memory fences—unsupported in WebGPU—making it infeasible in certain environments.\footnote{Our implementation includes a fenceless occupancy discovery method suitable for WebGPU, but its description is beyond the scope of this work.}
-
 \raph{"Fixed launch" means launching a fixed number of workgroups and having each workgroup process multiple blocks, generally total blocks / num workgroups?
@@
 
-Hardware-level considerations further underscore the limitations of fixed-launches.
-On GPUs where the global memory data cache is indexed by physical addresses---and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)---fixed-launches can lead to performance degradation due to cache thrashing.
-This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups.
-Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns---or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations---may experience significant performance degradation.
-
-Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture.
-By leveraging the workgroup scheduler, \emph{Chained-Scan} dynamically distributes work across architectures without requiring developer intervention.
-Furthermore, its single-kernel design and serial ordering make it suitable for \emph{in-situ} operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging.
-
-However, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in both theoretical and practical terms, it has a critical limitation: its reliance on a forward-progress guarantee.
-While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that Apple’s M-series and ARM GPUs do not.
-Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
+\thomas{Edited this section in response to above comments.
+Distinguished Chained Scan as specifically eliding barrier based sychronization.
+Linked fixed launch more clearly to manual tuning and occupancy discovery.
+Shortened section.
+OK to remove comments?}
 
@@
 Sorensen et al.~\cite{sorensen2018,sorensen2021} provide the most comprehensive analysis of GPU schedulers, classifying \emph{Chained-Scan} as FPG-reliant.
-However, their work primarily focuses on formalizing progress models rather than their direct application to specific algorithms.
-
-\raph{I think the deep-dive into the nature of deadlock can be trimmed back here, but chatting with John, such editing for concision can be done later, when we have a better sense about page count.}
-
-The challenges faced by \emph{Chained-Scan} resemble the \emph{Dining Philosophers Problem}, where competing tasks share limited resources, risking deadlock.
-Coffman et al.~\cite{10.1145/356586.356588} define four conditions necessary for deadlock:
-\begin{enumerate}
-  \item \textbf{Mutual Exclusion}: Tasks claim exclusive control of resources they require.
-  \item \textbf{Resource Holding}: Tasks hold acquired resources while waiting.
-  \item \textbf{No Preemption}: A resource cannot be removed from a task before completion.
-  \item \textbf{Circular Dependency}:
-        \begin{enumerate}
-          \item Dependency Relationships: A dependency exists such that each task holds one or more resources that are required by proceeding tasks.
-          \item Circular Ordering: A circular chain of tasks exists.
-        \end{enumerate}
-\end{enumerate}
-Applying this schema to \emph{Chained-Scan}, workgroups correspond to "tasks" and work-tile reductions to "resources." Conditions 1–3 hold due to the one-to-one mapping of work tiles to workgroups and the constraints of the GPU programming model.
-Condition 4a is satisfied by the inter-tile dependencies, leaving condition 4b---\emph{circular ordering}---as the key concern.
-
-Although the serial assignment of work tiles ensures that predecessor tiles are scheduled or completed, it does not guarantee their reductions will be available when a workgroup begins \emph{Chained Propagation}.
-This uncertainty arises because workgroups are scheduled \emph{individually}, meaning a predecessor may not have had the opportunity to compute and post its reduction.
-In such cases, the dependent workgroup spins, consuming execution resources while waiting.
-Because workgroups may share execution resources by being co-resident on the same multiprocessor\john{Question, and maybe this is worth a footnote.
+\begin{itemize}
+  \item Introduce Sorensen
+  \item Introduce Progress models
+  \item Explain why OBE + data
+  \item Explain why one-to-one mapping doesn't solve problem.
+\end{itemize}
+
+\john{Question, and maybe this is worth a footnote.
 What if we guaranteed that no multiprocessor ever had more than one workgroup scheduled to it, say by always allocating the maximum amount of workgroup memory?
@@
 But does that solve the problem?
-Or do we descend into an analogous issue with subgroup scheduling?}, without FPG, the scheduler may indefinitely prioritize the spinning workgroup over its predecessor, preventing progress and creating a circular dependency, satisfying condition 4b and resulting in deadlock.
+Or do we descend into an analogous issue with subgroup scheduling?}
 
@@
 Also, taking another look at "Specifying and Testing," I notice that they use "livelock" but not deadlock to describe situations where system progress depends on FPG.
-I'm not finding a cite for this now, but if memory serves, a deadlock condition is one in which no progress is possible under *any* scheduling condition, while in livelock there is some scheduler condition that could resolve the livelock, but it is not resolved under all such conditions.} \john{The reason that I ask this question is that I take the perspective of a reviewer who decides, oh, I can solve this FPG problem by doing something simple, like ``scheduling only one workgroup per processor''.
+I'm not finding a cite for this now, but if memory serves, a deadlock condition is one in which no progress is possible under *any* scheduling condition, while in livelock there is some scheduler condition that could resolve the livelock, but it is not resolved under all such conditions.}
+
+\john{The reason that I ask this question is that I take the perspective of a reviewer who decides, oh, I can solve this FPG problem by doing something simple, like ``scheduling only one workgroup per processor''.
 The reviewer, of course, doesn't actually implement this, but it totally kills the paper when a reviewer gets it into his/her head that there is another way to go about this and why didn't we study THAT way, reject, go back and incorporate my amazing idea.
 We need to explicitly close the door emphatically in the paper to obvious, simple, unworkable solutions; we can't give reviewers any headroom to substitute their own judgment when we know better.}
-
-For example, consider a scan operation requiring two workgroups on a GPU with a single multiprocessor.
-Both workgroups are co-resident, but the multiprocessor has only one scheduling unit, which lacks FPG\@.
-Both workgroups launch and acquire their respective work tiles.
-The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, repeatedly querying the memory address for the reduction of \emph{Work Tile 0}.
-However, because the scheduler lacks FPG, it continues scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction.
-As a result, \emph{Work Tile 1} spins indefinitely, potentially triggering \emph{timeout detection and recovery}\footnote{Operating systems have different mechanisms for handling unresponsive programs.
-We borrow here Window's terminology.
-On Windows, the Timeout Detection and Recovery system resets the GPU after 2 seconds by default.
-iOS has a timeout watchdog, which, interestingly, does not terminate a spinning kernel.
-On Linux, the behavior depends on the GPU driver.} (TDR) and crashing the host program.
 
@@
 
-  $\textit{part\_id} \gets \textit{wg\_broadcast}$\;
-
-  \textit{local\_vectors}: $\textnormal{array}\langle \textnormal{vec4}\langle u32\rangle, \textit{p} \rangle$\;
-
-  \textbf{Load}($\textit{in}$, $\textit{part\_id}$, \textit{local\_vectors})\;
-  \textbf{SubgroupRake}(\textit{local\_vectors})\;
-
-  \If{$\textnormal{highestRankLane}(\textit{threadid})$}{
-    $\textit{wg\_partials}[\textit{subgroup\_id}] \gets \textit{local\_vectors}[\textit{p} - 1]$\;
-  }
+  $\textit{tile\_id} \gets \textit{wg\_broadcast}$\;
+
+  $\textit{local\_vectors}$ \gets $\textnormal{array}\langle \textnormal{vec4}\langle u32\rangle, \textit{p} \rangle$\;
+
+  \textbf{Load}($\textit{in}$, $\textit{tile\_id}$, \textit{local\_vectors})\;
+  \textbf{SubgroupRake}($\textit{local\_vectors}$, $\textit{wg\_partials}$)\;
   \textbf{barrier()}\;
@@
   \tcp{Chained Propagation, Decoupled Fallback}
+  \textbf{PostLocalReduction}($\textit{spine}$, $\textit{tile\_id}$, $\textit{wg\_partials}$)\;
   \While{$\textit{wg\_control} = \textnormal{LOCKED}$}{
-    \textbf{Lookback}($\textit{spine}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
+    \textbf{Lookback}($\textit{spine}$, $\textit{tile\_id}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
     \textbf{barrier()}\;
@@
 
-  \textbf{Write}($\textit{out}$, $\textit{part\_id}$, \textit{local\_vectors}, $\textit{wg\_partials}$, $\textit{wg\_prev\_reduction}$)\;
+  \textbf{Write}($\textit{out}$, $\textit{tile\_id}$, $\textit{local\_vectors}$, $\textit{wg\_partials}$, $\textit{wg\_prev\_reduction}$)\;
 
@@
   \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
-  \item \textbf{Chained Propagation, Decoupled Fallback}: The first subgroup in the workgroup performs a lookback operation to acquire the reduction result of all previous workgroups, engaging in fallback operations as necessary.
+  \item \textbf{Chained Propagation, Decoupled Fallback}: The local reduction is posted to global memory.
+Then, the first subgroup in the workgroup performs a lookback operation to acquire the reduction result of all previous workgroups, engaging in fallback operations as necessary.
 Once completed, this subgroup posts the results to shared memory for the workgroup to consume.
@@
 Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking.
-Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.\footnote{Since we also require a workgroup-wide pure reduce operation, we obtain this by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}
+Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.\footnote{We may also obtain a workgroup-wide reduce operation by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}
 
@@
 The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it.
-This primitive is well studied in the context of memory models~\cite{}.
+This primitive is well studied in the context of memory models~\cite{Alglave0, MCMutants}.
 \raph{The earliest publication of the MP litmus test appears to be section 2.1 of "Intel 64 Architecture Memory Ordering White Paper".
-I'm not sure that's the best cite; the MC mutants paper cites https://arxiv.org/pdf/1207.7264} The key requirement is both control synchronization, meaning the reader does not proceed until the writer sends the message; and data synchronization, meaning that the reader reads the correct value.
+I'm not sure that's the best cite; the MC mutants paper cites https://arxiv.org/pdf/1207.7264} \thomas{Added MC Mutants + the Algave paper.
+OK?
+Or should cite more?
+Alglave has tons of papers.} The key requirement is both control synchronization, meaning the reader does not proceed until the writer sends the message; and data synchronization, meaning that the reader reads the correct value.
 The latter is especially challenging in relaxed memory models as are employed on GPUs.
@@
 However, if the tile state and data are stored in two different memory locations, then relaxed atomics are not sufficient and some other sychronization is necessary.
-On CPUs, it is common to upgrade the memory semantics from relaxed to acquire/release, which then provides the necessary guarantee.
-\raph{Here's a good place to cite the MP litmus test because that's exactly what it's about.} Unfortunately, on GPUs, acquire/release semantics are neither portable (they are only available on Vulkan, and optional until version 1.3), nor are they necessarily efficient even when they are available; often the implementation involves expensive cache flushes.
+On CPUs, it is common to upgrade the memory semantics from relaxed to acquire/release, which then provides the necessary guarantee~\cite[Section 10.2]{intel_sdm_2024} \thomas{I've cited the latest publication by Intel.
+The section is nearly identical to 2007.} .
+Unfortunately, on GPUs, acquire/release semantics are neither portable (they are only available on Vulkan, and optional until version 1.3), nor are they necessarily efficient even when they are available; often the implementation involves expensive cache flushes.
 A device-scoped memory barrier has similar problems.
@@
 We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$.
-Through testing, \john{across all devices (?)} \thomas{Only rigorously tested on the 2080, everything else was very light.}we find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks.
+We find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks.
 As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4.
@@
 We first compare \emph{Decoupled Fallback} (\emph{CSDLDF}) to the \emph{Decoupled Lookback} (\emph{CSDL}) implementation found in NVIDIA’s CUB library, using the 2080 Super only.
-To assess the impact of differing APIs, we also implement \emph{CSDLDF} in CUDA.\footnote{Due to WebGPU’s 128MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.}
+To assess the impact of differing APIs, we also provide the results of an identical \emph{CSDLDF} in CUDA.\footnote{Due to WebGPU’s 128MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.}
 
@@
   \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
-  \caption{Performance versus CUB (NVIDIA 2080 Super)}
+  \caption{(This graph can be provided with or without Memcpy speeds. I am thinking without, as it's currently very busy.) Performance versus CUB (NVIDIA 2080 Super)}
 \end{figure}
@@
 \john{I have no idea what this previous phrase means.} \thomas{In my head, if you read from location A, and then write immediately to B, you need room in your cache for A and B because the workgroup doesn't execute in lockstep.
-If you read, do just the right amount of work, then write B, you only need to hold A or B because the compute latency ensures that all the work from A is done before you get to B.
-Not sure if this is correct.}On the ARM Mali, \emph{CSDLDF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%.
+If you read, do just the right amount of work, then write B, you only need to hold A or B because the compute latency ensures that all the work from A is done before you get to B.}On the ARM Mali, \emph{CSDLDF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%.
 However, as its performance is nearly identical to \emph{Memcpy}, this indicates that the implementation is memory-bandwidth-bound, limiting further gains.
@@
 \subsection{Decoupled Fallback Statistics}
-To better understand how differing devices interact with \emph{CSDLDF}, we also collect a number of statistics during its execution:
+To better understand how devices interact with \emph{CSDLDF}, we also collect a number of statistics during its execution:
 \begin{itemize}
@@
 
-As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at a 100\% fallback rate across all devices—except for the Intel HD620, where only a 25\% fallback rate is induced.
+As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near 100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced.
 Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%.
@@
 (4) Additionally, in compute-bound in-situ compaction algorithms, the cost of fallbacks may outweigh the savings from reduced global memory traffic.
-(5) Finally, because the control flow of \emph{Decoupled Fallback} introduces barriers latency incurred during lookback cannot be hidden by workgroup computation.
+(5) Lastly, because the control flow of \emph{Decoupled Fallback} requires barriers, latency incurred during lookback cannot be hidden by workgroup computation.
 
 \subsection{Future Work}
+Sorting is one of the most notable beneficiaries of single-pass scan and, by extension, \emph{Decoupled Fallback}.
+The current state-of-the-art algorithm, Adinet and Merrill’s \emph{Onesweep}~\cite{adinets2022onesweepfastersignificantdigit}, is an 8-bit least-significant-digit radix sort, where single-pass scan minimizes data movement during digit-binning passes.
+However, due to its compute-intensive nature and large $radixDigit$-sized monoid, it is unclear whether \emph{Decoupled Fallback} maps well to sorting or if new techniques are needed.
+
+Our reexamination of \emph{Chained Scan} with dynamic assignment raises fundamental questions about GPU progress models: Can all HSA algorithms be converted to OBE via dynamic assignment?
+Are there cases, excluding atomic memory errors, where LOBE guarantees termination but dynamic assignment with OBE does not?
+\textbf{Is OBE all you need}?
 
 \subsection{Conclusion}
+In this paper, we presented \emph{Decoupled Fallback}, a single-pass \emph{Chained-Scan} capable of reaching SOL performance under all progress models and suitable for implementation within the constraints of contemporary shading languages.
+Our approach mantains minimal data movement while ensuring robust execution across diverse GPU architectures.
 

### 2025-02-17 artifact/CUDA/main.cu

@@ -50,201 +50,265 @@ void WriteVectorToCSV(const std::vector<float>& data, const std::string& filenam
         std::cerr << "Error: Could not open file " << filename << " for writing." << std::endl;
         return;
     }
-
-    csvFile << "throughput" << std::endl;
-    for (size_t i = 0; i < data.size(); ++i) {
-        csvFile << data[i] << std::endl;
+    csvFile << "time\n";  // CSV header
+    for (float val : data) {
+        csvFile << val << "\n";
     }
     csvFile.close();
     std::cout << "CSV data written to " << filename << std::endl;
 }
 
-void BatchTimingCubChainedScan(uint32_t size, uint32_t batchCount, uint32_t* scanIn,
-                               uint32_t* scanOut, uint32_t* err, bool shouldRecord,
-                               std::string csvName) {
-    std::vector<float> time;
-    if (shouldRecord) {
-        time.resize(batchCount);
-    }
+class CubTestRunner {
+public:
+    // Run the test for a given size and batch count.
+    // Returns a vector of individual iteration times (in ms), excluding the warmup.
+    std::vector<float> runTest(uint32_t size, uint32_t batchCount,
+                               uint32_t* scanIn, uint32_t* scanOut, uint32_t* err) {
+        std::vector<float> times(batchCount);
+        printf("Running CUB test for size %u, batch count %u\n", size, batchCount);
 
-    printf("Beginning CUB inclusive batch timing test at:\n");
-    printf("Size: %u\n", size);
-    printf("Test batch size: %u\n", batchCount);
-
-    void* d_temp_storage = nullptr;
-    size_t temp_storage_bytes = 0;
-    cub::DeviceScan::InclusiveSum(d_temp_storage, temp_storage_bytes, scanIn, size);
-    cudaMalloc(&d_temp_storage, temp_storage_bytes);
-
-    cudaEvent_t start;
-    cudaEvent_t stop;
-    cudaEventCreate(&start);
-    cudaEventCreate(&stop);
-
-    uint32_t testsPassed = 0;
-    float totalTime = 0.0f;
-    for (uint32_t i = 0; i <= batchCount; ++i) {
-        InitOne<<<256, 256>>>(scanIn, size);
-        cudaDeviceSynchronize();
-        cudaEventRecord(start);
+        // Warm-up: determine temporary storage size.
[…]

### 2025-02-17 artifact/Dawn/main.cpp

@@ -68,6 +68,16 @@ struct DataStruct {
     std::vector<double> lookbackLength;
     std::vector<uint32_t> fallbacksInitiated;
     std::vector<uint32_t> successfulInsertions;
+
+    DataStruct(const TestArgs& args) {
+        if(args.shouldRecord){
+            time.resize(args.batchSize);
+            totalSpins.resize(args.batchSize);
+            lookbackLength.resize(args.batchSize);
+            fallbacksInitiated.resize(args.batchSize);
+            successfulInsertions.resize(args.batchSize);
+        }
+    }
 };
 
 void GetGPUContext(GPUContext* context, uint32_t timestampCount) {
@@ -96,7 +106,7 @@ void GetGPUContext(GPUContext* context, uint32_t timestampCount) {
             adaptPromise.set_value();
         });
     std::future<void> adaptFuture = adaptPromise.get_future();
-    while (adaptFuture.wait_for(std::chrono::microseconds(100)) ==
+    while (adaptFuture.wait_for(std::chrono::nanoseconds(100)) ==
            std::future_status::timeout) {
         instance.ProcessEvents();
     }
@@ -153,7 +163,7 @@ void GetGPUContext(GPUContext* context, uint32_t timestampCount) {
             devPromise.set_value();
         });
     std::future<void> devFuture = devPromise.get_future();
-    while (devFuture.wait_for(std::chrono::microseconds(100)) ==
+    while (devFuture.wait_for(std::chrono::nanoseconds(100)) ==
            std::future_status::timeout) {
         instance.ProcessEvents();
     }
@@ -403,7 +413,7 @@ void CreateShaderFromSource(const GPUContext& gpu, const GPUBuffers& buffs,
             promise.set_value();
         });
     std::future<void> future = promise.get_future();
-    while (future.wait_for(std::chrono::microseconds(100)) ==
+    while (future.wait_for(std::chrono::nanoseconds(100)) ==
            std::future_status::timeout) {
         gpu.instance.ProcessEvents();
     }
@@ -486,7 +496,7 @@ void QueueSync(const GPUContext& gpu) {
             promise.set_value();
         });
     std::future<void> future = promise.get_future();
[…]

### 2025-02-17 paper.tex
First attempt at rewriting barrier section
Also clarify that reduce-then-scan relies on pipeline barriers.

I think more work is needed to really clearly explain the memory model requirements and reliance on forward progress guarantees, but hopefully this is a step in the right direction.

@@
 \subsubsection{Inter-Workgroup Barriers}
-The virtualization of processors introduces an emergent high-level limitation: the absence of an inter-workgroup barrier.
-As previously mentioned, workgroups cannot context switch.
-Although contemporary GPU hardware~\cite{} supports workgroup preemption for prioritizing real-time tasks or managing multi-process workloads, this functionality is deliberately excluded from the GPU programming model.
-This stems from the prohibitively high overhead in latency and storage associated with moving workgroup contexts on and off chip.
-In the oversubscription paradigm, where the number of workgroups is theoretically unbounded and typically far exceeds the available multiprocessors, every invocation of such a barrier would necessitate context-switching between all launched workgroups.
-Due to this high cost, no GPU programming environment furnishes developers with a true unbounded-launch inter-workgroup barrier.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent-Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} Instead, when inter-workgroup synchronization is necessary, developers must rely on kernel launches as synchronization points.
-However, kernel launches are suboptimal as barriers: they incur overheads due to the heterogeneous nature of the operation, and all execution context is lost between launches.
-
-\raph{I think the above is conflating memory barriers with control barriers.
-All modern GPU shading languages (since Metal 3.2) provide memory barriers with inter-workgroup (device) scope, but the above paragraph makes more sense interpreted as a control barrier.
-It might also be worth mentioning that kernel launches do not necessarily sychronize; in Vulkan it's really the pipeline barrier that does the synchronization.
-I can take a whack at re-writing this.}
+Parallel programs in general require synchronization between multiple threads (those that don't are considered embarrassingly parallel).
+A standard technique for synchronization is \emph{barriers.} GPUs provide a limited palette of barrier primitives.
+At the coarsest granularity, \emph{pipeline barriers} fully synchronize both control and memory between multiple kernel launches.
+However, pipeline barriers are extremely expensive and efficient algorithms strictly minimize their use.
+At a much finer granularity, \emph{workgroup barriers} provide reasonably efficient synchronization between the threads in a workgroup.
+However, synchronization between workgroups is more limited.
+Standard GPU programming environments provide no control barrier between workgroups.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent-Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} The availability of a memory-only barrier varies across GPU environments; it is present in Direct3D and Vulkan, but was missing in versions of Metal prior to 3.2.
+Because of its non-portability, it is also absent from the WebGPU specification.
+Even when available, the use of such memory barriers causes performance issues as they are usually implemented as a full cache flush.
 
@@
 \emph{Chained Scan} and \emph{Reduce-Then-Scan} represent two contrasting paradigms in scan design.
-By avoiding explicit barrier-based inter-workgroup synchronization, \emph{Chained Scan} aligns with the oversubscription paradigm, delegating occupancy management to the workgroup scheduler.
+By avoiding explicit pipeline barrier based inter-workgroup synchronization, \emph{Chained Scan} aligns with the oversubscription paradigm, delegating occupancy management to the workgroup scheduler.
 In contrast, \emph{Reduce-Then-Scan} employs workgroup raking, requiring fixed launches and explicit occupancy control, forcing developers to manually tune launch sizes or implement occupancy discovery techniques.
@@
 Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
-
-\raph{I'm having trouble with the above paragraph.
-By my read, Reduce-Then-Scan avoids inter-workgroup synchronization other than the pipeline barriers between kernel launches.
-Chained-Scan and Decoupled-Lookback both rely on inter-workgroup synchronization.
-Not sure why the "in some cases" qualifier either; seems to me you need synchronization except in the trivial case the entire array fits in one workgroup.}
-
-\raph{"Fixed launch" means launching a fixed number of workgroups and having each workgroup process multiple blocks, generally total blocks / num workgroups?
-I think I get the argument here but it's more subtle and not an inherent problem with RTS.
-You make a choice: either you do fixed launch and have to tune, *or* you can do a fixed number of blocks per workgroup, but in the latter case you need to need vary the number of dispatches based on total number of blocks.
-That's annoying but not a showstopper.}
-
-\thomas{Edited this section in response to above comments.
-Distinguished Chained Scan as specifically eliding barrier based sychronization.
-Linked fixed launch more clearly to manual tuning and occupancy discovery.
-Shortened section.
-OK to remove comments?}
 
@@
 
-  $\textit{local\_vectors}$ \gets $\textnormal{array}\langle \textnormal{vec4}\langle u32\rangle, \textit{p} \rangle$\;
+  $\textit{local\_vectors} \gets \textnormal{array}\langle \textnormal{vec4}\langle u32\rangle, \textit{p} \rangle$\;
 

### 2025-02-19 paper.tex
fixing compile error
just added a few small comments

--- 2025-02-17 paper.tex
@@
 \subsubsection{Scan-then-Propagate}
-Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2011} extends intra-workgroup scans to large inputs by resolving inter-workgroup dependencies in three phases:
+Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2011} extends intra-workgroup scans to large inputs by resolving inter-workgroup dependencies in three phases:
 \begin{enumerate}
@@
         $\textit{cond1} \gets \textit{fanout\_index} < \textit{spine\_length}$\;
-        $\textit{cond2} \gets (\textit{fanout\_index} \& (\textit{j} - 1)) \geq \textit{reduced\_stride}$\;
+        $\textit{cond2} \gets (\textit{fanout\_index} \,\&\, (\textit{j} - 1)) \geq \textit{reduced\_stride}$\;
 
@@
   \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
-  \caption{(This graph can be provided with or without Memcpy speeds. I am thinking without, as it's currently very busy.) Performance versus CUB (NVIDIA 2080 Super)}
+  \caption{(This graph can be provided with or without Memcpy speeds.
+I am thinking without, as it's currently very busy.
+\john{It's OK to include it, I think.
+I would make the memcpy lines two different shades of gray (not colored) and I would also make them dashed, just to ease the visual clutter.
+I would also consider a dotted horizontal line at the speed of light of the device.
+Also I would pick colors for the CSDLDF that are not so close together (you will dig colorbrewer.org).}) Performance versus CUB (NVIDIA 2080 Super)}
 \end{figure}
@@
     \includegraphics[width=\linewidth]{graphics/time_plot.pdf}
-    \caption{Throughput vs Baseline}
+    \caption{Throughput vs.\ Baseline}
     \label{fig:execution_time}
@@
 
-  \caption{Simulated Blocking}
+  \caption{Simulated Blocking.
+\john{Please write a good caption here---this is an important caption to write!
+Probably writing subfigure captions is the right way to go (explain each of them individually).}}
   \label{fig:vertical_images}
@@
   \end{tabular}
-  \caption{Vendor-Specified Memory Bandwidth and Expected Throughput (32-bit Elements per Second)}
+  \caption{Vendor-Specified Memory Bandwidth and Expected Throughput (32-bit Elements per Second).
+\john{Suggest splitting the Intel line into two lines, one for single and one for dual?
+A cleaner table, I think.}}
   \label{tab:memory_bandwidth}

### 2025-02-23 paper.tex
Tighten language on barriers inline with Raph's feedback.

@@
   We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained-Scans} to run on hardware without \emph{Forward-Progress Guarantees} (FPG) while avoiding deadlock.
-Additionally, we introduce a tile state representation for \emph{Chained-Scans} that does not rely on 64-bit atomics or memory fences for correctness, along with a subgroup-size-agnostic intra-workgroup implementation.
+Additionally, we introduce a tile state representation for \emph{Chained-Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation.
 On an FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled-Fallback} achieves near-\emph{memcopy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach.
-We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by emulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
+We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
 \end{abstract}
@@
 \end{enumerate}
-The absence of inter-workgroup barriers necessitates separate kernel launches for dependency resolution, naturally enforcing this structure.
+The absence of inter-workgroup barriers necessitates separate kernel launches and pipeline barriers for dependency resolution, naturally enforcing this structure.
 To accommodate arbitrarily large inputs, Sengupta et al.
@@
 This constrains the spine from a variable, potentially recursive structure to a fixed size $c$, ensuring a recursion depth of at most two.
-As a result, global memory movement for the spine scan is reduced from $O(n/t)$ to $O(2c)$, and kernel launches are limited to three, yielding a total global memory movement of $O(3n + 4c) = O(3n)$.
+As a result, global memory movement for the spine scan is reduced from $O(n/t)$ to $O(2c)$, and kernel launches and pipeline barriers are limited to three, yielding a total global memory movement of $O(3n + 4c) = O(3n)$.
 
@@
 Recognizing that raking naturally aligns with this paradigm, Breitbart's method statically \emph{discovers} the workgroup occupancy and launches exactly $o$ kernels.
-Since multiprocessors are not oversubscribed, workgroups persist for the entire kernel duration, enabling synchronization at an atomics-based inter-workgroup barrier without context switching.
-While global memory movement remains $O(3n)$, this consolidates the scan into a single kernel, minimizing launch overheads.
+Since multiprocessors are not oversubscribed, workgroups persist for the entire kernel duration, enabling synchronization at an atomics-based inter-workgroup memory and control barrier without kernel launches or pipeline barriers.
+While global memory movement remains $O(3n)$, this consolidates the scan into a single kernel, minimizing launch and pipeline overheads.
 
@@
 First introduced by Yan et al.~\cite{10.1145/2442516.2442539} in \emph{Stream-Scan}, the key innovation of the \emph{Chained-Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level.
-Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data.
+Instead of using kernel launches and pipeline barriers as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data.
 To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than arbitrary assignment based on virtualized workgroup indices.
-This shifts dependency resolution from explicit synchronization via barrier-like structures to implicit synchronization via access order.
+This shifts dependency resolution from explicit synchronization via barriers to implicit synchronization via access order.
 As a result, each element is read and written exactly once, achieving the theoretical minimum of $O(2n)$.
@@
 \emph{Chained Scan} and \emph{Reduce-Then-Scan} represent two contrasting paradigms in scan design.
-By avoiding explicit pipeline barrier based inter-workgroup synchronization, \emph{Chained Scan} aligns with the oversubscription paradigm, delegating occupancy management to the workgroup scheduler.
+By eliding the barrier-idiom, \emph{Chained Scan} aligns with the oversubscription paradigm, delegating occupancy management to the workgroup scheduler.
 In contrast, \emph{Reduce-Then-Scan} employs workgroup raking, requiring fixed launches and explicit occupancy control, forcing developers to manually tune launch sizes or implement occupancy discovery techniques.
@@
 However, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in both theoretical and practical terms, it has a critical limitation: its reliance on a forward-progress guarantee.
-While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that Apple’s M-series GPUs do not \thomas{I've cut mentioning ARM here due to reclassification, but it kind of reduces the impact.
-I.e.
-"Oh it's just Apple that's missing, so this is a vendor hack?"}.
+While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple’s M-series GPUs do not.
 Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
@@
   }
-  \textbf{barrier()}\;
+  \textbf{workgroupBarrier()}\;
 
@@
   \textbf{SubgroupRake}($\textit{local\_vectors}$, $\textit{wg\_partials}$)\;
-  \textbf{barrier()}\;
+  \textbf{workgroupBarrier()}\;
 
   \textbf{WorkgroupWideScan}($\textit{wg\_partials}$)\;
-  \textbf{barrier()}\;
+  \textbf{workgroupBarrier()}\;
 
@@
     \textbf{Lookback}($\textit{spine}$, $\textit{tile\_id}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
-    \textbf{barrier()}\;
+    \textbf{workgroupBarrier()}\;
 
@@
       \textbf{FallbackInsertionAttempt}($\textit{spine}$, $\textit{wg\_fallback}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
-      \textbf{barrier()}\;
+      \textbf{workgroupBarrier()}\;
     }
@@
 
-      \textbf{barrier()}\;
+      \textbf{workgroupBarrier()}\;
 
@@
 This primitive is well studied in the context of memory models~\cite{Alglave0, MCMutants}.
-\raph{The earliest publication of the MP litmus test appears to be section 2.1 of "Intel 64 Architecture Memory Ordering White Paper".
-I'm not sure that's the best cite; the MC mutants paper cites https://arxiv.org/pdf/1207.7264} \thomas{Added MC Mutants + the Algave paper.
-OK?
-Or should cite more?
-Alglave has tons of papers.} The key requirement is both control synchronization, meaning the reader does not proceed until the writer sends the message; and data synchronization, meaning that the reader reads the correct value.
+The key requirement is both control synchronization, meaning the reader does not proceed until the writer sends the message; and data synchronization, meaning that the reader reads the correct value.
 The latter is especially challenging in relaxed memory models as are employed on GPUs.
@@
 The semantics of atomics guarantee, even with relaxed memory ordering, that the data is fresh.
-However, if the tile state and data are stored in two different memory locations, then relaxed atomics are not sufficient and some other sychronization is necessary.
-On CPUs, it is common to upgrade the memory semantics from relaxed to acquire/release, which then provides the necessary guarantee~\cite[Section 10.2]{intel_sdm_2024} \thomas{I've cited the latest publication by Intel.
-The section is nearly identical to 2007.} .
+However, if the tile state and data are stored in separate memory locations, relaxed atomics are insufficient because the compiler is free to reorder stores, so one update does not imply the other.
+On CPUs, it is common to upgrade the memory semantics from relaxed to acquire/release, which then provides the necessary guarantee~\cite[Section 10.2]{intel_sdm_2024}.
 Unfortunately, on GPUs, acquire/release semantics are neither portable (they are only available on Vulkan, and optional until version 1.3), nor are they necessarily efficient even when they are available; often the implementation involves expensive cache flushes.
@@
   \includegraphics[width=\linewidth]{graphics/split.pdf}
+  \caption{Splitting one 32-bit value into two 32-bit values with bit-packed tile state.}
 \end{figure}
@@
 In particular, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state.
-Although this could also be accomplished with shared memory (and barriers), as an optimization, we use two or four threads from the first subgroup, leveraging subgroup \emph{ballot} to verify state consistency, and subgroup \emph{shuffle} to join and split reductions.
-With one thread per 16 bits and a minimum subgroup size of four in WebGPU, this method supports both 32-bit and 64-bit values, enabling 64-bit scans without 64-bit atomics or fences.\footnote{Although native 64-bit values are not currently supported by the WebGPU specification, their future inclusion is anticipated.}
+Although this could also be accomplished with shared memory (and workgroup barriers), as an optimization, we use two or four threads from the first subgroup, leveraging subgroup \emph{ballot} to verify state consistency, and subgroup \emph{shuffle} to join and split reductions.
+With one thread per 16 bits and a minimum subgroup size of four in WebGPU, this method supports both 32-bit and 64-bit values, enabling 64-bit scans without 64-bit atomics or memory barriers.\footnote{Although native 64-bit values are not currently supported by the WebGPU specification, their future inclusion is anticipated.}
 
@@
 \subsection{Trade-offs}
-Decoupled Fallback introduces several tradeoffs that must be considered: (1) Without a memory fence, the scan cannot be performed in place and requires a separate output buffer to prevent false-positive fallback operations on post-scanned values, as the tile state may not update before the scan values.
+\emph{Decoupled Fallback} introduces several tradeoffs that must be considered: (1) Without a memory barrier, the scan cannot be performed in place and requires a separate output buffer to prevent false-positive fallback operations on post-scanned values, as the tile state may not update before the scan values.
 (2) For monoids larger than 64 bits, multiple subgroups are required, increasing synchronization overhead due to the need for shared memory and barriers.
@@
 Sorting is one of the most notable beneficiaries of single-pass scan and, by extension, \emph{Decoupled Fallback}.
-The current state-of-the-art algorithm, Adinet and Merrill’s \emph{Onesweep}~\cite{adinets2022onesweepfastersignificantdigit}, is an 8-bit least-significant-digit radix sort, where single-pass scan minimizes data movement during digit-binning passes.
-However, due to its compute-intensive nature and large $radixDigit$-sized monoid, it is unclear whether \emph{Decoupled Fallback} maps well to sorting or if new techniques are needed.
+The current state-of-the-art algorithm, Adinets and Merrill’s \emph{Onesweep}~\cite{adinets2022onesweepfastersignificantdigit}, is an 8-bit least-significant-digit radix sort, where single-pass scan minimizes data movement during digit-binning passes.
+However, due to its compute-intensive nature and large $radixDigit$-sized monoid, it is unclear whether \emph{Decoupled Fallback} will map well to sorting or if new techniques are needed.
 

### 2025-02-23 paper.tex
Add deadlocking section

@@
 However, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in both theoretical and practical terms, it has a critical limitation: its reliance on a forward-progress guarantee.
-While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple’s M-series GPUs do not.
+While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple’s GPUs do not.
 Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
@@
 \subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
-A \emph{forward-progress guarantee} (FPG) ensures that all occupant subgroups (from possibly different workgroups) eventually make progress toward termination, preventing indefinite starvation.
-Sorensen et al.~\cite{sorensen2018,sorensen2021} provide the most comprehensive analysis of GPU schedulers, classifying \emph{Chained-Scan} as FPG-reliant.
+A \emph{forward-progress guarantee} (FPG) ensures that all occupant subgroups (of possibly different workgroups) eventually make progress toward termination, preventing indefinite starvation.
+Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees.
+Instead, Sorensen et al.
+systematically defined and tested three progress models\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}:
 \begin{itemize}
-  \item Introduce Sorensen
-  \item Introduce Progress models
-  \item Explain why OBE + data
-  \item Explain why one-to-one mapping doesn't solve problem.
+  \item \textbf{Occupancy-Bound Execution (OBE)}: Workgroups that have made at least one execution step.
+  \item \textbf{Heterogeneous System Architecture (HSA)}: The workgroup with the lowest ID that has not yet terminated.
+  \item \textbf{Linear OBE (LOBE)}: Workgroups that have made at least one step, and any workgroup with a lower ID than a workgroup that has made at least one step.
 \end{itemize}
+Their findings demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior assumptions in \cite{sorensen2018}.
 
-\john{Question, and maybe this is worth a footnote.
-What if we guaranteed that no multiprocessor ever had more than one workgroup scheduled to it, say by always allocating the maximum amount of workgroup memory?
-Of course there would be performance issues.
-But does that solve the problem?
-Or do we descend into an analogous issue with subgroup scheduling?}
+This has direct implications for \emph{Chained Scan}, which was \emph{previously} classified as a \emph{one-way producer-consumer} (PC) HSA algorithm: a workgroup only consumes from lower-ID counterparts, ensuring that execution proceeds in a strictly ordered fashion.
+Under HSA, this structure guarantees termination, as the lowest ID workgroup is always given forward progress.
+If subgroups from the lowest ID workgroup are not guaranteed execution, then any dependent subgroup may spin indefinitely, waiting on an update that never arrives.
 
-\raph{Answering John's question here.
-Short answer no.
-One mental model that would illustrate this is a workgroup being preempted from a multiprocessor for system reasons, for example to perform compositing for scanout.
-I brought this up with a GPU engineer and they said this wasn't actually the reason for lack of FPG on their hardware, but wasn't able to explain the true story for proprietary reasons.
-In any case, while I am fine to a large extent talking operationally about scheduling etc, at the end of the day you need a formal model that tells you which computations will terminate.
-That's what Tyler's work is about and I think it's best to cite it and not try to re-litigate it, especially in oversimplified form.
-Also, taking another look at "Specifying and Testing," I notice that they use "livelock" but not deadlock to describe situations where system progress depends on FPG.
-I'm not finding a cite for this now, but if memory serves, a deadlock condition is one in which no progress is possible under *any* scheduling condition, while in livelock there is some scheduler condition that could resolve the livelock, but it is not resolved under all such conditions.}
+However, the HSA classification relies on the assumption that \emph{Chained Scan} follows scheduler-assigned virtualized IDs rather than dynamically acquired atomic-bumped IDs, as used by both \emph{StreamScan} and \emph{Decoupled Lookback}.
+Since virtual IDs do not necessarily match atomic IDs, HSA does not guarantee termination---the lowest virtual ID may not be scheduled first, potentially leading to starvation.
 
-\john{The reason that I ask this question is that I take the perspective of a reviewer who decides, oh, I can solve this FPG problem by doing something simple, like ``scheduling only one workgroup per processor''.
-The reviewer, of course, doesn't actually implement this, but it totally kills the paper when a reviewer gets it into his/her head that there is another way to go about this and why didn't we study THAT way, reject, go back and incorporate my amazing idea.
-We need to explicitly close the door emphatically in the paper to obvious, simple, unworkable solutions; we can't give reviewers any headroom to substitute their own judgment when we know better.}
+LOBE adds an additional condition on top of OBE: lower-ID workgroups must be scheduled before higher-ID ones.
+However, this condition is unnecessary for \emph{Chained Scan} for two reasons: (1) We already guarantee serial ordering of work items through atomic bumping, making OBE sufficient.
+(2) \emph{Chained Scan} only requires forward progress after a workgroup has atomically acquired its tile index.
+By definition, a workgroup that has acquired its index must be occupant on a physical processor (i.e., it has made at least one execution step); thus, OBE guarantees forward progress.
+
+\begin{table}
+  \small
+  \centering
+  \begin{tabular}{lcccc}
+    \toprule
+    Algorithm       & LOBE & OBE & HSA & Unfair \\
+    \midrule
+    CS (Workgroup ID)  & Yes & No  & Yes & No  \\
+    CS (Atomic ID)        & Yes & Yes & No  & No  \\
+    Ours               & Yes & Yes & Yes & Yes \\
+    \bottomrule
+  \end{tabular}
+  \caption{Comparison of Chained Scan (CS) Variants and Our Implementation Across Progress Models.}
+  \label{tab:progress_models}
+\end{table}
+
+\thomas{Is the one-to-one workgroup to multiprocessor rebuttal still neccessary here?}
 
@@
 The key advance is to utilize the parallelism available in the workgroup to perform the fallback computation, rather than performing it sequentially.
-To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained-Propagation} logic is typically performed by the first subgroup.
+To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained-Propagation} logic is typically performed by the first thread.
 

### 2025-02-24 paper.tex
Update Artifact

@@
 We provide our artifact in the following GitHub repository:
-\thomas{Will anonymize with GithubAnonymous once we are ready.}
-
-\subsection{Requirements}
-
-\subsubsection{Device Requirements}
-Any device supporting Dawn, with subgroup and timestamp query capabilities, and at least 384 MB of available device memory. Requires at least $\sim$2GB of disk memory to build Dawn.
-
-\subsubsection{Software Requirements}
-\begin{itemize}
-  \item Git
-  \item CMake 3.10.2+ or another C++ build tool
-  \item C++17-compliant compiler (Clang recommended)
-  \item Python for fetching dependencies
-\end{itemize}
-
-\subsection{Building}
-Although native WebGPU is supported on various operating systems—most notably Windows, macOS, and Linux—this section will focus on building it in Unix-like environments for the sake of brevity, using CMake as the example build tool, and Clang++ as the example compiler.
-
-\subsubsection{Building Dawn}
-\begin{enumerate}
-  \item[(0)] Dawn depends on the following packages (Ubuntu package names):
-        \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
-sudo apt install libxrandr-dev \
-  libxinerama-dev libxcursor-dev libxi-dev \
-  libx11-xcb-dev mesa-common-dev
-  \end{lstlisting}
-
-  \item Clone the Dawn repository:
-        \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
-git clone https://dawn.googlesource.com/dawn
-cd dawn
-git fetch --all
-git checkout chromium/6950
-  \end{lstlisting}
-
-  \item Build and install Dawn:
-        \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
-cmake -S . -B out/Release \
-  -DDAWN_FETCH_DEPENDENCIES=ON \
-  -DDAWN_ENABLE_INSTALL=ON \
-  -DCMAKE_BUILD_TYPE=Release \
-  -DDAWN_BUILD_TESTS=OFF \
-  -DDAWN_BUILD_SAMPLES=OFF \
-  -DTINT_BUILD_TESTS=OFF \
-  -DTINT_BUILD_CMD_TOOLS=OFF \
-  -DCMAKE_CXX_COMPILER=clang++
-
-cmake --build out/Release \
-  --parallel <num_threads>
-
-cmake --install out/Release \
-  --prefix install/Release
-  \end{lstlisting}
-\end{enumerate}
-
-\subsubsection{Building Artifact}
-After successfully building and installing Dawn, follow these steps to build the artifact:
-
-\begin{enumerate}
-  \item Navigate to the artifact directory:
-        \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
-cd Decoupled-Fallback-Paper/artifact/Dawn
-  \end{lstlisting}
-
-  \item Set the `CMAKE\_PREFIX\_PATH` to point to your Dawn installation directory:
-        \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
-export CMAKE_PREFIX_PATH=/path/to/dawn/ \
-  install/Release
-  \end{lstlisting}
-
-  \item Build the artifact:
-        \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
-cmake -S . -B out/Release \
-  -DCMAKE_BUILD_TYPE=Release \
-  -DCMAKE_CXX_COMPILER=clang++
-  \end{lstlisting}
-
-  \item Compile the project:
-        \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
-cmake --build out/Release
-  \end{lstlisting}
-\end{enumerate}
-
-\subsubsection{Running the Artifact}
-Once the artifact has been successfully built, you can run it using the following command:
-
-\begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
-./out/Release/dawn <TestType> [record] [deviceName]
-\end{lstlisting}
-
-where:
-\begin{itemize}
-  \item \texttt{<TestType>} specifies the test to run. Valid options are:
-        \begin{itemize}
-          \item \texttt{csdl} – Runs the \emph{Decoupled Lookback} test. \textbf{Warning}, this \emph{will} crash without FPG\@.
-          \item \texttt{csdldf} – Runs the \emph{Decoupled Fallback} test.
-          \item \texttt{full} – Runs a full set of tests, including multiple \emph{Decoupled Fallback} simulations.
-        \end{itemize}
-  \item \texttt{record} (optional) enables recording the test results to a CSV file.
-  \item \texttt{deviceName} (optional) appends the device name to recorded test results.
-\end{itemize}
+\thomas{Will anonymize with GithubAnonymous once we are ready}
+The repository contains per API requirements and building instructions.
 

### 2025-02-24 paper.tex
Incorporate discussion on max spin counts.

@@
 
-\section{Introduction (WIP)}
+\section{Introduction}
  (Tell readers we are targeting WebGPU)
@@
 Furthermore, its single-kernel design and serial ordering make it suitable for \emph{in-situ} operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging.
-However, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in both theoretical and practical terms, it has a critical limitation: its reliance on a forward-progress guarantee.
+However, it has a critical limitation: its reliance on a forward-progress guarantee.
 While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple’s GPUs do not.
@@
 The absence of FPG in key hardware segments poses a critical challenge for developers.
-Our testing shows that, at best, running \emph{Chained-Scan} without FPG results in severe performance degradation, and at worst, risks TDR and program crashes.
+Our testing shows that, at best, running \emph{Chained-Scan} without FPG results in severe performance degradation, and at worst, risks \emph{Timeout Device Recovery} and program crashes.
 \begin{figure}
@@
 (2) Next, we limit the number of spins a thread can perform while waiting on a tile; exceeding this threshold triggers a fallback.
-While this approach risks redundant fallbacks---since a late tile is indistinguishable from a genuinely blocking one---exposing the maximum spin count as a tunable parameter helps minimize false positives.
-\john{I can see it is sensible to have a maximum spin count but I do not see how the action of exposing the maximum spin count to the programmer minimizes false positives.} (3) Finally, after completing a fallback, the workgroup attempts to post the reduction result of the blocking tile, preventing redundant fallbacks by subsequent workgroups.
-
-\raph{To address John's comment: increasing the spin count reduces false positives; a false positive can be defined as the preceding tile posting the result late, meaning that there is some spin count for which it would proceed.
-However, it also delays progress in the true positive case.
-The deeper way to talk about this is to engage the cost model, which is done to some extent in the paragraph below.
-The cost of a false positive includes increased memory traffic and also compute opportunity cost of other workgroups which can be scheduled on the same multiprocessor.
-Indeed, with sufficient occupancy the cost of a workgroup spinning (or making scalar progress) is minimal.
-So optimal performance really depends on the *distribution* of spin counts.
-I put all this in a comment but think we can say it in the paper.}
-\john{My point was not about the relationship between a spin count and false positives, it was that the draft specifically said that exposing the maximum spin count to the programmer is what minimized false positives.
-It's not the exposing that minimizes false positives.}
+While this approach risks \emph{false positive} fallbacks---since a late tile is indistinguishable from a genuinely blocking one---the maximum spin count can be tuned to minimize redundant fallbacks.
+Increasing the spin count lowers false positives but delays progress in true-blocking scenarios.
+False positives not only increase memory traffic but also consume compute resources that could be scheduled on the same multiprocessor.
+(3) After completing a fallback, the workgroup attempts to post the reduction result of the blocking tile, preventing redundant fallbacks by subsequent workgroups.
 
 The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled-Lookback}.
-Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$\!, independent of FPG, due to \emph{false-positive} fallback operations.
+Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$\!, independent of FPG, due to false-positive fallback operations.
 As a result, memory bandwidth usage increases (and we expect throughput to decrease) by $O(fn)$ to $O((2 + f)n)$.
 However, just as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache.
-As we later empirically demonstrate(Simulated Blocking Section), this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile.
+As we later empirically demonstrate in \ref{sec:Sim-Blocking}, this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile.
 Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.
@@
     \toprule
-    Vendor & Device Name & OBE & Occupancy & Subgroup Size \\
+    Vendor & Device Name & LOBE & Occupancy & Subgroup Size \\
     \midrule
     Intel  & HD Graphics 620        & \textbf{Yes} & 9   & 8--32  \\
-    ARM    & Mali-G78 MP20          & \textbf{Yes} & 40  & 16  \\
+    ARM    & Mali-G78 MP20          & \textbf{No}  & 40  & 16  \\
     Apple  & M3                     & \textbf{No}  & 80  & 32  \\
@@
 
-\subsection{Simulated Blocking}
+\subsection{Simulated Blocking}\label{sec:Sim-Blocking}
 To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally forcing fallbacks through the omission of work tile reductions.
@@
 \section{Discussion}
-\subsection{Limits on Speed-of-Light Performance}
-
-In true blocking scenarios, the stalled workgroup is starved of progress, but execution resources are not lost; rather, they are redistributed to other workgroups.
-As a result, we would expect lookback lengths to be longer than those observed in our simulation.
-
-\thomas{Somewhere in here is perfect for Raph's commentary on the speed of atomics on different hardware. Once we have it, we can retool this section around it.}
 
@@
 (5) Lastly, because the control flow of \emph{Decoupled Fallback} requires barriers, latency incurred during lookback cannot be hidden by workgroup computation.
+
+\subsection{Limits on Speed-of-Light Performance}
 
@@
 \section{Artifact}
+\subsection{Availability}
+We provide our artifact in the following GitHub repository: \thomas{Will anonymize with GithubAnonymous once we are ready}. The repository contains per API requirements and building instructions.
+
 \subsection{Baseline Comparisons: \emph{Memcpy}}
 \label{sec:memcpy}
-\begin{table}[h!]
+\begin{table}
   \small
@@
     Intel HD620 (Single-/Dual-Channel)    & 17.1 / 34.1      & 2.13e9 / 4.26e9 \\
-    ARM Mali-G78 MP20 / Pixel 6a (Tensor) & 51.2             & 6.40e9          \\
+    ARM Mali-G78 MP20 / Tensor (Pixel 6a) & 51.2             & 6.40e9          \\
     Apple M3                              & 102.4            & 12.8e9          \\
@@
 
-\subsection{Availability}
-We provide our artifact in the following GitHub repository:
-\thomas{Will anonymize with GithubAnonymous once we are ready}
-The repository contains per API requirements and building instructions.
-
 \bibliographystyle{ACM-Reference-Format}

### 2025-02-25 paper.tex
New CUDA graph. ROUGH introduction.

@@
 \section{Introduction}
- (Tell readers we are targeting WebGPU)
- (Speed of light definition here)
-\subsection{Goals of a Scan Implementation}
-\begin{itemize}
-  \item \textbf{Minimal global memory access:} the scan should minimize global memory access to the theoretical minimum of $2n$.
-  \item \textbf{Fully saturates global memory bandwidth:} the scan should be capable of fully saturating the global memory bandwidth.
-  \item \textbf{Portability:} the scan should be portable across hardware vendors and architectures.
-\end{itemize}
-Our method retains the best aspects of the \emph{Chained-Scan} architecture, while eliminating the need for forward progress guarantees.
-It is designed to operate on arbitrary hardware, support arbitrary monoids, process inputs of arbitrary size, and support arbitrary 32-bit and 64-bit data types.
+
+What is the exact problem that is being solved?
+Why is this problem important?
+Why is this problem hard / why aren't the current solutions to this problem sufficient?
+(In our case, that's going to be an assertion that current techniques don't work without fpg, but we will certainly not explain it in detail at this point; that's going to be a whole section later in the paper.)
+What are the precise contributions of the work?
+
+Scan is one of the most ubiquitous problems in GPGPU computing, with applications including sorting, sparse matrix etc.
+Although the state of the art prefix scan, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware and implemented in CUDA, it relies on a set of architectural and language-specific guarantees that no longer hold once outside of the NVIDIA ecosystem, most notably a property of a device's scheduler known as a \emph{forward progress guarantee}, without which \emph{Decoupled Lookback} is liable to starve indefinitely.
+
+\emph{Decoupled Lookback} belongs to a class of scan architectures known as \emph{Chained Scan}, which hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$.
+When this algorithm fully saturates the device’s memory bandwidth, it reaches the \emph{Speed of Light} (SOL) performance limit, the theoretical peak throughput constrained by available bandwidth and memory latency.
+As the compute capabilities of major GPUs have grown, there has been increasing interest in bringing Chained Scan techniques into broader use.
+However, the absence of a critical scheduling guarantee has forced developers to revert to slower techniques to maintain compatibility across architectures.
+
 \subsection{Goals}
@@
 \end{enumerate}
-
-\noindent
-To ground these objectives, we select the WebGPU shading language (WGSL) and the WebGPU standard as our target environment.
-WGSL is a meta-level shading language which is translated and compiled as necessary for backend graphics APIs---D3D12, Vulkan, and Metal---and as such, WGSL implementations are bound by the minimum capability across all backends.
-Because WGSL must operate within this limited capability space, it inherently embodies the challenge of portability.
 
@@
 \subsection{Contributions}
-\begin{enumerate}
-  \item Decoupled-Fallback
-  \item Split-Method
-  \item Subgroup-Size Agnostic stuff
-\end{enumerate}
+We present \textit{Chained Scan with Decoupled Lookback and Decoupled Fallback}, a fully portable prefix scan capable of reaching speed of light performance without FPG, suitable for the shading language environment (graphics APIs?), and implemented in WGSL.
+(REWORD THIS) Additionally, we introduce a tile state representation for \emph{Chained-Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation.
 
@@
 \subsection{Performance Comparison}
-\subsubsection{Performance Versus Contemporary Implementations:}
-We first compare \emph{Decoupled Fallback} (\emph{CSDLDF}) to the \emph{Decoupled Lookback} (\emph{CSDL}) implementation found in NVIDIA’s CUB library, using the 2080 Super only.
-To assess the impact of differing APIs, we also provide the results of an identical \emph{CSDLDF} in CUDA.\footnote{Due to WebGPU’s 128MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.}
+\subsubsection{Cross-API Performance Versus Contemporary Implementations}
+We first compare \emph{Decoupled Fallback} (\emph{DF}) to two key benchmarks: \emph{Memcpy}\textdagger{~\ref{sec:memcpy}} and the \emph{Decoupled Lookback} (\emph{DL}) implementation found in NVIDIA’s CUB library, using the 2080 Super only.
+To isolate the impact of differing APIs, we implement and evaluate \emph{DF} in both CUDA and Dawn.\footnote{Due to WebGPU’s 128MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.} \emph{Memcpy} serves as a performance reference, as it shares the same $2n$ global memory movement as \emph{DF} but lacks compute overhead, ensuring it remains purely memory bandwidth bound.
 
-Cross-API comparisons remain challenging, as despite being nearly identical at a high level, the CUDA implementation consistently outperforms its Dawn counterpart.
-Additionally, the performance cost of \emph{Decoupled Fallback} is most evident in our CUDA implementation when the entire input fits within the 2080 Super’s 4MB L2 cache, where increased synchronization and compute overhead slightly reduce throughput.
-However, at larger input sizes, where memory accesses extend beyond the cache, \emph{CSDLDF} achieves parity with CUB.
+Cross-API comparisons remain challenging, as our CUDA implementation consistently outperforms its Dawn counterpart despite near-identical high-level implementations.
+The performance cost of \emph{DF} is most pronounced in CUDA when the entire input fits within the 2080 Super’s 4MB L2 cache, where reduced global memory latency exposes \emph{DF's} increased synchronization and compute overhead compared to CUB.
+However, at larger input sizes, where memory accesses exceed cache capacity, \emph{DF} reaches parity with CUB and achieves \emph{Memcpy} speeds.
 
@@
   \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
-  \caption{(This graph can be provided with or without Memcpy speeds.
-I am thinking without, as it's currently very busy.
-\john{It's OK to include it, I think.
-I would make the memcpy lines two different shades of gray (not colored) and I would also make them dashed, just to ease the visual clutter.
-I would also consider a dotted horizontal line at the speed of light of the device.
-Also I would pick colors for the CSDLDF that are not so close together (you will dig colorbrewer.org).}) Performance versus CUB (NVIDIA 2080 Super)}
+  \caption{Performance versus CUB (NVIDIA 2080 Super)}
 \end{figure}
@@
 \subsubsection{Same-API Performance Versus Baseline Measures}
-We next compare \emph{CSDLDF} against two key benchmarks: \emph{Memcpy}\textdagger{~\ref{sec:memcpy}} and \emph{Reduce then Scan}\textdagger{~\ref{sec:rts}} (\emph{RTS}).
-\emph{RTS} serves as a baseline to demonstrate the performance gains of our approach over the previous FPG-constrained technique.
-\emph{Memcpy} is included as a point of reference, as it shares the same $2n$ global memory movement as \emph{Decoupled Fallback}, and without compute overhead, is guaranteed to be memory bandwidth bound.
+We next compare \emph{DF} against \emph{Reduce then Scan}\textdagger{~\ref{sec:rts}} (\emph{RTS}), which serves as a baseline to highlight the performance improvements of our approach over the prior FPG-constrained technique.
+This evaluation is conducted using Dawn across the full range of tested devices.
 
@@
     Avg \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)  & 1.50   & 2.48  & 10.25  & 38.62  & 54.39  & 72.28  \\
-    Avg \emph{CSDLDF} (32-bit ele/sec $\cdot 10^9$)   & 1.49   & 2.79  & 10.22  & 38.84  & 53.29  & 94.60  \\
+    Avg \emph{DF} (32-bit ele/sec $\cdot 10^9$)   & 1.49   & 2.79  & 10.22  & 38.84  & 53.29  & 94.60  \\
     Avg \emph{RTS} (32-bit ele/sec $\cdot 10^9$)      & 1.06   & 2.02  & 6.81   & 25.13  & 35.90  & 61.49  \\
-    \emph{CSDLDF}/\emph{Memcpy}       & 99.1\%    & 112.8\%  & 99.8\%    & 100.6\%   & 98.0\%    & 130.9\%   \\
-    \emph{CSDLDF}/\emph{RTS}          & 140.2\%   & 138.5\%  & 150.2\%   & 154.5\%   & 148.4\%   & 153.8\%   \\
+    \emph{DF}/\emph{Memcpy}       & 99.1\%    & 112.8\%  & 99.8\%    & 100.6\%   & 98.0\%    & 130.9\%   \\
+    \emph{DF}/\emph{RTS}          & 140.2\%   & 138.5\%  & 150.2\%   & 154.5\%   & 148.4\%   & 153.8\%   \\
     Avg Spins per Wg    & 0.305    & 1.477    & 0.606    & 0.604    & 0.419    & 0.148    \\
@@
   \end{tabular*}
-  \caption{\emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, Input Size $\mathbf{2^{25}}$}
+  \caption{\emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements}
   \label{tab:results}
@@
 
-As shown in Table~\ref{tab:results}, \emph{CSDLDF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices—closely approaching the theoretically expected $O(3n)/O(2n)$ 50\% speedup, except on the ARM Mali.
-Since \emph{Memcpy} is purely memory-bandwidth-bound, \emph{CSDLDF} should ideally match its throughput.
-This holds across most devices, except for the AMD 7900 XT and ARM Mali, where \emph{CSDLDF} outperforms \emph{Memcpy}, suggesting that the additional compute latency counterintuitively improves memory access patterns by reducing paging.
-\john{I have no idea what this previous phrase means.} \thomas{In my head, if you read from location A, and then write immediately to B, you need room in your cache for A and B because the workgroup doesn't execute in lockstep.
-If you read, do just the right amount of work, then write B, you only need to hold A or B because the compute latency ensures that all the work from A is done before you get to B.}On the ARM Mali, \emph{CSDLDF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%.
-However, as its performance is nearly identical to \emph{Memcpy}, this indicates that the implementation is memory-bandwidth-bound, limiting further gains.
+As shown in Table~\ref{tab:results}, \emph{DF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices, closely approaching the theoretically expected $O(3n)/O(2n)$ 50\% speedup, except on the ARM Mali.
+\emph{DF} closely matches \emph{Memcpy} across most devices, deviating only on the AMD 7900 XT and ARM Mali, where \emph{DF} outperforms \emph{Memcpy} by 30.9\% and 12.8\%, respectively.
+This suggests that additional compute latency may mitigate paging effects by separating loads from stores.
+On the ARM Mali, \emph{DF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%.
+However, its nearly identical performance to \emph{Memcpy}, indicates that the implementation is memory-bandwidth-bound, limiting further gains.
 
 \subsection{Decoupled Fallback Statistics}
-To better understand how devices interact with \emph{CSDLDF}, we also collect a number of statistics during its execution:
+To better understand how devices interact with \emph{DF}, we also collect a number of statistics during its execution:
 \begin{itemize}
@@
 \subsection{Simulated Blocking}\label{sec:Sim-Blocking}
-To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally forcing fallbacks through the omission of work tile reductions.
-Beginning with infrequent omissions---1 in 512 work tiles---we progressively increase the rate to near-continuous levels, reaching 1 in 2 work tiles.
-\begin{figure*}[t]
+To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally omitting work tile reduction updates.
+Because these updates are skipped, the immediate successor workgroup reads \emph{Not Ready}, spinning and triggering a fallback.
+Starting with infrequent omissions---1 in 512 work tiles---we progressively increase the omission rate to near-continuous levels---1 in 2 work tiles----recording \emph{DF} statistics at each step.
+\begin{figure*}
   \centering
@@
 
-Our reexamination of \emph{Chained Scan} with dynamic assignment raises fundamental questions about GPU progress models: Can all HSA algorithms be converted to OBE via dynamic assignment?
+Our reexamination of \emph{Chained Scan} with dynamic assignment raises fundamental questions about the GPU progress models proprosed by Sorensen et al.
+in \cite{sorensen2021}: Can all HSA algorithms be converted to OBE via dynamic assignment?
 Are there cases, excluding atomic memory errors, where LOBE guarantees termination but dynamic assignment with OBE does not?
-\textbf{Is OBE all you need}?
+Is OBE all you need?
 
@@
     \midrule
-    Intel HD620 (Single-/Dual-Channel)    & 17.1 / 34.1      & 2.13e9 / 4.26e9 \\
+    Intel HD620 Single-Channel            & 17.1             & 2.13e9          \\
+    Intel HD620 Dual-Channel              & 34.1             & 4.26e9          \\
     ARM Mali-G78 MP20 / Tensor (Pixel 6a) & 51.2             & 6.40e9          \\
@@
   \end{tabular}
-  \caption{Vendor-Specified Memory Bandwidth and Expected Throughput (32-bit Elements per Second).
-\john{Suggest splitting the Intel line into two lines, one for single and one for dual?
-A cleaner table, I think.}}
+  \caption{Vendor-Specified Memory Bandwidth and Expected Throughput (32-bit Elements per Second).}
   \label{tab:memory_bandwidth}
 \end{table}
-Although all backend APIs support \emph{timestamp queries inside encoders}, Apple GPUs do not, so we are forced to implement our own \emph{Memcpy} kernel to measure memory bandwidth, as opposed to simply timing an API's copy buffer operation.
+Although all of Dawn's backend APIs support \emph{timestamp queries inside encoders}, Apple GPUs do not, so we are forced to implement our own \emph{Memcpy} kernel to measure memory bandwidth, as opposed to simply timing an API's copy buffer operation.
 It is well known that \emph{observed} peak memory bandwidth, tends to diverge from vendor-listed peak bandwidth, so for completeness we list these figures above.

### 2025-02-26 paper.tex
Introduction Done. Regenerate Graphs. Add Limits of SOL draft.

@@
 
+\usepackage{xcolor}
+
+\pagecolor[rgb]{0.5,0.5,0.5}
+\color[rgb]{1,1,1}
+
 \begin{document}
@@
 Additionally, we introduce a tile state representation for \emph{Chained-Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation.
-On an FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled-Fallback} achieves near-\emph{memcopy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach.
+On an FPG-lacking devices---Apple M1 Max and M3 ()---\emph{Decoupled-Fallback} achieves near-\emph{memcopy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach.
 We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
@@
 
-\received{20 February 2007}
-\received[revised]{12 March 2009}
-\received[accepted]{5 June 2009}
+%\received{?}
+%\received[revised]{?}
+%\received[accepted]{?}
 
@@
 \section{Introduction}
+Scan is one of the most ubiquitous operations in GPGPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more.
+The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture.
+This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$.
+\emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{Speed of Light} (SOL) performance limit---the theoretical maximum throughput constrained only by bandwidth and memory latency.
+However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward progress guarantee}; without this guarantee, it risks indefinite starvation.
 
-What is the exact problem that is being solved?
-Why is this problem important?
-Why is this problem hard / why aren't the current solutions to this problem sufficient?
-(In our case, that's going to be an assertion that current techniques don't work without fpg, but we will certainly not explain it in detail at this point; that's going to be a whole section later in the paper.)
-What are the precise contributions of the work?
+As the compute capabilities of major graphics APIs---Vulkan, Direct3D, and Metal---have grown, there has been increasing interest in adapting \emph{Decoupled Lookback} beyond CUDA for broader adoption in real-time graphics and GPGPU computing.
+However, portable implementation is constrained by both hardware and API limitations.
+In terms of hardware, some GPU schedulers—most notably Apple’s M series and ARM—lack forward progress guarantees, and subgroup sizes vary across architectures.
+In terms of API design, the absence of 64-bit atomic operations and explicit memory barriers further complicates implementation, as both are integral to \emph{Decoupled Lookback}.
 
-Scan is one of the most ubiquitous problems in GPGPU computing, with applications including sorting, sparse matrix etc.
-Although the state of the art prefix scan, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware and implemented in CUDA, it relies on a set of architectural and language-specific guarantees that no longer hold once outside of the NVIDIA ecosystem, most notably a property of a device's scheduler known as a \emph{forward progress guarantee}, without which \emph{Decoupled Lookback} is liable to starve indefinitely.
+This work is guided by two main goals.
+(1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained-Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward progress guarantees (FPG) or support for 64-bit atomics.
+\textbf{Performance:} We aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.
 
-\emph{Decoupled Lookback} belongs to a class of scan architectures known as \emph{Chained Scan}, which hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$.
-When this algorithm fully saturates the device’s memory bandwidth, it reaches the \emph{Speed of Light} (SOL) performance limit, the theoretical peak throughput constrained by available bandwidth and memory latency.
-As the compute capabilities of major GPUs have grown, there has been increasing interest in bringing Chained Scan techniques into broader use.
-However, the absence of a critical scheduling guarantee has forced developers to revert to slower techniques to maintain compatibility across architectures.
-
-\subsection{Goals}
-This work is guided by two main goals:
-\begin{enumerate}
-  \item \textbf{Portability}: we want to develop a scan implementation that retains the benefits of the \emph{Chained-Scan} architecture but is also suitable for a diverse range of hardware vendors and architectures, including those without FPG or support for 64-bit atomics.
-  \item \textbf{Performance}: we aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.
-\end{enumerate}
-
-\subsection{Non Goals}
-Although our goal is to create a fully portable and highly performant scan implementation, we are constrained by underlying hardware platforms and programming models.
-As discussed in more detail in \emph{Limits on the Speed-of-Light}, not all architectures offer atomic operations that are sufficiently fast enough or scheduling models that are sufficiently fair enough to attain speed-of-light performance, and thus we cannot guarantee such performance.
-Although we are cognizant of the risks posed by subgroup divergence, subgroup functions in shading languages do not allow explicit divergence control,\footnote{For example, CUDA allows developers to explicitly provision subgroup functions with a mask of the threads that will participate in them.} placing a solution to subgroup divergence issues outside our scope.
+However, certain aspects fall outside the scope of this work.
+\textbf{Performance guarantees} are not universally attainable, as not all architectures offer atomic operations that are sufficiently fast or scheduling models that are sufficiently fair to attain speed-of-light performance.
+\textbf{Subgroup divergence} is another challenge we acknowledge but do not address directly, as shading languages do not provide explicit divergence control like CUDA’s subgroup masking mechanisms.
 Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained-Scan}.
 
-\subsection{Contributions}
-We present \textit{Chained Scan with Decoupled Lookback and Decoupled Fallback}, a fully portable prefix scan capable of reaching speed of light performance without FPG, suitable for the shading language environment (graphics APIs?), and implemented in WGSL.
-(REWORD THIS) Additionally, we introduce a tile state representation for \emph{Chained-Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation.
+We present \emph{Decoupled Fallback}, a fully portable prefix scan capable of achieving speed-of-light performance without requiring forward progress guarantees (FPG).
+Our implementation is designed for compatibility with shading languages, making it suitable for modern graphics APIs and implemented in the WebGPU Shading Language (WGSL).
+Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.
 
@@
 The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \).
-To allow parallelization, \( \oplus \) must be associative, but it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}.
+To allow parallelization, \( \oplus \) must be associative, but it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{Clifford1961}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}.
 In a scan, the result at the \( n \)-th element is the reduction of the preceding subsequence of elements.
@@
 Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities.
-We empirically find that the Google Chrome native implementation of WebGPU (Dawn) provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU\@.\footnote{To keep the perforamance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
+We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU\@.\footnote{To keep the perforamance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
 \begin{table}
@@
     ARM    & Mali-G78 MP20          & \textbf{No}  & 40  & 16  \\
-    Apple  & M3                     & \textbf{No}  & 80  & 32  \\
-    Apple  & M1 Max                 & \textbf{No}  & 160 & 32  \\
+    Apple  & M3     (??? Cores)     & \textbf{No}  & 80  & 32  \\
+    Apple  & M1 Max (32 Cores)      & \textbf{No}  & 160 & 32  \\
     NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192 & 32  \\
@@
 We first compare \emph{Decoupled Fallback} (\emph{DF}) to two key benchmarks: \emph{Memcpy}\textdagger{~\ref{sec:memcpy}} and the \emph{Decoupled Lookback} (\emph{DL}) implementation found in NVIDIA’s CUB library, using the 2080 Super only.
-To isolate the impact of differing APIs, we implement and evaluate \emph{DF} in both CUDA and Dawn.\footnote{Due to WebGPU’s 128MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.} \emph{Memcpy} serves as a performance reference, as it shares the same $2n$ global memory movement as \emph{DF} but lacks compute overhead, ensuring it remains purely memory bandwidth bound.
+To isolate the impact of differing APIs, we implement and evaluate \emph{DF} in both CUDA and Dawn.\footnote{Due to WebGPU’s 128MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.} \emph{Memcpy} serves as a performance reference, as it shares the same $2n$ global memory movement as \emph{DF} but lacks compute overhead and is purely memory bandwidth bound.
 
@@
 \subsection{Decoupled Fallback Statistics}
-To better understand how devices interact with \emph{DF}, we also collect a number of statistics during its execution:
+To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution:
 \begin{itemize}
@@
 \end{itemize}
-Despite significantly different absolute occupancies, the number of waiting spins remains consistent across devices.
+Despite differing occupancy rates, the number of waiting spins remains consistent across devices.
 Likewise, lookback length is stable and remains well below the theoretical occupancy bound.
@@
 \subsection{Limits on Speed-of-Light Performance}
+Due to its increased latency, synchronization overhead, and longer critical path, \emph{Decoupled Fallback} exacerbates existing challenges faced by \emph{Decoupled Lookback}, particularly a lack of compute capacity and unfair scheduling.
+Devices that already lack sufficient compute resources to reach SOL performance are especially affected, as the additional compute overhead from fallback operations cannot be hidden by memory loads and stores.
+The worst-case scenario arises in devices that suffer from both low compute power and unfair scheduling.
+
+Another potential bottleneck is the speed of atomic operations.
+On some older GPUs—such as pre-Kepler NVIDIA architectures—atomics do not operate within the cache, resulting in significantly higher latency.
+Slower atomics lead to delayed tile updates, which in turn increase lookback lengths.
+While our simulated deadlocks provide an artificial means of stress-testing execution, real deadlocks manifest as an effective \emph{occupancy reduction}: execution resources remain available, but fewer subgroups can context-switch to mitigate latency.
 
@@
     ARM Mali-G78 MP20 / Tensor (Pixel 6a) & 51.2             & 6.40e9          \\
-    Apple M3                              & 102.4            & 12.8e9          \\
-    Apple M1 Max                          & 409.6            & 51.2e9          \\
+    Apple M3 (??? Cores)                  & 102.4            & 12.8e9          \\
+    Apple M1 Max (32 Cores)               & 409.6            & 51.2e9          \\
     Nvidia 2080 Super                     & 496              & 62.0e9          \\
@@
   \item \textbf{Nvidia 2080 Super:} Observed throughput matches expectations.
-  \item \textbf{AMD 7900 XT:} Our above result, where \emph{CSDLDF} was observed to be faster than \emph{Memcopy}, is likely due to compute latency counterintuitively improving memory access patterns, leading to reduced paging.
-Furthermore, we see that our \emph{CSDLDF} throughput is within spec and reasonable.
+  \item \textbf{AMD 7900 XT:} Our above result, where \emph{DF} was observed to be faster than \emph{Memcopy}, is likely due to compute latency counterintuitively improving memory access patterns, leading to reduced paging.
+Furthermore, we see that our \emph{DF} throughput is within spec and reasonable.
 \end{itemize}
@@
 \label{sec:rts}
-As we desire the most competitve baseline possible, our \emph{RTS} kernels employ the same intra-workgroup scan strategy and tuning parameters as \emph{CSDLDF}.
+As we desire the most competitve baseline possible, our \emph{RTS} kernels employ the same intra-workgroup scan strategy and tuning parameters as \emph{DF}.
 Initially, we adopted Merrill and Grimshaw's~\cite{Merrill2009} workgroup raking approach for inter-workgroup processing.

### 2025-02-26 paper.tex
JDO's MacBook Air M3 has 10 cores

@@
     ARM    & Mali-G78 MP20          & \textbf{No}  & 40  & 16  \\
-    Apple  & M3     (??? Cores)     & \textbf{No}  & 80  & 32  \\
+    Apple  & M3     (10 Cores)      & \textbf{No}  & 80  & 32  \\
     Apple  & M1 Max (32 Cores)      & \textbf{No}  & 160 & 32  \\

### 2025-02-26 paper.tex
making a couple of tables fit to page/column width

@@
   \small
+  \setlength{\tabcolsep}{1pt}
   \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} l l c r r}
@@
   \centering
+  \setlength{\tabcolsep}{1pt}
   \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r r}
@@
     ARM Mali-G78 MP20 / Tensor (Pixel 6a) & 51.2             & 6.40e9          \\
-    Apple M3 (??? Cores)                  & 102.4            & 12.8e9          \\
+    Apple M3 (10 Cores)                   & 102.4            & 12.8e9          \\
     Apple M1 Max (32 Cores)               & 409.6            & 51.2e9          \\

### 2025-02-26 paper.tex
linted (actually, chktex)

@@
 Because multiple workgroups may reside on a single multiprocessor, two key issues arise at the subgroup scheduling level: \emph{fairness}---how evenly execution resources are distributed among subgroups---and \emph{progress guarantees}---ensuring that subgroups eventually make progress towards termination.
-However, the term "fairness" is used differently across the literature, leading to potential confusion.
-Sorensen et al.~\cite{sorensen2016,sorensen2018,sorensen2021}, whose work provides the most comprehensive taxonomy of progress models to date, use "fairness" specifically to describe differing levels of progress guarantees.
-In contrast, NVIDIA publications~\cite{4523358,Merrill2016} define "fairness" in terms of the distribution of execution resources among subgroups.
+However, the term ``fairness'' is used differently across the literature, leading to potential confusion.
+Sorensen et al.~\cite{sorensen2016,sorensen2018,sorensen2021}, whose work provides the most comprehensive taxonomy of progress models to date, use ``fairness'' specifically to describe differing levels of progress guarantees.
+In contrast, NVIDIA publications~\cite{4523358,Merrill2016} define ``fairness'' in terms of the distribution of execution resources among subgroups.
 In this work, we adopt NVIDIA's terminology.
@@
 The absence of inter-workgroup barriers necessitates separate kernel launches and pipeline barriers for dependency resolution, naturally enforcing this structure.
-To accommodate arbitrarily large inputs, Sengupta et al.
-employ recursion, repeatedly applying the three-phase process until the spine fits within a single work tile.
+To accommodate arbitrarily large inputs, Sengupta et al.\ employ recursion, repeatedly applying the three-phase process until the spine fits within a single work tile.
 Thus, an input of size $n$ and a work tile of size $t$ results in a recursive depth of $\lceil \log_t n \rceil$, and $2\cdot\lceil \log_t n \rceil - 1$ kernel launches.
@@
 Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees.
-Instead, Sorensen et al.
-systematically defined and tested three progress models\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}:
+Instead, Sorensen et al.\ systematically defined and tested three progress models:\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}
 \begin{itemize}
@@
 \end{itemize}
-Their findings demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior assumptions in \cite{sorensen2018}.
+Their findings demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior assumptions~\cite{sorensen2018}.
 
@@
   \end{tabular}
-  \caption{Comparison of Chained Scan (CS) Variants and Our Implementation Across Progress Models.}
-  \label{tab:progress_models}
+  \caption{Comparison of Chained Scan (CS) Variants and Our Implementation Across Progress Models.\label{tab:progress_models}}
 \end{table}
@@
 However, just as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache.
-As we later empirically demonstrate in \ref{sec:Sim-Blocking}, this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile.
+As we later empirically demonstrate in Section~\ref{sec:Sim-Blocking}, this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile.
 Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.
@@
   \Return{$\textit{x}$}\;
-  \caption{Workgroup-Wide Scan.}
-  \label{alg:example}
+  \caption{Workgroup-Wide Scan.\label{alg:example}}
 \end{algorithm}
@@
   \includegraphics[width=\linewidth]{graphics/FlowChart.pdf}
-  \label{fig:decoupled_fallback}
-  \caption{Flow Chart Representation of Decoupled Fallback}
+  \caption{Flow Chart Representation of Decoupled Fallback.\label{fig:decoupled_fallback}}
 \end{figure}
@@
   \end{tabular*}
-  \caption{Hardware Specifications of Testing Devices.}
-  \label{tab:hardware}
+  \caption{Hardware Specifications of Testing Devices.\label{tab:hardware}}
 \end{table}
@@
 Cross-API comparisons remain challenging, as our CUDA implementation consistently outperforms its Dawn counterpart despite near-identical high-level implementations.
-The performance cost of \emph{DF} is most pronounced in CUDA when the entire input fits within the 2080 Super’s 4MB L2 cache, where reduced global memory latency exposes \emph{DF's} increased synchronization and compute overhead compared to CUB.
+The performance cost of \emph{DF} is most pronounced in CUDA when the entire input fits within the 2080 Super’s 4MB L2 cache, where reduced global memory latency exposes \emph{DF's} increased synchronization and compute overhead compared to CUB\@.
 However, at larger input sizes, where memory accesses exceed cache capacity, \emph{DF} reaches parity with CUB and achieves \emph{Memcpy} speeds.
@@
   \end{tabular*}
-  \caption{\emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements}
-  \label{tab:results}
+  \caption{\emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements.\label{tab:results}}
 \end{table*}
@@
 Because these updates are skipped, the immediate successor workgroup reads \emph{Not Ready}, spinning and triggering a fallback.
-Starting with infrequent omissions---1 in 512 work tiles---we progressively increase the omission rate to near-continuous levels---1 in 2 work tiles----recording \emph{DF} statistics at each step.
+Starting with infrequent omissions---1 in 512 work tiles---we progressively increase the omission rate to near-continuous levels---1 in 2 work tiles---recording \emph{DF} statistics at each step.
 \begin{figure*}
@@
     \includegraphics[width=\linewidth]{graphics/fallbacksInitiated_plot.pdf}
-    \caption{Fallbacks Initiated}
-    \label{fig:fallbacks_initiated}
+    \caption{Fallbacks Initiated.\label{fig:fallbacks_initiated}}
   \end{subfigure}\hfill
@@
     \includegraphics[width=\linewidth]{graphics/totalSpins_plot.pdf}
-    \caption{Total Spins}
-    \label{fig:total_spins}
+    \caption{Total Spins.\label{fig:total_spins}}
   \end{subfigure}
@@
     \includegraphics[width=\linewidth]{graphics/lookbackLength_plot.pdf}
-    \caption{Lookback Length}
-    \label{fig:lookback_length}
+    \caption{Lookback Length.\label{fig:lookback_length}}
   \end{subfigure}\hfill
@@
     \includegraphics[width=\linewidth]{graphics/time_plot.pdf}
-    \caption{Throughput vs.\ Baseline}
-    \label{fig:execution_time}
+    \caption{Throughput vs.\ Baseline.\label{fig:execution_time}}
   \end{subfigure}
@@
 \john{Please write a good caption here---this is an important caption to write!
-Probably writing subfigure captions is the right way to go (explain each of them individually).}}
-  \label{fig:vertical_images}
+Probably writing subfigure captions is the right way to go (explain each of them individually).}\label{fig:vertical_images}}
 \end{figure*}
@@
 Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%.
-While Figures~\ref{fig:fallbacks_initiated} and \ref{fig:total_spins} confirm that we are inducing fallbacks, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior.
+While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that we are inducing fallbacks, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior.
 The finer details of this distinction are discussed next.
@@
 
-Our reexamination of \emph{Chained Scan} with dynamic assignment raises fundamental questions about the GPU progress models proprosed by Sorensen et al.
-in \cite{sorensen2021}: Can all HSA algorithms be converted to OBE via dynamic assignment?
+Our reexamination of \emph{Chained Scan} with dynamic assignment raises fundamental questions about the GPU progress models proprosed by Sorensen et al.~\cite{sorensen2021}: Can all HSA algorithms be converted to OBE via dynamic assignment?
 Are there cases, excluding atomic memory errors, where LOBE guarantees termination but dynamic assignment with OBE does not?
@@
 
-\subsection{Baseline Comparisons: \emph{Memcpy}}
+\subsection{Baseline Comparisons: \emph{Memcpy}}%
 \label{sec:memcpy}
@@
   \end{tabular}
-  \caption{Vendor-Specified Memory Bandwidth and Expected Throughput (32-bit Elements per Second).}
-  \label{tab:memory_bandwidth}
+  \caption{Vendor-Specified Memory Bandwidth and Expected Throughput (32-bit Elements per Second).\label{tab:memory_bandwidth}}
 \end{table}
@@
 
-\subsection{Baseline Comparisons: \emph{Reduce then Scan}}
+\subsection{Baseline Comparisons: \emph{Reduce then Scan}}%
 \label{sec:rts}

### 2025-02-26 paper.tex
update to acmsmall format

@@
 
-\documentclass[sigconf]{acmart}
+\documentclass[acmsmall, manuscript, screen, review, anonymous]{acmart}
 %%
@@
 \setcopyright{acmlicensed}
-\copyrightyear{2018}
-\acmYear{2018}
+\copyrightyear{2025}
+\acmYear{2025}
 \acmDOI{XXXXXXX.XXXXXXX}
-\acmConference[Conference acronym 'XX]{Make sure to enter the correct
-  conference title from your rights confirmation emai}{June 03--05,
-  2018}{Woodstock, NY}
+\acmConference[SPAA '25]{37th ACM Symposium on Parallelism in Algorithms and Architectures}{July 28--August 1,
+  2025}{Portland, OR, USA}
 \acmISBN{978-1-4503-XXXX-X/18/06}
@@
 
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{bib}
+
 \clearpage
@@
 
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{bib}
-
 \end{document}

### 2025-02-26 paper.tex
mostly spacing for tables / figure, a few typography things too

@@
 However, portable implementation is constrained by both hardware and API limitations.
-In terms of hardware, some GPU schedulers—most notably Apple’s M series and ARM—lack forward progress guarantees, and subgroup sizes vary across architectures.
+In terms of hardware, some GPU schedulers—most notably Apple's M series and ARM—lack forward progress guarantees, and subgroup sizes vary across architectures.
 In terms of API design, the absence of 64-bit atomic operations and explicit memory barriers further complicates implementation, as both are integral to \emph{Decoupled Lookback}.
@@
 \textbf{Performance guarantees} are not universally attainable, as not all architectures offer atomic operations that are sufficiently fast or scheduling models that are sufficiently fair to attain speed-of-light performance.
-\textbf{Subgroup divergence} is another challenge we acknowledge but do not address directly, as shading languages do not provide explicit divergence control like CUDA’s subgroup masking mechanisms.
+\textbf{Subgroup divergence} is another challenge we acknowledge but do not address directly, as shading languages do not provide explicit divergence control like CUDA's subgroup masking mechanisms.
 Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained-Scan}.
@@
 
-\subsection{The Scan Primitive}\label{sec:the-scan-primitive}
+\subsection{The Scan Primitive}%
+\label{sec:the-scan-primitive}
 The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{5219801,10.5555/1098666}.
@@
 
-LOBE adds an additional condition on top of OBE: lower-ID workgroups must be scheduled before higher-ID ones.
+LOBE adds an additional condition on top of OBE\@: lower-ID workgroups must be scheduled before higher-ID ones.
 However, this condition is unnecessary for \emph{Chained Scan} for two reasons: (1) We already guarantee serial ordering of work items through atomic bumping, making OBE sufficient.
@@
   \small
-  \SetAlgoLined
+  \SetAlgoLined%
   \KwGlobalIn{%
@@
   \centering
-  \includegraphics[width=\linewidth]{graphics/FlowChart.pdf}
+  \includegraphics[height=\textheight]{graphics/FlowChart.pdf}
   \caption{Flow Chart Representation of Decoupled Fallback.\label{fig:decoupled_fallback}}
@@
   \centering
-  \small
-  \setlength{\tabcolsep}{1pt}
-  \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} l l c r r}
+  \begin{tabular}{l l c r r}
     \toprule
@@
     \bottomrule
-  \end{tabular*}
+  \end{tabular}
   \caption{Hardware Specifications of Testing Devices.\label{tab:hardware}}
@@
   \centering
+  \small
   \setlength{\tabcolsep}{1pt}
@@
 
-\subsection{Simulated Blocking}\label{sec:Sim-Blocking}
+\subsection{Simulated Blocking}%
+\label{sec:Sim-Blocking}
 To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally omitting work tile reduction updates.

### 2025-02-27 artifact/Dawn/build_dawn.sh
Add shell script for dawn

@@ -1,0 +1,53 @@
+#!/bin/bash
+set -e
+set -o pipefail
+
+NUM_THREADS=${1:-4}
+SCRIPT_DIR="$(dirname "$(realpath "$0")")"
+ROOT_DIR="$(realpath "$SCRIPT_DIR/../../..")"
+
+echo "Script directory: $SCRIPT_DIR"
+echo "Project root: $ROOT_DIR"
+
+command -v git >/dev/null 2>&1 || { echo "Error: git is required but not installed."; exit 1; }
+command -v cmake >/dev/null 2>&1 || { echo "Error: cmake is required but not installed."; exit 1; }
+command -v clang++ >/dev/null 2>&1 || { echo "Error: clang++ is required but not installed."; exit 1; }
+
+echo "=== Installing dependencies ==="
+sudo apt update
+sudo apt install -y libxrandr-dev libxinerama-dev libxcursor-dev libxi-dev libx11-xcb-dev mesa-common-dev
+
+echo "=== Cloning the Dawn repository ==="
+if [ ! -d "$ROOT_DIR/dawn" ]; then
+    git clone https://dawn.googlesource.com/dawn "$ROOT_DIR/dawn"
+else
+    echo "Dawn repository already exists; skipping clone."
+fi
+
+cd "$ROOT_DIR/dawn"
+git fetch --all
+git checkout chromium/6950
+
+echo "=== Building and installing Dawn ==="
+cmake -S . -B out/Release \
+  -DDAWN_FETCH_DEPENDENCIES=ON \
+  -DDAWN_ENABLE_INSTALL=ON \
+  -DCMAKE_BUILD_TYPE=Release \
+  -DDAWN_BUILD_TESTS=OFF \
+  -DDAWN_BUILD_SAMPLES=OFF \
+  -DTINT_BUILD_TESTS=OFF \
+  -DTINT_BUILD_CMD_TOOLS=OFF \
+  -DCMAKE_CXX_COMPILER=clang++
+cmake --build out/Release --parallel "${NUM_THREADS}"
+cmake --install out/Release --prefix install/Release
+
+echo "=== Building the artifact ==="
+cd "$SCRIPT_DIR"
+
+export CMAKE_PREFIX_PATH="$(realpath "$ROOT_DIR/dawn/install/Release")"
+
+cmake -S . -B out/Release -DCMAKE_BUILD_TYPE=Release -DCMAKE_CXX_COMPILER=clang++
[…]

### 2025-02-27 paper.tex
Grammar Mega-Pass

@@
 \begin{abstract}
-  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained-Scans} to run on hardware without \emph{Forward-Progress Guarantees} (FPG) while avoiding deadlock.
-Additionally, we introduce a tile state representation for \emph{Chained-Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation.
-On an FPG-lacking devices---Apple M1 Max and M3 ()---\emph{Decoupled-Fallback} achieves near-\emph{memcopy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach.
+  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{Forward-Progress Guarantees} (FPG) while avoiding deadlock.
+Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation.
+On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled-Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach.
 We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
@@
 This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$.
-\emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{Speed of Light} (SOL) performance limit---the theoretical maximum throughput constrained only by bandwidth and memory latency.
+\emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{Speed of Light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device.
 However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward progress guarantee}; without this guarantee, it risks indefinite starvation.
@@
 This work is guided by two main goals.
-(1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained-Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward progress guarantees (FPG) or support for 64-bit atomics.
+(1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward progress guarantees (FPG) or support for 64-bit atomics.
 \textbf{Performance:} We aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.
@@
 \textbf{Subgroup divergence} is another challenge we acknowledge but do not address directly, as shading languages do not provide explicit divergence control like CUDA's subgroup masking mechanisms.
-Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained-Scan}.
+Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained Scan}.
 
@@
 However, synchronization between workgroups is more limited.
-Standard GPU programming environments provide no control barrier between workgroups.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent-Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} The availability of a memory-only barrier varies across GPU environments; it is present in Direct3D and Vulkan, but was missing in versions of Metal prior to 3.2.
+Standard GPU programming environments provide no control barrier between workgroups.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} The availability of a memory-only barrier varies across GPU environments; it is present in Direct3D and Vulkan, but was missing in versions of Metal prior to 3.2.
 Because of its non-portability, it is also absent from the WebGPU specification.
@@
 Consequently, recent scan optimizations have shifted toward refining global strategies, particularly inter-workgroup coordination and synchronization.
-This section reviews three key approaches---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---analyzing their handling of inter-workgroup dependencies and efficiency in minimizing global memory access while maximizing bandwidth utilization.
+This section reviews three key approaches---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained Scan}---analyzing their handling of inter-workgroup dependencies and efficiency in minimizing global memory access while maximizing bandwidth utilization.
 
@@
 If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like scan can profit by trading additional computation for reduced memory traffic.
-This trade-off led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{10.1145/1375527.1375559, Merrill2009, 10.1109/TPDS.2012.336, 10.5555/2031978.2032029}: by storing only the work tile reduction in global memory during the first phase and performing a redundant scan in the third, it eliminates $n$ global data movement while preserving the three-phase structure:
+This trade-off led to a refinement of the \emph{Scan-then-Propagate} strategy called \emph{Reduce-then-Scan}~\cite{10.1145/1375527.1375559, Merrill2009, 10.1109/TPDS.2012.336, 10.5555/2031978.2032029}: by storing only the work tile reduction in global memory during the first phase and performing a redundant scan in the third, it eliminates $n$ global data movement while preserving the three-phase structure:
 \begin{enumerate}
@@
 
-\subsubsection{Chained-Scan, Stream-Scan}
-First introduced by Yan et al.~\cite{10.1145/2442516.2442539} in \emph{Stream-Scan}, the key innovation of the \emph{Chained-Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level.
-Instead of using kernel launches and pipeline barriers as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data.
+\subsubsection{Chained Scan, StreamScan}
+First introduced by Yan et al.~\cite{10.1145/2442516.2442539} in \emph{StreamScan}, the key innovation of the \emph{Chained Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level.
+Instead of using kernel launches and pipeline barriers as inter-workgroup synchronization points, \emph{Chained Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data.
 To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than arbitrary assignment based on virtualized workgroup indices.
@@
 
-In any \emph{Chained-Scan}, once a workgroup computes its local scan, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles.
+In any \emph{Chained Scan}, once a workgroup computes its local scan, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles.
 This process, which we refer to as \emph{Chained Propagation}, defines the set of inter-workgroup techniques that facilitate dependency resolution.
@@
 
-\subsubsection{Chained-Scan, Decoupled-Lookback}
-Introduced by Merrill and Garland, the \emph{Decoupled-Lookback} technique refines \emph{Chained Propagation} by separating reduction communication from dependency resolution, mitigating bottlenecks caused by propagation latency.
+\subsubsection{Chained Scan, Decoupled Lookback}
+Introduced by Merrill and Garland, the \emph{Decoupled Lookback} technique refines \emph{Chained Propagation} by separating reduction communication from dependency resolution, mitigating bottlenecks caused by propagation latency.
 In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction.
@@
 
-Although \emph{Decoupled-Lookback} increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing confines spine memory accesses to a roughly $o$-width sliding window, making them highly likely to be cache-resident.
+Although \emph{Decoupled Lookback} increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing confines spine memory accesses to a roughly $o$-width sliding window, making them highly likely to be cache-resident.
 As a result, the additional global memory traffic along the spine introduces negligible overhead.
-With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled-Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance without the work tile size tuning required by \emph{Stream-Scan}.
+With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance without the work tile size tuning required by \emph{Stream Scan}.
 
 \subsection{Ideal Scan Architecture}
-\emph{Chained Scan} and \emph{Reduce-Then-Scan} represent two contrasting paradigms in scan design.
+\emph{Chained Scan} and \emph{Reduce-then-Scan} represent two contrasting paradigms in scan design.
 By eliding the barrier-idiom, \emph{Chained Scan} aligns with the oversubscription paradigm, delegating occupancy management to the workgroup scheduler.
-In contrast, \emph{Reduce-Then-Scan} employs workgroup raking, requiring fixed launches and explicit occupancy control, forcing developers to manually tune launch sizes or implement occupancy discovery techniques.
+In contrast, \emph{Reduce-then-Scan} employs workgroup raking, requiring fixed launches and explicit occupancy control, forcing developers to manually tune launch sizes or implement occupancy discovery techniques.
 
@@
 While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple’s GPUs do not.
-Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
+Without forward progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
 
-\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
+\subsection{Why does \emph{Chained Scan} Rely on Forward-Progress Guarantees?}
 A \emph{forward-progress guarantee} (FPG) ensures that all occupant subgroups (of possibly different workgroups) eventually make progress toward termination, preventing indefinite starvation.
@@
 The absence of FPG in key hardware segments poses a critical challenge for developers.
-Our testing shows that, at best, running \emph{Chained-Scan} without FPG results in severe performance degradation, and at worst, risks \emph{Timeout Device Recovery} and program crashes.
+Our testing shows that, at best, running \emph{Chained Scan} without FPG results in severe performance degradation, and at worst, risks \emph{Timeout Device Recovery} and program crashes.
 \begin{figure}
@@
   \includegraphics[width=\linewidth]{graphics/Figure_1.pdf}
-  \caption{Execution times of \emph{Decoupled-Lookback (PLACEHOLDER DATA)}}
+  \caption{Execution times of \emph{Decoupled Lookback (PLACEHOLDER DATA)}}
 \end{figure}
@@
 This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware.
-In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-Then-Scan}, sacrificing performance for broader compatibility.
+In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.
 
 \subsection{Earlier Attempts at Portability}
-Our first attempt \raph{we'll add the cite post-review} at implementing a \emph{Decoupled-Lookback} scan without relying on FPG was \emph{Scalar Fallback}.
+Our first attempt \raph{we'll add the cite post-review} at implementing a \emph{Decoupled Lookback} scan without relying on FPG was \emph{Scalar Fallback}.
 Each time a workgroup spins waiting on the reduction from a preceding tile, it advances a \emph{fallback} operation by processing one scalar element from that tile.
-This eliminates the possibility of deadlock because the number of such operations is bouded; the wait will eventually terminate, possibly at the expense of doing redundant work.
+This eliminates the possibility of deadlock because the number of such operations is bounded; the wait will eventually terminate, possibly at the expense of doing redundant work.
 
@@
 Second, once the fallback completes, the lookback thread does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback.
-While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce then Scan}.
+While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce-then-Scan}.
 
@@
 
-The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled-Lookback}.
+The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled Lookback}.
 Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$\!, independent of FPG, due to false-positive fallback operations.
@@
 To reiterate, the goal of \emph{Chained Propagation} is twofold: first, to determine the reduction of all preceding work tiles, and second, to relay those values to succeeding tiles to accelerate their own lookback operations.
-Following the \emph{Decoupled-Lookback} technique, we model the inter-workgroup spine as a finite-state machine, where each work tile is represented by a reduction value and a state, stored in global memory.
+Following the \emph{Decoupled Lookback} technique, we model the inter-workgroup spine as a finite-state machine, where each work tile is represented by a reduction value and a state, stored in global memory.
 Each tile may exist in one of three states:
@@
 \subsection{Tile State Implementation}
-In all variants of \emph{Chained-Scan}, there is communication and synchronization between workgroups.
+In all variants of \emph{Chained Scan}, there is communication and synchronization between workgroups.
 The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it.
@@
 Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities.
-We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU\@.\footnote{To keep the perforamance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
+We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU\@.\footnote{To keep the performance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
 \begin{table}
@@
 \subsubsection{Same-API Performance Versus Baseline Measures}
-We next compare \emph{DF} against \emph{Reduce then Scan}\textdagger{~\ref{sec:rts}} (\emph{RTS}), which serves as a baseline to highlight the performance improvements of our approach over the prior FPG-constrained technique.
+We next compare \emph{DF} against \emph{Reduce-then-Scan}\textdagger{~\ref{sec:rts}} (\emph{RTS}), which serves as a baseline to highlight the performance improvements of our approach over the prior FPG-constrained technique.
 This evaluation is conducted using Dawn across the full range of tested devices.
@@
 
-Our reexamination of \emph{Chained Scan} with dynamic assignment raises fundamental questions about the GPU progress models proprosed by Sorensen et al.~\cite{sorensen2021}: Can all HSA algorithms be converted to OBE via dynamic assignment?
+Our reexamination of \emph{Chained Scan} with dynamic assignment raises fundamental questions about the GPU progress models proposed by Sorensen et al.~\cite{sorensen2021}: Can all HSA algorithms be converted to OBE via dynamic assignment?
 Are there cases, excluding atomic memory errors, where LOBE guarantees termination but dynamic assignment with OBE does not?
@@
 \subsection{Conclusion}
-In this paper, we presented \emph{Decoupled Fallback}, a single-pass \emph{Chained-Scan} capable of reaching SOL performance under all progress models and suitable for implementation within the constraints of contemporary shading languages.
-Our approach mantains minimal data movement while ensuring robust execution across diverse GPU architectures.
+In this paper, we presented \emph{Decoupled Fallback}, a single-pass \emph{Chained Scan} capable of reaching SOL performance under all progress models and suitable for implementation within the constraints of contemporary shading languages.
+Our approach maintains minimal data movement while ensuring robust execution across diverse GPU architectures.
 
@@
 Although all of Dawn's backend APIs support \emph{timestamp queries inside encoders}, Apple GPUs do not, so we are forced to implement our own \emph{Memcpy} kernel to measure memory bandwidth, as opposed to simply timing an API's copy buffer operation.
-It is well known that \emph{observed} peak memory bandwidth, tends to diverge from vendor-listed peak bandwidth, so for completeness we list these figures above.
+It is well known that \emph{observed} peak memory bandwidth tends to diverge from vendor-listed peak bandwidth, so for completeness we list these figures above.
 We make a number of observations:
@@
   \item \textbf{Nvidia 2080 Super:} Observed throughput matches expectations.
-  \item \textbf{AMD 7900 XT:} Our above result, where \emph{DF} was observed to be faster than \emph{Memcopy}, is likely due to compute latency counterintuitively improving memory access patterns, leading to reduced paging.
+  \item \textbf{AMD 7900 XT:} Our above result, where \emph{DF} was observed to be faster than \emph{Memcpy}, is likely due to compute latency counterintuitively improving memory access patterns, leading to reduced paging.
 Furthermore, we see that our \emph{DF} throughput is within spec and reasonable.
@@
 
-\subsection{Baseline Comparisons: \emph{Reduce then Scan}}%
+\subsection{Baseline Comparisons: \emph{Reduce-then-Scan}}
 \label{sec:rts}
-As we desire the most competitve baseline possible, our \emph{RTS} kernels employ the same intra-workgroup scan strategy and tuning parameters as \emph{DF}.
+As we desire the most competitive baseline possible, our \emph{RTS} kernels employ the same intra-workgroup scan strategy and tuning parameters as \emph{DF}.
 Initially, we adopted Merrill and Grimshaw's~\cite{Merrill2009} workgroup raking approach for inter-workgroup processing.

### 2025-02-27 paper.tex
Edits on the Starvation/Deadlocking Section

@@
 
-\usepackage{xcolor}
-
-\pagecolor[rgb]{0.5,0.5,0.5}
-\color[rgb]{1,1,1}
-
 \begin{document}
@@
 \begin{abstract}
-  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{Forward-Progress Guarantees} (FPG) while avoiding deadlock.
+  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{forward-progress guarantees} (FPG) while avoiding deadlock.
 Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation.
@@
 \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{Speed of Light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device.
-However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward progress guarantee}; without this guarantee, it risks indefinite starvation.
+However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite starvation.
 
@@
 However, portable implementation is constrained by both hardware and API limitations.
-In terms of hardware, some GPU schedulers—most notably Apple's M series and ARM—lack forward progress guarantees, and subgroup sizes vary across architectures.
+In terms of hardware, some GPU schedulers—most notably Apple's M series and ARM—lack forward-progress guarantees, and subgroup sizes vary across architectures.
 In terms of API design, the absence of 64-bit atomic operations and explicit memory barriers further complicates implementation, as both are integral to \emph{Decoupled Lookback}.
@@
 This work is guided by two main goals.
-(1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward progress guarantees (FPG) or support for 64-bit atomics.
+(1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward-progress guarantees (FPG) or support for 64-bit atomics.
 \textbf{Performance:} We aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.
@@
 
-We present \emph{Decoupled Fallback}, a fully portable prefix scan capable of achieving speed-of-light performance without requiring forward progress guarantees (FPG).
+We present \emph{Decoupled Fallback}, a fully portable prefix scan capable of achieving speed-of-light performance without requiring forward-progress guarantees (FPG).
 Our implementation is designed for compatibility with shading languages, making it suitable for modern graphics APIs and implemented in the WebGPU Shading Language (WGSL).
@@
 In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume.
-This approach depends on a \emph{forward progress guarantee} of the GPU's execution model.
+This approach depends on a forward-progress guarantee of the GPU's execution model.
 Given such a guarantee, the predecessor will eventually produce a result and broadcast it.
@@
 As a result, the additional global memory traffic along the spine introduces negligible overhead.
-With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance without the work tile size tuning required by \emph{Stream Scan}.
+With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance without the work tile size tuning required by \emph{StreamScan}.
 
@@
 While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple’s GPUs do not.
-Without forward progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
+Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
 
-\subsection{Why does \emph{Chained Scan} Rely on Forward-Progress Guarantees?}
-A \emph{forward-progress guarantee} (FPG) ensures that all occupant subgroups (of possibly different workgroups) eventually make progress toward termination, preventing indefinite starvation.
+\subsection{Progress Guarantees and Their Impact on \emph{Chained Scan}}
+A \emph{forward-progress guarantee} ensures that all occupant subgroups---of possibly different workgroups---eventually make progress toward termination, preventing indefinite starvation.
 Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees.
-Instead, Sorensen et al.\ systematically defined and tested three progress models:\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}
+Instead, Sorensen et al.\ systematically proposed and tested three progress models:\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}
 \begin{itemize}
   \item \textbf{Occupancy-Bound Execution (OBE)}: Workgroups that have made at least one execution step.
-  \item \textbf{Heterogeneous System Architecture (HSA)}: The workgroup with the lowest ID that has not yet terminated.
+  \item \textbf{Heterogeneous System Architecture~\cite{} (HSA)}: The workgroup with the lowest ID that has not yet terminated.
   \item \textbf{Linear OBE (LOBE)}: Workgroups that have made at least one step, and any workgroup with a lower ID than a workgroup that has made at least one step.
 \end{itemize}
-Their findings demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior assumptions~\cite{sorensen2018}.
+Their findings in in \cite{sorensen2021} demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior hypothesis in \cite{sorensen2018}.
 
-This has direct implications for \emph{Chained Scan}, which was \emph{previously} classified as a \emph{one-way producer-consumer} (PC) HSA algorithm: a workgroup only consumes from lower-ID counterparts, ensuring that execution proceeds in a strictly ordered fashion.
-Under HSA, this structure guarantees termination, as the lowest ID workgroup is always given forward progress.
-If subgroups from the lowest ID workgroup are not guaranteed execution, then any dependent subgroup may spin indefinitely, waiting on an update that never arrives.
+Sorensen et al.
+classify \emph{Chained Scan} as a \emph{one-way producer-consumer} (PC) HSA algorithm: a workgroup only consumes from lower-ID counterparts, ensuring that execution proceeds in a strictly ordered fashion.
+HSA guarantees termination, as the lowest ID workgroup is always given forward progress.
+If subgroups from the lowest ID workgroup are not guaranteed execution, then dependent workgroups may spin indefinitely, waiting on an update that never arrives.
 
@@
 
-\begin{table}
+Thus, by dynamically acquiring IDs through atomics, \emph{Chained Scan} shifts its progress model requirement from HSA to OBE.
+While LOBE encompasses both HSA and OBE, its stricter ordering guarantees are unnecessary for correctness.
+Ultimately, regardless of a device’s specific progress model, any architecture lacking forward-progress guarantees will fail to execute \emph{Chained Scan}.
+In short:
+
+\begin{table}[h!]
   \small
@@
     \toprule
-    Algorithm         & LOBE & OBE & HSA & Unfair \\
+    Algorithm         & LOBE & OBE & HSA & None \\
     \midrule
-    CS (Workgroup ID) & Yes  & No  & Yes & No     \\
-    CS (Atomic ID)    & Yes  & Yes & No  & No     \\
+    \emph{Chained Scan} with Workgroup ID & Yes  & No  & Yes & No     \\
+    \emph{Chained Scan} with Atomic ID    & Yes  & Yes & No  & No     \\
     Ours                                  & Yes  & Yes & Yes & Yes    \\
@@
   \end{tabular}
-  \caption{Comparison of Chained Scan (CS) Variants and Our Implementation Across Progress Models.\label{tab:progress_models}}
+  \caption{Comparison of \emph{Chained Scan} Variants and Our Implementation Across Progress Models.\label{tab:progress_models}}
 \end{table}
 
-\thomas{Is the one-to-one workgroup to multiprocessor rebuttal still neccessary here?}
-
-\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
+\subsection{Why is reliance on Forward-Progress Guarantees a Portability Problem?}
 The absence of FPG in key hardware segments poses a critical challenge for developers.
@@
 Another potential bottleneck is the speed of atomic operations.
-On some older GPUs—such as pre-Kepler NVIDIA architectures—atomics do not operate within the cache, resulting in significantly higher latency.
+On some older GPUs---such as pre-Fermi NVIDIA architectures---atomics do not operate within the cache, resulting in significantly higher latency.
 Slower atomics lead to delayed tile updates, which in turn increase lookback lengths.

### 2025-02-27 paper.tex
full paper read, added some comments esp. on results

@@
 Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation.
-On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled-Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach.
-We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
+On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach.
+We further demonstrate the resilience of \emph{Decoupled Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
 \end{abstract}
@@
 \section{Introduction}
-Scan is one of the most ubiquitous operations in GPGPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more.
+Scan is one of the most ubiquitous operations in GPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more.
 The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture.
@@
 \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{Speed of Light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device.
-However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite starvation.
+However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite starvation and will never complete.
+In practice, it is easy to demonstrate this behavior.
 
-As the compute capabilities of major graphics APIs---Vulkan, Direct3D, and Metal---have grown, there has been increasing interest in adapting \emph{Decoupled Lookback} beyond CUDA for broader adoption in real-time graphics and GPGPU computing.
+As the compute capabilities of major graphics APIs---Vulkan, Direct3D, and Metal---have grown, there has been increasing interest in adapting \emph{Decoupled Lookback} beyond CUDA for broader adoption in real-time graphics and GPU computing.
 However, portable implementation is constrained by both hardware and API limitations.
@@
 We present \emph{Decoupled Fallback}, a fully portable prefix scan capable of achieving speed-of-light performance without requiring forward-progress guarantees (FPG).
-Our implementation is designed for compatibility with shading languages, making it suitable for modern graphics APIs and implemented in the WebGPU Shading Language (WGSL).
-Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.
+Our implementation is designed for compatibility with shading languages, making it suitable for modern graphics APIs and implemented in the WebGPU Shading Language (WGSL)\@.
+\john{I don't love ``compatibility with shading languages'' but then we name a single shading language.
+Instead, maybe something like ``Our implementation targets portability across modern GPU hardware and software platforms'', that sort of thing?} Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.
 
@@
 
-On GPU hardware, scheduling is divided into two levels: a \emph{workgroup} scheduler, which manages kernel launches and maps virtual processors, workgroups, to physical processors, \emph{multiprocessors}, and a \emph{subgroup} scheduler, responsible for selecting which of the currently \emph{occupant} subgroups on a multiprocessor for execution.
+On GPU hardware, scheduling is divided into two levels: a \emph{workgroup} scheduler, which manages kernel launches and maps virtual processors, workgroups, to physical processors, \emph{multiprocessors}, and a \emph{subgroup} scheduler, responsible for selecting a subset of the currently \emph{occupant} subgroups on a multiprocessor for execution.
 The GPU programming model virtualizes processors to ensure GPU programs---\emph{kernels}---remain portable across hardware with differing physical resources.
@@
 As a result, a single multiprocessor may host multiple workgroups simultaneously.
-However, as a consequence of virtualization, the developer relinquishes control over the order in which workgroups are launched, and once a workgroup begins execution on a multiprocessor, it must run to completion and cannot context switch.
+However, as a consequence of virtualization, the developer relinquishes control over the order in which workgroups are launched, and once a workgroup begins execution on a multiprocessor, it must run to completion and cannot context-switch.
 
@@
 Although this can be mitigated by increasing the size of the work tile, limited on-chip resources preclude scaling the tile size indefinitely.
-Moreover, this approach is unsuitable for a portable environment, as excessively large work tiles risk reducing occupancy on less capable hardware.
+Moreover, this approach is unsuitable from a portability perspective, as excessively large work tiles risk reducing occupancy on less capable hardware.
 
@@
 
-Hardware-level considerations underscore the limitations of fixed-launches.
-On GPUs where the global memory data cache is indexed by physical addresses---and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)---fixed-launches can lead to performance degradation due to cache thrashing.
+Hardware-level considerations underscore the limitations of the fixed-launch approach.
+On GPUs where the global memory data cache is indexed by physical addresses---and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)---fixed launches can lead to performance degradation due to cache thrashing.
 This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups.
@@
 Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture.
-Furthermore, its single-kernel design and serial ordering make it suitable for \emph{in-situ} operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging.
+Furthermore, its single-kernel design and serial ordering make it suitable for in-situ operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging.
 However, it has a critical limitation: its reliance on a forward-progress guarantee.
@@
   \item \textbf{Occupancy-Bound Execution (OBE)}: Workgroups that have made at least one execution step.
-  \item \textbf{Heterogeneous System Architecture~\cite{} (HSA)}: The workgroup with the lowest ID that has not yet terminated.
+  \item \textbf{Heterogeneous System Architecture~\cite[Section~2.11]{HSASysArch12} (HSA)}: The workgroup with the lowest ID that has not yet terminated.
   \item \textbf{Linear OBE (LOBE)}: Workgroups that have made at least one step, and any workgroup with a lower ID than a workgroup that has made at least one step.
 \end{itemize}
-Their findings in in \cite{sorensen2021} demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior hypothesis in \cite{sorensen2018}.
+Their 2021 findings~\cite{sorensen2021} demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior hypothesis~\cite{sorensen2018}.
 
-Sorensen et al.
-classify \emph{Chained Scan} as a \emph{one-way producer-consumer} (PC) HSA algorithm: a workgroup only consumes from lower-ID counterparts, ensuring that execution proceeds in a strictly ordered fashion.
+Sorensen et al.\ classify \emph{Chained Scan} as a \emph{one-way producer-consumer} (PC) HSA algorithm: a workgroup only consumes from lower-ID counterparts, ensuring that execution proceeds in a strictly ordered fashion.
 HSA guarantees termination, as the lowest ID workgroup is always given forward progress.
@@
 
-\subsection{Why is reliance on Forward-Progress Guarantees a Portability Problem?}
+\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
 The absence of FPG in key hardware segments poses a critical challenge for developers.
@@
 \emph{Decoupled Fallback} addresses both inefficiencies.
-The key advance is to utilize the parallelism available in the workgroup to perform the fallback computation, rather than performing it sequentially.
-To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained-Propagation} logic is typically performed by the first thread.
+Our key advance is to utilize the parallelism available in the workgroup to perform the fallback computation, rather than performing it sequentially.
+To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained Propagation} logic is typically performed by the first thread.
 
@@
 
-The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled Lookback}.
+The viability of \emph{Decoupled Fallback} relies on the same cache locality properties as \emph{Decoupled Lookback}.
 Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$\!, independent of FPG, due to false-positive fallback operations.
@@
 Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy.
-Phase 4, \emph{Decoupled-Fallback}, is responsible for managing all inter-workgroup dependencies.
+Phase 4, \emph{Decoupled Fallback}, is responsible for managing all inter-workgroup dependencies.
 As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations.
@@
 Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking.
-Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.\footnote{We may also obtain a workgroup-wide reduce operation by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}
+Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.\footnote{We may also obtain a workgroup-wide reduce operation by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.\john{I don't get this footnote.
+``May also''?
+What are we actually doing?
+What's the point of this footnote?}}
 
-\subsection{Decoupled-Fallback}
+\subsection{Decoupled Fallback}
 To reiterate, the goal of \emph{Chained Propagation} is twofold: first, to determine the reduction of all preceding work tiles, and second, to relay those values to succeeding tiles to accelerate their own lookback operations.
@@
 To coordinate this process across the workgroup, each workgroup initializes a shared memory \emph{control flag} during partition index acquisition.
-This flag governs both the initiation of fallback operations and the entry and exit of the entire \emph{Decoupled-Fallback} phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.
+This flag governs both the initiation of fallback operations and the entry and exit of the entire \emph{Decoupled Fallback} phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.
 
-We now detail the \emph{Decoupled-Fallback} procedure, presenting the algorithm from the perspective of a single workgroup.
+We now detail the \emph{Decoupled Fallback} procedure, presenting the algorithm from the perspective of a single workgroup.
 The details of tile state encoding and atomic updates are discussed in the following section and may vary depending on the data type, so for simplicity, we describe the operation as being performed by the \emph{first thread} of the workgroup:
@@
           \item If it is the first work tile, its local reduction \emph{is} its inclusive reduction.
-In this case, the workgroup updates its tile state directly to \emph{Inclusive} and bypasses the \emph{Chained-Propagation} phase entirely.
+In this case, the workgroup updates its tile state directly to \emph{Inclusive} and bypasses the \emph{Chained Propagation} phase entirely.
         \end{itemize}
@@
         \begin{itemize}
-          \item The control flag, initialized to \emph{Locked}, signals that the workgroup should enter the \emph{Chained-Propagation} phase.
+          \item The control flag, initialized to \emph{Locked}, signals that the workgroup should enter the \emph{Chained Propagation} phase.
           \item The first thread initializes a \emph{previous reduction} variable in registers with $\varepsilon$ to hold the accumulated reductions of preceding tiles.
@@
 The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it.
-This primitive is well studied in the context of memory models~\cite{Alglave0, MCMutants}.
-The key requirement is both control synchronization, meaning the reader does not proceed until the writer sends the message; and data synchronization, meaning that the reader reads the correct value.
+This primitive is well-studied in the context of memory models~\cite{Alglave0, MCMutants}.
+The key requirements here are both control synchronization (meaning the reader does not proceed until the writer sends the message) and data synchronization (meaning that the reader reads the correct value).
 The latter is especially challenging in relaxed memory models as are employed on GPUs.
@@
 Without some other mechanism, the requirement to pack tile state and data into a single word limits the data to 30 bits (assuming 2 bits for the tile state).
-That, in turn, makes it impossible to represent common ordinary 32-bit floating point values without loss of precision.
-While some GPUs support 64 bit atomics (and it is likely that WebGPU will include them as an optional extension), that is also sadly not portable, and merely raises the payload limit to 62 bits without eliminating it altogether.
+That, in turn, makes it impossible to represent ordinary 32-bit floating point values without loss of precision.
+While some GPUs support 64 bit atomics (and it is likely that WebGPU will someday include them as an optional extension), that is also sadly not portable, and merely raises the payload limit to 62 bits without eliminating it altogether.
 
@@
 For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum.
-Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities.
+Our test hardware, listed POSITIONAL-RELATION\john{what is $\leftarrow$ POSITIONAL-RELATION\@?
+Reference the table!}, spans a wide range of FPG and performance capabilities.
 We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU\@.\footnote{To keep the performance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
@@
 We first compare \emph{Decoupled Fallback} (\emph{DF}) to two key benchmarks: \emph{Memcpy}\textdagger{~\ref{sec:memcpy}} and the \emph{Decoupled Lookback} (\emph{DL}) implementation found in NVIDIA’s CUB library, using the 2080 Super only.
-To isolate the impact of differing APIs, we implement and evaluate \emph{DF} in both CUDA and Dawn.\footnote{Due to WebGPU’s 128MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.} \emph{Memcpy} serves as a performance reference, as it shares the same $2n$ global memory movement as \emph{DF} but lacks compute overhead and is purely memory bandwidth bound.
+To isolate the impact of differing APIs, we implement and evaluate \emph{DF} in both CUDA and Dawn.\footnote{Due to WebGPU’s 128~MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.} \emph{Memcpy} serves as a performance reference, as it shares the same $2n$ global memory movement as \emph{DF} but lacks compute overhead and is purely memory bandwidth bound.
 
 Cross-API comparisons remain challenging, as our CUDA implementation consistently outperforms its Dawn counterpart despite near-identical high-level implementations.
-The performance cost of \emph{DF} is most pronounced in CUDA when the entire input fits within the 2080 Super’s 4MB L2 cache, where reduced global memory latency exposes \emph{DF's} increased synchronization and compute overhead compared to CUB\@.
+The performance cost of \emph{DF} is most pronounced in CUDA when the entire input fits within the 2080 Super’s 4~MB L2 cache, where reduced global memory latency exposes \emph{DF's} increased synchronization and compute overhead compared to CUB\@.
 However, at larger input sizes, where memory accesses exceed cache capacity, \emph{DF} reaches parity with CUB and achieves \emph{Memcpy} speeds.
@@
     \toprule
-    Metric & Intel HD620 & ARM Mali-G78 MP20 & Apple M3 & Apple M1 Max & Nvidia 2080 Super & AMD 7900 XT \\
+    Metric & \!\!\!\!\!\!Intel HD620 & ARM Mali-G78 MP20 & Apple M3 & Apple M1 Max & Nvidia 2080 Super & AMD 7900 XT \\
     \midrule
@@
     \emph{DF}/\emph{RTS}          & 140.2\%   & 138.5\%  & 150.2\%   & 154.5\%   & 148.4\%   & 153.8\%   \\
-    Avg Spins per Wg    & 0.305    & 1.477    & 0.606    & 0.604    & 0.419    & 0.148    \\
-    Avg Lookback Length per Wg  & 1.726    & 3.537    & 3.415    & 10.782   & 13.046   & 20.722   \\
+    Avg Spins per workgroup    & 0.305    & 1.477    & 0.606    & 0.604    & 0.419    & 0.148    \\
+    Avg Lookback Length per workgroup& 1.726    & 3.537    & 3.415    & 10.782   & 13.046   & 20.722   \\
     Avg Total Fallbacks Initiated      & 254.662   & 2504.704  & 365.546   & 990.962   & 86.400   & 9.308   \\
@@
 On the ARM Mali, \emph{DF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%.
-However, its nearly identical performance to \emph{Memcpy}, indicates that the implementation is memory-bandwidth-bound, limiting further gains.
+However, its nearly identical performance to \emph{Memcpy} indicates that the implementation is memory-bandwidth-bound, limiting further gains.
 
 \subsection{Decoupled Fallback Statistics}
-To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution:
+To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution: \john{Below, I advocate that we say, for each of the four bullets, what we expect / what would be a good result / what the hardware actually does in practice.
+The reader doesn't deeply know what to expect / what is good.}
 \begin{itemize}
   \item \textbf{Spins per Workgroup:} How many times does a workgroup make a waiting spin?
-  \item \textbf{Lookback Length per Workgroup:} How many work tiles does a workgroup traverse to complete \emph{Chained-Propagation}?
+  \item \textbf{Lookback Length per Workgroup:} How many work tiles does a workgroup traverse to complete \emph{Chained Propagation}?
+\john{Example for above point: something like ``We hope that even with unfair hardware schedulers, this traversal is short (only a few tiles), but we also want to ensure that DF is robust and delivers good performance for even long traversals.''}
   \item \textbf{Total Fallbacks Initiated:} What is the total number of fallback operations initiated across all workgroups?
-  \item \textbf{Total Successful Insertions:} What is total number of fallback operations that successfully update a blocking (or late) tile to \emph{Ready}?
+  \item \textbf{Total Successful Insertions:} What is the total number of fallback operations that successfully update a blocking (or late) tile to \emph{Ready}?
 \end{itemize}
 Despite differing occupancy rates, the number of waiting spins remains consistent across devices.
-Likewise, lookback length is stable and remains well below the theoretical occupancy bound.
+\john{Is this a small number?
+Large?
+Tell don't show.} Likewise, lookback length is stable and remains well below the theoretical occupancy bound.
 Expressed as a percentage of occupancy, it remains under 10\% for all devices, except for the Intel HD620, where its elevated value is likely an artifact of its low absolute occupancy.
@@
 As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near 100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced.
-Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%.
+Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%; \emph{DF} is consistently faster than the previous standard for portable scans.
 While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that we are inducing fallbacks, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior.
@@
 \subsection{Conclusion}
-In this paper, we presented \emph{Decoupled Fallback}, a single-pass \emph{Chained Scan} capable of reaching SOL performance under all progress models and suitable for implementation within the constraints of contemporary shading languages.
+In this paper, we presented \emph{Decoupled Fallback}, a single-pass \emph{Chained Scan} capable of reaching SOL performance under all progress models and suitable for implementation within the constraints of contemporary portable shading languages.
 Our approach maintains minimal data movement while ensuring robust execution across diverse GPU architectures.

### 2025-02-27 paper.tex
#18 stab at CCS and keywords

@@
   <concept>
-  <concept_id>00000000.0000000.0000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
+  <concept_id>10010147.10010169.10010170.10010174</concept_id>
+  <concept_desc>Computing methodologies~Massively parallel algorithms</concept_desc>
   <concept_significance>500</concept_significance>
@@
   <concept>
-  <concept_id>00000000.00000000.00000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-  <concept_significance>300</concept_significance>
+  <concept_id>10010520.10010521.10010528</concept_id>
+  <concept_desc>Computer systems organization~Parallel architectures</concept_desc>
+  <concept_significance>500</concept_significance>
   </concept>
   <concept>
-  <concept_id>00000000.00000000.00000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-  <concept_significance>100</concept_significance>
+  <concept_id>10003752.10003753.10003761.10003762</concept_id>
+  <concept_desc>Theory of computation~Parallel computing models</concept_desc>
+  <concept_significance>500</concept_significance>
   </concept>
   <concept>
-  <concept_id>00000000.00000000.00000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-  <concept_significance>100</concept_significance>
+  <concept_id>10003752.10003753.10003761</concept_id>
+  <concept_desc>Theory of computation~Concurrency</concept_desc>
+  <concept_significance>500</concept_significance>
+  </concept>
+  <concept>
+  <concept_id>10010147.10010169.10010175</concept_id>
+  <concept_desc>Computing methodologies~Parallel programming languages</concept_desc>
+  <concept_significance>500</concept_significance>
   </concept>
@@
 
-\ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-\ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-\ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-\ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
+\ccsdesc[500]{Computing methodologies~Massively parallel algorithms}
+\ccsdesc[500]{Computer systems organization~Parallel architectures}
+\ccsdesc[500]{Theory of computation~Parallel computing models}
+\ccsdesc[500]{Theory of computation~Concurrency}
+\ccsdesc[500]{Computing methodologies~Parallel programming languages}
 
-\keywords{Do, Not, Us, This, Code, Put, the, Correct, Terms, for,
-  Your, Paper}
+\keywords{scan, parallel prefix, prefix sum, forward-progress guarantee, GPU, WGSL, decoupled fallback}
 

### 2025-02-27 paper.tex
pics are smaller
caption for figure 1

@@
 However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite starvation and will never complete.
-In practice, it is easy to demonstrate this behavior.
+In practice, it is easy to demonstrate this highly undesirable behavior.
 
@@
   \centering
-  \includegraphics[width=\linewidth]{graphics/Figure_1.pdf}
-  \caption{Execution times of \emph{Decoupled Lookback (PLACEHOLDER DATA)}}
+  \includegraphics[width=0.6\linewidth]{graphics/Figure_1.pdf}
+  \caption{Execution times of \emph{Decoupled Lookback} (PLACEHOLDER DATA).
+The TDR threshold represents the duration after which the OS, assuming a non-terminating kernel, kills the driver.
+\john{This figure needs to be cited in the text.}}
 \end{figure}
@@
   \centering
-  \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
-  \caption{Performance versus CUB (NVIDIA 2080 Super)}
+  \includegraphics[width=0.6\linewidth]{graphics/cuda_plot.pdf}
+  \caption{Performance versus CUB (NVIDIA 2080 Super).}
 \end{figure}

### 2025-02-28 paper.tex
Add distribution graph. Rollback ternary changes.

@@
 
-\subsection{The Scan Primitive}%
+\subsection{The Scan Primitive}
 \label{sec:the-scan-primitive}
@@
 The absence of FPG in key hardware segments poses a critical challenge for developers.
-Our testing shows that, at best, running \emph{Chained Scan} without FPG results in severe performance degradation, and at worst, risks \emph{Timeout Device Recovery} and program crashes.
+Our testing ~\ref{fig:dist} shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\footnote{Operating systems handle unresponsive programs differently.
+On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds.
+macOS has a timeout watchdog that, interestingly, does not terminate a spinning kernel---allowing us to obtain these results.
+On Linux, behavior varies by GPU driver.} (TDR), crashing the program.
 \begin{figure}
   \centering
-  \includegraphics[width=0.6\linewidth]{graphics/Figure_1.pdf}
-  \caption{Execution times of \emph{Decoupled Lookback} (PLACEHOLDER DATA).
-The TDR threshold represents the duration after which the OS, assuming a non-terminating kernel, kills the driver.
-\john{This figure needs to be cited in the text.}}
+  \includegraphics[width=0.6\linewidth]{graphics/dist.pdf}
+  \caption{Execution times of \emph{Decoupled Lookback} on an FPG-lacking device, Apple M1 Max.
+The TDR threshold represents the duration after which the OS, assuming a non-terminating kernel, kills the driver.\label{fig:dist}}
 \end{figure}
@@
 Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking.
-Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.\footnote{We may also obtain a workgroup-wide reduce operation by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.\john{I don't get this footnote.
-``May also''?
-What are we actually doing?
-What's the point of this footnote?}}
+Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.
 
@@
           \item Using the tile index provided by the first thread, the full workgroup computes the reduction of the blocking tile by directly processing the corresponding elements.
-This process involves a \emph{workgroup-wide reduction},\footnote{See footnote NUMBER.} utilizing additional shared memory to aggregate partial subgroup results.
+This process involves a \emph{workgroup-wide reduction},\footnote{We obtain the workgroup-wide reduce by omitting the fanout operation of the workgroup-wide scan and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.} utilizing additional shared memory to aggregate partial subgroup results.
           \item Once the fallback reduction is complete, the workgroup \emph{attempts} to post the reduction to global memory and update the tile's state to \emph{Ready}.
@@
 For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum.
-Our test hardware, listed POSITIONAL-RELATION\john{what is $\leftarrow$ POSITIONAL-RELATION\@?
-Reference the table!}, spans a wide range of FPG and performance capabilities.
-We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU\@.\footnote{To keep the performance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
+Our test hardware, shown in ~\ref{tab:hardware}, spans a wide range of FPG and performance capabilities.
+We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires native Android, necessitating the use of WGPU\@.\footnote{To keep the performance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
 \begin{table}

### 2025-02-28 paper.tex
adding 'Figure' to our figure refs

@@
 However, it has a critical limitation: its reliance on a forward-progress guarantee.
-While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple’s GPUs do not.
+While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple's GPUs do not.
 Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
@@
 While LOBE encompasses both HSA and OBE, its stricter ordering guarantees are unnecessary for correctness.
-Ultimately, regardless of a device’s specific progress model, any architecture lacking forward-progress guarantees will fail to execute \emph{Chained Scan}.
+Ultimately, regardless of a device's specific progress model, any architecture lacking forward-progress guarantees will fail to execute \emph{Chained Scan}.
 In short:
@@
 The absence of FPG in key hardware segments poses a critical challenge for developers.
-Our testing ~\ref{fig:dist} shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\footnote{Operating systems handle unresponsive programs differently.
+Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\footnote{Operating systems handle unresponsive programs differently.
 On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds.
@@
 For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum.
-Our test hardware, shown in ~\ref{tab:hardware}, spans a wide range of FPG and performance capabilities.
+Our test hardware, shown in Figure~\ref{tab:hardware}, spans a wide range of FPG and performance capabilities.
 We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires native Android, necessitating the use of WGPU\@.\footnote{To keep the performance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
@@
 \subsubsection{Cross-API Performance Versus Contemporary Implementations}
-We first compare \emph{Decoupled Fallback} (\emph{DF}) to two key benchmarks: \emph{Memcpy}\textdagger{~\ref{sec:memcpy}} and the \emph{Decoupled Lookback} (\emph{DL}) implementation found in NVIDIA’s CUB library, using the 2080 Super only.
-To isolate the impact of differing APIs, we implement and evaluate \emph{DF} in both CUDA and Dawn.\footnote{Due to WebGPU’s 128~MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.} \emph{Memcpy} serves as a performance reference, as it shares the same $2n$ global memory movement as \emph{DF} but lacks compute overhead and is purely memory bandwidth bound.
+We first compare \emph{Decoupled Fallback} (\emph{DF}) to two key benchmarks: \emph{Memcpy}\textdagger{~\ref{sec:memcpy}} and the \emph{Decoupled Lookback} (\emph{DL}) implementation found in NVIDIA's CUB library, using the 2080 Super only.
+To isolate the impact of differing APIs, we implement and evaluate \emph{DF} in both CUDA and Dawn.\footnote{Due to WebGPU's 128~MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.} \emph{Memcpy} serves as a performance reference, as it shares the same $2n$ global memory movement as \emph{DF} but lacks compute overhead and is purely memory bandwidth bound.
 
 Cross-API comparisons remain challenging, as our CUDA implementation consistently outperforms its Dawn counterpart despite near-identical high-level implementations.
-The performance cost of \emph{DF} is most pronounced in CUDA when the entire input fits within the 2080 Super’s 4~MB L2 cache, where reduced global memory latency exposes \emph{DF's} increased synchronization and compute overhead compared to CUB\@.
+The performance cost of \emph{DF} is most pronounced in CUDA when the entire input fits within the 2080 Super's 4~MB L2 cache, where reduced global memory latency exposes \emph{DF's} increased synchronization and compute overhead compared to CUB\@.
 However, at larger input sizes, where memory accesses exceed cache capacity, \emph{DF} reaches parity with CUB and achieves \emph{Memcpy} speeds.
@@
 Sorting is one of the most notable beneficiaries of single-pass scan and, by extension, \emph{Decoupled Fallback}.
-The current state-of-the-art algorithm, Adinets and Merrill’s \emph{Onesweep}~\cite{adinets2022onesweepfastersignificantdigit}, is an 8-bit least-significant-digit radix sort, where single-pass scan minimizes data movement during digit-binning passes.
+The current state-of-the-art algorithm, Adinets and Merrill's \emph{Onesweep}~\cite{adinets2022onesweepfastersignificantdigit}, is an 8-bit least-significant-digit radix sort, where single-pass scan minimizes data movement during digit-binning passes.
 However, due to its compute-intensive nature and large $radixDigit$-sized monoid, it is unclear whether \emph{Decoupled Fallback} will map well to sorting or if new techniques are needed.

### 2025-02-28 paper.tex
making fig 1 caption accurate

@@
 The absence of FPG in key hardware segments poses a critical challenge for developers.
-Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\footnote{Operating systems handle unresponsive programs differently.
+Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently.
 On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds.
@@
   \caption{Execution times of \emph{Decoupled Lookback} on an FPG-lacking device, Apple M1 Max.
-The TDR threshold represents the duration after which the OS, assuming a non-terminating kernel, kills the driver.\label{fig:dist}}
+The TDR threshold here is Windows's threshold for resetting the GPU\@; as noted in Section~\ref{sec:tdr}, MacOS handles this differently.\label{fig:dist}}
 \end{figure}

### 2025-02-28 paper.tex
Add comparison graph (slam dunk). Loosen TDR language.

@@
 On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds.
-macOS has a timeout watchdog that, interestingly, does not terminate a spinning kernel---allowing us to obtain these results.
-On Linux, behavior varies by GPU driver.} (TDR), crashing the program.
+Other operating systems---such as macOS and Linux---employ their own timeout mechanisms with varying behaviors.
+For consistency, we adopt Windows’ 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program.
 \begin{figure}
   \centering
-  \includegraphics[width=0.6\linewidth]{graphics/dist.pdf}
-  \caption{Execution times of \emph{Decoupled Lookback} on an FPG-lacking device, Apple M1 Max.
-The TDR threshold here is Windows's threshold for resetting the GPU\@; as noted in Section~\ref{sec:tdr}, MacOS handles this differently.\label{fig:dist}}
+  \begin{subfigure}[b]{0.45\linewidth}
+    \includegraphics[width=\linewidth]{graphics/decoupled_lookback.pdf}
+    \caption{\emph{Decoupled Lookback}}
+    \label{fig:decoupled}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.45\linewidth}
+    \includegraphics[width=\linewidth]{graphics/ours.pdf}
+    \caption{Ours}
+    \label{fig:ours}
+  \end{subfigure}
+  \caption{Histogram plots, Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements, on an FPG-lacking device, Apple M1 Max.
+The TDR threshold is Windows's threshold for resetting the GPU\@; as noted in Section~\ref{sec:tdr}, MacOS handles this differently.}
+  \label{fig:dist}
 \end{figure}
@@
   \end{tabular*}
-  \caption{\emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements.\label{tab:results}}
+  \caption{\emph{Decoupled Fallback} performance, Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements.\label{tab:results}}
 \end{table*}

### 2025-02-28 paper.tex
massaging caption, great plot

@@
     \includegraphics[width=\linewidth]{graphics/decoupled_lookback.pdf}
-    \caption{\emph{Decoupled Lookback}}
+    \caption{\emph{Decoupled Lookback.}}
     \label{fig:decoupled}
@@
     \includegraphics[width=\linewidth]{graphics/ours.pdf}
-    \caption{Ours}
+    \caption{\emph{Decoupled Fallback (ours).}}
     \label{fig:ours}
   \end{subfigure}
-  \caption{Histogram plots, Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements, on an FPG-lacking device, Apple M1 Max.
+  \caption{Histogram plots, 500 runs of Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements, on an FPG-lacking device, Apple M1 Max.
+Bar heights ($y$ axis) indicate the number of runs that completed in the execution time indicated on the $x$ axis.
 The TDR threshold is Windows's threshold for resetting the GPU\@; as noted in Section~\ref{sec:tdr}, MacOS handles this differently.}

### 2025-02-28 paper.tex
Respond to John Feedback results section.

@@
 This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$.
-\emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{Speed of Light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device.
+\emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{speed-of-light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device.
 However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite starvation and will never complete.
@@
 However, portable implementation is constrained by both hardware and API limitations.
-In terms of hardware, some GPU schedulers—most notably Apple's M series and ARM—lack forward-progress guarantees, and subgroup sizes vary across architectures.
+In terms of hardware, some GPU schedulers---most notably Apple's M series and ARM---lack forward-progress guarantees, and subgroup sizes vary across architectures.
 In terms of API design, the absence of 64-bit atomic operations and explicit memory barriers further complicates implementation, as both are integral to \emph{Decoupled Lookback}.
@@
 (1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward-progress guarantees (FPG) or support for 64-bit atomics.
-\textbf{Performance:} We aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.
+\textbf{Performance:} We aim for neear SOL execution to the greatest extent possible allowed by the underlying hardware and programming model.
 
 However, certain aspects fall outside the scope of this work.
-\textbf{Performance guarantees} are not universally attainable, as not all architectures offer atomic operations that are sufficiently fast or scheduling models that are sufficiently fair to attain speed-of-light performance.
+\textbf{Performance guarantees} are not universally attainable, as not all architectures offer atomic operations that are sufficiently fast or scheduling models that are sufficiently fair to attain SOL performance.
 \textbf{Subgroup divergence} is another challenge we acknowledge but do not address directly, as shading languages do not provide explicit divergence control like CUDA's subgroup masking mechanisms.
@@
 
-We present \emph{Decoupled Fallback}, a fully portable prefix scan capable of achieving speed-of-light performance without requiring forward-progress guarantees (FPG).
-Our implementation is designed for compatibility with shading languages, making it suitable for modern graphics APIs and implemented in the WebGPU Shading Language (WGSL)\@.
-\john{I don't love ``compatibility with shading languages'' but then we name a single shading language.
-Instead, maybe something like ``Our implementation targets portability across modern GPU hardware and software platforms'', that sort of thing?} Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.
+We present \emph{Decoupled Fallback}, a fully portable prefix scan capable of achieving SOL performance without requiring forward-progress guarantees (FPG).
+Our implementation targets portability across modern GPU hardware and software platforms.
+It is implemented in the WebGPU Shading Language (WGSL)---an apex-level shading language that transpiles to backend API languages---ensuring seamless integration with modern graphics APIs.
+Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.
 
@@
 
-While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently saturate global memory bandwidth, preventing it from achieving speed-of-light performance.
+While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently saturate global memory bandwidth, preventing it from achieving SOL performance.
 This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the result of its immediate predecessor.
@@
 As a result, the additional global memory traffic along the spine introduces negligible overhead.
-With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance without the work tile size tuning required by \emph{StreamScan}.
+With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled Lookback} achieves full utilization of device memory bandwidth and delivers SOL performance without the work tile size tuning required by \emph{StreamScan}.
 
@@
 Second, once the fallback completes, the lookback thread does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback.
-While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce-then-Scan}.
+While \emph{Scalar Fallback} achieved SOL performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce-then-Scan}.
 
@@
   \includegraphics[height=\textheight]{graphics/FlowChart.pdf}
-  \caption{Flow Chart Representation of Decoupled Fallback.\label{fig:decoupled_fallback}}
+  \caption{Flow Chart Representation of \emph{Decoupled Fallback}.\label{fig:decoupled_fallback}}
 \end{figure}
@@
 \subsection{Decoupled Fallback Statistics}
-To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution: \john{Below, I advocate that we say, for each of the four bullets, what we expect / what would be a good result / what the hardware actually does in practice.
-The reader doesn't deeply know what to expect / what is good.}
+To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution:
+
 \begin{itemize} 
-  \item \textbf{Spins per Workgroup:} How many times does a workgroup make a waiting spin?
+  \item \textbf{Spins per Workgroup:} How many times does a workgroup spin while waiting?
+With perfectly fair scheduling, no waiting is necessary.
+Since no real-world scheduler is perfectly fair, we want minimal waiting.
+Despite varying occupancy rates, the average spin rate across tested devices remains low (typically one or fewer), indicating that most workgroups experience minimal delays.
+
   \item \textbf{Lookback Length per Workgroup:} How many work tiles does a workgroup traverse to complete \emph{Chained Propagation}?
-\john{Example for above point: something like ``We hope that even with unfair hardware schedulers, this traversal is short (only a few tiles), but we also want to ensure that DF is robust and delivers good performance for even long traversals.''}
-  \item \textbf{Total Fallbacks Initiated:} What is the total number of fallback operations initiated across all workgroups?
-  \item \textbf{Total Successful Insertions:} What is the total number of fallback operations that successfully update a blocking (or late) tile to \emph{Ready}?
+Ideally, this traversal should be minimal, but longer traversal may result from unfair scheduling.
+Empirically, however, our data shows that lookback length remains well below the theoretical $O(o)$ bound---typically under 10\% of occupancy---meaning that most workgroups complete \emph{Chained Propagation} in relatively few steps compared to occupancy.
+
+  \item \textbf{Total Fallbacks Initiated:} The total number of fallback operations initiated across all workgroups.
+Ideally, this number would be zero on devices with FPG; because multiple workgroups may perform a fallback on the same blocking tile, this number can exceed the number of workgroups launched.
+In practice, this count is highly dependent on the fairness of the hardware scheduler.
+On most devices tested, initiated fallbacks remain modest---about 10\% or fewer of the total workgroups launched.
+
+  \item \textbf{Total Successful Insertions:} The total number of fallback operations that successfully update a blocking (or late) tile to \emph{Ready}.
+Due to false-positive fallbacks, this count does not precisely represent the number of true starvations, but instead provides an upper bound.
+Ideally, this number would also be zero on devices with FPG, but in practice, it is again dependent on the fairness of the scheduler.
+Unlike the number of fallback initiations, this statistic is extremely low---typically well under 1\% even on devices without FPG---suggesting that true starvation is rare but catastrophic.
 \end{itemize}
-Despite differing occupancy rates, the number of waiting spins remains consistent across devices.
-\john{Is this a small number?
-Large?
-Tell don't show.} Likewise, lookback length is stable and remains well below the theoretical occupancy bound.
-Expressed as a percentage of occupancy, it remains under 10\% for all devices, except for the Intel HD620, where its elevated value is likely an artifact of its low absolute occupancy.
-Most notably, the devices lacking FPG---Apple M3 and M1 Max---do not exhibit significantly higher successful insertion rates than the others, suggesting that true deadlocking is a rare occurrence.
+
+Furthermore, this data reveals two notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, indicating that fairness and FPG are not necessarily correlated.
+(2) The ARM Mali-G78 MP20 exhibits the most unfair scheduling among all tested devices, with a fallback rate of 30.8\% relative to workgroups launched.
+However, the average number of spins per workgroup is only 1.477---far below our fallback trigger threshold of 4 spins.
+This suggests a bimodal distribution of waiting times, where workgroups either proceed immediately with minimal delay or experience prolonged stalls that trigger fallbacks.
 
@@
     \includegraphics[width=\linewidth]{graphics/fallbacksInitiated_plot.pdf}
-    \caption{Fallbacks Initiated.\label{fig:fallbacks_initiated}}
+    \caption{Total number of fallback operations triggered as simulated blocking rates increase.
+This statistic verifies that the simulation effectively induces fallbacks as intended.\label{fig:fallbacks_initiated}}
   \end{subfigure}\hfill
@@
     \includegraphics[width=\linewidth]{graphics/totalSpins_plot.pdf}
-    \caption{Total Spins.\label{fig:total_spins}}
+    \caption{Average number of waiting spins per workgroup as simulated blocking rates increase.
+Higher spin counts correlate with increased fallback operations, reinforcing the behavior observed in Figure~\ref{fig:fallbacks_initiated}.\label{fig:total_spins}}
   \end{subfigure}
@@
     \includegraphics[width=\linewidth]{graphics/lookbackLength_plot.pdf}
-    \caption{Lookback Length.\label{fig:lookback_length}}
+    \caption{Average number of work tiles traversed before a workgroup completes \emph{Chained Propagation}.
+Longer lookback lengths indicate less efficient \emph{Chained Propagation}.\label{fig:lookback_length}}
   \end{subfigure}\hfill
@@
     \includegraphics[width=\linewidth]{graphics/time_plot.pdf}
-    \caption{Throughput vs.\ Baseline.\label{fig:execution_time}}
+    \caption{Relative execution time of \emph{DF} under increasing blocking rates compared to an unblocked baseline.
+Decreasing throughput reflects the growing overhead of fallback operations.\label{fig:execution_time}}
   \end{subfigure}
 
-  \caption{Simulated Blocking.
-\john{Please write a good caption here---this is an important caption to write!
-Probably writing subfigure captions is the right way to go (explain each of them individually).}\label{fig:vertical_images}}
+  \caption{\emph{Decoupled Fallback} Statistics Across Increasing Simulated Blocking Rates.
+Each subfigure highlights a key performance metric affected by the artificial omission of work tile updates, illustrating the resilience of \emph{DF} under extreme scheduling conditions.}
+  \label{fig:vertical_images}
 \end{figure*}
@@
 Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%; \emph{DF} is consistently faster than the previous standard for portable scans.
-While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that we are inducing fallbacks, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior.
-The finer details of this distinction are discussed next.
+While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that fallbacks are being induced, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior.
+This is because workgroups are still able to proceed after an omission.
+In true starvation, a blocking workgroup stalls before posting its reduction, preventing subsequent workgroups from making progress.
+Unlike our simulation, which allows workgroups to eventually continue, true starvation results in increased lookback lengths, as blocked workgroups cannot complete \emph{Chained Propagation} and update their flags to inclusive.
+Furthermore, starvation effectively leads to \emph{occupancy reduction}: execution resources remain available, but fewer subgroups can context-switch to mitigate latency
 \section{Discussion}
@@
 Devices that already lack sufficient compute resources to reach SOL performance are especially affected, as the additional compute overhead from fallback operations cannot be hidden by memory loads and stores.
-The worst-case scenario arises in devices that suffer from both low compute power and unfair scheduling.
-
-Another potential bottleneck is the speed of atomic operations.
-On some older GPUs---such as pre-Fermi NVIDIA architectures---atomics do not operate within the cache, resulting in significantly higher latency.
+The worst-case scenario arises in devices that suffer from both low compute power and unfair scheduling.Another potential bottleneck is the speed of atomic operations.
+On some older GPUs atomics do not operate within the cache, resulting in significantly higher latency.
 Slower atomics lead to delayed tile updates, which in turn increase lookback lengths.
-While our simulated deadlocks provide an artificial means of stress-testing execution, real deadlocks manifest as an effective \emph{occupancy reduction}: execution resources remain available, but fewer subgroups can context-switch to mitigate latency.
 

### 2025-03-01 paper.tex
Beat me to caption fix. Quick fix on input size.
typo

@@
 (1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward-progress guarantees (FPG) or support for 64-bit atomics.
-\textbf{Performance:} We aim for neear SOL execution to the greatest extent possible allowed by the underlying hardware and programming model.
+\textbf{Performance:} We aim for near-SOL execution to the greatest extent possible allowed by the underlying hardware and programming model.
 
@@
   \end{subfigure}
-  \caption{Histogram plots, 500 runs of Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements, on an FPG-lacking device, Apple M1 Max.
+  \caption{Histogram plots, 500 runs of Inclusive Prefix Sum, $\mathbf{2^{20}}$ 32-bit elements, on an FPG-lacking device, Apple M1 Max.
 Bar heights ($y$ axis) indicate the number of runs that completed in the execution time indicated on the $x$ axis.

### 2025-03-01 appendix.tex
Split Appendix

--- 2025-02-28 paper.tex
@@
 \usepackage{makecell}
-\newcommand{\thomas}[1]{{\footnotesize\color{orange}[Thomas: #1]}}
-\newcommand{\john}[1]{{\footnotesize\color{cyan}[John: #1]}}
-\newcommand{\raph}[1]{{\footnotesize\color{magenta}[Raph: #1]}}
 
@@
 
-\title{Decoupled Fallback: A Portable Single-Pass GPU Scan}
-
-\author{Thomas Smith}
-\affiliation{%
-  \institution{Google}
-  \country{USA}}
-
-\author{Raph Levien}
-\affiliation{%
-  \institution{Google}
-  \country{USA}
-}
-
-\author{John D. Owens}
-\affiliation{%
-  \institution{University of California, Davis}
-  \country{USA}}
-\additionalaffiliation{%
-  \institution{Google}
-  \country{USA}
-}
-
-\renewcommand{\shortauthors}{Smith et al.}
-
-\begin{abstract}
-  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{forward-progress guarantees} (FPG) while avoiding deadlock.
-Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation.
-On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach.
-We further demonstrate the resilience of \emph{Decoupled Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
-\end{abstract}
-
-\begin{CCSXML}
-  <ccs2012>
-  <concept>
-  <concept_id>10010147.10010169.10010170.10010174</concept_id>
-  <concept_desc>Computing methodologies~Massively parallel algorithms</concept_desc>
-  <concept_significance>500</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10010520.10010521.10010528</concept_id>
-  <concept_desc>Computer systems organization~Parallel architectures</concept_desc>
-  <concept_significance>500</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10003752.10003753.10003761.10003762</concept_id>
-  <concept_desc>Theory of computation~Parallel computing models</concept_desc>
-  <concept_significance>500</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10003752.10003753.10003761</concept_id>
-  <concept_desc>Theory of computation~Concurrency</concept_desc>
-  <concept_significance>500</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10010147.10010169.10010175</concept_id>
-  <concept_desc>Computing methodologies~Parallel programming languages</concept_desc>
-  <concept_significance>500</concept_significance>
-  </concept>
-  </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Computing methodologies~Massively parallel algorithms}
-\ccsdesc[500]{Computer systems organization~Parallel architectures}
-\ccsdesc[500]{Theory of computation~Parallel computing models}
-\ccsdesc[500]{Theory of computation~Concurrency}
-\ccsdesc[500]{Computing methodologies~Parallel programming languages}
-
-\keywords{scan, parallel prefix, prefix sum, forward-progress guarantee, GPU, WGSL, decoupled fallback}
-
-%\received{?}
-%\received[revised]{?}
-%\received[accepted]{?}
-
-\maketitle
-
-\section{Introduction}
-Scan is one of the most ubiquitous operations in GPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more.
-The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture.
-This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$.
-\emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{Speed of Light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device.
-However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite starvation and will never complete.
-In practice, it is easy to demonstrate this highly undesirable behavior.
-
-As the compute capabilities of major graphics APIs---Vulkan, Direct3D, and Metal---have grown, there has been increasing interest in adapting \emph{Decoupled Lookback} beyond CUDA for broader adoption in real-time graphics and GPU computing.
-However, portable implementation is constrained by both hardware and API limitations.
-In terms of hardware, some GPU schedulers—most notably Apple's M series and ARM—lack forward-progress guarantees, and subgroup sizes vary across architectures.
-In terms of API design, the absence of 64-bit atomic operations and explicit memory barriers further complicates implementation, as both are integral to \emph{Decoupled Lookback}.
-
-This work is guided by two main goals.
-(1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward-progress guarantees (FPG) or support for 64-bit atomics.
-\textbf{Performance:} We aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.
-
-However, certain aspects fall outside the scope of this work.
-\textbf{Performance guarantees} are not universally attainable, as not all architectures offer atomic operations that are sufficiently fast or scheduling models that are sufficiently fair to attain speed-of-light performance.
-\textbf{Subgroup divergence} is another challenge we acknowledge but do not address directly, as shading languages do not provide explicit divergence control like CUDA's subgroup masking mechanisms.
-Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained Scan}.
-
-We present \emph{Decoupled Fallback}, a fully portable prefix scan capable of achieving speed-of-light performance without requiring forward-progress guarantees (FPG).
-Our implementation is designed for compatibility with shading languages, making it suitable for modern graphics APIs and implemented in the WebGPU Shading Language (WGSL)\@.
-\john{I don't love ``compatibility with shading languages'' but then we name a single shading language.
-Instead, maybe something like ``Our implementation targets portability across modern GPU hardware and software platforms'', that sort of thing?} Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.
-
-\section{Background}
-\subsection{Virtualization and Scheduling in the GPU programming model}
-Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency.
-As comprehensive descriptions of the GPU programming model~\cite{10.1145/1365490.1365500} already exist, we focus on the aspects most relevant to this work: scheduling and synchronization.
-
-On GPU hardware, scheduling is divided into two levels: a \emph{workgroup} scheduler, which manages kernel launches and maps virtual processors, workgroups, to physical processors, \emph{multiprocessors}, and a \emph{subgroup} scheduler, responsible for selecting a subset of the currently \emph{occupant} subgroups on a multiprocessor for execution.
-The GPU programming model virtualizes processors to ensure GPU programs---\emph{kernels}---remain portable across hardware with differing physical resources.
-Effective virtualization requires \emph{oversubscription}, a programming paradigm where tasks are (over)partitioned in such a way that kernels request more virtual processors than are physically available.
-As a result, a single multiprocessor may host multiple workgroups simultaneously.
-However, as a consequence of virtualization, the developer relinquishes control over the order in which workgroups are launched, and once a workgroup begins execution on a multiprocessor, it must run to completion and cannot context-switch.
-
-\subsubsection{Inter-Workgroup Barriers}
-Parallel programs in general require synchronization between multiple threads (those that don't are considered embarrassingly parallel).
-A standard technique for synchronization is \emph{barriers.} GPUs provide a limited palette of barrier primitives.
-At the coarsest granularity, \emph{pipeline barriers} fully synchronize both control and memory between multiple kernel launches.
-However, pipeline barriers are extremely expensive and efficient algorithms strictly minimize their use.
-At a much finer granularity, \emph{workgroup barriers} provide reasonably efficient synchronization between the threads in a workgroup.
-However, synchronization between workgroups is more limited.
-Standard GPU programming environments provide no control barrier between workgroups.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} The availability of a memory-only barrier varies across GPU environments; it is present in Direct3D and Vulkan, but was missing in versions of Metal prior to 3.2.
-Because of its non-portability, it is also absent from the WebGPU specification.
-Even when available, the use of such memory barriers causes performance issues as they are usually implemented as a full cache flush.
-
-\subsubsection{Fairness and Progress}
-Because multiple workgroups may reside on a single multiprocessor, two key issues arise at the subgroup scheduling level: \emph{fairness}---how evenly execution resources are distributed among subgroups---and \emph{progress guarantees}---ensuring that subgroups eventually make progress towards termination.
-However, the term ``fairness'' is used differently across the literature, leading to potential confusion.
-Sorensen et al.~\cite{sorensen2016,sorensen2018,sorensen2021}, whose work provides the most comprehensive taxonomy of progress models to date, use ``fairness'' specifically to describe differing levels of progress guarantees.
-In contrast, NVIDIA publications~\cite{4523358,Merrill2016} define ``fairness'' in terms of the distribution of execution resources among subgroups.
-In this work, we adopt NVIDIA's terminology.
-
-\subsection{The Scan Primitive}
-\label{sec:the-scan-primitive}
-The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{5219801,10.5555/1098666}.
-A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \).
-The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \).
-To allow parallelization, \( \oplus \) must be associative, but it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{Clifford1961}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}.
-In a scan, the result at the \( n \)-th element is the reduction of the preceding subsequence of elements.
-If the subsequence includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}.
-The most common scan type is the prefix sum, where \( \oplus \) is addition.
-For example:
-\[
-  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1]
-\]
-\[
-  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
-\]
-\[
-  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
-\]
-Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science.
-Harris~\cite{1292373} offers a taxonomy that relates depth, fanout, and wire tracks, while Hinze~\cite{10.1007/978-3-540-27764-4_11} develops an algebraic framework for scans.
-Snir~\cite{10.1016/0196-67748690003-9} proved that depth $d$ and size $s$ are related by $s + d \ge 2n - 2$, and Fich~\cite{10.1145/800061.808738} proved that among minimum depth scans Ladner-Fischer~\cite{10.1145/322217.322232} networks have optimal size.
-Blelloch~\cite{Blelloch:1989:SAP} adapted scan to the PRAM model and popularized its use as an algorithmic primitive.
-Finally, Merrill and Garland~\cite{Merrill2016} provide a review of GPU scan implementations, complementing earlier work by Merrill and Grimshaw~\cite{Merrill2009}.
-
-\subsection{Evolution of Inter-Workgroup Scan Architectures}
-Contemporary GPU scans depart from early PRAM-like scans~\cite{Horn, Hensley, Sengupta2006, Gress} by leveraging the GPU memory hierarchy, which consists of progressively faster but increasingly private tiers.
-Efficient scans integrate fine-grained intra-workgroup (\emph{local}) strategies in shared memory and registers with coarse-grained inter-workgroup (\emph{global}) strategies in global memory.
-Given their low arithmetic intensity, local scans are inherently memory-bound, as demonstrated by Merrill and Grimshaw~\cite{Merrill2009}.
-Consequently, recent scan optimizations have shifted toward refining global strategies, particularly inter-workgroup coordination and synchronization.
-This section reviews three key approaches---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained Scan}---analyzing their handling of inter-workgroup dependencies and efficiency in minimizing global memory access while maximizing bandwidth utilization.
-
-\subsubsection{Scan-then-Propagate}
-Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2011} extends intra-workgroup scans to large inputs by resolving inter-workgroup dependencies in three phases:
-\begin{enumerate}
-  \item \textbf{Intra-Workgroup (Local) Scan:}  Each workgroup performs an inclusive scan on its assigned work tile, storing local results and reductions in global memory.
-  \item \textbf{Spine Scan:} The workgroup reductions, collectively forming the \emph{spine}, are gathered and scanned to compute the \emph{root}, resolving dependencies between work tiles.
-For large scans, this stage is an insignificant amount of work compared to the first and third.
-  \item \textbf{Propagation:} Each workgroup retrieves its portion of the root and adds it to the local scan output.
-\end{enumerate}
-The absence of inter-workgroup barriers necessitates separate kernel launches and pipeline barriers for dependency resolution, naturally enforcing this structure.
-To accommodate arbitrarily large inputs, Sengupta et al.\ employ recursion, repeatedly applying the three-phase process until the spine fits within a single work tile.
-Thus, an input of size $n$ and a work tile of size $t$ results in a recursive depth of $\lceil \log_t n \rceil$, and $2\cdot\lceil \log_t n \rceil - 1$ kernel launches.
-Each recursive step $k$ processes $n/t^k$ tiles, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps.
-Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total global data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.
-
-\subsubsection{Reduce-then-Scan}
-If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like scan can profit by trading additional computation for reduced memory traffic.
-This trade-off led to a refinement of the \emph{Scan-then-Propagate} strategy called \emph{Reduce-then-Scan}~\cite{10.1145/1375527.1375559, Merrill2009, 10.1109/TPDS.2012.336, 10.5555/2031978.2032029}: by storing only the work tile reduction in global memory during the first phase and performing a redundant scan in the third, it eliminates $n$ global data movement while preserving the three-phase structure:
-\begin{enumerate}
-  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result to global memory.
-  \item \textbf{Spine Scan:} The spine is scanned to produce the root.
-  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating its portion of the root as it writes to global memory.
-\end{enumerate}
-
-Merrill and Grimshaw~\cite{Merrill2009} introduced \emph{workgroup raking} to reduce recursion overhead for arbitrarily large inputs.
-Instead of recursively reducing the spine, the input is partitioned among a fixed number of workgroups, $c$, where $c \geq o$, the total workgroup occupancy.
-This constrains the spine from a variable, potentially recursive structure to a fixed size $c$, ensuring a recursion depth of at most two.
-As a result, global memory movement for the spine scan is reduced from $O(n/t)$ to $O(2c)$, and kernel launches and pipeline barriers are limited to three, yielding a total global memory movement of $O(3n + 4c) = O(3n)$.
-
-Breitbart~\cite{10.5555/2031978.2032029} applied Gupta et al.'s \emph{Persistent-Threads}~\cite{gupta2012} to scan, eliminating kernel launch overheads.
-Recognizing that raking naturally aligns with this paradigm, Breitbart's method statically \emph{discovers} the workgroup occupancy and launches exactly $o$ kernels.
-Since multiprocessors are not oversubscribed, workgroups persist for the entire kernel duration, enabling synchronization at an atomics-based inter-workgroup memory and control barrier without kernel launches or pipeline barriers.
-While global memory movement remains $O(3n)$, this consolidates the scan into a single kernel, minimizing launch and pipeline overheads.
-
-\subsubsection{Chained Scan, StreamScan}
-First introduced by Yan et al.~\cite{10.1145/2442516.2442539} in \emph{StreamScan}, the key innovation of the \emph{Chained Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level.
-Instead of using kernel launches and pipeline barriers as inter-workgroup synchronization points, \emph{Chained Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data.
-To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than arbitrary assignment based on virtualized workgroup indices.
-This shifts dependency resolution from explicit synchronization via barriers to implicit synchronization via access order.
-As a result, each element is read and written exactly once, achieving the theoretical minimum of $O(2n)$.
-
-In any \emph{Chained Scan}, once a workgroup computes its local scan, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles.
-This process, which we refer to as \emph{Chained Propagation}, defines the set of inter-workgroup techniques that facilitate dependency resolution.
-In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume.
-This approach depends on a forward-progress guarantee of the GPU's execution model.
-Given such a guarantee, the predecessor will eventually produce a result and broadcast it.
-On devices lacking such a guarantee, the workgroup may wait indefinitely, resulting in deadlock.
-
-While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently saturate global memory bandwidth, preventing it from achieving speed-of-light performance.
-This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the result of its immediate predecessor.
-While this results in efficient $O(n/t)$ communication of reductions, and $O(2n+ n/t)= O(2n)$ total data movement, the sequential dependency creates a bottleneck that limits performance to the rate at which reductions propagate through device memory.
-Although this can be mitigated by increasing the size of the work tile, limited on-chip resources preclude scaling the tile size indefinitely.
-Moreover, this approach is unsuitable from a portability perspective, as excessively large work tiles risk reducing occupancy on less capable hardware.
-
-\subsubsection{Chained Scan, Decoupled Lookback}
-Introduced by Merrill and Garland, the \emph{Decoupled Lookback} technique refines \emph{Chained Propagation} by separating reduction communication from dependency resolution, mitigating bottlenecks caused by propagation latency.
-In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction.
-Work and memory accesses are constant-bounded to the workgroup occupancy $o$ by posting the inclusive reduction after the lookback.
-However, granting each workgroup control over its dependency resolution does not eliminate reliance on \emph{progress guarantees}, as workgroups must still spin on tiles that have yet to post their reductions, risking deadlock.
-
-Although \emph{Decoupled Lookback} increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing confines spine memory accesses to a roughly $o$-width sliding window, making them highly likely to be cache-resident.
-As a result, the additional global memory traffic along the spine introduces negligible overhead.
-With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance without the work tile size tuning required by \emph{StreamScan}.
-
-\subsection{Ideal Scan Architecture}
-\emph{Chained Scan} and \emph{Reduce-then-Scan} represent two contrasting paradigms in scan design.
-By eliding the barrier-idiom, \emph{Chained Scan} aligns with the oversubscription paradigm, delegating occupancy management to the workgroup scheduler.
-In contrast, \emph{Reduce-then-Scan} employs workgroup raking, requiring fixed launches and explicit occupancy control, forcing developers to manually tune launch sizes or implement occupancy discovery techniques.
-
-Hardware-level considerations underscore the limitations of the fixed-launch approach.
-On GPUs where the global memory data cache is indexed by physical addresses---and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)---fixed launches can lead to performance degradation due to cache thrashing.
-This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups.
-Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns---or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations---may experience significant performance degradation.
-
-Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture.
-Furthermore, its single-kernel design and serial ordering make it suitable for in-situ operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging.
-However, it has a critical limitation: its reliance on a forward-progress guarantee.
-While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple's GPUs do not.
-Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
-
-\subsection{Progress Guarantees and Their Impact on \emph{Chained Scan}}
-A \emph{forward-progress guarantee} ensures that all occupant subgroups---of possibly different workgroups---eventually make progress toward termination, preventing indefinite starvation.
-Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees.
-Instead, Sorensen et al.\ systematically proposed and tested three progress models:\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}
-\begin{itemize}
-  \item \textbf{Occupancy-Bound Execution (OBE)}: Workgroups that have made at least one execution step.
-  \item \textbf{Heterogeneous System Architecture~\cite[Section~2.11]{HSASysArch12} (HSA)}: The workgroup with the lowest ID that has not yet terminated.
-  \item \textbf{Linear OBE (LOBE)}: Workgroups that have made at least one step, and any workgroup with a lower ID than a workgroup that has made at least one step.
-\end{itemize}
-Their 2021 findings~\cite{sorensen2021} demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior hypothesis~\cite{sorensen2018}.
-
-Sorensen et al.\ classify \emph{Chained Scan} as a \emph{one-way producer-consumer} (PC) HSA algorithm: a workgroup only consumes from lower-ID counterparts, ensuring that execution proceeds in a strictly ordered fashion.
-HSA guarantees termination, as the lowest ID workgroup is always given forward progress.
-If subgroups from the lowest ID workgroup are not guaranteed execution, then dependent workgroups may spin indefinitely, waiting on an update that never arrives.
-
-However, the HSA classification relies on the assumption that \emph{Chained Scan} follows scheduler-assigned virtualized IDs rather than dynamically acquired atomic-bumped IDs, as used by both \emph{StreamScan} and \emph{Decoupled Lookback}.
-Since virtual IDs do not necessarily match atomic IDs, HSA does not guarantee termination---the lowest virtual ID may not be scheduled first, potentially leading to starvation.
-
-LOBE adds an additional condition on top of OBE\@: lower-ID workgroups must be scheduled before higher-ID ones.
-However, this condition is unnecessary for \emph{Chained Scan} for two reasons: (1) We already guarantee serial ordering of work items through atomic bumping, making OBE sufficient.
-(2) \emph{Chained Scan} only requires forward progress after a workgroup has atomically acquired its tile index.
-By definition, a workgroup that has acquired its index must be occupant on a physical processor (i.e., it has made at least one execution step); thus, OBE guarantees forward progress.
-
-Thus, by dynamically acquiring IDs through atomics, \emph{Chained Scan} shifts its progress model requirement from HSA to OBE.
-While LOBE encompasses both HSA and OBE, its stricter ordering guarantees are unnecessary for correctness.
-Ultimately, regardless of a device's specific progress model, any architecture lacking forward-progress guarantees will fail to execute \emph{Chained Scan}.
-In short:
-
-\begin{table}[h!]
-  \small
-  \centering
-  \begin{tabular}{lcccc}
-    \toprule
-    Algorithm                             & LOBE & OBE & HSA & None \\
-    \midrule
-    \emph{Chained Scan} with Workgroup ID & Yes  & No  & Yes & No   \\
-    \emph{Chained Scan} with Atomic ID    & Yes  & Yes & No  & No   \\
-    Ours                                  & Yes  & Yes & Yes & Yes  \\
-    \bottomrule
-  \end{tabular}
-  \caption{Comparison of \emph{Chained Scan} Variants and Our Implementation Across Progress Models.\label{tab:progress_models}}
-\end{table}
-
-\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
-The absence of FPG in key hardware segments poses a critical challenge for developers.
-Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently.
-On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds.
-Other operating systems---such as macOS and Linux---employ their own timeout mechanisms with varying behaviors.
-For consistency, we adopt Windows’ 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program.
-\begin{figure}
-  \centering
-  \begin{subfigure}[b]{0.45\linewidth}
-    \includegraphics[width=\linewidth]{graphics/decoupled_lookback.pdf}
-    \caption{\emph{Decoupled Lookback.}}
-    \label{fig:decoupled}
-  \end{subfigure}
-  \hfill
-  \begin{subfigure}[b]{0.45\linewidth}
-    \includegraphics[width=\linewidth]{graphics/ours.pdf}
-    \caption{\emph{Decoupled Fallback (ours).}}
-    \label{fig:ours}
-  \end{subfigure}
-  \caption{Histogram plots, 500 runs of Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements, on an FPG-lacking device, Apple M1 Max.
-Bar heights ($y$ axis) indicate the number of runs that completed in the execution time indicated on the $x$ axis.
-The TDR threshold is Windows's threshold for resetting the GPU\@; as noted in Section~\ref{sec:tdr}, MacOS handles this differently.}
-  \label{fig:dist}
-\end{figure}
-Exacerbating this issue, no major graphics API provides a mechanism to query FPG support.
-This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware.
-In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.
-
-\subsection{Earlier Attempts at Portability}
-Our first attempt \raph{we'll add the cite post-review} at implementing a \emph{Decoupled Lookback} scan without relying on FPG was \emph{Scalar Fallback}.
-Each time a workgroup spins waiting on the reduction from a preceding tile, it advances a \emph{fallback} operation by processing one scalar element from that tile.
-This eliminates the possibility of deadlock because the number of such operations is bounded; the wait will eventually terminate, possibly at the expense of doing redundant work.
-
-\emph{Scalar Fallback} suffers from two primary inefficiencies.
-First, the fallback reduction is performed serially by a single thread, progressing only when the lookback thread performs a waiting spin.
-This results in a $t$-sized blocking tile requiring $t$ spins to complete.
-Second, once the fallback completes, the lookback thread does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback.
-While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce-then-Scan}.
-
-\section{Decoupled Fallback}
-\emph{Decoupled Fallback} addresses both inefficiencies.
-Our key advance is to utilize the parallelism available in the workgroup to perform the fallback computation, rather than performing it sequentially.
-To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained Propagation} logic is typically performed by the first thread.
-
-\begin{algorithm}
-  \small
-  \SetAlgoLined%
-  \KwGlobalIn{%
-    $\textit{in}[]$,
-    $\textit{out}[]$,
-    $\textit{spine}[]$,
-    $\textit{tileBump}$,
-  }
-  \KwSharedIn{%
-    $\textit{wg\_partials}[]$,
-    $\textit{wg\_fallback}[]$,
-    $\textit{wg\_prev\_reduction}$,
-    $\textit{wg\_broadcast}$,
-    $\textit{wg\_control}$,
-  }
-  \KwGlobalOut{$\textit{out}[]$ inclusive scan of $\textit{in}[]$}
-  \KwConstants{Vectors per thread $\textit{p}$}
-
-  \If{$\textit{threadid} = 0$}{
-    $\textit{wg\_broadcast} \gets \textnormal{atomicAdd}(\&\textit{tileBump}, 1)$\;
-    $\textit{wg\_control} \gets \textnormal{LOCKED}$\;
-  }
-  \textbf{workgroupBarrier()}\;
-
-  $\textit{tile\_id} \gets \textit{wg\_broadcast}$\;
-
-  $\textit{local\_vectors} \gets \textnormal{array}\langle \textnormal{vec4}\langle u32\rangle, \textit{p} \rangle$\;
-
-  \textbf{Load}($\textit{in}$, $\textit{tile\_id}$, \textit{local\_vectors})\;
-  \textbf{SubgroupRake}($\textit{local\_vectors}$, $\textit{wg\_partials}$)\;
-  \textbf{workgroupBarrier()}\;
-
-  \textbf{WorkgroupWideScan}($\textit{wg\_partials}$)\;
-  \textbf{workgroupBarrier()}\;
-
-  \tcp{Chained Propagation, Decoupled Fallback}
-  \textbf{PostLocalReduction}($\textit{spine}$, $\textit{tile\_id}$, $\textit{wg\_partials}$)\;
-  \While{$\textit{wg\_control} = \textnormal{LOCKED}$}{
-    \textbf{Lookback}($\textit{spine}$, $\textit{tile\_id}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
-    \textbf{workgroupBarrier()}\;
-
-    \If{$\textit{wg\_control} = \textnormal{LOCKED}$}{
-      \textbf{FallbackReduce}($\textit{in}$, $\textit{wg\_fallback}$)\;
-      \textbf{FallbackInsertionAttempt}($\textit{spine}$, $\textit{wg\_fallback}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
-      \textbf{workgroupBarrier()}\;
-    }
-  }
-
-  \textbf{Write}($\textit{out}$, $\textit{tile\_id}$, $\textit{local\_vectors}$, $\textit{wg\_partials}$, $\textit{wg\_prev\_reduction}$)\;
-
-  \caption{High-Level Scan Kernel}
-\end{algorithm}
-
-To optimize the fallback operation, we employ three strategies.
-(1) First, a shared memory \emph{control flag} enables threads performing the lookback to signal the entire workgroup when a fallback is necessary.
-(2) Next, we limit the number of spins a thread can perform while waiting on a tile; exceeding this threshold triggers a fallback.
-While this approach risks \emph{false positive} fallbacks---since a late tile is indistinguishable from a genuinely blocking one---the maximum spin count can be tuned to minimize redundant fallbacks.
-Increasing the spin count lowers false positives but delays progress in true-blocking scenarios.
-False positives not only increase memory traffic but also consume compute resources that could be scheduled on the same multiprocessor.
-(3) After completing a fallback, the workgroup attempts to post the reduction result of the blocking tile, preventing redundant fallbacks by subsequent workgroups.
-
-The viability of \emph{Decoupled Fallback} relies on the same cache locality properties as \emph{Decoupled Lookback}.
-Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$\!, independent of FPG, due to false-positive fallback operations.
-As a result, memory bandwidth usage increases (and we expect throughput to decrease) by $O(fn)$ to $O((2 + f)n)$.
-However, just as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache.
-As we later empirically demonstrate in Section~\ref{sec:Sim-Blocking}, this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile.
-Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.
-
-\section{Implementation}
-At a high level, our algorithm follows a five-phase construction:
-\begin{enumerate}
-  \item[(0)] \textbf {Initialization}: The first thread in the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
-  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering.
-For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
-  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking Kogge-Stone~\cite{5009159} subgroup scans,
-        allowing the subgroup-block scan to remain barrier-free and in registers.
-The subgroup-block reduction is then posted to shared memory.
-For a workgroup of size $w$, subgroup-raking reduces the size of the local spine to $w/s$.
-  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
-  \item \textbf{Chained Propagation, Decoupled Fallback}: The local reduction is posted to global memory.
-Then, the first subgroup in the workgroup performs a lookback operation to acquire the reduction result of all previous workgroups, engaging in fallback operations as necessary.
-Once completed, this subgroup posts the results to shared memory for the workgroup to consume.
-  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output.
-\end{enumerate}
-Phase 0 ensures that the work tiles are consumed in serial order.
-Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy.
-Phase 4, \emph{Decoupled Fallback}, is responsible for managing all inter-workgroup dependencies.
-As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations.
-
-\begin{algorithm}
-  \small
-  \SetAlgoLined
-  \KwSharedIn{ Array of subgroup partial reductions $\textit{x}$ }
-  \KwSharedOut{ Inclusive-scanned $\textit{x}$ }
-  \KwConstants{ Workgroup Size $\textit{W}$, Subgroup Size $\textit{S}$ }
-
-  $\textit{spine\_length} \gets \textit{W} / \textit{S}$\;
-  $\textit{alignment} \gets 1 << \text{divRoundUp}(\log_2(\textit{spine\_length}), \log_2(\textit{S})) \cdot \log_2(\textit{S})$\;
-
-  $\textit{stride} \gets 1$\;
-  $\textit{top\_offset} \gets 0$\;
-
-  \ForEach{$\textit{thread\_id}$ \textbf{in} $\textit{W}$ \textbf{in parallel}}{
-    \For{$\textit{j} \gets \textit{S}$ \KwTo $\textit{alignment}$ \textbf{with} $\textit{j} \gets \textit{j} \cdot \textit{S}$}{
-      \tcp{Brent-Kung Upsweep}
-
-      $\textit{step} \gets \textit{spine\_length} / \textit{stride}$\;
-
-      \If{$\textit{thread\_id} < \textit{step}$}{
-        $\textit{temp} \gets \text{subgroupInclusiveScan}(\textit{x}[\textit{thread\_id} + \textit{top\_offset}])$\;
-        $\textit{x}[\textit{thread\_id} + \textit{top\_offset}] \gets \textit{temp}$\;
-        \If{$\text{\normalfont highestRankLane}(\textit{thread\_id})$}{
-          $\textit{x}[(\textit{thread\_id} / \textit{S}) + \textit{step} + \textit{top\_offset}] \gets \textit{temp}$\;
-        }
-      }
-
-      \textbf{workgroupBarrier()}\;
-
-      \tcp{Fanout}
-
-      \If{$\textit{j} \neq \textit{S}$}{
-        $\textit{reduced\_stride} \gets \textit{j} / \textit{stride}$\;
-        $\textit{fanout\_index} \gets \textit{thread\_id} + \textit{reduced\_stride}$\;
-
-        $\textit{cond1} \gets \textit{fanout\_index} < \textit{spine\_length}$\;
-        $\textit{cond2} \gets (\textit{fanout\_index} \,\&\, (\textit{j} - 1)) \geq \textit{reduced\_stride}$\;
-
-        \If{$\textit{cond1} \textbf{ \&\& } \textit{cond2}$}{
-          $\textit{x}[\textit{fanout\_index}] \gets \textit{x}[\textit{fanout\_index}] + \textit{x}[((\textit{fanout\_index} / \textit{stride}) + \textit{top\_offset}) - 1]$\;
-        }
-      }
-      $\textit{stride} \gets \textit{stride} \cdot \textit{S}$\;
-      $\textit{top\_offset} \gets \textit{top\_offset} + \textit{step}$\;
-    }
-    \textbf{barrier()}\;
-  }
-  \Return{$\textit{x}$}\;
-  \caption{Workgroup-Wide Scan.\label{alg:example}}
-\end{algorithm}
-
-\subsection{Workgroup-Wide Scan}
-The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches.
-To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{10.1145/322217.322232} scan network embedded with Kogge-Stone~\cite{5009159} subgroup scans to perform the scan across the local spine.
-Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts.
-To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$.
-Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction.
-Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index.
-However, such access is interpreted by the hardware as a subgroup \emph{broadcast} operation, resulting in no conflicts.
-Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking.
-Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.
-
-\subsection{Decoupled Fallback}
-To reiterate, the goal of \emph{Chained Propagation} is twofold: first, to determine the reduction of all preceding work tiles, and second, to relay those values to succeeding tiles to accelerate their own lookback operations.
-Following the \emph{Decoupled Lookback} technique, we model the inter-workgroup spine as a finite-state machine, where each work tile is represented by a reduction value and a state, stored in global memory.
-Each tile may exist in one of three states:
-\begin{itemize}
-  \item \textbf{Not Ready}: The tile has not yet been processed or written to global memory.
-  \item \textbf{Ready}: The tile has completed its local reduction and posted it to global memory.
-  \item \textbf{Inclusive}: The tile has completed its lookback and updated its posting to reflect the inclusive reduction.
-\end{itemize}
-Workgroups operate independently; each is responsible for updating its own tile state but may also opportunistically update preceding tiles in the event of a fallback.
-Tile states transition in a strictly forward direction---\emph{Not Ready} $\rightarrow$ \emph{Ready} $\rightarrow$ \emph{Inclusive}---and never revert.
-Before kernel launch, all tiles are initialized to \emph{Not Ready} with reduction values set to $\varepsilon$.
-State transitions imply an update of the reduction value: a transition to \emph{Ready} results in the posting of the local workgroup reduction, and to \emph{Inclusive}, the reduction of all previous workgroup reductions.
-To coordinate this process across the workgroup, each workgroup initializes a shared memory \emph{control flag} during partition index acquisition.
-This flag governs both the initiation of fallback operations and the entry and exit of the entire \emph{Decoupled Fallback} phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.
-
-We now detail the \emph{Decoupled Fallback} procedure, presenting the algorithm from the perspective of a single workgroup.
-The details of tile state encoding and atomic updates are discussed in the following section and may vary depending on the data type, so for simplicity, we describe the operation as being performed by the \emph{first thread} of the workgroup:
-\begin{figure}
-  \centering
-  \includegraphics[height=\textheight]{graphics/FlowChart.pdf}
-  \caption{Flow Chart Representation of Decoupled Fallback.\label{fig:decoupled_fallback}}
-\end{figure}
-\begin{enumerate}
-  \item[(0)] \textbf{Posting Local Reduction:}
-        \begin{itemize}
-          \item Each workgroup posts its local reduction---the result of the workgroup-wide scan over partials---to global memory and updates its tile state to \emph{Ready}.
-          \item If it is the first work tile, its local reduction \emph{is} its inclusive reduction.
-In this case, the workgroup updates its tile state directly to \emph{Inclusive} and bypasses the \emph{Chained Propagation} phase entirely.
-        \end{itemize}
-
-  \item \textbf{Initialization:}
-        \begin{itemize}
-          \item The control flag, initialized to \emph{Locked}, signals that the workgroup should enter the \emph{Chained Propagation} phase.
-          \item The first thread initializes a \emph{previous reduction} variable in registers with $\varepsilon$ to hold the accumulated reductions of preceding tiles.
-          \item The first thread initializes a \emph{current tile} pointer to its immediate predecessor, with which it begins to traverse backwards through the spine.
-        \end{itemize}
-
-  \item \textbf{Lookback:}
-        \begin{itemize}
-          \item The first thread queries the state of \emph{current tile}.
-                \begin{itemize}
-                  \item If the tile is in the \emph{Ready} state, the workgroup ingests the value and sets \emph{current tile} to the preceding tile, repeating this step until reaching a tile that is \emph{Not Ready}.
-                  \item If the tile is in the \emph{Inclusive} state, the workgroup ingests the value and proceeds directly to step~4.
-                  \item If the tile is in the \emph{Not Ready} state, the first thread spins for a limited number of iterations, defined by the \emph{Maximum Spin Count}.
-                        \begin{itemize}
-                          \item If this count is exceeded, it exits the lookback loop and proceeds to step~3, leaving the control flag \emph{Locked} to signal to the workgroup that a fallback is required.
-Additionally, it copies \emph{current tile} into shared memory to broadcast the fallback tile identity to the workgroup.
-                          \item Otherwise, it continues spinning until the tile transitions to a \emph{Ready} or \emph{Inclusive} state.
-Without FPG, \emph{Decoupled Lookback} risks spinning indefinitely and deadlocking at this stage.
-                        \end{itemize}
-                \end{itemize}
-        \end{itemize}
-
-  \item \textbf{Fallback:}
-        \begin{itemize}
-          \item If the blocking tile is still \emph{Not Ready} after waiting, the workgroup initiates a fallback operation.
-          \item Using the tile index provided by the first thread, the full workgroup computes the reduction of the blocking tile by directly processing the corresponding elements.
-This process involves a \emph{workgroup-wide reduction},\footnote{We obtain the workgroup-wide reduce by omitting the fanout operation of the workgroup-wide scan and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.} utilizing additional shared memory to aggregate partial subgroup results.
-          \item Once the fallback reduction is complete, the workgroup \emph{attempts} to post the reduction to global memory and update the tile's state to \emph{Ready}.
-          \item As this update attempt is performed atomically, the workgroup simultaneously reads back the blocking tile's most recent state:
-                \begin{itemize}
-                  \item If that most recent state is \emph{Inclusive}, the fallback value computed by the workgroup is discarded, and the first thread ingests the returned value before proceeding to step 4.
-                  \item Otherwise, the first thread ingests the fallback value.
-                        \begin{itemize}
-                          \item If the blocking tile is the first in the spine, the workgroup has reached the start of the spine and has thus completed the \emph{Chained Propagation}.
-The first thread proceeds to step 4.
-                          \item Otherwise, the first thread updates \emph{current tile} to the preceding tile, returning to the lookback process in step~2.
-                        \end{itemize}
-                \end{itemize}
-        \end{itemize}
-
-  \item \textbf{Finalization:}
-        \begin{itemize}
-          \item To prioritize inter-workgroup reduction propagation, the first thread immediately adds the local reduction to \emph{previous reduction}, posts the inclusive reduction to global memory, and transitions the tile state to \emph{Inclusive}.
-          \item It then writes \emph{previous reduction} to shared memory, allowing all threads to combine it with their \emph{workgroup-wide} and private scan computations.
-          \item Finally, the first thread sets the control flag to \emph{Unlocked}, signaling the rest of the workgroup that \emph{Chained Propagation} has completed, and allowing the final write phase to begin.
-        \end{itemize}
-\end{enumerate}
-
-\subsection{Tile State Implementation}
-In all variants of \emph{Chained Scan}, there is communication and synchronization between workgroups.
-The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it.
-This primitive is well-studied in the context of memory models~\cite{Alglave0, MCMutants}.
-The key requirements here are both control synchronization (meaning the reader does not proceed until the writer sends the message) and data synchronization (meaning that the reader reads the correct value).
-The latter is especially challenging in relaxed memory models as are employed on GPUs.
-
-Combined control and data synchronization can be accomplished by packing both the tile state and the reduced value into a single word, using relaxed atomic store to send the message, and relaxed atomic load to receive the message, spinning as necessary until the tile state is updated.
-The semantics of atomics guarantee, even with relaxed memory ordering, that the data is fresh.
-However, if the tile state and data are stored in separate memory locations, relaxed atomics are insufficient because the compiler is free to reorder stores, so one update does not imply the other.
-On CPUs, it is common to upgrade the memory semantics from relaxed to acquire/release, which then provides the necessary guarantee~\cite[Section 10.2]{intel_sdm_2024}.
-Unfortunately, on GPUs, acquire/release semantics are neither portable (they are only available on Vulkan, and optional until version 1.3), nor are they necessarily efficient even when they are available; often the implementation involves expensive cache flushes.
-A device-scoped memory barrier has similar problems.
-While it would provide the necessary guarantee, it is also not available in WebGPU (or in versions of Metal prior to 3.2), and is expensive.
-
-Without some other mechanism, the requirement to pack tile state and data into a single word limits the data to 30 bits (assuming 2 bits for the tile state).
-That, in turn, makes it impossible to represent ordinary 32-bit floating point values without loss of precision.
-While some GPUs support 64 bit atomics (and it is likely that WebGPU will someday include them as an optional extension), that is also sadly not portable, and merely raises the payload limit to 62 bits without eliminating it altogether.
-
-A key innovation of this paper is to \emph{split} the reduced values into multiple atomics, each carrying its own copy of the tile state.
-The message is received when relaxed atomic loads of all the parts agree on their tile state.
-This way, the payload is reconstructed intact, even in the absence of any guarantees from the memory model stronger than relaxed atomics.
-
-\begin{figure}[h!]
-  \centering
-  \includegraphics[width=\linewidth]{graphics/split.pdf}
-  \caption{Splitting one 32-bit value into two 32-bit values with bit-packed tile state.}
-\end{figure}
-
-In particular, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state.
-Although this could also be accomplished with shared memory (and workgroup barriers), as an optimization, we use two or four threads from the first subgroup, leveraging subgroup \emph{ballot} to verify state consistency, and subgroup \emph{shuffle} to join and split reductions.
-With one thread per 16 bits and a minimum subgroup size of four in WebGPU, this method supports both 32-bit and 64-bit values, enabling 64-bit scans without 64-bit atomics or memory barriers.\footnote{Although native 64-bit values are not currently supported by the WebGPU specification, their future inclusion is anticipated.}
-
-Fallbacks require a tile state representation and an atomic operation that can safely transition a tile to \emph{Ready} without overwriting a more advanced state.
-When an update attempt is made, a blocking tile may be in any of the three states due to either of two scenarios: (1) a false-positive fallback on a tile that transitions to \emph{Ready} or \emph{Inclusive} before fallback completion, or (2) another workgroup completing its fallback first, leaving the tile in \emph{Ready}.
-If the tile state is \emph{Inclusive} then the write must be skipped, especially because races in writes for the parts may leave the tile permanently in an unreadable state.
-Thus, a simple atomic store is not suitable.
-An atomic compare-and-swap would have the correct semantics, but WebGPU lacks this primitive (it does have \texttt{atomicCompareExchangeWeak}, which could possibly be made to work but makes reasoning about correctness and performance more difficult).
-Instead, we encode the tile state in the most significant bits of the atomic value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}.
-This allows an \texttt{atomicMaximum}, which successfully updates to a more advanced state but prevents a state regression, to correctly perform updates.
-
-The atomic splitting technique generalizes to monoids of arbitrary size.
-
-\section{Evaluation}
-\subsection{Experimental Setup}
-We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$.
-We find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks.
-As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4.
-Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@.
-For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum.
-Our test hardware, shown in Figure~\ref{tab:hardware}, spans a wide range of FPG and performance capabilities.
-We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires native Android, necessitating the use of WGPU\@.\footnote{To keep the performance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
-\begin{table}
-  \centering
-  \begin{tabular}{l l c r r}
-    \toprule
-    Vendor & Device Name            & LOBE         & Occupancy & Subgroup Size \\
-    \midrule
-    Intel  & HD Graphics 620        & \textbf{Yes} & 9         & 8--32         \\
-    ARM    & Mali-G78 MP20          & \textbf{No}  & 40        & 16            \\
-    Apple  & M3     (10 Cores)      & \textbf{No}  & 80        & 32            \\
-    Apple  & M1 Max (32 Cores)      & \textbf{No}  & 160       & 32            \\
-    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192       & 32            \\
-    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420       & 32--64        \\
-    \bottomrule
-  \end{tabular}
-  \caption{Hardware Specifications of Testing Devices.\label{tab:hardware}}
-\end{table}
-
-\subsection{Performance Comparison}
-\subsubsection{Cross-API Performance Versus Contemporary Implementations}
-We first compare \emph{Decoupled Fallback} (\emph{DF}) to two key benchmarks: \emph{Memcpy}\textdagger{~\ref{sec:memcpy}} and the \emph{Decoupled Lookback} (\emph{DL}) implementation found in NVIDIA's CUB library, using the 2080 Super only.
-To isolate the impact of differing APIs, we implement and evaluate \emph{DF} in both CUDA and Dawn.\footnote{Due to WebGPU's 128~MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.} \emph{Memcpy} serves as a performance reference, as it shares the same $2n$ global memory movement as \emph{DF} but lacks compute overhead and is purely memory bandwidth bound.
-
-Cross-API comparisons remain challenging, as our CUDA implementation consistently outperforms its Dawn counterpart despite near-identical high-level implementations.
-The performance cost of \emph{DF} is most pronounced in CUDA when the entire input fits within the 2080 Super's 4~MB L2 cache, where reduced global memory latency exposes \emph{DF's} increased synchronization and compute overhead compared to CUB\@.
-However, at larger input sizes, where memory accesses exceed cache capacity, \emph{DF} reaches parity with CUB and achieves \emph{Memcpy} speeds.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.6\linewidth]{graphics/cuda_plot.pdf}
-  \caption{Performance versus CUB (NVIDIA 2080 Super).}
-\end{figure}
-
-\subsubsection{Same-API Performance Versus Baseline Measures}
-We next compare \emph{DF} against \emph{Reduce-then-Scan}\textdagger{~\ref{sec:rts}} (\emph{RTS}), which serves as a baseline to highlight the performance improvements of our approach over the prior FPG-constrained technique.
-This evaluation is conducted using Dawn across the full range of tested devices.
-
-\begin{table*}
-  \centering
-  \small
-  \setlength{\tabcolsep}{1pt}
-  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r r}
-    \toprule
-    Metric & \!\!\!\!\!\!Intel HD620 & ARM Mali-G78 MP20 & Apple M3 & Apple M1 Max & Nvidia 2080 Super & AMD 7900 XT \\
-    \midrule
-    Avg \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)  & 1.50   & 2.48  & 10.25  & 38.62  & 54.39  & 72.28  \\
-    Avg \emph{DF} (32-bit ele/sec $\cdot 10^9$)   & 1.49   & 2.79  & 10.22  & 38.84  & 53.29  & 94.60  \\
-    Avg \emph{RTS} (32-bit ele/sec $\cdot 10^9$)      & 1.06   & 2.02  & 6.81   & 25.13  & 35.90  & 61.49  \\
-    \emph{DF}/\emph{Memcpy}       & 99.1\%    & 112.8\%  & 99.8\%    & 100.6\%   & 98.0\%    & 130.9\%   \\
-    \emph{DF}/\emph{RTS}          & 140.2\%   & 138.5\%  & 150.2\%   & 154.5\%   & 148.4\%   & 153.8\%   \\
-    Avg Spins per workgroup    & 0.305    & 1.477    & 0.606    & 0.604    & 0.419    & 0.148    \\
-    Avg Lookback Length per workgroup& 1.726    & 3.537    & 3.415    & 10.782   & 13.046   & 20.722   \\
-    Avg Total Fallbacks Initiated      & 254.662   & 2504.704  & 365.546   & 990.962   & 86.400   & 9.308   \\
-    Avg Total Successful Insertions    & 6.170     & 122.436   & 4.612     & 0.012    & 0.284    & 0.020    \\
-    \bottomrule
-  \end{tabular*}
-  \caption{\emph{Decoupled Fallback} performance, Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements.\label{tab:results}}
-\end{table*}
-
-As shown in Table~\ref{tab:results}, \emph{DF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices, closely approaching the theoretically expected $O(3n)/O(2n)$ 50\% speedup, except on the ARM Mali.
-\emph{DF} closely matches \emph{Memcpy} across most devices, deviating only on the AMD 7900 XT and ARM Mali, where \emph{DF} outperforms \emph{Memcpy} by 30.9\% and 12.8\%, respectively.
-This suggests that additional compute latency may mitigate paging effects by separating loads from stores.
-On the ARM Mali, \emph{DF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%.
-However, its nearly identical performance to \emph{Memcpy} indicates that the implementation is memory-bandwidth-bound, limiting further gains.
-
-\subsection{Decoupled Fallback Statistics}
-To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution: \john{Below, I advocate that we say, for each of the four bullets, what we expect / what would be a good result / what the hardware actually does in practice.
-The reader doesn't deeply know what to expect / what is good.}
-\begin{itemize}
-  \item \textbf{Spins per Workgroup:} How many times does a workgroup make a waiting spin?
-  \item \textbf{Lookback Length per Workgroup:} How many work tiles does a workgroup traverse to complete \emph{Chained Propagation}?
-\john{Example for above point: something like ``We hope that even with unfair hardware schedulers, this traversal is short (only a few tiles), but we also want to ensure that DF is robust and delivers good performance for even long traversals.''}
-  \item \textbf{Total Fallbacks Initiated:} What is the total number of fallback operations initiated across all workgroups?
-  \item \textbf{Total Successful Insertions:} What is the total number of fallback operations that successfully update a blocking (or late) tile to \emph{Ready}?
-\end{itemize}
-Despite differing occupancy rates, the number of waiting spins remains consistent across devices.
-\john{Is this a small number?
-Large?
-Tell don't show.} Likewise, lookback length is stable and remains well below the theoretical occupancy bound.
-Expressed as a percentage of occupancy, it remains under 10\% for all devices, except for the Intel HD620, where its elevated value is likely an artifact of its low absolute occupancy.
-Most notably, the devices lacking FPG---Apple M3 and M1 Max---do not exhibit significantly higher successful insertion rates than the others, suggesting that true deadlocking is a rare occurrence.
-
-\subsection{Simulated Blocking}%
-\label{sec:Sim-Blocking}
-To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally omitting work tile reduction updates.
-Because these updates are skipped, the immediate successor workgroup reads \emph{Not Ready}, spinning and triggering a fallback.
-Starting with infrequent omissions---1 in 512 work tiles---we progressively increase the omission rate to near-continuous levels---1 in 2 work tiles---recording \emph{DF} statistics at each step.
-\begin{figure*}
-  \centering
-  \begin{subfigure}[t]{0.48\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{graphics/fallbacksInitiated_plot.pdf}
-    \caption{Fallbacks Initiated.\label{fig:fallbacks_initiated}}
-  \end{subfigure}\hfill
-  \begin{subfigure}[t]{0.48\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{graphics/totalSpins_plot.pdf}
-    \caption{Total Spins.\label{fig:total_spins}}
-  \end{subfigure}
-
-  \vspace{0.2em}
-
-  \begin{subfigure}[t]{0.48\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{graphics/lookbackLength_plot.pdf}
-    \caption{Lookback Length.\label{fig:lookback_length}}
-  \end{subfigure}\hfill
-  \begin{subfigure}[t]{0.48\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{graphics/time_plot.pdf}
-    \caption{Throughput vs.\ Baseline.\label{fig:execution_time}}
-  \end{subfigure}
-
-  \caption{Simulated Blocking.
-\john{Please write a good caption here---this is an important caption to write!
-Probably writing subfigure captions is the right way to go (explain each of them individually).}\label{fig:vertical_images}}
-\end{figure*}
-
-As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near 100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced.
-Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%; \emph{DF} is consistently faster than the previous standard for portable scans.
-While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that we are inducing fallbacks, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior.
-The finer details of this distinction are discussed next.
-\section{Discussion}
-
-\subsection{Trade-offs}
-\emph{Decoupled Fallback} introduces several tradeoffs that must be considered: (1) Without a memory barrier, the scan cannot be performed in place and requires a separate output buffer to prevent false-positive fallback operations on post-scanned values, as the tile state may not update before the scan values.
-(2) For monoids larger than 64 bits, multiple subgroups are required, increasing synchronization overhead due to the need for shared memory and barriers.
-(3) In certain in-situ compaction algorithms where fallback work-tiles cannot be easily reduced and must be held in registers, Decoupled Fallback may reduce occupancy.
-(4) Additionally, in compute-bound in-situ compaction algorithms, the cost of fallbacks may outweigh the savings from reduced global memory traffic.
-(5) Lastly, because the control flow of \emph{Decoupled Fallback} requires barriers, latency incurred during lookback cannot be hidden by workgroup computation.
-
-\subsection{Limits on Speed-of-Light Performance}
-Due to its increased latency, synchronization overhead, and longer critical path, \emph{Decoupled Fallback} exacerbates existing challenges faced by \emph{Decoupled Lookback}, particularly a lack of compute capacity and unfair scheduling.
-Devices that already lack sufficient compute resources to reach SOL performance are especially affected, as the additional compute overhead from fallback operations cannot be hidden by memory loads and stores.
-The worst-case scenario arises in devices that suffer from both low compute power and unfair scheduling.
-
-Another potential bottleneck is the speed of atomic operations.
-On some older GPUs---such as pre-Fermi NVIDIA architectures---atomics do not operate within the cache, resulting in significantly higher latency.
-Slower atomics lead to delayed tile updates, which in turn increase lookback lengths.
-While our simulated deadlocks provide an artificial means of stress-testing execution, real deadlocks manifest as an effective \emph{occupancy reduction}: execution resources remain available, but fewer subgroups can context-switch to mitigate latency.
-
-\subsection{Future Work}
-Sorting is one of the most notable beneficiaries of single-pass scan and, by extension, \emph{Decoupled Fallback}.
-The current state-of-the-art algorithm, Adinets and Merrill's \emph{Onesweep}~\cite{adinets2022onesweepfastersignificantdigit}, is an 8-bit least-significant-digit radix sort, where single-pass scan minimizes data movement during digit-binning passes.
-However, due to its compute-intensive nature and large $radixDigit$-sized monoid, it is unclear whether \emph{Decoupled Fallback} will map well to sorting or if new techniques are needed.
-
-Our reexamination of \emph{Chained Scan} with dynamic assignment raises fundamental questions about the GPU progress models proposed by Sorensen et al.~\cite{sorensen2021}: Can all HSA algorithms be converted to OBE via dynamic assignment?
-Are there cases, excluding atomic memory errors, where LOBE guarantees termination but dynamic assignment with OBE does not?
-Is OBE all you need?
-
-\subsection{Conclusion}
-In this paper, we presented \emph{Decoupled Fallback}, a single-pass \emph{Chained Scan} capable of reaching SOL performance under all progress models and suitable for implementation within the constraints of contemporary portable shading languages.
-Our approach maintains minimal data movement while ensuring robust execution across diverse GPU architectures.
-
-\begin{acks}
-\end{acks}
-
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{bib}
-
-\clearpage
 \appendix
-\section{Artifact}
-\subsection{Availability}
-We provide our artifact in the following GitHub repository: \thomas{Will anonymize with GithubAnonymous once we are ready}. The repository contains per API requirements and building instructions.
-
+\section{Artifact Notes}
 \subsection{Baseline Comparisons: \emph{Memcpy}}%
@@
 The cause of this discrepancy remains unclear, but we note other developers have observed similar speeds.
-  \item \textbf{Apple M1 Pro Max:} Observed throughput matches expectations.
+  \item \textbf{Apple M3:} Observed throughput matches expectations.
+  \item \textbf{Apple M1 Max:} Observed throughput matches expectations.
   \item \textbf{Nvidia 2080 Super:} Observed throughput matches expectations.
-  \item \textbf{AMD 7900 XT:} Our above result, where \emph{DF} was observed to be faster than \emph{Memcpy}, is likely due to compute latency counterintuitively improving memory access patterns, leading to reduced paging.
-Furthermore, we see that our \emph{DF} throughput is within spec and reasonable.
+  \item \textbf{AMD 7900 XT:} Our above result, where \emph{Decoupled Fallback} was observed to be faster than \emph{Memcpy}, is likely due to compute latency counterintuitively improving memory access patterns, leading to reduced paging.
+Furthermore, we see that our \emph{Decoupled Fallback} throughput is within spec and reasonable.
 \end{itemize}
@@
 \label{sec:rts}
-As we desire the most competitive baseline possible, our \emph{RTS} kernels employ the same intra-workgroup scan strategy and tuning parameters as \emph{DF}.
+As we desire the most competitive baseline possible, our \emph{Reduce-then-Scan} kernels employ the same intra-workgroup scan strategy and tuning parameters as \emph{Decoupled Fallback}.
 Initially, we adopted Merrill and Grimshaw's~\cite{Merrill2009} workgroup raking approach for inter-workgroup processing.
-However, empirical testing revealed that raking was consistently outperformed—by up to $\sim$15\% on lower-end devices—by a simpler approach that processes the spine serially with a single workgroup.
-Furthermore, our \emph{RTS} performance closely aligns with the theoretically expected $O(3n)$ global memory movement.
+However, empirical testing revealed that raking was consistently outperformed---by up to $\sim$15\% on lower-end devices---by a simpler approach that processes the spine serially with a single workgroup.
+We confirm the efficiency of our \emph{Reduce-then-Scan} implementation by observing that its performance closely aligns with the theoretically expected $O(3n)$ global memory throughput across all tested devices.
+
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{bib}
 

### 2025-03-01 paper.tex
Split Appendix

@@
 \subsection{Earlier Attempts at Portability}
-Our first attempt \raph{we'll add the cite post-review} at implementing a \emph{Decoupled Lookback} scan without relying on FPG was \emph{Scalar Fallback}.
+Our first attempt at implementing a \emph{Decoupled Lookback} scan without relying on FPG was \emph{Scalar Fallback}.
 Each time a workgroup spins waiting on the reduction from a preceding tile, it advances a \emph{fallback} operation by processing one scalar element from that tile.
@@
     \If{$\textit{wg\_control} = \textnormal{LOCKED}$}{
-      \textbf{FallbackReduce}($\textit{in}$, $\textit{wg\_fallback}$)\;
+      \textbf{Fallback}($\textit{in}$, $\textit{wg\_fallback}$)\;
       \textbf{FallbackInsertionAttempt}($\textit{spine}$, $\textit{wg\_fallback}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
@@
 Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking.
-Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.
+Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes:
 
@@
   \centering
-  \includegraphics[width=\linewidth]{graphics/split.pdf}
+  \includegraphics[width=.7\linewidth]{graphics/split.pdf}
   \caption{Splitting one 32-bit value into two 32-bit values with bit-packed tile state.}
@@
 \subsubsection{Cross-API Performance Versus Contemporary Implementations}
-We first compare \emph{Decoupled Fallback} (\emph{DF}) to two key benchmarks: \emph{Memcpy}\textdagger{~\ref{sec:memcpy}} and the \emph{Decoupled Lookback} (\emph{DL}) implementation found in NVIDIA's CUB library, using the 2080 Super only.
-To isolate the impact of differing APIs, we implement and evaluate \emph{DF} in both CUDA and Dawn.\footnote{Due to WebGPU's 128~MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.} \emph{Memcpy} serves as a performance reference, as it shares the same $2n$ global memory movement as \emph{DF} but lacks compute overhead and is purely memory bandwidth bound.
+We first compare \emph{Decoupled Fallback} (\emph{DF}) to two key benchmarks: \emph{Memcpy}\textdagger{A1} and the \emph{Decoupled Lookback} (\emph{DL}) implementation found in NVIDIA's CUB library, using the 2080 Super only.
+To isolate the impact of differing APIs, we implement and evaluate \emph{DF} in both CUDA and Dawn.\footnote{Due to WebGPU's 128~MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.} \emph{Memcpy} serves as a baseline reference, as it shares the same $2n$ global memory movement as \emph{DF} but lacks compute overhead and is purely memory bandwidth bound.
 
@@
 \subsubsection{Same-API Performance Versus Baseline Measures}
-We next compare \emph{DF} against \emph{Reduce-then-Scan}\textdagger{~\ref{sec:rts}} (\emph{RTS}), which serves as a baseline to highlight the performance improvements of our approach over the prior FPG-constrained technique.
+We next compare \emph{DF} against \emph{Reduce-then-Scan}\textdagger{A2} (\emph{RTS}), which serves as a baseline to highlight the performance improvements of our approach over the prior FPG-constrained technique.
 This evaluation is conducted using Dawn across the full range of tested devices.
@@
 
-Furthermore, this data reveals two notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, indicating that fairness and FPG are not necessarily correlated.
+Furthermore, this data reveals two notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, demonstrating that fairness and FPG are not necessarily correlated.
 (2) The ARM Mali-G78 MP20 exhibits the most unfair scheduling among all tested devices, with a fallback rate of 30.8\% relative to workgroups launched.
@@
 Devices that already lack sufficient compute resources to reach SOL performance are especially affected, as the additional compute overhead from fallback operations cannot be hidden by memory loads and stores.
-The worst-case scenario arises in devices that suffer from both low compute power and unfair scheduling.Another potential bottleneck is the speed of atomic operations.
-On some older GPUs atomics do not operate within the cache, resulting in significantly higher latency.
-Slower atomics lead to delayed tile updates, which in turn increase lookback lengths.
+The worst-case scenario arises in devices that suffer from both low compute power and unfair scheduling.
+Another potential bottleneck is the speed of atomic operations.
+On GPUs where atomics do not operate within the cache, the resulting higher latency leads to delayed tile updates, which in turn increases lookback lengths.
 
@@
 
-\begin{acks}
-\end{acks}
-
 \bibliographystyle{ACM-Reference-Format}
 \bibliography{bib}
-
-\clearpage
-\appendix
-\section{Artifact}
-\subsection{Availability}
-We provide our artifact in the following GitHub repository: \thomas{Will anonymize with GithubAnonymous once we are ready}. The repository contains per API requirements and building instructions.
-
-\subsection{Baseline Comparisons: \emph{Memcpy}}%
-\label{sec:memcpy}
-\begin{table}
-  \small
-  \centering
-  \begin{tabular}{l r r}
-    \toprule
-    Device                                & \makecell{Memory           \\ Bandwidth \\ (GB/s)} & \makecell{Expected \\ Throughput \\ (32-bit ele/sec)} \\
-    \midrule
-    Intel HD620 Single-Channel            & 17.1             & 2.13e9  \\
-    Intel HD620 Dual-Channel              & 34.1             & 4.26e9  \\
-    ARM Mali-G78 MP20 / Tensor (Pixel 6a) & 51.2             & 6.40e9  \\
-    Apple M3 (10 Cores)                   & 102.4            & 12.8e9  \\
-    Apple M1 Max (32 Cores)               & 409.6            & 51.2e9  \\
-    Nvidia 2080 Super                     & 496              & 62.0e9  \\
-    AMD 7900 XT                           & 800              & 100.0e9 \\
-    \bottomrule
-  \end{tabular}
-  \caption{Vendor-Specified Memory Bandwidth and Expected Throughput (32-bit Elements per Second).\label{tab:memory_bandwidth}}
-\end{table}
-Although all of Dawn's backend APIs support \emph{timestamp queries inside encoders}, Apple GPUs do not, so we are forced to implement our own \emph{Memcpy} kernel to measure memory bandwidth, as opposed to simply timing an API's copy buffer operation.
-It is well known that \emph{observed} peak memory bandwidth tends to diverge from vendor-listed peak bandwidth, so for completeness we list these figures above.
-We make a number of observations:
-\begin{itemize}
-  \item \textbf{Intel HD620:} As our HD620 laptop is only equipped with a single memory DIMM, its peak bandwidth is limited to the lower bound, 2.13e9 ele/sec.
-Thus, our observed bandwidth of 1.498e9 is reasonable.
-  \item \textbf{ARM Mali-G78 MP20 / Pixel 6a (Tensor):} The observed throughput is significantly lower than expected, despite a vendor-listed bandwidth of 6.40e9 ele/sec.
-The cause of this discrepancy remains unclear, but we note other developers have observed similar speeds.
-  \item \textbf{Apple M1 Pro Max:} Observed throughput matches expectations.
-  \item \textbf{Nvidia 2080 Super:} Observed throughput matches expectations.
-  \item \textbf{AMD 7900 XT:} Our above result, where \emph{DF} was observed to be faster than \emph{Memcpy}, is likely due to compute latency counterintuitively improving memory access patterns, leading to reduced paging.
-Furthermore, we see that our \emph{DF} throughput is within spec and reasonable.
-\end{itemize}
-
-\subsection{Baseline Comparisons: \emph{Reduce-then-Scan}}
-\label{sec:rts}
-As we desire the most competitive baseline possible, our \emph{RTS} kernels employ the same intra-workgroup scan strategy and tuning parameters as \emph{DF}.
-Initially, we adopted Merrill and Grimshaw's~\cite{Merrill2009} workgroup raking approach for inter-workgroup processing.
-However, empirical testing revealed that raking was consistently outperformed—by up to $\sim$15\% on lower-end devices—by a simpler approach that processes the spine serially with a single workgroup.
-Furthermore, our \emph{RTS} performance closely aligns with the theoretically expected $O(3n)$ global memory movement.
 

### 2025-03-01 paper.tex
Put back math for workgroup-wide scan.

@@
 Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking.
-Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes:
+
+For a subgroup of size $s$, the Kogge-Stone network has a work complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
+\begin{equation}
+  \begin{aligned}[b]
+    \text{Work} & = \underbrace{\vphantom{\sum_{k=0}^{\lceil \log_s n \rceil}}n}_{\text{fanout}}
+    +
+    \underbrace{\sum_{k=1}^{\lceil \log_s n \rceil}}_{\text{iterations}}
+    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}\frac{n}{s^k}}_{\text{calls per iteration}}
+    \cdot
+    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}s \log_2 s}_{\text{work per call}} \\
+                & = n + s \log_2 s \cdot \sum_{k=1}^{\lceil \log_s n \rceil} \frac{n}{s^k}       \\
+                & = O\left(n + n \log_2 s\right)                                                 \\
+                & = O(n \log_2 s)                                                                \\
+                & = O(n).
+  \end{aligned}
+\end{equation}
+Since the subgroup size remains constant within each kernel launch, the final work complexity is $O(n)$. The Kogge-Stone scan has a depth of $\log_2 s$; the total depth of the workgroup-wide scan is:
+\begin{equation}
+  \begin{aligned}[b]
+    \text{Depth} & = \underbrace{\lceil \log_s n \rceil}_{\text{loop iterations}}
+    \cdot \underbrace{\log_2 s}_{\text{depth per iteration}}                           \\
+                 & = \left\lceil \frac{\log_2 n}{\log_2 s} \right\rceil \cdot \log_2 s \\
+                 & = \log_2 n + O(1).
+  \end{aligned}
+\end{equation}
+Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.
+However, while minimal depth is preserved regardless of the subgroup size, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a workgroup-wide barrier at each iteration.
+Consequently, while the theoretical depth remains constant across subgroup sizes, smaller subgroups tend to perform worse in practice due to the increased relative cost of synchronization and the overhead of additional iterations in the main loop.
 

### 2025-03-01 paper.tex
Conclusion edit.

@@
 
-As the compute capabilities of major graphics APIs---Vulkan, Direct3D, and Metal---have grown, there has been increasing interest in adapting \emph{Decoupled Lookback} beyond CUDA for broader adoption in real-time graphics and GPU computing.
+As the compute capabilities of major graphics APIs have grown, there has been increasing interest in adapting \emph{Decoupled Lookback} beyond CUDA for broader adoption in real-time graphics and GPU computing.
 However, portable implementation is constrained by both hardware and API limitations.
@@
 Our implementation targets portability across modern GPU hardware and software platforms.
-It is implemented in the WebGPU Shading Language (WGSL)---an apex-level shading language that transpiles to backend API languages---ensuring seamless integration with modern graphics APIs.
+It is implemented in the WebGPU Shading Language (WGSL)---an apex-level shading language that transpiles to backend API languages---ensuring compatibility with all backend graphics APIs: Vulkan, Direct3D, and Metal.
 Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.
@@
 Unlike our simulation, which allows workgroups to eventually continue, true starvation results in increased lookback lengths, as blocked workgroups cannot complete \emph{Chained Propagation} and update their flags to inclusive.
-Furthermore, starvation effectively leads to \emph{occupancy reduction}: execution resources remain available, but fewer subgroups can context-switch to mitigate latency
+Furthermore, starvation effectively leads to \emph{occupancy reduction}: execution resources remain available, but fewer subgroups can context-switch to mitigate latency.
 \section{Discussion}
@@
 \subsection{Conclusion}
-In this paper, we presented \emph{Decoupled Fallback}, a single-pass \emph{Chained Scan} capable of reaching SOL performance under all progress models and suitable for implementation within the constraints of contemporary portable shading languages.
-Our approach maintains minimal data movement while ensuring robust execution across diverse GPU architectures.
+In this paper, we introduced \emph{Decoupled Fallback}, a single-pass \emph{Chained Scan} capable of achieving SOL performance across all progress models while remaining compatible with modern graphics APIs and hardware constraints.
+Our approach enables any workgroup to compute the reduction of any work tile, eliminating the risk of starvation.
+We reexamined the progress models proposed by Sorensen et al.~\cite{sorensen2021} in the context of dynamically assigned work, demonstrating that Chained Scan implementations require, at minimum, the OBE guarantee.
+Additionally, we highlighted the risks of running \emph{Decoupled Lookback} without FPG.
+Furthermore, we introduced a subgroup-accelerated yet subgroup-size-agnostic scan based on a generalized radix-$s$ Ladner-Fischer scan, along with a tile state representation that eliminates reliance on 64-bit atomics and memory barriers by distributing tile data across multiple values.
+
+Our artifact is available in the following GitHub repository: TODO. The repository contains per API requirements and building instructions.
 

### 2025-03-01 paper.tex
Deadlock -> Starvation

@@
 \begin{abstract}
-  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{forward-progress guarantees} (FPG) while avoiding deadlock.
+  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{forward-progress guarantees} (FPG) while avoiding \emph{starvation}.
 Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation.
@@
 \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{speed-of-light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device.
-However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite starvation and will never complete.
+However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite \emph{starvation} and will never complete.
 In practice, it is easy to demonstrate this highly undesirable behavior.
@@
 
-We present \emph{Decoupled Fallback}, a fully portable prefix scan capable of achieving SOL performance without requiring forward-progress guarantees (FPG).
+We present \emph{Decoupled Fallback}, a fully portable \emph{Chained Scan} capable of achieving SOL performance without requiring forward-progress guarantees (FPG).
 Our implementation targets portability across modern GPU hardware and software platforms.
@@
 Given such a guarantee, the predecessor will eventually produce a result and broadcast it.
-On devices lacking such a guarantee, the workgroup may wait indefinitely, resulting in deadlock.
+On devices lacking such a guarantee, the workgroup may continuously spin without making progress, resulting in \emph{starvation}---consuming execution resources that would otherwise allow the blocking tile to complete its reduction and stalling the kernel indefinitely.
 
@@
 Work and memory accesses are constant-bounded to the workgroup occupancy $o$ by posting the inclusive reduction after the lookback.
-However, granting each workgroup control over its dependency resolution does not eliminate reliance on \emph{progress guarantees}, as workgroups must still spin on tiles that have yet to post their reductions, risking deadlock.
+However, granting each workgroup control over its dependency resolution does not eliminate reliance on \emph{progress guarantees}, as workgroups must still spin on tiles that have yet to post their reductions, risking starvation.
 
@@
 While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple's GPUs do not.
-Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
+Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite starvation, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
 
 \subsection{Progress Guarantees and Their Impact on \emph{Chained Scan}}
-A \emph{forward-progress guarantee} ensures that all occupant subgroups---of possibly different workgroups---eventually make progress toward termination, preventing indefinite starvation.
+A \emph{forward-progress guarantee} ensures that all occupant subgroups---of possibly different workgroups---are eventually given execution and make progress toward termination, preventing indefinite starvation.
 Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees.
@@
 Each time a workgroup spins waiting on the reduction from a preceding tile, it advances a \emph{fallback} operation by processing one scalar element from that tile.
-This eliminates the possibility of deadlock because the number of such operations is bounded; the wait will eventually terminate, possibly at the expense of doing redundant work.
+This eliminates the possibility of starvation because the number of such operations is bounded; the wait will eventually terminate, possibly at the expense of doing redundant work.
 
@@
                           \item Otherwise, it continues spinning until the tile transitions to a \emph{Ready} or \emph{Inclusive} state.
-Without FPG, \emph{Decoupled Lookback} risks spinning indefinitely and deadlocking at this stage.
+Without FPG, \emph{Decoupled Lookback} risks spinning indefinitely at this stage.
                         \end{itemize}

### 2025-03-01 paper.tex
Remove Macros
Add anonymous repo.
Final Edits.

@@
 
-We present \emph{Decoupled Fallback}, a fully portable \emph{Chained Scan} capable of achieving SOL performance without requiring forward-progress guarantees (FPG).
+We present \emph{Decoupled Fallback}, a fully portable \emph{Chained Scan} capable of achieving SOL performance without requiring forward-progress guarantees.
 Our implementation targets portability across modern GPU hardware and software platforms.
@@
 In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume.
-This approach depends on a forward-progress guarantee of the GPU's execution model.
+This approach depends on a forward-progress guarantee of the GPU's scheduling model.
 Given such a guarantee, the predecessor will eventually produce a result and broadcast it.
@@
 (2) \emph{Chained Scan} only requires forward progress after a workgroup has atomically acquired its tile index.
-By definition, a workgroup that has acquired its index must be occupant on a physical processor (i.e., it has made at least one execution step); thus, OBE guarantees forward progress.
+By definition, a workgroup that has acquired its index must be occupant on a physical processor (i.e., it has made at least one execution step); thus, the OBE guarantee is sufficient for forward progress.
 
@@
 
-For a subgroup of size $s$, the Kogge-Stone network has a work complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
+For a subgroup of size $s$, the Kogge-Stone subgroup network has a work complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
 \begin{equation}
@@
 Unlike our simulation, which allows workgroups to eventually continue, true starvation results in increased lookback lengths, as blocked workgroups cannot complete \emph{Chained Propagation} and update their flags to inclusive.
-Furthermore, starvation effectively leads to \emph{occupancy reduction}: execution resources remain available, but fewer subgroups can context-switch to mitigate latency.
+Furthermore, starvation effectively leads to \emph{occupancy reduction}: execution resources remain available, but fewer subgroups are scheduled for execution, reducing opportunities for context switching to mitigate latency.
 \section{Discussion}
@@
 Due to its increased latency, synchronization overhead, and longer critical path, \emph{Decoupled Fallback} exacerbates existing challenges faced by \emph{Decoupled Lookback}, particularly a lack of compute capacity and unfair scheduling.
-Devices that already lack sufficient compute resources to reach SOL performance are especially affected, as the additional compute overhead from fallback operations cannot be hidden by memory loads and stores.
+Devices that already lack sufficient compute resources to reach SOL performance are especially affected, as the additional compute overhead from fallback operations cannot be hidden by global loads and stores.
 The worst-case scenario arises in devices that suffer from both low compute power and unfair scheduling.
@@
 In this paper, we introduced \emph{Decoupled Fallback}, a single-pass \emph{Chained Scan} capable of achieving SOL performance across all progress models while remaining compatible with modern graphics APIs and hardware constraints.
-Our approach enables any workgroup to compute the reduction of any work tile, eliminating the risk of starvation.
+Our approach enables any workgroup to compute the reduction of any blocking tile, eliminating the risk of starvation.
 We reexamined the progress models proposed by Sorensen et al.~\cite{sorensen2021} in the context of dynamically assigned work, demonstrating that Chained Scan implementations require, at minimum, the OBE guarantee.
@@
 
-Our artifact is available in the following GitHub repository: TODO. The repository contains per API requirements and building instructions.
+Our artifact is available in the following GitHub repository: https://anonymous.4open.science/status/Decoupled-Fallback-Paper-A33C.
+The repository contains per API requirements and building instructions.
 

### 2025-06-14 artifact/Dawn/main.cpp
Update artifact (not done yet).

@@ -5,6 +5,7 @@
 #include <fstream>
 #include <future>
 #include <iostream>
+#include <map>
 #include <sstream>
 #include <string>
 #include <vector>
@@ -23,7 +24,8 @@ struct ComputeShader {
 };
 
 struct Shaders {
-    ComputeShader init;
+    ComputeShader initPass;
+    ComputeShader initBatch;
     ComputeShader reduce;
     ComputeShader spineScan;
     ComputeShader downsweep;
@@ -47,14 +49,27 @@ struct GPUBuffers {
     wgpu::Buffer misc;
 };
 
+enum TestType {
+    Rts,
+    Csdl,
+    Csdldf,
+    Full,
+    SizeCsdldf,
+    SizeMemcpy,
+};
+
 struct TestArgs {
     GPUContext& gpu;
     GPUBuffers& buffs;
     Shaders& shaders;
     uint32_t size;
-    uint32_t batchSize;
     uint32_t workTiles;
     uint32_t readbackSize;
+    uint32_t warmupRuns = 1;
+    uint32_t totalRuns;
+    uint32_t batchSize;
+    uint32_t maxQueryEntries;
+    uint32_t miscStride;
     bool shouldValidate = false;
     bool shouldReadback = false;
     bool shouldTime = false;
@@ -62,27 +77,38 @@ struct TestArgs {
     bool shouldRecord = false;
 };
 
+struct TestStatistics {
+    uint64_t totalTime = 0ULL;
+    uint64_t maxTime = 0ULL;
+    uint64_t minTime = ~0ULL;
+    std::map<uint64_t, unsigned int> timeMap;
+    uint64_t totalSpins = 0;
+    uint64_t totalLookbackLength = 0;
+    uint64_t totalFallbacksInitiated = 0;
+    uint64_t totalSuccessfulInsertions = 0;
+};
+
 struct DataStruct {
     std::vector<double> time;
-    std::vector<double> totalSpins;
-    std::vector<double> lookbackLength;
+    std::vector<uint32_t> totalSpins;
+    std::vector<uint32_t> lookbackLength;
     std::vector<uint32_t> fallbacksInitiated;
     std::vector<uint32_t> successfulInsertions;
 
     DataStruct(const TestArgs& args) {
-        if(args.shouldRecord){
-            time.resize(args.batchSize);
-            totalSpins.resize(args.batchSize);
-            lookbackLength.resize(args.batchSize);
[…]

### 2025-06-14 artifact/Dawn/main_latest.cpp

@@ -1,1082 +1,0 @@
-#include <dawn/webgpu_cpp.h>
-
-#include <cstdlib>
-#include <cstring>
-#include <fstream>
-#include <future>
-#include <iostream>
-#include <sstream>
-#include <string>
-#include <vector>
-
-struct GPUContext {
-    wgpu::Instance instance;
-    wgpu::Device device;
-    wgpu::Queue queue;
-    wgpu::QuerySet querySet;
-};
-
-struct ComputeShader {
-    wgpu::BindGroup bindGroup;
-    wgpu::ComputePipeline computePipeline;
-    std::string label;
-};
-
-struct Shaders {
-    ComputeShader init;
-    ComputeShader reduce;
-    ComputeShader spineScan;
-    ComputeShader downsweep;
-    ComputeShader csdl;
-    ComputeShader csdldf;
-    ComputeShader csdldfOcc;
-    ComputeShader csdldfSimulate;
-    ComputeShader memcpy;
-    ComputeShader validate;
-};
-
-struct GPUBuffers {
-    wgpu::Buffer info;
-    wgpu::Buffer scanIn;
-    wgpu::Buffer scanOut;
-    wgpu::Buffer scanBump;
-    wgpu::Buffer reduction;
-    wgpu::Buffer timestamp;
-    wgpu::Buffer readbackTimestamp;
-    wgpu::Buffer readback;
-    wgpu::Buffer misc;
-};
-
-struct TestArgs {
-    GPUContext& gpu;
-    GPUBuffers& buffs;
-    Shaders& shaders;
-    uint32_t size;
-    uint32_t batchSize;
-    uint32_t workTiles;
-    uint32_t readbackSize;
-    bool shouldValidate = false;
-    bool shouldReadback = false;
-    bool shouldTime = false;
-    bool shouldGetStats = false;
-    bool shouldRecord = false;
-};
-
-struct DataStruct {
-    std::vector<double> time;
-    std::vector<double> totalSpins;
-    std::vector<double> lookbackLength;
-    std::vector<uint32_t> fallbacksInitiated;
-    std::vector<uint32_t> successfulInsertions;
-
-    DataStruct(const TestArgs& args) {
-        if (args.shouldRecord) {
-            time.resize(args.batchSize);
-            totalSpins.resize(args.batchSize);
-            lookbackLength.resize(args.batchSize);
-            fallbacksInitiated.resize(args.batchSize);
-            successfulInsertions.resize(args.batchSize);
-        }
-    }
-};
[…]

### 2025-06-14 artifact/Dawn/main.cpp
Working, just need to check csv generation

@@ -65,7 +65,7 @@ struct TestArgs {
     uint32_t size;
     uint32_t workTiles;
     uint32_t readbackSize;
-    uint32_t warmupRuns = 1;
+    uint32_t warmupRuns;
     uint32_t totalRuns;
     uint32_t batchSize;
     uint32_t maxQueryEntries;
@@ -122,8 +122,7 @@ void GetGPUContext(GPUContext* context, uint32_t maxQueryEntries) {
     std::promise<void> adaptPromise;
     instance.RequestAdapter(
         &options, wgpu::CallbackMode::AllowProcessEvents,
-        [&](wgpu::RequestAdapterStatus status, wgpu::Adapter adapt,
-            wgpu::StringView) {
+        [&](wgpu::RequestAdapterStatus status, wgpu::Adapter adapt, wgpu::StringView) {
             if (status == wgpu::RequestAdapterStatus::Success) {
                 adapter = adapt;
             } else {
@@ -132,30 +131,21 @@ void GetGPUContext(GPUContext* context, uint32_t maxQueryEntries) {
             adaptPromise.set_value();
         });
     std::future<void> adaptFuture = adaptPromise.get_future();
-    while (adaptFuture.wait_for(std::chrono::nanoseconds(100)) ==
-           std::future_status::timeout) {
+    while (adaptFuture.wait_for(std::chrono::nanoseconds(100)) == std::future_status::timeout) {
         instance.ProcessEvents();
     }
 
     wgpu::AdapterInfo info{};
     adapter.GetInfo(&info);
-    std::cout << "VendorID: " << std::hex << info.vendorID << std::dec
+    std::cout << "VendorID: " << std::hex << info.vendorID << std::dec << std::endl;
+    std::cout << "Vendor: " << std::string(info.vendor.data, info.vendor.length) << std::endl;
+    std::cout << "Architecture: " << std::string(info.architecture.data, info.architecture.length)
               << std::endl;
-    std::cout << "Vendor: " << std::string(info.vendor.data, info.vendor.length)
-              << std::endl;
-    std::cout << "Architecture: "
-              << std::string(info.architecture.data, info.architecture.length)
-              << std::endl;
-    std::cout << "DeviceID: " << std::hex << info.deviceID << std::dec
-              << std::endl;
-    std::cout << "Name: " << std::string(info.device.data, info.device.length)
-              << std::endl;
[…]

### 2025-06-14 artifact/Dawn/main.cpp
Validated csv generation

@@ -82,16 +82,16 @@ struct TestStatistics {
     uint64_t maxTime = 0ULL;
     uint64_t minTime = ~0ULL;
     std::map<uint64_t, unsigned int> timeMap;
-    uint64_t totalSpins = 0;
-    uint64_t totalLookbackLength = 0;
+    double totalSpins = 0.0;
+    double totalLookbackLength = 0.0;
     uint64_t totalFallbacksInitiated = 0;
     uint64_t totalSuccessfulInsertions = 0;
 };
 
 struct DataStruct {
     std::vector<double> time;
-    std::vector<uint32_t> totalSpins;
-    std::vector<uint32_t> lookbackLength;
+    std::vector<double> totalSpins;
+    std::vector<double> lookbackLength;
     std::vector<uint32_t> fallbacksInitiated;
     std::vector<uint32_t> successfulInsertions;
 
@@ -678,8 +678,7 @@ void RecordToCSV(const TestArgs& args, const DataStruct& data, const std::string
 
     if (args.shouldGetStats) {
         // Write full headers
-        file << "time,totalSpins,lookbackLength,fallbacksInitiated,"
-                "successfulInsertions\n";
+        file << "time,totalSpins,lookbackLength,fallbacksInitiated,successfulInsertions\n";
 
         // Write full data
         size_t rows = data.time.size();
@@ -722,10 +721,17 @@ void ProcessBatchResults(const TestArgs& args, const std::vector<uint64_t>& batc
 
         if (args.shouldGetStats) {
             uint32_t statsBaseIndex = k * args.miscStride;
-            uint32_t currentSpins = batchShaderStats[statsBaseIndex];
+            double currentSpins =
+                static_cast<double>(batchShaderStats[statsBaseIndex]) / args.workTiles;
             uint32_t currentFallbacks = batchShaderStats[statsBaseIndex + 1];
             uint32_t currentInsertions = batchShaderStats[statsBaseIndex + 2];
-            uint32_t currentLookback = batchShaderStats[statsBaseIndex + 3];
+            double currentLookback =
+                static_cast<double>(batchShaderStats[statsBaseIndex + 3]) / args.workTiles;;
+
+            stats.totalSpins += currentSpins;
+            stats.totalFallbacksInitiated += currentFallbacks;
+            stats.totalSuccessfulInsertions += currentInsertions;
+            stats.totalLookbackLength += currentLookback;
 
             if (args.shouldRecord) {
                 uint32_t runId = runsExecutedCount + k;
[…]

### 2025-06-14 artifact/Dawn/main.cpp
Fixes to arg parsing

@@ -980,21 +980,21 @@ struct CommandLineArgs {
     bool shouldRecord;
     std::string deviceName;
 };
-
 void ParseArguments(int argc, char* argv[], CommandLineArgs& args) {
     auto printUsage = []() {
         std::cerr
-            << "Usage: <TestType> [options] [deviceName]\n\n"
+            << "Usage: <TestType> [options]\n\n" // Removed [deviceName] from here
             << "  <TestType>: rts | csdl | csdldf | full | sizecsdldf | sizememcpy\n\n"
             << "Options:\n"
+            << "  --label <name>       Optional name to prefix output files (e.g., 'MyCoolGPU').\n"
             << "  --record             Enable recording to CSV.\n"
             << "  --runs <N>           Set the number of timed runs (default: 4192).\n"
             << "  --warmup <N>         Set the number of warmup runs (default: 1000).\n"
             << "  --batch <N>          Set the max batch size per submission (default: 2048).\n"
             << "  --size <N>           Set the problem size as a power of 2 (i.e., 1 << N) "
                "(default: 25).\n"
-            << "  -h, --help           Show this help message.\n\n"
-            << "  [deviceName]         Optional name to prefix output files (e.g., 'MyCoolGPU').\n";
+            << "  -h, --help           Show this help message.\n";
+            // Removed the separate [deviceName] description
     };
 
     if (argc < 2) {
@@ -1022,41 +1022,48 @@ void ParseArguments(int argc, char* argv[], CommandLineArgs& args) {
         throw std::invalid_argument("Invalid TestType: " + testTypeStr);
     }
 
-    int flags_end_index = argc;
-    if (argc > 2 && argv[argc - 1][0] != '-') {
-        args.deviceName = std::string(argv[argc - 1]) + "_";
-        flags_end_index = argc - 1;
+    std::vector<std::string> arguments;
+    for (int i = 2; i < argc; ++i) {
+        arguments.push_back(argv[i]);
     }
 
-    for (int i = 2; i < flags_end_index; ++i) {
-        std::string arg = argv[i];
+    for (size_t i = 0; i < arguments.size(); ++i) {
+        std::string arg = arguments[i];
         if (arg == "--record") {
             args.shouldRecord = true;
         } else if (arg == "--runs") {
-            if (i + 1 < flags_end_index)
-                args.totalRuns = std::stoul(argv[++i]);
-            else {
+            if (i + 1 < arguments.size()) {
[…]

### 2025-06-15 artifact/Dawn/main.cpp
Remove WGPU, now bench dawn on android

@@ -496,13 +496,17 @@ void SetComputePassTimed(const ComputeShader& cs, wgpu::CommandEncoder* comEncod
 
 void QueueSync(const GPUContext& gpu) {
     std::promise<void> promise;
-    gpu.queue.OnSubmittedWorkDone(wgpu::CallbackMode::AllowProcessEvents,
-                                  [&](wgpu::QueueWorkDoneStatus status) {
-                                      if (status != wgpu::QueueWorkDoneStatus::Success) {
-                                          std::cerr << "uh oh" << std::endl;
-                                      }
-                                      promise.set_value();
-                                  });
+    gpu.queue.OnSubmittedWorkDone(
+        wgpu::CallbackMode::AllowProcessEvents,
+        [&](wgpu::QueueWorkDoneStatus status, wgpu::StringView message) {
+            if (status != wgpu::QueueWorkDoneStatus::Success) {
+                std::cerr << "Queue work done failed! Status: " << static_cast<int>(status)
+                          << ", Message: " << std::string(message.data, message.length)
+                          << std::endl;
+            }
+            promise.set_value();
+        });
+
     std::future<void> future = promise.get_future();
     while (future.wait_for(std::chrono::nanoseconds(100)) == std::future_status::timeout) {
         gpu.instance.ProcessEvents();

### 2025-06-15 artifact/WGPU/src/main.rs

@@ -1,1125 +1,0 @@
-use std::{
-    env,
-    fs::File,
-    io::{self, Write},
-    vec,
-};
-
-fn div_round_up(x: u32, y: u32) -> u32 {
-    (x + y - 1) / y
-}
-
-struct GPUContext {
-    device: wgpu::Device,
-    queue: wgpu::Queue,
-    query_set: wgpu::QuerySet,
-    timestamp_freq: f32,
-}
-
-impl GPUContext {
-    async fn init() -> Self {
-        let instance = wgpu::Instance::new(&wgpu::InstanceDescriptor {
-            backends: wgpu::Backends::all(),
-            flags: wgpu::InstanceFlags::empty(),
-            backend_options: wgpu::BackendOptions::default(),
-        });
-
-        let adapter = instance
-            .request_adapter(&wgpu::RequestAdapterOptions {
-                power_preference: wgpu::PowerPreference::HighPerformance,
-                compatible_surface: None,
-                force_fallback_adapter: false,
-            })
-            .await
-            .expect("Failed to find an appropriate adapter");
-
-        let adapter_info = adapter.get_info();
-        println!("Adapter Info:");
-        println!("  Name: {}", adapter_info.name);
-        println!("  Vendor: {}", adapter_info.vendor);
-        println!("  Device: {}", adapter_info.device);
-        println!("  Backend: {:?}", adapter_info.backend);
-        println!("  Driver: {}", adapter_info.driver);
-        println!("  Driver Info: {}", adapter_info.driver_info);
-
-        let (device, queue) = adapter
-            .request_device(
-                &wgpu::DeviceDescriptor {
-                    label: None,
-                    required_features: wgpu::Features::TIMESTAMP_QUERY
-                        | wgpu::Features::SUBGROUP
-                        | wgpu::Features::SPIRV_SHADER_PASSTHROUGH,
-                    required_limits: wgpu::Limits::default(),
-                    memory_hints: wgpu::MemoryHints::Performance,
-                },
-                None,
-            )
-            .await
-            .expect("Failed to create device");
-
-        let query_set = device.create_query_set(&wgpu::QuerySetDescriptor {
-            label: Some("Timestamp Query Set"),
-            count: 6,
-            ty: wgpu::QueryType::Timestamp,
-        });
-
-        let timestamp_freq = queue.get_timestamp_period();
-
-        GPUContext {
-            device,
-            queue,
[…]

### 2025-06-16 paper.tex
changed ACM params to match camera ready requirements
just changing some spacing/layout, since we have room

@@
 
-\documentclass[acmsmall, manuscript, screen, review, anonymous]{acmart}
+\documentclass[sigconf,screen]{acmart}
+\settopmatter{printacmref=false, printfolios=false}
 %%
@@
 Other operating systems---such as macOS and Linux---employ their own timeout mechanisms with varying behaviors.
-For consistency, we adopt Windows’ 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program.
-\begin{figure}
+For consistency, we adopt Windows’s 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program.
+\begin{figure*}
   \centering
@@
   \label{fig:dist}
-\end{figure}
+\end{figure*}
 Exacerbating this issue, no major graphics API provides a mechanism to query FPG support.
@@
   \centering
-  \includegraphics[height=\textheight]{graphics/FlowChart.pdf}
+  \includegraphics[width=\columnwidth]{graphics/FlowChart.pdf}
   \caption{Flow Chart Representation of \emph{Decoupled Fallback}.\label{fig:decoupled_fallback}}
@@
 We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires native Android, necessitating the use of WGPU\@.\footnote{To keep the performance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
-\begin{table}
+\begin{table*}
   \centering
@@
   \caption{Hardware Specifications of Testing Devices.\label{tab:hardware}}
-\end{table}
+\end{table*}
 
@@
   \centering
-  \includegraphics[width=0.6\linewidth]{graphics/cuda_plot.pdf}
+  \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
   \caption{Performance versus CUB (NVIDIA 2080 Super).}

### 2025-06-19 paper.tex
Add new diagrams, overhaul some old diagrams

@@
 
+\begin{figure*}[!t]
+  \centering
+  \makebox[\textwidth][c]{
+    \begin{subfigure}[c]{0.48\textwidth}
+      \centering
+      \includegraphics[height=7.5cm,keepaspectratio]{graphics/ScanThenPropagate.pdf}
+      \caption{\emph{Scan-then-Propagate.}}
+      \label{fig:scan-then-propagate}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[c]{0.48\textwidth}
+      \centering
+      \includegraphics[height=6cm,keepaspectratio]{graphics/ReduceThenScan.pdf}
+      \caption{\emph{Raking Reduce-then-Scan.}}
+      \label{fig:reduce-then-scan}
+    \end{subfigure}%
+  }\\[1em]
+
+  \makebox[\textwidth][c]{
+    \begin{subfigure}[c]{0.48\textwidth}
+      \centering
+      \includegraphics[height=6cm,keepaspectratio]{graphics/StreamScan.pdf}
+      \caption{\emph{Stream Scan.}}
+      \label{fig:stream-scan}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[c]{0.48\textwidth}
+      \centering
+      \includegraphics[height=6cm,keepaspectratio]{graphics/DecoupledLookback.pdf}
+      \caption{\emph{Decoupled Lookback.}}
+      \label{fig:decoupled-lookback}
+    \end{subfigure}%
+  }
+
+  \caption{The four global scan techniques described in \ref{sec:Evolution}. Dashed grey boxes indicate individual kernel launches.}
+  \label{fig:four-techniques}
+\end{figure*}
+
 \subsection{Evolution of Inter-Workgroup Scan Architectures}
+\label{sec:Evolution}
 Contemporary GPU scans depart from early PRAM-like scans~\cite{Horn, Hensley, Sengupta2006, Gress} by leveraging the GPU memory hierarchy, which consists of progressively faster but increasingly private tiers.
@@
 
+\begin{figure*}[htbp]
+  \centering
+  \includegraphics[width=.95\textwidth]{graphics/DecoupledFallback.pdf}
+  \caption{\emph{Decoupled Fallback}. For brevity, the chained propagation phase is condensed, see \ref{fig:decoupled-lookback} for the full behavior.}
+  \label{fig:decoupled-fallback}
+\end{figure*}
+
 \section{Decoupled Fallback}
@@
 To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained Propagation} logic is typically performed by the first thread.
+
+To optimize the fallback operation, we employ three strategies.
+(1) First, a shared memory \emph{control flag} enables threads performing the lookback to signal the entire workgroup when a fallback is necessary.
+(2) Next, we limit the number of spins a thread can perform while waiting on a tile; exceeding this threshold triggers a fallback.
+While this approach risks \emph{false positive} fallbacks---since a late tile is indistinguishable from a genuinely blocking one---the maximum spin count can be tuned to minimize redundant fallbacks.
+Increasing the spin count lowers false positives but delays progress in true-blocking scenarios.
+False positives not only increase memory traffic but also consume compute resources that could be scheduled on the same multiprocessor.
+(3) After completing a fallback, the workgroup attempts to post the reduction result of the blocking tile, preventing redundant fallbacks by subsequent workgroups.
+
+The viability of \emph{Decoupled Fallback} relies on the same cache locality properties as \emph{Decoupled Lookback}.
+Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$\!, independent of FPG, due to false-positive fallback operations.
+As a result, memory bandwidth usage increases (and we expect throughput to decrease) by $O(fn)$ to $O((2 + f)n)$.
+However, just as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache.
+As we later empirically demonstrate in Section~\ref{sec:Sim-Blocking}, this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile.
+Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.
 
@@
 \end{algorithm}
-
-To optimize the fallback operation, we employ three strategies.
-(1) First, a shared memory \emph{control flag} enables threads performing the lookback to signal the entire workgroup when a fallback is necessary.
-(2) Next, we limit the number of spins a thread can perform while waiting on a tile; exceeding this threshold triggers a fallback.
-While this approach risks \emph{false positive} fallbacks---since a late tile is indistinguishable from a genuinely blocking one---the maximum spin count can be tuned to minimize redundant fallbacks.
-Increasing the spin count lowers false positives but delays progress in true-blocking scenarios.
-False positives not only increase memory traffic but also consume compute resources that could be scheduled on the same multiprocessor.
-(3) After completing a fallback, the workgroup attempts to post the reduction result of the blocking tile, preventing redundant fallbacks by subsequent workgroups.
-
-The viability of \emph{Decoupled Fallback} relies on the same cache locality properties as \emph{Decoupled Lookback}.
-Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$\!, independent of FPG, due to false-positive fallback operations.
-As a result, memory bandwidth usage increases (and we expect throughput to decrease) by $O(fn)$ to $O((2 + f)n)$.
-However, just as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache.
-As we later empirically demonstrate in Section~\ref{sec:Sim-Blocking}, this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile.
-Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.
 
@@
   \centering
-  \includegraphics[width=.7\linewidth]{graphics/split.pdf}
+  \includegraphics[width=1.03\linewidth]{graphics/split.pdf}
   \caption{Splitting one 32-bit value into two 32-bit values with bit-packed tile state.}

### 2025-06-19 paper.tex
Rebenched data.

@@
 \keywords{scan, parallel prefix, prefix sum, forward-progress guarantee, GPU, WGSL, decoupled fallback}
-
-%\received{?}
-%\received[revised]{?}
-%\received[accepted]{?}
 
@@
   \end{subfigure}
-  \caption{Histogram plots, 500 runs of Inclusive Prefix Sum, $\mathbf{2^{20}}$ 32-bit elements, on an FPG-lacking device, Apple M1 Max.
+  \caption{Histogram plots, 500 runs of Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements, on an FPG-lacking device, Apple M1 Max.
 Bar heights ($y$ axis) indicate the number of runs that completed in the execution time indicated on the $x$ axis.
+The Reduce-then-Scan threshold is the expected execution time of RTS at the same input size; runs to the left of this threshold indicate a performance uplift and vice versa.
 The TDR threshold is Windows's threshold for resetting the GPU\@; as noted in Section~\ref{sec:tdr}, MacOS handles this differently.}
@@
     \midrule
-    Avg \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)  & 1.50   & 2.48  & 10.25  & 38.62  & 54.39  & 72.28  \\
-    Avg \emph{DF} (32-bit ele/sec $\cdot 10^9$)   & 1.49   & 2.79  & 10.22  & 38.84  & 53.29  & 94.60  \\
-    Avg \emph{RTS} (32-bit ele/sec $\cdot 10^9$)      & 1.06   & 2.02  & 6.81   & 25.13  & 35.90  & 61.49  \\
-    \emph{DF}/\emph{Memcpy}       & 99.1\%    & 112.8\%  & 99.8\%    & 100.6\%   & 98.0\%    & 130.9\%   \\
-    \emph{DF}/\emph{RTS}          & 140.2\%   & 138.5\%  & 150.2\%   & 154.5\%   & 148.4\%   & 153.8\%   \\
-    Avg Spins per workgroup    & 0.305    & 1.477    & 0.606    & 0.604    & 0.419    & 0.148    \\
-    Avg Lookback Length per workgroup& 1.726    & 3.537    & 3.415    & 10.782   & 13.046   & 20.722   \\
-    Avg Total Fallbacks Initiated      & 254.662   & 2504.704  & 365.546   & 990.962   & 86.400   & 9.308   \\
-    Avg Total Successful Insertions    & 6.170     & 122.436   & 4.612     & 0.012    & 0.284    & 0.020    \\
+    Avg \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)       & 1.48    & 2.48     & 10.82   & 37.42    & 50.95    & 73.68    \\
+    Avg \emph{DF} (32-bit ele/sec $\cdot 10^9$)           & 1.51    & 2.19     & 10.87   & 36.85    & 51.25    & 83.27    \\
+    Avg \emph{RTS} (32-bit ele/sec $\cdot 10^9$)          & 1.06    & 1.93     & 7.46    & 25.75    & 34.36    & 62.57    \\
+    \emph{DF}/\emph{Memcpy}                              & 102.0\% & 88.4\%   & 100.5\% & 98.4\%   & 100.6\%  & 113.0\%  \\
+    \emph{DF}/\emph{RTS}                                 & 142.6\% & 113.6\%  & 145.7\% & 143.1\%  & 149.1\%  & 133.1\%  \\
+    Avg Spins per workgroup                              & 0.420   & 0.303    & 0.719   & 0.752    & 0.679    & 0.533    \\
+    Avg Lookback Length per workgroup                    & 1.714   & 4.036    & 4.598   & 10.737   & 12.350   & 22.648   \\
+    Avg Total Fallbacks Initiated                        & 446.183 & 245.404  & 449.961 & 1046.216 & 333.324  & 266.641  \\
+    Avg Total Successful Insertions                      & 21.475  & 0.733    & 23.887  & 4.822    & 11.378   & 46.288   \\
     \bottomrule
@@
 
-  \caption{\emph{Decoupled Fallback} Statistics Across Increasing Simulated Blocking Rates.
-Each subfigure highlights a key performance metric affected by the artificial omission of work tile updates, illustrating the resilience of \emph{DF} under extreme scheduling conditions.}
+  \caption{\emph{Decoupled Fallback} Statistics across increasing simulated blocking Rates.
+Each subfigure highlights a key performance metric affected by the artificial omission of work tile updates.}
   \label{fig:vertical_images}

### 2025-06-19 paper.tex
Ideal scan architecture fix

@@
 \begin{abstract}
-  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{forward-progress guarantees} (FPG) while avoiding \emph{starvation}.
+  We present \emph{Decoupled Fallback}, a work-stealing method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{forward-progress guarantees} (FPG) while avoiding \emph{starvation}.
 Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation.
-On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach.
-We further demonstrate the resilience of \emph{Decoupled Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
+On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WebGPU native and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach.
+We demonstrate that work-stealing is particularly well suited to \emph{Chained Scan}.
+By imposing blocking rates of up to 50\% to simulate unfair schedulers, we show that \emph{Decoupled Fallback} outperforms \emph{Reduce-then-Scan} even under extreme contention.
 \end{abstract}
@@
 Scan is one of the most ubiquitous operations in GPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more.
-The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture.
+The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}~\cite{Merrill2016}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture.
 This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$.
@@
 
-This work is guided by two main goals.
-(1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward-progress guarantees (FPG) or support for 64-bit atomics.
+This work is guided by two main goals: \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward-progress guarantees (FPG) or support for 64-bit atomics.
 \textbf{Performance:} We aim for near-SOL execution to the greatest extent possible allowed by the underlying hardware and programming model.
 
-However, certain aspects fall outside the scope of this work.
-\textbf{Performance guarantees} are not universally attainable, as not all architectures offer atomic operations that are sufficiently fast or scheduling models that are sufficiently fair to attain SOL performance.
+However, certain aspects fall outside the scope of this work: \textbf{Performance guarantees} are not universally attainable, as not all architectures offer atomic operations that are sufficiently fast or scheduling models that are sufficiently fair to attain SOL performance.
 \textbf{Subgroup divergence} is another challenge we acknowledge but do not address directly, as shading languages do not provide explicit divergence control like CUDA's subgroup masking mechanisms.
@@
 
-\begin{figure*}[!t]
+\begin{figure*}
   \centering
@@
 \subsection{Ideal Scan Architecture}
-\emph{Chained Scan} and \emph{Reduce-then-Scan} represent two contrasting paradigms in scan design.
-By eliding the barrier-idiom, \emph{Chained Scan} aligns with the oversubscription paradigm, delegating occupancy management to the workgroup scheduler.
-In contrast, \emph{Reduce-then-Scan} employs workgroup raking, requiring fixed launches and explicit occupancy control, forcing developers to manually tune launch sizes or implement occupancy discovery techniques.
+In terms of performance, an ideal global scan algorithm should minimize global memory movement, fully utilize the device’s memory bandwidth, and avoid costly global barriers.
+By these criteria alone, the \emph{Chained Scan} architecture emerges as the preferred solution; moreover, its single-kernel design offers compelling portability advantages across diverse hardware.
+By eliminating global barriers, \emph{Chained Scan} naturally conforms to the GPU’s oversubscription model, allowing occupancy control to be delegated the hardware's workgroup scheduler.
+In contrast, hierarchical approaches such as \emph{Reduce-then-Scan} require workgroup raking and fixed-occupancy persistent threads to bound the spine’s size, necessitating explicit kernel launches and manual tuning of launch configurations to maintain optimal occupancy.
 
 Hardware-level considerations underscore the limitations of the fixed-launch approach.
-On GPUs where the global memory data cache is indexed by physical addresses---and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)---fixed launches can lead to performance degradation due to cache thrashing.
+On GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—fixed launches can lead to performance degradation due to cache thrashing.
 This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups.
-Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns---or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations---may experience significant performance degradation.
+Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns—or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations—may experience significant performance degradation.
 
-Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture.
-Furthermore, its single-kernel design and serial ordering make it suitable for in-situ operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging.
-However, it has a critical limitation: its reliance on a forward-progress guarantee.
-While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple's GPUs do not.
-Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite starvation, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
+Moreover, the fused nature of \emph{Chained Scan} enables in-situ compaction: algorithms requiring compaction can merge that work directly into the scan kernel, avoiding redundant writes or secondary compaction passes.
+This fusion benefits applications ranging from data deduplication to compression.
+However, \emph{Chained Scan} has a critical limitation: its reliance on a forward-progress guarantee.
+While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple GPUs do not.
+Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient—it risks indefinite starvation, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
 
 \subsection{Progress Guarantees and Their Impact on \emph{Chained Scan}}
-A \emph{forward-progress guarantee} ensures that all occupant subgroups---of possibly different workgroups---are eventually given execution and make progress toward termination, preventing indefinite starvation.
  Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees.
@@
 \subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
+A \emph{forward-progress guarantee} ensures that all occupant subgroups---of possibly different workgroups---are eventually given execution and make progress toward termination, preventing indefinite starvation.
+In the most comprehensive
 The absence of FPG in key hardware segments poses a critical challenge for developers.
@@
 
-\begin{figure*}[htbp]
+\begin{figure*}
   \centering

### 2025-06-19 paper.tex
Reorder figures, remove progress classification section.

@@
 
-\setcopyright{acmlicensed}
-\copyrightyear{2025}
-\acmYear{2025}
-\acmDOI{XXXXXXX.XXXXXXX}
-\acmConference[SPAA '25]{37th ACM Symposium on Parallelism in Algorithms and Architectures}{July 28--August 1,
-  2025}{Portland, OR, USA}
-\acmISBN{978-1-4503-XXXX-X/18/06}
+\acmYear{2025}\copyrightyear{2025}
+\setcopyright{rightsretained}
+\acmConference[SPAA '25]{37th ACM Symposium on Parallelism in Algorithms and Architectures}{July 28--August 1, 2025}{Portland, OR, USA}
+\acmBooktitle{37th ACM Symposium on Parallelism in Algorithms and Architectures (SPAA '25), July 28--August 1, 2025, Portland, OR, USA}
+\acmDOI{10.1145/3694906.3743326}
+\acmISBN{979-8-4007-1258-6/25/07}
 
@@
 
-\begin{figure*}
+\begin{figure*}[!t]
   \centering
@@
 
-\subsection{Ideal Scan Architecture}
-In terms of performance, an ideal global scan algorithm should minimize global memory movement, fully utilize the device’s memory bandwidth, and avoid costly global barriers.
-By these criteria alone, the \emph{Chained Scan} architecture emerges as the preferred solution; moreover, its single-kernel design offers compelling portability advantages across diverse hardware.
-By eliminating global barriers, \emph{Chained Scan} naturally conforms to the GPU’s oversubscription model, allowing occupancy control to be delegated the hardware's workgroup scheduler.
-In contrast, hierarchical approaches such as \emph{Reduce-then-Scan} require workgroup raking and fixed-occupancy persistent threads to bound the spine’s size, necessitating explicit kernel launches and manual tuning of launch configurations to maintain optimal occupancy.
-
-Hardware-level considerations underscore the limitations of the fixed-launch approach.
-On GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—fixed launches can lead to performance degradation due to cache thrashing.
-This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups.
-Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns—or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations—may experience significant performance degradation.
-
-Moreover, the fused nature of \emph{Chained Scan} enables in-situ compaction: algorithms requiring compaction can merge that work directly into the scan kernel, avoiding redundant writes or secondary compaction passes.
-This fusion benefits applications ranging from data deduplication to compression.
-However, \emph{Chained Scan} has a critical limitation: its reliance on a forward-progress guarantee.
-While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple GPUs do not.
-Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient—it risks indefinite starvation, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
-
-\subsection{Progress Guarantees and Their Impact on \emph{Chained Scan}}
- Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees.
-Instead, Sorensen et al.\ systematically proposed and tested three progress models:\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}
-\begin{itemize}
-  \item \textbf{Occupancy-Bound Execution (OBE)}: Workgroups that have made at least one execution step.
-  \item \textbf{Heterogeneous System Architecture~\cite[Section~2.11]{HSASysArch12} (HSA)}: The workgroup with the lowest ID that has not yet terminated.
-  \item \textbf{Linear OBE (LOBE)}: Workgroups that have made at least one step, and any workgroup with a lower ID than a workgroup that has made at least one step.
-\end{itemize}
-Their 2021 findings~\cite{sorensen2021} demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior hypothesis~\cite{sorensen2018}.
-
-Sorensen et al.\ classify \emph{Chained Scan} as a \emph{one-way producer-consumer} (PC) HSA algorithm: a workgroup only consumes from lower-ID counterparts, ensuring that execution proceeds in a strictly ordered fashion.
-HSA guarantees termination, as the lowest ID workgroup is always given forward progress.
-If subgroups from the lowest ID workgroup are not guaranteed execution, then dependent workgroups may spin indefinitely, waiting on an update that never arrives.
-
-However, the HSA classification relies on the assumption that \emph{Chained Scan} follows scheduler-assigned virtualized IDs rather than dynamically acquired atomic-bumped IDs, as used by both \emph{StreamScan} and \emph{Decoupled Lookback}.
-Since virtual IDs do not necessarily match atomic IDs, HSA does not guarantee termination---the lowest virtual ID may not be scheduled first, potentially leading to starvation.
-
-LOBE adds an additional condition on top of OBE\@: lower-ID workgroups must be scheduled before higher-ID ones.
-However, this condition is unnecessary for \emph{Chained Scan} for two reasons: (1) We already guarantee serial ordering of work items through atomic bumping, making OBE sufficient.
-(2) \emph{Chained Scan} only requires forward progress after a workgroup has atomically acquired its tile index.
-By definition, a workgroup that has acquired its index must be occupant on a physical processor (i.e., it has made at least one execution step); thus, the OBE guarantee is sufficient for forward progress.
-
-Thus, by dynamically acquiring IDs through atomics, \emph{Chained Scan} shifts its progress model requirement from HSA to OBE.
-While LOBE encompasses both HSA and OBE, its stricter ordering guarantees are unnecessary for correctness.
-Ultimately, regardless of a device's specific progress model, any architecture lacking forward-progress guarantees will fail to execute \emph{Chained Scan}.
-In short:
-
-\begin{table}[h!]
-  \small
-  \centering
-  \begin{tabular}{lcccc}
-    \toprule
-    Algorithm                             & LOBE & OBE & HSA & None \\
-    \midrule
-    \emph{Chained Scan} with Workgroup ID & Yes  & No  & Yes & No   \\
-    \emph{Chained Scan} with Atomic ID    & Yes  & Yes & No  & No   \\
-    Ours                                  & Yes  & Yes & Yes & Yes  \\
-    \bottomrule
-  \end{tabular}
-  \caption{Comparison of \emph{Chained Scan} Variants and Our Implementation Across Progress Models.\label{tab:progress_models}}
-\end{table}
-
-\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
-A \emph{forward-progress guarantee} ensures that all occupant subgroups---of possibly different workgroups---are eventually given execution and make progress toward termination, preventing indefinite starvation.
-In the most comprehensive
-The absence of FPG in key hardware segments poses a critical challenge for developers.
-Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently.
-On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds.
-Other operating systems---such as macOS and Linux---employ their own timeout mechanisms with varying behaviors.
-For consistency, we adopt Windows’s 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program.
 \begin{figure*}
@@
 \end{figure*}
+
+\subsection{Ideal Scan Architecture}
+In terms of performance, an ideal global scan algorithm should minimize global memory movement, fully utilize the device’s memory bandwidth, and avoid costly global barriers.
+By these criteria alone, the \emph{Chained Scan} architecture emerges as the preferred solution; moreover, its single-kernel design offers compelling portability advantages across diverse hardware.
+By eliminating global barriers, \emph{Chained Scan} naturally conforms to the GPU’s oversubscription model, allowing occupancy control to be delegated the hardware's workgroup scheduler.
+In contrast, hierarchical approaches such as \emph{Reduce-then-Scan} require workgroup raking and fixed-occupancy persistent threads to bound the spine’s size, necessitating explicit kernel launches and manual tuning of launch configurations to maintain optimal occupancy.
+
+Hardware-level considerations underscore the limitations of the fixed-launch approach.
+On GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—fixed launches can lead to performance degradation due to cache thrashing.
+This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups.
+Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns—or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations—may experience significant performance degradation.
+
+Moreover, the fused nature of \emph{Chained Scan} enables in-situ compaction: algorithms requiring compaction can merge that work directly into the scan kernel, avoiding redundant writes or secondary compaction passes.
+This fusion benefits applications ranging from data deduplication to compression.
+However, \emph{Chained Scan} has a critical limitation: its reliance on a forward-progress guarantee.
+While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple GPUs do not.
+Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite starvation, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.
+
+\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
+\emph{Chained Scans} are \emph{one-way producer-consumer} algorithms; each workgroup consumes reductions only from lower-ID counterparts, thereby enforcing the ordering required to elide barriers.
+For such algorithms, guaranteeing execution of the lowest-ID workgroup is sufficient to ensure termination.
+If subgroups from the lowest-ID workgroup are not guaranteed execution, dependent workgroups may spin indefinitely, waiting for updates that never arrive.
+A \emph{forward-progress guarantee} ensures that all occupant subgroups---potentially from different workgroups---are eventually granted execution, thus preventing indefinite starvation.
+
+The absence of FPG in key hardware segments poses a critical challenge for developers.
+Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently.
+On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds.
+Other operating systems—such as macOS and Linux—employ their own timeout mechanisms with varying behaviors.
+For consistency, we adopt Windows’s 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program.
 Exacerbating this issue, no major graphics API provides a mechanism to query FPG support.
@@
 
+\begin{figure*}[t]
+  \centering
+  \includegraphics[width=.95\textwidth]{graphics/DecoupledFallback.pdf}
+  \caption{\emph{Decoupled Fallback}. For brevity, the chained propagation phase is condensed, see \ref{fig:decoupled-lookback} for the full behavior.}
+  \label{fig:decoupled-fallback}
+\end{figure*}
+
 \emph{Scalar Fallback} suffers from two primary inefficiencies.
@@
 While \emph{Scalar Fallback} achieved SOL performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce-then-Scan}.
-
-\begin{figure*}
-  \centering
-  \includegraphics[width=.95\textwidth]{graphics/DecoupledFallback.pdf}
-  \caption{\emph{Decoupled Fallback}. For brevity, the chained propagation phase is condensed, see \ref{fig:decoupled-lookback} for the full behavior.}
-  \label{fig:decoupled-fallback}
-\end{figure*}
 
@@
 Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.
+
+\section{Implementation}
+At a high level, our algorithm follows a five-phase construction:
+\begin{enumerate}
+  \item[(0)] \textbf {Initialization}: The first thread in the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
+  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering.
+For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
+  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking Kogge-Stone~\cite{5009159} subgroup scans,
+        allowing the subgroup-block scan to remain barrier-free and in registers.
+The subgroup-block reduction is then posted to shared memory.
+For a workgroup of size $w$, subgroup-raking reduces the size of the local spine to $w/s$.
+  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
+  \item \textbf{Chained Propagation, Decoupled Fallback}: The local reduction is posted to global memory.
+Then, the first subgroup in the workgroup performs a lookback operation to acquire the reduction result of all previous workgroups, engaging in fallback operations as necessary.
+Once completed, this subgroup posts the results to shared memory for the workgroup to consume.
+  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output.
+\end{enumerate}
+Phase 0 ensures that the work tiles are consumed in serial order.
+Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy.
+Phase 4, \emph{Decoupled Fallback}, is responsible for managing all inter-workgroup dependencies.
+As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations.
 
@@
 
-\section{Implementation}
-At a high level, our algorithm follows a five-phase construction:
-\begin{enumerate}
-  \item[(0)] \textbf {Initialization}: The first thread in the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
-  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering.
-For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
-  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking Kogge-Stone~\cite{5009159} subgroup scans,
-        allowing the subgroup-block scan to remain barrier-free and in registers.
-The subgroup-block reduction is then posted to shared memory.
-For a workgroup of size $w$, subgroup-raking reduces the size of the local spine to $w/s$.
-  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
-  \item \textbf{Chained Propagation, Decoupled Fallback}: The local reduction is posted to global memory.
-Then, the first subgroup in the workgroup performs a lookback operation to acquire the reduction result of all previous workgroups, engaging in fallback operations as necessary.
-Once completed, this subgroup posts the results to shared memory for the workgroup to consume.
-  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output.
-\end{enumerate}
-Phase 0 ensures that the work tiles are consumed in serial order.
-Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy.
-Phase 4, \emph{Decoupled Fallback}, is responsible for managing all inter-workgroup dependencies.
-As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations.
+\subsection{Workgroup-Wide Scan}
+The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches.
+To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{10.1145/322217.322232} scan network embedded with Kogge-Stone~\cite{5009159} subgroup scans to perform the scan across the local spine.
+Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts.
+To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$.
+Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction.
+Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index.
+However, such access is interpreted by the hardware as a subgroup \emph{broadcast} operation, resulting in no conflicts.
+Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking.
+
+For a subgroup of size $s$, the Kogge-Stone subgroup network has a work complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
+\begin{equation}
+  \begin{aligned}[b]
+    \text{Work} & = \underbrace{\vphantom{\sum_{k=0}^{\lceil \log_s n \rceil}}n}_{\text{fanout}}
+    +
+    \underbrace{\sum_{k=1}^{\lceil \log_s n \rceil}}_{\text{iterations}}
+    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}\frac{n}{s^k}}_{\text{calls per iteration}}
+    \cdot
+    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}s \log_2 s}_{\text{work per call}} \\
+                & = n + s \log_2 s \cdot \sum_{k=1}^{\lceil \log_s n \rceil} \frac{n}{s^k}       \\
+                & = O\left(n + n \log_2 s\right)                                                 \\
+                & = O(n \log_2 s)                                                                \\
+                & = O(n).
+  \end{aligned}
+\end{equation}
+Since the subgroup size remains constant within each kernel launch, the final work complexity is $O(n)$. The Kogge-Stone scan has a depth of $\log_2 s$; the total depth of the workgroup-wide scan is:
+\begin{equation}
+  \begin{aligned}[b]
+    \text{Depth} & = \underbrace{\lceil \log_s n \rceil}_{\text{loop iterations}}
+    \cdot \underbrace{\log_2 s}_{\text{depth per iteration}}                           \\
+                 & = \left\lceil \frac{\log_2 n}{\log_2 s} \right\rceil \cdot \log_2 s \\
+                 & = \log_2 n + O(1).
+  \end{aligned}
+\end{equation}
+Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.
+However, while minimal depth is preserved regardless of the subgroup size, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a workgroup-wide barrier at each iteration.
+Consequently, while the theoretical depth remains constant across subgroup sizes, smaller subgroups tend to perform worse in practice due to the increased relative cost of synchronization and the overhead of additional iterations in the main loop.
 
@@
 
-\subsection{Workgroup-Wide Scan}
-The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches.
-To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{10.1145/322217.322232} scan network embedded with Kogge-Stone~\cite{5009159} subgroup scans to perform the scan across the local spine.
-Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts.
-To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$.
-Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction.
-Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index.
-However, such access is interpreted by the hardware as a subgroup \emph{broadcast} operation, resulting in no conflicts.
-Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking.
-
-For a subgroup of size $s$, the Kogge-Stone subgroup network has a work complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
-\begin{equation}
-  \begin{aligned}[b]
-    \text{Work} & = \underbrace{\vphantom{\sum_{k=0}^{\lceil \log_s n \rceil}}n}_{\text{fanout}}
-    +
-    \underbrace{\sum_{k=1}^{\lceil \log_s n \rceil}}_{\text{iterations}}
-    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}\frac{n}{s^k}}_{\text{calls per iteration}}
-    \cdot
-    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}s \log_2 s}_{\text{work per call}} \\
-                & = n + s \log_2 s \cdot \sum_{k=1}^{\lceil \log_s n \rceil} \frac{n}{s^k}       \\
-                & = O\left(n + n \log_2 s\right)                                                 \\
-                & = O(n \log_2 s)                                                                \\
-                & = O(n).
-  \end{aligned}
-\end{equation}
-Since the subgroup size remains constant within each kernel launch, the final work complexity is $O(n)$. The Kogge-Stone scan has a depth of $\log_2 s$; the total depth of the workgroup-wide scan is:
-\begin{equation}
-  \begin{aligned}[b]
-    \text{Depth} & = \underbrace{\lceil \log_s n \rceil}_{\text{loop iterations}}
-    \cdot \underbrace{\log_2 s}_{\text{depth per iteration}}                           \\
-                 & = \left\lceil \frac{\log_2 n}{\log_2 s} \right\rceil \cdot \log_2 s \\
-                 & = \log_2 n + O(1).
-  \end{aligned}
-\end{equation}
-Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.
-However, while minimal depth is preserved regardless of the subgroup size, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a workgroup-wide barrier at each iteration.
-Consequently, while the theoretical depth remains constant across subgroup sizes, smaller subgroups tend to perform worse in practice due to the increased relative cost of synchronization and the overhead of additional iterations in the main loop.
-
 \subsection{Decoupled Fallback}
@@
 The details of tile state encoding and atomic updates are discussed in the following section and may vary depending on the data type, so for simplicity, we describe the operation as being performed by the \emph{first thread} of the workgroup:
-\begin{figure}
-  \centering
-  \includegraphics[width=\columnwidth]{graphics/FlowChart.pdf}
-  \caption{Flow Chart Representation of \emph{Decoupled Fallback}.\label{fig:decoupled_fallback}}
-\end{figure}
 \begin{enumerate}
@@
 \end{enumerate}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\columnwidth]{graphics/FlowChart.pdf}
+  \caption{Flow Chart Representation of \emph{Decoupled Fallback}.\label{fig:decoupled_fallback}}
+\end{figure}
 

### 2025-06-20 paper.tex
upload

@@
 Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation.
-On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WebGPU native and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach.
-We demonstrate that work-stealing is particularly well suited to \emph{Chained Scan}.
-By imposing blocking rates of up to 50\% to simulate unfair schedulers, we show that \emph{Decoupled Fallback} outperforms \emph{Reduce-then-Scan} even under extreme contention.
+On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WebGPU native and approaches the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach.
+We further demonstrate the resilience of \emph{Decoupled Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
 \end{abstract}
@@
 However, synchronization between workgroups is more limited.
-Standard GPU programming environments provide no control barrier between workgroups.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} The availability of a memory-only barrier varies across GPU environments; it is present in Direct3D and Vulkan, but was missing in versions of Metal prior to 3.2.
+Standard GPU programming environments provide no control barrier between workgroups.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent Thread}~\cite{gupta2012} paradigm, rather than serving as a global barrier across an unbounded launch.} The availability of a memory-only barrier varies across GPU environments; it is present in Direct3D and Vulkan, but was missing in versions of Metal prior to 3.2.
 Because of its non-portability, it is also absent from the WebGPU specification.
@@
 However, just as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache.
-As we later empirically demonstrate in Section~\ref{sec:Sim-Blocking}, this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile.
+This cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile.
 Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.
@@
 Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.
-However, while minimal depth is preserved regardless of the subgroup size, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a workgroup-wide barrier at each iteration.
+However, while minimal depth is preserved regardless of the subgroup size, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a barrier at each iteration.
 Consequently, while the theoretical depth remains constant across subgroup sizes, smaller subgroups tend to perform worse in practice due to the increased relative cost of synchronization and the overhead of additional iterations in the main loop.
@@
       \If{$\textit{thread\_id} < \textit{step}$}{
-        $\textit{temp} \gets \text{subgroupInclusiveScan}(\textit{x}[\textit{thread\_id} + \textit{top\_offset}])$\;
+        $\textit{temp} \gets \text{subgroupKoggeStoneScan}(\textit{x}[\textit{thread\_id} + \textit{top\_offset}])$\;
         $\textit{x}[\textit{thread\_id} + \textit{top\_offset}] \gets \textit{temp}$\;
@@
 Although this could also be accomplished with shared memory (and workgroup barriers), as an optimization, we use two or four threads from the first subgroup, leveraging subgroup \emph{ballot} to verify state consistency, and subgroup \emph{shuffle} to join and split reductions.
-With one thread per 16 bits and a minimum subgroup size of four in WebGPU, this method supports both 32-bit and 64-bit values, enabling 64-bit scans without 64-bit atomics or memory barriers.\footnote{Although native 64-bit values are not currently supported by the WebGPU specification, their future inclusion is anticipated.}
+With one thread per 16 bits and a minimum subgroup size of four in WebGPU, this method supports both 32-bit and 64-bit values, enabling 64-bit scans without 64-bit atomics or memory barriers.
 
@@
 Thus, a simple atomic store is not suitable.
-An atomic compare-and-swap would have the correct semantics, but WebGPU lacks this primitive (it does have \texttt{atomicCompareExchangeWeak}, which could possibly be made to work but makes reasoning about correctness and performance more difficult).
+An atomic compare-and-swap would have the correct semantics, but WebGPU lacks this primitive (it does have \emph{atomicCompareExchangeWeak}, which could possibly be made to work but makes reasoning about correctness and performance more difficult).
 Instead, we encode the tile state in the most significant bits of the atomic value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}.
-This allows an \texttt{atomicMaximum}, which successfully updates to a more advanced state but prevents a state regression, to correctly perform updates.
+This allows an \emph{atomicMaximum}, which successfully updates to a more advanced state but prevents a state regression, to correctly perform updates.
 
@@
 Our test hardware, shown in Figure~\ref{tab:hardware}, spans a wide range of FPG and performance capabilities.
-We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires native Android, necessitating the use of WGPU\@.\footnote{To keep the performance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
+We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing.
 \begin{table*}
@@
     \toprule
-    Vendor & Device Name            & LOBE         & Occupancy & Subgroup Size \\
+    Vendor & Device Name            & FPG         & Occupancy & Subgroup Size \\
     \midrule
@@
   \end{tabular}
-  \caption{Hardware Specifications of Testing Devices.\label{tab:hardware}}
+  \caption{Hardware Specifications of Testing Devices; FPG determinations are based on the data collected by Sorensen et. al.~\cite{sorensen2021}\label{tab:hardware}}
 \end{table*}
@@
 \subsubsection{Cross-API Performance Versus Contemporary Implementations}
-We first compare \emph{Decoupled Fallback} (\emph{DF}) to two key benchmarks: \emph{Memcpy}\textdagger{A1} and the \emph{Decoupled Lookback} (\emph{DL}) implementation found in NVIDIA's CUB library, using the 2080 Super only.
+We first compare \emph{Decoupled Fallback} (\emph{DF}) to two key benchmarks: \emph{Memcpy} and the \emph{Decoupled Lookback} (\emph{DL}) implementation found in NVIDIA's CUB library, using the 2080 Super only.
 To isolate the impact of differing APIs, we implement and evaluate \emph{DF} in both CUDA and Dawn.\footnote{Due to WebGPU's 128~MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.} \emph{Memcpy} serves as a baseline reference, as it shares the same $2n$ global memory movement as \emph{DF} but lacks compute overhead and is purely memory bandwidth bound.
@@
 
-\subsubsection{Same-API Performance Versus Baseline Measures}
-We next compare \emph{DF} against \emph{Reduce-then-Scan}\textdagger{A2} (\emph{RTS}), which serves as a baseline to highlight the performance improvements of our approach over the prior FPG-constrained technique.
-This evaluation is conducted using Dawn across the full range of tested devices.
-
 \begin{table*}
@@
     \midrule
-    Avg \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)       & 1.48    & 2.48     & 10.82   & 37.42    & 50.95    & 73.68    \\
-    Avg \emph{DF} (32-bit ele/sec $\cdot 10^9$)           & 1.51    & 2.19     & 10.87   & 36.85    & 51.25    & 83.27    \\
-    Avg \emph{RTS} (32-bit ele/sec $\cdot 10^9$)          & 1.06    & 1.93     & 7.46    & 25.75    & 34.36    & 62.57    \\
-    \emph{DF}/\emph{Memcpy}                              & 102.0\% & 88.4\%   & 100.5\% & 98.4\%   & 100.6\%  & 113.0\%  \\
-    \emph{DF}/\emph{RTS}                                 & 142.6\% & 113.6\%  & 145.7\% & 143.1\%  & 149.1\%  & 133.1\%  \\
-    Avg Spins per workgroup                              & 0.420   & 0.303    & 0.719   & 0.752    & 0.679    & 0.533    \\
-    Avg Lookback Length per workgroup                    & 1.714   & 4.036    & 4.598   & 10.737   & 12.350   & 22.648   \\
-    Avg Total Fallbacks Initiated                        & 446.183 & 245.404  & 449.961 & 1046.216 & 333.324  & 266.641  \\
-    Avg Total Successful Insertions                      & 21.475  & 0.733    & 23.887  & 4.822    & 11.378   & 46.288   \\
+    Avg \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)      & 1.48    & 2.50     & 10.82   & 37.42    & 50.95    & 73.68    \\
+    Avg \emph{DF} (32-bit ele/sec $\cdot 10^9$)          & 1.51    & 2.65     & 10.87   & 36.85    & 51.25    & 83.27    \\
+    Avg \emph{RTS} (32-bit ele/sec $\cdot 10^9$)         & 1.06    & 1.94     & 7.46    & 25.75    & 34.36    & 62.57    \\
+    \emph{DF}/\emph{Memcpy}                              & 102.0\% & 106.0\%  & 100.5\% & 98.4\%   & 100.6\%  & 113.0\%  \\
+    \emph{DF}/\emph{RTS}                                 & 142.6\% & 136.6\%  & 145.7\% & 143.1\%  & 149.1\%  & 133.1\%  \\
+    Avg Spins per workgroup                              & 0.420   & 2.52     & 0.719   & 0.752    & 0.679    & 0.533    \\
+    Avg Lookback Length per workgroup                    & 1.714   & 3.52     & 4.598   & 10.737   & 12.350   & 22.648   \\
+    Avg Total Fallbacks Initiated                        & 446.183 & 4739.91  & 449.961 & 1046.216 & 333.324  & 266.641  \\
+    Avg Total Successful Insertions                      & 21.475  & 571.24   & 23.887  & 4.822    & 11.378   & 46.288   \\
     \bottomrule
@@
 
-As shown in Table~\ref{tab:results}, \emph{DF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices, closely approaching the theoretically expected $O(3n)/O(2n)$ 50\% speedup, except on the ARM Mali.
-\emph{DF} closely matches \emph{Memcpy} across most devices, deviating only on the AMD 7900 XT and ARM Mali, where \emph{DF} outperforms \emph{Memcpy} by 30.9\% and 12.8\%, respectively.
-This suggests that additional compute latency may mitigate paging effects by separating loads from stores.
-On the ARM Mali, \emph{DF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%.
-However, its nearly identical performance to \emph{Memcpy} indicates that the implementation is memory-bandwidth-bound, limiting further gains.
+\subsubsection{Same-API Performance Versus Baseline Measures}
+We next compare \emph{DF} against \emph{Reduce-then-Scan} (\emph{RTS}), which serves as a baseline to highlight the performance improvements of our approach over the prior FPG-constrained technique.
+This evaluation is conducted using Dawn across the full range of tested devices.
+
+As shown in Table~\ref{tab:results}, \emph{DF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices, closely approaching the theoretically expected $O(3n)/O(2n)$ 50\% speedup, except on the 7900XT and Mali-G78.
+This result initially appears confusing, given that \emph{DF} outperforms \emph{Memcpy} by 6\% and 13\%, respectively.
+However, we hypothesized and confirmed a secondary effect of \emph{Chained Scans}: they enforce a queue-like behavior among workgroups.
+In-flight workgroups with a lower relative rank traverse fewer tiles during their lookbacks, enabling them to exit earlier than higher-ranked workgroups.
+Because this ordering emerges naturally from differing amounts of computation rather than explicit synchronization, the exit sequence is \emph{k-sorted} or \emph{nearly ordered}, rather than strictly ordered.
+Scheduling fairness and message-passing visibility can further influence this sequence---for example, an unfair scheduler might allocate more resources to a higher-ranked workgroup, causing it to exit before a lower-ranked counterpart, or a higher-ranked workgroup might observe an updated flag sooner, enabling earlier completion.
+We confirmed this behavior experimentally by using atomics to record and print the order in which workgroups completed their lookback phase.
+Unlike a simple \emph{Memcpy} kernel, where all in-flight workgroups begin writing immediately after completing reads, this “ordered exit” pattern facilitates more efficient global memory accesses during the write phase of the scan, as adjacent workgroups are more likely to share memory pages, reducing premature page evictions.
+We hypothesize that the 7900XT and Mali-G78 have insufficient memory systems to efficiently accommodate all in-flight workgroups, making them particularly sensitive to this effect.
+To test this hypothesis, we augmented our \emph{Memcpy} kernel with a simplified lookback routine---excluding the scan itself---and observed a performance improvement of approximately 10\% on the 7900XT.
+Thus, the unexpectedly slow baseline \emph{Memcpy} performance on these devices explains why \emph{DF}, despite its strong overall performance, does not fully achieve the theoretical 50\% speedup over \emph{RTS}.
 
@@
 
-\begin{itemize}
-  \item \textbf{Spins per Workgroup:} How many times does a workgroup spin while waiting?
-With perfectly fair scheduling, no waiting is necessary.
-Since no real-world scheduler is perfectly fair, we want minimal waiting.
-Despite varying occupancy rates, the average spin rate across tested devices remains low (typically one or fewer), indicating that most workgroups experience minimal delays.
-
-  \item \textbf{Lookback Length per Workgroup:} How many work tiles does a workgroup traverse to complete \emph{Chained Propagation}?
-Ideally, this traversal should be minimal, but longer traversal may result from unfair scheduling.
-Empirically, however, our data shows that lookback length remains well below the theoretical $O(o)$ bound---typically under 10\% of occupancy---meaning that most workgroups complete \emph{Chained Propagation} in relatively few steps compared to occupancy.
-
-  \item \textbf{Total Fallbacks Initiated:} The total number of fallback operations initiated across all workgroups.
-Ideally, this number would be zero on devices with FPG; because multiple workgroups may perform a fallback on the same blocking tile, this number can exceed the number of workgroups launched.
-In practice, this count is highly dependent on the fairness of the hardware scheduler.
-On most devices tested, initiated fallbacks remain modest---about 10\% or fewer of the total workgroups launched.
-
-  \item \textbf{Total Successful Insertions:} The total number of fallback operations that successfully update a blocking (or late) tile to \emph{Ready}.
-Due to false-positive fallbacks, this count does not precisely represent the number of true starvations, but instead provides an upper bound.
-Ideally, this number would also be zero on devices with FPG, but in practice, it is again dependent on the fairness of the scheduler.
-Unlike the number of fallback initiations, this statistic is extremely low---typically well under 1\% even on devices without FPG---suggesting that true starvation is rare but catastrophic.
-\end{itemize}
-
-Furthermore, this data reveals two notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, demonstrating that fairness and FPG are not necessarily correlated.
-(2) The ARM Mali-G78 MP20 exhibits the most unfair scheduling among all tested devices, with a fallback rate of 30.8\% relative to workgroups launched.
-However, the average number of spins per workgroup is only 1.477---far below our fallback trigger threshold of 4 spins.
-This suggests a bimodal distribution of waiting times, where workgroups either proceed immediately with minimal delay or experience prolonged stalls that trigger fallbacks.
-
-\subsection{Simulated Blocking}%
 \label{sec:Sim-Blocking}
@@
 
-As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near 100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced.
+\begin{itemize}
+  \item \textbf{Spins per Workgroup:} How many times does a workgroup spin while waiting?
+With perfectly fair scheduling, no waiting is necessary.
+Since no real-world scheduler is perfectly fair, we want minimal waiting.
+Despite varying occupancy rates, the average spin rate across tested devices remains low (typically one or fewer), indicating that most workgroups experience minimal delays.
+
+  \item \textbf{Lookback Length per Workgroup:} How many work tiles does a workgroup traverse to complete \emph{Chained Propagation}?
+Ideally, this traversal should be minimal, but longer traversal may result from unfair scheduling.
+Empirically, however, our data shows that lookback length remains well below the theoretical $O(o)$ bound---typically under 10\% of occupancy---meaning that most workgroups complete \emph{Chained Propagation} in relatively few steps compared to occupancy.
+
+  \item \textbf{Total Fallbacks Initiated:} The total number of fallback operations initiated across all workgroups.
+Ideally, this number would be zero on devices with FPG; because multiple workgroups may perform a fallback on the same blocking tile, this number can exceed the number of workgroups launched.
+In practice, this count is highly dependent on the fairness of the hardware scheduler.
+On most devices tested, initiated fallbacks remain modest---about 10\% or fewer of the total workgroups launched.
+
+  \item \textbf{Total Successful Insertions:} The total number of fallback operations that successfully update a blocking (or late) tile to \emph{Ready}.
+Due to false-positive fallbacks, this count does not precisely represent the number of true starvations, but instead provides an upper bound.
+Ideally, this number would also be zero on devices with FPG, but in practice, it is again dependent on the fairness of the scheduler.
+Unlike the number of fallback initiations, this statistic is extremely low---typically well under 1\% even on devices without FPG---suggesting that true starvation is rare but catastrophic.
+\end{itemize}
+
+Furthermore, this data reveals two notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, demonstrating that fairness and FPG are not necessarily correlated.
+(2) The ARM Mali-G78 MP20 exhibits the most unfair scheduling among all tested devices, with a fallback rate of 57.8\% relative to workgroups launched.
+However, the average number of spins per workgroup is only 2.52---below our fallback trigger threshold of 4 spins.
+This suggests a bimodal distribution of waiting times, where workgroups either proceed immediately with minimal delay or experience prolonged stalls that trigger fallbacks.
+
+\subsection{Simulated Blocking}%
+As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near-100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced.
 Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%; \emph{DF} is consistently faster than the previous standard for portable scans.
@@
 In true starvation, a blocking workgroup stalls before posting its reduction, preventing subsequent workgroups from making progress.
-Unlike our simulation, which allows workgroups to eventually continue, true starvation results in increased lookback lengths, as blocked workgroups cannot complete \emph{Chained Propagation} and update their flags to inclusive.
+Unlike our simulation, which allows workgroups to eventually continue, true starvation results in increased lookback lengths, as blocked workgroups cannot complete \emph{Chained Propagation} or update their flags to inclusive.
 Furthermore, starvation effectively leads to \emph{occupancy reduction}: execution resources remain available, but fewer subgroups are scheduled for execution, reducing opportunities for context switching to mitigate latency.
+The slight downward trend in Figure~\ref{fig:lookback_length} shows that as workgroups get bogged down in fallbacks, lookback lengths decrease as dependency resolution reverts to tile-to-adjacent-tile propagation of readiness information.
+This emergent behavior caused by blocking can be viewed as a computationally enforced variant of \emph{StreamScan}.
+We connect this behavior to \emph{StreamScan} because we posit that its strictly enforced adjacent-tile dependency propagation yields perfectly ordered workgroup exits, which could further enhance global write efficiency, possibly explaining the slight uptick in performance seen in Figure~\ref{fig:execution_time} on the Mali-G78.
+
 \section{Discussion}
@@
 
-Our reexamination of \emph{Chained Scan} with dynamic assignment raises fundamental questions about the GPU progress models proposed by Sorensen et al.~\cite{sorensen2021}: Can all HSA algorithms be converted to OBE via dynamic assignment?
-Are there cases, excluding atomic memory errors, where LOBE guarantees termination but dynamic assignment with OBE does not?
-Is OBE all you need?
-
 \subsection{Conclusion}
@@
 Our approach enables any workgroup to compute the reduction of any blocking tile, eliminating the risk of starvation.
-We reexamined the progress models proposed by Sorensen et al.~\cite{sorensen2021} in the context of dynamically assigned work, demonstrating that Chained Scan implementations require, at minimum, the OBE guarantee.
-Additionally, we highlighted the risks of running \emph{Decoupled Lookback} without FPG.
-Furthermore, we introduced a subgroup-accelerated yet subgroup-size-agnostic scan based on a generalized radix-$s$ Ladner-Fischer scan, along with a tile state representation that eliminates reliance on 64-bit atomics and memory barriers by distributing tile data across multiple values.
+We highlighted the risks of running \emph{Decoupled Lookback} without FPG, and introduced a subgroup-accelerated yet subgroup-size-agnostic scan based on a generalized radix-$s$ Ladner-Fischer scan, along with a tile state representation that eliminates reliance on 64-bit atomics and memory barriers by distributing tile data across multiple values.
 
-Our artifact is available in the following GitHub repository: https://anonymous.4open.science/status/Decoupled-Fallback-Paper-A33C.
-The repository contains per API requirements and building instructions.
+Our artifact is available in the following GitHub repository: https://github.com/b0nes164/Decoupled-Fallback-Paper. The repository contains per API requirements and building instructions.
+
+\begin{acks}
+We thank Elias Naur for his initial work on Scalar Fallback, Ken Russell for facilitating this colloboration, Duane Merrill for answering our questions about scan terminology, and Oliver Le for providing the initial testing on Apple devices.
+\end{acks}
 

### 2025-06-20 paper.tex
final grammar/typography check

@@
 The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}~\cite{Merrill2016}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture.
-This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$.
-\emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{speed-of-light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device.
+This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$.\ \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{speed-of-light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device.
 However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite \emph{starvation} and will never complete.
@@
 
-  \caption{The four global scan techniques described in \ref{sec:Evolution}. Dashed grey boxes indicate individual kernel launches.}
+  \caption{The four global scan techniques described in Section~\ref{sec:Evolution}. Dashed grey boxes indicate individual kernel launches.}
   \label{fig:four-techniques}
@@
 This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups.
-Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns—or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations—may experience significant performance degradation.
+Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns—or applications like radix sorting, where the scattering phase of each pass results in $\textit{radixDigits} \cdot o$ unique memory locations—may experience significant performance degradation.
 
@@
   \includegraphics[width=.95\textwidth]{graphics/DecoupledFallback.pdf}
-  \caption{\emph{Decoupled Fallback}. For brevity, the chained propagation phase is condensed, see \ref{fig:decoupled-lookback} for the full behavior.}
+  \caption{\emph{Decoupled Fallback}. For brevity, the chained propagation phase is condensed; see Figure~\ref{fig:decoupled-lookback} for the full behavior.}
   \label{fig:decoupled-fallback}
@@
   \end{tabular}
-  \caption{Hardware Specifications of Testing Devices; FPG determinations are based on the data collected by Sorensen et. al.~\cite{sorensen2021}\label{tab:hardware}}
+  \caption{Hardware Specifications of Testing Devices; FPG determinations are based on the data collected by Sorensen et al.~\cite{sorensen2021}.\label{tab:hardware}}
 \end{table*}
@@
 We hypothesize that the 7900XT and Mali-G78 have insufficient memory systems to efficiently accommodate all in-flight workgroups, making them particularly sensitive to this effect.
-To test this hypothesis, we augmented our \emph{Memcpy} kernel with a simplified lookback routine---excluding the scan itself---and observed a performance improvement of approximately 10\% on the 7900XT.
+To test this hypothesis, we augmented our \emph{Memcpy} kernel with a simplified lookback routine---excluding the scan itself---and observed a performance improvement of approximately 10\% on the 7900XT\@.
 Thus, the unexpectedly slow baseline \emph{Memcpy} performance on these devices explains why \emph{DF}, despite its strong overall performance, does not fully achieve the theoretical 50\% speedup over \emph{RTS}.
@@
   \item \textbf{Total Fallbacks Initiated:} The total number of fallback operations initiated across all workgroups.
-Ideally, this number would be zero on devices with FPG; because multiple workgroups may perform a fallback on the same blocking tile, this number can exceed the number of workgroups launched.
+Ideally, this number would be zero on devices with FPG\@; because multiple workgroups may perform a fallback on the same blocking tile, this number can exceed the number of workgroups launched.
 In practice, this count is highly dependent on the fairness of the hardware scheduler.
@@
 The current state-of-the-art algorithm, Adinets and Merrill's \emph{Onesweep}~\cite{adinets2022onesweepfastersignificantdigit}, is an 8-bit least-significant-digit radix sort, where single-pass scan minimizes data movement during digit-binning passes.
-However, due to its compute-intensive nature and large $radixDigit$-sized monoid, it is unclear whether \emph{Decoupled Fallback} will map well to sorting or if new techniques are needed.
+However, due to its compute-intensive nature and large $\textit{radixDigit}$-sized monoid, it is unclear whether \emph{Decoupled Fallback} will map well to sorting or if new techniques are needed.
 
@@
 
-Our artifact is available in the following GitHub repository: https://github.com/b0nes164/Decoupled-Fallback-Paper. The repository contains per API requirements and building instructions.
+Our artifact is available in the following GitHub repository: https://github.com/b0nes164/Decoupled-Fallback-Paper. The repository contains per-API requirements and building instructions.
 
 \begin{acks}
-We thank Elias Naur for his initial work on Scalar Fallback, Ken Russell for facilitating this colloboration, Duane Merrill for answering our questions about scan terminology, and Oliver Le for providing the initial testing on Apple devices.
+We thank Elias Naur for his initial work on Scalar Fallback, Ken Russell for facilitating this collaboration, Duane Merrill for answering our questions about scan terminology, and Oliver Le for providing the initial testing on Apple devices.
 \end{acks}

### 2025-06-21 paper.tex
More fixes!

@@
 \usepackage{makecell}
+\usepackage{stfloats}
 
@@
 Blelloch~\cite{Blelloch:1989:SAP} adapted scan to the PRAM model and popularized its use as an algorithmic primitive.
-Finally, Merrill and Garland~\cite{Merrill2016} provide a review of GPU scan implementations, complementing earlier work by Merrill and Grimshaw~\cite{Merrill2009}.
+Merrill and Garland~\cite{Merrill2016} provide a review of GPU scan implementations, complementing earlier work by Merrill and Grimshaw~\cite{Merrill2009}.
 
@@
       \includegraphics[height=6cm,keepaspectratio]{graphics/StreamScan.pdf}
-      \caption{\emph{Stream Scan.}}
+      \caption{\emph{StreamScan.}}
       \label{fig:stream-scan}
@@
 By these criteria alone, the \emph{Chained Scan} architecture emerges as the preferred solution; moreover, its single-kernel design offers compelling portability advantages across diverse hardware.
-By eliminating global barriers, \emph{Chained Scan} naturally conforms to the GPU’s oversubscription model, allowing occupancy control to be delegated the hardware's workgroup scheduler.
-In contrast, hierarchical approaches such as \emph{Reduce-then-Scan} require workgroup raking and fixed-occupancy persistent threads to bound the spine’s size, necessitating explicit kernel launches and manual tuning of launch configurations to maintain optimal occupancy.
+By eliminating global barriers, \emph{Chained Scan} naturally conforms to the GPU’s oversubscription model, allowing occupancy control to be delegated to the hardware's workgroup scheduler.
+In contrast, hierarchical approaches such as \emph{Reduce-then-Scan} require workgroup raking and fixed-occupancy persistent threads to bound the spine’s size, necessitating manual tuning of launch configurations to maintain optimal occupancy.
 
 Hardware-level considerations underscore the limitations of the fixed-launch approach.
-On GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—fixed launches can lead to performance degradation due to cache thrashing.
+On GPUs where the global memory data cache is indexed by physical addresses---and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)---fixed launches can lead to performance degradation due to cache thrashing.
 This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups.
-Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns—or applications like radix sorting, where the scattering phase of each pass results in $\textit{radixDigits} \cdot o$ unique memory locations—may experience significant performance degradation.
+Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns---or applications like radix sorting, where the scattering phase of each pass results in $\textit{radixDigits} \cdot o$ unique memory locations---may experience significant performance degradation.
 
-Moreover, the fused nature of \emph{Chained Scan} enables in-situ compaction: algorithms requiring compaction can merge that work directly into the scan kernel, avoiding redundant writes or secondary compaction passes.
+Moreover, the fused nature of \emph{Chained Scan} enables \emph{in-situ compaction}: algorithms requiring compaction can merge that work directly into the scan kernel, avoiding redundant writes or secondary compaction passes.
 This fusion benefits applications ranging from data deduplication to compression.
@@
 \subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
-\emph{Chained Scans} are \emph{one-way producer-consumer} algorithms; each workgroup consumes reductions only from lower-ID counterparts, thereby enforcing the ordering required to elide barriers.
-For such algorithms, guaranteeing execution of the lowest-ID workgroup is sufficient to ensure termination.
-If subgroups from the lowest-ID workgroup are not guaranteed execution, dependent workgroups may spin indefinitely, waiting for updates that never arrive.
-A \emph{forward-progress guarantee} ensures that all occupant subgroups---potentially from different workgroups---are eventually granted execution, thus preventing indefinite starvation.
+\emph{Chained Scans} are \emph{one-way producer-consumer} algorithms; each workgroup consumes reductions only from \emph{lower-rank}\footnote{We specifically use the term \emph{rank} here to generalize between which ID---atomically acquired or scheduler-assigned---is used to determine this ordering.
+This distinction matters because the choice of ID determines which progress model, as defined by Sorensen et al.~\cite{sorensen2021}, is required to guarantee termination.
+Working under the assumption that the scheduler-assigned ID is used, Sorensen et al.~\cite{sorensen2021, sorensen2018} classify \emph{Chained Scan} as requiring \emph{HSA} (Heterogeneous System Architecture), a model that guarantees progress for the lowest-ranked workgroup.
+However, if the atomically acquired ID determines rank, then the HSA guarantee no longer suffices, as the lowest scheduler-assigned ID may not correspond to the lowest atomic ID.
+In this case, termination requires only \emph{OBE} (Occupancy-Bound Execution), which guarantees progress once a workgroup begins execution.} counterparts, thereby enforcing the ordering required to elide barriers.
+For such algorithms, guaranteeing execution of the lowest-rank workgroup is sufficient to ensure termination.
+If subgroups from the lowest-rank workgroup are not guaranteed execution, dependent workgroups may spin indefinitely, waiting for updates that never arrive.
+A \emph{forward-progress guarantee}, in the sense required by \emph{Chained Scan}, ensures that all occupant subgroups---potentially from different workgroups---are eventually granted execution after becoming active, thus preventing indefinite starvation.
 
 The absence of FPG in key hardware segments poses a critical challenge for developers.
-Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently.
+Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Detection and Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently.
 On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds.
-Other operating systems—such as macOS and Linux—employ their own timeout mechanisms with varying behaviors.
+Other operating systems---such as macOS and Linux---employ their own timeout mechanisms with varying behaviors.
 For consistency, we adopt Windows’s 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program.
 Exacerbating this issue, no major graphics API provides a mechanism to query FPG support.
-This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware.
-In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.
+This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on their target hardware.
+In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing the advantages of \emph{Chained Scan} for broader compatibility.
 
@@
 This eliminates the possibility of starvation because the number of such operations is bounded; the wait will eventually terminate, possibly at the expense of doing redundant work.
-
-\begin{figure*}[t]
-  \centering
-  \includegraphics[width=.95\textwidth]{graphics/DecoupledFallback.pdf}
-  \caption{\emph{Decoupled Fallback}. For brevity, the chained propagation phase is condensed; see Figure~\ref{fig:decoupled-lookback} for the full behavior.}
-  \label{fig:decoupled-fallback}
-\end{figure*}
 
@@
 
+\begin{figure*}
+  \centering
+  \includegraphics[width=.95\textwidth]{graphics/DecoupledFallback.pdf}
+  \caption{\emph{Decoupled Fallback}. For brevity, the lookback is condensed; see Figure~\ref{fig:decoupled-lookback} for the full behavior, which is shared with \emph{Decoupled Lookback}.}
+  \label{fig:decoupled-fallback}
+\end{figure*}
 \section{Decoupled Fallback}
@@
 \begin{enumerate}
-  \item[(0)] \textbf {Initialization}: The first thread in the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
+  \item[(0)] \textbf {Initialization}: The first thread of the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
   \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering.
@@
 Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts.
-To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$.
+To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network---in our case, $s$.
 Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction.
@@
       \If{$\textit{thread\_id} < \textit{step}$}{
-        $\textit{temp} \gets \text{subgroupKoggeStoneScan}(\textit{x}[\textit{thread\_id} + \textit{top\_offset}])$\;
+        $\textit{temp} \gets \text{subgroupKoggeStone}(\textit{x}[\textit{thread\_id} + \textit{top\_offset}])$\;
         $\textit{x}[\textit{thread\_id} + \textit{top\_offset}] \gets \textit{temp}$\;
@@
         $\textit{cond1} \gets \textit{fanout\_index} < \textit{spine\_length}$\;
-        $\textit{cond2} \gets (\textit{fanout\_index} \,\&\, (\textit{j} - 1)) \geq \textit{reduced\_stride}$\;
+        $\textit{cond2} \gets \textit{fanout\_index} \bmod \textit{j} \geq \textit{reduced\_stride}$\;
 
@@
 Instead, we encode the tile state in the most significant bits of the atomic value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}.
-This allows an \emph{atomicMaximum}, which successfully updates to a more advanced state but prevents a state regression, to correctly perform updates.
+This allows an \emph{atomicMaximum} to successfully update to a more advanced state while preventing state regression.
 
@@
 For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum.
-Our test hardware, shown in Figure~\ref{tab:hardware}, spans a wide range of FPG and performance capabilities.
+Our test hardware, shown in Table~\ref{tab:hardware}, spans a wide range of FPG and performance capabilities.
 We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing.
@@
     Intel  & HD Graphics 620          & \textbf{Yes} & 9         & 8--32         \\
-    ARM    & Mali-G78 MP20          & \textbf{No}  & 40        & 16            \\
+    ARM    & Mali-G78 MP20 (Pixel 6a) & \textbf{No}  & 40        & 16            \\
     Apple  & M3     (10 Cores)        & \textbf{No}  & 80        & 32            \\
@@
 
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
+  \caption{Performance versus CUB (NVIDIA 2080 Super).}
+  \label{fig:cuda}
+\end{figure}
+
 \subsection{Performance Comparison}
@@
 
-Cross-API comparisons remain challenging, as our CUDA implementation consistently outperforms its Dawn counterpart despite near-identical high-level implementations.
+As can be seen in Figure \ref{fig:cuda}, cross-API comparisons remain challenging, as our CUDA implementation consistently outperforms its Dawn counterpart despite near-identical high-level implementations.
 The performance cost of \emph{DF} is most pronounced in CUDA when the entire input fits within the 2080 Super's 4~MB L2 cache, where reduced global memory latency exposes \emph{DF's} increased synchronization and compute overhead compared to CUB\@.
 However, at larger input sizes, where memory accesses exceed cache capacity, \emph{DF} reaches parity with CUB and achieves \emph{Memcpy} speeds.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
-  \caption{Performance versus CUB (NVIDIA 2080 Super).}
-\end{figure}
 
@@
 \subsubsection{Same-API Performance Versus Baseline Measures}
+\label{sec:SameVsBase}
 We next compare \emph{DF} against \emph{Reduce-then-Scan} (\emph{RTS}), which serves as a baseline to highlight the performance improvements of our approach over the prior FPG-constrained technique.
@@
 \subsection{Decoupled Fallback Statistics}
-To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution:
-
-\label{sec:Sim-Blocking}
-To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally omitting work tile reduction updates.
-Because these updates are skipped, the immediate successor workgroup reads \emph{Not Ready}, spinning and triggering a fallback.
-Starting with infrequent omissions---1 in 512 work tiles---we progressively increase the omission rate to near-continuous levels---1 in 2 work tiles---recording \emph{DF} statistics at each step.
 \begin{figure*}
@@
 
+To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution:
+
 \begin{itemize}
@@
 
-Furthermore, this data reveals two notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, demonstrating that fairness and FPG are not necessarily correlated.
+This data reveals two notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, demonstrating that fairness and FPG are not necessarily correlated.
 (2) The ARM Mali-G78 MP20 exhibits the most unfair scheduling among all tested devices, with a fallback rate of 57.8\% relative to workgroups launched.
@@
 \subsection{Simulated Blocking}%
+\label{sec:Sim-Blocking}
+To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally omitting work tile reduction updates.
+Because these updates are skipped, the immediate successor workgroup reads \emph{Not Ready}, spinning and triggering a fallback.
+Starting with infrequent omissions---1 in 512 work tiles---we progressively increase the omission rate to near-continuous levels---1 in 2 work tiles---recording \emph{DF} statistics at each step.
+
 As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near-100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced.
 Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%; \emph{DF} is consistently faster than the previous standard for portable scans.
-While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that fallbacks are being induced, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior.
-This is because workgroups are still able to proceed after an omission.
-In true starvation, a blocking workgroup stalls before posting its reduction, preventing subsequent workgroups from making progress.
-Unlike our simulation, which allows workgroups to eventually continue, true starvation results in increased lookback lengths, as blocked workgroups cannot complete \emph{Chained Propagation} or update their flags to inclusive.
-Furthermore, starvation effectively leads to \emph{occupancy reduction}: execution resources remain available, but fewer subgroups are scheduled for execution, reducing opportunities for context switching to mitigate latency.
-The slight downward trend in Figure~\ref{fig:lookback_length} shows that as workgroups get bogged down in fallbacks, lookback lengths decrease as dependency resolution reverts to tile-to-adjacent-tile propagation of readiness information.
-This emergent behavior caused by blocking can be viewed as a computationally enforced variant of \emph{StreamScan}.
-We connect this behavior to \emph{StreamScan} because we posit that its strictly enforced adjacent-tile dependency propagation yields perfectly ordered workgroup exits, which could further enhance global write efficiency, possibly explaining the slight uptick in performance seen in Figure~\ref{fig:execution_time} on the Mali-G78.
+While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that fallbacks are being induced, this test is limited in that it only measures the cost of increased fallback operations rather than truly simulating blocking behavior.
+In true starvation, a blocking workgroup stalls before posting its reduction and cannot continue until it receives execution from the scheduler.
+By contrast, our simulation allows workgroups to proceed and eventually post their inclusive reductions to the spine.
+Thus, we would expect true starvation to result in increased lookback lengths---unlike the slight downward trend observed in our simulation.
+Lastly, starvation leads to effective \emph{occupancy reduction}, something we cannot simulate dynamically: execution resources remain available, but fewer subgroups are scheduled, reducing opportunities for context switching to mitigate latency.
+
+The slight downward trend in lookback lengths observed in Figure~\ref{fig:lookback_length} reflects the shift in dependency resolution strategy as workgroups become increasingly bogged down by fallbacks.
+Rather than resolving dependencies through lookback traversals, workgroups revert to tile-to-adjacent-tile propagation of readiness information.
+This emergent behavior, caused by blocking, can be viewed as a computationally enforced variant of \emph{StreamScan}.
+We connect this behavior to \emph{StreamScan} because we posit that its strictly enforced adjacent-tile dependency propagation yields perfectly ordered workgroup exits, which could further enhance the global write efficiency behavior observed in Section~\ref{sec:SameVsBase}, possibly explaining the slight uptick in performance seen in Figure~\ref{fig:execution_time} on the Mali-G78.
 
@@
 \begin{acks}
-We thank Elias Naur for his initial work on Scalar Fallback, Ken Russell for facilitating this collaboration, Duane Merrill for answering our questions about scan terminology, and Oliver Le for providing the initial testing on Apple devices.
+We thank Elias Naur for his initial work on \emph{Scalar Fallback}, Ken Russell for facilitating this collaboration, Duane Merrill for answering our questions about scan terminology, and Oliver Le for providing the initial testing on Apple devices.
 \end{acks}

### 2025-06-21 paper.tex
Fix bad mali figures

@@
 \subsubsection{Chained Scan, Decoupled Lookback}
-Introduced by Merrill and Garland, the \emph{Decoupled Lookback} technique refines \emph{Chained Propagation} by separating reduction communication from dependency resolution, mitigating bottlenecks caused by propagation latency.
+Introduced by Merrill and Garland~\cite{Merrill2016}, the \emph{Decoupled Lookback} technique refines \emph{Chained Propagation} by separating reduction communication from dependency resolution, mitigating bottlenecks caused by propagation latency.
 In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction.
@@
 \subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
-\emph{Chained Scans} are \emph{one-way producer-consumer} algorithms; each workgroup consumes reductions only from \emph{lower-rank}\footnote{We specifically use the term \emph{rank} here to generalize between which ID---atomically acquired or scheduler-assigned---is used to determine this ordering.
+\emph{Chained Scans} are \emph{one-way producer-consumer} algorithms; each workgroup consumes reductions only from lower-\emph{rank} counterparts, thereby enforcing the ordering required to elide barriers.
+For such algorithms, guaranteeing execution of the lowest-rank workgroup is sufficient to ensure termination.
+If subgroups from the lowest-rank workgroup are not guaranteed execution, dependent workgroups may spin indefinitely, waiting for updates that never arrive.
+A \emph{forward-progress guarantee}, in the sense required by \emph{Chained Scan}, ensures that all occupant subgroups---from potentially different workgroups---are eventually granted execution after becoming active, thus preventing indefinite starvation.\footnote{We specifically use the term \emph{rank} here to generalize which ID---atomically acquired or scheduler-assigned---is used to determine this ordering.
 This distinction matters because the choice of ID determines which progress model, as defined by Sorensen et al.~\cite{sorensen2021}, is required to guarantee termination.
 Working under the assumption that the scheduler-assigned ID is used, Sorensen et al.~\cite{sorensen2021, sorensen2018} classify \emph{Chained Scan} as requiring \emph{HSA} (Heterogeneous System Architecture), a model that guarantees progress for the lowest-ranked workgroup.
-However, if the atomically acquired ID determines rank, then the HSA guarantee no longer suffices, as the lowest scheduler-assigned ID may not correspond to the lowest atomic ID.
-In this case, termination requires only \emph{OBE} (Occupancy-Bound Execution), which guarantees progress once a workgroup begins execution.} counterparts, thereby enforcing the ordering required to elide barriers.
-For such algorithms, guaranteeing execution of the lowest-rank workgroup is sufficient to ensure termination.
-If subgroups from the lowest-rank workgroup are not guaranteed execution, dependent workgroups may spin indefinitely, waiting for updates that never arrive.
-A \emph{forward-progress guarantee}, in the sense required by \emph{Chained Scan}, ensures that all occupant subgroups---potentially from different workgroups---are eventually granted execution after becoming active, thus preventing indefinite starvation.
+However, if the atomically acquired ID determines rank---as used by \emph{Decoupled Lookback}~\cite[Section 4.4]{Merrill2016}, \emph{StreamScan}~\cite[Section 3.3.4]{10.1145/2442516.2442539}, and this work---then the HSA guarantee no longer suffices, as the lowest scheduler-assigned ID may not correspond to the lowest atomic ID.
+In this case, termination requires \emph{OBE} (Occupancy-Bound Execution), which guarantees progress once a workgroup begins execution.}
 
@@
   \includegraphics[width=.95\textwidth]{graphics/DecoupledFallback.pdf}
-  \caption{\emph{Decoupled Fallback}. For brevity, the lookback is condensed; see Figure~\ref{fig:decoupled-lookback} for the full behavior, which is shared with \emph{Decoupled Lookback}.}
+  \caption{\emph{Decoupled Fallback}. For brevity, the lookback is condensed; see Figure~\ref{fig:decoupled-lookback} for full lookback behavior.}
   \label{fig:decoupled-fallback}
@@
     \midrule
-    Avg \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)      & 1.48    & 2.50     & 10.82   & 37.42    & 50.95    & 73.68    \\
-    Avg \emph{DF} (32-bit ele/sec $\cdot 10^9$)          & 1.51    & 2.65     & 10.87   & 36.85    & 51.25    & 83.27    \\
-    Avg \emph{RTS} (32-bit ele/sec $\cdot 10^9$)         & 1.06    & 1.94     & 7.46    & 25.75    & 34.36    & 62.57    \\
-    \emph{DF}/\emph{Memcpy}                              & 102.0\% & 106.0\%  & 100.5\% & 98.4\%   & 100.6\%  & 113.0\%  \\
-    \emph{DF}/\emph{RTS}                                 & 142.6\% & 136.6\%  & 145.7\% & 143.1\%  & 149.1\%  & 133.1\%  \\
-    Avg Spins per workgroup                              & 0.420   & 2.52     & 0.719   & 0.752    & 0.679    & 0.533    \\
-    Avg Lookback Length per workgroup                    & 1.714   & 3.52     & 4.598   & 10.737   & 12.350   & 22.648   \\
-    Avg Total Fallbacks Initiated                        & 446.183 & 4739.91  & 449.961 & 1046.216 & 333.324  & 266.641  \\
-    Avg Total Successful Insertions                      & 21.475  & 571.24   & 23.887  & 4.822    & 11.378   & 46.288   \\
+    Avg \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$) & 1.481   & 2.626   & 10.82   & 37.42    & 50.95   & 73.68   \\
+    Avg \emph{DF} (32-bit ele/sec $\cdot 10^9$)     & 1.511   & 2.725   & 10.87   & 36.85    & 51.25   & 83.27   \\
+    Avg \emph{RTS} (32-bit ele/sec $\cdot 10^9$)    & 1.06    & 2.025   & 7.46    & 25.75    & 34.36   & 62.57   \\
+    \emph{DF}/\emph{Memcpy}                         & 102.0\% & 103.8\% & 100.5\% & 98.4\%   & 100.6\% & 113.0\% \\
+    \emph{DF}/\emph{RTS}                            & 142.6\% & 134.6\% & 145.7\% & 143.1\%  & 149.1\% & 133.1\% \\
+    Avg Spins per workgroup                         & 0.42    & 2.504   & 0.719   & 0.752    & 0.679   & 0.533   \\
+    Avg Lookback Length per workgroup               & 1.714   & 3.519   & 4.598   & 10.737   & 12.35   & 22.648  \\
+    Avg Total Fallbacks Initiated                   & 446.183 & 4702.01 & 449.961 & 1046.216 & 333.324 & 266.641 \\
+    Avg Total Successful Insertions                 & 21.475  & 560.15  & 23.887  & 4.822    & 11.378  & 46.288  \\
     \bottomrule

### 2025-06-21 paper.tex
Analysis fixes

@@
 
-As shown in Table~\ref{tab:results}, \emph{DF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices, closely approaching the theoretically expected $O(3n)/O(2n)$ 50\% speedup, except on the 7900XT and Mali-G78.
-This result initially appears confusing, given that \emph{DF} outperforms \emph{Memcpy} by 6\% and 13\%, respectively.
-However, we hypothesized and confirmed a secondary effect of \emph{Chained Scans}: they enforce a queue-like behavior among workgroups.
-In-flight workgroups with a lower relative rank traverse fewer tiles during their lookbacks, enabling them to exit earlier than higher-ranked workgroups.
-Because this ordering emerges naturally from differing amounts of computation rather than explicit synchronization, the exit sequence is \emph{k-sorted} or \emph{nearly ordered}, rather than strictly ordered.
-Scheduling fairness and message-passing visibility can further influence this sequence---for example, an unfair scheduler might allocate more resources to a higher-ranked workgroup, causing it to exit before a lower-ranked counterpart, or a higher-ranked workgroup might observe an updated flag sooner, enabling earlier completion.
-We confirmed this behavior experimentally by using atomics to record and print the order in which workgroups completed their lookback phase.
-Unlike a simple \emph{Memcpy} kernel, where all in-flight workgroups begin writing immediately after completing reads, this “ordered exit” pattern facilitates more efficient global memory accesses during the write phase of the scan, as adjacent workgroups are more likely to share memory pages, reducing premature page evictions.
-We hypothesize that the 7900XT and Mali-G78 have insufficient memory systems to efficiently accommodate all in-flight workgroups, making them particularly sensitive to this effect.
-To test this hypothesis, we augmented our \emph{Memcpy} kernel with a simplified lookback routine---excluding the scan itself---and observed a performance improvement of approximately 10\% on the 7900XT\@.
-Thus, the unexpectedly slow baseline \emph{Memcpy} performance on these devices explains why \emph{DF}, despite its strong overall performance, does not fully achieve the theoretical 50\% speedup over \emph{RTS}.
+As shown in Table~\ref{tab:results}, \emph{DF} achieves at least a $\sim$30\% speedup over \emph{RTS} across all tested devices, closely approaching the theoretically expected $O(3n)/O(2n)$ 50\% speedup on most devices.
+Most strikingly, \emph{DF} slightly exceeds the performance of \emph{Memcpy} on the 7900XT and Mali-G78, despite performing additional computation.
+While this result appears counterintuitive—since we expect \emph{Memcpy} to be the upper bound for memory-bound throughput—we attribute this behavior to a secondary effect of \emph{Chained Scans}: they impose a queue-like structure on workgroup progression.
+Despite entering the chained propagation phase at approximately the same time—subject to scheduler fairness—in-flight workgroups with lower relative rank traverse fewer tiles and therefore exit the phase earlier than higher-ranked workgroups.
+Because this ordering is computationally rather than logically enforced, the resulting sequence of completions is \emph{k-sorted} or \emph{nearly ordered}, rather than \emph{perfectly ordered}.
+This emergent ordering is further shaped by scheduler fairness and message-passing latency: for instance, an unfair scheduler might accelerate a higher-ranked workgroup’s completion, or message passing timing might allow a higher-ranked workgroup to observe readiness earlier than its lower-ranked peers.
+
+We confirmed this queue-like progression experimentally by using atomics to record and print the order in which workgroups exited their lookback phase.
+This 'ordered exit' behavior is notably different from that of a \emph{Memcpy} kernel, where all in-flight workgroups typically begin writing immediately after completing reads, often leading to high-concurrency, uncoordinated global writes.
+In contrast, \emph{DF}'s staggered exit pattern implicitly throttles the number of concurrent writers, reducing memory pressure and improving locality.
+Workgroups that exit in near-order are more likely to access adjacent memory regions, reducing page thrashing and avoiding premature eviction of active pages.
+On devices whose memory systems may be insufficient to support highly concurrent writes---such as the 7900XT and Mali-G78---these effects plausibly contribute to scan throughput that exceeds \emph{Memcpy}, despite the added arithmetic.
+To test this hypothesis, we augmented our \emph{Memcpy} kernel with a simplified lookback routine---excluding the scan itself---and observed a $\sim$10\% throughput increase on the 7900XT.
+These findings suggest that \emph{Chained Scan}'s computational structure not only enables progress without explicit synchronization, but also improves global memory efficiency by shaping the temporal pattern of writes, effectively translating computational work into emergent memory coordination.
 
@@
 To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution:
-
 \begin{itemize}
@@
 
-This data reveals two notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, demonstrating that fairness and FPG are not necessarily correlated.
+This data reveals three notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, demonstrating that fairness and FPG are not necessarily correlated.
 (2) The ARM Mali-G78 MP20 exhibits the most unfair scheduling among all tested devices, with a fallback rate of 57.8\% relative to workgroups launched.
 However, the average number of spins per workgroup is only 2.52---below our fallback trigger threshold of 4 spins.
-This suggests a bimodal distribution of waiting times, where workgroups either proceed immediately with minimal delay or experience prolonged stalls that trigger fallbacks.
+This suggests a skewed distribution of waiting times, where workgroups either proceed with minimal delay or experience prolonged stalls that trigger fallbacks.
+(3) Despite this organically unfair scheduling, \emph{Decoupled Fallback} still achieves a 34.6\% speedup over \emph{RTS}, confirming its robustness to scheduler behavior in real-world conditions.
 
@@
 This emergent behavior, caused by blocking, can be viewed as a computationally enforced variant of \emph{StreamScan}.
-We connect this behavior to \emph{StreamScan} because we posit that its strictly enforced adjacent-tile dependency propagation yields perfectly ordered workgroup exits, which could further enhance the global write efficiency behavior observed in Section~\ref{sec:SameVsBase}, possibly explaining the slight uptick in performance seen in Figure~\ref{fig:execution_time} on the Mali-G78.
+We connect this behavior to \emph{StreamScan} because we posit that its strictly enforced adjacent-tile dependency propagation yields \emph{perfectly ordered} workgroup exits, which could further enhance the global write efficiency behavior observed in Section~\ref{sec:SameVsBase}, possibly explaining the slight uptick in performance seen in Figure~\ref{fig:execution_time} on the Mali-G78.
 

### 2025-06-21 paper.tex
Small fixes

@@
 Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation.
-On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WebGPU native and approaches the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach.
+On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum using the native \emph{Dawn} implementation of the WebGPU standard, and approaches the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach.
 We further demonstrate the resilience of \emph{Decoupled Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
@@
 Our first attempt at implementing a \emph{Decoupled Lookback} scan without relying on FPG was \emph{Scalar Fallback}.
-Each time a workgroup spins waiting on the reduction from a preceding tile, it advances a \emph{fallback} operation by processing one scalar element from that tile.
+A work-stealing method, each time a workgroup spins waiting on the reduction from a preceding tile, it advances a \emph{fallback} operation by processing one scalar element from that tile.
 This eliminates the possibility of starvation because the number of such operations is bounded; the wait will eventually terminate, possibly at the expense of doing redundant work.
@@
 We confirmed this queue-like progression experimentally by using atomics to record and print the order in which workgroups exited their lookback phase.
-This 'ordered exit' behavior is notably different from that of a \emph{Memcpy} kernel, where all in-flight workgroups typically begin writing immediately after completing reads, often leading to high-concurrency, uncoordinated global writes.
+This 'ordered exit' behavior differs from that of a \emph{Memcpy} kernel, where all in-flight workgroups typically begin writing immediately after completing reads, often leading to high-concurrency, uncoordinated global writes.
 In contrast, \emph{DF}'s staggered exit pattern implicitly throttles the number of concurrent writers, reducing memory pressure and improving locality.
@@
 In practice, this count is highly dependent on the fairness of the hardware scheduler.
-On most devices tested, initiated fallbacks remain modest---about 10\% or fewer of the total workgroups launched.
+On most devices tested, initiated fallbacks remain modest---typically affecting fewer than 10\% of launched workgroups
 
@@
 
-As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near-100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced.
-Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%; \emph{DF} is consistently faster than the previous standard for portable scans.
-While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that fallbacks are being induced, this test is limited in that it only measures the cost of increased fallback operations rather than truly simulating blocking behavior.
-In true starvation, a blocking workgroup stalls before posting its reduction and cannot continue until it receives execution from the scheduler.
-By contrast, our simulation allows workgroups to proceed and eventually post their inclusive reductions to the spine.
-Thus, we would expect true starvation to result in increased lookback lengths---unlike the slight downward trend observed in our simulation.
-Lastly, starvation leads to effective \emph{occupancy reduction}, something we cannot simulate dynamically: execution resources remain available, but fewer subgroups are scheduled, reducing opportunities for context switching to mitigate latency.
+As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near-100\% fallback rates across all devices—except for the Intel HD620, where only a 25\% fallback rate is induced.
+Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%; \emph{DF} consistently outperforms the previous standard for portable scans.
+While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that fallbacks are being induced, this test is limited: it measures the cost of increased fallback operations, but does not fully replicate true blocking behavior.
+In actual starvation, a blocking workgroup stalls before posting its reduction and cannot proceed until it is granted execution by the scheduler.
+By contrast, our simulation allows all workgroups to eventually progress and post their inclusive reductions to the spine.
+Thus, true starvation would be expected to increase lookback lengths—contrary to the slight downward trend observed in our simulation.
+Finally, starvation leads to effective \emph{occupancy reduction}, which we cannot dynamically simulate: execution resources remain available, but fewer subgroups are scheduled, reducing opportunities for context switching to mitigate latency.
 

### 2025-06-21 paper.tex
Final^2 changes

@@
 \subsection{Decoupled Fallback Statistics}
+To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution:
+\begin{itemize}
+  \item \textbf{Spins per Workgroup:} How many times does a workgroup spin while waiting?
+With perfectly fair scheduling, no waiting is necessary.
+Since no real-world scheduler is perfectly fair, we want minimal waiting.
+Despite varying occupancy rates, the average spin rate across tested devices remains low (typically one or fewer), indicating that most workgroups experience minimal delays.
+
+  \item \textbf{Lookback Length per Workgroup:} How many work tiles does a workgroup traverse to complete \emph{Chained Propagation}?
+Ideally, this traversal should be minimal, but longer traversal may result from unfair scheduling.
+Empirically, however, our data shows that lookback length remains well below the theoretical $O(o)$ bound---typically under 10\% of occupancy---meaning that most workgroups complete \emph{Chained Propagation} in relatively few steps compared to occupancy.
+
+  \item \textbf{Total Fallbacks Initiated:} The total number of fallback operations initiated across all workgroups.
+Ideally, this number would be zero on devices with FPG\@; because multiple workgroups may perform a fallback on the same blocking tile, this number can exceed the number of workgroups launched.
+In practice, this count is highly dependent on the fairness of the hardware scheduler.
+On most devices tested, initiated fallbacks remain modest---typically affecting fewer than 10\% of launched workgroups
+
+  \item \textbf{Total Successful Insertions:} The total number of fallback operations that successfully update a blocking (or late) tile to \emph{Ready}.
+Due to false-positive fallbacks, this count does not precisely represent the number of true starvations, but instead provides an upper bound.
+Ideally, this number would also be zero on devices with FPG, but in practice, it is again dependent on the fairness of the scheduler.
+Unlike the number of fallback initiations, this statistic is extremely low---typically well under 1\% even on devices without FPG---suggesting that true starvation is rare but catastrophic.
+\end{itemize}
+This data reveals three notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, demonstrating that fairness and FPG are not necessarily correlated.
+(2) The ARM Mali-G78 MP20 exhibits the most unfair scheduling among all tested devices, with a fallback rate of 57.8\% relative to workgroups launched.
+However, the average number of spins per workgroup is only 2.52---below our fallback trigger threshold of 4 spins.
+This suggests a skewed distribution of waiting times, where workgroups either proceed with minimal delay or experience prolonged stalls that trigger fallbacks.
+(3) Despite this organically unfair scheduling, \emph{Decoupled Fallback} still achieves a 34.6\% speedup over \emph{RTS}, confirming its robustness to scheduler behavior in real-world conditions.
+
+\subsection{Simulated Blocking}%
+\label{sec:Sim-Blocking}
+To further demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally omitting work tile reduction updates.
+Because these updates are skipped, the immediate successor workgroup reads \emph{Not Ready}, spinning and triggering a fallback.
+Starting with infrequent omissions---1 in 512 work tiles---we progressively increase the omission rate to near-continuous levels---1 in 2 work tiles---recording \emph{DF} statistics at each step.
+
 \begin{figure*}
@@
 \end{figure*}
-
-To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution:
-\begin{itemize}
-  \item \textbf{Spins per Workgroup:} How many times does a workgroup spin while waiting?
-With perfectly fair scheduling, no waiting is necessary.
-Since no real-world scheduler is perfectly fair, we want minimal waiting.
-Despite varying occupancy rates, the average spin rate across tested devices remains low (typically one or fewer), indicating that most workgroups experience minimal delays.
-
-  \item \textbf{Lookback Length per Workgroup:} How many work tiles does a workgroup traverse to complete \emph{Chained Propagation}?
-Ideally, this traversal should be minimal, but longer traversal may result from unfair scheduling.
-Empirically, however, our data shows that lookback length remains well below the theoretical $O(o)$ bound---typically under 10\% of occupancy---meaning that most workgroups complete \emph{Chained Propagation} in relatively few steps compared to occupancy.
-
-  \item \textbf{Total Fallbacks Initiated:} The total number of fallback operations initiated across all workgroups.
-Ideally, this number would be zero on devices with FPG\@; because multiple workgroups may perform a fallback on the same blocking tile, this number can exceed the number of workgroups launched.
-In practice, this count is highly dependent on the fairness of the hardware scheduler.
-On most devices tested, initiated fallbacks remain modest---typically affecting fewer than 10\% of launched workgroups
-
-  \item \textbf{Total Successful Insertions:} The total number of fallback operations that successfully update a blocking (or late) tile to \emph{Ready}.
-Due to false-positive fallbacks, this count does not precisely represent the number of true starvations, but instead provides an upper bound.
-Ideally, this number would also be zero on devices with FPG, but in practice, it is again dependent on the fairness of the scheduler.
-Unlike the number of fallback initiations, this statistic is extremely low---typically well under 1\% even on devices without FPG---suggesting that true starvation is rare but catastrophic.
-\end{itemize}
-
-This data reveals three notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, demonstrating that fairness and FPG are not necessarily correlated.
-(2) The ARM Mali-G78 MP20 exhibits the most unfair scheduling among all tested devices, with a fallback rate of 57.8\% relative to workgroups launched.
-However, the average number of spins per workgroup is only 2.52---below our fallback trigger threshold of 4 spins.
-This suggests a skewed distribution of waiting times, where workgroups either proceed with minimal delay or experience prolonged stalls that trigger fallbacks.
-(3) Despite this organically unfair scheduling, \emph{Decoupled Fallback} still achieves a 34.6\% speedup over \emph{RTS}, confirming its robustness to scheduler behavior in real-world conditions.
-
-\subsection{Simulated Blocking}%
-\label{sec:Sim-Blocking}
-To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally omitting work tile reduction updates.
-Because these updates are skipped, the immediate successor workgroup reads \emph{Not Ready}, spinning and triggering a fallback.
-Starting with infrequent omissions---1 in 512 work tiles---we progressively increase the omission rate to near-continuous levels---1 in 2 work tiles---recording \emph{DF} statistics at each step.
 
